### 2023-12-08 Paper.Rmd
First draft

---
title: "Технология использования открытых данных Федеральной налоговой службы России для экономико-географического анализа"
output: odt_document
bibliography: References.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Аннотация:
Ключевые слова:
A Method of Using Open Data Provided by Federal Tax Service of Russia in Spatial Economy Analysis
Abstract:
Keywords:

# Введение

Современные общественно-географические исследования часто опираются на работу с данными. Некоторые специалисты используют на агрегированную экономическую и демографическую статистику. Так, О. Л. Виноградова, анализируя изменение интенсивности земледелия в Европе, опирается на показатели стран в разные годы [@Виноградова2020]; Т. А. Балина с соавторам, рассматривая воспрос о территориях опережающего социально-экономического развития, приводит подсчёты по субъектам Российской Федерации [@Балина2021]; Г. М. Лаппо, описывая свою концепцию разнообразия городов, опирается на сведения о числе жителей в населённых пунктах в разные периоды времени [@Лаппо2019]. Другие исследователи используют большие массивы дезагрегированных данных, собирая их вручную в общедоступных источниках и используя сложные методы их обработки. Например, А. И. и М. В. Красновы для своей работы, посвящённой трансформации сети расселения в Псковской области, собирали публичные сведения о местоположении пользователей социальной сети «ВКонтакте» и открытые данные о поисковых запросах к «Яндексу» [@Красновы2020]. Есть работы, опирающиеся на обобщённую статистику, но использующие продвинутые методы её анализа. В частности, А. В. Федорченко взял данные Международной организации автопроизводителей, рассчитал на их основе индексы и дополнил их картографической визуализацией, что позволило эффективно проанализировать территориальные сдвиги в мировой автомобильной промышленности [@Федорченко2017]. Таким образом, наблюдается общая тенденция: в том или ином виде используются данные, хотя объём, тип, качество этих данных и технологии их обработки отличаются. «Большие данные» и методы машинного обучения также проникают в общественно-географическую науку, что отметили ещё десять лет назад М. Грэм и Т. Шелтон [@Graham2013]. Недавний пример такого рода исследований — работа Вэй Лю и Вэньцзэ У, которые используют сведения о пользователях социальной сети WeChat для изучения связей между типом землепользования и деятельностью людей [@Liu2020].

Необходимость использования данных закономерно ведёт к вопросу об источниках их получения, поскольку прежде чем обрабатывать данные для исследований и делать выводы из них, необходимо эти данные где-то взять: найти в готовом или пригодном для обработки виде или собрать самостоятельно. Поиск качественных данных и их предварительная обработка во многом определяют результативность научной деятельности: если данных нет, они некачественные или подготовлены неправильно, то исследование с высокой вероятностью окажется неудачным или приведёт к заблуждениям. Используемые источники данных и технологии их обработки, таким образом, являются важным прикладным аспектом большинства экономико-географических исследований. Практическая важность данных для исследований подтверждается, в частности, различными инициативами, связанными с обеспечением доступа к таким данным. Среди этих инициатив есть как государственные, так и частные.

В российских реалиях наиболее типичным и массово используемым источником данных для экономико-географических исследований являются ресурсы Росстата. Они включают в себя официальные статистические публикации, статистические данные, размещённые на официальном сайте, базу данных показателей муниципальных образований (БДМО) и веб-портал единой межведомственной информационно-статистической системы (ЕМИСС). Причины такой популярности Росстата понятны. Во-первых, Росстат — это государственный орган, уполномоченный на сбор информации от хозяйствующих субъектов, в том числе конфиденциальной, имеющий необходимые права, обязанности и гарантии, несущий ответственность за своевременность, полноту и качество предоставляемых и распространяемых сведений. Во-вторых, деятельность Росстата и его предшественников охватывает всю территорию страны, все отрасли экономики, все виды организаций и значительный временной промежуток, поэтому по количеству собранной информации он превосходит иные источники, а в каких-то вопросах является единственным прямым источником информации. В-третьих, ресурсы Росстата широко известны, относительно просты для использования (даже если эта простота порой кажущаяся), в основном доступны без регистрации и платы.

Тем не менее, есть причины, которые могут побудить к поиску и использованию других источников данных. К числу общих недостатков Росстата как источника данных можно отнести запутанность и неочевидность интерфейсов, описаний, каталогов и классификаций, частые и взаимно несовместимые изменения схем и форматов публикации сведений, низкую пригодность данных к автоматизированной обработке, и кроме того, порой обоснованные сомнения в достоверности информации. Если же говорить конкретно о географических исследованиях, особенно в сфере экономической географии, то ключевым недостатком является низкая детализация данных по разным параметрам. Так, официальные статистические показатели обычно представлены в разрезе регионов (субъектов федерации). Даже муниципальный уровень доступен не всегда: обычно по нему есть только некоторые показатели или наиболее новые данные. Более точная географическая привязка отсутствует, и это порой может становиться проблемой. В частности, в геоинформационных системах типично представление с использованием географических координат, но официальная статистическая информация не содержит координат, в результате чего любое картографирование и визуализация таких сведений неизбежно сводится к уровню субъектов федерации или муниципальных образований, что снижает пространственную детализацию и потенциально может скрывать некоторые закономерности. Временное разрешение официальных статистических данных — год, хотя в некоторых случаях этого недостаточно. Классификация по сферам экономики также довольно обширна: отрасли либо не дифференцированы вовсе, либо сведены к широким группам классификатора видов экономической деятельности, что, конечно, достаточно для изучения традиционных отраслей, но мало при фокусировании внимания на относительно новых или сильно дифференцированных сфер деятельности, например, многих видов услуг. Причины такого положения дел отчасти понятны: во-первых, низкая детализация снижает риски раскрытия первичных статистических сведений, которые по закону конфиденциальны, а во-вторых, потребителей у информации Росстата много, и вряд ли он будет фокусироваться на удовлетворении довольно специфических запросов некоторых специалистов по экономической географии. Тем не менее, в результате актуален вопрос о возможностях и технологиях использования иных источников данных для экономико-географического анализа.

В статье представлена технология, позволяющая использовать некоторые наборы открытых данных Федеральной налоговой службы в качестве источника информации об организациях и индивидуальных предпринимателях, осуществляющих деятельность в том или ином месте в той или иной сфере экономики, а также о доходах, расходах и численности работников этих организаций. Несмотря на свою относительную концептуальную простоту, технология является достаточно универсальной и потенциально полезной, особенно учитывая нехватку наборов данных для исследований, особенно в машиночитаемом виде. Она позволяет получить данные за период с середины 2016 года с детализацией по месяцам, отдельным населённым пунктам (городским и сельским) и конкретным кодам классификатора видов экономической деятельности. Кроме того, технология обеспечивает детализацию до конкретных физических лиц и организаций, что открывает возможности для углублённого анализа (например, миграции отдельных компаний и сопутствующая динамика). Полнота сведений отличается в зависимости от цели использования:
а) применительно к малому и среднему предпринимательству технология позволяет получить полный список соответствующих организаций и индивидуальных предпринимателей;
б) применительно к экономике в целом технология позволяет получить выборку организаций, которая является репрезентативной в пространственном аспекте и в аспекте классификации по видам экономической деятельности.

В следующем разделе, «Данные и методы», приводится описание источников «сырых» данных и методов их обработки в общем виде, то есть на концептуальном уровне формулируется алгоритм, составляющий суть технологии. Кроме того, даётся краткое описание авторской реализации данной технологии и  инструментов, использованных для этой реализации, а также ссылка на исходный код компьютерной программы. Раздел «Результаты» описывает результаты пробного использования технологии для получения информации о компаниях и физических лицах, осуществляющих деятельность в сфере юридических услуг, а также результаты валидации технологии с использованием статистики Росстата. В разделе «Обсуждение» анализируются возможные варианты и ограничения использования технологии, а также некоторые аспекты её практической реализации. «Выводы» подводят итоги статьи.

# Данные и методы

Исходные данные
В качестве исходных данных, которые обрабатываются алгоритмом, используются три набора открытых данных Федеральной налоговой службы (ФНС) России.
Первый, основной — это набор «Единый реестр субъектов малого и среднего предпринимательства». Технически он представляет собой коллекцию zip-архивов xml-файлов. Zip-архив представляет собой «срез» (выгрузку) из реестра по состоянию на определённую дату. Одна дата представлена ровно одним архивом. В каждом архиве находится множество (несколько тысяч) xml-файлов. Каждый xml-файл, в свою очередь, содержит в себе информацию о множестве (обычно ровно 900) организациях или индивидуальных предпринимателях, включённых в реестр. Одна организация или один индивидуальный предприниматель в xml-файле представлен элементом «Документ». Отдельные характеристики организации или предпринимателя (название, ИНН, коды ОКВЭД, местонахождение и другие) представлены как вложенные элементы и их атрибуты. Общая схема указанной структуры данных представлена на рисунке.
Указанный набор открытых данных доступен за период с августа 2016 г. по настоящее время. Новые выгрузки публикуются ежемесячно. Таким образом, на момент написания статьи доступно около 85 выгрузок. Каждая выгрузка (архив) имеет объём от 2 до 3 Гб в сжатом виде и от 30 до 50 Гб в распакованном. Общий объём исходных данных этого набора в несжатом виде, таким образом, оценивается в несколько терабайт (тысяч гигабайт), что приводит к необходимости использования специальных методов оптимизации и конкретных технологических решений, адаптированных к эффективной работе с «большими данными», о которых будет сказано ниже.
Второй, вспомогательный набор — это «Сведения о суммах доходов и расходов по данным бухгалтерской (финансовой) отчетности организации». Технически он также представляет собой коллекцию zip-архивов xml-файлов. Каждый архив представляет собой выгрузку на определённую дату, содержит несколько тысяч xml-файлов, каждый из которых, в свою очередь, содержит информацию о нескольких сотнях организаций. Данные обновляются с разной периодичностью, но фактическая частота обновления — ежегодно: если в течение года публикуется несколько выгрузок, то все они содержат данные за предыдущий год и отличаются лишь в небольшой части записей, что, вероятно, представляет собой исправление ошибок в ранее размещённых выгрузках. В момент подготовки статьи были доступны выгрузки с 2019 г. В текущий момент выгрузки за предыдущие даты исчезли с сайта ФНС России, и причины этого непонятны.
Третий набор, также вспомогательный — «Сведения о среднесписочной численности работников». Его структура и периодичность обновления аналогичны предыдущему. В момент подготовки статьи были доступны выгрузки с 2019 г. В текущий момент выгрузки за предыдущие даты исчезли с сайта ФНС России, и причины этого непонятны.
В каждом из трёх наборов указан ИНН организации или индивидуального предпринимателя. Поскольку ИНН является уникальным идентификатором, то он позволяет связать сведения из трёх наборов данных.
Открытые данные по определению являются официальными и свободно распространяемыми, что позволяет рассматривать их как достаточно надёжный и удобный для использования источник информации. Тем не менее, количество данных и формат их распространения порождают необходимость использовать специальные методы обработки, о которых говорится далее.
Общий алгоритм обработки данных
Шаг 1. Загрузка данных из источника на локальный компьютер или в облачное хранилище. Смысл этапа — получить собственную копию исходных данных, которая будет обрабатываться. Этот шаг выполняет две функции: во-первых, ускоряет дальнейшую обработку за счёт того, что исходные архивы уже загружены на устройство, где они будут обрабатываться, или в облачное хранилище, где передача данных между сервисом хранения и сервисом вычисления происходит быстро; во-вторых, страхует от внезапного исчезновения исходных архивов и позволяет проводить обработку данных даже в том случае, если они будут сняты с публикации, как случилось с двумя вспомогательными наборами. Под облачным хранилищем может пониматься как пользовательский сервис хранения данных наподобие «Яндекс-диска», так и специализированное объектное хранилище с S3-совместимым интерфейсом наподобие Object Storage Яндекс-облака.
Шаг 2. Фильтрующее извлечение данных. Смысл этапа — извлечь информацию из архивов, выбирая при этом только те сведения, которые необходимы конечному пользователю данных. Чтобы понять преимущество данного этапа, нужно сравнить его с более классическим алгоритмом. Более классический вариант работы с xml-файлами, запакованными во множество архивов, выглядел бы следующим образом: извлечение xml-файлов из архива — конвертация xml-файлов в табличный формат — сохранение табличных данных — фильтрация табличных данных. Несмотря на «каноничность» такого решения, у него есть существенный недостаток: оно приводит к формированию большого количества промежуточных файлов, которые занимают большой объём хранилища, но не имеют практической ценности. При небольшом объёме исходных данных проблема промежуточных файлов несущественна, однако, как говорилось выше, объём первого, самого крупного набора данных в распакованном виде оценивается в несколько терабайт, и примерно столько же будут занимать промежуточные файлы. Как следствие, для реализации алгоритма требуется либо компьютер, выходящий за рамки стандартного потребительского класса, либо аренда облачных ресурсов. И то, и другое дорого и неоптимально с точки зрения результата. Фильтрующее извлечение позволяет распаковывать и извлекать данные «на лету», сводя к минимуму образование промежуточных файлов. Сохраняются при этом только те сведения, которые необходимы конечному пользователю. В частности, можно отфильтровать данные по коду ОКВЭД или субъекту федерации. Итоговое табличное представление данных является более компактным, чем исходный набор xml-файлов, примерно в 10–20 раз, а при фильтрации — и более.
Шаг 3. Агрегация данных. Смысл этапа — построить хронологию на основе повторяющихся записей. Исходные данные представляют собой выгрузки с определённой периодичностью. За время, прошедшее между выгрузками, информация о конкретной организации или конкретном индивидуальном предпринимателе могла остаться прежней, в результате чего в табличном представлении может остаться несколько строк с одинаковыми данными, различающиеся только временной меткой. Суть агрегации в том, чтобы заменить несколько повторяющихся строк одной, для которой будет указано время актуальности. Поскольку проверка того, изменилась ли информация или нет, должна учитывать потенциальные несущественные технические изменения и ошибки в исходных данных, то агрегация также включает в себя базовую подготовку данных: приведение регистра, проверку написания кодов и номеров и тому подобное.
Шаг 4. Геопривязка данных. Смысл этапа — для каждой организации или каждого индивидуального предпринимателя указать адрес, координаты и код ОКТМО. Исходные данные содержат информацию об адресе согласно Федеральной информационной адресной системе исходя из административно-территориального деления: субъект федерации, район, город, населённый пункт. В процессе геопривязки адреса нормализуются. Исходя из нормализованного адреса, указывается код ОКТМО и географические координаты.
Шаг 5. Обогащение данных. На этом этапе базовая информация об организациях и индивидуальных предпринимателях дополняется вспомогательной. В качестве источника вспомогательной информации используются второй и третий наборы данных. Опционально возможно использовать и другие источники данных. Ключевым условием является наличие в дополнительных источниках данных ИНН, поскольку именно он является уникальным идентификатором.
Шаг 6 (опциональный). Подготовка панельного представления. Суть этапа — преобразовать полученные данные в более привычный формат. После прохождения описанных выше пяти этапов данные представлены в виде строк, каждая из которых содержит сведения о хозяйствующем субъекте за определённый период времени, причём период времени может быть разным. При этом начало и конец периода указаны с точностью до месяца и года. В географических исследованиях более применимы данные, которые характеризуют положение дел на какой-то год. Таким образом, в ходе данного шага данные преобразуются так, что для каждой организации или индивидуального предпринимателя за каждый год указывается преобладающее значение той или иной характеристики. Например, если в течение 10 месяцев в году организация находилась в Москве, а остальные два месяца — в Московской области, то считается, что весь год организация находилась в Москве.
Реализация алгоритма
Описанный выше алгоритм реализован как несколько скриптов на языке программирования Python. Для хранения данных использовался локальный компьютер. Для извлечения данных использовались стандартные модули Python и библиотека lxml. Агрегация данных производилась с помощью системы для обработки больших данных Spark. Геопривязка осуществлялась с помощью специально разработанного алгоритма на базе двух справочных таблиц: одной по городам России, другой по населённым пунктам России (преимущественно сельским). Обогащение данных и создание панели также реализованы как скрипты на Python. Исходный код опубликован по адресу: ...
Валидация данных
Включает в себя две стадии. Первая — это формальная валидация: проверка типа данных и заполненности разных полей итоговой таблицы. Вторая — это содержательная валидация. Очевидно, что реестр субъектов малого и среднего предпринимательства не содержит информации обо всех организациях в России и представляет собой выборку из него, причём в общем смысле не репрезентативную. Тем не менее, есть основания полагать, что выборка всё же является достаточно репрезентативной в отношении географического положения хозяйствующих субъектов и видов деятельности. Соответственно, на второй стадии валидации важно сравнить получившийся набор данных с генеральной совокупностью. В качестве генеральной совокупности используется статистический регистр Росстата. Для сравнения берётся сводная информация о количестве организаций в разрезе субъектов федерации и кодов ОКВЭД за 2019–2022 гг. Поскольку при географических исследованиях важным зачастую является не точное совпадение абсолютных чисел (которое в данном случае недостижимо), а совпадение относительного положения тех или иных мест или категорий, то в качестве метрики для оценки используется ранговый коэффициент корреляции Спирмена в разрезе регионов или кодов ОКВЭД.
Результаты
Для тестирования алгоритм был запущен со следующими параметрами:
1) временной диапазон входных данных — с августа 2016 г. по ноябрь 2023 г. (все доступные на сайте ФНС России данные);
2) группа ОКВЭД — «А» (сельское, лесное хозяйство, охота, рыболовство, рыболовство). Она объединяет несколько отраслей экономики, которые в экономическое географии обычно рассматриваются изолированно. Тем не менее, поскольку при тестировании важно показать принцип работы алгоритма (извлечение по группам или кодам ОКВЭД), то было отдано предпочтение именно классификатору, а не группировке, принятой в научной сфере. Кроме того, такой вариант запуска показывает, что алгоритм позволяет одновременно извлечь сведения о нескольких связанных отраслях экономики.
Сводная статистика по работе алгоритма
Примерное время подготовки набора данных составило 15 часов, вычисления выполнялись на обычном персональном компьютере с примерно 300 Гб свободного места на диске и 16 Гб оперативной памяти.
На схеме показано уменьшение количества обрабатываемых данных от шага к шагу алгоритма. Входные данные — это 88 архивов суммарным объёмом 193 Гб в сжатом виде, содержащие около 600 тысяч xml-файлов и 550 миллионов исходных записей. После фильтрующего извлечения промежуточный набор данных состоял из 88 csv-файлов объёмом 6 Гб и суммарно содержал 15 миллионов строк. В результате агрегации получился один датасет (csv-файл) объёмом 133 Мб, содержащий 500 тысяч строк. Геопривязанный датасет содержит почти столько же строк (их количество немного уменьшилось из-за того, что часть адресов после нормализации оказалась дублями) и занимает чуть больше 200 Мб (объём вырос из-за добавления нормализованных адресов к «сырым»). Панельное представление также является одним csv-файлов объёмом около 600 Мб, содержащим несколько миллионов строк (панельное представление менее компактно, так как данные, не меняющиеся год от года, дублируются для сохранения временной структуры). Таким образом, в результате обработки объём данных уменьшился с нескольких терабайт до нескольких сотен мегабайт (с промежуточными файлами — нескольких гигабайт), и из полного списка организаций, содержащего несколько сотен миллионов записей, было выбрано несколько сотен тысяч интересующих. Кроме того, данные обогатились геопривязкой и кодами ОКТМО. В результате получился компактный датасет, который можно обработать на обычном компьютере без установки специальных программ для работы с «большими данными» — в отличие от исходного набора данных на сайте ФНС России.
Итоговый датасет содержит информацию о 300 тысячах индивидуальных предпринимателей и 100 тысячах огранизаций, осуществляющих или осуществлявших деятельность в сфере сельского, лесного хозяйства, охоты, рыболовства и рыбоводства. Он позволяет искать, группировать, агрегировать и анализировать данные в нескольких разных аспектах. Приведём основные из них.
Временная динамика
Набор данных охватывает период с 2016 по 2023 г. с точностью до месяца. Соответственно, он позволяет проследить динамику численности хозяйствующих субъектов в том или ином регионе или той или иной сфере деятельности. На графике 1.1 показаны первые 10 регионов с наибольшим количеством организаций в группе А за каждый месяц из анализируемого периода, а на графике 1.2 — изменение числа хозяйствующих субъектов в отдельных укрупнённых сферах экономики: сельское хозяйство, лесное хозяйство, охота, рыболовство и рыбоводство.
Пространственная структура по субъектам федерации
Набор данных охватывает все субъекты России, для которых имеются исходные данные. Благодаря нормализованному (приведённому в соответствие с Конституцией России) названию субъекта федерации данные можно легко агрегировать и тем самым подсчитать количество хозяйствующих субъектов в каждом регионе в целом или по отдельным кодам ОКВЭД. На картограмме 2.1 показано распределение регионов по количеству хозяйствующих субъектов в сфере сельского хозяйства, а на картограмме 2.2 — лесного хозяйства. Видно, что география указанных отраслей ожидаемо отличается. Глубокая детализация по кодам ОКВЭД позволяет выбрать и визуализировать распределение хозяйствующих субъектов по узким и довольно специфичным кодам. Так, картограмма 2.3 показывает распределение организаций с кодом 1.6.22 «Выращивание подсолнечника», а картограмма 2.4 — (что-то про лесное хозяйство).
Пространственная структура по населённым пунктам
Набор данных содержит информацию о местоположении хозяйствующих субъектов с точностью до населённого пункта, а также координаты этих пунктов и коды ОКТМО соответствующих муниципальных образований. Это позволяет агрегировать данные по населённым пунктам, визуализировать их на картах и схемах, а также связывать со статистикой по муниципальным образованиям.
На картограмме 3.1 показано распределение организаций по территории России в виде точек, изображающих населённые пункты. Картограмма 3.2 представляет собой часть картограммы 1 и показывает число и соотношение видов организаций в г. Екатеринбурге и окрестностях.
В таблице 3.2 приведены результаты простейшего анализа данных во взаимосвязи с внешними ресурсами: количество организаций в городах в пересчёте на тысячу жителей.
Отслеживание отдельных хозяйствующих субъектов
Набор данных детилизирован до отдельных организаций и индивидуальных предпринимателей, что позволяет проследить динамику каждого из них, а также агрегировать информацию об изменениях. В таблице 4.1 показано число изменений юридического адреса у хозяйствующих субъектов. На картограмме 4.2 показаны «регионы-доноры» и «регионы-реципиенты» хозяйствующих субъектов за период с 2017 по 2021 г.
Анализ наименований
Набор данных содержит сведения о наименованиях организаций. Наименования сами по себе могут быть ценным источником информации, если обрабатывать их с использованием методов автоматизированной обработки текстов. На схемах 5.1 и 5.2 показаны «облака слов» из названий организаций в сфере сельского и лесного хозяйства, соответственно.
Дополнительные атрибуты
Расширение набора данных за счёт информации о численности работников организаций и доходах с расходами позволяет перейти от простого подсчёта к «взвешенному» подсчёту и анализировать территориальное, временное или структурное распределение с учётом прибыли и числа занятых. На схемах 6.1–6.3 показано распределение соответственно числа организаций, их суммарной прибыли и числа работников в сфере «сельское хозяйство».
Обсуждение
Рассуждая о предложенном алгоритме, необходимо обратить внимание на два его аспекта: техническую ценность и исследовательскую ценность. Говоря об исследовательской ценности, важно обратить внимание на два аспекта: частный (о применимости получаемого датасета для исследований в области малого и среднего предпринимательства) и общий (возможность использования данных для выводов об экономике в целом).
Техническая ценность предложенного алгоритма связана с тем, что он позволяет достаточно эффективно (без существенных финансовых затрат и за сравнительно короткий период времени) обработать массивы исходных данных ФНС России и представить их в виде, подходящем для географических исследований, то есть плоских таблиц с геопривязкой. Благодаря алгоритму и его прикладной реализации несколько терабайт исходной информации фильтруются и компактно представляются в виде относительно небольшой таблицы, которую можно использовать на стандартном компьютере без облачных технологий и инструментов работы с большими данными. Программный код находится в открытом доступе и может использоваться любым специалистом, владеющим языком программирования Python, для создания собственных производных решений.
Исследовательская ценность алгоритма и набора данных в контексте исследований малого и среднего предпринимательства заключается в том, что датасет представляет собой максимально полный и детализированный из доступных набор сведений об организациях и предпринимателях, относящихся к МСП. Он объединяет в себе три измерения: пространственное (с точностью до населённого пункта), временное (с точностью до месяца или года) и структурное (с точностью до отдельных кодов ОКВЭД) и представлен в наиболее дезагрегированном виде: в отличие от традиционных статистических данных, исходная единица в нём — это конкретный хозяйствующий субьект. Таким образом, набор данных может рассматриваться как квази-источник первичных статистических сведений. Поскольку вся представленная в нём информация об организациях представляет собой изъятие из режима налоговой тайны и по закону имеет открытый характер, то использование датасета не нарушает права частных лиц и законодательство России.
Помимо организаций, набор данных содержит также сведения об индивидуальных предпринимателях. Его ценность ниже, так как для них отсутствуют дополнительные данные о доходах, расходах и числе работников. Кроме того, если информация о названиях организаций обладает самостоятельной исследовательской ценностью, то имена физических лиц такой ценностью вряд ли характеризуются. Кроме того, сведения в датасете относятся к информации об определённом физическом лице, то есть являются персональными данными, что поднимает вопрос о юридических рисках их использования, особенно в свете ужесточения законодательства о персональных данных.
Вопрос об исследовательской ценности набора данных для анализа экономики России в целом более сложен. С одной стороны, очевидно, что субъекты малого и среднего предпринимательства — это не все хозяйствующие субъекты в России. Более того, понятно также, что они не являются случайной выборкой из генеральной совокупности таких субъектов, поскольку отобраны по конкретным критериям: число работников и доход. Иначе говоря, реестр и производный от него датасет содержат сведения только о части фирм, причём о самых небольших из них, а крупный бизнес при этом выпадает из поля зрения. Тем не менее, есть некоторые доказательства, что как минимум отчасти такой датасет может быть репрезентативным в географическом и структурном отношении. Как показано в разделе с валидацией, агрегированная статистика по регионам и группам деятельности, полученная на базе датасета, высоко скоррелирована с данными, полученными на базе статистического регистра Росстата, который является генеральной совокупностью. Коэффициент корреляции не очень высокий разве что в отдельных специфических группах, что объяснимо (большая часть образовательных организаций и организаций в сфере публичной власти — не частные, а государственные или муниципальные), в остальных случаях он превышает 0.9. Иначе говоря, датасет на базе реестра МСП (как выборка) хорошо скоррелирован с генеральной совокупностью в географическом (с точностью до регионов) и структурном (с точностью до большинства групп ОКВЭД) разрезах. Это позволяет говорить о его репрезентативности как минимум в этих двух аспектах. Так, при любом ранжировании регионов или видов деятельности результат, полученный на базе реестра МСП, будет в основном совпадать с результатом, полученным на базе сплошного обследования всех фирм в России. Это не говорит о том, что датасет репрезентативен в более «узких» разрезах (на уровне отдельных кодов ОКВЭД или конкретных населённых пунктов), но в отсутствие иных источников информации позволяет использовать его в качестве первого — и наиболее адекватного из доступных — приближений к генеральной совокупности в тех случаях, когда требуются данные по экономическим субъектам в России с высокой детализацией. Тем не менее, вопрос о репрезентативности желательно при каждом конкретном исследовании анализировать отдельно.
Два сопутствующих вопроса, касающихся исследовательской ценности в любом аспекте, касаются соответствия между номинальным и реальным положением дел. В реестре МСП указаны юридический (номинальный) адрес и номинальный же код ОКВЭД. Очевидно, что они могут отличаться от реальных. Также следует заметить, что отличия могут быть разной степени выраженности в зависимости от места и вида деятельности. Так, можно предположить, что место регистрации фирмы будет как минимум отчасти объясняться налоговыми или престижными соображениями, а место регистрации фирм, осуществляющих деятельность в интернете или в нескольких регионах, будет мало связано с местом их регистрации. Автору не известны какие-либо исследования на эту тему, равно как и надёжные и легко реализуемые методы, позволяющие проверить соответствие между номинальными данными и реальностью. Понятно лишь то, что датасет и любые производные от него статистические данные будут сочетать в себе как реальные закономерности, так и искажения, вызванные различием номинального и реального, причём оценить масштаб этих расхождений априори не представляется возможным — требуется проверка в каждом конкретном случае. Поскольку суть статьи сводится к описанию алгоритма и технологии, то остаётся лишь оставить здесь соответствующее предостережение для других исследователей. Потенциальными источниками «реальных» сведений могут быть, в частности, материалы открытых или частных картографических сервисов либо данные выборочных обследований Росстата.
Выводы

### 2023-12-09 Paper.Rmd
Introduction, 2nd paragraph

@@
 Федорченко взял данные Международной организации автопроизводителей, рассчитал на их основе индексы и дополнил их картографической визуализацией, что позволило эффективно проанализировать территориальные сдвиги в мировой автомобильной промышленности [@Федорченко2017].
-Таким образом, наблюдается общая тенденция: в том или ином виде используются данные, хотя объём, тип, качество этих данных и технологии их обработки отличаются.
-«Большие данные» и методы машинного обучения также проникают в общественно-географическую науку, что отметили ещё десять лет назад М.
+«Большие данные» также проникают в общественно-географическую науку, что отметили ещё десять лет назад М.
 Грэм и Т.
@@
 Недавний пример такого рода исследований — работа Вэй Лю и Вэньцзэ У, которые используют сведения о пользователях социальной сети WeChat для изучения связей между типом землепользования и деятельностью людей [@Liu2020].
+Таким образом, общая тенденция такова: для исследований используются данные, хотя объём, тип, качество этих данных и технологии их обработки отличаются.
 
-Необходимость использования данных закономерно ведёт к вопросу об источниках их получения, поскольку прежде чем обрабатывать данные для исследований и делать выводы из них, необходимо эти данные где-то взять: найти в готовом или пригодном для обработки виде или собрать самостоятельно.
-Поиск качественных данных и их предварительная обработка во многом определяют результативность научной деятельности: если данных нет, они некачественные или подготовлены неправильно, то исследование с высокой вероятностью окажется неудачным или приведёт к заблуждениям.
-Используемые источники данных и технологии их обработки, таким образом, являются важным прикладным аспектом большинства экономико-географических исследований.
+Прежде чем использовать данные, необходимо их получить.
+Поиск качественных данных в достаточном количестве и их предварительная обработка во многом определяют результативность научной деятельности: проблемы в данных или в их анализе часто приводят к неправильным выводам, а сделать исследование при нехватке данных зачастую невозможно.
+Следовательно, источники данных и информационные технологии — это важный прикладной аспект общественной географии как науки.
+В.
+Н.
+Лаженцев, например, отмечает, что «совершенствование статистической науки в значительной мере предопределяет уровень экономико-географических исследований» [@Лаженцев2015], а К.
+Брунсдон рассуждает о важности изложения точных деталей анализа данных в контексте воспроизводимости географических исследований [@Brunsdon2016].
 Практическая важность данных для исследований подтверждается, в частности, различными инициативами, связанными с обеспечением доступа к таким данным.
-Среди этих инициатив есть как государственные, так и частные.
+Так, П.
+Арцбергер с соавторами ещё 20 лет назад предложили следующий принцип: результаты исследований, финансируемых на публичном уровне, должны быть максимально общедоступными, поскольку это, в частности, способствует новым исследованиям и тем самым увеличивает научный потенциал и отдачу от инвестиций в науку [@Arzberger2006], а Д.
+Кард с коллегами указывают на важность доступа учёных к административным данным, то есть информации, которая накапливается органами исполнительной власти при осуществлении их функций [@Card2010].
+Для географических исследований особую роль играет такая разновидность данных, как пространственные данные, обеспечение доступа к которым в недавние годы также вызывает интерес во многих странах, включая США, Великобританию, государства ЕС, Сингапур [@Jozefowicz2020].
 

### 2023-12-10 Paper.Rmd
Introduction, paragraphs 3 and 4

@@
 Прежде чем использовать данные, необходимо их получить.
-Поиск качественных данных в достаточном количестве и их предварительная обработка во многом определяют результативность научной деятельности: проблемы в данных или в их анализе часто приводят к неправильным выводам, а сделать исследование при нехватке данных зачастую невозможно.
+Поиск качественных данных в достаточном количестве и их предварительная обработка во многом определяют результативность научной деятельности: проблемы в данных или в их анализе приводят к неправильным выводам, а сделать исследование при нехватке данных зачастую невозможно.
 Следовательно, источники данных и информационные технологии — это важный прикладной аспект общественной географии как науки.
@@
 
-В российских реалиях наиболее типичным и массово используемым источником данных для экономико-географических исследований являются ресурсы Росстата.
-Они включают в себя официальные статистические публикации, статистические данные, размещённые на официальном сайте, базу данных показателей муниципальных образований (БДМО) и веб-портал единой межведомственной информационно-статистической системы (ЕМИСС).
-Причины такой популярности Росстата понятны.
-Во-первых, Росстат — это государственный орган, уполномоченный на сбор информации от хозяйствующих субъектов, в том числе конфиденциальной, имеющий необходимые права, обязанности и гарантии, несущий ответственность за своевременность, полноту и качество предоставляемых и распространяемых сведений.
-Во-вторых, деятельность Росстата и его предшественников охватывает всю территорию страны, все отрасли экономики, все виды организаций и значительный временной промежуток, поэтому по количеству собранной информации он превосходит иные источники, а в каких-то вопросах является единственным прямым источником информации.
-В-третьих, ресурсы Росстата широко известны, относительно просты для использования (даже если эта простота порой кажущаяся), в основном доступны без регистрации и платы.
+Типичным источником данных для исследований общественной географии России являются ресурсы Федеральной службы государственной статитстики (Росстата).
+К ним относятся, в частности, размещённые на официальном сайте этого органа данные официальной статистики, переписей и статистических обследований, база данных показателей муниципальных образований (БДМО) и единая межведомственная информационно-статистическая система (ЕМИСС).
+О.
+В.
+Кузнецова при исследовании структуры экономики российских регионов использует статистику валового регионального продукта и объёма отгруженных товаров [@Кузнецова2018]; Г.
+М.
+Фёдоров в своём анализе экономики регионов России на Балтике опирается на данные Росстата и Федеральной таможенной службы [@Fedorov2022]; С.
+Г.
+Сафронов изучает третичный сектор экономики России на базе данных Eurostat, Росстата и Федеральной налоговой службы [@Сафронов2021]; Ю.
+В.
+Петров на базе региональной, муниципальной статистики и данных иных органов власти формулирует предложения по социально-экономическому развитию юга Тюменской области [@Petrov2021]; М.
+А.
+Саранча оценивает развитие гостиничного и ресторанного бизнеса, задействуя результаты сплошного наблюдения Росстата за малым и среднем предпринимательством [@Sarancha2014].
+Данные Росстата также используются для создания производных наборов данных, как в работе И.
+Ю.
+Пекер [@Peker2022].
+Таким образом, большинство общественно-географических исследований не обходятся без применения данных Росстата как единственного источника информации или как одного из источников.
 
-Тем не менее, есть причины, которые могут побудить к поиску и использованию других источников данных.
-К числу общих недостатков Росстата как источника данных можно отнести запутанность и неочевидность интерфейсов, описаний, каталогов и классификаций, частые и взаимно несовместимые изменения схем и форматов публикации сведений, низкую пригодность данных к автоматизированной обработке, и кроме того, порой обоснованные сомнения в достоверности информации.
+Тем не менее, в ряде случаев данных Росстата недостаточно.
+Как следствие, возникает необходимость в альтернативных источниках информации, которые либо дополняют официальную статистику, либо заменяют её.
+Причины этого бывают разные.
+В некоторых случаях альтернативные источники информации используются для проверки или корректировки данных Росстата.
+Например, Р.
+А.
+Бабкин, основываясь на данных сотовых операторов, доказывает, что официальная численность населения московской агломерации занижена примерно на 10% из-за погрешности и инертности статистического учёта и «двудомности» значительной части москвичей [@Babkin2020].
+Другая, более частая причина — отсутствие нужных сведений.
+Например, К.
+В.
+Ростислав, анализируя влияние местоположения на прибыльность предприятий, учитывает только организации с численностью работников меньше 250 человек и для проверки этого условия использует открытые данные ФНС России, так как Росстат такие сведения не предоставляет [@Rostislav2021].
+С.
+С.
+Лачининский с соавторами оценивает спрол городских территорий Санкт-Петербурга с опорой на спутниковые снимки Landsat, так как на базе сведений официальной статистики провести необходимые подсчёты невозможно [@Lachininskiy2023].
+Данные Росстата в силу закона являются сводными, в то время как для ряда исследований нужна дезагрегированная информация.
+Кроме того числу общих недостатков Росстата как источника данных можно отнести запутанность и неочевидность интерфейсов, описаний, каталогов и классификаций, частые и взаимно несовместимые изменения схем и форматов публикации сведений, низкую пригодность данных к автоматизированной обработке, и кроме того, порой обоснованные сомнения в достоверности информации.
 Если же говорить конкретно о географических исследованиях, особенно в сфере экономической географии, то ключевым недостатком является низкая детализация данных по разным параметрам.

### 2023-12-11 Paper.Rmd
Continue editing introduction

@@
 Лачининский с соавторами оценивает спрол городских территорий Санкт-Петербурга с опорой на спутниковые снимки Landsat, так как на базе сведений официальной статистики провести необходимые подсчёты невозможно [@Lachininskiy2023].
-Данные Росстата в силу закона являются сводными, в то время как для ряда исследований нужна дезагрегированная информация.
-Кроме того числу общих недостатков Росстата как источника данных можно отнести запутанность и неочевидность интерфейсов, описаний, каталогов и классификаций, частые и взаимно несовместимые изменения схем и форматов публикации сведений, низкую пригодность данных к автоматизированной обработке, и кроме того, порой обоснованные сомнения в достоверности информации.
-Если же говорить конкретно о географических исследованиях, особенно в сфере экономической географии, то ключевым недостатком является низкая детализация данных по разным параметрам.
-Так, официальные статистические показатели обычно представлены в разрезе регионов (субъектов федерации).
-Даже муниципальный уровень доступен не всегда: обычно по нему есть только некоторые показатели или наиболее новые данные.
-Более точная географическая привязка отсутствует, и это порой может становиться проблемой.
-В частности, в геоинформационных системах типично представление с использованием географических координат, но официальная статистическая информация не содержит координат, в результате чего любое картографирование и визуализация таких сведений неизбежно сводится к уровню субъектов федерации или муниципальных образований, что снижает пространственную детализацию и потенциально может скрывать некоторые закономерности.
-Временное разрешение официальных статистических данных — год, хотя в некоторых случаях этого недостаточно.
-Классификация по сферам экономики также довольно обширна: отрасли либо не дифференцированы вовсе, либо сведены к широким группам классификатора видов экономической деятельности, что, конечно, достаточно для изучения традиционных отраслей, но мало при фокусировании внимания на относительно новых или сильно дифференцированных сфер деятельности, например, многих видов услуг.
-Причины такого положения дел отчасти понятны: во-первых, низкая детализация снижает риски раскрытия первичных статистических сведений, которые по закону конфиденциальны, а во-вторых, потребителей у информации Росстата много, и вряд ли он будет фокусироваться на удовлетворении довольно специфических запросов некоторых специалистов по экономической географии.
-Тем не менее, в результате актуален вопрос о возможностях и технологиях использования иных источников данных для экономико-географического анализа.
+К преимуществам альтернативных источников данных относят повышенную точность и своевременность [@Воробьев2021], более высокое пространственное и временное разрешение [@Бабкин2021], возможность точного и детального анализа на микроуровне [@Бочкарёв2017], отмечая при этом несвоевременность, неполноту и потенциальную недостоверность официальной статистики [@Кузьминов2021].
+Альтернативные источники данных дают доступ к генеральной совокупности, вследствие чего «отпадает необходимость в использовании выборочных, агрегированных и косвенных» источников сведений [@Блануца2017].
+При использовании методов машинного обучения, что также становится актуальным в современной географии, важны объём выборки, дезагрегированность, количество признаков в данных, что также побуждает обращаться к альтернативным источникам [@Ростислав2020].
+Таким образом, хотя традиционные статистические сведения по-прежнему важны в общественно-географических исследованиях, а порой незаменимы, есть нарастающая потребность в других информационных ресурсах.
+Запрос со стороны исследователей идёт на повышение объёма, оперативности предоставления, точности, пространственного и временного разрешения, расширение набора доступных показателей, а также на возможность валидации статистических сведений.
+Он проявляется и в части традиционных научных методов, так и методов, связанных с машинным обучением, хотя более актуален именно для последних, наиболее чувствительных к объёму данных, их дезагрегированности, детализации, удобным форматам представления.
+
+Одним из альтернативных источников информации, которые входят в обиход географов-обществоведов, являются административные данные Федеральной налоговой службы России (ФНС России), которые публикуются на её официальном сайте или связанных с ним ресурсах.
+Используются, в частности, сведения о доходах физических лиц [@Петров2021], о налоговых поступлениях в разрезе субъектов федерации или видов экономической деятельности [@Ростислав2020; @Сафронов2021], о среднесписочной численности работников организаций [@Ростислав2021], о количестве учтённых жилых помещений [@Lachininsky2021], о количестве малых и средних предприятий [@Гуменюк2022].
+Внимание исследователей к данным, публикуемым ФНС России, связано, по-видимому, с многофункциональностью этого органа власти, из-за которой он собирает огромный объём информации о разных сторонах российской экономики (хозяйствующие субъекты, имущество, доходы, расходы).
+Как правило, ФНС России публикует сводную (агрегированную) информацию.
+Тем не менее, есть несколько наборов данных, содержащих дезагрегированные (исходные, первичные, «сырые») данные, из которых наибольшее внимание привлекает реестр субъектов малого и среднего предпринимательства, который полностью и регулярно размещается на сайте службы и содержит сведения о конкретных организациях и индивидуальных предпринимателях.
 

### 2023-12-12 Paper.Rmd
Finish introduction, split introduction into introduction and literature review, start writing data sources part, add stats helper

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(include = FALSE)
+```
+
+```{r imports}
+source("StatsHelper.R")
 ```
@@
 
+В статье предложена технология, позволяющая формировать геопривязанный табличный набор данных о малых и средних фирмах в России.
+В качестве основы используется реестр субъектов малого и среднего предпринимательства (реестр МСП), размещённый в форме открытых данных на сайте Федеральной налоговой службы России (ФНС России).
+Набор данных охватывает всю территорию России, все виды экономической деятельности и период с середины 2016 г.
+по настоящее время.
+Его пространственное разрешение — до конкретного населённого пункта, временное — до месяца, структурное — до конкретного кода по общероссийскому классификатору видов экономической деятельности (ОКВЭД).
+Сведения являются дезагрегированными, то есть детализированными до уровня отдельных фирм (организаций и индивидуальных предпринимателей).
+Помимо структурированных адресов, набор содержит географические координаты и коды муниципальных образований.
+В отношении субъектов малого и среднего предпринимательства (МСП) набор данных представляет собой генеральную совокупность.
+Кроме того, в статье обосновывается, что указанный набор данных может использоваться как достаточно репрезентативная выборка для анализа экономики в целом, а не только малого и среднего предпринимательства.
+
+Статья, помимо введения, включает в себя несколько разделов.
+В обзоре литературы приводится краткая информация об использовании различных источников данных в общественно-географических исследованиях и через это обосновывается необходимость в предложенной технологии.
+Раздел «Исходные данные» содержит описание первичных наборов данных, которые задействуются в технологии, а раздел «Методика обработки данных» излагает саму технологию.
+В разделе «Тестирование технологии» описываются и визуализируются процесс и результаты опробования технологии в отношении сведений о фирмах, ведущих деятельность в сфере сельского и лесного хозяйства, охоты и рыболовства.
+Раздел «Возможность генерализации результатов» посвящён оценке репрезентативности полученного набора данных применительно к экономике России в целом.
+В «Обсуждении» затрагиваются ключевые спорные моменты касательно предложенной технологии.
+«Выводы» подводят итоги работы.
+
+# Обзор литературы: источники данных в общественной географии
+
 Типичным источником данных для исследований общественной географии России являются ресурсы Федеральной службы государственной статитстики (Росстата).
@@
 С.
-Лачининский с соавторами оценивает спрол городских территорий Санкт-Петербурга с опорой на спутниковые снимки Landsat, так как на базе сведений официальной статистики провести необходимые подсчёты невозможно [@Lachininskiy2023].
+Лачининский с соавторами оценивает спрол городских территорий Санкт-Петербурга с опорой на спутниковые снимки Landsat, так как на базе сведений официальной статистики провести необходимые подсчёты невозможно [@Lachininsky2023].
 К преимуществам альтернативных источников данных относят повышенную точность и своевременность [@Воробьев2021], более высокое пространственное и временное разрешение [@Бабкин2021], возможность точного и детального анализа на микроуровне [@Бочкарёв2017], отмечая при этом несвоевременность, неполноту и потенциальную недостоверность официальной статистики [@Кузьминов2021].
@@
 
-Одним из альтернативных источников информации, которые входят в обиход географов-обществоведов, являются административные данные Федеральной налоговой службы России (ФНС России), которые публикуются на её официальном сайте или связанных с ним ресурсах.
-Используются, в частности, сведения о доходах физических лиц [@Петров2021], о налоговых поступлениях в разрезе субъектов федерации или видов экономической деятельности [@Ростислав2020; @Сафронов2021], о среднесписочной численности работников организаций [@Ростислав2021], о количестве учтённых жилых помещений [@Lachininsky2021], о количестве малых и средних предприятий [@Гуменюк2022].
-Внимание исследователей к данным, публикуемым ФНС России, связано, по-видимому, с многофункциональностью этого органа власти, из-за которой он собирает огромный объём информации о разных сторонах российской экономики (хозяйствующие субъекты, имущество, доходы, расходы).
-Как правило, ФНС России публикует сводную (агрегированную) информацию.
-Тем не менее, есть несколько наборов данных, содержащих дезагрегированные (исходные, первичные, «сырые») данные, из которых наибольшее внимание привлекает реестр субъектов малого и среднего предпринимательства, который полностью и регулярно размещается на сайте службы и содержит сведения о конкретных организациях и индивидуальных предпринимателях.
+Одним из альтернативных источников информации для географов-обществоведов являются административные данные Федеральной налоговой службы России (ФНС России).
+Они публикуются на официальном сайте этого органа власти или связанных с ним ресурсах.
+В географических работах используются, в частности, сведения о доходах физических лиц [@Petrov2021], о налоговых поступлениях в разрезе субъектов федерации или видов экономической деятельности [@Ростислав2020; @Сафронов2021], о среднесписочной численности работников организаций [@Rostislav2021], о количестве учтённых жилых помещений [@Lachininsky2021], о количестве малых и средних предприятий и мерах их поддержки [@Гуменюк2022].
+Внимание исследователей к данным, публикуемым ФНС России, связано, по-видимому, с многофункциональностью этого органа власти, из-за которой он собирает огромный объём информации о разных сторонах российской экономики: хозяйствующих субъектах, имуществе, доходах, расходах.
+Обычно ФНС России, как и Росстат, публикует сводную (агрегированную) информацию.
+Тем не менее, есть несколько наборов данных, содержащих дезагрегированные (исходные, первичные, «сырые») сведения.
+Наибольшее внимание среди них привлекает реестр субъектов малого и среднего предпринимательства (реестра МСП), который регулярно и полностью размещается на сайте службы и содержит сведения о конкретных организациях и индивидуальных предпринимателях.
 
-В статье представлена технология, позволяющая использовать некоторые наборы открытых данных Федеральной налоговой службы в качестве источника информации об организациях и индивидуальных предпринимателях, осуществляющих деятельность в том или ином месте в той или иной сфере экономики, а также о доходах, расходах и численности работников этих организаций.
-Несмотря на свою относительную концептуальную простоту, технология является достаточно универсальной и потенциально полезной, особенно учитывая нехватку наборов данных для исследований, особенно в машиночитаемом виде.
-Она позволяет получить данные за период с середины 2016 года с детализацией по месяцам, отдельным населённым пунктам (городским и сельским) и конкретным кодам классификатора видов экономической деятельности.
-Кроме того, технология обеспечивает детализацию до конкретных физических лиц и организаций, что открывает возможности для углублённого анализа (например, миграции отдельных компаний и сопутствующая динамика).
-Полнота сведений отличается в зависимости от цели использования:
-а) применительно к малому и среднему предпринимательству технология позволяет получить полный список соответствующих организаций и индивидуальных предпринимателей;
-б) применительно к экономике в целом технология позволяет получить выборку организаций, которая является репрезентативной в пространственном аспекте и в аспекте классификации по видам экономической деятельности.
+Для некоторых экономико-географических и смежных исследований важны сведения, детализированные до уровня конкретных фирм.
+К таким сведениям относятся, например, местонахождение, выручка, расходы, число работников, наименование и другие характеристики.
+Указанные данные могут использоваться напрямую, как в работах К.
+В.
+Ростислава [@Rostislav2021], Е.
+А.
+Коломак и А.
+И.
+Шерубнёвой [@Коломак2023], которые анализирует агломерационные эффекты на основе показателей предприятий.
+Кроме того, на их основе можно считать статистику с произвольной пространственной или временной группировкой, в том числе такой, которую невозможно, сложно или неудобно извлекать из данных Росстата.
+Например, И.
+С.
+Гуменюк оперирует подсчётами абсолютного и относительного числа малых и средних предприятий по муниципалитетам Калининградской области [@Гуменюк2022]; М.
+А.
+Саранча включает количество предприятий в интегральный индекс уровня развития малого предпринимательства на территории муниципального образования [@Sarancha2014]; а М.
+А.
+Макушин с соавторами использует данные о предприятиях в городах БАМа, чтобы оценить структуру промышленности [@Макушин2023].
+Дополнительную важность сведения о конкретных фирмах приобретают при исследованиях в масштабе городов [@Lachininsky2021; Никоноров2021; @Kozhevnikov2023], при изучении узких секторов экономики, которые не фигурируют обособленно в статистике [@Моисеева2016; @Pankratov2023], при анализе отдельно взятой отрасли [@Кузьминов2021].
 
-В следующем разделе, «Данные и методы», приводится описание источников «сырых» данных и методов их обработки в общем виде, то есть на концептуальном уровне формулируется алгоритм, составляющий суть технологии.
-Кроме того, даётся краткое описание авторской реализации данной технологии и  инструментов, использованных для этой реализации, а также ссылка на исходный код компьютерной программы.
-Раздел «Результаты» описывает результаты пробного использования технологии для получения информации о компаниях и физических лицах, осуществляющих деятельность в сфере юридических услуг, а также результаты валидации технологии с использованием статистики Росстата.
-В разделе «Обсуждение» анализируются возможные варианты и ограничения использования технологии, а также некоторые аспекты её практической реализации.
-«Выводы» подводят итоги статьи.
+Общепринятого источника геопривязанной информации о фирмах в России нет.
+Исследователи и прикладные специалисты обычно опираются на данные коммерческих [@Lachininsky2021] или открытых геоинформационных сервисов, систем бизнес-аналитики [@Моисеева2016; @Кузьминов2021], уже упомянутого реестра МСП [@Гуменюк2022].
+Как правило, использование этих способов сопряжено с двумя проблемами: данные предоставляются за плату либо требуется приложить значительные усилия для их приведения в подходящий для анализа формат.
+Системы бизнес-аналитики, такие как Spark и Ruslana, содержат много разнообразной информации, но платные.
+Выгрузка данных из частных геоинформационных систем наподобие 2GIS и Яндекс.Карт также стоит денег.
+Открытые сервисы наподобие OpenStreetMap или реестра МСП бесплатны, но выгрузки из них имеют большой объём и сложную структуру, поэтому исследователю приходится тратить много усилий на техническую сторону работы, привлекать сторонних специалистов или отказываться от использования такого источника данных.
+Формально полный список российских организаций и индивидуальных предпринимателей содержится в двух государственных реестрах (соответственно Едином государственном реестре юридических лиц и Едином государственном реестре индивидуальных предпринимателей), однако ни один из них не распространяется в полном виде — можно лишь извлекать отдельные записи вручную через веб-сервис ФНС России либо автоматизированно за плату через программный интерфейс, что не даёт доступ к генеральной совокупности компаний и вряд ли позволяет построить репрезентативную выборку без чрезмерных затрат и усилий.
+Таким образом, актуален вопрос об удобном способе, который позволил бы бесплатно и легально получить максимально полный набор геопривязанных данных о российских фирмах.
 
-# Данные и методы
+# Исходные данные
 
-Исходные данные
-В качестве исходных данных, которые обрабатываются алгоритмом, используются три набора открытых данных Федеральной налоговой службы (ФНС) России.
-Первый, основной — это набор «Единый реестр субъектов малого и среднего предпринимательства».
-Технически он представляет собой коллекцию zip-архивов xml-файлов.
-Zip-архив представляет собой «срез» (выгрузку) из реестра по состоянию на определённую дату.
-Одна дата представлена ровно одним архивом.
-В каждом архиве находится множество (несколько тысяч) xml-файлов.
-Каждый xml-файл, в свою очередь, содержит в себе информацию о множестве (обычно ровно 900) организациях или индивидуальных предпринимателях, включённых в реестр.
-Одна организация или один индивидуальный предприниматель в xml-файле представлен элементом «Документ».
-Отдельные характеристики организации или предпринимателя (название, ИНН, коды ОКВЭД, местонахождение и другие) представлены как вложенные элементы и их атрибуты.
-Общая схема указанной структуры данных представлена на рисунке.
-Указанный набор открытых данных доступен за период с августа 2016 г.
-по настоящее время.
-Новые выгрузки публикуются ежемесячно.
-Таким образом, на момент написания статьи доступно около 85 выгрузок.
-Каждая выгрузка (архив) имеет объём от 2 до 3 Гб в сжатом виде и от 30 до 50 Гб в распакованном.
-Общий объём исходных данных этого набора в несжатом виде, таким образом, оценивается в несколько терабайт (тысяч гигабайт), что приводит к необходимости использования специальных методов оптимизации и конкретных технологических решений, адаптированных к эффективной работе с «большими данными», о которых будет сказано ниже.
+Технология опирается на один основной источник исходных данных и несколько вспомогательных источников.
+
+## Основной источник: реестр МСП
+
+Реестр МСП — это государственная информационная система, содержащая информацию о субъектах малого и среднего предпринимательства.
+Она предусмотрена ст.
+4.1 ФЗ «О развитии малого и среднего предпринимательства в Российской Федерации».
+Субъекты малого и среднего предпринимательства (субъекты МСП) — это коммерческие юридические лица и индивидуальные предприниматели, соответствующие критериям по доходу, числу работников и некоторым другим показателям.
+За реестр отвечает ФНС России.
+Сведения в реестр попадают автоматически, то есть без участия самого субъекта МСП, благодаря чему обеспечивается максимально возможная полнота данных.
+По закону, сведения реестра ежемесячно размещаются в Интернете и являются общедоступными в течение пяти лет с года опубликования.
+Норма о реестре появилась в законе в конце 2015 г.
+На практике реестр размещён на сайте ФНС России в форме открытых данных в виде ежемесячных полных выгрузок, первая из которых датируется августом 2016 г.
+
+Технически реестр МСП — это коллекция zip-архивов xml-файлов.
+Каждый zip-архив — это «срез» (выгрузка, дамп, dump) реестра на определённую дату.
+Коллекция zip-архивов тем самым образует временной ряд (time series).
+Каждый zip-архив содержит множество (несколько тысяч) xml-файлов.
+Каждый xml-файл включает в себя информацию о множестве (обычно ровно 900) субъектов МСП.
+Один субъект МСП в xml-файле представлен элементом «Документ».
+Отдельные характеристики субъекта МСП представлены вложенными элементами, атрибутами и значениями в элементе «Документ».
+К числу таких характеристик относятся ИНН, ОГРН, адрес регистрации, наименование или имя, коды ОКВЭД, вид и категория субъекта МСП, дата включения в реестр и другие.
+
+На момент подготовки статьи (декабрь 2023 г.) было доступно `r raw_files_count` zip-архивов.
+Их суммарный объём — `r format(raw_files_size_gb, digits = 0)` Гб в сжатом виде и `r format (raw_files_uncompressed_size_tb, digits = 2)` Тб в распакованном.
+Из-за большого объёма исходных данных требуется оптимизировать алгоритм их обработки и использовать специализированные инструменты для работы с «большими данными», о которых будет сказано ниже.
+При этом проделанные оптимизации позволяют сделать все необходимые вычисления на обычном персональном компьютере, не задействуя облачные платформы.
+
+## Вспомогательные источники: доходы, расходы, число работников, геоданные
+
 Второй, вспомогательный набор — это «Сведения о суммах доходов и расходов по данным бухгалтерской (финансовой) отчетности организации».

### 2023-12-12 StatsHelper.R

@@ -1,0 +1,16 @@
+raw_data_path <- "../tax-service-opendata/rsmp/xml"
+
+raw_files <- list.files(raw_data_path, pattern = "*.zip", full.names = TRUE)
+raw_files_count <- length(raw_files)
+raw_files_count
+
+raw_files_size_gb <- sum(file.size(raw_files)) / 2 ** 30
+raw_files_size_gb
+
+raw_files_uncompressed_size_tb <- sum(sapply(
+  raw_files,
+  function(path) {
+    sum(unzip(path, list = TRUE)$Length)
+})) / 2 ** 40
+raw_files_uncompressed_size_tb
+prettyNum(raw_files_uncompressed_size_tb, digits = 2)

### 2023-12-14 Paper.Rmd
Write about auxiliary sources, start describing methods

@@
 
-Второй, вспомогательный набор — это «Сведения о суммах доходов и расходов по данным бухгалтерской (финансовой) отчетности организации».
-Технически он также представляет собой коллекцию zip-архивов xml-файлов.
-Каждый архив представляет собой выгрузку на определённую дату, содержит несколько тысяч xml-файлов, каждый из которых, в свою очередь, содержит информацию о нескольких сотнях организаций.
-Данные обновляются с разной периодичностью, но фактическая частота обновления — ежегодно: если в течение года публикуется несколько выгрузок, то все они содержат данные за предыдущий год и отличаются лишь в небольшой части записей, что, вероятно, представляет собой исправление ошибок в ранее размещённых выгрузках.
-В момент подготовки статьи были доступны выгрузки с 2019 г.
-В текущий момент выгрузки за предыдущие даты исчезли с сайта ФНС России, и причины этого непонятны.
-Третий набор, также вспомогательный — «Сведения о среднесписочной численности работников».
-Его структура и периодичность обновления аналогичны предыдущему.
-В момент подготовки статьи были доступны выгрузки с 2019 г.
-В текущий момент выгрузки за предыдущие даты исчезли с сайта ФНС России, и причины этого непонятны.
-В каждом из трёх наборов указан ИНН организации или индивидуального предпринимателя. Поскольку ИНН является уникальным идентификатором, то он позволяет связать сведения из трёх наборов данных.
-Открытые данные по определению являются официальными и свободно распространяемыми, что позволяет рассматривать их как достаточно надёжный и удобный для использования источник информации.
-Тем не менее, количество данных и формат их распространения порождают необходимость использовать специальные методы обработки, о которых говорится далее.
-Общий алгоритм обработки данных
-Шаг 1.
-Загрузка данных из источника на локальный компьютер или в облачное хранилище.
-Смысл этапа — получить собственную копию исходных данных, которая будет обрабатываться.
-Этот шаг выполняет две функции: во-первых, ускоряет дальнейшую обработку за счёт того, что исходные архивы уже загружены на устройство, где они будут обрабатываться, или в облачное хранилище, где передача данных между сервисом хранения и сервисом вычисления происходит быстро; во-вторых, страхует от внезапного исчезновения исходных архивов и позволяет проводить обработку данных даже в том случае, если они будут сняты с публикации, как случилось с двумя вспомогательными наборами.
-Под облачным хранилищем может пониматься как пользовательский сервис хранения данных наподобие «Яндекс-диска», так и специализированное объектное хранилище с S3-совместимым интерфейсом наподобие Object Storage Яндекс-облака.
-Шаг 2.
-Фильтрующее извлечение данных.
-Смысл этапа — извлечь информацию из архивов, выбирая при этом только те сведения, которые необходимы конечному пользователю данных.
-Чтобы понять преимущество данного этапа, нужно сравнить его с более классическим алгоритмом.
-Более классический вариант работы с xml-файлами, запакованными во множество архивов, выглядел бы следующим образом: извлечение xml-файлов из архива — конвертация xml-файлов в табличный формат — сохранение табличных данных — фильтрация табличных данных.
-Несмотря на «каноничность» такого решения, у него есть существенный недостаток: оно приводит к формированию большого количества промежуточных файлов, которые занимают большой объём хранилища, но не имеют практической ценности.
-При небольшом объёме исходных данных проблема промежуточных файлов несущественна, однако, как говорилось выше, объём первого, самого крупного набора данных в распакованном виде оценивается в несколько терабайт, и примерно столько же будут занимать промежуточные файлы.
-Как следствие, для реализации алгоритма требуется либо компьютер, выходящий за рамки стандартного потребительского класса, либо аренда облачных ресурсов.
-И то, и другое дорого и неоптимально с точки зрения результата.
-Фильтрующее извлечение позволяет распаковывать и извлекать данные «на лету», сводя к минимуму образование промежуточных файлов.
-Сохраняются при этом только те сведения, которые необходимы конечному пользователю.
-В частности, можно отфильтровать данные по коду ОКВЭД или субъекту федерации.
-Итоговое табличное представление данных является более компактным, чем исходный набор xml-файлов, примерно в 10–20 раз, а при фильтрации — и более.
+В качестве вспомогательных источников информации используются ещё два реестра открытых данных ФНС России и два набора данных о населённых пунктах России.
+
+1.
+Набор открытых данных ФНС России «Сведения о суммах доходов и расходов по данным бухгалтерской (финансовой) отчетности организаций».
+Содержит ИНН, суммарные доходы и расходы за год, предшествующий году публикации.
+Обновляется несколько раз в год, но последующие обновления в течение одного года, по-видимому, представляют собой уточнение ранее опубликованных сведений, поэтому практическая периодичность обновления — это год.
+Данные доступны с 2019 г.
+по настоящее время[^1].
+В отличие от реестра МСП, содержит только информацию об органзациях, но не об индивидуальных предпринимателях.
+Технически аналогичен реестру МСП: представляет собой временной ряд zip-архивов xml-файлов, где каждый архив соответствует определённой дате, а xml-файл внутри архива содержит информацию о множестве организаций.
+
+[^1]: На момент подготовки статьи на сайте ФНС России была опубликована только последняя версия этого и следующего набора данных.
+Раньше были доступны предыдущие.
+Они сохранились в личном архиве автора и доступны по запросу.
+
+2.
+Набор открытых данных ФНС России «Сведения о среднесписочной численности работников организации».
+Содержит ИНН и среднесписочную численность работников за год, предшествующий году публикации сведений.
+По остальным характеристикам идентичен предыдущему.
+
+3.
+Набор данных «Населенные пункты России: численность населения и географические координаты».
+Подготовлен проектом «Инфраструктура научно-исследовательских данных» на базе открытых данных Росстата и сведений региональных медицинских информационно-аналитических центров.
+Содержит перечень городских и сельских населённых пунктов с указанием субъекта федерации и муниципального образования, кода ОКТМО и географических координат.
+
+4.
+Набор данных «Города России».
+Подготовлен компанией «Дадата» на базе собственных разработок.
+Содержит перечень городских населённых пунктов с информацией о их месте в административно-территориальном делении, коде ОКТМО и географических координатах.
+Поскольку некоторые населённые пункты в нём отсутствуют, автор статьи добавил их вручную, опираясь на сведения Федеральной информационной адресной системы (ФИАС).
+
+Все перечисленные выше источники данных являются общедоступными и бесплатными для использования.
+
+# Методика обработки данных
+
+Предложенный алгоритм обработки данных состоит из шести шагов.
+
+## Шаг 1: загрузка данных ФНС России
+
+Наборы открытых данных загружаются с сайта ФНС России на устройство пользователя.
+Загрузка ускоряет дальнейшую обработку.
+Кроме того, благоря ей создаётся резервная копия исходных данных, что страхует от риска удаления наборов с сайта налоговой службы.
+Теоретически возможна загрузка и обработка данных в облачном сервисе, таком как Yandex Cloud, но в текущей реализации алгоритма она не представлена.
+
+# Шаг 2а: фильтрующее извлечение данных реестра МСП
+
+Каждый zip-архив преобразуется в csv-таблицу.
+Каждая строка таблицы содержит сведения об одном субъекте МСП.
+Извлекаются следующие характеристики: вид, категория, ИНН, ОГРН, имя или наименование, дата включения в реестр, дата выгрузки сведений в набор открытых данных, юридический адрес (с точностью до населённого пункта), основной код ОКВЭД.
+При извлечении данные могут фильтроваться по одному или нескольким кодам или группе кодов ОКВЭД — тогда таблица будет содержать сведения только о части субъектов МСП.
+
+Реализация данного шага оптимизирована, чтобы уменьшить время обработки, расход оперативной памяти и места для хранения временных файлов.
+Наивный подход к преобразованию исходных архивов в таблицу выглядел бы следующим образом: извлечь xml-файлы из архива, прочитать их полностью, сконвертировать в таблицу, выбрать необходимые признаки и наблюдения.
+Он привёл бы к образованию большого объёма временных файлов (десятки гигабайт) и увеличению времени на их обработку (чтение и конвертация в таблицу осуществлялись бы для всех данных).
+Оптимизированный подход, реализованный автором, предполагает чтение отдельных xml-файлов из архива с распаковкой «на лету» (вместо распаковки всего архива), фильтрацию по коду ОКВЭД в процессе обработки xml-файла (вместо последующей фильтрации) и извлечение нужных признаков напрямую из xml-файла (вместо последующего отбора).
+
+## Шаг 2б: извлечение данных о доходах и расходах
+
+Каждый zip-архив набора открытых данных о доходах и расходах преобразуется в csv-таблицу.
+Строка таблицы соответствует одному субъекту МСП.
+Извлекаются следующие признаки: ИНН, дата выгрузки сведений, доходы, расходы.
+Фильтрация не производится.
+Реализация оптимизирована: архивы читаются напрямую без распаковки на диск, из xml-файлов сразу берутся необходимые признаки.
+
+## Шаг 2в: извлечение данных о среднесписочной численности работников
+
+Аналогичен шагу 2б, отличается входными данными (архивы из набора о среднесписочной численности работников) и признаками в результирующей таблице (ИНН, дата выгрузки, среднесписочная численность работников).
+
 Шаг 3.

### 2023-12-18 Paper.Rmd
Finish methods, start testing section

@@
 
-Шаг 3.
-Агрегация данных.
-Смысл этапа — построить хронологию на основе повторяющихся записей.
-Исходные данные представляют собой выгрузки с определённой периодичностью.
-За время, прошедшее между выгрузками, информация о конкретной организации или конкретном индивидуальном предпринимателе могла остаться прежней, в результате чего в табличном представлении может остаться несколько строк с одинаковыми данными, различающиеся только временной меткой.
-Суть агрегации в том, чтобы заменить несколько повторяющихся строк одной, для которой будет указано время актуальности.
-Поскольку проверка того, изменилась ли информация или нет, должна учитывать потенциальные несущественные технические изменения и ошибки в исходных данных, то агрегация также включает в себя базовую подготовку данных: приведение регистра, проверку написания кодов и номеров и тому подобное.
-Шаг 4.
-Геопривязка данных.
-Смысл этапа — для каждой организации или каждого индивидуального предпринимателя указать адрес, координаты и код ОКТМО.
-Исходные данные содержат информацию об адресе согласно Федеральной информационной адресной системе исходя из административно-территориального деления: субъект федерации, район, город, населённый пункт.
-В процессе геопривязки адреса нормализуются.
-Исходя из нормализованного адреса, указывается код ОКТМО и географические координаты.
-Шаг 5.
-Обогащение данных.
-На этом этапе базовая информация об организациях и индивидуальных предпринимателях дополняется вспомогательной.
-В качестве источника вспомогательной информации используются второй и третий наборы данных.
-Опционально возможно использовать и другие источники данных.
-Ключевым условием является наличие в дополнительных источниках данных ИНН, поскольку именно он является уникальным идентификатором.
-Шаг 6 (опциональный).
-Подготовка панельного представления.
-Суть этапа — преобразовать полученные данные в более привычный формат.
-После прохождения описанных выше пяти этапов данные представлены в виде строк, каждая из которых содержит сведения о хозяйствующем субъекте за определённый период времени, причём период времени может быть разным.
-При этом начало и конец периода указаны с точностью до месяца и года.
-В географических исследованиях более применимы данные, которые характеризуют положение дел на какой-то год.
-Таким образом, в ходе данного шага данные преобразуются так, что для каждой организации или индивидуального предпринимателя за каждый год указывается преобладающее значение той или иной характеристики.
-Например, если в течение 10 месяцев в году организация находилась в Москве, а остальные два месяца — в Московской области, то считается, что весь год организация находилась в Москве.
-Реализация алгоритма
-Описанный выше алгоритм реализован как несколько скриптов на языке программирования Python.
-Для хранения данных использовался локальный компьютер.
-Для извлечения данных использовались стандартные модули Python и библиотека lxml.
-Агрегация данных производилась с помощью системы для обработки больших данных Spark.
-Геопривязка осуществлялась с помощью специально разработанного алгоритма на базе двух справочных таблиц: одной по городам России, другой по населённым пунктам России (преимущественно сельским).
-Обогащение данных и создание панели также реализованы как скрипты на Python.
-Исходный код опубликован по адресу: ...
-Валидация данных
-Включает в себя две стадии.
-Первая — это формальная валидация: проверка типа данных и заполненности разных полей итоговой таблицы.
-Вторая — это содержательная валидация.
-Очевидно, что реестр субъектов малого и среднего предпринимательства не содержит информации обо всех организациях в России и представляет собой выборку из него, причём в общем смысле не репрезентативную.
-Тем не менее, есть основания полагать, что выборка всё же является достаточно репрезентативной в отношении географического положения хозяйствующих субъектов и видов деятельности.
-Соответственно, на второй стадии валидации важно сравнить получившийся набор данных с генеральной совокупностью.
-В качестве генеральной совокупности используется статистический регистр Росстата.
-Для сравнения берётся сводная информация о количестве организаций в разрезе субъектов федерации и кодов ОКВЭД за 2019–2022 гг.
-Поскольку при географических исследованиях важным зачастую является не точное совпадение абсолютных чисел (которое в данном случае недостижимо), а совпадение относительного положения тех или иных мест или категорий, то в качестве метрики для оценки используется ранговый коэффициент корреляции Спирмена в разрезе регионов или кодов ОКВЭД.
-Результаты
-Для тестирования алгоритм был запущен со следующими параметрами:
-1) временной диапазон входных данных — с августа 2016 г. по ноябрь 2023 г. (все доступные на сайте ФНС России данные);
-2) группа ОКВЭД — «А» (сельское, лесное хозяйство, охота, рыболовство, рыболовство).
-Она объединяет несколько отраслей экономики, которые в экономическое географии обычно рассматриваются изолированно.
-Тем не менее, поскольку при тестировании важно показать принцип работы алгоритма (извлечение по группам или кодам ОКВЭД), то было отдано предпочтение именно классификатору, а не группировке, принятой в научной сфере.
-Кроме того, такой вариант запуска показывает, что алгоритм позволяет одновременно извлечь сведения о нескольких связанных отраслях экономики.
-Сводная статистика по работе алгоритма
+## Шаг 3а: агрегация данных о субъектах МСП
+
+Таблицы в csv-формате, получившиеся на шаге 2а, сводятся в одну таблицу, из которой удаляются строки, в которых совпадает набор значимых признаков: ИНН, ОГРН, имя или наименование, вид, категория, код ОКВЭД, адрес регистрации.
+После удаления дублей в каждой строке простравляются две даты: дата начала действия и дата окончания действия.
+Эти даты указывают на период, в течение которого актуальна информация, содержащаяся в строке.
+Например, если организация с 10 августа 2016 г.
+по 10 марта 2021 г.
+была известна под одним наименованием, а с 10 апреля 2021 г.
+— под другим, то в агрегированной таблице она будет представлена двумя строками.
+У первой дата начала — 10.08.2016, дата окончания — 10.03.2021; у второй дата начала — 10.04.2021, дата окончания — 10.11.2023 (условная дата последнего архива исходных данных).
+Кроме того, на данном этапе происходят вспомогательные преобразования, цель которых — исправить неточности в исходых данных и более точно определить дублирующиеся строки.
+Например, в реестре МСП тип населённого пункта может быть указан полностью («город») или сокращённо («г.»), и вспомогательная обработка перед агрегацией позволяет игнорировать такие различия и воспринимать адрес как неизменный.
+
+## Шаг 3б: агрегация данных о доходах, расходах
+
+Таблицы в csv-формате, получившиеся на шаге 2б, объединяются в одну так, что для каждой организации указаны доходы и расходы за конкретный год.
+Объединённая таблица фильтруется по ИНН так, что в ней остаются только те организации, которые есть в агрегированных данных из реестра МСП, полученных на шаге 3а.
+Результат сохраняется в итоговый csv-файл.
+
+## Шаг 3в: агрегация данных о работниках
+
+Аналогичен шагу 3б.
+
+## Шаг 4: Геопривязка субъектов МСП
+
+Каждая запись в csv-таблице, полученной на этапе 3а, дополняется кодом ОКТМО и географическими координатами населённого пункта, в котором находится субъект МСП.
+Для этого используется адрес регистрации и вспомогательные наборы геоданных.
+Кроме того, вместо исходного адреса в наборе данных появляется нормализованный, а исходный адрес в виде структурированной текстовой строки сохраняется как отдельный атрибут каждой записи.
+После нормализации адресов набор данных повторно проверяется на дубли, и вновь образовавшиеся дубли удаляются, а даты актуальности сведений пересчитываются.
+
+## Шаг 5: создание панельного представления
+
+Таблица, полученная на этапе 4, преобразуется в панельный временной ряд, где представлены характеристики каждого субъекта МСП за каждый год с 2016 по 2023.
+Содержательно новая таблица не отличается от старой, но более удобна для анализа, поскольку вместо компактного, но не слишком интуитивного и весьма технологизированного формата с датами актуальности сведения в ней идут по годам.
+Такой вид более привычен для исследователей, а также упрощает объединение набора данных с официальной статистикой, которая также обычно даётся в разрезе календарных лет.
+
+## Реализация алгоритма
+Описанный выше алгоритм реализован как набор скриптов на языке программирования Python.
+Чтение xml осуществляется с помощью высокоэффективной библиотеки lxml.
+Для агрегации используется система обработки больших данных Apache Spark.
+Исходный код опубликован по адресу: https://github.com/PavelSyomin/russian-legal-companies
+
+# Тестирование
+
+Для тестирования с помощью представленной технологии был подготовлен набор данных по субъектам МСП, у которых основной код ОКВЭД находится в группе А — сельское, лесное хозяйство, охота, рыболовство, рыболовство.
+Использовались исходные данные за весь доступный период.
+
+Время работы алгоритма составило около 15 часов, из которых подавляющую часть заняло извлечение информации из архивов (этап 2а).
+
 Примерное время подготовки набора данных составило 15 часов, вычисления выполнялись на обычном персональном компьютере с примерно 300 Гб свободного места на диске и 16 Гб оперативной памяти.
@@
 На схемах 6.1–6.3 показано распределение соответственно числа организаций, их суммарной прибыли и числа работников в сфере «сельское хозяйство».
+
+Валидация данных
+Включает в себя две стадии.
+Первая — это формальная валидация: проверка типа данных и заполненности разных полей итоговой таблицы.
+Вторая — это содержательная валидация.
+Очевидно, что реестр субъектов малого и среднего предпринимательства не содержит информации обо всех организациях в России и представляет собой выборку из него, причём в общем смысле не репрезентативную.
+Тем не менее, есть основания полагать, что выборка всё же является достаточно репрезентативной в отношении географического положения хозяйствующих субъектов и видов деятельности.
+Соответственно, на второй стадии валидации важно сравнить получившийся набор данных с генеральной совокупностью.
+В качестве генеральной совокупности используется статистический регистр Росстата.
+Для сравнения берётся сводная информация о количестве организаций в разрезе субъектов федерации и кодов ОКВЭД за 2019–2022 гг.
+Поскольку при географических исследованиях важным зачастую является не точное совпадение абсолютных чисел (которое в данном случае недостижимо), а совпадение относительного положения тех или иных мест или категорий, то в качестве метрики для оценки используется ранговый коэффициент корреляции Спирмена в разрезе регионов или кодов ОКВЭД.
+
 Обсуждение

### 2023-12-19 Paper.Rmd
Continue writing testing (demo) section: tech stats

@@
 knitr::opts_chunk$set(include = FALSE)
+options(knitr.kable.NA = "—")
 ```
@@
 
-В статье предложена технология, позволяющая формировать геопривязанный табличный набор данных о малых и средних фирмах в России.
+В статье предложена технология (алгоритм), позволяющая формировать геопривязанный табличный набор данных о малых и средних фирмах в России.
 В качестве основы используется реестр субъектов малого и среднего предпринимательства (реестр МСП), размещённый в форме открытых данных на сайте Федеральной налоговой службы России (ФНС России).
@@
 
-# Тестирование
+# Демонстрация работы технологии
 
-Для тестирования с помощью представленной технологии был подготовлен набор данных по субъектам МСП, у которых основной код ОКВЭД находится в группе А — сельское, лесное хозяйство, охота, рыболовство, рыболовство.
-Использовались исходные данные за весь доступный период.
+С помощью описанного алгоритма был подготовлен набор данных по субъектам МСП, ведущих деятельность в сфере сельского, лесного хозяйства, охоты, рыболовства и рыбоводства.
+Для этого использовалось следующее условие фильтрации: основной код ОКВЭД организации или предпринимателя находится в группе А классификатора в редакции 2014 г.
+На «вход» алгоритм принимал три набора данных ФНС России за весь доступный на момент подготовки статьи период: с августа 2016 г.
+для реестра МСП, с октября 2019 г.
+для сведений о доходах и расходах организаций; с марта 2020 г.
+для сведений о среднесписочной численности работников организаций (дата окончания везде — ноябрь 2023 г.).
+Результат работы алгоритма — несколько csv-файлов.
+Далее в подразделах представлена техническая статистика выполнения алгоритма с акцентом на изменение объёма обрабатываемых данных и описание базовых возможностей по анализу полученных наборов данных с упором на визуализацию, дающую наглядное представление о перспективах применения технологии как инструмента создания наборов данных для экономико-географических исследований.
 
-Время работы алгоритма составило около 15 часов, из которых подавляющую часть заняло извлечение информации из архивов (этап 2а).
+## Техническая статистика работы алгоритма
 
-Примерное время подготовки набора данных составило 15 часов, вычисления выполнялись на обычном персональном компьютере с примерно 300 Гб свободного места на диске и 16 Гб оперативной памяти.
-На схеме показано уменьшение количества обрабатываемых данных от шага к шагу алгоритма.
-Входные данные — это 88 архивов суммарным объёмом 193 Гб в сжатом виде, содержащие около 600 тысяч xml-файлов и 550 миллионов исходных записей.
-После фильтрующего извлечения промежуточный набор данных состоял из 88 csv-файлов объёмом 6 Гб и суммарно содержал 15 миллионов строк.
-В результате агрегации получился один датасет (csv-файл) объёмом 133 Мб, содержащий 500 тысяч строк.
-Геопривязанный датасет содержит почти столько же строк (их количество немного уменьшилось из-за того, что часть адресов после нормализации оказалась дублями) и занимает чуть больше 200 Мб (объём вырос из-за добавления нормализованных адресов к «сырым»).
-Панельное представление также является одним csv-файлов объёмом около 600 Мб, содержащим несколько миллионов строк (панельное представление менее компактно, так как данные, не меняющиеся год от года, дублируются для сохранения временной структуры).
-Таким образом, в результате обработки объём данных уменьшился с нескольких терабайт до нескольких сотен мегабайт (с промежуточными файлами — нескольких гигабайт), и из полного списка организаций, содержащего несколько сотен миллионов записей, было выбрано несколько сотен тысяч интересующих.
-Кроме того, данные обогатились геопривязкой и кодами ОКТМО.
-В результате получился компактный датасет, который можно обработать на обычном компьютере без установки специальных программ для работы с «большими данными» — в отличие от исходного набора данных на сайте ФНС России.
-Итоговый датасет содержит информацию о 300 тысячах индивидуальных предпринимателей и 100 тысячах огранизаций, осуществляющих или осуществлявших деятельность в сфере сельского, лесного хозяйства, охоты, рыболовства и рыбоводства.
-Он позволяет искать, группировать, агрегировать и анализировать данные в нескольких разных аспектах.
-Приведём основные из них.
+Время работы алгоритма — около 15 часов. Больше всего заняло извлечение данных из архивов (этап 2).
+
+В таблице 1 приведена информация об уменьшении объёма данных в процессе работы алгоритма, а в таблице 2 — о фильтрации записей.
+
+```{r sizes_funnel}
+bytes_in_mb <- 2 ** 20
+sizes_funnel <-  sizes_funnel / bytes_in_mb
+knitr::kable(
+  sizes_funnel,
+  digits = 0,
+  format.args = list(scientific = FALSE),
+  col.names = c("Реестр МСП", "Доходы+расходы", "Работники", "Всего"))
+```
+
+```{r counts_funnel}
+counts_funnel <- counts_funnel / 1000 # in thousands
+knitr::kable(
+  counts_funnel,
+  digits = 0,
+  format.args = list(scientific = FALSE),
+  col.names = c("Реестр МСП", "Доходы+расходы", "Работники"))
+```
+
+Из таблиц следует, что технология многократно уменьшает объём обрабатываемых сведений: от сотен гигабайт исходных до нескольких десятков или сотен мегабайт итоговых наборов данных.
+Одновременно уменьшается и количество записей в данных с сотен миллионов до сотен тысяч.
+Таким образом, вместо десятков объёмных архивов с исходными данными в многомерном xml-формате исследователь может работать с относительно привычной и компактной двумерной таблицей.
+Итоговые таблицы имеют небольшой объём и поэтому могут обрабатываться без инструментов анализа больших данных.
+Панельное представление слегка теряет в компактности по сравнению с итоговыми таблицами, но выигрывает благодаря более удобной форме представления временного ряда.
+
+Итоговый набор данных содержит информацию примерно о `r round(out_org_count / 1000)` тысячах организаций и `r round(out_ind_count / 1000)` тысячах индивидуальных предпринимателей.
+
 Временная динамика

### 2023-12-19 StatsHelper.R

@@ -1,4 +1,4 @@
-raw_data_path <- "../tax-service-opendata/rsmp/xml"
+rsmp_raw_data_path <- "../tax-service-opendata/rsmp/xml"
 
 raw_files <- list.files(raw_data_path, pattern = "*.zip", full.names = TRUE)
 raw_files_count <- length(raw_files)
@@ -13,4 +13,84 @@ raw_files_uncompressed_size_tb <- sum(sapply(
     sum(unzip(path, list = TRUE)$Length)
 })) / 2 ** 40
 raw_files_uncompressed_size_tb
-prettyNum(raw_files_uncompressed_size_tb, digits = 2)
+
+# Algorithm execution stats for group A demonstration
+rsmp_raw_data_path <- "../tax-service-opendata/rsmp/xml"
+revexp_raw_data_path <- "../tax-service-opendata/revexp/xml"
+sshr_raw_data_path <- "../tax-service-opendata/sshr/xml"
+rsmp_source_files <- list.files(rsmp_raw_data_path, pattern = "*.zip", full.names = TRUE)
+revexp_source_files <- list.files(revexp_raw_data_path, pattern = "*.zip", full.names = TRUE)
+sshr_source_files <- list.files(sshr_raw_data_path, pattern = "*.zip", full.names = TRUE)
+rsmp_source_files_size <- sum(file.size(rsmp_source_files))
+revexp_source_files_size <- sum(file.size(revexp_source_files))
+sshr_source_files_size <- sum(file.size(sshr_source_files))
+
+rsmp_intermediate_data_path <- "../tax-service-opendata/rsmp/reestr_group_A/"
+revexp_intermediate_data_path <- "../tax-service-opendata/revexp/csv"
+sshr_intermediate_data_path <- "../tax-service-opendata/sshr/csv"
+rsmp_intermediate_data_files <- list.files(rsmp_intermediate_data_path, pattern = "^data-", full.names = TRUE)
+revexp_intermediate_data_files <- list.files(revexp_intermediate_data_path, pattern = "^data-", full.names = TRUE)
+sshr_intermediate_data_files <- list.files(sshr_intermediate_data_path, pattern = "^data-", full.names = TRUE)
+rsmp_intermediate_data_files_size <- sum(file.size(rsmp_intermediate_data_files))
[…]

### 2023-12-21 Paper.Rmd
Continue results section: plots and maps

@@
 ```{r imports}
+source("Plots.R")
 source("StatsHelper.R")
@@
 
-Временная динамика
-Набор данных охватывает период с 2016 по 2023 г.
-с точностью до месяца.
-Соответственно, он позволяет проследить динамику численности хозяйствующих субъектов в том или ином регионе или той или иной сфере деятельности.
-На графике 1.1 показаны первые 10 регионов с наибольшим количеством организаций в группе А за каждый месяц из анализируемого периода, а на графике 1.2 — изменение числа хозяйствующих субъектов в отдельных укрупнённых сферах экономики: сельское хозяйство, лесное хозяйство, охота, рыболовство и рыбоводство.
-Пространственная структура по субъектам федерации
-Набор данных охватывает все субъекты России, для которых имеются исходные данные.
-Благодаря нормализованному (приведённому в соответствие с Конституцией России) названию субъекта федерации данные можно легко агрегировать и тем самым подсчитать количество хозяйствующих субъектов в каждом регионе в целом или по отдельным кодам ОКВЭД.
-На картограмме 2.1 показано распределение регионов по количеству хозяйствующих субъектов в сфере сельского хозяйства, а на картограмме 2.2 — лесного хозяйства.
-Видно, что география указанных отраслей ожидаемо отличается.
-Глубокая детализация по кодам ОКВЭД позволяет выбрать и визуализировать распределение хозяйствующих субъектов по узким и довольно специфичным кодам.
-Так, картограмма 2.3 показывает распределение организаций с кодом 1.6.22 «Выращивание подсолнечника», а картограмма 2.4 — (что-то про лесное хозяйство).
-Пространственная структура по населённым пунктам
-Набор данных содержит информацию о местоположении хозяйствующих субъектов с точностью до населённого пункта, а также координаты этих пунктов и коды ОКТМО соответствующих муниципальных образований.
-Это позволяет агрегировать данные по населённым пунктам, визуализировать их на картах и схемах, а также связывать со статистикой по муниципальным образованиям.
-На картограмме 3.1 показано распределение организаций по территории России в виде точек, изображающих населённые пункты.
-Картограмма 3.2 представляет собой часть картограммы 1 и показывает число и соотношение видов организаций в г.
-Екатеринбурге и окрестностях.
-В таблице 3.2 приведены результаты простейшего анализа данных во взаимосвязи с внешними ресурсами: количество организаций в городах в пересчёте на тысячу жителей.
-Отслеживание отдельных хозяйствующих субъектов
-Набор данных детилизирован до отдельных организаций и индивидуальных предпринимателей, что позволяет проследить динамику каждого из них, а также агрегировать информацию об изменениях.
-В таблице 4.1 показано число изменений юридического адреса у хозяйствующих субъектов.
-На картограмме 4.2 показаны «регионы-доноры» и «регионы-реципиенты» хозяйствующих субъектов за период с 2017 по 2021 г.
-Анализ наименований
-Набор данных содержит сведения о наименованиях организаций.
-Наименования сами по себе могут быть ценным источником информации, если обрабатывать их с использованием методов автоматизированной обработки текстов.
-На схемах 5.1 и 5.2 показаны «облака слов» из названий организаций в сфере сельского и лесного хозяйства, соответственно.
+## Временная динамика
+
+Набор данных позволяет отследить те или иные количественные характерстики с точностью до месяца.
+На графике ниже показаны 20 субъектов России с наибольшим количеством зарегистрированных субъектов МСП и изменение этого количества с августа 2016 г.
+по ноябрь 2023 г.
+
+```{r time_series_plot}
+time_sereis_plot
+```
+
+## Распределение по субъектам федерации
+
+Нормализованные названия позволяют агрегировать информацию по субъектам федерации.
+На картограмах ниже представлено количество субъектов МСП, которые вели деятельность в сфере лесного (слева) и сельского (справа) хозяйства в 2021 г.
+Распределение ожидаемо разное: наибольшее количество фирм в сфере сельского хозяйства расположено в Черноземье и на юге России, также заметны животноводческие регионы; лесное хозяйство сосредоточено на севере европейской части страны, в Восточной Сибири и на Дальнем Востоке.
+
+```{r regions_agri_forest}
+regions_plot_agri_forestry
+```
+
+В отличие от официальной статистики, получившийся набор данных позволяет проанализировать пространственное распределение с точностью до конкретных видов деятельности, а не крупных агломератов.
+На картограммах ниже представлено количество субъектов МСП, ведущих деятельность по выращиванию пшеницы (код 01.11.11) и картофеля (код 01.13.31).
+Здесь заметны некоторые странности (например, в лидерах по картофелю оказалась Якутия), которые, видимо, связаны с небольшим количеством субъектов МСП в каждом регионе: обычно у фирм в качестве основного указан более общий код, в то время как картограмма строилась по максимально детализированным.
+
+```{r wheat_potato}
+regions_plot_wheat_potato
+```
+
+## Распределение по населённым пунктам
+
+Набор данных, помимо названий субъектов содержит также географические координаты, название населённого пункта и территориальной единицы (района, муниципалитета) внутри субъекта.
+Это позволяет выполнять более детализированные агрегации и картографически визуализировать их.
+На картограмме ниже показано количество субъектов МСП в разных населённых пунктах, а также преимущественная специализация населённых пунктов исходя из того, каких фирм в нём больше всего.
+Показаны только населённые пункты более чем с 50 фирмами.
+Ещё одна аналогичная картограмма изображает количество субъектов МСП по видам деятельности в населённых пунктах Свердловской области, причём без ограничения на количество фирм.
+
+```{r settlements_russia}
+activity_by_settlements
+```
+
+```{r settlements_svr}
+activity_by_settlements_svr
+```
+
+## Отслеживание отдельных фирм
+
+Благодаря детальности набор данных позволяет анализировать изменения в конкретных фирмах, например, юридический «переезд» из одного региона в другой.
+На картограмме показаны «регионы-доноры» и «регионы-реципиенты» субъектов МСП за весь доступный период.
+
+```{r migrations}
+migrations_plot
+```
+
 Дополнительные атрибуты
+
 Расширение набора данных за счёт информации о численности работников организаций и доходах с расходами позволяет перейти от простого подсчёта к «взвешенному» подсчёту и анализировать территориальное, временное или структурное распределение с учётом прибыли и числа занятых.
@@
 
-Валидация данных
+# Валидация данных
+
 Включает в себя две стадии.
@@
 
-Обсуждение
+# Обсуждение
+
 Рассуждая о предложенном алгоритме, необходимо обратить внимание на два его аспекта: техническую ценность и исследовательскую ценность.

### 2023-12-21 Plots.R

@@ -1,0 +1,161 @@
+library(dplyr)
+library(forcats)
+library(ggplot2)
+library(readr)
+library(scales)
+library(sf)
+library(tidyr)
+
+rsmp_data_path <- "../tax-service-opendata/rsmp/reestr_group_A/data_product.csv"
+rsmp_data <- read_csv(rsmp_data_path)
+
+rsmp_data %>% 
+  count(region, start_date, end_date)
+# Regions time series
+monthly_counts <- do.call(rbind, lapply(
+  seq.Date(min(rsmp_data$start_date), max(rsmp_data$end_date), by = "month"),
+  function(x) {
+    filter(rsmp_data, start_date <= x, end_date >= x) %>% 
+      count(region) %>% 
+      mutate(date = x) %>% 
+      arrange(-n)
+  }
+))
+
+time_series_plot <- expand(monthly_counts, region, date) %>% 
+  left_join(monthly_counts, by = c("region", "date")) %>% 
+  mutate(region = fct_reorder2(region, date, n, first2)) %>% 
+  filter(n > 1000) %>% 
+  ggplot(aes(x = date, y = n, color = region)) +
+  geom_line(na.rm = TRUE, show.legend = FALSE) +
+  #scale_x_continuous(limits = c(2016, 2024)) + 
+  scale_y_continuous(trans = "log10") +
+  scale_color_grey(name = "Субъект России") +
+  labs(x = "Год", y = "Количество субъектов МСП") +
+  theme_bw()
+time_series_plot
+
+# Maps
+regions_boundaries <- st_read("geoBoundaries-RUS-ADM1_simplified.geojson")
+ru_crs <- st_crs("+proj=aea +lat_0=0 +lon_0=100 +lat_1=68 +lat_2=44 +x_0=0 +y_0=0 +ellps=krass +towgs84=28,-130,-95,0,0,0,0 +units=m +no_defs")
+regions <- read_csv("../tax-service-opendata/assets/regions.csv")
+
+regions <- regions_boundaries %>% 
+  left_join(regions, by = c("shapeISO" = "iso_code")) %>% 
+  select(name)
+
+ac_code_mapping <- c("01" = "Сельское хозяйство", "02" = "Лесное хозяйство")
+regions_plot_agri_forestry <- rsmp_data %>%
+  mutate(
+    ac_start = substr(activity_code_main, 1, 2),
+    ac_type = ac_code_mapping[ac_start]) %>% 
+  filter(
[…]

### 2023-12-22 Paper.Rmd
One more plot to results, start discussion

@@
 
-Дополнительные атрибуты
+## Анализ прибыли и числа работников
 
-Расширение набора данных за счёт информации о численности работников организаций и доходах с расходами позволяет перейти от простого подсчёта к «взвешенному» подсчёту и анализировать территориальное, временное или структурное распределение с учётом прибыли и числа занятых.
-На схемах 6.1–6.3 показано распределение соответственно числа организаций, их суммарной прибыли и числа работников в сфере «сельское хозяйство».
+Вспомогательные данные о доходах, расходах и числе работников позволяют дополнить анализ, перейдя от простого числа компаний к характеристикам их деятельности.
+На трёх картограммах ниже показано соответственно общее число фирм, их суммарная прибыль и суммарное число работников по регионам.
+Видно, что распределение хотя и похожее, но всё же отличается в зависимости от визуализируемой характеристики.
 
-# Валидация данных
+```{r many_variables_by_region}
+count_profit_employees_by_region_plot
+```
+
+# Обсуждение
+
+Предложенный алгоритм имеет техническую ценность: он позволяет обработать большой объём сложноструктутированных данных с сайта ФНС России и сделать на их основе несколько сравнительно компактных таблиц в формате, который традиционно используется исследователями при анализе данных.
+Кроме того, технология нормализует адреса, содержащиеся в исходных данных, и обогащает их географическими координатами и кодами ОКТМО, тем самым делая их пригодными для обработки в геоинформационных системах (ГИС) и пространственного анализа в целом.
+Для выполнения алгоритма, несмотря на то что в его реализации используется Apache Spark, достаточно персонального компьютера средней конфигурации — использовать облачные вычислительные ресурсы не нужно.
+Таким образом, для конечного пользователя технология представляет собой способ работать с big data, обойдя при этом сопутствующие сложности и повышенные требования к ресурсам и навыкам и с помощью нескольких оптимизаций превратив её в работу с обычными «малыми» данными.
+Благодаря открытому исходному коду и использованию свободного программного обеспечения алгоритм может пригодиться не только пользователям, но и разработчикам как база для создания собственных продуктов.
+
+Набор данных, сформированный с помощью описанной технологии — это информация о генеральной совокупности субъектов МСП в России.
+Его можно использовать для сплошного исследования малого и среднего предпринимательства в стране.
+В некотором роде это квази-источник первичных статистических сведений, которые по общему правилу относятся к информации ограниченного доступа[^1], но в данном случае являются открытыми в силу прямого указания в Налоговом кодексе[^2].
+Кроме того, набор данных является полным геопривязанным перечнем индивидуальных предпринимателей.
+Тем не менее, поскольку ФНС России не публикует сведения о доходах, расходах и числе работников применительно к индивидуальным предпринимателям, исследовательская ценность этих данных меньше, чем информации об организациях.
+Также следует отметить, что анализ сведений об индивидуальных предпринимателях может рассматриваться как обработка персональных данных физических лиц, что создаёт правовые риски, особенно актуальные в свете регулярного ужесточения юридической ответственности за нарушения в этой сфере.
+
+^1: Ч. 1 ст. 9 Федерального закона от 29.11.2007 № 282-ФЗ «Об официальном статистическом учете и системе государственной статистики в Российской Федерации».
+
+^2: П. 1.1 ст. 102 Налогового кодекса России.
 
@@
 
-Рассуждая о предложенном алгоритме, необходимо обратить внимание на два его аспекта: техническую ценность и исследовательскую ценность.
-Говоря об исследовательской ценности, важно обратить внимание на два аспекта: частный (о применимости получаемого датасета для исследований в области малого и среднего предпринимательства) и общий (возможность использования данных для выводов об экономике в целом).
-Техническая ценность предложенного алгоритма связана с тем, что он позволяет достаточно эффективно (без существенных финансовых затрат и за сравнительно короткий период времени) обработать массивы исходных данных ФНС России и представить их в виде, подходящем для географических исследований, то есть плоских таблиц с геопривязкой.
-Благодаря алгоритму и его прикладной реализации несколько терабайт исходной информации фильтруются и компактно представляются в виде относительно небольшой таблицы, которую можно использовать на стандартном компьютере без облачных технологий и инструментов работы с большими данными.
-Программный код находится в открытом доступе и может использоваться любым специалистом, владеющим языком программирования Python, для создания собственных производных решений.
-Исследовательская ценность алгоритма и набора данных в контексте исследований малого и среднего предпринимательства заключается в том, что датасет представляет собой максимально полный и детализированный из доступных набор сведений об организациях и предпринимателях, относящихся к МСП.
-Он объединяет в себе три измерения: пространственное (с точностью до населённого пункта), временное (с точностью до месяца или года) и структурное (с точностью до отдельных кодов ОКВЭД) и представлен в наиболее дезагрегированном виде: в отличие от традиционных статистических данных, исходная единица в нём — это конкретный хозяйствующий субьект.
-Таким образом, набор данных может рассматриваться как квази-источник первичных статистических сведений.
-Поскольку вся представленная в нём информация об организациях представляет собой изъятие из режима налоговой тайны и по закону имеет открытый характер, то использование датасета не нарушает права частных лиц и законодательство России.
-Помимо организаций, набор данных содержит также сведения об индивидуальных предпринимателях.
-Его ценность ниже, так как для них отсутствуют дополнительные данные о доходах, расходах и числе работников.
-Кроме того, если информация о названиях организаций обладает самостоятельной исследовательской ценностью, то имена физических лиц такой ценностью вряд ли характеризуются.
-Кроме того, сведения в датасете относятся к информации об определённом физическом лице, то есть являются персональными данными, что поднимает вопрос о юридических рисках их использования, особенно в свете ужесточения законодательства о персональных данных.
 Вопрос об исследовательской ценности набора данных для анализа экономики России в целом более сложен.

### 2023-12-22 Plots.R

@@ -159,3 +159,38 @@ migrations_plot <- region_changes %>% ungroup() %>%
   coord_sf(crs = ru_crs) +
   scale_fill_distiller(palette = "PiYG", direction = 1)
 migrations_plot
+
+# With revexp and employees
+revexp_data_path <- "../tax-service-opendata/revexp/csv/group_A.csv"
+revexp_data <- read_csv(revexp_data_path)
+sshr_data_path <- "../tax-service-opendata/sshr/csv/group_A.csv"
+sshr_data <- read_csv(sshr_data_path)
+
+count_profit_employees_by_region_plot <- rsmp_data %>%
+  filter(
+    start_date <= "2021-12-31",
+    end_date >= "2021-12-31",
+    substr(activity_code_main, 1, 2) == "01"
+  ) %>% 
+  left_join(filter(revexp_data, year == 2021), by = "tin") %>% 
+  left_join(filter(sshr_data, year == 2021), by = "tin") %>% 
+  mutate(profit = revenue - expediture) %>% 
+  select(tin, region, profit, employees_count) %>% 
+  group_by(region) %>% 
+  summarise(
+    count = n(),
+    profit = sum(profit, na.rm = TRUE),
+    employees = sum(employees_count, na.rm = TRUE)
+  ) %>% 
+  pivot_longer(-region, names_to = "option", values_to = "value") %>% 
+  right_join(regions, by = c("region" = "name")) %>% 
+  st_as_sf() %>% 
+  group_by(option) %>% 
+  mutate(value = rescale(value)) %>% 
+  ungroup() %>% 
+  ggplot() +
+  geom_sf(aes(fill = value), size = .1) +
+  coord_sf(crs = ru_crs) +
+  facet_wrap(~option, ncol = 2) +
+  scale_fill_distiller(palette = "YlGn", direction = 1)
+count_profit_employees_by_region_plot

### 2023-12-25 Paper.Rmd
Finish discussion

@@
 Предложенный алгоритм имеет техническую ценность: он позволяет обработать большой объём сложноструктутированных данных с сайта ФНС России и сделать на их основе несколько сравнительно компактных таблиц в формате, который традиционно используется исследователями при анализе данных.
-Кроме того, технология нормализует адреса, содержащиеся в исходных данных, и обогащает их географическими координатами и кодами ОКТМО, тем самым делая их пригодными для обработки в геоинформационных системах (ГИС) и пространственного анализа в целом.
+Кроме того, технология нормализует адреса, содержащиеся в исходных данных, и обогащает их географическими координатами и кодами ОКТМО, тем самым делая их пригодными для обработки в геоинформационных системах (ГИС) и пространственного анализа в целом, а также упрощая объединение с муниципальной статистикой.
 Для выполнения алгоритма, несмотря на то что в его реализации используется Apache Spark, достаточно персонального компьютера средней конфигурации — использовать облачные вычислительные ресурсы не нужно.
@@
 
-Включает в себя две стадии.
-Первая — это формальная валидация: проверка типа данных и заполненности разных полей итоговой таблицы.
-Вторая — это содержательная валидация.
-Очевидно, что реестр субъектов малого и среднего предпринимательства не содержит информации обо всех организациях в России и представляет собой выборку из него, причём в общем смысле не репрезентативную.
-Тем не менее, есть основания полагать, что выборка всё же является достаточно репрезентативной в отношении географического положения хозяйствующих субъектов и видов деятельности.
-Соответственно, на второй стадии валидации важно сравнить получившийся набор данных с генеральной совокупностью.
-В качестве генеральной совокупности используется статистический регистр Росстата.
-Для сравнения берётся сводная информация о количестве организаций в разрезе субъектов федерации и кодов ОКВЭД за 2019–2022 гг.
-Поскольку при географических исследованиях важным зачастую является не точное совпадение абсолютных чисел (которое в данном случае недостижимо), а совпадение относительного положения тех или иных мест или категорий, то в качестве метрики для оценки используется ранговый коэффициент корреляции Спирмена в разрезе регионов или кодов ОКВЭД.
+Вопрос о том, можно ли использовать технологию и создаваемый с её помощью набор данных для изучения не только малого и среднего бизнеса, но и экономики в целом, является сложным.
+Как отмечалось в обзоре литературы, потребность в данных, детализированных до уровня отдельных фирм, существует.
+Поскольку получить такие данные сложно или дорого, возникает желание использовать сведения о субъектах МСП в качестве заместителя (прокси), особенно при математическом моделировании.
+Это может показаться разумным, ведь субъекты МСП составляют почти 98% в общем числе фирм в России, дают 43% ВВП и обеспечивают занятостью 38% работников [@Barinova2019].
+В то же время очевидно, что реестр МСП не является случайной выборкой из генеральной совокупности хозяйствующих субъектов, поскольку его элементы отобраны исходя из жёстких лимитов, ограничивающих максимальный доход и число работников.
+Соответственно, реестр МСП даёт информацию только о небольших фирмах, и выводы, полученные на его основе, нельзя распространить на все компании в России.
+Во всяком случае, такой вывод следует, если оценивать репрезентативность реестра МСП исходя из процедурного критерия, то есть рассуждать теоретически.
+Возможен, однако, и альтернативный эмпирический путь: чтобы понять, является ли реестр МСП достаточно хорошим «прокси» для генеральной совокупности фирм, надо сравнить распределение фирм в нём с распределением в генеральной совокупности.
+Обычно при анализе репрезентативности сделать такое сравнение невозможно, поскольку характеристики генеральной совокупности неизвестны.
+В случае с фирмами в России ситуация другая: статистический регистр Росстата представляет собой относительно полный источник сведений о генеральной совокупности.
 
-# Обсуждение
+Для оценки репрезентативности реестра МСП можно использовать корреляционный анализ.
+Единая медведомственная информационно-статистическая система (ЕМИСС) содержит данные о числе организаций в разрезе регионов и групп ОКВЭД за 2019–2022 гг[^1].
+Такая же статистика рассчитывается на базе реестра МСП.
+В результате для каждого года, каждого региона и каждой группы ОКВЭД известно два числа: общее количество организаций и количество субъектов МСП.
+Далее вычисляются:
+- коэффециенты корреляции в каждом году;
+- коэффициенты корреляции в каждом регионе;
+- коэффициенты корреляции в каждой группе ОКВЭД;
+Используется ранговый коэффициент корреляции Спирмена, так как он более устойчив к выбросам, чем коэффициент корреляции Пирсона.
 
-Вопрос об исследовательской ценности набора данных для анализа экономики России в целом более сложен.
-С одной стороны, очевидно, что субъекты малого и среднего предпринимательства — это не все хозяйствующие субъекты в России.
-Более того, понятно также, что они не являются случайной выборкой из генеральной совокупности таких субъектов, поскольку отобраны по конкретным критериям: число работников и доход.
-Иначе говоря, реестр и производный от него датасет содержат сведения только о части фирм, причём о самых небольших из них, а крупный бизнес при этом выпадает из поля зрения.
-Тем не менее, есть некоторые доказательства, что как минимум отчасти такой датасет может быть репрезентативным в географическом и структурном отношении.
-Как показано в разделе с валидацией, агрегированная статистика по регионам и группам деятельности, полученная на базе датасета, высоко скоррелирована с данными, полученными на базе статистического регистра Росстата, который является генеральной совокупностью.
-Коэффициент корреляции не очень высокий разве что в отдельных специфических группах, что объяснимо (большая часть образовательных организаций и организаций в сфере публичной власти — не частные, а государственные или муниципальные), в остальных случаях он превышает 0.9.
-Иначе говоря, датасет на базе реестра МСП (как выборка) хорошо скоррелирован с генеральной совокупностью в географическом (с точностью до регионов) и структурном (с точностью до большинства групп ОКВЭД) разрезах.
-Это позволяет говорить о его репрезентативности как минимум в этих двух аспектах.
-Так, при любом ранжировании регионов или видов деятельности результат, полученный на базе реестра МСП, будет в основном совпадать с результатом, полученным на базе сплошного обследования всех фирм в России.
-Это не говорит о том, что датасет репрезентативен в более «узких» разрезах (на уровне отдельных кодов ОКВЭД или конкретных населённых пунктов), но в отсутствие иных источников информации позволяет использовать его в качестве первого — и наиболее адекватного из доступных — приближений к генеральной совокупности в тех случаях, когда требуются данные по экономическим субъектам в России с высокой детализацией.
-Тем не менее, вопрос о репрезентативности желательно при каждом конкретном исследовании анализировать отдельно.
-Два сопутствующих вопроса, касающихся исследовательской ценности в любом аспекте, касаются соответствия между номинальным и реальным положением дел.
-В реестре МСП указаны юридический (номинальный) адрес и номинальный же код ОКВЭД.
-Очевидно, что они могут отличаться от реальных.
-Также следует заметить, что отличия могут быть разной степени выраженности в зависимости от места и вида деятельности.
+^1: https://fedstat.ru/indicator/58109
+
+Результаты анализа показывают высокую согласованность между числом организаций в целом и числом организаций в реестре МСП.
+Общий коэффициент корреляции в течение рассмотренных четырёх лет составляет чуть больше 0.8 и почти не изменяется (таблица 1).
+Коэффициенты корреляции по группам ОКВЭД в регионах различаются довольно сильно (см.
+график), но в основном составляют от 0.7 до 0.9.
+При анализе отдельных групп ОКВЭД коэффициент корреляции в большинстве случаев не меньше 0.9, а иногда приближается к единице (таблица 2).
+Лишь три группы характеризуются сравнительно небольшой корреляцией: госуправление (O), образование (P) и экстерриториальная деятельность (U).
+Это объяснимо, потому что в госуправлении и образовании подавляющее большинство организаций являются государственными или муниципальными (органы власти, подведомственные им учреждения, школы, детские сады, вузы), а количество организаций, ведущих экстерриториальную деятельность, слишком мало, и даже небольшие абсолютные расхождения в числах резко меняют общую картину.
+Если повторить анализ по годам и регионам, исключив три наиболее «проблемные» группы, то общие коэффициенты корреляции по годам повысятся до 0.94–0.95, а по регионам сместятся в сторону 0.9–1.0.
+(таблица 1 и график).
+Таким образом, в большинстве случаев реестр МСП может рассматриваться как достаточно качественный «прокси» как минимум в двух аспектах: он корректно отражает географическое распределение количества организаций по субъектам России в подавляющем большинстве групп ОКВЭД и структуру экономики по группам ОКВЭД в значительной части субъектов.
+Это не гарантирует репрезентативность на более низком уровне (конкретные коды ОКВЭД, муниципальные образования, населённые пункты) или по другим характеристикам (число работников, прибыль), но позволяет сделать осторожный вывод о том, что как минимум в ряде задач технология позволяет подготовить такой набор данных, который можно использовать — в отсутствие подходящих альтернатив — не только при изучении МСП как такового, но и при географическом исследовании экономической активности всех организаций.
+
+```{r corr_by_year}
+knitr::kable(corr_by_year, digits = 2, col.names = c("Год", "Корр. по всем группам ОКВЭД", "Корр. без O, P, U"))
+```
+
+```{r corr_by_region}
+corr_by_region_plot
+```
+
+```{r corr_by_group}
+knitr::kable(corr_by_group, digits = 2, col.names = c("Группа ОКВЭД", "$\\rho_{Spearman}$"))
+```
+
+Ещё одной проблемой является различие между реальным и формальным положением дел.
+Юридический адрес может отличаться от места ведения деятельности, а ОКВЭД, указанный в документах, может не соответствовать фактической деятельности.
+Сложности усиливаются из-за того, что расхождения могут проявляться в разной степени в зависимости от расположения и отрасли экономики: так, в особых экономических зонах будет больше номинально зарегистрированных организаций, а место деятельности интернет-компаний будет явно не совпадать с местом нахождения их офиса.
 Так, можно предположить, что место регистрации фирмы будет как минимум отчасти объясняться налоговыми или престижными соображениями, а место регистрации фирм, осуществляющих деятельность в интернете или в нескольких регионах, будет мало связано с местом их регистрации.
-Автору не известны какие-либо исследования на эту тему, равно как и надёжные и легко реализуемые методы, позволяющие проверить соответствие между номинальными данными и реальностью.
-Понятно лишь то, что датасет и любые производные от него статистические данные будут сочетать в себе как реальные закономерности, так и искажения, вызванные различием номинального и реального, причём оценить масштаб этих расхождений априори не представляется возможным — требуется проверка в каждом конкретном случае.
-Поскольку суть статьи сводится к описанию алгоритма и технологии, то остаётся лишь оставить здесь соответствующее предостережение для других исследователей.
-Потенциальными источниками «реальных» сведений могут быть, в частности, материалы открытых или частных картографических сервисов либо данные выборочных обследований Росстата.
-Выводы
+Вряд ли эту проблему можно решить системно, поэтому технология презюмирует достоверность данных о местонахождении и коде ОКВЭД и не проверяет их.
+В некоторых случаях объём данных сам по себе снижает эффект от неизбежных расхождений, но в других ситуациях задача исследователя, использующего технологию и произведённый с её помощью набор данных — самостоятельно проверить соответствие между номинальным и фактическим состоянием.
+Для этого можно, например, использовать открытые картографические сервисы, сайты организаций и здравый смысл.
+
+# Выводы

### 2023-12-25 StatsHelper.R

@@ -1,3 +1,8 @@
+library(dplyr)
+library(ggplot2)
+library(readr)
+library(tidyr)
+
 rsmp_raw_data_path <- "../tax-service-opendata/rsmp/xml"
 
 raw_files <- list.files(raw_data_path, pattern = "*.zip", full.names = TRUE)
@@ -94,3 +99,45 @@ counts_funnel <- data.frame(
 out_data <- read_csv(rsmp_out_file)
 out_org_count <- table(out_data$kind)["1"]
 out_ind_count <- table(out_data$kind)["2"]
+
+# Validation
+val_stats <- read_csv("../tax-service-opendata/Stats_for_validation.csv")
+
+corr_by_year_all <- val_stats %>% 
+  group_by(year) %>% 
+  summarise(cor = cor(count_reestr, count_rosstat, method = "spearman"))
+corr_by_year_filtered <- val_stats %>% 
+  filter(group != "O", group != "P", group != "U") %>% 
+  group_by(year) %>% 
+  summarise(cor = cor(count_reestr, count_rosstat, method = "spearman"))
+corr_by_year <- left_join(
+  yearly_corr_all,
+  yearly_corr_filtered,
+  by = "year",
+  suffix = c("_all", "_filtered")
+)
+
+corr_by_region_all <- val_stats %>% 
+  group_by(region) %>% 
+  summarise(cor = cor(count_reestr, count_rosstat, method = "spearman")) %>% 
+  arrange(cor)
+corr_by_region_filtered <- val_stats %>% 
+  filter(group != "O", group != "P", group != "U") %>% 
+  group_by(region) %>% 
+  summarise(cor = cor(count_reestr, count_rosstat, method = "spearman")) %>% 
+  arrange(cor)
+corr_by_region <- left_join(
+  corr_by_region_all, corr_by_region_filtered,
+  by = "region", suffix = c("_all", "_filtered")) %>% 
+  pivot_longer(-region, names_to = "option", values_to = "cor")
+corr_by_region_plot <- ggplot(corr_by_region, aes(x = cor, color = option)) +
+  geom_freqpoly(binwidth = .05) +
+  scale_color_discrete(name = "Группы ОКВЭД", labels = c("Все", "Все, кроме O, P, U")) +
+  labs(x = "Коэффициент корреляции Спирмена", y = "Число субъектов России")
+corr_by_region_plot
+
+corr_by_group <- val_stats %>% 
[…]

### 2023-12-25 Paper.Rmd
Add conclusion

@@
 # Выводы
+
+В статье описана технология, позволяющая на основе открытых данных ФНС России создать геопривязанный набор данных о субъектах МСП, ведущих деятельность определённого вида.
+Получаемый набор охватывает период с середины 2016 г.
+по конец 2023 г., содержит сведения о наименовании, местонахождении фирмы, её виде и категории, доходах, расходах и числе работников.
+В пространственном отношении он детализирован до населённых пунктов, во временном — до лет (в части доходов, расходов и числа работников) или месяцев (по остальным характеристикам), в структурном — до кода вида деятельности по ОКВЭД.
+Местонахождение фирмы указано как нормализованный адрес населённого пункта, его географические координаты и код ОКТМО.
+Технология реализована на языке программирования Python с использованием свободного программного обеспечения и позволяет обрабатывать большие (несколько терабайт) объёмы исходных данных на компьютере средней конфигурации за разумное время (обычно не более суток).
+
+Технология может использоваться как готовый способ получения данных для экономико-географических исследований или как база для создания производных решений.
+Поскольку получаемый набор данных представляет собой сплошной геопривязанный перечень субъектов МСП в России за несколько лет, то он может применяться для пространственного анализа малого и среднего предпринимательства.
+Благодаря относительной репрезентативности реестра МСП по отношению к генеральной совокупности российских фирм в географическом и структурном плане, технология может использоваться как источник данных везде, где требуется геопривязанная информация об экономической активности с детализацией до конкретных организаций или населённых пунктов.
+Технология позволяет самостоятельно получить данные, аналогичные тем, которые содержатся в коммерческих системах бизнес-аналитики или в справочных картографических сервисах.
+Поскольку создаваемый набор данных содержит координаты для каждой записи (элемента), то он удобен также для картографической визуализации.
+
+К ограничениям технологии можно отнести неизбежные расхождения между номинальными сведениями, указанными в реестрах, и фактическим положением дел; охват только части фирм, причём не обязательно репрезентативной во всех отношениях; правовые риски при анализе данных об индивидуальных предпринимателях, которые одновременно являются персональными данными.
+Из-за неполноты адресных баз данных и изменчивости классификатора муниципального деления геопривязка также происходит корректно не во всех случах.
+Кроме того, с учётом тенденции к снижению открытости административных данных есть опасность, что исходные наборы данных на сайте ФНС России перестанут быть доступными.

### 2023-12-26 Paper.Rmd
Add data flow diagram

@@
 
-## Вспомогательные источники: доходы, расходы, число работников, геоданные
+## Вспомогательные источники: доходы, расходы, число работников
 
-В качестве вспомогательных источников информации используются ещё два реестра открытых данных ФНС России и два набора данных о населённых пунктах России.
+В качестве вспомогательных источников информации используются ещё два реестра открытых данных ФНС России.
 
@@
 
-3.
+## Геоданные
+Для нормализации адресов, расстановки координат и кодов ОКТМО используются два справочника.
+
+1.
 Набор данных «Населенные пункты России: численность населения и географические координаты».
@@
 
-4.
+2.
 Набор данных «Города России».
@@
 
-Предложенный алгоритм обработки данных состоит из шести шагов.
+Предложенный алгоритм обработки данных состоит из шести шагов. Он изображён на схеме 1 и описан ниже.
+
+```{r diagram}
+diagram
+```
 

### 2023-12-26 Diagram.R

@@ -1,0 +1,69 @@
+library(DiagrammeR)
+
+diagram <- grViz("digraph {
+  
+graph[layout = dot, fontname = 'Ubuntu', fontsize = 18]
+
+subgraph cluster_a {
+  rank = same;
+  node[shape = box, width = 2.5, height = .5, fixedsize = True]
+  D1a[label='Реестр МСП\n(N zip-архивов)']
+  D2a[label='Извлечённые данные\n(N csv-файлов)']
+  D3a[label='Объединённая таблица\n(csv-файл)']
+  D4a[label='Геопривязанная таблица\n(csv-файл)']
+  
+  node[shape = none, width = 2, height = .5, fixedsize = True]
+  A2a[label = 'Извлечение, фильтрация,\nтрансформация (2а)']
+  A3a[label = 'Агрегирование (3а)']
+  A4a[label = 'Геопривязка (4)']
+  
+  D1a -> A2a -> D2a -> A3a -> D3a -> A4a -> D4a
+}
+
+subgraph cluster_b {
+  rank = same;
+  node[shape = box, width = 2.5, height = .5, fixedsize = True]
+  D1b[label='Доходах и расходы\n(M zip-архивов)']
+  D2b[label='Извлечённые данные\n(M csv-файлов)']
+  D4b[label='Отфильтрованная таблица\n(csv-файл)']
+  
+  node[shape = none, width = 2, height = .5, fixedsize = True]
+  A2b[label = 'Извлечение и трансформация (2б)']
+  A3b[label = 'Агрегирование и фильтрация(3б)']
+  
+  D1b -> A2b -> D2b -> A3b -> D4b
+}
+
+subgraph cluster_c {
+  rank = min;
+  node[shape = box, width = 2.5, height = .5, fixedsize = True]
+  D1c[label='Число работников\n(K zip-архивов)']
+  D2c[label='Извлечённые данные\n(K csv-файлов)']
+  D4c[label='Отфильтрованная таблица\n(csv-файл)']
+  
+  node[shape = none, width = 2, height = .5, fixedsize = True]
+  A2c[label = 'Извлечение, трансформация (2в)']
+  A3c[label = 'Агрегирование и фильтрация (3в)']
+  
+  D1c -> A2c -> D2c -> A3c -> D4c
+}
+
+# nodes
+# D stands for 'data', A stands for 'action'
+node[shape = box, width = 3, height = .5, fixedsize = True]
[…]

### 2023-12-28 Paper.Rmd
Start preparing for submission to Vestnik SPBU

@@
 ---
-title: "Технология использования открытых данных Федеральной налоговой службы России для экономико-географического анализа"
-output: odt_document
-bibliography: References.bib
+output:
+  odt_document:
+    number_sections: true
+    reference_odt: "assets/reference.odt"
+bibliography: "assets/biblio_ru.bib"
+csl: "assets/citations.csl"
 ---
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(include = FALSE)
+knitr::opts_chunk$set(
+  warning = FALSE, 
+  message = FALSE, 
+  echo = FALSE, 
+  fig.width = 6.5, 
+  fig.height = 5)
 options(knitr.kable.NA = "—")
@@
 
-```{r imports}
+```{r imports, cache = TRUE, include = FALSE}
 source("Plots.R")
 source("StatsHelper.R")
+source("Diagram.R")
 ```
 
-Аннотация:
-Ключевые слова:
-A Method of Using Open Data Provided by Federal Tax Service of Russia in Spatial Economy Analysis
-Abstract:
-Keywords:
+УДК 911.3, 519.688
 
-# Введение
+**Использование открытых данных ФНС России для создания геопривязанного набора данных о субъектах малого и среднего предпринимательства в России**
+
+Сёмин Павел Олегович
+
+Пермский государственный национальный исследовательский университет, Россия, Пермский край, г. Пермь, ул. Букирева, 15, 614068
+
+Сёмин Павел Олегович: ntsp\@ya.ru, ORCID 0000-0002-4015-9206
+
+Контактное лицо:  
+- Сёмин Павел Олегович  
+- Телефон +7 9222 06 77 01  
+- Email: ntsp\@ya.ru  
+- Адрес: Россия, 622002, Свердловская обл., г. Нижний Тагил, ул. Космонавтов, д. 35, кв. 62
+
+# Аннотация {-}
+
+В статье представлена технология (алгоритм, приложение), которая создаёт геопривязанный табличный набор данных о субъектах малого и среднего предпринимательства (МСП) в России на основе открытых данных Федеральной налоговой службы (ФНС России).
+Получаемый набор данных охватывает период с 2016 по 2023 гг.
+и всю территорию России.
+Он представляет собой полный список (генеральную совокупность) субъектов МСП и содержит информацию о категории, имени или наименовании, месте жительства или местонахождении, виде деятельности, доходах, расходах и численности работников организации или индивидуального предпринимателя.
+В пространственном отношении набор детализирован до населённых пунктов с указанием географических координат и кода ОКТМО; во временном отношении — до месяца или года в зависимости от характеристики (признака субъекта МСП); в отношении вида деятельности — до конкретных кодов ОКВЭД.
+В отличие от классической data paper, статья содержит не описание данных как таковых, а рассказывает о методике их получения и демонстрацию одного из возможных результатов применения технологии.
+Используя описанный алгоритм и его предложенную реализацию на языке Python, исследователь может самостоятельно обработать исходные данные и получить актуальную версию набора данных, адаптированную для его задач.
+Алгоритм оптимизирован так, чтобы для его выполнения было достаточно компьютера средней конфигурации, несмотря на большой (около 2 Тб) объём исходных данных.
+Создаваемый с помощью алгоритма набор данных может применяться как непосредственно для пространственного анализа МСП в России, так как и в качестве  относительно репрезентативного источника сведений об экономике России в разрезе конкретных организаций, выступая как открытая и бесплатная альтернатива коммерческим системам наподобие Spark и Контур.Фокус.
+Кроме того, благодаря координатам он может без дополнительной обработки использоваться для картографирования, а наличие кодов ОКТМО позволяет объединять его с официальными статистическими сведениями.
+
+**Ключевые слова:** открытые данные, пространственный анализ, геоданные, административные данные, малое и среднее предпринимательство, реестр организаций, набор данных, big data, Apache Spark, ФНС России
+
+# 1. Введение
 
@@
 
-# Обзор литературы: источники данных в общественной географии
+# 2. Обзор литературы: источники данных в общественной географии
 
@@
 
-# Исходные данные
+# 3. Исходные данные
 
@@
 
-## Основной источник: реестр МСП
+## 3.1. Основной источник: реестр МСП
 
@@
 
-## Вспомогательные источники: доходы, расходы, число работников
+## 3.2. Вспомогательные источники: доходы, расходы, число работников
 
@@
 
-## Геоданные
+## 3.3. Геоданные
 Для нормализации адресов, расстановки координат и кодов ОКТМО используются два справочника.
@@
 
-# Методика обработки данных
+# 4. Методика обработки данных
 
-Предложенный алгоритм обработки данных состоит из шести шагов. Он изображён на схеме 1 и описан ниже.
+Предложенный алгоритм обработки данных состоит из шести шагов. Он изображён на рис. 1 и описан ниже.
 
-```{r diagram}
-diagram
-```
-
-## Шаг 1: загрузка данных ФНС России
+## 4.1. Шаг 1: загрузка данных ФНС России
 
@@
 
-# Шаг 2а: фильтрующее извлечение данных реестра МСП
+## 4.2.1. Шаг 2а: фильтрующее извлечение данных реестра МСП
 
@@
 
-## Шаг 2б: извлечение данных о доходах и расходах
+## 4.2.2. Шаг 2б: извлечение данных о доходах и расходах
 
@@
 
-## Шаг 2в: извлечение данных о среднесписочной численности работников
+## 4.2.3. Шаг 2в: извлечение данных о среднесписочной численности работников
 
@@
 
-## Шаг 3а: агрегация данных о субъектах МСП
+## 4.3.1. Шаг 3а: агрегация данных о субъектах МСП
 
@@
 
-## Шаг 3б: агрегация данных о доходах, расходах
+## 4.3.2. Шаг 3б: агрегация данных о доходах, расходах
 
@@
 
-## Шаг 3в: агрегация данных о работниках
+## 4.3.3. Шаг 3в: агрегация данных о работниках
 
@@
 
-## Шаг 4: Геопривязка субъектов МСП
+## 4.4. Шаг 4: Геопривязка субъектов МСП
 
@@
 
-## Шаг 5: создание панельного представления
+## 4.5. Шаг 5: создание панельного представления
 
@@
 
-## Реализация алгоритма
+## 4.6. Реализация алгоритма
 Описанный выше алгоритм реализован как набор скриптов на языке программирования Python.
@@
 
-# Демонстрация работы технологии
+# 5. Демонстрация работы технологии
 
@@
 
-## Техническая статистика работы алгоритма
+## 5.1. Техническая статистика работы алгоритма
 
@@
 В таблице 1 приведена информация об уменьшении объёма данных в процессе работы алгоритма, а в таблице 2 — о фильтрации записей.
-
-```{r sizes_funnel}
-bytes_in_mb <- 2 ** 20
-sizes_funnel <-  sizes_funnel / bytes_in_mb
-knitr::kable(
-  sizes_funnel,
-  digits = 0,
-  format.args = list(scientific = FALSE),
-  col.names = c("Реестр МСП", "Доходы+расходы", "Работники", "Всего"))
-```
-
-```{r counts_funnel}
-counts_funnel <- counts_funnel / 1000 # in thousands
-knitr::kable(
-  counts_funnel,
-  digits = 0,
-  format.args = list(scientific = FALSE),
-  col.names = c("Реестр МСП", "Доходы+расходы", "Работники"))
-```
-
-Из таблиц следует, что технология многократно уменьшает объём обрабатываемых сведений: от сотен гигабайт исходных до нескольких десятков или сотен мегабайт итоговых наборов данных.
+Из таблиц видно, что технология многократно уменьшает объём обрабатываемых сведений: от сотен гигабайт исходных до нескольких десятков или сотен мегабайт итоговых наборов данных.
 Одновременно уменьшается и количество записей в данных с сотен миллионов до сотен тысяч.
@@
 
-## Временная динамика
+## 5.2. Временная динамика
 
 Набор данных позволяет отследить те или иные количественные характерстики с точностью до месяца.
-На графике ниже показаны 20 субъектов России с наибольшим количеством зарегистрированных субъектов МСП и изменение этого количества с августа 2016 г.
+На рис.
+2 показаны 20 субъектов России с наибольшим количеством зарегистрированных субъектов МСП и изменение этого количества с августа 2016 г.
 по ноябрь 2023 г.
 
-```{r time_series_plot}
-time_sereis_plot
-```
-
-## Распределение по субъектам федерации
+## 5.3. Распределение по субъектам федерации
 
 Нормализованные названия позволяют агрегировать информацию по субъектам федерации.
-На картограмах ниже представлено количество субъектов МСП, которые вели деятельность в сфере лесного (слева) и сельского (справа) хозяйства в 2021 г.
+На картограмах (рис.
+3) представлено количество субъектов МСП, которые вели деятельность в сфере лесного (слева) и сельского (справа) хозяйства в 2021 г.
 Распределение ожидаемо разное: наибольшее количество фирм в сфере сельского хозяйства расположено в Черноземье и на юге России, также заметны животноводческие регионы; лесное хозяйство сосредоточено на севере европейской части страны, в Восточной Сибири и на Дальнем Востоке.
 
-```{r regions_agri_forest}
-regions_plot_agri_forestry
-```
-
 В отличие от официальной статистики, получившийся набор данных позволяет проанализировать пространственное распределение с точностью до конкретных видов деятельности, а не крупных агломератов.
-На картограммах ниже представлено количество субъектов МСП, ведущих деятельность по выращиванию пшеницы (код 01.11.11) и картофеля (код 01.13.31).
+На картограммах (рис.
+4) представлено количество субъектов МСП, ведущих деятельность по выращиванию пшеницы (код 01.11.11) и картофеля (код 01.13.31).
 Здесь заметны некоторые странности (например, в лидерах по картофелю оказалась Якутия), которые, видимо, связаны с небольшим количеством субъектов МСП в каждом регионе: обычно у фирм в качестве основного указан более общий код, в то время как картограмма строилась по максимально детализированным.
 
-```{r wheat_potato}
-regions_plot_wheat_potato
-```
-
-## Распределение по населённым пунктам
+## 5.4. Распределение по населённым пунктам
 
@@
 Это позволяет выполнять более детализированные агрегации и картографически визуализировать их.
-На картограмме ниже показано количество субъектов МСП в разных населённых пунктах, а также преимущественная специализация населённых пунктов исходя из того, каких фирм в нём больше всего.
+На картограмме (рис.
+5) показано количество субъектов МСП в разных населённых пунктах, а также преимущественная специализация населённых пунктов исходя из того, каких фирм в нём больше всего.
 Показаны только населённые пункты более чем с 50 фирмами.
-Ещё одна аналогичная картограмма изображает количество субъектов МСП по видам деятельности в населённых пунктах Свердловской области, причём без ограничения на количество фирм.
+Ещё одна аналогичная картограмма (рис.
+6) изображает количество субъектов МСП по видам деятельности в населённых пунктах Свердловской области, причём без ограничения на количество фирм.
 
-```{r settlements_russia}
-activity_by_settlements
-```
-
-```{r settlements_svr}
-activity_by_settlements_svr
-```
-
-## Отслеживание отдельных фирм
+## 5.5. Отслеживание отдельных фирм
 
 Благодаря детальности набор данных позволяет анализировать изменения в конкретных фирмах, например, юридический «переезд» из одного региона в другой.
-На картограмме показаны «регионы-доноры» и «регионы-реципиенты» субъектов МСП за весь доступный период.
+На картограмме (рис.
+7) показаны «регионы-доноры» и «регионы-реципиенты» субъектов МСП за весь доступный период.
 
-```{r migrations}
-migrations_plot
-```
-
-## Анализ прибыли и числа работников
+## 5.6. Анализ прибыли и числа работников
 
 Вспомогательные данные о доходах, расходах и числе работников позволяют дополнить анализ, перейдя от простого числа компаний к характеристикам их деятельности.
-На трёх картограммах ниже показано соответственно общее число фирм, их суммарная прибыль и суммарное число работников по регионам.
+На трёх картограммах (рис.
+8) показано соответственно общее число фирм, их суммарная прибыль и суммарное число работников по регионам.
 Видно, что распределение хотя и похожее, но всё же отличается в зависимости от визуализируемой характеристики.
 
-```{r many_variables_by_region}
-count_profit_employees_by_region_plot
-```
-
-# Обсуждение
+# 6. Обсуждение
 
@@
 
-^1: Ч. 1 ст. 9 Федерального закона от 29.11.2007 № 282-ФЗ «Об официальном статистическом учете и системе государственной статистики в Российской Федерации».
+[^1]: Ч. 1 ст. 9 Федерального закона от 29.11.2007 № 282-ФЗ «Об официальном статистическом учете и системе государственной статистики в Российской Федерации».
 
-^2: П. 1.1 ст. 102 Налогового кодекса России.
+[^2]: П. 1.1 ст. 102 Налогового кодекса России.
 
@@
 
-^1: https://fedstat.ru/indicator/58109
+[^1]: https://fedstat.ru/indicator/58109
 
 Результаты анализа показывают высокую согласованность между числом организаций в целом и числом организаций в реестре МСП.
-Общий коэффициент корреляции в течение рассмотренных четырёх лет составляет чуть больше 0.8 и почти не изменяется (таблица 1).
+Общий коэффициент корреляции в течение рассмотренных четырёх лет составляет чуть больше 0.8 и почти не изменяется (таблица 3).
 Коэффициенты корреляции по группам ОКВЭД в регионах различаются довольно сильно (см.
-график), но в основном составляют от 0.7 до 0.9.
-При анализе отдельных групп ОКВЭД коэффициент корреляции в большинстве случаев не меньше 0.9, а иногда приближается к единице (таблица 2).
+рис.
+9), но в основном составляют от 0.7 до 0.9.
+При анализе отдельных групп ОКВЭД коэффициент корреляции в большинстве случаев не меньше 0.9, а иногда приближается к единице (таблица 4).
 Лишь три группы характеризуются сравнительно небольшой корреляцией: госуправление (O), образование (P) и экстерриториальная деятельность (U).
@@
 Если повторить анализ по годам и регионам, исключив три наиболее «проблемные» группы, то общие коэффициенты корреляции по годам повысятся до 0.94–0.95, а по регионам сместятся в сторону 0.9–1.0.
-(таблица 1 и график).
+(таблица 3 и рис.
+9).
 Таким образом, в большинстве случаев реестр МСП может рассматриваться как достаточно качественный «прокси» как минимум в двух аспектах: он корректно отражает географическое распределение количества организаций по субъектам России в подавляющем большинстве групп ОКВЭД и структуру экономики по группам ОКВЭД в значительной части субъектов.
 Это не гарантирует репрезентативность на более низком уровне (конкретные коды ОКВЭД, муниципальные образования, населённые пункты) или по другим характеристикам (число работников, прибыль), но позволяет сделать осторожный вывод о том, что как минимум в ряде задач технология позволяет подготовить такой набор данных, который можно использовать — в отсутствие подходящих альтернатив — не только при изучении МСП как такового, но и при географическом исследовании экономической активности всех организаций.
-
-```{r corr_by_year}
-knitr::kable(corr_by_year, digits = 2, col.names = c("Год", "Корр. по всем группам ОКВЭД", "Корр. без O, P, U"))
-```
-
-```{r corr_by_region}
-corr_by_region_plot
-```
-
-```{r corr_by_group}
-knitr::kable(corr_by_group, digits = 2, col.names = c("Группа ОКВЭД", "$\\rho_{Spearman}$"))
-```
 
@@
 
-# Выводы
+# 7. Выводы
 
@@
 Кроме того, с учётом тенденции к снижению открытости административных данных есть опасность, что исходные наборы данных на сайте ФНС России перестанут быть доступными.
+
+# Литература {-}
+
+<div id="refs"></div>
+
+![Рис. 1. Схема обработки данных. Пояснения в тексте](assets/diagram.png)
+
+```{r sizes_funnel}
+bytes_in_mb <- 2 ** 20
+sizes_funnel <-  sizes_funnel / bytes_in_mb
+knitr::kable(
+  sizes_funnel,
+  digits = 0,
+  format.args = list(scientific = FALSE),
+  col.names = c("Реестр МСП", "Доходы+расходы", "Работники", "Всего"),
+  caption = "Таблица 1. Изменение объёма данных в процессе работы алгоритма. Значения в мегабайтах (Мб). Строки — этапы работы алгоритма, столбцы — вид исходных данных и сумма")
+```
+
+```{r counts_funnel}
+counts_funnel <- counts_funnel / 1000 # in thousands
+knitr::kable(
+  counts_funnel,
+  digits = 0,
+  format.args = list(scientific = FALSE),
+  col.names = c("Реестр МСП", "Доходы+расходы", "Работники"),
+  caption = "Таблица 2. Изменение количества записей в процессе работы алгоритма. Значения в тысячах штук. Строки — этапы работы алгоритма, столбцы — вид исходных данных")
+```
+
+```{r time_series_plot, fig.cap = "Рис.
+2.
+Изменение числа субъектов МСП в демонстрационном наборе данных в 2016–2023 гг.
+с детализацией по месяцам.
+Показаны только первые 15 регионов (субъектов федерации) по количеству записей в соответствующем месяце"}
+time_series_plot
+```
+
+```{r regions_agri_forest, fig.cap = "Рис.
+3.
+Распределение количества субъектов МСП, ведущих деятельность в сфере сельского  и лесного хозяйства по регионам (субъектам федерации).
+Цвет показывает число, масштабированное так, что 0 — минимум, 1 — максимум"}
+regions_plot_agri_forestry
+```
+
+```{r wheat_potato, fig.cap = "Рис.
+4.
+Распределение количества субъектов МСП, ведущих деятельность по выращиванию пшеницы и картофеля по регионам (субъектам федерации).
+Цвет показывает число, масштабированное так, что 0 — минимум, 1 — максимум.
+Серый цвет — отсутствие данных"}
+regions_plot_wheat_potato
+```
+
+```{r settlements_russia, fig.cap = "Рис.
+5.
+Число субъектов МСП по населённым пунктам России и преобладающий вид деятельности.
+Размер кругов соответствует числу субъектов МСП, цвет — одной из трёх групп видов деятельности по классификатору ОКВЭД (см.
+легенду).
+Показаны только населённые пункты, в которых зарегистрировано более 100 фирм"}
+activity_by_settlements
+```
+
+```{r settlements_svr, fig.cap = "Рис.
+6.
+Число  субъектов МСП в населённых пунктах Свердловской области и их преобладающий вид деятельности.
+Размер кругов соответствует числу субъектов МСП, цвет — одной из трёх групп видов деятельности по классификатору ОКВЭД (см.
+легенду).
+Показаны все населённые пункты без ограничений по количеству зарегистрированных фирм"}
+activity_by_settlements_svr
+```
+
+```{r migrations, fig.cap = "Рис.
+7.
+Распределение количества субъектов МСП из демонстрационного набора, прибывших или покинувших территорию региона (субъекта федерации) за 2016–2023 г.
+Абсолютные числа масштабированы так, что -1 — максимальная убыль, 1 — максимальный прирост"}
+migrations_plot
+```
+
+```{r many_variables_by_region, fig.cap = "Распределение числа зарегистрированных субъектов МСП, их суммарной прибыли и суммарной численности работников в демонстрационном наборе в 2021 г.
+по регионам (субъектам федерации).
+Абсолютные значения масштабированы по шкале от 0 до 1, где 0 — минимум, 1 — максимум"}
+count_profit_employees_by_region_plot
+```
+
+```{r corr_by_year}
+knitr::kable(
+  corr_by_year,
+  digits = 2, 
+  col.names = c("Год", "Корр. по всем группам ОКВЭД", "Корр. без O, P, U"),
+  caption = "Таблица 3. Корреляция между числом организаций в реестре МСП и общим числом организаций в разрезе регионов (субъектов федерации) и групп деятельности по годам")
+```
+
+```{r corr_by_region, fig.cap = "Рис. 9. Распределение регионов (субъектов федерации) по корреляции между числом организаций в реестре МСП и общим числом организаций"}
+corr_by_region_plot
+```
+
+```{r corr_by_group}
+knitr::kable(
+  corr_by_group,
+  digits = 2,
+  col.names = c("Группа ОКВЭД", "$\\rho_{Spearman}$"),
+  caption = "Таблица 4. Корреляция между числом организаций в реестре МСП и общим числом организаций в разрезе регионов (субъектов федерации) и годов по группам деятельности")
+```

### 2023-12-28 Diagram.R

@@ -1,4 +1,6 @@
 library(DiagrammeR)
+library(DiagrammeRsvg)
+library(rsvg)
 
 diagram <- grViz("digraph {
   
@@ -67,3 +69,6 @@ D3a -> {A3b, A3c}
 {D4a, D4b, D4c} -> A5 -> D5 
 }")
 
+diagram %>% export_svg() %>% charToRaw() %>% rsvg_png("assets/diagram.png")
+
+            

### 2023-12-28 StatsHelper.R

@@ -5,7 +5,7 @@ library(tidyr)
 
 rsmp_raw_data_path <- "../tax-service-opendata/rsmp/xml"
 
-raw_files <- list.files(raw_data_path, pattern = "*.zip", full.names = TRUE)
+raw_files <- list.files(rsmp_raw_data_path, pattern = "*.zip", full.names = TRUE)
 raw_files_count <- length(raw_files)
 raw_files_count
 
@@ -111,8 +111,8 @@ corr_by_year_filtered <- val_stats %>%
   group_by(year) %>% 
   summarise(cor = cor(count_reestr, count_rosstat, method = "spearman"))
 corr_by_year <- left_join(
-  yearly_corr_all,
-  yearly_corr_filtered,
+  corr_by_year_all,
+  corr_by_year_filtered,
   by = "year",
   suffix = c("_all", "_filtered")
 )

### 2024-01-06 Paper.Rmd
Add English translation, some other minor fixes

@@
     number_sections: true
-    reference_odt: "assets/reference.odt"
-bibliography: "assets/biblio_ru.bib"
+    reference_odt: assets/reference.odt
+bibliography: assets/biblio_ru.bib
 csl: "assets/citations.csl"
+link-bibliography: false
 ---
@@
   fig.width = 6.5, 
-  fig.height = 5)
+  fig.height = 5,
+  fig.id = FALSE,
+  dpi = 300)
 options(knitr.kable.NA = "—")
@@
 
+<br>
+
 **Использование открытых данных ФНС России для создания геопривязанного набора данных о субъектах малого и среднего предпринимательства в России**
+
+<br>
 
@@
 
+<br>
+
 Пермский государственный национальный исследовательский университет, Россия, Пермский край, г. Пермь, ул. Букирева, 15, 614068
+
+<br>
 
@@
 
+<br>
+
 Контактное лицо:
-- Сёмин Павел Олегович  
-- Телефон +7 9222 06 77 01  
-- Email: ntsp\@ya.ru  
-- Адрес: Россия, 622002, Свердловская обл., г. Нижний Тагил, ул. Космонавтов, д. 35, кв. 62
+
+Сёмин Павел Олегович
+
+Телефон +7 9222 06 77 01
+
+Email: ntsp\@ya.ru
+
+Адрес: Россия, 622002, Свердловская обл., г. Нижний Тагил, ул. Космонавтов, д. 35, кв. 62
 
@@
 
-# 1. Введение
+# Введение
 
@@
 Л.
-Виноградова, анализируя изменение интенсивности земледелия в Европе, опирается на показатели стран в разные годы [@Виноградова2020]; Т.
+Виноградова, анализируя изменение интенсивности земледелия в Европе, опирается на показатели стран в разные годы [@Vinogradova2020]; Т.
 А.
-Балина с соавторам, рассматривая воспрос о территориях опережающего социально-экономического развития, приводит подсчёты по субъектам Российской Федерации [@Балина2021]; Г.
+Балина с соавторам, рассматривая воспрос о территориях опережающего социально-экономического развития, приводит подсчёты по субъектам Российской Федерации [@Balina2021]; Г.
 М.
-Лаппо, описывая свою концепцию разнообразия городов, опирается на сведения о числе жителей в населённых пунктах в разные периоды времени [@Лаппо2019].
+Лаппо, описывая свою концепцию разнообразия городов, опирается на сведения о числе жителей в населённых пунктах в разные периоды времени [@Lappo2019].
 Другие исследователи используют большие массивы дезагрегированных данных, собирая их вручную в общедоступных источниках и используя сложные методы их обработки.
@@
 В.
-Красновы для своей работы, посвящённой трансформации сети расселения в Псковской области, собирали публичные сведения о местоположении пользователей социальной сети «ВКонтакте» и открытые данные о поисковых запросах к «Яндексу» [@Красновы2020].
+Красновы для своей работы, посвящённой трансформации сети расселения в Псковской области, собирали публичные сведения о местоположении пользователей социальной сети «ВКонтакте» и открытые данные о поисковых запросах к «Яндексу» [@Krasnov2020].
 Есть работы, опирающиеся на обобщённую статистику, но использующие продвинутые методы её анализа.
@@
 В.
-Федорченко взял данные Международной организации автопроизводителей, рассчитал на их основе индексы и дополнил их картографической визуализацией, что позволило эффективно проанализировать территориальные сдвиги в мировой автомобильной промышленности [@Федорченко2017].
+Федорченко взял данные Международной организации автопроизводителей, рассчитал на их основе индексы и дополнил их картографической визуализацией, что позволило эффективно проанализировать территориальные сдвиги в мировой автомобильной промышленности [@Fedorchenko2017].
 «Большие данные» также проникают в общественно-географическую науку, что отметили ещё десять лет назад М.
@@
 Н.
-Лаженцев, например, отмечает, что «совершенствование статистической науки в значительной мере предопределяет уровень экономико-географических исследований» [@Лаженцев2015], а К.
+Лаженцев, например, отмечает, что «совершенствование статистической науки в значительной мере предопределяет уровень экономико-географических исследований» [@Lazhentsev2015], а К.
 Брунсдон рассуждает о важности изложения точных деталей анализа данных в контексте воспроизводимости географических исследований [@Brunsdon2016].
@@
 
-# 2. Обзор литературы: источники данных в общественной географии
+# Обзор литературы: источники данных в общественной географии
 
@@
 В.
-Кузнецова при исследовании структуры экономики российских регионов использует статистику валового регионального продукта и объёма отгруженных товаров [@Кузнецова2018]; Г.
+Кузнецова при исследовании структуры экономики российских регионов использует статистику валового регионального продукта и объёма отгруженных товаров [@Kuznetsova2018]; Г.
 М.
@@
 Г.
-Сафронов изучает третичный сектор экономики России на базе данных Eurostat, Росстата и Федеральной налоговой службы [@Сафронов2021]; Ю.
+Сафронов изучает третичный сектор экономики России на базе данных Eurostat, Росстата и Федеральной налоговой службы [@Safronov2021]; Ю.
 В.
@@
 Лачининский с соавторами оценивает спрол городских территорий Санкт-Петербурга с опорой на спутниковые снимки Landsat, так как на базе сведений официальной статистики провести необходимые подсчёты невозможно [@Lachininsky2023].
-К преимуществам альтернативных источников данных относят повышенную точность и своевременность [@Воробьев2021], более высокое пространственное и временное разрешение [@Бабкин2021], возможность точного и детального анализа на микроуровне [@Бочкарёв2017], отмечая при этом несвоевременность, неполноту и потенциальную недостоверность официальной статистики [@Кузьминов2021].
-Альтернативные источники данных дают доступ к генеральной совокупности, вследствие чего «отпадает необходимость в использовании выборочных, агрегированных и косвенных» источников сведений [@Блануца2017].
-При использовании методов машинного обучения, что также становится актуальным в современной географии, важны объём выборки, дезагрегированность, количество признаков в данных, что также побуждает обращаться к альтернативным источникам [@Ростислав2020].
+К преимуществам альтернативных источников данных относят повышенную точность и своевременность [@Vorobyev2021], более высокое пространственное и временное разрешение [@Babkin2021], возможность точного и детального анализа на микроуровне [@Bochkaryov2017], отмечая при этом несвоевременность, неполноту и потенциальную недостоверность официальной статистики [@Kuzminov2021].
+Альтернативные источники данных дают доступ к генеральной совокупности, вследствие чего «отпадает необходимость в использовании выборочных, агрегированных и косвенных» источников сведений [@Blanutsa2017].
+При использовании методов машинного обучения, что также становится актуальным в современной географии, важны объём выборки, дезагрегированность, количество признаков в данных, что также побуждает обращаться к альтернативным источникам [@Rostislav2020].
 Таким образом, хотя традиционные статистические сведения по-прежнему важны в общественно-географических исследованиях, а порой незаменимы, есть нарастающая потребность в других информационных ресурсах.
@@
 Они публикуются на официальном сайте этого органа власти или связанных с ним ресурсах.
-В географических работах используются, в частности, сведения о доходах физических лиц [@Petrov2021], о налоговых поступлениях в разрезе субъектов федерации или видов экономической деятельности [@Ростислав2020; @Сафронов2021], о среднесписочной численности работников организаций [@Rostislav2021], о количестве учтённых жилых помещений [@Lachininsky2021], о количестве малых и средних предприятий и мерах их поддержки [@Гуменюк2022].
+В географических работах используются, в частности, сведения о доходах физических лиц [@Petrov2021], о налоговых поступлениях в разрезе субъектов федерации или видов экономической деятельности [@Rostislav2020; @Safronov2021], о среднесписочной численности работников организаций [@Rostislav2021], о количестве учтённых жилых помещений [@Lachininsky2021], о количестве малых и средних предприятий и мерах их поддержки [@Gumenyuk2022].
 Внимание исследователей к данным, публикуемым ФНС России, связано, по-видимому, с многофункциональностью этого органа власти, из-за которой он собирает огромный объём информации о разных сторонах российской экономики: хозяйствующих субъектах, имуществе, доходах, расходах.
@@
 И.
-Шерубнёвой [@Коломак2023], которые анализирует агломерационные эффекты на основе показателей предприятий.
+Шерубнёвой [@Kolomak2023], которые анализирует агломерационные эффекты на основе показателей предприятий.
 Кроме того, на их основе можно считать статистику с произвольной пространственной или временной группировкой, в том числе такой, которую невозможно, сложно или неудобно извлекать из данных Росстата.
@@
 С.
-Гуменюк оперирует подсчётами абсолютного и относительного числа малых и средних предприятий по муниципалитетам Калининградской области [@Гуменюк2022]; М.
+Гуменюк оперирует подсчётами абсолютного и относительного числа малых и средних предприятий по муниципалитетам Калининградской области [@Gumenyuk2022]; М.
 А.
@@
 А.
-Макушин с соавторами использует данные о предприятиях в городах БАМа, чтобы оценить структуру промышленности [@Макушин2023].
-Дополнительную важность сведения о конкретных фирмах приобретают при исследованиях в масштабе городов [@Lachininsky2021; Никоноров2021; @Kozhevnikov2023], при изучении узких секторов экономики, которые не фигурируют обособленно в статистике [@Моисеева2016; @Pankratov2023], при анализе отдельно взятой отрасли [@Кузьминов2021].
+Макушин с соавторами использует данные о предприятиях в городах БАМа, чтобы оценить структуру промышленности [@Makushin2023].
+Дополнительную важность сведения о конкретных фирмах приобретают при исследованиях в масштабе городов [@Lachininsky2021; @Nikonorov2021; @Kozhevnikov2023], при изучении узких секторов экономики, которые не фигурируют обособленно в статистике [@Moiseeva2016; @Pankratov2023], при анализе отдельно взятой отрасли [@Kuzminov2021].
 
 Общепринятого источника геопривязанной информации о фирмах в России нет.
-Исследователи и прикладные специалисты обычно опираются на данные коммерческих [@Lachininsky2021] или открытых геоинформационных сервисов, систем бизнес-аналитики [@Моисеева2016; @Кузьминов2021], уже упомянутого реестра МСП [@Гуменюк2022].
+Исследователи и прикладные специалисты обычно опираются на данные коммерческих [@Lachininsky2021] или открытых геоинформационных сервисов, систем бизнес-аналитики [@Moiseeva2016; @Kuzminov2021], уже упомянутого реестра МСП [@Gumenyuk2022].
 Как правило, использование этих способов сопряжено с двумя проблемами: данные предоставляются за плату либо требуется приложить значительные усилия для их приведения в подходящий для анализа формат.
@@
 
-# 3. Исходные данные
+# Исходные данные
 
@@
 
-## 3.1. Основной источник: реестр МСП
+## Основной источник: реестр МСП
 
@@
 Норма о реестре появилась в законе в конце 2015 г.
-На практике реестр размещён на сайте ФНС России в форме открытых данных в виде ежемесячных полных выгрузок, первая из которых датируется августом 2016 г.
+На практике реестр размещён на сайте ФНС России в форме открытых данных в виде ежемесячных полных выгрузок, первая из которых датируется августом 2016 г.[^1]
+
+[^1]: https://www.nalog.gov.ru/opendata/7707329152-rsmp/
 
@@
 
-## 3.2. Вспомогательные источники: доходы, расходы, число работников
+## Вспомогательные источники: доходы, расходы, число работников
 
@@
 1.
-Набор открытых данных ФНС России «Сведения о суммах доходов и расходов по данным бухгалтерской (финансовой) отчетности организаций».
+Набор открытых данных ФНС России «Сведения о суммах доходов и расходов по данным бухгалтерской (финансовой) отчетности организаций»[^2].
 Содержит ИНН, суммарные доходы и расходы за год, предшествующий году публикации.
@@
 Данные доступны с 2019 г.
-по настоящее время[^1].
+по настоящее время[^3].
 В отличие от реестра МСП, содержит только информацию об органзациях, но не об индивидуальных предпринимателях.
@@
 
-[^1]: На момент подготовки статьи на сайте ФНС России была опубликована только последняя версия этого и следующего набора данных.
+[^2]: https://www.nalog.gov.ru/opendata/7707329152-revexp/
+
+[^3]: На момент подготовки статьи на сайте ФНС России была опубликована только последняя версия этого и следующего набора данных.
 Раньше были доступны предыдущие.
@@
 2.
-Набор открытых данных ФНС России «Сведения о среднесписочной численности работников организации».
+Набор открытых данных ФНС России «Сведения о среднесписочной численности работников организации»[^4].
 Содержит ИНН и среднесписочную численность работников за год, предшествующий году публикации сведений.
@@
 
-## 3.3. Геоданные
+[^4]: https://www.nalog.gov.ru/opendata/7707329152-sshr2019/
+
+## Геоданные
 Для нормализации адресов, расстановки координат и кодов ОКТМО используются два справочника.
@@
 1.
-Набор данных «Населенные пункты России: численность населения и географические координаты».
+Набор данных «Населенные пункты России: численность населения и географические координаты»[^5].
 Подготовлен проектом «Инфраструктура научно-исследовательских данных» на базе открытых данных Росстата и сведений региональных медицинских информационно-аналитических центров.
@@
 
+[^5]: https://data.rcsi.science/data-catalog/datasets/160/
+
 2.
-Набор данных «Города России».
+Набор данных «Города России»[^6].
 Подготовлен компанией «Дадата» на базе собственных разработок.
@@
 
+[^6]: https://github.com/hflabs/city
+
 Все перечисленные выше источники данных являются общедоступными и бесплатными для использования.
 
-# 4. Методика обработки данных
+# Методика обработки данных
 
@@
 
-## 4.1. Шаг 1: загрузка данных ФНС России
+## Шаг 1: загрузка данных ФНС России
 
@@
 
-## 4.2.1. Шаг 2а: фильтрующее извлечение данных реестра МСП
+## Шаг 2а: фильтрующее извлечение данных реестра МСП
 
@@
 
-## 4.2.2. Шаг 2б: извлечение данных о доходах и расходах
+## Шаг 2б: извлечение данных о доходах и расходах
 
@@
 
-## 4.2.3. Шаг 2в: извлечение данных о среднесписочной численности работников
+## Шаг 2в: извлечение данных о среднесписочной численности работников
 
@@
 
-## 4.3.1. Шаг 3а: агрегация данных о субъектах МСП
+## Шаг 3а: агрегация данных о субъектах МСП
 
@@
 
-## 4.3.2. Шаг 3б: агрегация данных о доходах, расходах
+## Шаг 3б: агрегация данных о доходах, расходах
 
@@
 
-## 4.3.3. Шаг 3в: агрегация данных о работниках
+## Шаг 3в: агрегация данных о работниках
 
@@
 
-## 4.4. Шаг 4: Геопривязка субъектов МСП
+## Шаг 4: Геопривязка субъектов МСП
 
@@
 
-## 4.5. Шаг 5: создание панельного представления
+## Шаг 5: создание панельного представления
 
@@
 
-## 4.6. Реализация алгоритма
+## Реализация алгоритма
 Описанный выше алгоритм реализован как набор скриптов на языке программирования Python.
@@
 Для агрегации используется система обработки больших данных Apache Spark.
-Исходный код опубликован по адресу: https://github.com/PavelSyomin/russian-legal-companies
+Исходный код опубликован по адресу: https://github.com/PavelSyomin/russian-smb-companies
 
-# 5. Демонстрация работы технологии
+# Демонстрация работы технологии
 
@@
 
-## 5.1. Техническая статистика работы алгоритма
+## Техническая статистика работы алгоритма
 
@@
 
-## 5.2. Временная динамика
+## Временная динамика
 
@@
 
-## 5.3. Распределение по субъектам федерации
+## Распределение по субъектам федерации
 
@@
 
-## 5.4. Распределение по населённым пунктам
+## Распределение по населённым пунктам
 
@@
 
-## 5.5. Отслеживание отдельных фирм
+## Отслеживание отдельных фирм
 
@@
 
-## 5.6. Анализ прибыли и числа работников
+## Анализ прибыли и числа работников
 
@@
 
-# 6. Обсуждение
+# Обсуждение
 
@@
 Его можно использовать для сплошного исследования малого и среднего предпринимательства в стране.
-В некотором роде это квази-источник первичных статистических сведений, которые по общему правилу относятся к информации ограниченного доступа[^1], но в данном случае являются открытыми в силу прямого указания в Налоговом кодексе[^2].
+В некотором роде это квази-источник первичных статистических сведений, которые по общему правилу относятся к информации ограниченного доступа[^7], но в данном случае являются открытыми в силу прямого указания в Налоговом кодексе[^8].
 Кроме того, набор данных является полным геопривязанным перечнем индивидуальных предпринимателей.
@@
 
-[^1]: Ч. 1 ст. 9 Федерального закона от 29.11.2007 № 282-ФЗ «Об официальном статистическом учете и системе государственной статистики в Российской Федерации».
+[^7]: Ч. 1 ст. 9 Федерального закона от 29.11.2007 № 282-ФЗ «Об официальном статистическом учете и системе государственной статистики в Российской Федерации».
 
-[^2]: П. 1.1 ст. 102 Налогового кодекса России.
+[^8]: П. 1.1 ст. 102 Налогового кодекса России.
 
@@
 Для оценки репрезентативности реестра МСП можно использовать корреляционный анализ.
-Единая медведомственная информационно-статистическая система (ЕМИСС) содержит данные о числе организаций в разрезе регионов и групп ОКВЭД за 2019–2022 гг[^1].
+Единая медведомственная информационно-статистическая система (ЕМИСС) содержит данные о числе организаций в разрезе регионов и групп ОКВЭД за 2019–2022 гг[^9].
 Такая же статистика рассчитывается на базе реестра МСП.
@@
 
-[^1]: https://fedstat.ru/indicator/58109
+[^9]: https://fedstat.ru/indicator/58109
 
@@
 
-# 7. Выводы
+# Выводы
 
@@
 
-![Рис. 1. Схема обработки данных. Пояснения в тексте](assets/diagram.png)
+\newpage
 
-```{r sizes_funnel}
+```{r fig-1-diagram, fig.cap = "Рис. 1. Схема обработки данных"}
+grid::grid.raster(diagram_raster)
+```
+
+<br>
+
+```{r tab-1-sizes-funnel}
 bytes_in_mb <- 2 ** 20
@@
 
-```{r counts_funnel}
+<br>
+
+```{r tab-2-counts-funnel}
 counts_funnel <- counts_funnel / 1000 # in thousands
@@
 
-```{r time_series_plot, fig.cap = "Рис.
+<br>
+
+```{r fig-2-time-series-plot, fig.cap = "Рис.
 2.
@@
 
-```{r regions_agri_forest, fig.cap = "Рис.
+<br>
+
+```{r fig-3-regions-agri-forest, fig.cap = "Рис.
 3.
@@
 
-```{r wheat_potato, fig.cap = "Рис.
+<br>
+
+```{r fig-3-regions-wheat-potato, fig.cap = "Рис.
 4.
@@
 
-```{r settlements_russia, fig.cap = "Рис.
+<br>
+
+```{r fig-5-activity-by-settlements-russia, fig.cap = "Рис.
 5.
@@
 легенду).
-Показаны только населённые пункты, в которых зарегистрировано более 100 фирм"}
+Показаны только населённые пункты, в которых зарегистрировано более 50 фирм"}
 activity_by_settlements
@@
 
-```{r settlements_svr, fig.cap = "Рис.
+<br>
+
+```{r fig-6-activity-by-settlements_svr, fig.cap = "Рис.
 6.
@@
 
-```{r migrations, fig.cap = "Рис.
+<br>
+
+```{r fig-7-migrations, fig.cap = "Рис.
 7.
@@
 
-```{r many_variables_by_region, fig.cap = "Распределение числа зарегистрированных субъектов МСП, их суммарной прибыли и суммарной численности работников в демонстрационном наборе в 2021 г.
-по регионам (субъектам федерации).
-Абсолютные значения масштабированы по шкале от 0 до 1, где 0 — минимум, 1 — максимум"}
+<br>
+
+```{r fig-8-count-employees-profit-by-region, fig.cap = "Распределение числа зарегистрированных субъектов МСП, их суммарной прибыли и суммарной численности работников в демонстрационном наборе в 2021 г.
+по регионам (субъектам федерации)"}
 count_profit_employees_by_region_plot
@@
 
-```{r corr_by_year}
+<br>
+
+```{r tab-3-corr-by-year}
 knitr::kable(
@@
   digits = 2, 
-  col.names = c("Год", "Корр. по всем группам ОКВЭД", "Корр. без O, P, U"),
+  col.names = c("Год", "$\\rho_{Spearman}$, все группы ОКВЭД", "$\\rho_{Spearman}$, без O, P, U"),
   caption = "Таблица 3. Корреляция между числом организаций в реестре МСП и общим числом организаций в разрезе регионов (субъектов федерации) и групп деятельности по годам")
@@
 
-```{r corr_by_region, fig.cap = "Рис. 9. Распределение регионов (субъектов федерации) по корреляции между числом организаций в реестре МСП и общим числом организаций"}
+<br>
+
+```{r fig-9-corr-by-region, fig.cap = "Рис. 9. Распределение регионов (субъектов федерации) по корреляции между числом организаций в реестре МСП и общим числом организаций"}
 corr_by_region_plot
@@
 
-```{r corr_by_group}
+<br>
+
+```{r tab-4-corr_by_group}
 knitr::kable(

### 2024-01-06 Paper_meta_en.Rmd
Add English translation, some other minor fixes

---
output:
  odt_document:
    number_sections: true
    reference_odt: assets/reference.odt
bibliography: assets/biblio_en.bib
csl: "assets/citations.csl"
link-bibliography: false
nocite: |
  @*
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE, 
  message = FALSE, 
  echo = FALSE, 
  fig.width = 6.5, 
  fig.height = 5,
  fig.id = FALSE,
  dpi = 300)
options(knitr.kable.NA = "—")
```

```{r imports, cache = TRUE, include = FALSE}
source("Plots.R")
source("StatsHelper.R")
source("Diagram.R")
```

UDC 911.3, 519.688

<br>

**Technology of Using Federal Taxpayer Service's Open Data for Making a Geo-Referenced Dataset about Small and Medium-Sized Businesses in Russia**

<br>

Syomin Pavel Olegovich

<br>

Perm State National University, 15 Bukireva st., Perm, 614068, Russian Federation

<br>

Syomin Pavel Olegovich: ntsp\@ya.ru, ORCID 0000-0002-4015-9206

<br>

Contact:

Syomin Pavel Olegovich

Phone +7 9222 06 77 01

Email: ntsp\@ya.ru

Address: 35-62 Kosmonavtov st., Sverdlovsk obl., Nizhniy Tagil, 622002, Russia

# Abstract {-}

The paper features a technology (algorithm, application) for making of a tabular georeferenced dataset about small and medium-sized bisunesses (SMBs) in Russia based on open data of Federal Taxpayer Service (FTS) of Russia. The resulting dataset covers the period from 2016 to 2023 and the whole area of Russia. It can be considered as a complete list, or population, of SMBs and contains information about category, name, address, activity code, revenue, expediture, number of employees of an organization or an individual entrepreneur. The spatial resolution of the dataset is by settlement, including geographical coordinates and municipality code; the temporal resolution is up to a year or month depending on the features; the resolution regarding the activity type is by concrete activity code according to the state classifier. In comparison to the typical data paper, this article contains the description of the methods used to produce the data and demonstrates the application of the proposed technology rather than the description of the dataset itself. Using the algorithm and its sample implementation on Python, a researcher can perform the data processing independently and obtain the up-to-date copy of the resulting data, optionally tailored to its specific purposes. The algorithm is optimized so that it runs on a typical PC despite the relatively large (~2Tb) size of unpacked source data. The dataset produced by the proposed method can be used for spatial analysis of SMBs in Russia as well as a proxy data source which is quite relevant for analysis of Russia's economy. The detalization up to a specific organizations enables the use of the produced dataset as a free partial alternative to commercial business intelligence systems like Spark and Kontur.Focus. Also, the presence of geographical coordinates makes mapping easy, and municipality codes may be used to join the dataset with official statistical databases.

**Keywords:** open data, spatial analysis, spatial data, administrative data, small&medium business, business registry, dataset, big data, Apache Spark, FTS of Russia

# Bibliography {-}

<div id="refs"></div>

\newpage

```{r fig-1-diagram, fig.cap = "Fig. 1. Data Processing Diagram"}
grid::grid.raster(diagram_raster)
```

<br>

```{r tab-1-sizes-funnel}
bytes_in_mb <- 2 ** 20
sizes_funnel <-  sizes_funnel / bytes_in_mb
knitr::kable(
  sizes_funnel,
  digits = 0,
  format.args = list(scientific = FALSE),
  col.names = c("SMB Registry", "Revenue+Expediture", "Employees", "Total"),
  caption = "Table 1. Change in the data volumes during program execution (in megabytes). Rows are stages, columns are source datasets and totals")
```

<br>

```{r tab-2-counts-funnel}
counts_funnel <- counts_funnel / 1000 # in thousands
knitr::kable(
  counts_funnel,
  digits = 0,
  format.args = list(scientific = FALSE),
  col.names = c("SMB Registry", "Revenue+Expediture", "Employees"),
  caption = "Table 2. Change in the number of records during program execution (in thousands). Rows are stages, columns are source datasets")
```

<br>

```{r fig-2-time-series-plot, fig.cap = "Fig. 2. Monthly changes in the number of SMBs in the demo dataset in 2016–2023. Only top 15 regions are shown"}
time_series_plot
```

<br>

```{r fig-3-regions-agri-forest, fig.cap = "Fig. 3. Regional distribution of SMBs which main activity is agriculture or forestry. Color represents the min-max scaled value"}
regions_plot_agri_forestry
```

<br>

```{r fig-3-regions-wheat-potato, fig.cap = "Fig. 4. Regional distribution of SMBs which main activity is growing of wheat or potato. Color represents the min-max scaled value, gray is no data"}
regions_plot_wheat_potato
```

<br>

```{r fig-5-activity-by-settlements-russia, fig.cap = "Fig. 5. Count of SMBs in some Russian settlements with dominant activity type. The counts are represented with circle size, and the most popular activity group in the settlement is displayed by color. Only settlements with 50+ SMBs are shown"}
activity_by_settlements
```

<br>

```{r fig-6-activity-by-settlements_svr, fig.cap = "Fig. 6. Count of SMBs in settlements of Sverdlovsk region with dominant activity type. The counts are represented with circle size, and the most popular activity group in the settlement is displayed by color. All settlements regardless of the counts are shown"}
activity_by_settlements_svr
```

<br>

```{r fig-7-migrations, fig.cap = "Fig. 7. Regional distribution of SMBs which have arrived to or left the region during 2016–2023. The color is min-max scaled value, -1 is the top decline, while +1 is the top growth"}
migrations_plot
```

<br>

```{r fig-8-count-employees-profit-by-region, fig.cap = "Fig. 8. Regional distribution of counts of SMBs, their total profit and total number of employees in the demo dataset in 2021"}
count_profit_employees_by_region_plot
```

<br>

```{r tab-3-corr-by-year}
knitr::kable(
  corr_by_year,
  digits = 2, 
  col.names = c("Year", "$\\rho_{Spearman}$, all activity groups", "$\\rho_{Spearman}$, excluding groups O, P, U"),
  caption = "Table 3. The correlation between the count of SMBs and total count of businesses by year")
```

<br>

```{r fig-9-corr-by-region, fig.cap = "Fig. 9. Distribution of regions by the correlation coefficient between the SMBs counts and total businesses counts"}
corr_by_region_plot
```

<br>

```{r tab-4-corr_by_group}
knitr::kable(
  corr_by_group,
  digits = 2,
  col.names = c("Activity group", "$\\rho_{Spearman}$"),
  caption = "Table 4. The correlation between the count of SMBs and total count of businesses by activity group")
```

### 2024-01-06 Diagram.R

@@ -69,6 +69,4 @@ D3a -> {A3b, A3c}
 {D4a, D4b, D4c} -> A5 -> D5 
 }")
 
-diagram %>% export_svg() %>% charToRaw() %>% rsvg_png("assets/diagram.png")
-
-            
+diagram_raster <- diagram %>% export_svg() %>% charToRaw() %>% rsvg_nativeraster(width = 2000)

### 2024-01-06 Plots.R

@@ -9,8 +9,6 @@ library(tidyr)
 rsmp_data_path <- "../tax-service-opendata/rsmp/reestr_group_A/data_product.csv"
 rsmp_data <- read_csv(rsmp_data_path)
 
-rsmp_data %>% 
-  count(region, start_date, end_date)
 # Regions time series
 monthly_counts <- do.call(rbind, lapply(
   seq.Date(min(rsmp_data$start_date), max(rsmp_data$end_date), by = "month"),
@@ -18,7 +16,8 @@ monthly_counts <- do.call(rbind, lapply(
     filter(rsmp_data, start_date <= x, end_date >= x) %>% 
       count(region) %>% 
       mutate(date = x) %>% 
-      arrange(-n)
+      arrange(-n) %>% 
+      head(15)
   }
 ))
 
@@ -27,12 +26,11 @@ time_series_plot <- expand(monthly_counts, region, date) %>%
   mutate(region = fct_reorder2(region, date, n, first2)) %>% 
   filter(n > 1000) %>% 
   ggplot(aes(x = date, y = n, color = region)) +
-  geom_line(na.rm = TRUE, show.legend = FALSE) +
-  #scale_x_continuous(limits = c(2016, 2024)) + 
+  geom_line(na.rm = TRUE) +
   scale_y_continuous(trans = "log10") +
-  scale_color_grey(name = "Субъект России") +
+  scale_color_discrete(name = "Субъект России") +
   labs(x = "Год", y = "Количество субъектов МСП") +
-  theme_bw()
+  theme_bw(base_size = 12, base_family = "Times New Roman")
 time_series_plot
 
 # Maps
@@ -66,7 +64,9 @@ regions_plot_agri_forestry <- rsmp_data %>%
     labels = function(breaks) c("Min", rep("", length(breaks) - 2), "Max"),
     palette = "YlGn",
     direction = 1) +
-  facet_wrap(~ac_type)
+  facet_wrap(~ac_type) +
+  theme_bw(base_size = 12, base_family = "Times New Roman") +
+  theme(legend.position = "bottom")
 regions_plot_agri_forestry
 
 ac_code_mapping_wp <- c("01.11.11" = "Пшеница", "01.13.31" = "Картофель")
@@ -92,10 +92,12 @@ regions_plot_wheat_potato <- rsmp_data %>%
     labels = function(breaks) c("Min", rep("", length(breaks) - 2), "Max"),
     palette = "YlGn",
     direction = 1) +
-  facet_wrap(~ac_type)
[…]

### 2024-01-07 Paper.Rmd
Translate plots, tables and diagram to English, start proofreading

@@
 options(knitr.kable.NA = "—")
+LOCALE = "ru"
 ```
@@
 
-**Использование открытых данных ФНС России для создания геопривязанного набора данных о субъектах малого и среднего предпринимательства в России**
+**Использование открытых данных налоговой службы для создания геопривязанного набора данных о субъектах малого и среднего предпринимательства в России**
 
@@
 
-В статье представлена технология (алгоритм, приложение), которая создаёт геопривязанный табличный набор данных о субъектах малого и среднего предпринимательства (МСП) в России на основе открытых данных Федеральной налоговой службы (ФНС России).
-Получаемый набор данных охватывает период с 2016 по 2023 гг.
-и всю территорию России.
-Он представляет собой полный список (генеральную совокупность) субъектов МСП и содержит информацию о категории, имени или наименовании, месте жительства или местонахождении, виде деятельности, доходах, расходах и численности работников организации или индивидуального предпринимателя.
-В пространственном отношении набор детализирован до населённых пунктов с указанием географических координат и кода ОКТМО; во временном отношении — до месяца или года в зависимости от характеристики (признака субъекта МСП); в отношении вида деятельности — до конкретных кодов ОКВЭД.
-В отличие от классической data paper, статья содержит не описание данных как таковых, а рассказывает о методике их получения и демонстрацию одного из возможных результатов применения технологии.
-Используя описанный алгоритм и его предложенную реализацию на языке Python, исследователь может самостоятельно обработать исходные данные и получить актуальную версию набора данных, адаптированную для его задач.
-Алгоритм оптимизирован так, чтобы для его выполнения было достаточно компьютера средней конфигурации, несмотря на большой (около 2 Тб) объём исходных данных.
-Создаваемый с помощью алгоритма набор данных может применяться как непосредственно для пространственного анализа МСП в России, так как и в качестве  относительно репрезентативного источника сведений об экономике России в разрезе конкретных организаций, выступая как открытая и бесплатная альтернатива коммерческим системам наподобие Spark и Контур.Фокус.
-Кроме того, благодаря координатам он может без дополнительной обработки использоваться для картографирования, а наличие кодов ОКТМО позволяет объединять его с официальными статистическими сведениями.
+В статье представлена технология для создания геопривязанного табличного набора данных о субъектах малого и среднего предпринимательства (МСП) в России на основе открытых данных Федеральной налоговой службы (ФНС России).
+Получаемый набор данных охватывает всю территорию России за 2016–2023 гг.
+Он представляет собой полный список субъектов МСП и содержит информацию о регистрационных номерах, категории, имени или наименовании, месте жительства или местонахождении, виде деятельности, доходах, расходах и численности работников.
+В пространственном отношении набор детализирован до населённых пунктов с указанием географических координат и кода муниципального образования (ОКТМО); во временном отношении — до месяцев или лет в зависимости от конкретного признака; в отношении вида деятельности — до отдельных кодов общероссийского классификатора (ОКВЭД).
+В отличие от классической data paper, статья не описывает итоговые данные как таковые, а излагает воспроизводимую методику их получения и демонстрирует её использование для получения набора данных о фирмах в сфере сельского, лесного хозяйства и рыболовства.
+С помощью предложенного алгоритма и его реализации на языке Python исследователь может самостоятельно обработать исходные данные и получить актуальную версию набора данных, адаптированную для его задач.
+Алгоритм оптимизирован так, чтобы для его выполнения было достаточно компьютера средней конфигурации, несмотря на сравнительно большой (около 2 Тб в несжатом виде) объём исходных данных.
+Получаемый с помощью технологии набор данных может применяться как непосредственно для пространственного анализа МСП в России, так и в качестве открытого, бесплатного и относительно репрезентативного источника сведений об экономике России в разрезе конкретных организаций, позволяя отчасти заменить коммерческие системы бизнес-аналитики наподобие Spark и Контур.Фокус.
+Кроме того, благодаря координатам набор данных может без дополнительной обработки использоваться для картографирования, а наличие кодов ОКТМО позволяет объединять его с официальными статистическими сведениями.
 
@@
 
-Современные общественно-географические исследования часто опираются на работу с данными.
-Некоторые специалисты используют на агрегированную экономическую и демографическую статистику.
+Современные общественно-географические исследования часто основаны на работе с данными.
+Некоторые специалисты используют агрегированную экономическую и демографическую статистику.
 Так, О.
@@
 А.
-Балина с соавторам, рассматривая воспрос о территориях опережающего социально-экономического развития, приводит подсчёты по субъектам Российской Федерации [@Balina2021]; Г.
+Балина с соавторам, рассматривая вопрос о территориях опережающего социально-экономического развития, приводит подсчёты по субъектам Российской Федерации [@Balina2021]; Г.
 М.
-Лаппо, описывая свою концепцию разнообразия городов, опирается на сведения о числе жителей в населённых пунктах в разные периоды времени [@Lappo2019].
+Лаппо, описывая свою концепцию разнообразия городов, оперирует сведениями о числе жителей в населённых пунктах в разные периоды времени [@Lappo2019].
 Другие исследователи используют большие массивы дезагрегированных данных, собирая их вручную в общедоступных источниках и используя сложные методы их обработки.
@@
 В.
-Красновы для своей работы, посвящённой трансформации сети расселения в Псковской области, собирали публичные сведения о местоположении пользователей социальной сети «ВКонтакте» и открытые данные о поисковых запросах к «Яндексу» [@Krasnov2020].
+Красновы для своей работы, посвящённой трансформации сети расселения в Псковской области, собирали публичные сведения о местоположении пользователей социальной сети «ВКонтакте» и открытую информацию о поисковых запросах к «Яндексу» [@Krasnov2020].
 Есть работы, опирающиеся на обобщённую статистику, но использующие продвинутые методы её анализа.
@@
 В.
-Федорченко взял данные Международной организации автопроизводителей, рассчитал на их основе индексы и дополнил их картографической визуализацией, что позволило эффективно проанализировать территориальные сдвиги в мировой автомобильной промышленности [@Fedorchenko2017].
+Федорченко взял данные Международной организации автопроизводителей, рассчитал производные индексы и дополнил их картографической визуализацией, что позволило эффективно проанализировать территориальные сдвиги в мировой автомобильной промышленности [@Fedorchenko2017].
 «Большие данные» также проникают в общественно-географическую науку, что отметили ещё десять лет назад М.
@@
 Недавний пример такого рода исследований — работа Вэй Лю и Вэньцзэ У, которые используют сведения о пользователях социальной сети WeChat для изучения связей между типом землепользования и деятельностью людей [@Liu2020].
-Таким образом, общая тенденция такова: для исследований используются данные, хотя объём, тип, качество этих данных и технологии их обработки отличаются.
+Таким образом, общая тенденция такова: географы используют данные для исследований, хотя объём, тип, качество этих данных и технологии их обработки отличаются.
 
@@
 Сведения являются дезагрегированными, то есть детализированными до уровня отдельных фирм (организаций и индивидуальных предпринимателей).
-Помимо структурированных адресов, набор содержит географические координаты и коды муниципальных образований.
+Помимо структурированных адресов, набор содержит географические координаты и коды ОКТМО муниципальных образований.
 В отношении субъектов малого и среднего предпринимательства (МСП) набор данных представляет собой генеральную совокупность.
@@
 Статья, помимо введения, включает в себя несколько разделов.
-В обзоре литературы приводится краткая информация об использовании различных источников данных в общественно-географических исследованиях и через это обосновывается необходимость в предложенной технологии.
-Раздел «Исходные данные» содержит описание первичных наборов данных, которые задействуются в технологии, а раздел «Методика обработки данных» излагает саму технологию.
-В разделе «Тестирование технологии» описываются и визуализируются процесс и результаты опробования технологии в отношении сведений о фирмах, ведущих деятельность в сфере сельского и лесного хозяйства, охоты и рыболовства.
-Раздел «Возможность генерализации результатов» посвящён оценке репрезентативности полученного набора данных применительно к экономике России в целом.
-В «Обсуждении» затрагиваются ключевые спорные моменты касательно предложенной технологии.
+В обзоре литературы приводится краткая информация об использовании различных источников данных в общественно-географических исследованиях и через это обосновывается необходимость предложенной технологии.
+Раздел «Исходные данные» содержит описание первичных наборов данных, с которыми работает алгоритм, а раздел «Методика обработки данных» излагает саму технологию.
+В разделе «Демонстрация работы технологии» описываются и визуализируются процесс и результаты опробования технологии в отношении сведений о фирмах, ведущих деятельность в сфере сельского и лесного хозяйства, охоты и рыболовства.
+В «Обсуждении» затрагиваются ключевые спорные моменты касательно предложенной технологии и анализируется репрезентативность получаемого с её помощью набора данных.
 «Выводы» подводят итоги работы.
@@
 
-Типичным источником данных для исследований общественной географии России являются ресурсы Федеральной службы государственной статитстики (Росстата).
+Типичным источником данных для исследований общественной географии России являются ресурсы Федеральной службы государственной статистики (Росстата).
 К ним относятся, в частности, размещённые на официальном сайте этого органа данные официальной статистики, переписей и статистических обследований, база данных показателей муниципальных образований (БДМО) и единая межведомственная информационно-статистическая система (ЕМИСС).
-О.
+Примеров их использования в географии множество: О.
 В.
@@
 В.
-Петров на базе региональной, муниципальной статистики и данных иных органов власти формулирует предложения по социально-экономическому развитию юга Тюменской области [@Petrov2021]; М.
+Петров с помощью региональной, муниципальной статистики и данных иных органов власти формулирует предложения по социально-экономическому развитию юга Тюменской области [@Petrov2021]; М.
 А.
 Саранча оценивает развитие гостиничного и ресторанного бизнеса, задействуя результаты сплошного наблюдения Росстата за малым и среднем предпринимательством [@Sarancha2014].
-Данные Росстата также используются для создания производных наборов данных, как в работе И.
+Сведения Росстата также используются для создания производных наборов данных, как в работе И.
 Ю.
@@
 Причины этого бывают разные.
-В некоторых случаях альтернативные источники информации используются для проверки или корректировки данных Росстата.
+В некоторых случаях альтернативные источники информации используются для проверки или корректировки Росстата.
 Например, Р.
@@
 
-Одним из альтернативных источников информации для географов-обществоведов являются административные данные Федеральной налоговой службы России (ФНС России).
+Одним из альтернативных источников информации для географов-обществоведов являются административные данные ФНС России.
 Они публикуются на официальном сайте этого органа власти или связанных с ним ресурсах.
@@
 Тем не менее, есть несколько наборов данных, содержащих дезагрегированные (исходные, первичные, «сырые») сведения.
-Наибольшее внимание среди них привлекает реестр субъектов малого и среднего предпринимательства (реестра МСП), который регулярно и полностью размещается на сайте службы и содержит сведения о конкретных организациях и индивидуальных предпринимателях.
+Наибольшее внимание среди них привлекает реестр субъектов малого и среднего предпринимательства (реестр МСП), который регулярно и полностью размещается на сайте службы и содержит сведения о конкретных организациях и индивидуальных предпринимателях.
 
@@
 Общепринятого источника геопривязанной информации о фирмах в России нет.
-Исследователи и прикладные специалисты обычно опираются на данные коммерческих [@Lachininsky2021] или открытых геоинформационных сервисов, систем бизнес-аналитики [@Moiseeva2016; @Kuzminov2021], уже упомянутого реестра МСП [@Gumenyuk2022].
+Исследователи и прикладные специалисты обычно опираются на данные коммерческих [@Lachininsky2021] или открытых геоинформационных сервисов, систем бизнес-аналитики [@Moiseeva2016; @Kuzminov2021], а также упомянутого выше реестра МСП [@Gumenyuk2022].
 Как правило, использование этих способов сопряжено с двумя проблемами: данные предоставляются за плату либо требуется приложить значительные усилия для их приведения в подходящий для анализа формат.
-Системы бизнес-аналитики, такие как Spark и Ruslana, содержат много разнообразной информации, но платные.
-Выгрузка данных из частных геоинформационных систем наподобие 2GIS и Яндекс.Карт также стоит денег.
-Открытые сервисы наподобие OpenStreetMap или реестра МСП бесплатны, но выгрузки из них имеют большой объём и сложную структуру, поэтому исследователю приходится тратить много усилий на техническую сторону работы, привлекать сторонних специалистов или отказываться от использования такого источника данных.
+Системы бизнес-аналитики, такие как Spark, Контур.Фокус и Ruslana, содержат много разнообразной информации, но платные.
+Выгрузка данных из частных геоинформационных систем наподобие 2GIS и Яндекс.Карт также стоит денег, а кроме того, для неё нет удобного интерфейса.
+Открытые платформы наподобие OpenStreetMap или реестра МСП бесплатны, но выгрузки из них имеют большой объём и сложную структуру, поэтому исследователю приходится тратить много усилий на техническую сторону работы, привлекать сторонних специалистов или отказываться от использования такого источника данных.
 Формально полный список российских организаций и индивидуальных предпринимателей содержится в двух государственных реестрах (соответственно Едином государственном реестре юридических лиц и Едином государственном реестре индивидуальных предпринимателей), однако ни один из них не распространяется в полном виде — можно лишь извлекать отдельные записи вручную через веб-сервис ФНС России либо автоматизированно за плату через программный интерфейс, что не даёт доступ к генеральной совокупности компаний и вряд ли позволяет построить репрезентативную выборку без чрезмерных затрат и усилий.
-Таким образом, актуален вопрос об удобном способе, который позволил бы бесплатно и легально получить максимально полный набор геопривязанных данных о российских фирмах.
+Таким образом, актуален вопрос об удобном инструменте, который позволил бы бесплатно и легально получить максимально полный набор геопривязанных данных о российских фирмах.
 
@@
 
-Технология опирается на один основной источник исходных данных и несколько вспомогательных источников.
+Технология опирается на один основной источник исходных данных и несколько вспомогательных источников. Все они являются общедоступными и бесплатными для использования.
 
@@
 Она предусмотрена ст.
-4.1 ФЗ «О развитии малого и среднего предпринимательства в Российской Федерации».
+4.1 Федерального закона от 24.07.2007 № 209-ФЗ «О развитии малого и среднего предпринимательства в Российской Федерации».
 Субъекты малого и среднего предпринимательства (субъекты МСП) — это коммерческие юридические лица и индивидуальные предприниматели, соответствующие критериям по доходу, числу работников и некоторым другим показателям.
@@
 Норма о реестре появилась в законе в конце 2015 г.
-На практике реестр размещён на сайте ФНС России в форме открытых данных в виде ежемесячных полных выгрузок, первая из которых датируется августом 2016 г.[^1]
+На практике реестр размещён на сайте ФНС России в форме открытых данных в виде ежемесячных полных выгрузок, первая из которых датируется августом 2016 г., а последняя доступная на момент подготовки статьи — декабрём 2023 г.[^1]
 
@@
 
-Технически реестр МСП — это коллекция zip-архивов xml-файлов.
+Технически реестр МСП в форме открытых данных — это коллекция zip-архивов xml-файлов.
 Каждый zip-архив — это «срез» (выгрузка, дамп, dump) реестра на определённую дату.
 Коллекция zip-архивов тем самым образует временной ряд (time series).
-Каждый zip-архив содержит множество (несколько тысяч) xml-файлов.
+Каждый zip-архив содержит множество (обычно 6-7 тысяч) xml-файлов.
 Каждый xml-файл включает в себя информацию о множестве (обычно ровно 900) субъектов МСП.
@@
 
-В качестве вспомогательных источников информации используются ещё два реестра открытых данных ФНС России.
+В качестве вспомогательных источников информации используются ещё два набора открытых данных ФНС России.
 
 1.
-Набор открытых данных ФНС России «Сведения о суммах доходов и расходов по данным бухгалтерской (финансовой) отчетности организаций»[^2].
+Набор «Сведения о суммах доходов и расходов по данным бухгалтерской (финансовой) отчетности организаций»[^2].
 Содержит ИНН, суммарные доходы и расходы за год, предшествующий году публикации.
@@
 по настоящее время[^3].
-В отличие от реестра МСП, содержит только информацию об органзациях, но не об индивидуальных предпринимателях.
-Технически аналогичен реестру МСП: представляет собой временной ряд zip-архивов xml-файлов, где каждый архив соответствует определённой дате, а xml-файл внутри архива содержит информацию о множестве организаций.
+В отличие от реестра МСП, содержит только информацию об организациях, но не об индивидуальных предпринимателях.
+Технически аналогичен реестру МСП: представляет собой временной ряд zip-архивов xml-файлов, где каждый архив соответствует определённой дате, а каждый xml-файл внутри архива содержит информацию о множестве организаций.
 
@@
 2.
-Набор открытых данных ФНС России «Сведения о среднесписочной численности работников организации»[^4].
+Набор «Сведения о среднесписочной численности работников организации»[^4].
 Содержит ИНН и среднесписочную численность работников за год, предшествующий году публикации сведений.
@@
 ## Геоданные
+
 Для нормализации адресов, расстановки координат и кодов ОКТМО используются два справочника.
@@
 
-Все перечисленные выше источники данных являются общедоступными и бесплатными для использования.
-
 # Методика обработки данных
 
-Предложенный алгоритм обработки данных состоит из шести шагов. Он изображён на рис. 1 и описан ниже.
+Предложенный алгоритм обработки данных состоит из пяти шагов, некоторые из которых разбиты на подзадачи. Он изображён на рис. 1 и описан ниже.
 
-## Шаг 1: загрузка данных ФНС России
+## Загрузка данных ФНС России (1)
 
@@
 Загрузка ускоряет дальнейшую обработку.
-Кроме того, благоря ей создаётся резервная копия исходных данных, что страхует от риска удаления наборов с сайта налоговой службы.
-Теоретически возможна загрузка и обработка данных в облачном сервисе, таком как Yandex Cloud, но в текущей реализации алгоритма она не представлена.
+Кроме того, благодаря ей создаётся резервная копия исходных данных, что страхует от риска удаления наборов с сайта налоговой службы.
+Теоретически возможна загрузка и обработка данных в облачном сервисе, таком как Yandex Cloud.
 
-## Шаг 2а: фильтрующее извлечение данных реестра МСП
+## Фильтрующее извлечение данных реестра МСП (2a)
 
@@
 Извлекаются следующие характеристики: вид, категория, ИНН, ОГРН, имя или наименование, дата включения в реестр, дата выгрузки сведений в набор открытых данных, юридический адрес (с точностью до населённого пункта), основной код ОКВЭД.
-При извлечении данные могут фильтроваться по одному или нескольким кодам или группе кодов ОКВЭД — тогда таблица будет содержать сведения только о части субъектов МСП.
+При извлечении данные могут фильтроваться по одному или нескольким кодам или группе кодов ОКВЭД.
+Отфильтрованная таблица будет содержать сведения только из необходимой части реестра МСП, что ускоряет работу и сокращает использование дискового пространства.
 
 Реализация данного шага оптимизирована, чтобы уменьшить время обработки, расход оперативной памяти и места для хранения временных файлов.
-Наивный подход к преобразованию исходных архивов в таблицу выглядел бы следующим образом: извлечь xml-файлы из архива, прочитать их полностью, сконвертировать в таблицу, выбрать необходимые признаки и наблюдения.
+Наивный подход к преобразованию исходных архивов в таблицу выглядел бы следующим образом: извлечь все xml-файлы из архива, прочитать каждый из них полностью, преобразовать в таблицу, выбрать необходимые признаки и наблюдения.
 Он привёл бы к образованию большого объёма временных файлов (десятки гигабайт) и увеличению времени на их обработку (чтение и конвертация в таблицу осуществлялись бы для всех данных).
 Оптимизированный подход, реализованный автором, предполагает чтение отдельных xml-файлов из архива с распаковкой «на лету» (вместо распаковки всего архива), фильтрацию по коду ОКВЭД в процессе обработки xml-файла (вместо последующей фильтрации) и извлечение нужных признаков напрямую из xml-файла (вместо последующего отбора).
+Он в несколько раз быстрее и не образует временные распакованные файлы.
 
-## Шаг 2б: извлечение данных о доходах и расходах
+## Извлечение данных о доходах и расходах (2b)
 
@@
 
-## Шаг 2в: извлечение данных о среднесписочной численности работников
+## Извлечение данных о среднесписочной численности работников (2c)
 
-Аналогичен шагу 2б, отличается входными данными (архивы из набора о среднесписочной численности работников) и признаками в результирующей таблице (ИНН, дата выгрузки, среднесписочная численность работников).
+Аналогичен шагу 2b, отличается входными данными (zip-архивы из набора о среднесписочной численности работников) и признаками в результирующей таблице (ИНН, дата выгрузки, среднесписочная численность работников).
 
-## Шаг 3а: агрегация данных о субъектах МСП
+## Агрегация данных о субъектах МСП (3a)
 
-Таблицы в csv-формате, получившиеся на шаге 2а, сводятся в одну таблицу, из которой удаляются строки, в которых совпадает набор значимых признаков: ИНН, ОГРН, имя или наименование, вид, категория, код ОКВЭД, адрес регистрации.
-После удаления дублей в каждой строке простравляются две даты: дата начала действия и дата окончания действия.
+Таблицы в csv-формате, получившиеся на шаге 2a, сводятся в одну таблицу, из которой удаляются дубли, то есть строки, в которых совпадает набор значимых признаков: ИНН, ОГРН, имя или наименование, вид, категория, код ОКВЭД, адрес регистрации.
+После удаления дублей в каждой строке проставляются две даты: дата начала действия и дата окончания действия.
 Эти даты указывают на период, в течение которого актуальна информация, содержащаяся в строке.
@@
 — под другим, то в агрегированной таблице она будет представлена двумя строками.
-У первой дата начала — 10.08.2016, дата окончания — 10.03.2021; у второй дата начала — 10.04.2021, дата окончания — 10.11.2023 (условная дата последнего архива исходных данных).
-Кроме того, на данном этапе происходят вспомогательные преобразования, цель которых — исправить неточности в исходых данных и более точно определить дублирующиеся строки.
+У первой дата начала — 10.08.2016, дата окончания — 10.03.2021; у второй дата начала — 10.04.2021, дата окончания — 10.12.2023 (условная дата последнего архива исходных данных).
+Кроме того, на данном этапе происходят вспомогательные преобразования, цель которых — исправить неточности в исходных данных и корректно определить дубли.
 Например, в реестре МСП тип населённого пункта может быть указан полностью («город») или сокращённо («г.»), и вспомогательная обработка перед агрегацией позволяет игнорировать такие различия и воспринимать адрес как неизменный.
 
-## Шаг 3б: агрегация данных о доходах, расходах
+## Агрегация и фильтрация данных о доходах, расходах (3b)
 
-Таблицы в csv-формате, получившиеся на шаге 2б, объединяются в одну так, что для каждой организации указаны доходы и расходы за конкретный год.
-Объединённая таблица фильтруется по ИНН так, что в ней остаются только те организации, которые есть в агрегированных данных из реестра МСП, полученных на шаге 3а.
+Таблицы в csv-формате, получившиеся на шаге 2b, объединяются в одну так, что для каждой организации указаны доходы и расходы за каждый доступный год.
+Объединённая таблица фильтруется по ИНН так, что в ней остаются только те организации, которые есть в агрегированных данных из реестра МСП, полученных на шаге 3a.
 Результат сохраняется в итоговый csv-файл.
 
-## Шаг 3в: агрегация данных о работниках
+## Агрегация данных о работниках (3c)
 
-Аналогичен шагу 3б.
+Аналогичен шагу 3б, но на вход подаются таблицы, полученные на шаге 2c.
 
-## Шаг 4: Геопривязка субъектов МСП
+## Геопривязка субъектов МСП (4)
 
-Каждая запись в csv-таблице, полученной на этапе 3а, дополняется кодом ОКТМО и географическими координатами населённого пункта, в котором находится субъект МСП.
-Для этого используется адрес регистрации и вспомогательные наборы геоданных.
-Кроме того, вместо исходного адреса в наборе данных появляется нормализованный, а исходный адрес в виде структурированной текстовой строки сохраняется как отдельный атрибут каждой записи.
-После нормализации адресов набор данных повторно проверяется на дубли, и вновь образовавшиеся дубли удаляются, а даты актуальности сведений пересчитываются.
+Каждая запись в csv-таблице, полученной на этапе 3a, дополняется кодом ОКТМО и географическими координатами населённого пункта, в котором находится субъект МСП.
+Для этого используются адрес регистрации и вспомогательные наборы геоданных.
+Кроме того, исходный адрес в наборе данных заменяется на нормализованный и сохраняется как отдельный атрибут каждый записи в виде структурированной текстовой строки.
+После нормализации адресов набор данных повторно проверяется на дубли.
+Дубли, образовавшиеся из-за нормализации адресов, удаляются, а даты актуальности сведений пересчитываются.
 
-## Шаг 5: создание панельного представления
+## Создание панельного представления (5)
 
-Таблица, полученная на этапе 4, преобразуется в панельный временной ряд, где представлены характеристики каждого субъекта МСП за каждый год с 2016 по 2023.
-Содержательно новая таблица не отличается от старой, но более удобна для анализа, поскольку вместо компактного, но не слишком интуитивного и весьма технологизированного формата с датами актуальности сведения в ней идут по годам.
+Таблица, полученная на этапе 4, объединяется с таблицами, полученными на этапах 3b и 3c, и преобразуется в панельный временной ряд, где представлены характеристики каждого субъекта МСП за каждый год с 2016 по 2023.
+Содержательно такая таблица не отличается от первоначальных, но более удобна для анализа: она одна вместо трёх, и вместо компактного, но не слишком интуитивного и весьма технологизированного формата с датами актуальности информация в ней представлена по годам.
 Такой вид более привычен для исследователей, а также упрощает объединение набора данных с официальной статистикой, которая также обычно даётся в разрезе календарных лет.
@@
 ## Реализация алгоритма
+
 Описанный выше алгоритм реализован как набор скриптов на языке программирования Python.
@@
 Для агрегации используется система обработки больших данных Apache Spark.
+Геопривязка и создание панельного представления реализованы с помощью библиотеки Pandas.
 Исходный код опубликован по адресу: https://github.com/PavelSyomin/russian-smb-companies
@@
 
-```{r fig-8-count-employees-profit-by-region, fig.cap = "Распределение числа зарегистрированных субъектов МСП, их суммарной прибыли и суммарной численности работников в демонстрационном наборе в 2021 г.
+```{r fig-8-count-employees-profit-by-region, fig.cap = "Рис.
+8.
+Распределение числа зарегистрированных субъектов МСП, их суммарной прибыли и суммарной численности работников в демонстрационном наборе в 2021 г.
 по регионам (субъектам федерации)"}

### 2024-01-07 Paper_meta_en.Rmd
Translate plots, tables and diagram to English, start proofreading

@@
 link-bibliography: false
-nocite: |
-  @*
+nocite: "@*"
 ---
@@
 options(knitr.kable.NA = "—")
+LOCALE = "en"
 ```
@@
 
-Syomin Pavel Olegovich
+Pavel O. Syomin
 
@@
 
-Perm State National University, 15 Bukireva st., Perm, 614068, Russian Federation
+Perm State University, 15 Bukireva st., Perm, 614068, Russian Federation
 

### 2024-01-07 Diagram.R

@@ -2,6 +2,59 @@ library(DiagrammeR)
 library(DiagrammeRsvg)
 library(rsvg)
 
+labels <- data.frame(
+  ru = c(
+    "Реестр МСП\n(N zip-архивов)",
+    "Извлечённые данные\n(N csv-файлов)",
+    "Объединённая таблица\n(csv-файл)",
+    "Геопривязанная таблица\n(csv-файл)",
+    "Извлечение, фильтрация,\nтрансформация (2а)",
+    "Агрегирование (3а)",
+    "Геопривязка (4)",
+    "Доходы и расходы\n(M zip-архивов)",
+    "Извлечённые данные\n(M csv-файлов)",
+    "Отфильтрованная таблица\n(csv-файл)",
+    "Извлечение и трансформация (2б)",
+    "Агрегирование и фильтрация(3б)",
+    "Число работников\n(K zip-архивов)",
+    "Извлечённые данные\n(K csv-файлов)",
+    "Отфильтрованная таблица\n(csv-файл)",
+    "Извлечение, трансформация (2в)",
+    "Агрегирование и фильтрация (3в)",
+    "Открытые данные ФНС России",
+    "Панельное представление",
+    "Справочник\nнаселённых пунктов",
+    "Загрузка (1)",
+    "Объединение и трансформация (5)"
+  ),
+  en = c(
+    "SMB registry\n(N zip archives)",
+    "Extracted data\n(N csv files)",
+    "Joint table\n(a csv file)",
+    "Georeferenced table\n(a csv file)",
+    "Extract, filter,\ntransform (2а)",
+    "Aggregate (3а)",
+    "Georeference (4)",
+    "Revenue&expediture\n(M zip archives)",
+    "Extracted data\n(M csv files)",
+    "Filtered table\n(a csv file)",
+    "Extract and transform (2б)",
+    "Aggregate and filter(3б)",
+    "Employees count\n(K zip files)",
+    "Extracted data\n(K csv files)",
+    "Filtered table\n(a csv file)",
+    "Extract and transform (2в)",
+    "Aggregate and filte (3в)",
+    "FTS open data",
+    "Panel view",
+    "Settlements\nlookup table",
+    "Download (1)",
+    "Join and transform (5)"
+  )
+)
+
+labels <- labels[[LOCALE]]
[…]

### 2024-01-07 Plots.R

@@ -6,6 +6,47 @@ library(scales)
 library(sf)
 library(tidyr)
 
+labels <- data.frame(
+  ru = c(
+    "Субъект России",
+    "Год",
+    "Количество субъектов МСП",
+    "Сельское хозяйство",
+    "Лесное хозяйство",
+    "Пшеница",
+    "Картофель",
+    "Количество фирм",
+    "Группа ОКВЭД",
+    "Сельское хозяйство",
+    "Лесоводство и лесозаготовки",
+    "Рыболовство и рыбоводство",
+    "Изменение за 2016–2023",
+    "Количество фирм",
+    "Численность работников",
+    "Прибыль"
+  ),
+  en = c(
+    "Region",
+    "Year",
+    "SMBs count",
+    "Agriculture",
+    "Forestry",
+    "Wheat",
+    "Potato",
+    "SMBs count",
+    "Activity group",
+    "Agriculture",
+    "Forestry",
+    "Fishery",
+    "Change, 2016–2023",
+    "SMBs count",
+    "Employees count",
+    "Total profit"
+    
+  )
+)
+labels <- labels[[LOCALE]]
+
 rsmp_data_path <- "../tax-service-opendata/rsmp/reestr_group_A/data_product.csv"
 rsmp_data <- read_csv(rsmp_data_path)
 
@@ -28,8 +69,8 @@ time_series_plot <- expand(monthly_counts, region, date) %>%
   ggplot(aes(x = date, y = n, color = region)) +
   geom_line(na.rm = TRUE) +
   scale_y_continuous(trans = "log10") +
-  scale_color_discrete(name = "Субъект России") +
-  labs(x = "Год", y = "Количество субъектов МСП") +
+  scale_color_discrete(name = labels[1]) +
+  labs(x = labels[2], y = labels[3]) +
   theme_bw(base_size = 12, base_family = "Times New Roman")
 time_series_plot
 
@@ -42,7 +83,7 @@ regions <- regions_boundaries %>%
   left_join(regions, by = c("shapeISO" = "iso_code")) %>% 
   select(name)
 
-ac_code_mapping <- c("01" = "Сельское хозяйство", "02" = "Лесное хозяйство")
+ac_code_mapping <- c("01" = labels[4], "02" = labels[5])
 regions_plot_agri_forestry <- rsmp_data %>%
   mutate(
     ac_start = substr(activity_code_main, 1, 2),
@@ -69,7 +110,7 @@ regions_plot_agri_forestry <- rsmp_data %>%
[…]

### 2024-01-07 StatsHelper.R

@@ -3,6 +3,31 @@ library(ggplot2)
 library(readr)
 library(tidyr)
 
+labels <- data.frame(
+  ru = c(
+    "Исходные данные",
+    "Промежуточные файлы",
+    "Итоговые таблицы",
+    "Панельное представление",
+    "Группы ОКВЭД",
+    "Все",
+    "Все, кроме O, P, U",
+    "Коэффициент корреляции Спирмена",
+    "Число субъектов России"
+  ),
+  en = c(
+    "Source data",
+    "Intermediary files",
+    "Resulting tables",
+    "Panel view",
+    "Activity groups",
+    "All",
+    "All except for O, P, U",
+    "Spearman's rho",
+    "Number of regions"
+  )
+)
+labels <- labels[[LOCALE]]
 rsmp_raw_data_path <- "../tax-service-opendata/rsmp/xml"
 
 raw_files <- list.files(rsmp_raw_data_path, pattern = "*.zip", full.names = TRUE)
@@ -54,7 +79,7 @@ sizes_funnel <- data.frame(
   rsmp = c(rsmp_source_files_size, rsmp_intermediate_data_files_size, rsmp_out_file_size, rsmp_panel_file_size),
   revexp = c(revexp_source_files_size, revexp_intermediate_data_files_size, revexp_out_file_size, NA),
   sshr = c(sshr_source_files_size, sshr_intermediate_data_files_size, sshr_out_file_size, NA),
-  row.names = c("Исходные данные", "Промежуточные файлы", "Итоговые таблицы", "Панельное представление")
+  row.names = c(labels[1], labels[2], labels[3], labels[4])
 )
 sizes_funnel$total <- rowSums(sizes_funnel, na.rm = TRUE)
 
@@ -93,7 +118,7 @@ counts_funnel <- data.frame(
   rsmp = c(rsmp_source_obs_count, rsmp_intermediate_obs_count, rsmp_out_obs_count, rsmp_panel_obs_count),
   revexp = c(revexp_intermediate_obs_count, revexp_intermediate_obs_count, revexp_out_obs_count, NA),
   sshr = c(sshr_intermediate_obs_count, sshr_intermediate_obs_count, sshr_out_obs_count, NA),
-  row.names = c("Исходные данные", "Промежуточные файлы", "Итоговые таблицы", "Панельное представление")
+  row.names = c(labels[1], labels[2], labels[3], labels[4])
 )
 
[…]

### 2024-01-08 Paper.Rmd
Proofreading

@@
 
-Аналогичен шагу 3б, но на вход подаются таблицы, полученные на шаге 2c.
+Аналогичен шагу 3b, но на вход подаются таблицы, полученные на шаге 2c.
 
@@
 Геопривязка и создание панельного представления реализованы с помощью библиотеки Pandas.
-Исходный код опубликован по адресу: https://github.com/PavelSyomin/russian-smb-companies
+Исходный код опубликован в личном репозитории автора статьи: https://github.com/PavelSyomin/russian-smb-companies
 
@@
 
-С помощью описанного алгоритма был подготовлен набор данных по субъектам МСП, ведущих деятельность в сфере сельского, лесного хозяйства, охоты, рыболовства и рыбоводства.
-Для этого использовалось следующее условие фильтрации: основной код ОКВЭД организации или предпринимателя находится в группе А классификатора в редакции 2014 г.
-На «вход» алгоритм принимал три набора данных ФНС России за весь доступный на момент подготовки статьи период: с августа 2016 г.
+С помощью описанного алгоритма был подготовлен пробный набор данных о субъектах МСП, ведущих деятельность в сфере сельского, лесного хозяйства, охоты, рыболовства и рыбоводства.
+Фильтрация вида деятельности производилась по следующему условию: основной код ОКВЭД организации или предпринимателя находится в группе А классификатора.
+Использовались исходные данные за весь доступный период: с августа 2016 г.
 для реестра МСП, с октября 2019 г.
 для сведений о доходах и расходах организаций; с марта 2020 г.
-для сведений о среднесписочной численности работников организаций (дата окончания везде — ноябрь 2023 г.).
-Результат работы алгоритма — несколько csv-файлов.
-Далее в подразделах представлена техническая статистика выполнения алгоритма с акцентом на изменение объёма обрабатываемых данных и описание базовых возможностей по анализу полученных наборов данных с упором на визуализацию, дающую наглядное представление о перспективах применения технологии как инструмента создания наборов данных для экономико-географических исследований.
+для сведений о среднесписочной численности работников организаций.
+Далее в подразделах представлена техническая статистика выполнения алгоритма и показаны базовые возможности по анализу полученного набора данных.
+Акцент в технической части делается на изменении количества обрабатываемых данных от шага к шагу алгоритма: благодаря этому видно, что предложенный метод помогает справиться с большим объёмом исходных сведений и превратить сотни гигабайт «сырых» данных в несколько таблиц, размер которых меньше на порядки.
+Аналитическая часть сфокусирована на картографической визуализации, поскольку картографирование — один из наиболее характерных географических методов.
+Она призвана показать пространственную, временную и структурную детализацию полученного набора данных, возможность вычислять на его основе статистические показатели с произвольной группировкой и отслеживать конкретные фирмы.
 
@@
 
-Время работы алгоритма — около 15 часов. Больше всего заняло извлечение данных из архивов (этап 2).
+Время создания пробного набора данных — около 15 часов. Больше всего занимает извлечение информации из архивов (шаг 2).
 
-В таблице 1 приведена информация об уменьшении объёма данных в процессе работы алгоритма, а в таблице 2 — о фильтрации записей.
+В таблице 1 приведена информация об уменьшении объёма данных в процессе работы алгоритма, а в таблице 2 — об уменьшении количества записей.
 Из таблиц видно, что технология многократно уменьшает объём обрабатываемых сведений: от сотен гигабайт исходных до нескольких десятков или сотен мегабайт итоговых наборов данных.
-Одновременно уменьшается и количество записей в данных с сотен миллионов до сотен тысяч.
-Таким образом, вместо десятков объёмных архивов с исходными данными в многомерном xml-формате исследователь может работать с относительно привычной и компактной двумерной таблицей.
-Итоговые таблицы имеют небольшой объём и поэтому могут обрабатываться без инструментов анализа больших данных.
-Панельное представление слегка теряет в компактности по сравнению с итоговыми таблицами, но выигрывает благодаря более удобной форме представления временного ряда.
+Похожим образом уменьшается и количество записей.
+В результате вместо десятков больших архивов с исходными данными в многомерном xml-формате исследователь получает относительно небольшие двумерные таблицы.
+Панельное представление слегка уступает итоговым таблицам в компактности, но отличается более удобной формой представления временного ряда.
 
@@
 
-Набор данных позволяет отследить те или иные количественные характерстики с точностью до месяца.
+Набор данных позволяет отследить необходимые количественные характеристики с точностью до месяца.
 На рис.
-2 показаны 20 субъектов России с наибольшим количеством зарегистрированных субъектов МСП и изменение этого количества с августа 2016 г.
+2 показаны 15 субъектов России с наибольшим числом зарегистрированных субъектов МСП и изменение этого числа с августа 2016 г.
 по ноябрь 2023 г.
@@
 
-Нормализованные названия позволяют агрегировать информацию по субъектам федерации.
-На картограмах (рис.
-3) представлено количество субъектов МСП, которые вели деятельность в сфере лесного (слева) и сельского (справа) хозяйства в 2021 г.
+Нормализованные адреса позволяют агрегировать информацию по субъектам федерации.
+На рис.
+3 представлено количество субъектов МСП, которые вели деятельность в сфере лесного (слева) и сельского (справа) хозяйства в 2021 г.
 Распределение ожидаемо разное: наибольшее количество фирм в сфере сельского хозяйства расположено в Черноземье и на юге России, также заметны животноводческие регионы; лесное хозяйство сосредоточено на севере европейской части страны, в Восточной Сибири и на Дальнем Востоке.
@@
 В отличие от официальной статистики, получившийся набор данных позволяет проанализировать пространственное распределение с точностью до конкретных видов деятельности, а не крупных агломератов.
-На картограммах (рис.
-4) представлено количество субъектов МСП, ведущих деятельность по выращиванию пшеницы (код 01.11.11) и картофеля (код 01.13.31).
-Здесь заметны некоторые странности (например, в лидерах по картофелю оказалась Якутия), которые, видимо, связаны с небольшим количеством субъектов МСП в каждом регионе: обычно у фирм в качестве основного указан более общий код, в то время как картограмма строилась по максимально детализированным.
+На рис.
+4 показано количество субъектов МСП, ведущих деятельность по выращиванию пшеницы (код 01.11.11) и картофеля (код 01.13.31) в 2021 г.
+Здесь заметны некоторые странности (например, в лидерах по картофелю оказалась Якутия), которые, видимо, связаны с небольшим количеством субъектов МСП с указанными кодами: обычно у фирм в качестве основного указан более общий код, в то время как для демонстрации картограмма строилась по максимально детализированным.
 
@@
 
-Набор данных, помимо названий субъектов содержит также географические координаты, название населённого пункта и территориальной единицы (района, муниципалитета) внутри субъекта.
-Это позволяет выполнять более детализированные агрегации и картографически визуализировать их.
-На картограмме (рис.
-5) показано количество субъектов МСП в разных населённых пунктах, а также преимущественная специализация населённых пунктов исходя из того, каких фирм в нём больше всего.
+Набор данных помимо названия субъекта содержит географические координаты, название населённого пункта и территориальной единицы (района, муниципалитета) внутри субъекта.
+Это позволяет выполнять более детализированные агрегации и показывать их на карте.
+На рис.
+5 показано количество субъектов МСП в населённых пунктах России в 2021 г, а также преимущественная специализация населённых пунктов исходя из того, каких фирм в нём больше всего.
 Показаны только населённые пункты более чем с 50 фирмами.
-Ещё одна аналогичная картограмма (рис.
-6) изображает количество субъектов МСП по видам деятельности в населённых пунктах Свердловской области, причём без ограничения на количество фирм.
+Аналогичную картограмму можно построить и для региона: рис.
+6 изображает количество субъектов МСП по видам деятельности в населённых пунктах Свердловской области в 2021 г., причём без ограничения по числу фирм.
 
@@
 
-Благодаря детальности набор данных позволяет анализировать изменения в конкретных фирмах, например, юридический «переезд» из одного региона в другой.
-На картограмме (рис.
-7) показаны «регионы-доноры» и «регионы-реципиенты» субъектов МСП за весь доступный период.
+Набор данных позволяет анализировать изменения у конкретных организаций и предпринимателей, например, «переезд» (изменение адреса регистрации) из одного региона в другой.
+На рис.
+7 показано изменение количества субъектов МСП из пробного набора, зарегистрированных в субъектах России, за весь доступный период.
 
@@
 
-Вспомогательные данные о доходах, расходах и числе работников позволяют дополнить анализ, перейдя от простого числа компаний к характеристикам их деятельности.
+Вспомогательные данные о доходах, расходах и числе работников позволяют перейти от простого числа компаний к характеристикам их деятельности.
 На трёх картограммах (рис.
-8) показано соответственно общее число фирм, их суммарная прибыль и суммарное число работников по регионам.
-Видно, что распределение хотя и похожее, но всё же отличается в зависимости от визуализируемой характеристики.
+8) показано соответственно общее число фирм, их суммарная прибыль и суммарное число работников по субъектам России.
+Видно, что распределение похожее, но всё-таки разное в зависимости от визуализируемой характеристики.
 
@@
 
-Предложенный алгоритм имеет техническую ценность: он позволяет обработать большой объём сложноструктутированных данных с сайта ФНС России и сделать на их основе несколько сравнительно компактных таблиц в формате, который традиционно используется исследователями при анализе данных.
+Предложенный алгоритм имеет техническую ценность: он позволяет обработать большой массив данных, имеющих сложную структуру, и сделать на его основе несколько сравнительно компактных таблиц в формате, который традиционно используется исследователями.
 Кроме того, технология нормализует адреса, содержащиеся в исходных данных, и обогащает их географическими координатами и кодами ОКТМО, тем самым делая их пригодными для обработки в геоинформационных системах (ГИС) и пространственного анализа в целом, а также упрощая объединение с муниципальной статистикой.
-Для выполнения алгоритма, несмотря на то что в его реализации используется Apache Spark, достаточно персонального компьютера средней конфигурации — использовать облачные вычислительные ресурсы не нужно.
+Для выполнения алгоритма, несмотря на использование Apache Spark, достаточно персонального компьютера средней конфигурации — использовать облачные вычислительные ресурсы не обязательно.
 Таким образом, для конечного пользователя технология представляет собой способ работать с big data, обойдя при этом сопутствующие сложности и повышенные требования к ресурсам и навыкам и с помощью нескольких оптимизаций превратив её в работу с обычными «малыми» данными.
@@
 
-Набор данных, сформированный с помощью описанной технологии — это информация о генеральной совокупности субъектов МСП в России.
+Набор данных, формируемый с помощью описанной технологии — это информация о генеральной совокупности субъектов МСП в России.
 Его можно использовать для сплошного исследования малого и среднего предпринимательства в стране.
-В некотором роде это квази-источник первичных статистических сведений, которые по общему правилу относятся к информации ограниченного доступа[^7], но в данном случае являются открытыми в силу прямого указания в Налоговом кодексе[^8].
-Кроме того, набор данных является полным геопривязанным перечнем индивидуальных предпринимателей.
-Тем не менее, поскольку ФНС России не публикует сведения о доходах, расходах и числе работников применительно к индивидуальным предпринимателям, исследовательская ценность этих данных меньше, чем информации об организациях.
+В некотором роде это квази-источник первичных статистических сведений, которые, по общему правилу, относятся к информации ограниченного доступа[^7], но в данном случае являются открытыми в силу прямого указания в Налоговом кодексе[^8].
+
+Технология де-факто позволяет получить полный перечень индивидуальных предпринимателей с именами и адресами (с точностью до населённого пункта).
+ФНС России не публикует сведения о доходах, расходах и числе работников применительно к индивидуальным предпринимателям, из-за чего исследовательская ценность этих данных меньше, чем информации об организациях.
 Также следует отметить, что анализ сведений об индивидуальных предпринимателях может рассматриваться как обработка персональных данных физических лиц, что создаёт правовые риски, особенно актуальные в свете регулярного ужесточения юридической ответственности за нарушения в этой сфере.
@@
 Вопрос о том, можно ли использовать технологию и создаваемый с её помощью набор данных для изучения не только малого и среднего бизнеса, но и экономики в целом, является сложным.
-Как отмечалось в обзоре литературы, потребность в данных, детализированных до уровня отдельных фирм, существует.
-Поскольку получить такие данные сложно или дорого, возникает желание использовать сведения о субъектах МСП в качестве заместителя (прокси), особенно при математическом моделировании.
-Это может показаться разумным, ведь субъекты МСП составляют почти 98% в общем числе фирм в России, дают 43% ВВП и обеспечивают занятостью 38% работников [@Barinova2019].
-В то же время очевидно, что реестр МСП не является случайной выборкой из генеральной совокупности хозяйствующих субъектов, поскольку его элементы отобраны исходя из жёстких лимитов, ограничивающих максимальный доход и число работников.
-Соответственно, реестр МСП даёт информацию только о небольших фирмах, и выводы, полученные на его основе, нельзя распространить на все компании в России.
-Во всяком случае, такой вывод следует, если оценивать репрезентативность реестра МСП исходя из процедурного критерия, то есть рассуждать теоретически.
-Возможен, однако, и альтернативный эмпирический путь: чтобы понять, является ли реестр МСП достаточно хорошим «прокси» для генеральной совокупности фирм, надо сравнить распределение фирм в нём с распределением в генеральной совокупности.
+Как отмечалось в обзоре литературы, потребность в данных, детализированных до уровня отдельных фирм, существует, но получить такую информацию сложно или дорого.
+Может быть, сведений о субъектах МСП будет достаточно, чтобы делать правдоподобные выводы об экономике в целом?
+На первый взгляд, это разумное предположение, ведь малые и средние предприятия составляют почти 98% в общем числе фирм в России, дают 43% ВВП и обеспечивают занятостью 38% работников [@Barinova2019].
+В то же время очевидно, что реестр МСП не является случайной выборкой из генеральной совокупности хозяйствующих субъектов, поскольку его элементы выбраны не случайно, а исходя из ограничений по финансовым показателям и числу занятых.
+Следовательно, реестр МСП позволяет делать выводы о небольших фирмах, но не позволяет распространить их на все компании.
+Во всяком случае, такой вывод следует, если оценивать репрезентативность реестра МСП как выборки исходя из процедуры её формирования, то есть рассуждать теоретически.
+Возможен, однако, и альтернативный, эмпирический путь: чтобы понять, является ли реестр МСП достаточно хорошим «прокси» для генеральной совокупности фирм, надо сравнить распределение фирм в нём с распределением в генеральной совокупности.
 Обычно при анализе репрезентативности сделать такое сравнение невозможно, поскольку характеристики генеральной совокупности неизвестны.
@@
 Для оценки репрезентативности реестра МСП можно использовать корреляционный анализ.
-Единая медведомственная информационно-статистическая система (ЕМИСС) содержит данные о числе организаций в разрезе регионов и групп ОКВЭД за 2019–2022 гг[^9].
+Единая межведомственная информационно-статистическая система (ЕМИСС) содержит данные о числе организаций в разрезе субъектов России и групп ОКВЭД за 2019–2022 гг[^9].
 Такая же статистика рассчитывается на базе реестра МСП.
-В результате для каждого года, каждого региона и каждой группы ОКВЭД известно два числа: общее количество организаций и количество субъектов МСП.
-Далее вычисляются:
-- коэффециенты корреляции в каждом году;
-- коэффициенты корреляции в каждом регионе;
-- коэффициенты корреляции в каждой группе ОКВЭД;
+В результате для каждого года, каждого региона и каждой группы ОКВЭД известны два числа: общее количество организаций и количество организаций в реестре МСП.
+Далее вычисляются коэффициенты корреляции по годам, регионам и группам.
 Используется ранговый коэффициент корреляции Спирмена, так как он более устойчив к выбросам, чем коэффициент корреляции Пирсона.
@@
 Результаты анализа показывают высокую согласованность между числом организаций в целом и числом организаций в реестре МСП.
-Общий коэффициент корреляции в течение рассмотренных четырёх лет составляет чуть больше 0.8 и почти не изменяется (таблица 3).
-Коэффициенты корреляции по группам ОКВЭД в регионах различаются довольно сильно (см.
+Общий коэффициент корреляции в течение рассмотренных четырёх лет составляет чуть больше 0.8 и мало изменяется (таблица 3).
+Коэффициенты корреляции по регионам различаются довольно сильно (см.
 рис.
@@
 9).
-Таким образом, в большинстве случаев реестр МСП может рассматриваться как достаточно качественный «прокси» как минимум в двух аспектах: он корректно отражает географическое распределение количества организаций по субъектам России в подавляющем большинстве групп ОКВЭД и структуру экономики по группам ОКВЭД в значительной части субъектов.
-Это не гарантирует репрезентативность на более низком уровне (конкретные коды ОКВЭД, муниципальные образования, населённые пункты) или по другим характеристикам (число работников, прибыль), но позволяет сделать осторожный вывод о том, что как минимум в ряде задач технология позволяет подготовить такой набор данных, который можно использовать — в отсутствие подходящих альтернатив — не только при изучении МСП как такового, но и при географическом исследовании экономической активности всех организаций.
+
+Таким образом, в большинстве случаев реестр МСП может рассматриваться как достаточно качественная замена полному списку организаций как минимум в двух аспектах: он корректно отражает географическое распределение количества организаций по субъектам России в подавляющем большинстве групп ОКВЭД и структуру экономики по группам ОКВЭД в значительной части субъектов.
+Это не гарантирует репрезентативность на более низком уровне (конкретные коды ОКВЭД, муниципальные образования, населённые пункты) или по другим характеристикам (число работников, прибыль).
+Тем не менее, технология позволяет подготовить такой набор данных, который можно использовать — в ряде задач, в отсутствие альтернатив и при дополнительных проверках — не только при изучении малого и среднего предпринимательства как такового, но и при географическом исследовании экономической активности всех организаций.
+Кроме того, набор данных можно применять для разведочного анализа, по результатам которого искать полные данные для более качественной проверки гипотез.
 
@@
 Юридический адрес может отличаться от места ведения деятельности, а ОКВЭД, указанный в документах, может не соответствовать фактической деятельности.
-Сложности усиливаются из-за того, что расхождения могут проявляться в разной степени в зависимости от расположения и отрасли экономики: так, в особых экономических зонах будет больше номинально зарегистрированных организаций, а место деятельности интернет-компаний будет явно не совпадать с местом нахождения их офиса.
-Так, можно предположить, что место регистрации фирмы будет как минимум отчасти объясняться налоговыми или престижными соображениями, а место регистрации фирм, осуществляющих деятельность в интернете или в нескольких регионах, будет мало связано с местом их регистрации.
-Вряд ли эту проблему можно решить системно, поэтому технология презюмирует достоверность данных о местонахождении и коде ОКВЭД и не проверяет их.
-В некоторых случаях объём данных сам по себе снижает эффект от неизбежных расхождений, но в других ситуациях задача исследователя, использующего технологию и произведённый с её помощью набор данных — самостоятельно проверить соответствие между номинальным и фактическим состоянием.
-Для этого можно, например, использовать открытые картографические сервисы, сайты организаций и здравый смысл.
+Сложности усиливаются из-за того, что расхождения могут проявляться в разной степени в зависимости от расположения и отрасли экономики: так, в особых экономических зонах много номинально зарегистрированных организаций, а место деятельности интернет-компаний вряд ли совпадает с местом нахождения офиса.
+Вряд ли эту проблему можно решить системно, поэтому технология презюмирует достоверность данных о местонахождении и виде деятельности и не проверяет их.
+В некоторых случаях объём данных сам по себе снижает эффект от неизбежных расхождений.
+Если возникают сомнения, то целесообразно выборочно проверить соответствие между номинальным и фактическим положением дел.
+Для этого можно использовать открытые картографические сервисы, сайты организаций и здравый смысл.
 
@@
 
-В статье описана технология, позволяющая на основе открытых данных ФНС России создать геопривязанный набор данных о субъектах МСП, ведущих деятельность определённого вида.
+В статье описана технология, позволяющая на основе открытых данных ФНС России создать геопривязанный набор данных о субъектах МСП (организациях и индивидуальных предпринимателях), ведущих деятельность определённого вида.
 Получаемый набор охватывает период с середины 2016 г.
-по конец 2023 г., содержит сведения о наименовании, местонахождении фирмы, её виде и категории, доходах, расходах и числе работников.
+по настоящее время, содержит сведения о регистрационных номерах, наименовании, местонахождении фирмы, её виде и категории, доходах, расходах и числе работников.
 В пространственном отношении он детализирован до населённых пунктов, во временном — до лет (в части доходов, расходов и числа работников) или месяцев (по остальным характеристикам), в структурном — до кода вида деятельности по ОКВЭД.
@@
 Технология может использоваться как готовый способ получения данных для экономико-географических исследований или как база для создания производных решений.
-Поскольку получаемый набор данных представляет собой сплошной геопривязанный перечень субъектов МСП в России за несколько лет, то он может применяться для пространственного анализа малого и среднего предпринимательства.
+Поскольку получаемый набор данных представляет собой сплошной геопривязанный перечень субъектов МСП в России за несколько лет, то он может применяться для пространственного и временного анализа малого и среднего предпринимательства.
 Благодаря относительной репрезентативности реестра МСП по отношению к генеральной совокупности российских фирм в географическом и структурном плане, технология может использоваться как источник данных везде, где требуется геопривязанная информация об экономической активности с детализацией до конкретных организаций или населённых пунктов.
-Технология позволяет самостоятельно получить данные, аналогичные тем, которые содержатся в коммерческих системах бизнес-аналитики или в справочных картографических сервисах.
+Технология позволяет самостоятельно получить данные, аналогичные тем, которые содержатся в коммерческих системах бизнес-аналитики или в картографических сервисах.
 Поскольку создаваемый набор данных содержит координаты для каждой записи (элемента), то он удобен также для картографической визуализации.
@@
 К ограничениям технологии можно отнести неизбежные расхождения между номинальными сведениями, указанными в реестрах, и фактическим положением дел; охват только части фирм, причём не обязательно репрезентативной во всех отношениях; правовые риски при анализе данных об индивидуальных предпринимателях, которые одновременно являются персональными данными.
-Из-за неполноты адресных баз данных и изменчивости классификатора муниципального деления геопривязка также происходит корректно не во всех случах.
+Из-за неполноты адресных баз данных и изменчивости классификатора муниципального деления геопривязка также происходит корректно не во всех случаях.
 Кроме того, с учётом тенденции к снижению открытости административных данных есть опасность, что исходные наборы данных на сайте ФНС России перестанут быть доступными.
@@
 3.
-Распределение количества субъектов МСП, ведущих деятельность в сфере сельского  и лесного хозяйства по регионам (субъектам федерации).
-Цвет показывает число, масштабированное так, что 0 — минимум, 1 — максимум"}
+Распределение количества субъектов МСП, ведущих деятельность в сфере сельского и лесного хозяйства, по регионам (субъектам федерации).
+Цвет — масштабированное количество (0 — минимум, 1 — максимум)"}
 regions_plot_agri_forestry
@@
 4.
-Распределение количества субъектов МСП, ведущих деятельность по выращиванию пшеницы и картофеля по регионам (субъектам федерации).
-Цвет показывает число, масштабированное так, что 0 — минимум, 1 — максимум.
+Распределение количества субъектов МСП, ведущих деятельность по выращиванию пшеницы и картофеля, по регионам (субъектам федерации).
+Цвет — масштабированное количество (0 — минимум, 1 — максимум).
 Серый цвет — отсутствие данных"}
@@
 легенду).
-Показаны все населённые пункты без ограничений по количеству зарегистрированных фирм"}
+Показаны все населённые пункты без ограничения по количеству зарегистрированных фирм"}
 activity_by_settlements_svr
@@
   col.names = c("Год", "$\\rho_{Spearman}$, все группы ОКВЭД", "$\\rho_{Spearman}$, без O, P, U"),
-  caption = "Таблица 3. Корреляция между числом организаций в реестре МСП и общим числом организаций в разрезе регионов (субъектов федерации) и групп деятельности по годам")
+  caption = "Таблица 3. Корреляция между числом организаций в реестре МСП и общим числом организаций в каждом регионе(субъекте федерации) в каждой группе деятельности по годам")
 ```
@@
 
-```{r fig-9-corr-by-region, fig.cap = "Рис. 9. Распределение регионов (субъектов федерации) по корреляции между числом организаций в реестре МСП и общим числом организаций"}
+```{r fig-9-corr-by-region, fig.cap = "Рис.
+9.
+Распределение регионов (субъектов федерации) по корреляции между числом организаций в реестре МСП и общим числом организаций в каждом году в каждой группе деятельности"}
 corr_by_region_plot
@@
   col.names = c("Группа ОКВЭД", "$\\rho_{Spearman}$"),
-  caption = "Таблица 4. Корреляция между числом организаций в реестре МСП и общим числом организаций в разрезе регионов (субъектов федерации) и годов по группам деятельности")
+  caption = "Таблица 4. Корреляция между числом организаций в реестре МСП и общим числом организаций в каждом регионе (субъекте федерации) и каждом году по группам деятельности")
 ```

### 2024-01-08 Plots.R

@@ -22,7 +22,7 @@ labels <- data.frame(
     "Рыболовство и рыбоводство",
     "Изменение за 2016–2023",
     "Количество фирм",
-    "Численность работников",
+    "Число работников",
     "Прибыль"
   ),
   en = c(

### 2024-01-08 Paper.Rmd
Proofreading of English meta, title change

@@
 
-**Использование открытых данных налоговой службы для создания геопривязанного набора данных о субъектах малого и среднего предпринимательства в России**
+**Технология создания геопривязанного набора данных о субъектах малого и среднего предпринимательства в России на базе открытых данных налоговой службы**
 

### 2024-01-08 Paper_meta_en.Rmd
Proofreading of English meta, title change

@@
 
-**Technology of Using Federal Taxpayer Service's Open Data for Making a Geo-Referenced Dataset about Small and Medium-Sized Businesses in Russia**
+**Technology of Making a Geo-Referenced Dataset about Small and Medium-Sized Businesses in Russia Based on the Federal Tax Service's Open Data**
 
@@
 
-Syomin Pavel Olegovich: ntsp\@ya.ru, ORCID 0000-0002-4015-9206
+Pavel O. Syomin: ntsp\@ya.ru, ORCID 0000-0002-4015-9206
 
@@
   col.names = c("SMB Registry", "Revenue+Expediture", "Employees", "Total"),
-  caption = "Table 1. Change in the data volumes during program execution (in megabytes). Rows are stages, columns are source datasets and totals")
+  caption = "Table 1. Change in the data volume during program execution (in megabytes). Rows are stages, columns are source datasets and totals")
 ```
@@
 
-```{r fig-3-regions-agri-forest, fig.cap = "Fig. 3. Regional distribution of SMBs which main activity is agriculture or forestry. Color represents the min-max scaled value"}
+```{r fig-3-regions-agri-forest, fig.cap = "Fig. 3. Regional distribution of SMBs which main activity is agriculture or forestry, in 2021. Color represents the min-max scaled value"}
 regions_plot_agri_forestry
@@
 
-```{r fig-3-regions-wheat-potato, fig.cap = "Fig. 4. Regional distribution of SMBs which main activity is growing of wheat or potato. Color represents the min-max scaled value, gray is no data"}
+```{r fig-3-regions-wheat-potato, fig.cap = "Fig.
+4.
+Regional distribution of SMBs which main activity is growing of wheat or potato, in 2021.
+Color represents the min-max scaled value, gray is no data"}
 regions_plot_wheat_potato
@@
 5.
-Count of SMBs in some Russian settlements with dominant activity type.
-The counts are represented with circle size, and the most popular activity group in the settlement is displayed by color.
+Count of SMBs in some Russian settlements with dominant activity type, in 2021.
+The counts are represented by circle size, and the most popular activity group in the settlement is displayed by color.
 Only settlements with 50+ SMBs are shown"}
@@
 6.
-Count of SMBs in settlements of Sverdlovsk region with dominant activity type.
-The counts are represented with circle size, and the most popular activity group in the settlement is displayed by color.
+Count of SMBs in settlements of Sverdlovsk region with dominant activity type, in 2021.
+The counts are represented by circle size, and the most popular activity group in the settlement is displayed by color.
 All settlements regardless of the counts are shown"}
@@
   col.names = c("Year", "$\\rho_{Spearman}$, all activity groups", "$\\rho_{Spearman}$, excluding groups O, P, U"),
-  caption = "Table 3. The correlation between the count of SMBs and total count of businesses by year")
+  caption = "Table 3. The correlation between the count of SMB organizations and total count of commercial organizations by year")
 ```
@@
 
-```{r fig-9-corr-by-region, fig.cap = "Fig. 9. Distribution of regions by the correlation coefficient between the SMBs counts and total businesses counts"}
+```{r fig-9-corr-by-region, fig.cap = "Fig. 9. Distribution of regions by the correlation coefficient between the SMB organizations counts and total commercial organizations counts"}
 corr_by_region_plot
@@
   col.names = c("Activity group", "$\\rho_{Spearman}$"),
-  caption = "Table 4. The correlation between the count of SMBs and total count of businesses by activity group")
+  caption = "Table 4. The correlation between the count of SMB organizations and total count of commercial organizations by activity group")
 ```

### 2024-01-08 Paper.Rmd
Edit English abstract, change color legend of fig. 8, minor fixes

@@
 
-```{r fig-7-migrations, fig.cap = "Рис.
-7.
-Распределение количества субъектов МСП из демонстрационного набора, прибывших или покинувших территорию региона (субъекта федерации) за 2016–2023 г.
-Абсолютные числа масштабированы так, что -1 — максимальная убыль, 1 — максимальный прирост"}
+```{r fig-7-migrations, fig.cap = "Рис. 7. Распределение количества субъектов МСП из демонстрационного набора, прибывших или покинувших территорию региона (субъекта федерации) за 2016–2023 г."}
 migrations_plot

### 2024-01-08 Paper_meta_en.Rmd
Edit English abstract, change color legend of fig. 8, minor fixes

@@
 
-The paper features a technology (algorithm, application) for making of a tabular georeferenced dataset about small and medium-sized bisunesses (SMBs) in Russia based on open data of Federal Taxpayer Service (FTS) of Russia.
-The resulting dataset covers the period from 2016 to 2023 and the whole area of Russia.
-It can be considered as a complete list, or population, of SMBs and contains information about category, name, address, activity code, revenue, expediture, number of employees of an organization or an individual entrepreneur.
-The spatial resolution of the dataset is by settlement, including geographical coordinates and municipality code; the temporal resolution is up to a year or month depending on the features; the resolution regarding the activity type is by concrete activity code according to the state classifier.
-In comparison to the typical data paper, this article contains the description of the methods used to produce the data and demonstrates the application of the proposed technology rather than the description of the dataset itself.
-Using the algorithm and its sample implementation on Python, a researcher can perform the data processing independently and obtain the up-to-date copy of the resulting data, optionally tailored to its specific purposes.
-The algorithm is optimized so that it runs on a typical PC despite the relatively large (~2Tb) size of unpacked source data.
-The dataset produced by the proposed method can be used for spatial analysis of SMBs in Russia as well as a proxy data source which is quite relevant for analysis of Russia's economy.
-The detalization up to a specific organizations enables the use of the produced dataset as a free partial alternative to commercial business intelligence systems like Spark and Kontur.Focus.
-Also, the presence of geographical coordinates makes mapping easy, and municipality codes may be used to join the dataset with official statistical databases.
+The paper features a technology for creating a geo-referenced tabular dataset of small and medium-sized businesses (SMBs) in Russia based on open data from the Federal Tax Service (FTS of Russia).
+The resulting dataset covers the entire territory of Russia for 2016—2023.
+It is a complete list of SMBs and it contains information on registration numbers, category, name, address, activity type, income, expenses, and number of employees.
+In spatial terms, the dataset is detailed down to settlements, including geographic coordinates and municipal code (OKTMO); in a temporary sense — up to months or years, depending on the specific feature; in relation to the type of activity — up to individual codes of the activity classifier.
+Unlike a classic data paper, the article does not describe the resulting dataset as such, but outlines a reproducible methodology for creating it and demonstrates its use to obtain a dataset of firms in the agriculture, forestry and fisheries sectors.
+Using the proposed algorithm and its implementation in Python, the researcher can independently process the source data and get an up-to-date version of the dataset, tailored to his purposes.
+The algorithm is optimized so that it runs on a typical PC despite the relatively large (~2Tb uncompressed) volume of source data.
+The dataset produced by the proposed method can be used for spatial analysis of SMBs in Russia and as an open, free and relatively representative source of information about the Russian economy in the context of specific organizations, making it possible to partially replace commercial business intelligence systems like Spark and Kontur.Focus.
+In addition, thanks to the coordinates, the data set can be used for mapping without additional processing, and the presence of OKTMO codes allows it to be combined with official statistical information.
 
@@
   format.args = list(scientific = FALSE),
-  col.names = c("SMB Registry", "Revenue+Expediture", "Employees", "Total"),
+  col.names = c("SMB Registry", "Revenue+Expenditure", "Employees", "Total"),
   caption = "Table 1. Change in the data volume during program execution (in megabytes). Rows are stages, columns are source datasets and totals")
@@
   format.args = list(scientific = FALSE),
-  col.names = c("SMB Registry", "Revenue+Expediture", "Employees"),
+  col.names = c("SMB Registry", "Revenue+Expenditure", "Employees"),
   caption = "Table 2. Change in the number of records during program execution (in thousands). Rows are stages, columns are source datasets")
@@
 
-```{r fig-7-migrations, fig.cap = "Fig.
-7.
-Regional distribution of SMBs which have arrived to or left the region during 2016–2023.
-The color is min-max scaled value, -1 is the top decline, while +1 is the top growth"}
+```{r fig-7-migrations, fig.cap = "Fig. 7. Regional distribution of SMBs which have arrived to or left the region during 2016–2023"}
 migrations_plot

### 2024-01-08 Plots.R

@@ -202,7 +202,7 @@ migrations_plot <- region_changes %>% ungroup() %>%
   ggplot() +
   geom_sf(aes(fill = change), size = .1) + 
   coord_sf(crs = ru_crs) +
-  scale_fill_distiller(name = labels[13], palette = "PiYG", direction = 1) +
+  scale_fill_gradient2(name = labels[13], low = "#d01c8b", high = "#4dac26") +
   theme_bw(base_size = 12, base_family = "Times New Roman") +
   theme(legend.position = "bottom")
 migrations_plot

### 2024-01-08 helpers/Diagram.R
Move geojsons to assets, move helper scripts to helpers, make minor changes

@@ -57,7 +57,7 @@ labels <- labels[[LOCALE]]
 
 diagram <- grViz("digraph {
   
-graph[layout = dot, fontname = 'Ubuntu', fontsize = 18]
+graph[layout = dot, fontname = 'Times New Roman', fontsize = 18]
 
 subgraph cluster_a {
   rank = same;
@@ -145,6 +145,5 @@ D3a -> {A3b, A3c}
 [21]: labels[21]
 [22]: labels[22]
 ")
-diagram
+
 diagram_raster <- diagram %>% export_svg() %>% charToRaw() %>% rsvg_nativeraster(width = 2000)
-grid::grid.raster(diagram_raster)

### 2024-01-08 helpers/Plots.R

@@ -6,6 +6,7 @@ library(scales)
 library(sf)
 library(tidyr)
 
+# Localized labels
 labels <- data.frame(
   ru = c(
     "Субъект России",
@@ -47,8 +48,18 @@ labels <- data.frame(
 )
 labels <- labels[[LOCALE]]
 
+# Data
 rsmp_data_path <- "../tax-service-opendata/rsmp/reestr_group_A/data_product.csv"
 rsmp_data <- read_csv(rsmp_data_path)
+
+# Maps
+regions_boundaries <- st_read("assets/ru.geojson")
+ru_crs <- st_crs("+proj=aea +lat_0=0 +lon_0=100 +lat_1=68 +lat_2=44 +x_0=0 +y_0=0 +ellps=krass +towgs84=28,-130,-95,0,0,0,0 +units=m +no_defs")
+regions <- read_csv("../tax-service-opendata/assets/regions.csv")
+regions <- regions_boundaries %>% 
+  left_join(regions, by = c("shapeISO" = "iso_code")) %>% 
+  select(name, name_en = shapeName)
+ru_svr <- st_read("assets/ru_svr.geojson")
 
 # Regions time series
 monthly_counts <- do.call(rbind, lapply(
@@ -69,20 +80,23 @@ time_series_plot <- expand(monthly_counts, region, date) %>%
   ggplot(aes(x = date, y = n, color = region)) +
   geom_line(na.rm = TRUE) +
   scale_y_continuous(trans = "log10") +
-  scale_color_discrete(name = labels[1]) +
   labs(x = labels[2], y = labels[3]) +
   theme_bw(base_size = 12, base_family = "Times New Roman")
+
+if (LOCALE == "en") {
+  time_series_plot <- time_series_plot + 
+    scale_color_discrete(
+      name = labels[1],
+      labels = function(breaks) {
+        br <- data.frame(name = breaks)
+        left_join(br, regions) %>% select(name_en) %>% pull
+    })
+} else {
+  time_series_plot <- time_series_plot + scale_color_discrete(name = labels[1])
+}
 time_series_plot
 
-# Maps
-regions_boundaries <- st_read("geoBoundaries-RUS-ADM1_simplified.geojson")
[…]

### 2024-01-08 helpers/Stats.R

@@ -158,7 +158,8 @@ corr_by_region <- left_join(
 corr_by_region_plot <- ggplot(corr_by_region, aes(x = cor, color = option)) +
   geom_freqpoly(binwidth = .05) +
   scale_color_discrete(name = labels[5], labels = c(labels[6], labels[7])) +
-  labs(x = labels[8], y = labels[9])
+  labs(x = labels[8], y = labels[9]) +
+  theme_bw(base_size = 12, base_family = "Times New Roman")
 corr_by_region_plot
 
 corr_by_group <- val_stats %>% 

### 2024-01-26 argo-conf-2024-paper/Paper.Rmd
Paper for ARGO Russia conference 2024

---
output: 
  word_document:
    reference_docx: reference.docx
bibliography: refs.bib
csl: "gost-r-7-0-5-2008-numeric-alphabetical.csl"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE, 
  message = FALSE, 
  echo = FALSE, 
  fig.width = 6.4, 
  fig.height = 5,
  fig.id = FALSE,
  dpi = 300)
```

```{r imports, include = FALSE, cache = TRUE}
source("Script.R")
```

УДК 911.3, 519.688

<div custom-style="Author">
П. О. Сёмин

P. O. Syomin
</div>

<br>

<div custom-style="Title">
Консольное приложение для создания геопривязанного набора данных о субъектах малого и среднего предпринимательства в России на базе открытых данных Федеральной налоговой службы

A Python CLI application to generate a geo-referenced dataset on small and medium-sized businesses in Russia based on Federal Tax Service open data
</div>

<br>

<br>

<div custom-style="Abstract">
Представлено консольное приложение на языке Python, которое создаёт полный геопривязанный набор данных о субъектах малого и среднего предпринимательства в России. Источник сведений — открытые данные Федеральной налоговой службы. Системные требования — 12 Гб оперативной памяти, 300 Гб свободного места на диске. Создаваемый набор данных содержит: наименование, вид, категорию, ИНН, ОГРН, адрес (населённый пункт, его координаты, ОКТМО), код ОКВЭД, доходы, расходы, среднесписочную численность работников за каждый год c 2016 по 2023.

A command-line Python tool generates a complete geo-referenced dataset of small and medium-sized businesses in Russia. Open data of Federal Tax Service is used as a source. System requirements are 12 Gb of RAM and 300 Gb of free storage. The created dataset includes the following features for each company for each year from 2016 to 2023: name, kind, category, taxpayer number, registration number, address (settlement, geographic coordinates, municipal code), activity code, revenue, expenditure, count of employees.

<br>

Ключевые слова: открытые данные, малое и среднее предпринимательство, ФНС России, набор данных, big data.

Keywords: open data, small and medium-sized businesses, FTS of Russia, dataset, big data.
</div>

В современных экономико-географических исследованиях нередко возникает потребность в дезагрегированных данных. К ним, в частности, относятся сведения о конкретных организациях: наименование, местонахождение, прибыль, число работников, вид деятельности. Такая информация важна при построении математических моделей [@Rostislav2021], для оценки уровня развития территорий [@Gumenyuk2022], при анализе узких сфер деятельности, по которым нет отдельной статистики [@Moiseeva2016; @Pankratov2023], для исследований на локальном уровне (отдельные города, административные районы, муниципальные образования) [@Kozhevnikov2023; @Lachininsky2021]. Для получения дезагрегированных данных исследователи обычно обращаются к системам бизнес-аналитики («Спарк», «Контур.Фокус»), онлайн-картам  (коммерческим, таким как 2ГИС или Яндекс.Карты, или открытым, например, OpenStreetMap) [@Lachininsky2021]. У такого подхода есть существенные недостатки: если данные покупаются, то надо платить, заключать отдельные договоры, соблюдать ограничения, а если данные собираются с общедоступных карт, то их обработка требует существенных усилий и не всегда законна.

В статье представлен инструмент, позволяющий самостоятельно создать набор геопривязанных данных о субъектах малого и среднего предпринимательства (МСП) в России. Получаемый набор содержит такую информацию, как наименование и местонахождение фирмы, её регистрационные номера (ИНН, ОГРН), вид и категорию (физическое или юридическое лицо; микропредприятие, малое или среднее предприятие), код основного вида деятельности по ОКВЭД, доходы, расходы и среднесписочная численность работников за каждый год, начиная с 2019. Местонахождение дано с точностью до населённого пункта (городского или сельского), при этом указаны его географические координаты и код ОКТМО. Сведения доступны с середины 2016 г., что позволяет отслеживать временную динамику. Источник информации — открытые данные Федеральной налоговой службы (ФНС России). Использование открытых данных налоговой службы в экономико-географических исследованиях уже практиковалось рядом специалистов [@Gumenyuk2022; @Rostislav2021; @Lachininsky2021]. Обрабатывать открытые данные ФНС России сложно из-за большого объёма и запутанного формата. Предложенный инструмент упрощает эту задачу.

Приложение использует три набора открытых данных ФНС России: один основной и два вспомогательных. Основной — «Реестр субъектов малого и среднего предпринимательства», вспомогательные — «Сведения о доходах и расходах организаций» и «Сведения о среднесписочной численности работников организаций». Каждый из них — это коллекция zip-архивов xml-файлов, где один архив содержит данные на определённую дату. Наборы обновляются с разной периодичностью, но не реже раза в год. Каждое обновление — это публикация текущей версии данных полностью, а не только изменений, поэтому сведения в выгрузках частично дублируются.

Для геопривязки приложение использует две справочные таблицы населённых пунктов России: набор «Населённые пункты России» от проекта «Инфраструктура научно-исследовательских данных» и набор «Города России», бесплатно опубликованный коммерческой компанией DaData.

Приложение представляет собой консольную программу на языке программирования Python. Оно основано на нескольких популярных сторонних компонентах: Pandas (средство обработки табличных данных), lxml (библиотека для чтения xml-файлов), Apache Spark (инструмент для обработки больших объёмов данных), Typer (фреймворк для создания интерфейсов командной строки). Приложение работает на компьютере средней конфигурации: достаточно 12 Гб оперативной памяти и 300 Гб свободного места на жёстком диске.

Алгоритм обработки данных состоит из пяти шагов: загрузка данных, извлечение данных из архивов (с опциональной фильтрацией по виду деятельности), агрегирование данных для удаления повторяющихся записей, геопривязка, создание панельного представления (по годам). Соответственно, приложение включает в себя пять ключевых компонентов: загрузчик (Downloader), распаковщик (Extractor), агрегатор (Aggregator), геопривязчик (Georeferencer), генератор панельного представления (Panelizer). Командный интерфейс, в свою очередь, содержит пять основных команд: download, extract, aggregate, georeference, panelize. У каждой команды есть опции, описание которых имеется в документации. Поскольку предполагается, что основной сценарий использования приложения — это создание набора данных «под ключ», то интерфейс включает комбинированную команду process, которая берёт исходные данные, по очереди запускает все этапы обработки и производит панельный набор данных.

Приложение размещено в открытом репозитории по адресу: https://github.com/PavelSyomin/russian-smb-companies. Там же доступна документация с примерами использования.

Для проверки и демонстрации работы приложения был подготовлен набор данных о фирмах, ведущих деятельность в сфере сельского, лесного хозяйства, охоты, рыболовства и рыбоводства (группа «А» ОКВЭД). Объём исходных данных — чуть больше 200 Гб в сжатом виде и около 2.5 Тб в несжатом, размер итогового панельного представления — 630 Мб, то есть в сотни раз меньше. Время обработки — около 15 часов. Набор данных содержит сведения о `r total` фирмах.

Картограмма (рис. 1) показывает все организации в сфере сельского хозяйства в 2021 г., сведения о которых имеются в наборе данных. Таблица 1 — это пример одной записи из набора данных, который позволяет понять состав сведений.

```{r plot-1, fig.cap = "Рис. 1. Населённые пункты регистрации российских малых и средних организаций, ведущих деятельность в сфере сельского хозяйства (растениеводства и животноводства), в 2021 г. Если в одном и том же населённом пункте находится несколько организаций разной отрасли, то цвет показывает, каких организаций в этом месте больше"}
plot_1
```

```{r sample-table}
knitr::kable(
  sample_table, 
  col.names = c("Признак", "Значение"),
  caption = "Таблица 1. Пример одной записи из набора данных, созданного с помощью приложения"
  )
```

Таким образом, приложение помогает исследователю собрать и подготовить исходные данные для дальнейшего анализа. Оно может использоваться как частичная альтернатива коммерческим сервисам, предоставляющим дезагрегированные данные об организациях. Получаемый набор данных подходит для изучения малого и среднего предпринимательства как такового, а также анализа тех отраслей экономики и видов экономической деятельности, где малые и средние предприятия преобладают.

# Использованная литература

<div id="refs"></div>

# Сведения об авторе

**Сёмин Павел Олегович** — аспирант Пермского государственного национального исследовательского университета, бэкенд-разработчик ООО «АгроСофтвер», ntsp\@ya.ru.

**Pavel O. Syomin** (ntsp\@ya.ru) is a PhD student at Perm State University and a backend developer at AgroSoftware LLC.

### 2024-01-26 argo-conf-2024-paper/Script.R

@@ -1,0 +1,50 @@
+library(dplyr)
+library(ggplot2)
+library(readr)
+library(scatterpie)
+library(sf)
+library(tidyr)
+
+data_path <- "../ru-smb-companies/group_A/panel.csv"
+data <- read_csv(data_path)
+
+total <- data %>% distinct(tin) %>% nrow()
+
+sample_row <- data %>%
+  filter(year == 2021, kind == 1) %>% 
+  slice(1000:1001)
+sample_table <- data.frame(
+  var = colnames(sample_row),
+  value = as.character(sample_row[1, ])
+) %>% replace_na(list(value = "—"))
+
+regions_boundaries <- st_read("ru.geojson")
+ru_crs <- st_crs("+proj=aea +lat_0=0 +lon_0=100 +lat_1=68 +lat_2=44 +x_0=0 +y_0=0 +ellps=krass +towgs84=28,-130,-95,0,0,0,0 +units=m +no_defs")
+
+plot_1 <- data %>% 
+  filter(
+    year == 2021,
+    kind == 1, 
+    substr(activity_code_main, 1, 2) == "01") %>% 
+  mutate(activity = case_when(
+    substr(activity_code_main, 1, 4) %in% c("01.1", "01.2", "01.3") ~ "Растениеводство",
+    substr(activity_code_main, 1, 4) == "01.4" ~ "Животноводство")) %>% 
+  drop_na(activity) %>% 
+  count(lat, lon, activity) %>% 
+  drop_na(lat, lon) %>% 
+  group_by(lat, lon) %>% 
+  arrange(-n) %>% 
+  slice(1) %>% 
+  ungroup() %>% 
+  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>% 
+  ggplot() +
+  geom_sf(data = st_combine(regions_boundaries)) + 
+  geom_sf(aes(color = activity), size = .1) +
+  geom_sf(data = regions_boundaries, size = .1, fill = NA) + 
+  coord_sf(crs = ru_crs) +
+  scale_color_discrete(name = "Преобладающий вид деятельности") +
+  guides(color = guide_legend(override.aes = list(size = 1))) +
+  theme_bw(base_size = 12, base_family = "Times New Roman") +
+  theme(legend.position = "bottom", legend.direction = "horizontal")
+plot_1  
+

### 2024-01-29 argo-conf-2024-paper/Paper.Rmd
Add links to source datasets, fix typo in title, fix font in some styles, other minor improvements

@@
 
-A Python CLI application to generate a geo-referenced dataset on small and medium-sized businesses in Russia based on Federal Tax Service open data
+A Python CLI application to generate a geo-referenced dataset of small and medium-sized businesses in Russia based on Federal Tax Service open data
 </div>
@@
 Приложение использует три набора открытых данных ФНС России: один основной и два вспомогательных.
-Основной — «Реестр субъектов малого и среднего предпринимательства», вспомогательные — «Сведения о доходах и расходах организаций» и «Сведения о среднесписочной численности работников организаций».
+Основной — «Реестр субъектов малого и среднего предпринимательства»[^1], вспомогательные — «Сведения о доходах и расходах организаций»[^2] и «Сведения о среднесписочной численности работников организаций»[^3].
 Каждый из них — это коллекция zip-архивов xml-файлов, где один архив содержит данные на определённую дату.
 Наборы обновляются с разной периодичностью, но не реже раза в год.
-Каждое обновление — это публикация текущей версии данных полностью, а не только изменений, поэтому сведения в выгрузках частично дублируются.
+Каждое обновление — это публикация текущей версии данных полностью, а не только изменений, поэтому сведения в архивах частично дублируются.
 
-Для геопривязки приложение использует две справочные таблицы населённых пунктов России: набор «Населённые пункты России» от проекта «Инфраструктура научно-исследовательских данных» и набор «Города России», бесплатно опубликованный коммерческой компанией DaData.
+[^1]: https://www.nalog.gov.ru/opendata/7707329152-rsmp/
+[^2]: https://www.nalog.gov.ru/opendata/7707329152-revexp/
+[^3]: https://www.nalog.gov.ru/opendata/7707329152-sshr2019/
+
+Для геопривязки приложение использует две справочные таблицы населённых пунктов России: набор «Населённые пункты России»[^4] от проекта «Инфраструктура научно-исследовательских данных» и набор «Города России»[^5], бесплатно опубликованный коммерческой компанией DaData.
+
+[^4]: https://data.rcsi.science/data-catalog/datasets/160/
+[^5]: https://github.com/hflabs/city
 

### 2024-01-29 argo-conf-2024-paper/Script.R

@@ -5,7 +5,7 @@ library(scatterpie)
 library(sf)
 library(tidyr)
 
-data_path <- "../ru-smb-companies/group_A/panel.csv"
+data_path <- "../../ru-smb-companies/group_A/panel.csv"
 data <- read_csv(data_path)
 
 total <- data %>% distinct(tin) %>% nrow()

### 2024-02-09 maksakov-readings-conf-2024/Paper.Rmd
Helper script, base Rmd, aux data, introduction for Maksakov Readings 2024 conference

---
title: "Пространственный анализ миграций юридических компаний в России в 2016–2023 гг. по открытым данным налоговой службы"
author: П. О. Сёмин
output: 
  word_document:
    reference_docx: reference.docx
bibliography: refs.bib
csl: "gost-r-7-0-5-2008-numeric-alphabetical.csl"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE, 
  message = FALSE, 
  echo = FALSE, 
  fig.width = 6.4, 
  fig.height = 5,
  fig.id = FALSE,
  dpi = 300)
```

```{r imports}
library(dplyr)
library(ggplot2)
library(glue)
library(readr)
library(tidyr)
library(sf)
```

```{r load-data, include=FALSE, cache=TRUE}
data <- read_csv("../../ru-smb-companies/legal/smb.csv")
revexp <- read_csv("../../ru-smb-companies/legal/revexp.csv")
empl <- read_csv("../../ru-smb-companies/legal/empl.csv")
cities <- read_csv("cities.csv")
regions <- st_read("ru.geojson")
```

<div custom-style="Author">

Пермский государственный национальный исследовательский университет, г. Пермь
</div>

<div custom-style="Abstract">

*Аннотация:*

*Ключевые слова:* миграция компаний, малое и среднее предпринимательство, юридические компании, юридические услуги, география услуг, анализ данных, открытые данные, ФНС России

</div>

# Введение

Миграция, или релокация, компаний — важное экономико-географическое явление, играющее большую роль в экономическом развитии стран, регионов, территорий. Оно связано с мобильностью труда и капитала, инвестиционной привлекательностью, балансом выгод и издержек, агломерационными эффектами, глобализацией, экономическим ростом, налоговой оптимизацией и другими факторами, обстоятельствами, причинами, процессами и закономерностями. Исследование миграции компаний восходит к середине 20 века: одной из первых научных работ на эту тему считается книга *Why industry moves South* Г. МакЛафлина и С. Робока, в которой описывается перемещение промышленных производств с севера на юг США, вызванное особенностями рынка труда — меньшими зарплатами и активностью просфоюзов [@McLaughlin1949]. Релокация фирм — сложное явление, которое может рассматриваться не только в пространственном измерении, но и с точки зрения комплексного процесса принятия решений [@Carrincazeaux2015], восприниматься не только в исследовательском ключе, но и как прикладная задача [@Shortland1987]. Большой пласт исследований на протяжении десятилетий посвящён вопросу о причинах (факторах) миграции компаний [@Nakosteen1987; @Wu2007; @Rasel2019; @Pan2020] и о направлениях миграции [@Rupasingha2018; @Rossi2019; @Cybulska2020], а также о связи причин и направлений. Отдельные специалисты фиксируют внимание на более редких аспектах: дискурс о релокации [@Piotti2007], миграция в венчурные кластеры [@Plant2007]. Возникает и вопрос о неблагоприятных практических последствиях и мерах по их предотвращению, например, на налоговых потерях и желательности введения «налога на выход» [@Kubicova2016].

В России статьи о миграции компаний редки. В основном авторы немногочисленных имеющихся работ фокусируются на «негативной» стороне вопроса, анализируя миграцию налогоплательщиков-организаций как способ налоговой оптимизации [@Adigamova2011; @Romanovskaya2015; @Shcherbakova2016; @Sinenko2021; ]. Есть также исследование о релокации как факторе инвестиционной активности применительно к «компаниям-единорогам» [@Kutsenko2022]. Простой географический анализ миграций фирм при этом либо упускается из виду, либо имеет промежуточное значение на пути к более серьёзным выводам. Тем не менее, он тоже может быть полезен, поскольку понимание масштаба явления и его базовых географических характеристик необходимо для формирования более адекватной картины мира и выдвижения гипотез. В данном докладе проводится именно такой анализ по отношению к российским юридическим фирмам. Выбор вида деятельности анализируемых организаций объясняется личными интересами автора исследования.

# Данные и методы

# Результаты

# Обсуждение

# Выводы

# Литература

<div id="refs"></div>

# Приложения

<div custom-style="Title">
Spatial Analysis of Migration of Legal Companies in Russia (2016–2023) Based on Tax Service Open Data
<div>

<div custom-style="Author">
P. O. Syomin

Perm State University, Perm
</div>

<div custom-style="Abstract">

*Abstract:*

*Keywords:* companies migration, small and medium-sized businesses, legal companies, legal service, service geography, data analysis, open data, FTS of Russia

</div>

### 2024-02-09 maksakov-readings-conf-2024/Script.R

@@ -1,0 +1,267 @@
+library(dplyr)
+library(ggplot2)
+library(glue)
+library(readr)
+library(tidyr)
+library(sf)
+library(units)
+
+# Load the data
+data <- read_csv("../../ru-smb-companies/legal/smb.csv")
+revexp <- read_csv("../../ru-smb-companies/legal/revexp.csv")
+empl <- read_csv("../../ru-smb-companies/legal/empl.csv")
+cities <- read_csv("cities.csv")
+regions <- st_read("ru.geojson")
+
+# Calculate the absolute numbers and shares of companies
+# with changes in location
+# Location is defined as change either in region or settlement
+data %>% 
+  filter(kind == 1) %>% 
+  group_by(tin) %>%
+  distinct(region, settlement) %>% 
+  count() %>% 
+  ungroup() %>% 
+  count(n, name = "count") %>% 
+  mutate(share = 100 * count / sum(count))
+
+# Prepare the base dataset to analyze migrations of companies
+max_empl <- c(15, 100, 250)
+max_revenue <- c(120e6, 800e6, 2e9)
+migrations <- data %>% 
+  filter(kind == 1) %>% 
+  group_by(tin) %>% 
+  summarise(n = n_distinct(region, settlement)) %>%
+  filter(n > 1) %>% 
+  select(tin) %>% 
+  left_join(data) %>% 
+  group_by(tin) %>% 
+  distinct(region, settlement, .keep_all = TRUE) %>% 
+  arrange(start_date, end_date) %>% 
+  mutate(
+    next_region = lead(region),
+    next_settlement = lead(settlement),
+    next_lat = lead(lat),
+    next_lon = lead(lon),
+    date = lead(start_date),
+    year = as.numeric(format(date, "%Y"))
+  ) %>% 
+  drop_na(next_region) %>% 
+  ungroup() %>%
+  left_join(revexp, by = c("tin", "year")) %>% 
+  left_join(empl, by = c("tin", "year")) %>% 
+  replace_na(list(revenue = 0, expenditure = 0, employees_count = 0)) %>% 
+  mutate(
+    employees_count = ifelse(employees_count > 500, 0, employees_count),
+    revenue = ifelse(revenue > 1e10, 0, revenue)
+  ) %>% 
+  select(
+    tin,
+    region_from = region,
+    settlement_from = settlement,
+    lat_from = lat, 
+    lon_from = lon,
+    region_to = next_region, 
+    settlement_to = next_settlement,
[…]

### 2024-02-10 maksakov-readings-conf-2024/Conference-paper.Rmd
Write about data, methods, results; split into paper and conference paper

---
title: "Migrations of Russian Commercial Legal Companies in 2016–23"
author: P. O. Syomin
output: 
  word_document:
    reference_docx: conference-paper-reference.docx
bibliography: refs.bib
csl: "gost-r-7-0-5-2008-numeric-alphabetical.csl"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE, 
  message = FALSE, 
  echo = FALSE, 
  fig.width = 6.4, 
  fig.height = 5,
  fig.id = FALSE,
  dpi = 300)
```

```{r imports, include=FALSE, cache=TRUE}
library(knitr)
source("Script.R")
```

<div custom-style="Date">

Perm State University, 15 Bukireva st., Perm, 614068, Russia, ntsp\@ya.ru

</div>

<div custom-style="Abstract">

<Аннотация>

<br>

*Keywords:* companies migration, small and medium-sized businesses, legal companies, legal service, service geography, data analysis, open data, FTS of Russia

</div>

<br>

Migration of companies, also known as business relocation, is an important phenomenon of economics and geography. Its first studies date back to the middle of 20th century, starting from G. McLaughlin and S. Robock's book *Why industry moves South* that have described the migration of enterprises from the North of the US to the South and linked it to the cheaper labor and smaller activity of labor unions [@McLaughlin1949]. Business relocation may be seen from spatial, relational or decision-making point of view [@Carrincazeaux2015]. A vast amount of research is devoted to various aspects of companies migration, including factors (reasons) [@Nakosteen1987; @Wu2007], directions (places) [@Rupasingha2018; @Rossi2019], discourses [@Piotti2007], taxation effects [@Kubicova2016]. In Russia, there are few papers about migration of companies. Their authors focus mainly on the issue of relocation as an instrument of tax optimization, thus discussing only “negative” part of the problem [@Adigamova2011; @Shcherbakova2016] without diving into factors or places and mostly skipping the geographical aspects of the phenomenon. Thus, it appears that a simple spatial analysis of business relocation in Russia is important. The present study addresses this issue by describing and visualizing the migration of Russian commercial legal companies during the period from the middle of 2016 to the end of 2023. The selection of concrete area of services, i.e. legal firms, is quite a subjective choice, but looks like it is relatively representative. Moreover, it is likely that legal business is more mobile and more sensitive to non-economic factors, although it is just a hypothesis.

**Data.** For the analysis, a complete geo-referenced dataset of Russian small and medium-sized enterprises (SMEs) was used. It was generated from the open and freely available dumps of SME registry distributed by Federal Tax Service (FTS) of Russia [^1]. For the generation of the dataset, the author have developed an open source Python CLI application *ru-smb-companies* [^2]. To select legal companies only, the filtering by activity code 69.10 (*activity in the area of law* according to the All-Russia Activity Code Classifier) was used. The dataset covers the period from August 2016 to November 2023. It contains information about `r companies_count` businesses in tabular format. The data includes, in particular, the following fields: name of the company, its registration address, coordinates of the registration settlement, revenue, expenditure, count of employees (the three last entries are available for each year from 2019 to 2022). It is apperant that some legal companies are missing in the dataset, e.g. large firms which exceeds the limit on the revenue or number of employees for SMEs or advocacy organizations which are categorized as non-profit companies according to the Russian laws and so cannot be included to the list of SMEs. However, there are high chances that it nevertheless can be used as a relevant data source. According to the analysis by Instutite for the Rule of Law [@Moiseeva2016], by August 2016 in Russia there have been 47,000 commercial legal companies, and the dataset used here lists `r `round(companies_count_08_2016, 0)` companies for August 2016. So, we can conclude with high confidence that 3/4 of Russian commercial legal companies are present in our dataset, and this fact proves its reliability.

[^1]: https://www.nalog.gov.ru/opendata/7707329152-rsmp/
[^2]: https://github.com/PavelSyomin/russian-smb-companies

**Methods.** Here, the following definition of migration of company is used: migration is a change in address of registration [@Shcherbakova2016]. Maybe it is too narrow, because does not account for other ways of relocation, e. g. mergers or reorganization. Also, it can be criticized as formalistic, because the change of nominal address does not necessarily reflects the change of place of doing business. However, it looks the best choice for the data available, because the dataset does not contain information about mergers and reorganizations, and does not tell us anything about the actual location of business activity. The analysis itself is done by the methods of descriptive statistics, statistical tests, data visualization and mapping. The following research questions are asked:

1) how are the legal companies distrubuted by the number of migrations?
<br>
2) is there any relationship between the migration choice and the size of company?
<br>
3) what is the distribution of migrations by distance?
<br>
4) are there any temporal patterns in migration?
<br>
5) what is the migration saldo by regions?
<br>
6) which are the typical paths of inter-regional migration?
<br>
7) what is the ratio of inter-regional vs intra-regional migration?

To quantitatively estimate the migration, the first or all the three of the following variables are used: count of companies, sum of revenue, sum of count of employees. Technically, the analysis was performed with R programming language (including additional packages). The source code for the analysis is stored in public repository [^3].

[^3]: https://github.com/PavelSyomin/tax-service-opendata-paper/tree/main/maksakov-readings-conf-2024

A **distribution of companies by the number of migrations** is shown on *Fig. 1* (see below). Almost all the companies (97%) in 2016–2023 have not migrated. About 2.5% of companies have changed the location exactly once, and more frequent migrations are extremely rare. It is seen that the distribution is almost log10-linear: the count in each next group is ~10 times smaller.

<br>

```{r fig-1-mig-distr, fig.cap="*Fig. 1.* The distribution of companies by the number of migrations. Y axis is log10-scaled. The counts are shown above the bars, and percetrages are below the top border", fig.height=3}
mig_count_distr_plot
```

<br>

A **relationship between the migration choice and the size of company** is depicted on Fig. 2. Migration among small firms is more frequent in comparison to mircocompanies: according to exact Fisher test, `r ifelse(categories_test$p.value < 0.001, "p < 0.001", paste("p = ", categories_test$p.value))`, and odds ratio is `r paste0("1 to ", round(1 / categories_test$estimate, 0))`.

<br>

```{r fig-2-mig-by-company-size, fig.cap="*Fig. 2*. Migration choice depending on company's size (category)", fig.height=3}
mig_by_category_plot
```

<br>

A **distribution of migration distance** is shown on * Fig. 3*. The majority of businesses moves to small distance (about 10 km), but there is a significant group of long-migrating firms (the distance is about 1000 km).

<br>

```{r fig-3-distance, fig.cap="*Fig. 3.* Distribution of migration distance"}
mig_distance_plot
```

**Распределение миграций по времени** показано на графиках 2 и 3. Закономерности за весь рассмотренный период отсутствуют: интенсивность миграций выглядит случайной. Тем не менее, видно, что компании чаще меняют своё местонахождение летом.

**Распределение регионов по притоку (оттоку) компаний** изображено на картограмме 1. Общая закономерность отсутствует, при этом заметно, что показатели противоречат друг другу: регионы, лидирующие по сальдо исходя из числа компаний, могут проигрывать при расчёте по выручке или численности работников.

**Распределение населённых пунктов по притоку (оттоку) компаний** показано на картограмме 2. Здесь также сложно выявить закономерности, хотя лидируют по сальдо, измеренному как число компаний, города-спутники Москвы. Опять же, показатели противоречивые: лидерство по численному сальдо не означает лидерство по выручке или числу работников.

**Соотношение внутрирегиональной миграции и межрегиональной миграции** показано в таблице 3. Видно, что между регионами перемещается примерно треть фирм, а остальные две трети — внутри региона. При этом в пересчёте на число работников и выручку межрегиональная миграция масштабнее, чем внутрирегиональная, из чего следует, что компания, поменявшая регион, в среднем «везёт» с собой больше денег и людей, чем компания, поменявшая населённый пункт внутри региона.

**Пути межрегиональной миграции** отражены в таблице 4. Абсолютные числа переместившихся компаний, как правило, невелики. Лидируют миграции из Москвы и Санкт-Петербурга в прилегающие к ним области. Нет чёткой связи между числом переехавших компаний и «перевезённой» выручкой или числом работников.

**Сочетание централизации и децентрализации в межрегиональной миграции** (таблица 5) преимущественно равномерное. По числу работников и компаний преобладает центробежная тенденция, а по доходам — центростремительная. Треть потоков не может быть описани ни как центростремительные, ни как центробежные.

**Пути внутрирегиональной миграции** (таблица 6) ещё более неоднозначны, чем пути межрегиональной миграции, поскольку абсолютные перемещения почти во всех случаях измеряются единицами. Нет и связи между тремя показателями: число фирм, доходы, число работников.

**Сочетание централизации и децентрализации во внутрирегиональной миграции** (таблица 7) почти такое же, как и в случае с межрегиональной миграцией: перемещения к центру выигрывают по числу фирм и людей, но уступают в доходах, при этом треть релокаций не могут быть описаны ни как центростремительные, ни как центробежные.

# Bibliographical references

<div id="refs"></div>

```{r table-2-mig-cats}
kable(
  count_by_category,
  col.names = c("Категория фирмы", "Всего", "Из них мигрировавших"),
  caption = "Таблица 2. Распределение юридических компаний по категории и статусу миграции"
)
```

```{r table-3-inter-intra}
kable(
  inter_intra,
  col.names = c("Тип миграции", "Число", "Доходы", "Работники"),
  caption = "Таблица 3. Показатели межрегиональной и внутрирегиональной миграции юридических компаний: число фирм, суммарные доходы (млн рублей), суммарное число работников"
)
```

```{r table-4-reg-paths}
kable(
  rbind(head(regional_paths_combined, 5), tail(regional_paths_combined, 5)),
  col.names = c("Путь", "Количество", "Доходы, млн", "Работники"),
  caption = "Таблица 4. Маршруты миграции юридических компаний по субъектам России и их показатели: число фирм, суммарные доходы в миллионах рублей, суммарная численность работников. Поразаны первые и последние 5 маршрутов по числу фирм"
)
```

```{r table-5-reg-central}
kable(
  reg_central,
  col.names = c("Тип миграции", "Число", "Доходы", "Работники"),
  caption = "Таблица 5. Показатели центростремительных и центробежных межрегиональных миграций юридических компаний: число фирм, суммарные доходы в миллионах рублей, суммарная численность работников"
)
```

```{r table-6-settl-paths}
kable(
  rbind(head(settlement_paths_combined, 5), tail(settlement_paths_combined, 5)),
  col.names = c("Маршрут", "Количество", "Доходы, млн", "Работники"),
  caption = "Таблица 6. Маршруты миграции юридических компаний по населённым пунктам и их показатели: число фирм, суммарные доходы в миллионах рублей, суммарная численность работников. Поразаны первые и последние 5 маршрутов по числу фирм"
)
```

```{r table-7-settl-central}
kable(
  settl_central,
  col.names = c("Тип миграции", "Число", "Доходы", "Работники"),
  caption = "Таблица 7. Показатели центростремительных и центробежных внутрирегиональных миграций юридических компаний: число фирм, суммарные доходы в миллионах рублей, суммарная численность работников"
)
```

```{r plot-2-timeline}
mig_timeline_plot
```

```{r plot-3-monthly}
mig_by_month_plot
```

```{r map-1-regions}

```

```{r map-2-settlements}

```

### 2024-02-10 maksakov-readings-conf-2024/Script.R

@@ -13,17 +13,45 @@ empl <- read_csv("../../ru-smb-companies/legal/empl.csv")
 cities <- read_csv("cities.csv")
 regions <- st_read("ru.geojson")
 
+# Number of unique companies
+companies_count <- n_distinct(data[data$kind == 1, "tin"])
+
+# Number of companies as of 2016-08-10 (to compare with Moiseeva2016)
+companies_count_08_2016 <- nrow(filter(data, kind == 1, start_date == "2016-08-10"))
+
 # Calculate the absolute numbers and shares of companies
 # with changes in location
 # Location is defined as change either in region or settlement
-data %>% 
+migration_count_distribution <- data %>% 
   filter(kind == 1) %>% 
   group_by(tin) %>%
   distinct(region, settlement) %>% 
   count() %>% 
   ungroup() %>% 
   count(n, name = "count") %>% 
-  mutate(share = 100 * count / sum(count))
+  mutate(
+    n = n - 1,
+    share = 100 * count / sum(count))
+
+# Draw it on plot
+mig_count_distr_plot <- migration_count_distribution %>% 
+  ggplot(aes(x = n, y = count)) +
+  geom_col() +
+  geom_text(
+    aes(y = 1e6, label = scales::number(share, accuracy = 0.01, suffix = " %")),
+    color = "gray30", size = 3
+    ) + 
+  geom_text(aes(label = count), vjust = -.5) + 
+  scale_y_log10(limits = c(NA, 1e6), breaks = 10^(1:5), labels = ~ .x) +
+  labs(
+    x = "Count of migrations by a company", 
+    y = "Count of companies (log10 scale)") +
+  theme_bw() +
+  theme(
+    panel.grid.minor.x = element_blank(), 
+    panel.grid.major.x = element_blank(),
+    panel.grid.minor.y = element_blank())
+mig_count_distr_plot
 
 # Prepare the base dataset to analyze migrations of companies
 max_empl <- c(15, 100, 250)
@@ -78,29 +106,53 @@ migrations <- data %>%
   st_drop_geometry()
 
 # Distribution of companies' sizes by migration status
+smb_categories <- c("Микропредприятие", "Малое предприятие", "Среднее предприятие")
 c1 <- data %>% 
   distinct(tin, .keep_all = TRUE) %>% 
   count(category)
 c2 <- migrations %>% 
   distinct(tin, .keep_all = TRUE) %>% 
   count(category)
[…]

### 2024-02-11 maksakov-readings-conf-2024/Conference-paper.Rmd
Translate Russian biblio items to English, finish first draft of conference paper, adjust styles

@@
   dpi = 300)
+library(knitr)
 ```
@@
 ```{r imports, include=FALSE, cache=TRUE}
-library(knitr)
 source("Script.R")
@@
 
-<Аннотация>
+Migrations of Russian legal companies in 2016–23 are described and analysed.
+The study uses Federal Tax Service open data on small and medium-sized companies.
+Distributions by migration count and distance are estimated and visualized, the relationship between company size and migration decision is examined and statistically tested, net migration rates by region are mapped, and top migration paths are listed.
+It is found that migrations are rare, mostly short-distance (~10 km).
+Bigger companies tend to relocate more often.
+Moscow and St.
+Petersburg have probably lost small and inefficient businesses but gained larger and more profitable.
 
@@
 
-*Keywords:* companies migration, small and medium-sized businesses, legal companies, legal service, service geography, data analysis, open data, FTS of Russia
+*Keywords:* companies migration, small and medium-sized businesses, legal companies, legal service, service geography, data analysis, open data, Federal Tax Service.
 
@@
 In Russia, there are few papers about migration of companies.
-Their authors focus mainly on the issue of relocation as an instrument of tax optimization, thus discussing only “negative” part of the problem [@Adigamova2011; @Shcherbakova2016] without diving into factors or places and mostly skipping the geographical aspects of the phenomenon.
+Their authors focus mainly on the issue of relocation as an instrument of tax optimization, thus discussing only “negative” part of the problem [@Adigamova2011_en; @Shcherbakova2016_en] without diving into factors or places and mostly skipping the geographical aspects of the phenomenon.
 Thus, it appears that a simple spatial analysis of business relocation in Russia is important.
@@
 However, there are high chances that it nevertheless can be used as a relevant data source.
-According to the analysis by Instutite for the Rule of Law [@Moiseeva2016], by August 2016 in Russia there have been 47,000 commercial legal companies, and the dataset used here lists `r `round(companies_count_08_2016, 0)` companies for August 2016.
+According to the analysis by Instutite for the Rule of Law [@Moiseeva2016_en], by August 2016 in Russia there have been 47,000 commercial legal companies, and the dataset used here lists `r `round(companies_count_08_2016, 0)` companies for August 2016.
 So, we can conclude with high confidence that 3/4 of Russian commercial legal companies are present in our dataset, and this fact proves its reliability.
@@
 
-**Methods.** Here, the following definition of migration of company is used: migration is a change in address of registration [@Shcherbakova2016].
+**Methods.** Here, the following definition of migration of company is used: migration is a change in address of registration [@Shcherbakova2016_en].
 Maybe it is too narrow, because does not account for other ways of relocation, e.
@@
 However, it looks the best choice for the data available, because the dataset does not contain information about mergers and reorganizations, and does not tell us anything about the actual location of business activity.
-The analysis itself is done by the methods of descriptive statistics, statistical tests, data visualization and mapping.
-The following research questions are asked:
-
-1) how are the legal companies distrubuted by the number of migrations?
-<br>
-2) is there any relationship between the migration choice and the size of company?
-<br>
-3) what is the distribution of migrations by distance?
-<br>
-4) are there any temporal patterns in migration?
-<br>
-5) what is the migration saldo by regions?
-<br>
-6) which are the typical paths of inter-regional migration?
-<br>
-7) what is the ratio of inter-regional vs intra-regional migration?
+The analysis is done by the means of descriptive statistics, statistical tests, data visualization and mapping.
+It includes five elements: (1) distribution of legal companies by the count of migrations, (2) relationship between the migration decision and size of a company; (3) distribution of migration distance; (4) net migration rate by region; (5) net migration rate by inter-region migration paths.
 
@@
 
-```{r fig-2-mig-by-company-size, fig.cap="*Fig. 2*. Migration choice depending on company's size (category)", fig.height=3}
+```{r fig-2-mig-by-company-size, fig.cap="*Fig. 2*. Migration choice depending on company's size (category)", fig.height=2}
 mig_by_category_plot
@@
 
-A **distribution of migration distance** is shown on * Fig.
+A **distribution of migration distance** is shown on *Fig.
 3*.
@@
 
-```{r fig-3-distance, fig.cap="*Fig. 3.* Distribution of migration distance"}
+```{r fig-3-distance, fig.cap="*Fig. 3.* Distribution of migration distance", fig.height=3}
 mig_distance_plot
@@
 
-**Распределение миграций по времени** показано на графиках 2 и 3.
-Закономерности за весь рассмотренный период отсутствуют: интенсивность миграций выглядит случайной.
-Тем не менее, видно, что компании чаще меняют своё местонахождение летом.
+<br>
 
-**Распределение регионов по притоку (оттоку) компаний** изображено на картограмме 1.
-Общая закономерность отсутствует, при этом заметно, что показатели противоречат друг другу: регионы, лидирующие по сальдо исходя из числа компаний, могут проигрывать при расчёте по выручке или численности работников.
+**Net migration rate by region** is shown on map (see *Fig.
+4*).
+Apparently, most of the regions have net migration rate close to zero.
+However, there are some regions with substantial growth or decline in the number of registered legal companies.
+For example, Moscow region is the top by growth, and the moderate positive net migration rate is also characteristic of Saint Petersburg, Krasnodar, Rostov, Novosibirsk regions (oblasts), Perm region (krai) and Udmurt republic.
+The desrease in the number of legal businesses is typical for Volga region, the Urals, Western Siberia and Krasnoyarsk region (krai).
+It is worth mentioning that Moscow itself (as a city) also shows the decline.
+However, it is also obvious that the absolute numbers of growth or decline are small compared to the absolute number of registered companies and do not exceed 100 in any of the regions.
 
-**Распределение населённых пунктов по притоку (оттоку) компаний** показано на картограмме 2.
-Здесь также сложно выявить закономерности, хотя лидируют по сальдо, измеренному как число компаний, города-спутники Москвы.
-Опять же, показатели противоречивые: лидерство по численному сальдо не означает лидерство по выручке или числу работников.
+<br>
 
-**Соотношение внутрирегиональной миграции и межрегиональной миграции** показано в таблице 3.
-Видно, что между регионами перемещается примерно треть фирм, а остальные две трети — внутри региона.
-При этом в пересчёте на число работников и выручку межрегиональная миграция масштабнее, чем внутрирегиональная, из чего следует, что компания, поменявшая регион, в среднем «везёт» с собой больше денег и людей, чем компания, поменявшая населённый пункт внутри региона.
+```{r fig-4-mig-by-region-plot, fig.cap="*Fig. 4.* Net migration rate of legal companies by regions", fig.height=4}
+mig_by_region_plot
+```
 
-**Пути межрегиональной миграции** отражены в таблице 4.
-Абсолютные числа переместившихся компаний, как правило, невелики.
-Лидируют миграции из Москвы и Санкт-Петербурга в прилегающие к ним области.
-Нет чёткой связи между числом переехавших компаний и «перевезённой» выручкой или числом работников.
+<br>
 
-**Сочетание централизации и децентрализации в межрегиональной миграции** (таблица 5) преимущественно равномерное.
-По числу работников и компаний преобладает центробежная тенденция, а по доходам — центростремительная.
-Треть потоков не может быть описани ни как центростремительные, ни как центробежные.
+**The most popular migration paths** (found by the number of migrated companies) with corresponding total revenues and employee counts are listed in the **Table**.
+It is important to note that the first two of them represent the moving from two largest cities of Russia (including the capital) to the surrounding regions (oblasts).
+However, the migration from other regions to the capital is popular too.
+Besides, the increase in the number of registered businesses is not necessarily associated with the increase in income or labor force, and vise versa, that is, even if a net migration rate is above zero, the total revenue and employees count may have decreased, possibly meaning that arriving businesses are profitable and large, while leaving companies are small and inefficient.
+This is the case, for example, of the aforementioned Moscow-to-region and Saint-Petersburg-to-region paths, where the net change of the number of firms is in favor of regions, while the net change in revenue and labor is positive for the cities.
 
-**Пути внутрирегиональной миграции** (таблица 6) ещё более неоднозначны, чем пути межрегиональной миграции, поскольку абсолютные перемещения почти во всех случаях измеряются единицами.
-Нет и связи между тремя показателями: число фирм, доходы, число работников.
+<br>
 
-**Сочетание централизации и децентрализации во внутрирегиональной миграции** (таблица 7) почти такое же, как и в случае с межрегиональной миграцией: перемещения к центру выигрывают по числу фирм и людей, но уступают в доходах, при этом треть релокаций не могут быть описаны ни как центростремительные, ни как центробежные.
+```{r table}
+kable(
+  head(regional_paths_combined, 10),
+  col.names = c("Region", "Count", "Revenue", "Employees"),
+  caption = "Top 10 paths of inter-regional migration of Russian legal companies in 2016–2023 and their features: net count of relocated businesses, net total revenue, net total number of employees"
+)
+```
+
+**Conclusions.** Migration of legal companies in Russia is a rare phenomenon: just about 2.5% of all businesses in this activity area have changed their nominal location in the period of 2016–23.
+Small companies, in comparison to micro-businesses, are more likely to relocate: the chances are 4 times higher.
+Short-distance (~10 km) migrations are more common than long-distance (100–1000 km) ones.
+Moscow and St.
+Petersburg regions are at the top by net migration rate estimated by the count of relocated businesses, while Moscow have lost the highest number of companies.
+However, it looks like that while losing small and inefficient companies, Moscow and St.
+Patersburg attract more profitable businesses.
 
@@
 <div id="refs"></div>
-
-```{r table-2-mig-cats}
-kable(
-  count_by_category,
-  col.names = c("Категория фирмы", "Всего", "Из них мигрировавших"),
-  caption = "Таблица 2. Распределение юридических компаний по категории и статусу миграции"
-)
-```
-
-```{r table-3-inter-intra}
-kable(
-  inter_intra,
-  col.names = c("Тип миграции", "Число", "Доходы", "Работники"),
-  caption = "Таблица 3. Показатели межрегиональной и внутрирегиональной миграции юридических компаний: число фирм, суммарные доходы (млн рублей), суммарное число работников"
-)
-```
-
-```{r table-4-reg-paths}
-kable(
-  rbind(head(regional_paths_combined, 5), tail(regional_paths_combined, 5)),
-  col.names = c("Путь", "Количество", "Доходы, млн", "Работники"),
-  caption = "Таблица 4.
-Маршруты миграции юридических компаний по субъектам России и их показатели: число фирм, суммарные доходы в миллионах рублей, суммарная численность работников.
-Поразаны первые и последние 5 маршрутов по числу фирм"
-)
-```
-
-```{r table-5-reg-central}
-kable(
-  reg_central,
-  col.names = c("Тип миграции", "Число", "Доходы", "Работники"),
-  caption = "Таблица 5. Показатели центростремительных и центробежных межрегиональных миграций юридических компаний: число фирм, суммарные доходы в миллионах рублей, суммарная численность работников"
-)
-```
-
-```{r table-6-settl-paths}
-kable(
-  rbind(head(settlement_paths_combined, 5), tail(settlement_paths_combined, 5)),
-  col.names = c("Маршрут", "Количество", "Доходы, млн", "Работники"),
-  caption = "Таблица 6.
-Маршруты миграции юридических компаний по населённым пунктам и их показатели: число фирм, суммарные доходы в миллионах рублей, суммарная численность работников.
-Поразаны первые и последние 5 маршрутов по числу фирм"
-)
-```
-
-```{r table-7-settl-central}
-kable(
-  settl_central,
-  col.names = c("Тип миграции", "Число", "Доходы", "Работники"),
-  caption = "Таблица 7.
-Показатели центростремительных и центробежных внутрирегиональных миграций юридических компаний: число фирм, суммарные доходы в миллионах рублей, суммарная численность работников"
-)
-```
-
-```{r plot-2-timeline}
-mig_timeline_plot
-```
-
-```{r plot-3-monthly}
-mig_by_month_plot
-```
-
-```{r map-1-regions}
-
-```
-
-```{r map-2-settlements}
-
-```

### 2024-02-11 maksakov-readings-conf-2024/Script.R

@@ -11,7 +11,8 @@ data <- read_csv("../../ru-smb-companies/legal/smb.csv")
 revexp <- read_csv("../../ru-smb-companies/legal/revexp.csv")
 empl <- read_csv("../../ru-smb-companies/legal/empl.csv")
 cities <- read_csv("cities.csv")
-regions <- st_read("ru.geojson")
+regions_geo <- st_read("ru.geojson")
+regions <- read_csv("regions.csv")
 
 # Number of unique companies
 companies_count <- n_distinct(data[data$kind == 1, "tin"])
@@ -81,7 +82,8 @@ migrations <- data %>%
   replace_na(list(revenue = 0, expenditure = 0, employees_count = 0)) %>% 
   mutate(
     employees_count = ifelse(employees_count > 500, 0, employees_count),
-    revenue = ifelse(revenue > 1e10, 0, revenue)
+    revenue = ifelse(revenue > 1e10, 0, revenue),
+    revenue = revenue / 1e6
   ) %>% 
   select(
     tin,
@@ -182,7 +184,19 @@ migration_by_region <- migrations %>%
   mutate(change = region_to - region_from) %>% 
   select(region, var, change) %>% 
   pivot_wider(region, names_from = var, values_from = change) %>% 
-  arrange(-count)
+  arrange(-count) %>% 
+  left_join(select(regions, name, iso_code), by = c("region" = "name")) %>% 
+  right_join(select(regions_geo, shapeISO), by = c("iso_code" = "shapeISO")) %>% 
+  st_as_sf()
+
+ru_crs <- st_crs("+proj=aea +lat_0=0 +lon_0=100 +lat_1=68 +lat_2=44 +x_0=0 +y_0=0 +ellps=krass +towgs84=28,-130,-95,0,0,0,0 +units=m +no_defs")
+mig_by_region_plot <- migration_by_region %>% 
+  ggplot() +
+  geom_sf(aes(fill = count), size = .1) + 
+  coord_sf(crs = ru_crs) +
+  scale_fill_gradient2(name = "Net migration rate", low = "#d01c8b", high = "#4dac26") +
+  theme_bw(base_size = 12, base_family = "Times New Roman") +
+  theme(legend.position = "bottom")
 
 migration_msk_spb <- migration_by_region %>% 
   filter(region %in% c("Москва", "Санкт-Петербург")) %>% 

### 2024-02-12 maksakov-readings-conf-2024/Conference-paper.Rmd
Minor fixes in CSL translation and docx styles, compression of paper text, finish of translation of plots and tables, fix of a bug with paths table

@@
 ---
-title: "Migrations of Russian Commercial Legal Companies in 2016–23"
+title: "UDC 911.3"
+subtitle: "Migration of Russian Legal Companies in 2016–23"
 author: P. O. Syomin
@@
 bibliography: refs.bib
-csl: "gost-r-7-0-5-2008-numeric-alphabetical.csl"
+csl: "gost-r-7-0-5-2008-numeric-alphabetical-en.csl"
 ---
@@
 
-Migrations of Russian legal companies in 2016–23 are described and analysed.
-The study uses Federal Tax Service open data on small and medium-sized companies.
-Distributions by migration count and distance are estimated and visualized, the relationship between company size and migration decision is examined and statistically tested, net migration rates by region are mapped, and top migration paths are listed.
-It is found that migrations are rare, mostly short-distance (~10 km).
+Migration of Russian commercial legal companies in 2016–23 is described and analysed.
+Distributions by migration count and distance are estimated and visualized, the relationship between company size and migration decision is statistically tested, net migration rates by region are mapped, and top migration paths are listed.
+It is found that migrations are rare, mostly short-distance (10–50 km).
 Bigger companies tend to relocate more often.
-Moscow and St.
-Petersburg have probably lost small and inefficient businesses but gained larger and more profitable.
+Moscow has probably lost small and inefficient businesses but gained larger and more profitable.
 
@@
 
-*Keywords:* companies migration, small and medium-sized businesses, legal companies, legal service, service geography, data analysis, open data, Federal Tax Service.
+***Keywords:*** companies migration, small and medium-sized businesses, legal companies, legal service, service geography, data analysis, open data, Federal Tax Service.
 
@@
 
-Migration of companies, also known as business relocation, is an important phenomenon of economics and geography.
+Migration of companies (business relocation) is an important phenomenon of economy and geography.
 Its first studies date back to the middle of 20th century, starting from G.
 McLaughlin and S.
-Robock's book *Why industry moves South* that have described the migration of enterprises from the North of the US to the South and linked it to the cheaper labor and smaller activity of labor unions [@McLaughlin1949].
-Business relocation may be seen from spatial, relational or decision-making point of view [@Carrincazeaux2015].
-A vast amount of research is devoted to various aspects of companies migration, including factors (reasons) [@Nakosteen1987; @Wu2007], directions (places) [@Rupasingha2018; @Rossi2019], discourses [@Piotti2007], taxation effects [@Kubicova2016].
+Robock's book *Why industry moves South* that have described the migration of enterprises from the North of the US and linked it to the cheaper labor and smaller activity of labor unions [@McLaughlin1949].
+Various aspects of companies migration have been studied, including factors (reasons) [@Wu2007], directions (places) [@Rossi2019], taxation effects [@Kubicova2016].
+
 In Russia, there are few papers about migration of companies.
-Their authors focus mainly on the issue of relocation as an instrument of tax optimization, thus discussing only “negative” part of the problem [@Adigamova2011_en; @Shcherbakova2016_en] without diving into factors or places and mostly skipping the geographical aspects of the phenomenon.
-Thus, it appears that a simple spatial analysis of business relocation in Russia is important.
-The present study addresses this issue by describing and visualizing the migration of Russian commercial legal companies during the period from the middle of 2016 to the end of 2023.
-The selection of concrete area of services, i.e.
-legal firms, is quite a subjective choice, but looks like it is relatively representative.
-Moreover, it is likely that legal business is more mobile and more sensitive to non-economic factors, although it is just a hypothesis.
+Existing works focus  on relocation as an instrument of tax optimization [@Adigamova2011_en; @Shcherbakova2016_en], skipping the geographical overview and spatial aspects of the phenomenon.
+So, a spatial analysis of business relocation in Russia is important.
+This study describes and visualizes the migration of Russian commercial legal companies in 2016–23.
+The choice of particular area of economy, that is, legal service, is subjective.
 
-**Data.** For the analysis, a complete geo-referenced dataset of Russian small and medium-sized enterprises (SMEs) was used.
-It was generated from the open and freely available dumps of SME registry distributed by Federal Tax Service (FTS) of Russia [^1].
-For the generation of the dataset, the author have developed an open source Python CLI application *ru-smb-companies* [^2].
-To select legal companies only, the filtering by activity code 69.10 (*activity in the area of law* according to the All-Russia Activity Code Classifier) was used.
+**Data.** A geo-referenced dataset of Russian small and medium-sized businesses (SMBs) with main activity code equal to 69.10 (activity in the area of law) was generated from the open dumps of SMB registry[^1] with an open source Python CLI application *ru-smb-companies*[^2] developer by the author of this paper.
 The dataset covers the period from August 2016 to November 2023.
-It contains information about `r companies_count` businesses in tabular format.
-The data includes, in particular, the following fields: name of the company, its registration address, coordinates of the registration settlement, revenue, expenditure, count of employees (the three last entries are available for each year from 2019 to 2022).
-It is apperant that some legal companies are missing in the dataset, e.g.
-large firms which exceeds the limit on the revenue or number of employees for SMEs or advocacy organizations which are categorized as non-profit companies according to the Russian laws and so cannot be included to the list of SMEs.
-However, there are high chances that it nevertheless can be used as a relevant data source.
-According to the analysis by Instutite for the Rule of Law [@Moiseeva2016_en], by August 2016 in Russia there have been 47,000 commercial legal companies, and the dataset used here lists `r `round(companies_count_08_2016, 0)` companies for August 2016.
-So, we can conclude with high confidence that 3/4 of Russian commercial legal companies are present in our dataset, and this fact proves its reliability.
+According to [@Moiseeva2016_en], by August 2016 in Russia there have been 47000 commercial legal companies, and the dataset contains about `r round(companies_count_08_2016, -3)` companies with the same date.
+Some legal companies are missing, e.
+g.
+large ones that exceed the limit on the revenue or number of employees for SMBs, but 3/4 are present, so the dataset looks reliable and relevant.
 
@@
 
-**Methods.** Here, the following definition of migration of company is used: migration is a change in address of registration [@Shcherbakova2016_en].
-Maybe it is too narrow, because does not account for other ways of relocation, e.
+**Methods.** Company migration is seen as a change in address of registration [@Shcherbakova2016_en].
+It is narrow and does not account for other ways of relocation (e.
 g.
-mergers or reorganization.
-Also, it can be criticized as formalistic, because the change of nominal address does not necessarily reflects the change of place of doing business.
-However, it looks the best choice for the data available, because the dataset does not contain information about mergers and reorganizations, and does not tell us anything about the actual location of business activity.
-The analysis is done by the means of descriptive statistics, statistical tests, data visualization and mapping.
-It includes five elements: (1) distribution of legal companies by the count of migrations, (2) relationship between the migration decision and size of a company; (3) distribution of migration distance; (4) net migration rate by region; (5) net migration rate by inter-region migration paths.
-
-To quantitatively estimate the migration, the first or all the three of the following variables are used: count of companies, sum of revenue, sum of count of employees.
+mergers or reorganization).
+Also, it is formal, because the change of nominal address does not necessarily reflects the change of actual place of doing business.
+However, it looks like the best choice for the data available.
+Descriptive statistics, statistical tests, data visualization and mapping are used.
 Technically, the analysis was performed with R programming language (including additional packages).
-The source code for the analysis is stored in public repository [^3].
+The source code for the analysis is stored in public repository[^3].
 
@@
 A **distribution of companies by the number of migrations** is shown on *Fig.
-1* (see below).
-Almost all the companies (97%) in 2016–2023 have not migrated.
+1*.
+Almost all the companies (97%) in 2016–23 have not migrated.
 About 2.5% of companies have changed the location exactly once, and more frequent migrations are extremely rare.
-It is seen that the distribution is almost log10-linear: the count in each next group is ~10 times smaller.
+The distribution is almost log10-linear.
 
@@
 Y axis is log10-scaled.
-The counts are shown above the bars, and percetrages are below the top border", fig.height=3}
+The counts are shown above the bars, and percetrages are below the top border", fig.height=2.5}
 mig_count_distr_plot
@@
 
-A **relationship between the migration choice and the size of company** is depicted on Fig.
-2.
-Migration among small firms is more frequent in comparison to mircocompanies: according to exact Fisher test, `r ifelse(categories_test$p.value < 0.001, "p < 0.001", paste("p = ", categories_test$p.value))`, and odds ratio is `r paste0("1 to ", round(1 / categories_test$estimate, 0))`.
+A **relationship between the migration decision and the size of company** is depicted on *Fig.
+2*.
+Migration among small firms is more frequent in comparison to micro-businesses: according to Fisher's exact test, `r ifelse(categories_test$p.value < 0.001, "p < 0.001", paste("p = ", categories_test$p.value))`, and odds ratio is `r paste0("1 to ", round(1 / categories_test$estimate, 0))`.
 
@@
 
-```{r fig-2-mig-by-company-size, fig.cap="*Fig. 2*. Migration choice depending on company's size (category)", fig.height=2}
+```{r fig-2-mig-by-company-size, fig.cap="*Fig. 2*. Migration decision depending on company's size (category)", fig.height=1}
 mig_by_category_plot
@@
 3*.
-The majority of businesses moves to small distance (about 10 km), but there is a significant group of long-migrating firms (the distance is about 1000 km).
+The majority of businesses moves to small distance (10–50 km), but there is a significant group of long-migrating firms (the distance is about 1000 km).
 
@@
 
-```{r fig-3-distance, fig.cap="*Fig. 3.* Distribution of migration distance", fig.height=3}
+```{r fig-3-distance, fig.cap="*Fig. 3.* Distribution of migration distance", fig.height=2.5}
 mig_distance_plot
@@
 4*).
-Apparently, most of the regions have net migration rate close to zero.
-However, there are some regions with substantial growth or decline in the number of registered legal companies.
-For example, Moscow region is the top by growth, and the moderate positive net migration rate is also characteristic of Saint Petersburg, Krasnodar, Rostov, Novosibirsk regions (oblasts), Perm region (krai) and Udmurt republic.
-The desrease in the number of legal businesses is typical for Volga region, the Urals, Western Siberia and Krasnoyarsk region (krai).
-It is worth mentioning that Moscow itself (as a city) also shows the decline.
-However, it is also obvious that the absolute numbers of growth or decline are small compared to the absolute number of registered companies and do not exceed 100 in any of the regions.
+Most of the regions have net migration rate close to zero.
+However, there are some regions with substantial growth (Moscow, St.
+Petersburg, Krasnodar, Rostov, Novosibirsk oblast, Perm krai) or decline (Moscow as a city) in the number of registered legal companies.
+The absolute counts are small (<100) compared to the quantity of registered companies.
 
@@
 
-```{r fig-4-mig-by-region-plot, fig.cap="*Fig. 4.* Net migration rate of legal companies by regions", fig.height=4}
+```{r fig-4-mig-by-region-plot, fig.cap="*Fig. 4.* Net migration rate of legal companies by regions", fig.height=3}
 mig_by_region_plot
@@
 
-**The most popular migration paths** (found by the number of migrated companies) with corresponding total revenues and employee counts are listed in the **Table**.
-It is important to note that the first two of them represent the moving from two largest cities of Russia (including the capital) to the surrounding regions (oblasts).
-However, the migration from other regions to the capital is popular too.
-Besides, the increase in the number of registered businesses is not necessarily associated with the increase in income or labor force, and vise versa, that is, even if a net migration rate is above zero, the total revenue and employees count may have decreased, possibly meaning that arriving businesses are profitable and large, while leaving companies are small and inefficient.
-This is the case, for example, of the aforementioned Moscow-to-region and Saint-Petersburg-to-region paths, where the net change of the number of firms is in favor of regions, while the net change in revenue and labor is positive for the cities.
+**The most popular migration paths** (by the count of migrated companies) with summed up revenues and employee counts are listed in the **Table**.
+The first two paths represent the relocation from two largest cities of Russia to the surrounding regions.
+Net change in the number of businesses is relatively small.
+An increase in the number of registered businesses is not necessarily associated with the increase in income or labor force, and vise versa.
+Positive flow of companies with negative flow of revenue or labor probably means that arriving businesses are larger and more profitable than leaving.
 
@@
 kable(
-  head(regional_paths_combined, 10),
-  col.names = c("Region", "Count", "Revenue", "Employees"),
-  caption = "Top 10 paths of inter-regional migration of Russian legal companies in 2016–2023 and their features: net count of relocated businesses, net total revenue, net total number of employees"
+  head(regional_paths_combined_en, 10),
+  col.names = c("Path", "Count", "Revenue, M rub", "Employees"),
+  caption = "Top 10 paths of inter-regional migration of Russian legal companies in 2016–23",
+  digits = 0
 )
@@
 
-**Conclusions.** Migration of legal companies in Russia is a rare phenomenon: just about 2.5% of all businesses in this activity area have changed their nominal location in the period of 2016–23.
+<br>
+
+**Conclusion.** Migration of legal companies in Russia is a rare phenomenon: only about 2.5% of all businesses in this activity area have changed their nominal location in the period of 2016–23.
 Small companies, in comparison to micro-businesses, are more likely to relocate: the chances are 4 times higher.
-Short-distance (~10 km) migrations are more common than long-distance (100–1000 km) ones.
+Short-distance (10–50 km) migrations are more common than long-distance (~1000 km) ones.
 Moscow and St.
 Petersburg regions are at the top by net migration rate estimated by the count of relocated businesses, while Moscow have lost the highest number of companies.
-However, it looks like that while losing small and inefficient companies, Moscow and St.
-Patersburg attract more profitable businesses.
+However, it looks like that while losing small and inefficient companies, Moscow attracts more profitable businesses.
 

### 2024-02-12 maksakov-readings-conf-2024/Script.R

@@ -45,9 +45,9 @@ mig_count_distr_plot <- migration_count_distribution %>%
   geom_text(aes(label = count), vjust = -.5) + 
   scale_y_log10(limits = c(NA, 1e6), breaks = 10^(1:5), labels = ~ .x) +
   labs(
-    x = "Count of migrations by a company", 
-    y = "Count of companies (log10 scale)") +
-  theme_bw() +
+    x = "Number of migrations by a company", 
+    y = "Number of companies (log10 scale)") +
+  theme_bw(base_size = 12, base_family = "Times New Roman") +
   theme(
     panel.grid.minor.x = element_blank(), 
     panel.grid.major.x = element_blank(),
@@ -133,15 +133,13 @@ mig_by_category_plot <- count_by_category %>%
   pivot_longer(cols = mig:stay, names_to = "option", values_to = "count") %>% 
   ggplot(aes(x = category, y = count, fill = option)) +
     geom_col(width = 0.5) +
+    scale_x_discrete(labels = c("Микропредприятие"  = "Micro-business", "Малое предприятие" = "Small company")) +
     scale_y_continuous(labels = scales::label_percent()) +
     scale_fill_discrete(name = "", labels = c("Migrated", "Stayed")) +
     coord_flip() +
     labs(x = "", y = "Share of companies") +
-    theme_bw() +
-    theme(
-      legend.position = "bottom", 
-      legend.direction = "horizontal",
-      panel.grid.minor.x = element_blank())
+    theme_bw(base_size = 12, base_family = "Times New Roman") +
+    theme(panel.grid.minor.x = element_blank())
     
 # Distance of migrations
 mig_distance_plot <- migrations %>% 
@@ -194,9 +192,8 @@ mig_by_region_plot <- migration_by_region %>%
   ggplot() +
   geom_sf(aes(fill = count), size = .1) + 
   coord_sf(crs = ru_crs) +
-  scale_fill_gradient2(name = "Net migration rate", low = "#d01c8b", high = "#4dac26") +
-  theme_bw(base_size = 12, base_family = "Times New Roman") +
-  theme(legend.position = "bottom")
+  scale_fill_gradient2(name = "", low = "#d01c8b", high = "#4dac26") +
+  theme_bw(base_size = 12, base_family = "Times New Roman")
 
 migration_msk_spb <- migration_by_region %>% 
[…]

### 2024-02-12 maksakov-readings-conf-2024/Conference-paper.Rmd
Proofreading with Grammarly

@@
 
-Migration of Russian commercial legal companies in 2016–23 is described and analysed.
+The migration of Russian commercial legal companies in 2016–23 is described and analyzed.
 Distributions by migration count and distance are estimated and visualized, the relationship between company size and migration decision is statistically tested, net migration rates by region are mapped, and top migration paths are listed.
@@
 Bigger companies tend to relocate more often.
-Moscow has probably lost small and inefficient businesses but gained larger and more profitable.
+Moscow has probably lost small and inefficient businesses but gained larger and more profitable ones.
 
@@
 Migration of companies (business relocation) is an important phenomenon of economy and geography.
-Its first studies date back to the middle of 20th century, starting from G.
+Its first studies date back to the middle of the 20th century, starting from G.
 McLaughlin and S.
-Robock's book *Why industry moves South* that have described the migration of enterprises from the North of the US and linked it to the cheaper labor and smaller activity of labor unions [@McLaughlin1949].
-Various aspects of companies migration have been studied, including factors (reasons) [@Wu2007], directions (places) [@Rossi2019], taxation effects [@Kubicova2016].
+Robock's book *Why Industry Moves South* that have described the migration of enterprises from the North of the US and linked it to cheaper labor and smaller activity of labor unions [@McLaughlin1949].
+Various aspects of company migration have been studied, including factors (reasons) [@Wu2007], directions (places) [@Rossi2019], and taxation effects [@Kubicova2016].
 
-In Russia, there are few papers about migration of companies.
+In Russia, there are few papers about the migration of companies.
 Existing works focus on relocation as an instrument of tax optimization [@Adigamova2011_en; @Shcherbakova2016_en], skipping the geographical overview and spatial aspects of the phenomenon.
@@
 This study describes and visualizes the migration of Russian commercial legal companies in 2016–23.
-The choice of particular area of economy, that is, legal service, is subjective.
+The choice of a particular area of economy, that is, legal service, is subjective.
 
-**Data.** A geo-referenced dataset of Russian small and medium-sized businesses (SMBs) with main activity code equal to 69.10 (activity in the area of law) was generated from the open dumps of SMB registry[^1] with an open source Python CLI application *ru-smb-companies*[^2] developer by the author of this paper.
+**Data.** A geo-referenced dataset of Russian small and medium-sized businesses (SMBs) with a main activity code equal to 69.10 (activity in the area of law) was generated from the open dumps of SMB registry[^1] with an open source Python CLI application *ru-smb-companies*[^2] developed by the author of this paper.
 The dataset covers the period from August 2016 to November 2023.
-According to [@Moiseeva2016_en], by August 2016 in Russia there have been 47000 commercial legal companies, and the dataset contains about `r round(companies_count_08_2016, -3)` companies with the same date.
+According to [@Moiseeva2016_en], by August 2016 in Russia, there were 47000 commercial legal companies, and the dataset contains about `r round(companies_count_08_2016, -3)` companies with the same date.
 Some legal companies are missing, e.
@@
 
-**Methods.** Company migration is seen as a change in address of registration [@Shcherbakova2016_en].
+**Methods.** Company migration is seen as a change in the address of registration [@Shcherbakova2016_en].
 It is narrow and does not account for other ways of relocation (e.
@@
 mergers or reorganization).
-Also, it is formal, because the change of nominal address does not necessarily reflects the change of actual place of doing business.
+Also, it is formal, because the change of nominal address does not necessarily reflect the change of actual place of doing business.
 However, it looks like the best choice for the data available.
-Descriptive statistics, statistical tests, data visualization and mapping are used.
+Descriptive statistics, statistical tests, data visualization, and mapping are used.
 Technically, the analysis was performed with R programming language (including additional packages).
-The source code for the analysis is stored in public repository[^3].
+The source code for the analysis is stored in a public repository[^3].
 
@@
 
-A **distribution of companies by the number of migrations** is shown on *Fig.
+A **distribution of companies by the number of migrations** is shown in *Fig.
 1*.
@@
 Y axis is log10-scaled.
-The counts are shown above the bars, and percetrages are below the top border", fig.height=2.5}
+The counts are shown above the bars, and percentages are below the top border", fig.height=2.5}
 mig_count_distr_plot
@@
 
-A **relationship between the migration decision and the size of company** is depicted on *Fig.
+A **relationship between the migration decision and the size of the company** is depicted in *Fig.
 2*.
-Migration among small firms is more frequent in comparison to micro-businesses: according to Fisher's exact test, `r ifelse(categories_test$p.value < 0.001, "p < 0.001", paste("p = ", categories_test$p.value))`, and odds ratio is `r paste0("1 to ", round(1 / categories_test$estimate, 0))`.
+Migration among small firms is more frequent in comparison to micro-businesses: according to Fisher's exact test, `r ifelse(categories_test$p.value < 0.001, "p < 0.001", paste("p = ", categories_test$p.value))`, and the odds ratio is `r paste0("1 to ", round(1 / categories_test$estimate, 0))`.
 
@@
 
-A **distribution of migration distance** is shown on *Fig.
+A **distribution of migration distance** is shown in *Fig.
 3*.
-The majority of businesses moves to small distance (10–50 km), but there is a significant group of long-migrating firms (the distance is about 1000 km).
+The majority of businesses move to small distances (10–50 km), but there is a significant group of long-migrating firms (the distance is about 1000 km).
 
@@
 4*).
-Most of the regions have net migration rate close to zero.
+Most of the regions have a net migration rate close to zero.
 However, there are some regions with substantial growth (Moscow, St.
@@
 
-**The most popular migration paths** (by the count of migrated companies) with summed up revenues and employee counts are listed in the **Table**.
-The first two paths represent the relocation from two largest cities of Russia to the surrounding regions.
-Net change in the number of businesses is relatively small.
-An increase in the number of registered businesses is not necessarily associated with the increase in income or labor force, and vise versa.
-Positive flow of companies with negative flow of revenue or labor probably means that arriving businesses are larger and more profitable than leaving.
+**The most popular migration paths** (by the count of migrated companies) with summed revenues and employee counts are listed in the **Table**.
+The first two paths represent the relocation from the two largest cities of Russia to the surrounding regions.
+The net change in the number of businesses is relatively small.
+An increase in the number of registered businesses is not necessarily associated with the increase in income or labor force, and vice versa.
+A positive flow of companies with a negative flow of revenue or labor probably means that arriving businesses are larger and more profitable than those leaving.
 
@@
 
-**Conclusion.** Migration of legal companies in Russia is a rare phenomenon: only about 2.5% of all businesses in this activity area have changed their nominal location in the period of 2016–23.
+**Conclusion.** Migration of legal companies in Russia is a rare phenomenon: only about 2.5% of all businesses in this activity area have changed their nominal location in 2016–23.
 Small companies, in comparison to micro-businesses, are more likely to relocate: the chances are 4 times higher.
@@
 Moscow and St.
-Petersburg regions are at the top by net migration rate estimated by the count of relocated businesses, while Moscow have lost the highest number of companies.
+Petersburg regions are at the top by the net migration rate estimated by the count of relocated businesses, while Moscow has lost the highest number of companies.
 However, it looks like that while losing small and inefficient companies, Moscow attracts more profitable businesses.

### 2024-02-13 bel-geo-congress-2024/Conference-paper.Rmd
Proofreading, folder rename

--- 2024-02-12 maksakov-readings-conf-2024/Conference-paper.Rmd
@@
 bibliography: refs.bib
-csl: "gost-r-7-0-5-2008-numeric-alphabetical-en.csl"
+csl: biblio-style.csl
 ---
@@
 The migration of Russian commercial legal companies in 2016–23 is described and analyzed.
-Distributions by migration count and distance are estimated and visualized, the relationship between company size and migration decision is statistically tested, net migration rates by region are mapped, and top migration paths are listed.
+Distributions by migration count and distance are estimated and visualized, the relationship between company size and migration decision is statistically tested, net migration rate by region is mapped, and top migration paths are listed.
 It is found that migrations are rare, mostly short-distance (10–50 km).
@@
 
-***Keywords:*** companies migration, small and medium-sized businesses, legal companies, legal service, service geography, data analysis, open data, Federal Tax Service.
+***Keywords:*** company migration, small and medium-sized businesses, legal companies, legal service, service geography, data analysis, open data, Federal Tax Service.
 
@@
 McLaughlin and S.
-Robock's book *Why Industry Moves South* that have described the migration of enterprises from the North of the US and linked it to cheaper labor and smaller activity of labor unions [@McLaughlin1949].
+Robock's book *Why Industry Moves South* that has described the migration of enterprises from the North of the US and linked it to cheaper labor and weaker activity of labor unions [@McLaughlin1949].
 Various aspects of company migration have been studied, including factors (reasons) [@Wu2007], directions (places) [@Rossi2019], and taxation effects [@Kubicova2016].
@@
 In Russia, there are few papers about the migration of companies.
-Existing works focus on relocation as an instrument of tax optimization [@Adigamova2011_en; @Shcherbakova2016_en], skipping the geographical overview and spatial aspects of the phenomenon.
-So, a spatial analysis of business relocation in Russia is important.
+Existing works focus on relocation as an instrument of tax optimization [@Adigamova2011_en; @Shcherbakova2016_en] and do not pay much attention to the geographical overview and spatial aspects of the phenomenon.
+So, a spatial analysis of business relocation in Russia is needed.
 This study describes and visualizes the migration of Russian commercial legal companies in 2016–23.
@@
 The dataset covers the period from August 2016 to November 2023.
-According to [@Moiseeva2016_en], by August 2016 in Russia, there were 47000 commercial legal companies, and the dataset contains about `r round(companies_count_08_2016, -3)` companies with the same date.
+According to [@Moiseeva2016_en], by August 2016 in Russia, there were 47000 commercial legal companies, and the dataset contains about `r format(round(companies_count_08_2016, -3), scientific = FALSE)` companies with the same date.
 Some legal companies are missing, e.
 g.
-large ones that exceed the limit on the revenue or number of employees for SMBs, but 3/4 are present, so the dataset looks reliable and relevant.
+larger ones that exceed the limit on the revenue or number of employees for SMBs, but 3/4 are present, so the dataset looks reliable and relevant.
 
@@
 
-**Methods.** Company migration is seen as a change in the address of registration [@Shcherbakova2016_en].
-It is narrow and does not account for other ways of relocation (e.
+**Methods.** Company migration here is seen as a change in the address of registration, like in [@Shcherbakova2016_en].
+This definition is narrow and does not account for other ways of relocation (e.
 g.
@@
 
-[^3]: https://github.com/PavelSyomin/tax-service-opendata-paper/tree/main/maksakov-readings-conf-2024
+[^3]: https://github.com/PavelSyomin/tax-service-opendata-paper/tree/main/bel-geo-congress-2024
 
@@
 2*.
-Migration among small firms is more frequent in comparison to micro-businesses: according to Fisher's exact test, `r ifelse(categories_test$p.value < 0.001, "p < 0.001", paste("p = ", categories_test$p.value))`, and the odds ratio is `r paste0("1 to ", round(1 / categories_test$estimate, 0))`.
+Migration among small firms is more frequent in comparison to micro-businesses: according to Fisher's exact test, `r ifelse(categories_test$p.value < 0.001, "*p* < 0.001", paste("*p* = ", categories_test$p.value))`, and the odds ratio is approximately `r paste0("1 to ", round(1 / categories_test$estimate, 0))`.
 
@@
 3*.
-The majority of businesses move to small distances (10–50 km), but there is a significant group of long-migrating firms (the distance is about 1000 km).
+The majority of businesses move to small distance (10–50 km), but there is a significant group of long-migrating firms (the distance is about 1000 km).
 
@@
 
-**The most popular migration paths** (by the count of migrated companies) with summed revenues and employee counts are listed in the **Table**.
+**The most popular migration paths** (by the count of migrated companies) with summed revenue and employee count are listed in the **table**.
 The first two paths represent the relocation from the two largest cities of Russia to the surrounding regions.
@@
 **Conclusion.** Migration of legal companies in Russia is a rare phenomenon: only about 2.5% of all businesses in this activity area have changed their nominal location in 2016–23.
-Small companies, in comparison to micro-businesses, are more likely to relocate: the chances are 4 times higher.
+Small companies, in comparison to micro-businesses, are more likely to relocate: the probability is 4 times higher.
 Short-distance (10–50 km) migrations are more common than long-distance (~1000 km) ones.

### 2024-02-13 bel-geo-congress-2024/Paper.Rmd
Proofreading, folder rename

---
title: "Пространственный анализ миграций юридических компаний в России в 2016–2023 гг. по открытым данным налоговой службы"
author: П. О. Сёмин
output: 
  word_document:
    reference_docx: reference.docx
bibliography: refs.bib
csl: "gost-r-7-0-5-2008-numeric-alphabetical.csl"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE, 
  message = FALSE, 
  echo = FALSE, 
  fig.width = 6.4, 
  fig.height = 5,
  fig.id = FALSE,
  dpi = 300)
```

```{r imports, include=FALSE, cache=TRUE}
library(knitr)
source("Script.R")
```

<div custom-style="Author">

Пермский государственный национальный исследовательский университет, г. Пермь
</div>

<div custom-style="Abstract">

*Аннотация:*

*Ключевые слова:* миграция компаний, малое и среднее предпринимательство, юридические компании, юридические услуги, география услуг, анализ данных, открытые данные, ФНС России

</div>

# Введение

Миграция, или релокация, компаний — важное экономико-географическое явление, играющее большую роль в экономическом развитии стран, регионов, территорий. Оно связано с мобильностью труда и капитала, инвестиционной привлекательностью, балансом выгод и издержек, агломерационными эффектами, глобализацией, экономическим ростом, налоговой оптимизацией и другими факторами, обстоятельствами, причинами, процессами и закономерностями. Исследование миграции компаний восходит к середине 20 века: одной из первых научных работ на эту тему считается книга *Why industry moves South* Г. МакЛафлина и С. Робока, в которой описывается перемещение промышленных производств с севера на юг США, вызванное особенностями рынка труда — меньшими зарплатами и активностью просфоюзов [@McLaughlin1949]. Релокация фирм — сложное явление, которое может рассматриваться не только в пространственном измерении, но и с точки зрения комплексного процесса принятия решений [@Carrincazeaux2015], восприниматься не только в исследовательском ключе, но и как прикладная задача [@Shortland1987]. Большой пласт исследований на протяжении десятилетий посвящён вопросу о причинах (факторах) миграции компаний [@Nakosteen1987; @Wu2007; @Rasel2019; @Pan2020] и о направлениях миграции [@Rupasingha2018; @Rossi2019; @Cybulska2020], а также о связи причин и направлений. Отдельные специалисты фиксируют внимание на более редких аспектах: дискурс о релокации [@Piotti2007], миграция в венчурные кластеры [@Plant2007]. Возникает и вопрос о неблагоприятных практических последствиях и мерах по их предотвращению, например, на налоговых потерях и желательности введения «налога на выход» [@Kubicova2016].

В России статьи о миграции компаний редки. В основном авторы немногочисленных имеющихся работ фокусируются на «негативной» стороне вопроса, анализируя миграцию налогоплательщиков-организаций как способ налоговой оптимизации [@Adigamova2011; @Romanovskaya2015; @Shcherbakova2016; @Sinenko2021; ]. Есть также исследование о релокации как факторе инвестиционной активности применительно к «компаниям-единорогам» [@Kutsenko2022]. Простой географический анализ миграций фирм при этом либо упускается из виду, либо имеет промежуточное значение на пути к более серьёзным выводам. Тем не менее, он тоже может быть полезен, поскольку понимание масштаба явления и его базовых географических характеристик необходимо для формирования более адекватной картины мира и выдвижения гипотез. В данном докладе проводится именно такой анализ по отношению к российским юридическим фирмам. Выбор вида деятельности анализируемых организаций объясняется субъективными интересами автора исследования.

# Данные и методы

**Данные.** Для анализа использовался полный дезагрегированный геопривязанный набор данных о субъектах малого и среднего предпринимательства (МСП) в России, ведущих деятельность в области права. Этот набор данных был подготовлен на базе открытых выгрузок [^1] из реестра МСП Федеральной налоговой службы. Для создания набора данных автор самостоятельно разработал консольное приложение *ru-smb-companies* [^2] на языке программирования Python, которое позволяет получить подобный набор данных для любой сферы деятельности, не ограничиваясь юридическими фирмами. Набор данных охватывает период с августа 2016 г. по ноябрь 2023 г. Он содержит сведения о `r companies_count` организации, в частности: наименование, адреса регистрации с точностью до населённого пункта и его географических координат, доходы, расходы и число работников за каждый год с 2019. Очевидно, что набор данных включает информацию не обо всех юридических фирмах в в России: например, в него не попали те из них, которые не относятся к субъектам МСП из-за превышения лимита по доходам или числу работников, а также адвокатские образования, которые не считаются коммерческими организациями и поэтому тоже не вносятся в реестр МСП. Несмотря на эти ограничения, набор данных является наиболее полным из общедоступных и с высокой вероятностью географически репрезентативен. Так, согласно анализу рынка юридических услуг, проведённому в 2016 г. специалистами Института проблем правоприменения [@Moiseeva2016], в России на конец июля 2016 г. действовало 47 тысяч коммерческих юридических фирм, в то время как подсчёты на базе набора данных, используемого в текущем исследовании, дают оценку их количества в `r `round(companies_count_08_2016, 0)` тысячи. Следовательно, охвачены три четверти юридических фирм, и этого, вероятно, достаточно для надёжных выводов.

[^1]: https://www.nalog.gov.ru/opendata/7707329152-rsmp/
[^2]: https://github.com/PavelSyomin/russian-smb-companies

**Методы.** Миграция компании в данном исследовании понимается формально: как изменение адреса регистрации юридического лица. Такой подход не позволяет отследить все способы изменения местонахождения компании (в частности, «за кадром» остаются варианты реорганизации или слияния, при которых та же по сути фирма меняет местонахождение) и отождествляет номинальное и фактическое перемещение (хотя при изменении юридического местонахождения фактическая территория деятельности может остаться прежней), но максимально соответствует возможностям, которые даёт используемый набор данных. В пространственном отношении миграция отслеживается с точностью до региона (субъекта федерации) и населённого пункта, а во временном — с точностью до месяца. Анализ проводится с помощью методов описательной статистики, визуализации данных и картографирования. Он включает в себя следующие элементы (вопросы).

1. Распределение компаний по количеству миграций.
2. Связь между миграцией и размером компании.
3. Распределение миграций по расстоянию.
4. Распределение миграций по времени.
5. Распределение регионов по притоку (оттоку) компаний.
6. Распределение населённых пунктов по притоку (оттоку) компаний.
7. Соотношение внутрирегиональной миграции и межрегиональной миграции.
8. Пути межрегиональной миграции.
9. Сочетание централизации и децентрализации в межрегиональной миграции.
10. Пути внутрирегиональной миграции.
11. Сочетание централизации и децентрализации во внутрирегиональной миграции.

Для оценки интенсивности того или иного явления используется либо число компаний, либо число, суммарная выручка и суммарная численность работников компаний. Анализ проводился с помощью языка программирования R версии 4 с дополнительными пакетами. Исходный код доступен в репозитории статьи [^3].

[^3]: https://github.com/PavelSyomin/tax-service-opendata-paper/tree/main/maksakov-readings-conf-2024

# Результаты

**Распределение компаний по количеству миграций** показано в таблице 1 (здесь и далее см. приложения в конце статьи). Видно, что подавляющее большинство (97%) компаний за 2016–2023 гг. ни разу не мигрировали. Около 2.5% фирм поменяли местонахождение однократно, а многократные миграции крайне редки.

**Связь между миграцией и размером компаний** отражена в таблице 2. Хотя пропорции не слишком очевидны из-за дисбаланса категорий, малые предприятия мигрируют статистически значимо чаще, чем микропредприятия (по результатам точного теста Фишера `r ifelse(categories_test$p.value < 0.001, "p < 0.001", paste("p = ", categories_test$p.value))`, соотношение шансов `r paste0("1 к ", round(1 / categories_test$estimate, 0))`.

**Распределение миграций по расстоянию** показано на графике 1. Видно, что преобладают перемещения на небольшие расстояния, хотя есть очевидная вторая группа — миграции на расстояние около 1000 км.

**Распределение миграций по времени** показано на графиках 2 и 3. Закономерности за весь рассмотренный период отсутствуют: интенсивность миграций выглядит случайной. Тем не менее, видно, что компании чаще меняют своё местонахождение летом.

**Распределение регионов по притоку (оттоку) компаний** изображено на картограмме 1. Общая закономерность отсутствует, при этом заметно, что показатели противоречат друг другу: регионы, лидирующие по сальдо исходя из числа компаний, могут проигрывать при расчёте по выручке или численности работников.

**Распределение населённых пунктов по притоку (оттоку) компаний** показано на картограмме 2. Здесь также сложно выявить закономерности, хотя лидируют по сальдо, измеренному как число компаний, города-спутники Москвы. Опять же, показатели противоречивые: лидерство по численному сальдо не означает лидерство по выручке или числу работников.

**Соотношение внутрирегиональной миграции и межрегиональной миграции** показано в таблице 3. Видно, что между регионами перемещается примерно треть фирм, а остальные две трети — внутри региона. При этом в пересчёте на число работников и выручку межрегиональная миграция масштабнее, чем внутрирегиональная, из чего следует, что компания, поменявшая регион, в среднем «везёт» с собой больше денег и людей, чем компания, поменявшая населённый пункт внутри региона.

**Пути межрегиональной миграции** отражены в таблице 4. Абсолютные числа переместившихся компаний, как правило, невелики. Лидируют миграции из Москвы и Санкт-Петербурга в прилегающие к ним области. Нет чёткой связи между числом переехавших компаний и «перевезённой» выручкой или числом работников.

**Сочетание централизации и децентрализации в межрегиональной миграции** (таблица 5) преимущественно равномерное. По числу работников и компаний преобладает центробежная тенденция, а по доходам — центростремительная. Треть потоков не может быть описани ни как центростремительные, ни как центробежные.

**Пути внутрирегиональной миграции** (таблица 6) ещё более неоднозначны, чем пути межрегиональной миграции, поскольку абсолютные перемещения почти во всех случаях измеряются единицами. Нет и связи между тремя показателями: число фирм, доходы, число работников.

**Сочетание централизации и децентрализации во внутрирегиональной миграции** (таблица 7) почти такое же, как и в случае с межрегиональной миграцией: перемещения к центру выигрывают по числу фирм и людей, но уступают в доходах, при этом треть релокаций не могут быть описаны ни как центростремительные, ни как центробежные.

# Обсуждение

# Выводы

# Литература

<div id="refs"></div>

# Приложения

```{r table-1-mig-distr}
kable(
  migration_count_distribution,
  col.names = c("Количество миграций", "Количество компаний", "%"),
  caption = "Таблица 1. Распределение юридических компаний по количеству миграций")
```

```{r table-2-mig-cats}
kable(
  count_by_category,
  col.names = c("Категория фирмы", "Всего", "Из них мигрировавших"),
  caption = "Таблица 2. Распределение юридических компаний по категории и статусу миграции"
)
```

```{r table-3-inter-intra}
kable(
  inter_intra,
  col.names = c("Тип миграции", "Число", "Доходы", "Работники"),
  caption = "Таблица 3. Показатели межрегиональной и внутрирегиональной миграции юридических компаний: число фирм, суммарные доходы (млн рублей), суммарное число работников"
)
```

```{r table-4-reg-paths}
kable(
  rbind(head(regional_paths_combined, 5), tail(regional_paths_combined, 5)),
  col.names = c("Путь", "Количество", "Доходы, млн", "Работники"),
  caption = "Таблица 4. Маршруты миграции юридических компаний по субъектам России и их показатели: число фирм, суммарные доходы в миллионах рублей, суммарная численность работников. Поразаны первые и последние 5 маршрутов по числу фирм"
)
```

```{r table-5-reg-central}
kable(
  reg_central,
  col.names = c("Тип миграции", "Число", "Доходы", "Работники"),
  caption = "Таблица 5. Показатели центростремительных и центробежных межрегиональных миграций юридических компаний: число фирм, суммарные доходы в миллионах рублей, суммарная численность работников"
)
```

```{r table-6-settl-paths}
kable(
  rbind(head(settlement_paths_combined, 5), tail(settlement_paths_combined, 5)),
  col.names = c("Маршрут", "Количество", "Доходы, млн", "Работники"),
  caption = "Таблица 6. Маршруты миграции юридических компаний по населённым пунктам и их показатели: число фирм, суммарные доходы в миллионах рублей, суммарная численность работников. Поразаны первые и последние 5 маршрутов по числу фирм"
)
```

```{r table-7-settl-central}
kable(
  settl_central,
  col.names = c("Тип миграции", "Число", "Доходы", "Работники"),
  caption = "Таблица 7. Показатели центростремительных и центробежных внутрирегиональных миграций юридических компаний: число фирм, суммарные доходы в миллионах рублей, суммарная численность работников"
)
```

```{r plot-1-distance}
mig_distance_plot
```

```{r plot-2-timeline}
mig_timeline_plot
```

```{r plot-3-monthly}
mig_by_month_plot
```

```{r map-1-regions}

```

```{r map-2-settlements}

```

<div custom-style="Title">
Spatial Analysis of Migration of Legal Companies in Russia (2016–2023) Based on Tax Service Open Data
<div>

<div custom-style="Author">
P. O. Syomin

Perm State University, Perm
</div>

<div custom-style="Abstract">

*Abstract:*

*Keywords:* companies migration, small and medium-sized businesses, legal companies, legal service, service geography, data analysis, open data, FTS of Russia

</div>

### 2024-02-13 bel-geo-congress-2024/Script.R

@@ -46,7 +46,7 @@ mig_count_distr_plot <- migration_count_distribution %>%
   scale_y_log10(limits = c(NA, 1e6), breaks = 10^(1:5), labels = ~ .x) +
   labs(
     x = "Number of migrations by a company", 
-    y = "Number of companies (log10 scale)") +
+    y = "Number of companies (log10)") +
   theme_bw(base_size = 12, base_family = "Times New Roman") +
   theme(
     panel.grid.minor.x = element_blank(), 
@@ -147,8 +147,8 @@ mig_distance_plot <- migrations %>%
   ggplot(aes(x = distance)) +
   geom_freqpoly(bins = 50) +
   scale_x_continuous(trans = "log10", breaks = 10^(-1:3), labels = ~ .x) +
-  theme_bw(base_size = 14, base_family = "Times New Roman") +
-  labs(x = "Distance of migration, km", y = "Count of companies")
+  theme_bw(base_size = 12, base_family = "Times New Roman") +
+  labs(x = "Distance of migration, km (log10)", y = "Count of companies")
 
 # Date of migrations
 # Overall timeline

### 2024-02-15 legal-companies-migration/Paper.Rmd
Edit introduction and data sections, move plots and tables, rename biblio file and styles, add biblio items, add project file

--- 2024-02-13 bel-geo-congress-2024/Paper.Rmd
@@
 ---
-title: "Пространственный анализ миграций юридических компаний в России в 2016–2023 гг. по открытым данным налоговой службы"
+title: "Пространственный анализ миграции юридических компаний в России в 2016–2023 гг. по открытым данным налоговой службы"
 author: П. О. Сёмин
@@
     reference_docx: reference.docx
-bibliography: refs.bib
-csl: "gost-r-7-0-5-2008-numeric-alphabetical.csl"
+bibliography: biblio-ru.bib
+csl: "biblio-styles-ru.csl"
 ---
@@
 
-Миграция, или релокация, компаний — важное экономико-географическое явление, играющее большую роль в экономическом развитии стран, регионов, территорий.
-Оно связано с мобильностью труда и капитала, инвестиционной привлекательностью, балансом выгод и издержек, агломерационными эффектами, глобализацией, экономическим ростом, налоговой оптимизацией и другими факторами, обстоятельствами, причинами, процессами и закономерностями.
-Исследование миграции компаний восходит к середине 20 века: одной из первых научных работ на эту тему считается книга *Why industry moves South* Г.
+Миграция, или релокация, компаний (бизнесов) — важное экономико-географическое явление, играющее большую роль в экономическом развитии стран, регионов, территорий.
+Оно связано с мобильностью и стоимостью труда и капитала, инвестиционной привлекательностью, международной конкуренцией и торговлей, агломерационными эффектами, глобализацией, экономическим ростом, налоговой оптимизацией.
+Исследование миграции компаний восходит к середине 20 века: одной из первых научных работ на эту тему считается книга *Why Industry Moves South* Г.
 МакЛафлина и С.
-Робока, в которой описывается перемещение промышленных производств с севера на юг США, вызванное особенностями рынка труда — меньшими зарплатами и активностью просфоюзов [@McLaughlin1949].
-Релокация фирм — сложное явление, которое может рассматриваться не только в пространственном измерении, но и с точки зрения комплексного процесса принятия решений [@Carrincazeaux2015], восприниматься не только в исследовательском ключе, но и как прикладная задача [@Shortland1987].
-Большой пласт исследований на протяжении десятилетий посвящён вопросу о причинах (факторах) миграции компаний [@Nakosteen1987; @Wu2007; @Rasel2019; @Pan2020] и о направлениях миграции [@Rupasingha2018; @Rossi2019; @Cybulska2020], а также о связи причин и направлений.
-Отдельные специалисты фиксируют внимание на более редких аспектах: дискурс о релокации [@Piotti2007], миграция в венчурные кластеры [@Plant2007].
-Возникает и вопрос о неблагоприятных практических последствиях и мерах по их предотвращению, например, на налоговых потерях и желательности введения «налога на выход» [@Kubicova2016].
+Робока.
+В ней описывается перемещение промышленных производств с севера на юг США, вызванное особенностями рынка труда: меньшими зарплатами и меньшей активностью просфоюзов на юге [@McLaughlin1949].
 
-В России статьи о миграции компаний редки.
-В основном авторы немногочисленных имеющихся работ фокусируются на «негативной» стороне вопроса, анализируя миграцию налогоплательщиков-организаций как способ налоговой оптимизации [@Adigamova2011; @Romanovskaya2015; @Shcherbakova2016; @Sinenko2021; ].
-Есть также исследование о релокации как факторе инвестиционной активности применительно к «компаниям-единорогам» [@Kutsenko2022].
-Простой географический анализ миграций фирм при этом либо упускается из виду, либо имеет промежуточное значение на пути к более серьёзным выводам.
-Тем не менее, он тоже может быть полезен, поскольку понимание масштаба явления и его базовых географических характеристик необходимо для формирования более адекватной картины мира и выдвижения гипотез.
-В данном докладе проводится именно такой анализ по отношению к российским юридическим фирмам.
-Выбор вида деятельности анализируемых организаций объясняется субъективными интересами автора исследования.
+Релокация фирм — это сложный процесс принятия решений [@Carrincazeaux2015], и география является только одним из его аспектов.
+На протяжении десятилетий учёные изучают такие вопросы, как факторы (причины) миграции, пути и направления переезда, способы и виды релокации, её последствия на макроэкономическом (например, для налогообложения) и микроэкономическом (в частности, для доходов предприятия) уровне.
+Так, отмечается, что на миграцию влияют налоги и меры поддержки [@Wu2007; @Pan2020], размер компании, издержки на производство и уровень агломерации [@Rasel2019], а результатом перемещения часто оказывается рост продаж и числа работников у фирмы [@Nakosteen1987].
+Компании обычно мигрируют на небольшое расстояние и предпочитают урбанизированные районы [@Rossi2019; @Cybulska2020], причём похожие на прежнее место нахождения во многих аспектах [@Rupasingha2018].
+Заметную роль играют венчурные кластеры, которые «притягивают» фирмы своим финансированием [@Plant2007], и миграция основателей стартапов [@Kutsenko2022].
+Отдельные исследования посвящены более гуманитарным аспектам проблемы, таким как дискурс о релокации [@Piotti2007].
+Существенный пласт научных работ посвящён налоговым причинам и последствиям релокации [@Burwitz2006; @Chand2013; @Kubicova2016].
+Миграция бизнеса может рассматриваться не только с академической точки зрения, но и как прикладная задача [@Shortland1987].
 
-# Данные и методы
+Исследований, посвящённых миграции компаний в России, немного.
+Они касаются в основном налоговых вопросов: фиктивной миграции, её влиянии на собираемость налогов и налоговую задолженность, мер борьбы с ней [@Adigamova2011; @Shcherbakova2016], использовании миграции в особые экономические зоны в качестве инструмента налоговой оптимизации [@Sinenko2021].
+В таких работах косвенно затрагиваются пространственные аспекты (например, называются регионы, в которые перемещается большое количество фирм), но основной целью исследования они не являются.
+Географический анализ миграции компаний в России позволит дать ответы на довольно простые, но важные, базовые вопросы об этом явлении: сколько фирм перемещается, как часто и в какое время это происходит, в каком направлении, на какое расстояние, как сочетается с популярными представлении о центростремительных тенденциях.
+Именно этому посвящена данная статья.
 
-**Данные.** Для анализа использовался полный дезагрегированный геопривязанный набор данных о субъектах малого и среднего предпринимательства (МСП) в России, ведущих деятельность в области права.
-Этот набор данных был подготовлен на базе открытых выгрузок [^1] из реестра МСП Федеральной налоговой службы.
-Для создания набора данных автор самостоятельно разработал консольное приложение *ru-smb-companies* [^2] на языке программирования Python, которое позволяет получить подобный набор данных для любой сферы деятельности, не ограничиваясь юридическими фирмами.
+Вместо того чтобы рассматривать все компании в России, данное исследование фокусируется только на отдельной категории фирм — юридических.
+Такой выбор отчасти субъективен и объясняется личными академическими интересами автора.
+Тем не менее, у него есть и объективные причины.
+Во-первых, само понятие миграции фирм в большей степени адаптировано к небольшим организациям, расположенным в одном месте [@Plant2007], и юридические фирмы, как правило, именно такие.
+Во-вторых, сфера юридических услуг сравнительно слабо связано с набором «традиционных» географических факторов размещения, «привязывающих» фирмы к их первоначальному местонахождению, таких как минеральные, почвенные, лесные, климатические, энергетические ресурсы, и это повышает шанс найти закономерности, отличающиеся от классических.
+В-третьих, юридические услуги важны для бизнеса, и перемещение фирм, оказывающих их, косвенно связано с экономической активностью в целом.
+В-четвёртых, выбор одной конкретной группы компаний упрощает анализ, поскольку уменьшает количество исходных данных, которых, как будет показано ниже, в исходном виде довольно много.
+
+# Объекты и методы исследования
+
+**Данные.** Для анализа использовался дезагрегированный геопривязанный набор данных о субъектах малого и среднего предпринимательства (МСП) в России, ведущих деятельность в области права (код основного вида деятельности по ОКВЭД 69.10).
+Он был подготовлен на базе выгрузок из реестра МСП, размещённых на сайте Федеральной налоговой службы в формате открытых данных[^1].
+Объём исходных выгрузок составляет около 200 Гб в сжатом виде и более 2 Тб в несжатом, поэтому для их обработки автор самостоятельно разработал консольное приложение *ru-smb-companies*[^2] на языке программирования Python, которое обрабатывает исходную информацию и формирует на её базе компактную таблицу с фирмами, имеющими необходимый код ОКВЭД.
 Набор данных охватывает период с августа 2016 г.
 по ноябрь 2023 г.
-Он содержит сведения о `r companies_count` организации, в частности: наименование, адреса регистрации с точностью до населённого пункта и его географических координат, доходы, расходы и число работников за каждый год с 2019.
-Очевидно, что набор данных включает информацию не обо всех юридических фирмах в в России: например, в него не попали те из них, которые не относятся к субъектам МСП из-за превышения лимита по доходам или числу работников, а также адвокатские образования, которые не считаются коммерческими организациями и поэтому тоже не вносятся в реестр МСП.
+Он содержит сведения о `r companies_count` организации.
+Таблица имеет, в частности, такие поля (колонки): наименование компании, её категория (малое, среднее или микропредприятие), регион, административный район, населённый пункт регистрации, его географические координаты, доходы, расходы и число работников организации за каждый год с 2019.
+
+Набор данных не является выборочным, но очевидно, что он включает информацию не обо всех российских юридических фирмах.
+Так, в него не попали те организации, которые не относятся к субъектам МСП из-за превышения нормативного ограничения по доходам или числу работников.
+Кроме того, в нём не значатся адвокатские образования, которые не считаются коммерческими организациями и поэтому тоже не вносятся в реестр МСП.
 Несмотря на эти ограничения, набор данных является наиболее полным из общедоступных и с высокой вероятностью географически репрезентативен.
-Так, согласно анализу рынка юридических услуг, проведённому в 2016 г.
+Согласно анализу рынка юридических услуг, проведённому в 2016 г.
 специалистами Института проблем правоприменения [@Moiseeva2016], в России на конец июля 2016 г.
-действовало 47 тысяч коммерческих юридических фирм, в то время как подсчёты на базе набора данных, используемого в текущем исследовании, дают оценку их количества в `r `round(companies_count_08_2016, 0)` тысячи.
-Следовательно, охвачены три четверти юридических фирм, и этого, вероятно, достаточно для надёжных выводов.
+действовало 47 тысяч коммерческих юридических фирм, а подсчёты на базе набора данных, используемого в текущем исследовании, дают оценку их количества в `r `format(round(companies_count_08_2016, -3))` на 10 августа 2016 г.
+Иначе говоря, в наборе данных есть сведения о примерно трёх четвертях всех необходимых организаций.
+Более полный список можно было бы получить, например, имея доступ к Единому государственному реестра юридических лиц (ЕГРЮЛ), но, в отличие от реестра МСП, он в полном виде не публикуется.
 
@@
 
-# Результаты
+# Результаты и их обсуждение
 
@@
 
+```{r table-1-mig-distr}
+kable(
+  migration_count_distribution,
+  col.names = c("Количество миграций", "Количество компаний", "%"),
+  caption = "Таблица 1. Распределение юридических компаний по количеству миграций")
+```
+
 **Связь между миграцией и размером компаний** отражена в таблице 2.
@@
 
+```{r table-2-mig-cats}
+kable(
+  count_by_category,
+  col.names = c("Категория фирмы", "Всего", "Из них мигрировавших"),
+  caption = "Таблица 2. Распределение юридических компаний по категории и статусу миграции"
+)
+```
+
 **Распределение миграций по расстоянию** показано на графике 1. Видно, что преобладают перемещения на небольшие расстояния, хотя есть очевидная вторая группа — миграции на расстояние около 1000 км.
+
+```{r plot-1-distance}
+mig_distance_plot
+```
 
@@
 
+```{r plot-2-timeline}
+mig_timeline_plot
+```
+
+```{r plot-3-monthly}
+mig_by_month_plot
+```
+
 **Распределение регионов по притоку (оттоку) компаний** изображено на картограмме 1.
 Общая закономерность отсутствует, при этом заметно, что показатели противоречат друг другу: регионы, лидирующие по сальдо исходя из числа компаний, могут проигрывать при расчёте по выручке или численности работников.
+
+```{r fig-4-mig-by-region-plot, fig.cap="*Fig. 4.* Net migration rate of legal companies by regions", fig.height=3}
+mig_by_region_plot
+```
 
@@
 
-**Пути межрегиональной миграции** отражены в таблице 4.
-Абсолютные числа переместившихся компаний, как правило, невелики.
-Лидируют миграции из Москвы и Санкт-Петербурга в прилегающие к ним области.
-Нет чёткой связи между числом переехавших компаний и «перевезённой» выручкой или числом работников.
-
-**Сочетание централизации и децентрализации в межрегиональной миграции** (таблица 5) преимущественно равномерное.
-По числу работников и компаний преобладает центробежная тенденция, а по доходам — центростремительная.
-Треть потоков не может быть описани ни как центростремительные, ни как центробежные.
-
-**Пути внутрирегиональной миграции** (таблица 6) ещё более неоднозначны, чем пути межрегиональной миграции, поскольку абсолютные перемещения почти во всех случаях измеряются единицами.
-Нет и связи между тремя показателями: число фирм, доходы, число работников.
-
-**Сочетание централизации и децентрализации во внутрирегиональной миграции** (таблица 7) почти такое же, как и в случае с межрегиональной миграцией: перемещения к центру выигрывают по числу фирм и людей, но уступают в доходах, при этом треть релокаций не могут быть описаны ни как центростремительные, ни как центробежные.
-
-# Обсуждение
-
-# Выводы
-
-# Литература
-
-<div id="refs"></div>
-
-# Приложения
-
-```{r table-1-mig-distr}
-kable(
-  migration_count_distribution,
-  col.names = c("Количество миграций", "Количество компаний", "%"),
-  caption = "Таблица 1. Распределение юридических компаний по количеству миграций")
-```
-
-```{r table-2-mig-cats}
-kable(
-  count_by_category,
-  col.names = c("Категория фирмы", "Всего", "Из них мигрировавших"),
-  caption = "Таблица 2. Распределение юридических компаний по категории и статусу миграции"
-)
-```
-
 ```{r table-3-inter-intra}
@@
 ```
+
+**Пути межрегиональной миграции** отражены в таблице 4.
+Абсолютные числа переместившихся компаний, как правило, невелики.
+Лидируют миграции из Москвы и Санкт-Петербурга в прилегающие к ним области.
+Нет чёткой связи между числом переехавших компаний и «перевезённой» выручкой или числом работников.
 
@@
 
+**Сочетание централизации и децентрализации в межрегиональной миграции** (таблица 5) преимущественно равномерное.
+По числу работников и компаний преобладает центробежная тенденция, а по доходам — центростремительная.
+Треть потоков не может быть описани ни как центростремительные, ни как центробежные.
+
 ```{r table-5-reg-central}
@@
 ```
+
+**Пути внутрирегиональной миграции** (таблица 6) ещё более неоднозначны, чем пути межрегиональной миграции, поскольку абсолютные перемещения почти во всех случаях измеряются единицами.
+Нет и связи между тремя показателями: число фирм, доходы, число работников.
 
@@
 
+**Сочетание централизации и децентрализации во внутрирегиональной миграции** (таблица 7) почти такое же, как и в случае с межрегиональной миграцией: перемещения к центру выигрывают по числу фирм и людей, но уступают в доходах, при этом треть релокаций не могут быть описаны ни как центростремительные, ни как центробежные.
+
 ```{r table-7-settl-central}
@@
 
-```{r plot-1-distance}
-mig_distance_plot
-```
+# Выводы
 
-```{r plot-2-timeline}
-mig_timeline_plot
-```
+# Список литературы
 
-```{r plot-3-monthly}
-mig_by_month_plot
-```
-
-```{r map-1-regions}
-
-```
-
-```{r map-2-settlements}
-
-```
+<div id="refs"></div>
 

### 2024-02-16 legal-companies-migration/Paper.Rmd
Edit methods section

@@
 Он содержит сведения о `r companies_count` организации.
-Таблица имеет, в частности, такие поля (колонки): наименование компании, её категория (малое, среднее или микропредприятие), регион, административный район, населённый пункт регистрации, его географические координаты, доходы, расходы и число работников организации за каждый год с 2019.
+Таблица имеет, в частности, такие поля (колонки): наименование компании, её категория (малое, среднее или микропредприятие), ИНН, регион, административный район, населённый пункт регистрации, его географические координаты, доходы, расходы и число работников организации за каждый год с 2019.
 
@@
 
-**Методы.** Миграция компании в данном исследовании понимается формально: как изменение адреса регистрации юридического лица.
-Такой подход не позволяет отследить все способы изменения местонахождения компании (в частности, «за кадром» остаются варианты реорганизации или слияния, при которых та же по сути фирма меняет местонахождение) и отождествляет номинальное и фактическое перемещение (хотя при изменении юридического местонахождения фактическая территория деятельности может остаться прежней), но максимально соответствует возможностям, которые даёт используемый набор данных.
-В пространственном отношении миграция отслеживается с точностью до региона (субъекта федерации) и населённого пункта, а во временном — с точностью до месяца.
-Анализ проводится с помощью методов описательной статистики, визуализации данных и картографирования.
-Он включает в себя следующие элементы (вопросы).
+**Методы.** Миграция компании в данном исследовании понимается как изменение адреса регистрации юридического лица.
+Похожее определение принято, например, в работах Н.
+В.
+Щербаковой [@Shcherbakova2016] и Р.
+Планта [@Plant2007].
+Оно не позволяет отследить все способы изменения местонахождения компании: «за кадром» остаются варианты реорганизации или слияния, при которых та же по сути, но другая по регистрационному номеру фирма меняет местонахождение и которые также актуальны для России [@Adigamova2011; @Shcherbakova2016].
+Кроме того, оно отождествляет номинальное и фактическое перемещение, хотя при изменении юридического адреса фактическое место ведения бизнеса может остаться прежним.
+Тем не менее, оно наилучшим образом подходит к используемому набору данных, который содержит историю адресов регистрации организаций, даты их изменения и ИНН юридических лиц: если номер налогоплательщика не поменялся, а адрес стал другим, то можно считать, что компания мигрировала.
+В пространственном отношении миграция отслеживается с точностью до региона (субъекта федерации) и населённого пункта, а во временном — с точностью до месяца и года.
 
-1. Распределение компаний по количеству миграций.
-2. Связь между миграцией и размером компании.
-3. Распределение миграций по расстоянию.
-4. Распределение миграций по времени.
-5. Распределение регионов по притоку (оттоку) компаний.
-6. Распределение населённых пунктов по притоку (оттоку) компаний.
-7. Соотношение внутрирегиональной миграции и межрегиональной миграции.
-8. Пути межрегиональной миграции.
-9. Сочетание централизации и децентрализации в межрегиональной миграции.
-10. Пути внутрирегиональной миграции.
-11. Сочетание централизации и децентрализации во внутрирегиональной миграции.
+Анализ проводится с помощью методов описательной статистики, статистических тестов, визуализации данных и картографирования.
+Он включает в себя 11 вопросов, некоторые из которых взаимосвязаны.
+Каждому вопросу отводится свой небольшой подраздел в «Результатах и их обсуждении».
+Для начала рассматривается распределение компаний по количеству миграций (1) и связь между миграцией и размером компании (2).
+Далее анализируется распределение миграций по расстоянию (3) и по времени (4).
+Следующий шаг — картографирование распределения регионов (5) и населённых пунктов (6) по сальдо миграции.
+После этого изучается соотношение внутри- и межрегиональной миграци (7).
+Далее отдельно рассматриваются внутрирегиональная миграция: определяются пути миграции и их количественные характеристики (8) и анализируется сочетание централизации и децентрализации (9).
+Аналогичные вопросы поднимаются применительно к межрегиональной миграции (10 и 11).
 
-Для оценки интенсивности того или иного явления используется либо число компаний, либо число, суммарная выручка и суммарная численность работников компаний.
-Анализ проводился с помощью языка программирования R версии 4 с дополнительными пакетами.
-Исходный код доступен в репозитории статьи [^3].
+В качестве количественных характеристик используется одна или несколько из следующих трёх: число фирм, их суммарная выручка и суммарная численность работников.
+Под регионом условно понимается субъект России.
+Под централизацией в межрегиональном разрезе понимается перемещение в Москву из любого региона, в Московскую область из любого региона, кроме Москвы, в Санкт-Петербург из Ленинградской области.
+Централизация внутри региона рассматривается как перемещение в административный центр субъекта федерации.
+С технической точки зрения исследование реализовано с помощью языка программирования R с дополнительными пакетами.
+Исходный код доступен в репозитории статьи[^3].
 
-[^3]: https://github.com/PavelSyomin/tax-service-opendata-paper/tree/main/maksakov-readings-conf-2024
+[^3]: https://github.com/PavelSyomin/ru-smb-companies-papers/tree/main/legal-companies-migration
 

### 2024-02-16 legal-companies-migration/Paper.Rmd
Edit styles, edit results section, add settlements plot

@@
 ---
+
 title: "Пространственный анализ миграции юридических компаний в России в 2016–2023 гг. по открытым данным налоговой службы"
@@
 
-**Распределение компаний по количеству миграций** показано в таблице 1 (здесь и далее см.
-приложения в конце статьи).
+**Распределение компаний по количеству миграций** показано в таблице 1.
 Видно, что подавляющее большинство (97%) компаний за 2016–2023 гг.
 ни разу не мигрировали.
-Около 2.5% фирм поменяли местонахождение однократно, а многократные миграции крайне редки.
+Около 2.5% фирм поменяли местонахождение однократно, а многократные релокации крайне редки.
+Сопоставить полученные значения с результатами других исследований сложно из-за отсутствия аналогичных работ.
+В Нидерландах в 1998–2003 гг.
+ежегодно меняли местонахождение около 5% компаний, что в пересчёте на несколько лет примерно в 10 раз больше, чем получилось в данном анализе [@Pellenbarg2005].
 
@@
   migration_count_distribution,
+  digits = 3,
   col.names = c("Количество миграций", "Количество компаний", "%"),
@@
 Хотя пропорции не слишком очевидны из-за дисбаланса категорий, малые предприятия мигрируют статистически значимо чаще, чем микропредприятия (по результатам точного теста Фишера `r ifelse(categories_test$p.value < 0.001, "p < 0.001", paste("p = ", categories_test$p.value))`, соотношение шансов `r paste0("1 к ", round(1 / categories_test$estimate, 0))`.
+Делать выводы в отношении средних предприятий сложно, поскольку их в наборе данных мало.
 
@@
 
-**Распределение миграций по расстоянию** показано на графике 1. Видно, что преобладают перемещения на небольшие расстояния, хотя есть очевидная вторая группа — миграции на расстояние около 1000 км.
+**Распределение миграций по расстоянию** показано на графике 1.
+Видно, что преобладают перемещения на небольшие расстояния (10–50 км), хотя есть вторая заметная группа — миграции на расстояние около 1000 км.
+Более высокая частота релокаций на небольшое расстояние отмечалась и в других исследованиях [@Pellenbarg2005; @Rossi2019].
 
@@
 **Распределение миграций по времени** показано на графиках 2 и 3.
-Закономерности за весь рассмотренный период отсутствуют: интенсивность миграций выглядит случайной.
+Закономерности за весь рассмотренный период отсутствуют: интенсивность миграции выглядит случайной.
 Тем не менее, видно, что компании чаще меняют своё местонахождение летом.
+Возможно, это связано с тем, что летом спрос на юридические услуги и возможность их оказания уменьшаются из-за отпусков у клиентов и служащих государственных структур, и возникает благоприятное окно времени для переезда.
 

### 2024-02-16 legal-companies-migration/Script.R

@@ -6,6 +6,9 @@ library(tidyr)
 library(sf)
 library(units)
 
+# Base theme for plots
+theme <- theme_bw(base_size = 11, base_family = "Times New Roman")
+
 # Load the data
 data <- read_csv("../../ru-smb-companies/legal/smb.csv")
 revexp <- read_csv("../../ru-smb-companies/legal/revexp.csv")
@@ -197,10 +200,19 @@ mig_by_region_plot <- migration_by_region %>%
 
 migration_msk_spb <- migration_by_region %>% 
   filter(region %in% c("Москва", "Санкт-Петербург")) %>% 
-  rename(settlement = region)
+  rename(settlement = region) %>% 
+  select(-iso_code) %>% 
+  st_drop_geometry()
 
 # Migration by settlements
+settl_coords <- rbind(
+  select(distinct(data, settlement, .keep_all = TRUE), settlement, lat, lon),
+  select(distinct(data, region, .keep_all = TRUE), settlement = region, lat, lon)
+) %>% 
+  drop_na(lat, lon) %>% 
+  st_as_sf(coords = c("lon", "lat"), crs = "EPSG:4326")
 migration_by_settlement <- migrations %>% 
+  st_drop_geometry() %>% 
   select(settlement_from, settlement_to, revenue, empl) %>% 
   pivot_longer(settlement_from:settlement_to, names_to = "type", values_to = "settlement") %>% 
   drop_na(settlement) %>% 
@@ -212,7 +224,19 @@ migration_by_settlement <- migrations %>%
   select(settlement, var, change) %>% 
   pivot_wider(settlement, names_from = var, values_from = change) %>%
   arrange(-count) %>% 
-  rbind(migration_msk_spb)
+  rbind(migration_msk_spb) %>% 
+  left_join(settl_coords) %>% 
+  st_as_sf()
+migration_by_settlement_plot <- migration_by_settlement %>% 
+  filter(abs(count) > 1) %>% 
+  mutate(direction = count > 0, count = abs(count)) %>% 
+  ggplot() +
+  geom_sf(data = regions_geo, size = .5) +
+  geom_sf(aes(color = direction), size = 1) +
+  coord_sf(crs = ru_crs) +
+  scale_color_manual(name = "", labels = c("Отток", "Приток"), values = c("#d01c8b", "#4dac26")) +
+  theme_bw(base_size = 11, base_family = "Times New Roman")
+migration_by_settlement_plot
 
[…]

### 2024-02-20 legal-companies-migration/Paper.Rmd
Finish results and discussion, add conclusion, update author info, add abstract, change headings, add/edit table/figure captions

@@
 ---
-
-title: "Пространственный анализ миграции юридических компаний в России в 2016–2023 гг. по открытым данным налоговой службы"
-author: П. О. Сёмин
+title: "УДК 911.3"
+subtitle: "Пространственный анализ миграции юридических компаний в России в 2016–2023 гг. по открытым данным налоговой службы"
+author: Сёмин П. О.
 output: 
@@
 
-Пермский государственный национальный исследовательский университет, г. Пермь
+Пермский государственный национальный исследовательский университет
+
+Россия, 614068, г. Пермь, ул. Букирева, 15
+
+E-mail: ntsp\@ya.ru
+
 </div>
@@
 
-*Аннотация:*
+**Аннотация.** В статье представлены результаты анализа миграции юридических фирм в России в 2016–2023 гг.
+Миграция компании понимается как изменение адреса регистрации.
+Исследование основано на открытых данных Федеральной налоговой службы (ФНС России), преобразованных в панельный геопривязанный табличный набор данных.
+Методы анализа — описательная статистика, статистический тест Фишера, картографирование, визуализация данных.
+Показано, что миграции юридических фирм в России — редкое явление.
+Как правило, компании перемещаются на небольшое расстояние (10–50 км).
+Две трети организаций мигрируют в пределах субъекта федерации.
+Малые предприятия меняют адрес статистически значимо чаще, чем микропредприятия.
+Пространственная картина миграций разнообразная и не характеризуется чёткими тенденциями.
+Москва лидирует по оттоку фирм, а Московская область — по их притоку.
+Центростремительная и центробежные тенденции примерно уравновешивают друг друга по числу перемещающихся фирм и работников, но по движению доходов преобладает центростремительная.
 
-*Ключевые слова:* миграция компаний, малое и среднее предпринимательство, юридические компании, юридические услуги, география услуг, анализ данных, открытые данные, ФНС России
+**Ключевые слова:** миграция компаний, малое и среднее предпринимательство, юридические компании, юридические услуги, география услуг, анализ данных, открытые данные, ФНС России
+
+</div>
+
+<div custom-style="Title">
+Spatial Analysis of Migration of Legal Companies in Russia (2016–2023) Based on Tax Service Open Data
+<div>
+
+<div custom-style="Author">
+Pavel O. Syomin
+
+15 Bukireva st., Perm, 614068, Russia
+
+E-mail: ntsp\@ya.ru
+
+</div>
+
+<div custom-style="Abstract">
+
+**Abstract.** The paper features the results of analysis of migrations of Russian legal companies in 2016–23.
+Migration here is defined as a change in legal address of a company.
+The research is based on open data published by Federal Tax Service of Russia that has been processed to make a panel geo-referenced tabular dataset.
+Descriptive statistics, exact Fisher's test, data visualization, and mapping are used as methods.
+It is found that migration of legal companies is relatively a rare phenomenon.
+Typically, businesses move to a short distance (10–50 km), and two thirds of all the migrations are intra-regional, i.e.
+within the constituent entity of Russia.
+Small companies are more likely to migrate compared to microbusinesses.
+The spatial patterns and migrations flows are diverse and have few remarkable tendencies.
+Moscow city has the top decline in the number of legal companies, while Moscow region (oblast) has gained the largest number of businesses.
+The flows from or to the center are almost equal if measured by the number of companies or employees.
+However, if estimated by the total revenue, the migration from the periphery to the center dominates.
+
+**Keywords:** companies migration, small and medium-sized businesses, legal companies, legal service, service geography, data analysis, open data, FTS of Russia
 
@@
 
-**Данные.** Для анализа использовался дезагрегированный геопривязанный набор данных о субъектах малого и среднего предпринимательства (МСП) в России, ведущих деятельность в области права (код основного вида деятельности по ОКВЭД 69.10).
+## Данные
+
+Для анализа использовался дезагрегированный геопривязанный набор данных о субъектах малого и среднего предпринимательства (МСП) в России, ведущих деятельность в области права (код основного вида деятельности по ОКВЭД 69.10).
 Он был подготовлен на базе выгрузок из реестра МСП, размещённых на сайте Федеральной налоговой службы в формате открытых данных[^1].
@@
 
-**Методы.** Миграция компании в данном исследовании понимается как изменение адреса регистрации юридического лица.
+## Методы
+
+Миграция компании в данном исследовании понимается как изменение адреса регистрации юридического лица.
 Похожее определение принято, например, в работах Н.
@@
 
-**Распределение компаний по количеству миграций** показано в таблице 1.
-Видно, что подавляющее большинство (97%) компаний за 2016–2023 гг.
-ни разу не мигрировали.
+## Распределение компаний по количеству миграций
+
+Подавляющее большинство (97%) компаний за 2016–2023 гг.
+ни разу не мигрировали (табл.
+1).
 Около 2.5% фирм поменяли местонахождение однократно, а многократные релокации крайне редки.
@@
   col.names = c("Количество миграций", "Количество компаний", "%"),
-  caption = "Таблица 1. Распределение юридических компаний по количеству миграций")
+  caption = "Таблица 1. Распределение юридических компаний по количеству миграций <br> Distribution of legal companies by the count of migrations")
 ```
 
-**Связь между миграцией и размером компаний** отражена в таблице 2.
+## Связь между миграцией и размером компаний
+
+Вероятность миграции зависит от категории фирмы (табл.
+2).
 Хотя пропорции не слишком очевидны из-за дисбаланса категорий, малые предприятия мигрируют статистически значимо чаще, чем микропредприятия (по результатам точного теста Фишера `r ifelse(categories_test$p.value < 0.001, "p < 0.001", paste("p = ", categories_test$p.value))`, соотношение шансов `r paste0("1 к ", round(1 / categories_test$estimate, 0))`.
@@
   col.names = c("Категория фирмы", "Всего", "Из них мигрировавших"),
-  caption = "Таблица 2. Распределение юридических компаний по категории и статусу миграции"
+  caption = "Таблица 2. Распределение юридических компаний по категории и статусу миграции <br> Distribution of legal companies by category and migration status"
 )
@@
 
-**Распределение миграций по расстоянию** показано на графике 1.
-Видно, что преобладают перемещения на небольшие расстояния (10–50 км), хотя есть вторая заметная группа — миграции на расстояние около 1000 км.
+## Распределение миграций по расстоянию
+
+Преобладают перемещения на небольшие расстояния (10–50 км), хотя есть вторая заметная группа — миграции на расстояние около 1000 км (рис.
+1).
 Более высокая частота релокаций на небольшое расстояние отмечалась и в других исследованиях [@Pellenbarg2005; @Rossi2019].
 
-```{r plot-1-distance}
+```{r fig-1-distance, fig.cap="Рис. 1. Распределение расстояния миграций <br> Fig. 1. Distribution of migration distance"}
 mig_distance_plot
@@
 
-**Распределение миграций по времени** показано на графиках 2 и 3.
-Закономерности за весь рассмотренный период отсутствуют: интенсивность миграции выглядит случайной.
-Тем не менее, видно, что компании чаще меняют своё местонахождение летом.
+## Распределение миграций по времени
+
+Закономерности за весь рассмотренный период отсутствуют: интенсивность миграции выглядит случайной (рис.
+2).
+Тем не менее, видно, что компании чаще меняют своё местонахождение летом (рис.
+3), а осенью и зимой частота перемещений обычно понижается.
 Возможно, это связано с тем, что летом спрос на юридические услуги и возможность их оказания уменьшаются из-за отпусков у клиентов и служащих государственных структур, и возникает благоприятное окно времени для переезда.
 
-```{r plot-2-timeline}
+```{r fig-2-timeline, fig.cap="Рис. 2. Количество миграций за каждый месяц с августа 2016 г. по ноябрь 2023 г. <br> Fig. 2. Migration count for each month from August 2016 to November 2023"}
 mig_timeline_plot
@@
 
-```{r plot-3-monthly}
+```{r fig-3-monthly, fig.cap="Рис. 3. Количество миграций по месяцам года <br> Fig. 3. Migration count by month"}
 mig_by_month_plot
@@
 
-**Распределение регионов по притоку (оттоку) компаний** изображено на картограмме 1.
-Общая закономерность отсутствует, при этом заметно, что показатели противоречат друг другу: регионы, лидирующие по сальдо исходя из числа компаний, могут проигрывать при расчёте по выручке или численности работников.
+## Распределение регионов по притоку (оттоку) компаний
 
-```{r fig-4-mig-by-region-plot, fig.cap="*Fig. 4.* Net migration rate of legal companies by regions", fig.height=3}
+Картограмма (рис.
+4) показывает сальдо миграции по регионам (субъектам федерации).
+Однозначной тенденции не прослеживается, но заметно, что в Московскую область переехало больше всего фирм.
+Сама Москва при этом — абсолютный лидер по оттоку фирм.
+Это согласуется, в частности, с данными из Польши, где также характерен уход компаний из столицы в прилегающие районы [@Cybulska2020].
+В остальном сальдо миграции компаний распределено по регионам хаотично, при этом абсолютные значения (число переехавших организаций) невелики — до 25.
+
+```{r fig-4-mig-by-region-plot, fig.cap="Рис. 4. Сальдо миграции юридических фирм по регионам <br> *Fig. 4.* Net migration rate of legal companies by regions", fig.height=3}
 mig_by_region_plot
@@
 
-**Распределение населённых пунктов по притоку (оттоку) компаний** показано на картограмме 2.
-Здесь также сложно выявить закономерности, хотя лидируют по сальдо, измеренному как число компаний, города-спутники Москвы.
-Опять же, показатели противоречивые: лидерство по численному сальдо не означает лидерство по выручке или числу работников.
+## Распределение населённых пунктов по притоку (оттоку) компаний
 
-**Соотношение внутрирегиональной миграции и межрегиональной миграции** показано в таблице 3.
-Видно, что между регионами перемещается примерно треть фирм, а остальные две трети — внутри региона.
-При этом в пересчёте на число работников и выручку межрегиональная миграция масштабнее, чем внутрирегиональная, из чего следует, что компания, поменявшая регион, в среднем «везёт» с собой больше денег и людей, чем компания, поменявшая населённый пункт внутри региона.
+Картограмма, детализированная до населённых пунктов, показана на 
+рис.
+5.
+Здесь видно, что населённые пункты с положительным сальдо миграции компаний сосредоточены в основном вокруг Москвы и Санкт-Петербурга.
+Кроме того, можно заметить, что в значительной доле субъектов федерации есть как населённые пункты, принимающие фирмы, так и населённые пункты, теряющие их.
 
-```{r table-3-inter-intra}
+```{r fig-5-migration-by-settlement, fig.cap="Рис.
+5.
+Сальдо миграции юридических фирм (положительное или отрицательное) по населённым пунктам <br> Fig.
+5.
+Net migration rate of legal companies (positive or negative) by settlements"}
+migration_by_settlement_plot
+```
+
+## Соотношение внутрирегиональной миграции и межрегиональной миграции
+
+Между субъектами России перемещается примерно треть фирм, а остальные две трети — внутри (табл.
+3).
+Такая пропорция согласуется с распределением миграций по расстоянию: перемещения внутри региона обычно короткие.
+В пересчёте на число работников и выручку межрегиональная миграция масштабнее, чем внутрирегиональная.
+Можно сделать выводы, что компания, поменявшая регион, в среднем «перевозит» с собой больше денег и людей, чем компания, поменявшая населённый пункт внутри региона, и, наоборот, что для смены региона фирма должна обладать более существенными ресурсами, чем для перемещения внутри региона.
+
+```{r table-3-inter-intraregional-migration}
 kable(
@@
   col.names = c("Тип миграции", "Число", "Доходы", "Работники"),
-  caption = "Таблица 3. Показатели межрегиональной и внутрирегиональной миграции юридических компаний: число фирм, суммарные доходы (млн рублей), суммарное число работников"
+  caption = "Таблица 3.
+Показатели межрегиональной и внутрирегиональной миграции юридических компаний: число фирм, суммарные доходы (млн рублей), суммарное число работников <br> Metrics of inter- and intra-regional migration of legal companies: businesses count, total revenue, total number of employees"
 )
@@
 
-**Пути межрегиональной миграции** отражены в таблице 4.
+## Пути межрегиональной миграции
+
+Лидируют миграции из Москвы и Санкт-Петербурга в прилегающие к ним области (табл.
+4).
+Москва также играет роль в менее крупных миграционных потоках.
+Важное место столицы в маршрутах релокации фирм согласуется с результатами исследований в Польше [@Cybulska2020].
+Нет чёткой связи между числом переехавших компаний и «перевезённой» выручкой или числом работников.
 Абсолютные числа переместившихся компаний, как правило, невелики.
-Лидируют миграции из Москвы и Санкт-Петербурга в прилегающие к ним области.
-Нет чёткой связи между числом переехавших компаний и «перевезённой» выручкой или числом работников.
 
@@
 kable(
-  rbind(head(regional_paths_combined, 5), tail(regional_paths_combined, 5)),
+  head(regional_paths_combined, 10),
   col.names = c("Путь", "Количество", "Доходы, млн", "Работники"),
@@
 Маршруты миграции юридических компаний по субъектам России и их показатели: число фирм, суммарные доходы в миллионах рублей, суммарная численность работников.
-Поразаны первые и последние 5 маршрутов по числу фирм"
+Показаны первые 10 маршрутов по числу фирм <br> Routes of legal companies migration by regions and their metrics (businesses count, total revenue, total number of employees).
+Top 10 routes by businesses count are shown"
 )
@@
 
-**Сочетание централизации и децентрализации в межрегиональной миграции** (таблица 5) преимущественно равномерное.
+## Сочетание централизации и децентрализации в межрегиональной миграции
+
+Показатели центростремительных, центробежных и прочих миграций характеризуются преимущественно равномерным распределением (табл.
+5).
 По числу работников и компаний преобладает центробежная тенденция, а по доходам — центростремительная.
-Треть потоков не может быть описани ни как центростремительные, ни как центробежные.
+Можно сделать вывод, что в центр мигрируют более экономически успешные компании.
 
@@
   col.names = c("Тип миграции", "Число", "Доходы", "Работники"),
-  caption = "Таблица 5. Показатели центростремительных и центробежных межрегиональных миграций юридических компаний: число фирм, суммарные доходы в миллионах рублей, суммарная численность работников"
+  caption = "Таблица 5.
+Показатели центростремительных и центробежных межрегиональных миграций юридических компаний: число фирм, суммарные доходы в миллионах рублей, суммарная численность работников <br> Metrics of legal companies inter-regional migration (businesses count, total revenue, total number of employees) by general direction: from the center to peripheries, from peripheries to the center, other"
 )
@@
 
-**Пути внутрирегиональной миграции** (таблица 6) ещё более неоднозначны, чем пути межрегиональной миграции, поскольку абсолютные перемещения почти во всех случаях измеряются единицами.
-Нет и связи между тремя показателями: число фирм, доходы, число работников.
+## Пути внутрирегиональной миграции
+
+Типичных и массовых маршрутов внутри субъектов России нет — вместо этого есть разнообразие перемещений в разных направлениях (табл.
+6).
+Так же как и в случае с межрегиональной миграцией, не видно связи между движением компаний и движением доходов или числа работников.
 
@@
 kable(
-  rbind(head(settlement_paths_combined, 5), tail(settlement_paths_combined, 5)),
+  head(settlement_paths_combined, 10),
   col.names = c("Маршрут", "Количество", "Доходы, млн", "Работники"),
@@
 Маршруты миграции юридических компаний по населённым пунктам и их показатели: число фирм, суммарные доходы в миллионах рублей, суммарная численность работников.
-Поразаны первые и последние 5 маршрутов по числу фирм"
+Поразаны первые и последние 5 маршрутов по числу фирм <br> Routes of legal companies migration by settlements and their metrics (businesses count, total revenue, total number of employees).
+Top 10 routes by businesses count are shown"
 )
@@
 
-**Сочетание централизации и децентрализации во внутрирегиональной миграции** (таблица 7) почти такое же, как и в случае с межрегиональной миграцией: перемещения к центру выигрывают по числу фирм и людей, но уступают в доходах, при этом треть релокаций не могут быть описаны ни как центростремительные, ни как центробежные.
+## Сочетание централизации и децентрализации во внутрирегиональной миграции
+
+Централизация и децентрализация миграций внутри регионов отчасти похожа на ту, что наблюдается в случае с межрегиональной миграцией: распределение по числу фирм близко к равномерному, центростремительная тенденция уступает центробежной по количеству компаний и работников, но превосходит по доходам (табл.
+7).
 
@@
   caption = "Таблица 7.
-Показатели центростремительных и центробежных внутрирегиональных миграций юридических компаний: число фирм, суммарные доходы в миллионах рублей, суммарная численность работников"
+Показатели центростремительных и центробежных внутрирегиональных миграций юридических компаний: число фирм, суммарные доходы в миллионах рублей, суммарная численность работников <br> Metrics of legal companies intra-regional migration (businesses count, total revenue, total number of employees) by general direction: from the center to peripheries, from peripheries to the center, other"
 )
@@
 
-# Выводы
+# Заключение
+
+Миграция юридических фирм в России — сравнительно редкое явление: за 2016–2023 гг.
+лишь около 3% организаций поменяли адрес регистрации.
+Более крупные фирмы (малые предприятия) перемещаются чаще, чем мелкие (микропредприятия).
+Компании чаще всего мигрируют на небольшое расстояние (10–50 км) внутри региона, но типичны и межрегиональные миграции на относительно большое расстояние (1000 км).
+За рассмотренные годы не выявлено периодов резкого повышения миграционной активности, но при этом такая активность ежегодно возрастает летом и снижается осенью и зимой.
+
+Миграция юридических фирм в России приблизительно равномерно сочетает центростремительную (к Москве или административным центрам субъектов федерации) и центробежную тенденции.
+При этом при оценке по доходам центростремительная тенденция доминирует: можно предположить, что более экономически успешные компании стремятся в центр.
+Москва — лидер по абсолютному оттоку фирм, а Московская область — лидер по их абсолютному притоку, и маршрут «из Москвы в область» является наиболее мощным исходя из количества фирм.
+Заметными являются также миграционные потоки Москвы с другими субъектами России.
+Распределение иных регионов и населённых пунктов по числу миграций фирм, равно как и маршруты миграции, не характеризуются выраженными тенденциями и при этом невелики по количественным показателям.
+Положительное сальдо миграции исходя из числа фирм может сочетаться с отрицательным сальдо по доходам или числу работников.
 
@@
 
-<div custom-style="Title">
-Spatial Analysis of Migration of Legal Companies in Russia (2016–2023) Based on Tax Service Open Data
-<div>
+# Информация об авторе
 
-<div custom-style="Author">
-P. O. Syomin
+**Сёмин Павел Олегович**, аспирант, кафедра социально-экономической географии, Пермский государственный национальный исследовательский университет, г. Пермь, Россия
 
-Perm State University, Perm
-</div>
-
-<div custom-style="Abstract">
-
-*Abstract:*
-
-*Keywords:* companies migration, small and medium-sized businesses, legal companies, legal service, service geography, data analysis, open data, FTS of Russia
-
-</div>
+**Pavel O. Syomin**, PhD student, Department of Social and Economic Geography, Perm State University, Perm, Russia

### 2024-02-20 legal-companies-migration/Paper.Rmd
Edit styles of document and bibliography, add references in english, other improvements

@@
 title: "УДК 911.3"
-subtitle: "Пространственный анализ миграции юридических компаний в России в 2016–2023 гг. по открытым данным налоговой службы"
+subtitle: "Анализ миграции юридических компаний в России в 2016–2023 гг. по открытым данным налоговой службы"
 author: Сёмин П. О.
@@
 
+<br>
+
 <div custom-style="Abstract">
@@
 
+<br>
+
 **Ключевые слова:** миграция компаний, малое и среднее предпринимательство, юридические компании, юридические услуги, география услуг, анализ данных, открытые данные, ФНС России
@@
 
-<div custom-style="Title">
-Spatial Analysis of Migration of Legal Companies in Russia (2016–2023) Based on Tax Service Open Data
-<div>
+<div custom-style="Subtitle">
+Analysis of Migration of Legal Companies in Russia (2016–2023) Based on Tax Service Open Data
+</div>
 
@@
 </div>
+
+<br>
 
@@
 However, if estimated by the total revenue, the migration from the periphery to the center dominates.
+
+<br>
 

### 2024-02-20 legal-companies-migration/Script.R

@@ -4,10 +4,6 @@ library(glue)
 library(readr)
 library(tidyr)
 library(sf)
-library(units)
-
-# Base theme for plots
-theme <- theme_bw(base_size = 11, base_family = "Times New Roman")
 
 # Load the data
 data <- read_csv("../../ru-smb-companies/legal/smb.csv")
@@ -151,7 +147,8 @@ mig_distance_plot <- migrations %>%
   geom_freqpoly(bins = 50) +
   scale_x_continuous(trans = "log10", breaks = 10^(-1:3), labels = ~ .x) +
   theme_bw(base_size = 12, base_family = "Times New Roman") +
-  labs(x = "Distance of migration, km (log10)", y = "Count of companies")
+  labs(x = "Расстояние миграции, км (десятичный логарифм)",
+       y = "Число фирм")
 
 # Date of migrations
 # Overall timeline
@@ -162,16 +159,18 @@ mig_timeline_plot <- migrations %>%
     scale_x_date(
       date_breaks = "6 months", date_labels = "%m.%Y", 
       guide = guide_axis(angle = 90)) +
-  theme_bw() +
-  labs(x = "Месяц и год", y = "Количество компаний")
+  labs(x = "Месяц и год", y = "Количество компаний") +
+  theme_bw(base_size = 12, base_family = "Times New Roman") +
+  theme(panel.grid.minor.y = element_blank())
 
 # By month
 mig_by_month_plot <- migrations %>% 
   mutate(month = format(date, "%m")) %>% 
   ggplot(aes(x = month)) +
   geom_bar() +
-  theme_bw() +
-  labs(x = "Месяц", y = "Количество компаний")
+  labs(x = "Месяц", y = "Количество компаний") +
+  theme_bw(base_size = 12, base_family = "Times New Roman") +
+  theme(panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank())
 
 # Migration by regions
 migration_by_region <- migrations %>% 
@@ -227,15 +226,17 @@ migration_by_settlement <- migrations %>%
   rbind(migration_msk_spb) %>% 
   left_join(settl_coords) %>% 
   st_as_sf()
+
 migration_by_settlement_plot <- migration_by_settlement %>% 
   filter(abs(count) > 1) %>% 
   mutate(direction = count > 0, count = abs(count)) %>% 
   ggplot() +
-  geom_sf(data = regions_geo, size = .5) +
+  geom_sf(data = regions_geo, size = .5, fill = "white") +
[…]

### 2024-02-20 legal-companies-migration/Script.R
Change base size for plots from 12 to 11

@@ -46,7 +46,7 @@ mig_count_distr_plot <- migration_count_distribution %>%
   labs(
     x = "Number of migrations by a company", 
     y = "Number of companies (log10)") +
-  theme_bw(base_size = 12, base_family = "Times New Roman") +
+  theme_bw(base_size = 11, base_family = "Times New Roman") +
   theme(
     panel.grid.minor.x = element_blank(), 
     panel.grid.major.x = element_blank(),
@@ -137,7 +137,7 @@ mig_by_category_plot <- count_by_category %>%
     scale_fill_discrete(name = "", labels = c("Migrated", "Stayed")) +
     coord_flip() +
     labs(x = "", y = "Share of companies") +
-    theme_bw(base_size = 12, base_family = "Times New Roman") +
+    theme_bw(base_size = 11, base_family = "Times New Roman") +
     theme(panel.grid.minor.x = element_blank())
     
 # Distance of migrations
@@ -146,7 +146,7 @@ mig_distance_plot <- migrations %>%
   ggplot(aes(x = distance)) +
   geom_freqpoly(bins = 50) +
   scale_x_continuous(trans = "log10", breaks = 10^(-1:3), labels = ~ .x) +
-  theme_bw(base_size = 12, base_family = "Times New Roman") +
+  theme_bw(base_size = 11, base_family = "Times New Roman") +
   labs(x = "Расстояние миграции, км (десятичный логарифм)",
        y = "Число фирм")
 
@@ -160,7 +160,7 @@ mig_timeline_plot <- migrations %>%
       date_breaks = "6 months", date_labels = "%m.%Y", 
       guide = guide_axis(angle = 90)) +
   labs(x = "Месяц и год", y = "Количество компаний") +
-  theme_bw(base_size = 12, base_family = "Times New Roman") +
+  theme_bw(base_size = 11, base_family = "Times New Roman") +
   theme(panel.grid.minor.y = element_blank())
 
 # By month
@@ -169,7 +169,7 @@ mig_by_month_plot <- migrations %>%
   ggplot(aes(x = month)) +
   geom_bar() +
   labs(x = "Месяц", y = "Количество компаний") +
-  theme_bw(base_size = 12, base_family = "Times New Roman") +
+  theme_bw(base_size = 11, base_family = "Times New Roman") +
   theme(panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank())
 
 # Migration by regions
[…]

### 2024-02-21 legal-companies-migration/Paper.Rmd
Proofreading, minor fixes

@@
 csl: "biblio-styles-ru.csl"
+
 ---
@@
 <div custom-style="Subtitle">
-Analysis of Migration of Legal Companies in Russia (2016–2023) Based on Tax Service Open Data
+Analysis of Migration of Russian Legal Companies in 2016–23 Based on Tax Service Open Data
 </div>
@@
 Small companies are more likely to migrate compared to microbusinesses.
-The spatial patterns and migrations flows are diverse and have few remarkable tendencies.
+The spatial patterns and migration flows are diverse and have few remarkable tendencies.
 Moscow city has the top decline in the number of legal companies, while Moscow region (oblast) has gained the largest number of businesses.
@@
 Оно связано с мобильностью и стоимостью труда и капитала, инвестиционной привлекательностью, международной конкуренцией и торговлей, агломерационными эффектами, глобализацией, экономическим ростом, налоговой оптимизацией.
-Исследование миграции компаний восходит к середине 20 века: одной из первых научных работ на эту тему считается книга *Why Industry Moves South* Г.
-МакЛафлина и С.
-Робока.
-В ней описывается перемещение промышленных производств с севера на юг США, вызванное особенностями рынка труда: меньшими зарплатами и меньшей активностью просфоюзов на юге [@McLaughlin1949].
+Исследование миграции компаний восходит к середине 20 века: одной из первых научных работ на эту тему считается книга *Why Industry Moves South* [@McLaughlin1949].
+В ней описывается перемещение промышленных производств с севера на юг США, вызванное особенностями рынка труда: меньшими зарплатами и меньшей активностью профсоюзов на юге .
 
@@
 Заметную роль играют венчурные кластеры, которые «притягивают» фирмы своим финансированием [@Plant2007], и миграция основателей стартапов [@Kutsenko2022].
-Отдельные исследования посвящены более гуманитарным аспектам проблемы, таким как дискурс о релокации [@Piotti2007].
+Отдельные исследования посвящены гуманитарно-философским аспектам проблемы, таким как дискурс о релокации [@Piotti2007].
 Существенный пласт научных работ посвящён налоговым причинам и последствиям релокации [@Burwitz2006; @Chand2013; @Kubicova2016].
@@
 Исследований, посвящённых миграции компаний в России, немного.
-Они касаются в основном налоговых вопросов: фиктивной миграции, её влиянии на собираемость налогов и налоговую задолженность, мер борьбы с ней [@Adigamova2011; @Shcherbakova2016], использовании миграции в особые экономические зоны в качестве инструмента налоговой оптимизации [@Sinenko2021].
+Они касаются в основном налоговых вопросов: фиктивной миграции, её влияния на собираемость налогов и налоговую задолженность, мер борьбы с ней [@Adigamova2011; @Shcherbakova2016], использования миграции в особые экономические зоны в качестве инструмента налоговой оптимизации [@Sinenko2021].
 В таких работах косвенно затрагиваются пространственные аспекты (например, называются регионы, в которые перемещается большое количество фирм), но основной целью исследования они не являются.
@@
 В-третьих, юридические услуги важны для бизнеса, и перемещение фирм, оказывающих их, косвенно связано с экономической активностью в целом.
-В-четвёртых, выбор одной конкретной группы компаний упрощает анализ, поскольку уменьшает количество исходных данных, которых, как будет показано ниже, в исходном виде довольно много.
+В-четвёртых, выбор одной конкретной группы компаний упрощает анализ, поскольку уменьшает количество исходных данных, которых довольно много.
 
@@
 Для анализа использовался дезагрегированный геопривязанный набор данных о субъектах малого и среднего предпринимательства (МСП) в России, ведущих деятельность в области права (код основного вида деятельности по ОКВЭД 69.10).
-Он был подготовлен на базе выгрузок из реестра МСП, размещённых на сайте Федеральной налоговой службы в формате открытых данных[^1].
-Объём исходных выгрузок составляет около 200 Гб в сжатом виде и более 2 Тб в несжатом, поэтому для их обработки автор самостоятельно разработал консольное приложение *ru-smb-companies*[^2] на языке программирования Python, которое обрабатывает исходную информацию и формирует на её базе компактную таблицу с фирмами, имеющими необходимый код ОКВЭД.
+Он был подготовлен на базе выгрузок из реестра МСП, размещённых на сайте Федеральной налоговой службы в формате открытых данных[^1], а также данных о доходах и расходах[^2] и среднесписочной численности работников организаций[^3].
+Объём исходных выгрузок составляет около 200 Гб в сжатом виде и более 2 Тб в несжатом, поэтому для их обработки автор самостоятельно разработал консольное приложение *ru-smb-companies*[^4] на языке программирования Python, которое формирует на базе открытых компактную таблицу с фирмами, имеющими необходимый код ОКВЭД.
 Набор данных охватывает период с августа 2016 г.
@@
 Таблица имеет, в частности, такие поля (колонки): наименование компании, её категория (малое, среднее или микропредприятие), ИНН, регион, административный район, населённый пункт регистрации, его географические координаты, доходы, расходы и число работников организации за каждый год с 2019.
+
+[^1]: https://www.nalog.gov.ru/opendata/7707329152-rsmp/
+[^2]: https://www.nalog.gov.ru/opendata/7707329152-revexp/
+[^3]: https://www.nalog.gov.ru/opendata/7707329152-sshr2019/
+[^4]: https://github.com/PavelSyomin/russian-smb-companies
 
@@
 специалистами Института проблем правоприменения [@Moiseeva2016], в России на конец июля 2016 г.
-действовало 47 тысяч коммерческих юридических фирм, а подсчёты на базе набора данных, используемого в текущем исследовании, дают оценку их количества в `r `format(round(companies_count_08_2016, -3))` на 10 августа 2016 г.
+действовало 47 тысяч коммерческих юридических фирм, а подсчёты на базе набора данных, используемого в текущем исследовании, дают оценку их количества в `r format(round(companies_count_08_2016, -3))` на 10 августа 2016 г.
 Иначе говоря, в наборе данных есть сведения о примерно трёх четвертях всех необходимых организаций.
-Более полный список можно было бы получить, например, имея доступ к Единому государственному реестра юридических лиц (ЕГРЮЛ), но, в отличие от реестра МСП, он в полном виде не публикуется.
-
-[^1]: https://www.nalog.gov.ru/opendata/7707329152-rsmp/
-[^2]: https://github.com/PavelSyomin/russian-smb-companies
+Более полный список можно было бы получить, например, имея доступ к Единому государственному реестру юридических лиц (ЕГРЮЛ), но, в отличие от реестра МСП, он в полном виде не публикуется.
 
@@
 Он включает в себя 11 вопросов, некоторые из которых взаимосвязаны.
-Каждому вопросу отводится свой небольшой подраздел в «Результатах и их обсуждении».
+Каждому вопросу отводится подраздел в «Результатах и их обсуждении».
 Для начала рассматривается распределение компаний по количеству миграций (1) и связь между миграцией и размером компании (2).
@@
 Следующий шаг — картографирование распределения регионов (5) и населённых пунктов (6) по сальдо миграции.
-После этого изучается соотношение внутри- и межрегиональной миграци (7).
+После этого изучается соотношение внутри- и межрегиональной миграции (7).
 Далее отдельно рассматриваются внутрирегиональная миграция: определяются пути миграции и их количественные характеристики (8) и анализируется сочетание централизации и децентрализации (9).
@@
 С технической точки зрения исследование реализовано с помощью языка программирования R с дополнительными пакетами.
-Исходный код доступен в репозитории статьи[^3].
+Исходный код доступен в репозитории статьи[^5].
 
-[^3]: https://github.com/PavelSyomin/ru-smb-companies-papers/tree/main/legal-companies-migration
+[^5]: https://github.com/PavelSyomin/ru-smb-companies-papers/tree/main/legal-companies-migration
 
@@
 2).
-Хотя пропорции не слишком очевидны из-за дисбаланса категорий, малые предприятия мигрируют статистически значимо чаще, чем микропредприятия (по результатам точного теста Фишера `r ifelse(categories_test$p.value < 0.001, "p < 0.001", paste("p = ", categories_test$p.value))`, соотношение шансов `r paste0("1 к ", round(1 / categories_test$estimate, 0))`.
+Хотя пропорции не слишком очевидны из-за дисбаланса категорий, малые предприятия мигрируют статистически значимо чаще, чем микропредприятия (по результатам точного теста Фишера `r ifelse(categories_test$p.value < 0.001, "p < 0.001", paste("p = ", categories_test$p.value))`, соотношение шансов `r paste0("1 к ", round(1 / categories_test$estimate, 0))`).
 Делать выводы в отношении средних предприятий сложно, поскольку их в наборе данных мало.
@@
 Это согласуется, в частности, с данными из Польши, где также характерен уход компаний из столицы в прилегающие районы [@Cybulska2020].
-В остальном сальдо миграции компаний распределено по регионам хаотично, при этом абсолютные значения (число переехавших организаций) невелики — до 25.
+В остальном сальдо миграции компаний распределено по регионам хаотично, при этом абсолютные значения (число переехавших организаций) невелики — обычно до 25.
 
-```{r fig-4-mig-by-region-plot, fig.cap="Рис. 4. Сальдо миграции юридических фирм по регионам <br> *Fig. 4.* Net migration rate of legal companies by regions", fig.height=3}
+```{r fig-4-mig-by-region-plot, fig.cap="Рис. 4. Сальдо миграции юридических фирм по регионам <br> Fig. 4. Net migration rate of legal companies by regions", fig.height=3}
 mig_by_region_plot
@@
 
-## Соотношение внутрирегиональной миграции и межрегиональной миграции
+## Соотношение внутрирегиональной и межрегиональной миграции
 
@@
   inter_intra,
-  col.names = c("Тип миграции", "Число", "Доходы", "Работники"),
+  col.names = c("Тип миграции", "Число фирм", "Доходы, млн. руб.", "Число работников"),
   caption = "Таблица 3.
-Показатели межрегиональной и внутрирегиональной миграции юридических компаний: число фирм, суммарные доходы (млн рублей), суммарное число работников <br> Metrics of inter- and intra-regional migration of legal companies: businesses count, total revenue, total number of employees"
+Показатели межрегиональной и внутрирегиональной миграции юридических компаний <br> Metrics of inter- and intra-regional migration of legal companies: businesses count, total revenue, total number of employees"
 )
@@
   head(regional_paths_combined, 10),
-  col.names = c("Путь", "Количество", "Доходы, млн", "Работники"),
+  col.names = c("Путь", "Число фирм", "Доходы, млн. руб.", "Число работников"),
   caption = "Таблица 4.
@@
   reg_central,
-  col.names = c("Тип миграции", "Число", "Доходы", "Работники"),
+  col.names = c("Тип миграции", "Число фирм", "Доходы, млн. руб.", "Число работников"),
   caption = "Таблица 5.
@@
   head(settlement_paths_combined, 10),
-  col.names = c("Маршрут", "Количество", "Доходы, млн", "Работники"),
+  col.names = c("Маршрут", "Число фирм", "Доходы, млн. руб.", "Число работников"),
   caption = "Таблица 6.
@@
   settl_central,
-  col.names = c("Тип миграции", "Число", "Доходы", "Работники"),
+  col.names = c("Тип миграции", "Число фирм", "Доходы, млн. руб.", "Число работников"),
   caption = "Таблица 7.
-Показатели центростремительных и центробежных внутрирегиональных миграций юридических компаний: число фирм, суммарные доходы в миллионах рублей, суммарная численность работников <br> Metrics of legal companies intra-regional migration (businesses count, total revenue, total number of employees) by general direction: from the center to peripheries, from peripheries to the center, other"
+Показатели центростремительных и центробежных внутрирегиональных миграций юридических компаний <br> Metrics of legal companies intra-regional migration (businesses count, total revenue, total number of employees) by general direction: from the center to peripheries, from peripheries to the center, other"
 )
@@
 Заметными являются также миграционные потоки Москвы с другими субъектами России.
-Распределение иных регионов и населённых пунктов по числу миграций фирм, равно как и маршруты миграции, не характеризуются выраженными тенденциями и при этом невелики по количественным показателям.
+Распределение иных регионов и населённых пунктов по числу миграций фирм, равно как и маршруты миграции, не характеризуются выраженными тенденциями и при этом обладают небольшими количественными показателями.
 Положительное сальдо миграции исходя из числа фирм может сочетаться с отрицательным сальдо по доходам или числу работников.

### 2024-02-21 legal-companies-migration/Script.R

@@ -147,7 +147,7 @@ mig_distance_plot <- migrations %>%
   geom_freqpoly(bins = 50) +
   scale_x_continuous(trans = "log10", breaks = 10^(-1:3), labels = ~ .x) +
   theme_bw(base_size = 11, base_family = "Times New Roman") +
-  labs(x = "Расстояние миграции, км (десятичный логарифм)",
+  labs(x = "Расстояние миграции, км",
        y = "Число фирм")
 
 # Date of migrations
@@ -244,8 +244,8 @@ inter_intra <- migrations %>%
   mutate(
     type = ifelse(
       region_from != region_to, 
-      "inter-regional", 
-      "intra-regional"),
+      "Межрегиональная", 
+      "Внутрирегиональная"),
     count = 1) %>% 
   group_by(type) %>% 
   summarise(across(c("count", "revenue", "empl"), sum, na.rm = TRUE))

### 2024-02-27 legal-companies-by-cities/Paper.Rmd
Start writing paper about concentration and accessibility of legal companies in cities: script, template docx, biblio styles, metadata, introduction, data description

---
title: "UDC 911.3"
subtitle: "Does Concentration of Legal Companies Enable Accessibility of Legal Service: Empirical Evidence from Russian Cities"
output:
  word_document:
    reference: reference.docx
bibliography: biblio.bib
csl: biblio-style.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<div custom-style="Author">

Pavel O. Syomin

<br>

Perm State University

<br>

e-mail: ntsp\@ya.ru

<br>

The paper is published for the first time.

</div>

<br>

<div custom-style="Abstract">

**Abstract:**

**Keywords:**

</div>

<br>

Cities are known for being the centers of business activity. This is especially true, in particular, for large cities that concentrate many people and companies and thus usually act as service providers for the surrounding areas. It is often said that urbanization improves the quality of life compared to the rural neighborhoods by enhancing people's access to a variety of goods and services. Also, the co-location of companies in cities due to agglomeration effects gives economic benefits for business, thus further increasing the concentration. However, it is doubtful whether the concentration itself necessarily leads to better provision of urban services. Moreover, it is worth studying whether the high accessibility of services may be achieved by other means rather than concentration, because the top cities by the number of businesses are typically the largest cities, and large cities often have their own problems which partially offset their positive aspects.

In this paper, the relationship between the concentration of legal companies in Russian cities and people's access to legal services is empirically tested based on data about location of legal organizations if Russia. Four research questions are asked. The first one is: what is the relationship between city population and number of legal businesses and legal workers in the city? The second issue is: does the distribution of legal businesses and legal workers in Russian cities comply with well-known Zipf (rank-size) law? The third problem and the main question is: how does the accessibility of legal professionals depends on the concentration of legal companies with respect to city size by population and its relative size in the region (constituent entity) of Russia? The last, auxiliary issue is: are there any geographical patterns in the spatial distribution of cities by accessibility of legal services? This study partially continues the work by member of Institute for the Rule of Law [@Moiseeva2016] by testing whether the top of cities by count of legal businesses per population has changed over time or hot, but adds additional questions and tries to answer them.

**Data.** Company-wise disaggregated geo-referenced dataset on small and medium-sized legal companies in Russia was used. It was produced with Python CLI tool *ru-smb-companies* developer by the author of this paper to extract and georeference open data from Russian state registry of small and medium-sized businesses (SMBs) distributed by Federal Tax Service. For the analysis, data on 2021 year was used. Only the businesses with the main activity code equal to 69.10 (*activity in the area of law* according to All-Russia Classifier of Economic Activity Kinds) were treated as legal companies, and that was in line with previously mentioned study by Institute for the Rule of Law. Also, sole entrepreneurs were excluded from the dataset, as well as companies who had declared zero revenue of expenditure. Organizations registered in rural area were not included in the dataset, because the aim of the research is to study relationships with respect to cities.

**Methods.**

## References

<div id="refs"></div>

### 2024-02-27 legal-companies-by-cities/Script.R

@@ -1,0 +1,133 @@
+library(dplyr)
+library(ggplot2)
+library(readr)
+library(sf)
+library(tidyr)
+
+# Data on firms, and cities lookup table
+panel <- read_csv("../../ru-smb-companies/legal/panel.csv")
+cities <- rbind(
+  read_csv("cities.csv"),
+  read_csv("cities_additional.csv")
+)
+regions_geo <- st_read("ru.geojson")
+
+lcc <- panel %>% 
+  filter(
+    kind == 1, 
+    year == 2021, 
+    settlement_type == "г",
+    revenue > 0,
+    expenditure > 0
+  ) %>% 
+  mutate(city = case_when(
+    region == "Москва" & is.na(settlement) ~ "Москва",
+    region == "Санкт-Петербург" & is.na(settlement) ~ "Санкт-Петербург",
+    TRUE ~ settlement
+  )) %>% 
+  select(region, city, oktmo, empl = employees_count) %>% 
+  replace_na(list(empl = 0)) %>% 
+  mutate(empl = replace(empl, empl == 0, 1)) %>% 
+  group_by(region, city, oktmo) %>% 
+  summarise(count = n(), empl = sum(empl), .groups = "drop")
+
+cp <- cities %>% 
+  mutate(city = coalesce(cities$city, cities$area, cities$region)) %>% 
+  select(city, oktmo, population, lat = geo_lat, lon = geo_lon)
+
+data <- left_join(lcc, cp, by = c("oktmo", "city")) %>% 
+  drop_na(count, empl, population) %>% 
+  mutate(
+    count_per_100k = 1e5 * count / population,
+    empl_per_100k = 1e5 * empl / population,
+    concentration = empl / sum(empl),
+    city_size_group = cut(
+      population,
+      breaks = c(0, 5e4, 1e5, 2.5e5, 5e5, 1e6, 1e9),
+      labels = c(
+        "Small (<50k)", "Medium (50–100k)", "Big (100–250k)",
+        "Large (250–500k)", "Extra-large (500k–1M)", 
+        "Millionare (>1M)")
+    )) %>% 
+  filter(
+    !(city %in% c("Иннополис", "Кировск", "Бронницы", "Красноармейск"))) %>% 
+  group_by(region) %>% 
+  mutate(
+    city_region_rank = row_number(-population),
+    is_largest_in_region = city_region_rank == 1) %>% 
+  ungroup()
+
+ggplot(data, aes(
[…]

### 2024-02-28 legal-companies-by-cities/Paper.Rmd
Continue working on paper: add methods, results and discussion, and conclusion

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(
+  echo = FALSE,
+  message = FALSE,
+  warning = FALSE,
+  dpi = 300,
+  fig.id = FALSE,
+  fig.width = 6.4,
+  fig.height = 5)
+```
+
+```{r imports, include=FALSE, cache = TRUE}
+source("Script.R")
 ```
@@
 Organizations registered in rural area were not included in the dataset, because the aim of the research is to study relationships with respect to cities.
+During the exploratory data analysis, four cities (namely, Innopolis, Kirovsk, Bronnytsy, Krasnoarmeisk) were marked as outliers and excluded from the data.
 
-**Methods.**
+**Methods.** Concentration of legal companies in cities was estimated as the number of employees of legal companies in a given city divided by the total number of employees in legal companies in all the cities.
+Accessibility of legal services was measured as the number of employees of legal companies (here, the are also called legal workers) in the city divided by the city's population.
+Both these metrics are quite rough, but give a basic understanding of the underlying phenomena.
+To categorize the cities by population, the classification of Lappo [@Lappo1997] was used.
+Data visualization, including cartography, was used as the main method.
+In addition, the hypothesis of Zipf law distribution was tested with linear regression model.
+Technically, the analysis was performed with R programming language v.
+4.1.2.
+The source code for this paper is available at the public repository: https://github.com/PavelSyomin/ru-smb-companies-papers/tree/main/legal-companies-by-cities
+
+**Results and discussion.** The main result of the study, that is, the relationship between the concentration of legal companies and accessibility of legal services, is shown on Fig.
+1.
+Both axes of the plot are log-10 scaled to overcome the skewed distribution of underlying data.
+The relationship between two variables is close to linear, but the variance is high.
+However, if data is grouped by city size, the variance in each group decreases, and the larger the city, the smaller the variance of concentration.
+Also, the larger the city, the bigger the concentration.
+However, in terms of accessibility the picture is not as robust, because despite the general trend, the variance of accessibility in each population group is extremely high, and some small or medium cities have the better accessibility than large or even millionaire.
+It can be adder, though, that all the millionaire cities have relatively big number of legal workers per population, thus having the population over 1M people is always associated with high availability of legal services.
+Also, the largest city in every region usually have both high concentration of legal companies and high accessibility of legal services, despite the presence of several counter-examples.
+At the right top of the plot, there are two notable outliers, namely Moscow and St.
+Petersburg, that have the highest concentration and good accessibility.
+Their relative position (compared to the main line of the points cloud) implies that there is some sort of “accessibility saturation”, that is, the point where the marginal growth of concentration has no longer effect on accessibility.
+To sum it up, it may be said that high concentration and population, as well as the relative size inside the region, is somehow associated with accessibility, but at the same time there are various paths to high level of legal professionals per population, and some cities reach the high levels of accessibility without huge concentration.
+
+```{r fig-1-concentration-and-accessibility, fig.cap="", fig.height=5}
+concentration_and_accessibility_plot
+```
+
+The relationship between the count of legal companies and the population and between the count of legal workers and the population is close to linear (Fig.
+2).
+In fact, the error is high if population and counts are small, but if population is high, the error quickly decreases.
+
+```{r fig-2-counts-and-population, fig.cap=""}
+counts_and_population_plot
+```
+
+The distribution of cities by the count of legal companies and workers mostly complies with Zipf law (Fig.
+3).
+The relevant coefficients of determination (R-squared) are above 0.95 in both cases, and the coefficients of linear models are statistically significant with high confidence, and the slope is 0.5–0.6.
+There are some deviations from the rank-size log-linear trend for top-rank cities.
+
+```{r fig-3-rank-size, fig.cap=""}
+rank_size_plot
+```
+
+The spatial distribution of cities by legal services accessibility (Fig.
+4) proves that regional centers have high relative number of legal workers.
+Moscow and its satellite cities also have high accessibility, as well as St.
+Petersburg.
+Several other point of high availability of legal services can be noted, e.
+g.
+Krasnodar and Stavropol at the South, Yekaterinburg in the Urals, Tuymen in Western Siberia.
+There are regions with low accessibility of legal services, for example, bordering regions in the South of Siberia, Komi and Kalmyk republics.
+
+```{r fig-4-map, fig.cap=""}
+spatial_plot
+```
+
+**Conclusion.** The number of legal companies in Russian cities and the number of employees of these companies is positively linearly related to cities population.
+The distribution of cities by the count of legal companies and workers of these companies follows the Zipf (rank-size) law.
+The largest cities of regions usually have relatively high accessibility of legal services, and the same is true for the cities with population above 1 million.
+The concentration of legal companies in cities has a positive linear relationship with the number of lawyers per population, but the variance of variables is big.
+Empirical data suggests several paths for reaching high levels of legal services availability, where increasing concentration of legal companies in a city is just one of the options, and cities with relatively small concentration can achieve the level of accessibility similar to the largest ones.
+The growth of legal services availability probably saturates after some level of concentration.
 

### 2024-02-28 legal-companies-by-cities/Script.R

@@ -57,7 +57,7 @@ data <- left_join(lcc, cp, by = c("oktmo", "city")) %>%
     is_largest_in_region = city_region_rank == 1) %>% 
   ungroup()
 
-ggplot(data, aes(
+concentration_and_accessibility_plot <- ggplot(data, aes(
     x = empl_per_100k, y = concentration, 
     color = city_size_group, shape = is_largest_in_region)) +
   geom_point(size = 2) +
@@ -75,8 +75,9 @@ ggplot(data, aes(
     legend.position = c(.01, .99),
     legend.justification = c(0, 1),
     panel.grid.minor.y = element_blank())
+concentration_and_accessibility_plot
 
-data %>% 
+counts_and_population_plot <- data %>% 
   pivot_longer(
     cols = c("count", "empl"),
     names_to = "var",
@@ -91,8 +92,9 @@ data %>%
     labeller = as_labeller(
       c("count" = "Companies", "empl" = "Employees"))) +
   theme_bw(base_size = 14, base_family = "Times New Roman")
+counts_and_population_plot
 
-data %>% 
+rank_size_plot <- data %>% 
   pivot_longer(
     cols = c("count", "empl"),
     names_to = "var",
@@ -110,6 +112,7 @@ data %>%
   facet_wrap(~var, ncol = 2, labeller = as_labeller(
     c("count" = "Companies", "empl" = "Employees"))) +
   theme_bw(base_size = 14, base_family = "Times New Roman")
+rank_size_plot
 
 rank_empl_lm <- data %>% 
   mutate(rank = row_number(-empl)) %>% 
@@ -122,7 +125,7 @@ rank_count_lm <- data %>%
 summary(rank_count_lm)
 
 ru_crs <- st_crs("+proj=aea +lat_0=0 +lon_0=100 +lat_1=68 +lat_2=44 +x_0=0 +y_0=0 +ellps=krass +towgs84=28,-130,-95,0,0,0,0 +units=m +no_defs")
-data %>% 
+spatial_plot <- data %>% 
   st_as_sf(coords = c("lon", "lat"), crs = "EPSG:4326") %>% 
   ggplot() +
   geom_sf(data = regions_geo, color = "gray50", fill = "white") +
@@ -131,3 +134,4 @@ data %>%
   scale_color_binned(name = "Lawyers per 100k", n.breaks = 4, low = "#dadaeb", high = "#3f007d") +
[…]

### 2024-02-29 legal-companies-mapping/Paper.Rmd
First draft of paper about mapping

---
output:
  odt_document:
    reference_odt: assets/reference.odt
bibliography: assets/biblio_en.bib
csl: assets/citations.csl
link-bibliography: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE, 
  message = FALSE, 
  echo = FALSE, 
  fig.width = 6.5, 
  fig.height = 5,
  fig.id = FALSE,
  dpi = 300)
options(knitr.kable.NA = "—")
```

```{r imports, cache = TRUE, include = FALSE}
source("helpers/Plots.R")
```

УДК 911.3, 519.688

<div custom-style="Author">

**P. O. Syomin**

Perm State University, Perm, Russia

E-mail: ntsp\@ya.ru

</div>

<br>

<div custom-style="Title">

Mapping of Russian Small and Medium-Sized Businesses Using Tax Service Open Data

</div>

<br>

<div custom-style="Abstract">

**Keywords:** open data, spatial data, mapping, geocoding, small&medium business, dataset, big data, Federal Tax Service

</div>

<br>

<div custom-style="Author">

**П. О. Сёмин**

Пермский государственный национальный исследовательский университет, Пермь, Россия

E-mail: ntsp\@ya.ru

</div>

<br>

<div custom-style="Title">

Картографирование российских малых и средних предприятий по открытым данным налоговой службы

</div>

<br>

<div custom-style="Abstract">

В статье представлена технология для создания геопривязанного табличного набора данных о субъектах малого и среднего предпринимательства (МСП) в России на основе открытых данных Федеральной налоговой службы (ФНС России). Получаемый набор данных охватывает всю территорию России за 2016–2023 гг. Он представляет собой полный список субъектов МСП и содержит информацию о регистрационных номерах, категории, имени или наименовании, месте жительства или местонахождении, виде деятельности, доходах, расходах и численности работников. В пространственном отношении набор детализирован до населённых пунктов с указанием географических координат и кода муниципального образования (ОКТМО); во временном отношении — до месяцев или лет в зависимости от конкретного признака; в отношении вида деятельности — до отдельных кодов общероссийского классификатора (ОКВЭД). В отличие от классической data paper, статья не описывает итоговые данные как таковые, а излагает воспроизводимую методику их получения и демонстрирует её использование для получения набора данных о фирмах в сфере сельского, лесного хозяйства и рыболовства. С помощью предложенного алгоритма и его реализации на языке Python исследователь может самостоятельно обработать исходные данные и получить актуальную версию набора данных, адаптированную для его задач. Алгоритм оптимизирован так, чтобы для его выполнения было достаточно компьютера средней конфигурации, несмотря на сравнительно большой (около 2 Тб в несжатом виде) объём исходных данных. Получаемый с помощью технологии набор данных может применяться как непосредственно для пространственного анализа МСП в России, так и в качестве открытого, бесплатного и относительно репрезентативного источника сведений об экономике России в разрезе конкретных организаций, позволяя отчасти заменить коммерческие системы бизнес-аналитики наподобие Spark и Контур.Фокус. Кроме того, благодаря координатам набор данных может без дополнительной обработки использоваться для картографирования, а наличие кодов ОКТМО позволяет объединять его с официальными статистическими сведениями.

**Ключевые слова:** открытые данные, пространственные данные, картографирование, геокодирование, малое и среднее предпринимательство, набор данных, большие данные, Федеральная налоговая служба.

**Introduction.** Contemporary mapping solutions used for scientific and other purposes usually require well-structured easy-to-use spatial datasets. Here, “spatial” typically means “with coordinates of features provided in one of popular reference systems”. However, getting such the datasets may become a challenging task which can significantly slow down the research. The public request for high-quality spatial data has stimulated many governments to launch the state programs enabling access to it [@Jozefowicz2020]. In Russia, governmental data is sometimes distributed with coordinates, but in general, its availability is low. Despite the fact that many publicly available datasets include geographical reference of some sort, it is hard to use them in digital mapping solutions, because the spatial aspects are represented by human-readible addresses rather than by machine-readible geographical coordinates. To convert human-readible addresses into geographical coordinates, a procedure called geocoding is used. There are several geocoding service providers in Russia[^1], and their solutions have considerably good precision. However, almost all good geocoding services can be used only by calling special online application programming interfaces (APIs) which are available with a paid subscription. Thus, a researcher trying to create a map based on available datasets usually have to overcome two problems, one of them is the need to write a computer program for geocoding via online API, and another is the requirement to pay for this API. Both these issues can be solved with some effort, but nevertheless, they distract from the research itself, and anyway, a better solution is desirable.

Among various govenrmental datasets which may be of high interest for the researchers, the data about small and medium-sized businesses (SMBs) attracts attention. The state SMB registry was created in the end of 2015 according to Art. 4.1 of Federal Law of July 24, 2007 No 207-FZ “On the development of small and medium-sized businesses in the Russian Federation”. It is operated by the Federal Tax Service (FTS) of Russia and contains information about all Russian SMBs. The registry is updated automatically, that is, no action from the business entity is required, thus it can be treated as the most complete list of all commercial organizations and individual entrepreneurs that meet several requirements for the number of employees and yearly income. It can not be used as a complete list of all commercial companies in the country, because the complete list of all registered organizations is contained in another state registry, namely Unified State Registry of Legal Entities. However, in comparison to the latter, the full dumps of SMB registry are distributed as open data free of charge and with no restrictions to use, while the Unified State Registry is accessible only via a paid subscription with several use restrictions or via online access to the data on particular companies. Thus, the SMB registry seems to be the most complete openly available collection of Russian commercial companies. The necessity of company-level data for geographical research has been noted by many scientists (see literature review below), and the open collection of such data seems to have high potential value. For example, it can be used to map companies with high spatial resolution (up to cities). However, the dumps of SMB registry do not contain coordinates — only human-readible addresses are included. Also, these dumps are of large size (hundreds of gigabytes if archived and some terabytes if unpacked) and have complex structure. To make them usable by a regular developer, a simplification, geocoding and transformation is required.

**Objective.** In this paper, an application that helps to build a ready-to-mapping spatial dataset of Russian small and medium-sized businesses is featured and discussed. The source data is open dumps of SMB registry. A spatial dataset contains geographical coordinates of each entity with spatial resolution up to a settlement (urban or rural). Application is a Python command-line (CLI) tool with easy-to-use interface that can be run on a medium personal computer without the need to use special tools for big data processing. In the paper, the application is first briefly described, and after that, a demonstration of its use for mapping the Russian legal companies is performed.

**Literature review.** A typical source of data for studies of social geography in Russia is the resources of the Federal State Statistics Service (Rosstat). These include, in particular, the data of official statistics, censuses and statistical surveys posted on the official website of this body, the database of indicators of municipalities (BDMO) and the unified interdepartmental information and statistical system (EMISS). There are many examples of their use in geography: O. V. Kuznetsova, when studying the structure of the economy of Russian regions, uses statistics of gross regional product and the volume of goods shipped [@Kuznetsova2018]; G. M. Fedorov, in his analysis of the economy of the Russian regions in the Baltic, relies on data from Rosstat and the Federal Customs Service [@Fedorov2022]; S. G. Safronov studies the tertiary sector of the Russian economy on the basis of data from Eurostat, Rosstat and the Federal Tax Service [@Safronov2021]; Yu. V. Petrov, with the help of regional, municipal statistics and data from other authorities, formulates proposals for the socio-economic development of the south of the Tyumen region [@Petrov2021]; M. A. Sarancha assesses the development of the hotel and restaurant business, using the results of Rosstat’s continuous monitoring of small and medium-sized businesses [@Sarancha2014]. Rosstat information is also used to create derived data sets, as in the work of I. Yu. Peker [@Peker2022]. Thus, most socio-geographical studies cannot do without using Rosstat data as the only source of information or as one of the sources.

However, in some cases, Rosstat data is insufficient. As a result, there is a need for alternative sources of information that either complement official statistics or replace them. The reasons for this vary. In some cases, alternative sources of information are used to verify or correct Rosstat. For example, R. A. Babkin, based on data from cellular operators, proves that the official population of the Moscow agglomeration is underestimated by approximately 10% due to the error and inertia of statistical accounting and the “diocomy” of a significant part of Muscovites [@Babkin2020]. Another, more common reason is the lack of necessary information. For example, K.V. Rostislav, analyzing the impact of location on the profitability of enterprises, takes into account only organizations with fewer than 250 employees and uses open data from the Federal Tax Service of Russia to check this condition, since Rosstat does not provide such information [@Rostislav2021]. S. S. Lachininsky and co-authors estimate the spread of urban areas of St. Petersburg based on Landsat satellite images, since it is impossible to make the necessary calculations based on official statistics [@Lachininsky2023]. The advantages of alternative data sources include increased accuracy and timeliness [@Vorobyev2021], higher spatial and temporal resolution [@Babkin2021], the possibility of accurate and detailed analysis at the micro level [@Bochkaryov2017], while noting the untimeliness, incompleteness and potential unreliability of official statistics [@Kuzminov2021]. Alternative data sources provide access to the general population, as a result of which “the need for the use of selective, aggregated and indirect” sources of information is eliminated [@Blanutsa2017]. When using machine learning methods, which is also becoming relevant in modern geography, the sample size, disaggregation, and number of features in the data are important, which also encourages turning to alternative sources [@Rostislav2020]. Thus, although traditional statistical information is still important in socio-geographical research, and sometimes irreplaceable, there is a growing need for other information resources. The request from researchers is to increase the volume, speed of provision, accuracy, spatial and temporal resolution, expand the set of available indicators, as well as the possibility of validating statistical information. It manifests itself both in terms of traditional scientific methods and methods related to machine learning, although it is more relevant specifically for the latter, which are the most sensitive to the volume of data, their disaggregation, detail, and convenient presentation formats.

One of the alternative sources of information for social geographers is administrative data from the Federal Tax Service of Russia. They are published on the official website of this authority or related resources. Geographical works use, in particular, information on the income of individuals [@Petrov2021], on tax revenues by federal subjects or types of economic activity [@Rostislav2020; @Safronov2021], on the average number of employees of organizations [@Rostislav2021], on the number of registered residential premises [@Lachininsky2021], on the number of small and medium-sized enterprises and measures to support them [@Gumenyuk2022]. The attention to the data published by the Federal Tax Service of Russia is apparently due to the multifunctionality of this government body, due to which it collects a huge amount of information about different aspects of the Russian economy: business entities, property, income, expenses. Typically, the Federal Tax Service of Russia, like Rosstat, publishes summary (aggregated) information. However, there are several data sets that contain disaggregated (raw) information. Among them, the most attention is drawn to the register of small and medium-sized businesses (SME register), which is regularly and completely posted on the service’s website and contains information about specific organizations and individual entrepreneurs.

For some economic-geographical and related studies, information detailed down to the level of specific firms is important. Such information includes, for example, location, revenue, expenses, number of employees, name and other characteristics. The specified data can be used directly, as in the works of K. V. Rostislav [@Rostislav2021], E. A. Kolomak and A. I. Sherubneva [@Kolomak2023], who analyze agglomeration effects based on enterprise indicators. In addition, on their basis, statistics with arbitrary spatial or temporal grouping can be considered, including those that are impossible, difficult or inconvenient to extract from Rosstat data. For example, I. S. Gumenyuk operates with calculations of the absolute and relative number of small and medium-sized enterprises in the municipalities of the Kaliningrad region [@Gumenyuk2022]; M. A. Sarancha includes the number of enterprises in the integral index of the level of development of small businesses in the territory of the municipality [@Sarancha2014]; and M.A. Makushin and co-authors use data on enterprises in the cities of BAM to assess the structure of industry [@Makushin2023]. Information about specific firms acquires additional importance in city-scale studies [@Lachininsky2021; @Nikonorov2021; @Kozhevnikov2023], when studying narrow sectors of the economy that do not appear separately in statistics [@Moiseeva2016; @Pankratov2023], when analyzing a single industry [@Kuzminov2021].

There is no generally accepted source of geo-referenced information about companies in Russia. Researchers and applied specialists usually rely on data from commercial [@Lachininsky2021] or open geoinformation services, business intelligence systems [@Moiseeva2016; @Kuzminov2021], as well as the above-mentioned SME register [@Gumenyuk2022]. Typically, using these methods poses two problems: the data is provided for a fee, or considerable effort is required to bring it into a format suitable for analysis. Business intelligence systems such as Spark, Kontur.Focus and Ruslana contain a lot of varied information, but are paid. Downloading data from private geographic information systems like 2GIS and Yandex.Maps also costs money, and in addition, there is no convenient interface for it. Open platforms like OpenStreetMap or the SME registry are free, but the downloads from them are large and complex, so the researcher has to spend a lot of effort on the technical side of the work, involve third-party specialists, or refuse to use such a data source. Formally, a complete list of Russian organizations and individual entrepreneurs is contained in two state registers (respectively, the Unified State Register of Legal Entities and the Unified State Register of Individual Entrepreneurs), however, neither of them is distributed in full - you can only extract individual records manually through the Federal Tax Service web service Russia or automatically for a fee through a software interface, which does not provide access to the general population of companies and is unlikely to allow constructing a representative sample without excessive costs and effort. Thus, the question of a convenient tool that would allow free and legal obtaining the most complete set of geo-referenced data on Russian companies is relevant.

**Source data.** For the mapping, one main data source, two additional data sources, and two lookup tables are used. The main data source is SMB registry open data dumps[^1]. They are published with monthly interval starting from August 2016. Each dump contains the full registry for the date of publication, so a substantial portion of the data is duplicated. Technically, a dump is a ZIP archive with thousands of XML files. Each XML file contains information about 900 organizations or individuals. The data attributes of XML file include, for example, name of the business, its registration address (region, district, city, settlement), activity codes according to All-Russia Classifier of Economic Activity Kinds.

Additional data sources are two FTS open datasets. The first of them is called “Information about revenue and expenditure of organizations by their accounting (financial) reporting”[^3], and the second is named “Information about average list count of employees of organizations”[^4]. Both of them are available starting from approximately 2019 and are updated irregularly, but at least once a year. Technically, they are also ZIP archives of XML files, where each archive contains data for a particular date, and each XML file lists 900 organizations. Individuals are not included in these datasets due to personal data issues.

[^3]: https://www.nalog.gov.ru/opendata/7707329152-revexp/
[^4]: https://www.nalog.gov.ru/opendata/7707329152-sshr2019/

The lookup tables used for the mapping are “Settlements of Russia: population and geographic coordinates” dataset provided by “Infrastructure of the scientific data” project»[^5], and “Cities of Russia” dataset published by commercial company HFLabs[^6]. Both of them contain addresses and geographical coordinates of Russian cities and settlements. In addition, the author created a tiny supplement to the “Cities of Russia” dataset with information about missing cities extracted manually from Federal Address Information System (FIAS)[^7].

[^5]: https://data.rcsi.science/data-catalog/datasets/160/
[^6]: https://github.com/hflabs/city
[^7]: https://fias.ru

**Methods.** A Python command-line (CLI) application was developed to process the source data. Technically, it is based on Python standard library and several third-party packages, including fast XML processing library called lxml, iconic tabular data processing tool named Pandas, popular big data framework Apache Spark, and easy-to-use CLI framework Typer. This application was intended to be published as open-source tool for other researchers, so its source code with brief documentation is available at author's personal GitHub account[^8]. Readers who are familiar with Python programming language may go to the repository and look at the code thoroughly. Here, a top-level structure of the application is described, and the data processing flow for the mapping is almost the same.

[^8]:

*Step 1* is data download. The source ZIP archives were downloaded from FTS website and stored locally. *Step 2* is data extraction and filtering. The data was extracted from ZIP archives and stored in CSV tables, so that one ZIP archive was transformed into 1 CSV table. A lot of unused data attributes were dropped, so the resulting size of CSV tables is far smaller than the size of original archives. Also, the filtering by activity code was performed on this step to select data only about legal companies. According to the classifier, the main activity code equal to 69.10 was used as a filtering clause. The application itself allows to filter by other codes or groups of classifier or disable filtering at all, thus providing the opportunity to generate various slices of companies (e.g. agricultural, forestry, health services, heavy industries etc) with various structural detalization (i. e. up to groups or particular low-level codes of classifier). *Step 3* is data aggregation: due to the specifics of data publication, a lot of information in archives and corresponding CSV tables is duplicated, and thus duplicates have to be dropped. This step dropped the duplicates and thus reduced the volume of data even further. *Step 4* is geocoding. The addresses were converted to geographical coordinates using the look-up tables, and ≈98% of all addresses were successfully matched. Because the address in the source data is detalized up to settlements, the geographical coordinates refer to cities or settlements, and this is the lower bound of spatial resolution of the map. Optional *step 5* combines SMB registry data with data on revenue, expenditure and employees and transforms it to produce a yearly panel table. Here, it was not performed, because the geocoded dataset was sufficient for mapping.

After these preprocessing, the tabular dataset was filtered by year (2021 was chosen). The maps were plotted with a script written on R programming language with additional packages, including, in particular, ggplot2 and sf. The source code for the paper is available at the separate repository[^9].

[^9]:

**Results.** The spatial resolution of the dataset is up to cities or settlements, but regions and their subdivisions are also included, so the aggregation and mapping are also possible. Figure 1 shows the count of legal companies by region. Some spatial patterns are clearly seen from this visaulization, for instance, the high concentration of legal firms in the capital and the surrounding region, the “Urals-Volga-Krasnodar” belt, and small values in the Central economic region.

```{r fig-1-settlements-russia, fig.cap = "Рис. 1. Число субъектов МСП по населённым пунктам России и преобладающий вид деятельности. Размер кругов соответствует числу субъектов МСП, цвет — одной из трёх групп видов деятельности по классификатору ОКВЭД (см. легенду). Показаны только населённые пункты, в которых зарегистрировано более 50 фирм"}
regions_plot
```

Figure 2 shows the results of the mapping of legal companies by settlements. This is high-resolution map which gives a quick overview of legal business in Russia. The spatial distribution a bit resembles the region-scale map, but city-level data allows to notice other tendencies, e. g. concentration of businesses in regional centres. Also, the high concentration in Moscow and surroundings is clear, and the similar pattern for St. Petersburg is visible.

```{r fig-2-settlements-russia, fig.cap = "Рис. 2. Число субъектов МСП по населённым пунктам России и преобладающий вид деятельности. Размер кругов соответствует числу субъектов МСП, цвет — одной из трёх групп видов деятельности по классификатору ОКВЭД (см. легенду). Показаны только населённые пункты, в которых зарегистрировано более 50 фирм"}
settlements_map
```

The similar map may be drawn for a particular region, e. g. for Sverdlovsk oblast, as in the Figure 3. A pattern of high concentration is apparent again.

```{r fig-3-settlements_svr, fig.cap = "Рис. 3. Число  субъектов МСП в населённых пунктах Свердловской области и их преобладающий вид деятельности. Размер кругов соответствует числу субъектов МСП, цвет — одной из трёх групп видов деятельности по классификатору ОКВЭД (см. легенду). Показаны все населённые пункты без ограничения по количеству зарегистрированных фирм"}
settlements_map_svr
```

The presence of additional data about financial metrics and count of employees allows us to draw more maps. On Figure 4, the spatial distribution of legal companies profit is displayed, and on Figure 5, the similar distribution with respect to tne number of employees is shown. Both maps include only companies with non-zero profit or number of employees. The maps looks like the counts map (see Figure 2), but exact spatial patterns vary.

```{r fig-4-profit-map, fig.cap = "Рис. 4. Распределение числа зарегистрированных субъектов МСП, их суммарной прибыли и суммарной численности работников в демонстрационном наборе в 2021 г. по регионам (субъектам федерации)"}
settlements_map_profit
```

<br>

```{r fig-5-employees-map, fig.cap = "Рис. 5. Распределение числа зарегистрированных субъектов МСП, их суммарной прибыли и суммарной численности работников в демонстрационном наборе в 2021 г. по регионам (субъектам федерации)"}
settlements_map_empl
```

**Conclusion.**

# References

<div id="refs"></div>

<br>

# About the author

Pavel Olegovich Syomin, PhD Student, Department of Social and Economic Geography, Perm State University, Perm, Russia; Software Developer, AgroSoftware LLC, Moscow, Russia.

E-mail: ntsp\@ya.ru

ORCID: 0000-0002-4015-9206

# Сведения об авторе

Сёмин Павел Олегович, аспирант кафедры социально-экономической географии Пермского государственного национального исследовательского университета, г. Пермь, Россия; разработчик в ООО «АгроСофтвер», г. Москва, Россия.

E-mail: ntsp\@ya.ru

ORCID: 0000-0002-4015-9206

### 2024-02-29 legal-companies-mapping/helpers/Plots.R

@@ -1,0 +1,124 @@
+library(dplyr)
+library(forcats)
+library(ggplot2)
+library(readr)
+library(scales)
+library(sf)
+library(tidyr)
+
+# Data
+smb_data <- read_csv("assets/smb.csv")
+revexp_data <- read_csv("assets/revexp.csv")
+empl_data <- read_csv("assets/empl.csv")
+
+# Maps
+regions_boundaries <- st_read("assets/ru.geojson")
+ru_crs <- st_crs("+proj=aea +lat_0=0 +lon_0=100 +lat_1=68 +lat_2=44 +x_0=0 +y_0=0 +ellps=krass +towgs84=28,-130,-95,0,0,0,0 +units=m +no_defs")
+regions <- read_csv("assets/regions.csv")
+regions <- regions_boundaries %>% 
+  left_join(regions, by = c("shapeISO" = "iso_code")) %>% 
+  select(name, name_en = shapeName)
+ru_svr <- st_read("assets/ru_svr.geojson")
+
+# Regional distribution
+regions_plot <- smb_data %>%
+  filter(
+    start_date <= "2021-12-31",
+    end_date >= "2021-12-31") %>% 
+  count(region) %>% 
+  left_join(regions, by = c("region" = "name")) %>% 
+  st_as_sf() %>% 
+  ggplot() +
+  geom_sf(aes(fill = n), size = .1) +
+  coord_sf(crs = ru_crs) +
+  scale_fill_distiller(
+    name = "Count of legal companies",
+    palette = "YlGn",
+    direction = 1) +
+  theme_bw(base_size = 11, base_family = "Times New Roman") +
+  theme(legend.position = "bottom")
+regions_plot
+
+settlements_df <- smb_data %>%
+  filter(
+    start_date <= "2021-12-31",
+    end_date >= "2021-12-31"
+  ) %>% 
+  count(region, area, settlement, lat, lon) %>% 
+  drop_na(lat, lon) %>% 
+  st_as_sf(coords = c("lon", "lat"), crs = 4326)
+
+settlements_map <- regions %>% ggplot() +
+  geom_sf(size = .1) +
+  geom_sf(data = settlements_df, aes(color = log10(n)), size = .1) +
+  coord_sf(crs = ru_crs) +
+  scale_color_distiller(
+    name = "Count of legal companies (log10)",
+    palette = "YlGn",
+    direction = 1) +
+  theme_bw(base_size = 11, base_family = "Times New Roman") +
[…]

### 2024-02-29 legal-companies-mapping/Paper.Rmd
Finish the paper about mapping

@@
   echo = FALSE, 
-  fig.width = 6.5, 
-  fig.height = 5,
+  fig.width = 5.7, 
+  fig.height = 4,
   fig.id = FALSE,
   dpi = 300)
-options(knitr.kable.NA = "—")
 ```
@@
 
-Mapping of Russian Small and Medium-Sized Businesses Using Tax Service Open Data
+Mapping Russian Small and Medium-Sized Businesses Using Tax Service Open Data
 
@@
 
-<div custom-style="Abstract">
+*This paper presents a flexible and reproducible workflow for creating highly detailed maps of small and medium-sized businesses (SMBs) in Russia.
+Currently, there is a lack of data sources for mapping the Russian economy, and existing official datasets often provide location information in human-readable addresses rather than machine-readable geographical coordinates.
+The proposed set of tools aims to solve this problem and simplify the mapping process.
+The study utilizes three open datasets published by the Federal Tax Service, including open dumps of SMB registry, as its data source.
+To transform the source data into tables and normalize and geocode the addresses, a Python CLI application was developed.
+Despite the large volume of the initial data archives, the application can run on a regular modern personal computer without the need for cloud computing resources.
+The workflow's usability is demonstrated by mapping a small subset of data, specifically legal companies in Russia.*
 
-**Keywords:** open data, spatial data, mapping, geocoding, small&medium business, dataset, big data, Federal Tax Service
-
-</div>
+***Keywords:*** *open data, spatial data, mapping, geocoding, small and medium-sized business, dataset, big data, Federal Tax Service.*
 
@@
 
-<div custom-style="Abstract">
+*В статье представлена гибкая и воспроизводимая методика картографирования малых и средних предприятий (МСП) в России.
+В настоящий момент не хватает качественных наборов данных для визуализации характеристик экономики страны на карте.
+Существующие официальные источники содержат информацию о местоположении тех или иных объектов в виде человекочитаемых адресов, в то время как ГИС-платформы требуют машиночитаемые пространственные данные с координатами.
+Представленный в работе набор инструментов — это попытка решить проблему и упростить подготовку данных для картографирования.
+В качестве источника сведений используются три набора открытых данных Федеральной налоговой службы, в частности, выгрузки реестра МСП.
+Для их обработки, нормализации адресов и геокодирования было разработано специальное приложение командной строки на языке программирования Python.
+Несмотря на большой объём исходных данных, приложение работает на среднем современном компьютере и не требует обращения к облачным вычислительным мощностям.
+Работоспособность инструментов продемонстрирована путём картографирования небольшой части исходного набора данных — сведений о юридических фирмах.*
 
-В статье представлена технология для создания геопривязанного табличного набора данных о субъектах малого и среднего предпринимательства (МСП) в России на основе открытых данных Федеральной налоговой службы (ФНС России).
-Получаемый набор данных охватывает всю территорию России за 2016–2023 гг.
-Он представляет собой полный список субъектов МСП и содержит информацию о регистрационных номерах, категории, имени или наименовании, месте жительства или местонахождении, виде деятельности, доходах, расходах и численности работников.
-В пространственном отношении набор детализирован до населённых пунктов с указанием географических координат и кода муниципального образования (ОКТМО); во временном отношении — до месяцев или лет в зависимости от конкретного признака; в отношении вида деятельности — до отдельных кодов общероссийского классификатора (ОКВЭД).
-В отличие от классической data paper, статья не описывает итоговые данные как таковые, а излагает воспроизводимую методику их получения и демонстрирует её использование для получения набора данных о фирмах в сфере сельского, лесного хозяйства и рыболовства.
-С помощью предложенного алгоритма и его реализации на языке Python исследователь может самостоятельно обработать исходные данные и получить актуальную версию набора данных, адаптированную для его задач.
-Алгоритм оптимизирован так, чтобы для его выполнения было достаточно компьютера средней конфигурации, несмотря на сравнительно большой (около 2 Тб в несжатом виде) объём исходных данных.
-Получаемый с помощью технологии набор данных может применяться как непосредственно для пространственного анализа МСП в России, так и в качестве открытого, бесплатного и относительно репрезентативного источника сведений об экономике России в разрезе конкретных организаций, позволяя отчасти заменить коммерческие системы бизнес-аналитики наподобие Spark и Контур.Фокус.
-Кроме того, благодаря координатам набор данных может без дополнительной обработки использоваться для картографирования, а наличие кодов ОКТМО позволяет объединять его с официальными статистическими сведениями.
+***Ключевые слова:*** *открытые данные, пространственные данные, картографирование, геокодирование, малое и среднее предпринимательство, набор данных, большие данные, Федеральная налоговая служба.*
 
-**Ключевые слова:** открытые данные, пространственные данные, картографирование, геокодирование, малое и среднее предпринимательство, набор данных, большие данные, Федеральная налоговая служба.
+<br>
 
-**Introduction.** Contemporary mapping solutions used for scientific and other purposes usually require well-structured easy-to-use spatial datasets.
-Here, “spatial” typically means “with coordinates of features provided in one of popular reference systems”.
-However, getting such the datasets may become a challenging task which can significantly slow down the research.
-The public request for high-quality spatial data has stimulated many governments to launch the state programs enabling access to it [@Jozefowicz2020].
-In Russia, governmental data is sometimes distributed with coordinates, but in general, its availability is low.
-Despite the fact that many publicly available datasets include geographical reference of some sort, it is hard to use them in digital mapping solutions, because the spatial aspects are represented by human-readible addresses rather than by machine-readible geographical coordinates.
-To convert human-readible addresses into geographical coordinates, a procedure called geocoding is used.
-There are several geocoding service providers in Russia[^1], and their solutions have considerably good precision.
-However, almost all good geocoding services can be used only by calling special online application programming interfaces (APIs) which are available with a paid subscription.
-Thus, a researcher trying to create a map based on available datasets usually have to overcome two problems, one of them is the need to write a computer program for geocoding via online API, and another is the requirement to pay for this API.
-Both these issues can be solved with some effort, but nevertheless, they distract from the research itself, and anyway, a better solution is desirable.
+**Introduction.** Contemporary mapping solutions used for scientific and other purposes usually require sources of well-structured spatial data.
+“Spatial” refers to data that is provided in an appropriate format with feature coordinates provided in one of the standardized coordinate reference systems.
+Obtaining datasets suitable for the mapping can be a challenging task that may significantly slow down research or increase its cost.
+The public demand for high-quality spatial data has led many governments to create geoportals or to facilitate the distribution of such data through other means [@Jozefowicz2020].
 
-Among various govenrmental datasets which may be of high interest for the researchers, the data about small and medium-sized businesses (SMBs) attracts attention.
-The state SMB registry was created in the end of 2015 according to Art.
+In Russia, official data is occasionally published in machine-readable formats with coordinates, but generally, open spatial data is not widely available.
+This lack of availability may pose a challenge for mapping various aspects of society, including the economy, demography, and culture.
+Datasets often include a geographical reference in the form of a human-readable address.
+However, GIS software requires machine-readable geographical coordinates.
+To convert addresses into coordinates, a process called geocoding is used.
+Several geocoders have relatively high accuracy for Russian addresses[^1].
+However, most of these tools are only available through paid online APIs or have legal terms that prohibit the storage and use of geocoding results.
+Therefore, researchers attempting to map official data often have to pay for a geocoding API and write a program to convert addresses to coordinates.
+It would be much better to have a user-friendly tool to assist in mapping valuable open datasets.
+
+[^1]: Examples are Yandex or Google geocoders, DaData geocoding service, and OpenStreetMap Nominatim tool.
+
+The data from the small and medium-sized businesses (SMB) registry, among various official datasets published by Russian state bodies, is of high value for researchers in economic geography.
+The state SMB registry was created at the end of 2015 in accordance with Art.
 4.1 of Federal Law of July 24, 2007 No 207-FZ “On the development of small and medium-sized businesses in the Russian Federation”.
-It is operated by the Federal Tax Service (FTS) of Russia and contains information about all Russian SMBs.
-The registry is updated automatically, that is, no action from the business entity is required, thus it can be treated as the most complete list of all commercial organizations and individual entrepreneurs that meet several requirements for the number of employees and yearly income.
-It can not be used as a complete list of all commercial companies in the country, because the complete list of all registered organizations is contained in another state registry, namely Unified State Registry of Legal Entities.
-However, in comparison to the latter, the full dumps of SMB registry are distributed as open data free of charge and with no restrictions to use, while the Unified State Registry is accessible only via a paid subscription with several use restrictions or via online access to the data on particular companies.
-Thus, the SMB registry seems to be the most complete openly available collection of Russian commercial companies.
-The necessity of company-level data for geographical research has been noted by many scientists (see literature review below), and the open collection of such data seems to have high potential value.
-For example, it can be used to map companies with high spatial resolution (up to cities).
-However, the dumps of SMB registry do not contain coordinates — only human-readible addresses are included.
-Also, these dumps are of large size (hundreds of gigabytes if archived and some terabytes if unpacked) and have complex structure.
-To make them usable by a regular developer, a simplification, geocoding and transformation is required.
+The SMB registry is operated by the Federal Tax Service (FTS) of Russia and includes information about all Russian SMBs that meet the inclusion criteria.
+These criteria include limits on revenue and the number of employees, and all eligible organizations and individual entrepreneurs are automatically included in the registry without any action required on their part.
 
-**Objective.** In this paper, an application that helps to build a ready-to-mapping spatial dataset of Russian small and medium-sized businesses is featured and discussed.
-The source data is open dumps of SMB registry.
-A spatial dataset contains geographical coordinates of each entity with spatial resolution up to a settlement (urban or rural).
-Application is a Python command-line (CLI) tool with easy-to-use interface that can be run on a medium personal computer without the need to use special tools for big data processing.
-In the paper, the application is first briefly described, and after that, a demonstration of its use for mapping the Russian legal companies is performed.
+The SMB registry seems to be the most comprehensive openly available collection of Russian commercial companies.
+Other sources of such data, e.
+g.
+business intelligence systems or the web portal of the Unified State Registry of Legal Entities, are either paid or have usage restrictions.
+The full dumps of the SMB registry are published as open data free of charge, making them a valuable data source for detailed high-resolution mapping of Russian business entities.
+However, the distributed data only contain human-readable location information and do not include geographical coordinates.
+Another problem is that these dumps have a large volume, reaching hundreds of gigabytes when archived and several terabytes when unpacked.
+Additionally, their structure is complex.
 
-**Literature review.** A typical source of data for studies of social geography in Russia is the resources of the Federal State Statistics Service (Rosstat).
-These include, in particular, the data of official statistics, censuses and statistical surveys posted on the official website of this body, the database of indicators of municipalities (BDMO) and the unified interdepartmental information and statistical system (EMISS).
-There are many examples of their use in geography: O.
-V.
-Kuznetsova, when studying the structure of the economy of Russian regions, uses statistics of gross regional product and the volume of goods shipped [@Kuznetsova2018]; G.
-M.
-Fedorov, in his analysis of the economy of the Russian regions in the Baltic, relies on data from Rosstat and the Federal Customs Service [@Fedorov2022]; S.
-G.
-Safronov studies the tertiary sector of the Russian economy on the basis of data from Eurostat, Rosstat and the Federal Tax Service [@Safronov2021]; Yu.
-V.
-Petrov, with the help of regional, municipal statistics and data from other authorities, formulates proposals for the socio-economic development of the south of the Tyumen region [@Petrov2021]; M.
-A.
-Sarancha assesses the development of the hotel and restaurant business, using the results of Rosstat’s continuous monitoring of small and medium-sized businesses [@Sarancha2014].
-Rosstat information is also used to create derived data sets, as in the work of I.
-Yu.
-Peker [@Peker2022].
-Thus, most socio-geographical studies cannot do without using Rosstat data as the only source of information or as one of the sources.
+**Objective.** The purpose of this work is to present a flexible and reproducible workflow for mapping Russian SMB using FTS open data.
+Reproducibility in this context refers to the development and publication of an open-source application for processing and geocoding FTS data, which can be used by any researcher.
+The paper provides a limited example of SMB mapping, focusing only on legal companies.
+However, the workflow's flexibility allows for the mapping of any SMB data.
 
-However, in some cases, Rosstat data is insufficient.
-As a result, there is a need for alternative sources of information that either complement official statistics or replace them.
-The reasons for this vary.
-In some cases, alternative sources of information are used to verify or correct Rosstat.
-For example, R.
-A.
-Babkin, based on data from cellular operators, proves that the official population of the Moscow agglomeration is underestimated by approximately 10% due to the error and inertia of statistical accounting and the “diocomy” of a significant part of Muscovites [@Babkin2020].
-Another, more common reason is the lack of necessary information.
-For example, K.V.
-Rostislav, analyzing the impact of location on the profitability of enterprises, takes into account only organizations with fewer than 250 employees and uses open data from the Federal Tax Service of Russia to check this condition, since Rosstat does not provide such information [@Rostislav2021].
-S.
-S.
-Lachininsky and co-authors estimate the spread of urban areas of St.
-Petersburg based on Landsat satellite images, since it is impossible to make the necessary calculations based on official statistics [@Lachininsky2023].
-The advantages of alternative data sources include increased accuracy and timeliness [@Vorobyev2021], higher spatial and temporal resolution [@Babkin2021], the possibility of accurate and detailed analysis at the micro level [@Bochkaryov2017], while noting the untimeliness, incompleteness and potential unreliability of official statistics [@Kuzminov2021].
-Alternative data sources provide access to the general population, as a result of which “the need for the use of selective, aggregated and indirect” sources of information is eliminated [@Blanutsa2017].
-When using machine learning methods, which is also becoming relevant in modern geography, the sample size, disaggregation, and number of features in the data are important, which also encourages turning to alternative sources [@Rostislav2020].
-Thus, although traditional statistical information is still important in socio-geographical research, and sometimes irreplaceable, there is a growing need for other information resources.
-The request from researchers is to increase the volume, speed of provision, accuracy, spatial and temporal resolution, expand the set of available indicators, as well as the possibility of validating statistical information.
-It manifests itself both in terms of traditional scientific methods and methods related to machine learning, although it is more relevant specifically for the latter, which are the most sensitive to the volume of data, their disaggregation, detail, and convenient presentation formats.
+**Literature review.** A common source of data for economic geography studies in Russia is the resources provided by the Federal State Statistics Service (Rosstat).
+These resources include official statistics, censuses, and statistical surveys available on the official website of Rosstat, as well as the database of municipality indicators and the unified interdepartmental information and statistical system.
+These data sources are used in various studies [@Sarancha2014; @Kuznetsova2018; @Petrov2021; @Safronov2021; @Fedorov2022].
+Rosstat is a crucial source of data in economic geography research, and its data is used explicitly or implicitly in almost all studies.
+This data is not suitable for cartography due to the lack of coordinates.
+However, it does provide information about location in terms of regions, municipalities, and cities.
+This information can be combined with spatial data from other sources, such as the popular GADM or Natural Earth databases, to create a map.
 
-One of the alternative sources of information for social geographers is administrative data from the Federal Tax Service of Russia.
-They are published on the official website of this authority or related resources.
-Geographical works use, in particular, information on the income of individuals [@Petrov2021], on tax revenues by federal subjects or types of economic activity [@Rostislav2020; @Safronov2021], on the average number of employees of organizations [@Rostislav2021], on the number of registered residential premises [@Lachininsky2021], on the number of small and medium-sized enterprises and measures to support them [@Gumenyuk2022].
-The attention to the data published by the Federal Tax Service of Russia is apparently due to the multifunctionality of this government body, due to which it collects a huge amount of information about different aspects of the Russian economy: business entities, property, income, expenses.
-Typically, the Federal Tax Service of Russia, like Rosstat, publishes summary (aggregated) information.
-However, there are several data sets that contain disaggregated (raw) information.
-Among them, the most attention is drawn to the register of small and medium-sized businesses (SME register), which is regularly and completely posted on the service’s website and contains information about specific organizations and individual entrepreneurs.
+Economic geographers may also use administrative data from the Federal Tax Service of Russia as an alternative source of information.
+It is published on the official website of this authority or its satellite web resources.
+Geographers use, in particular, information on the income of individuals [@Petrov2021], on tax revenues by regions or types of economic activity [@Rostislav2020; @Safronov2021], on the average number of employees of organizations [@Rostislav2021], on the number of SMBs and measures to support them [@Gumenyuk2022].
+FTS of Russia is not a specialized statistical body, but it collects a vast amount of information about various aspects of the Russian economy, including business entities, property, income, and expenses.
+The geographical location in FTS data is described with address elements rather than coordinates, similar to Rosstat.
+Additionally, it is worth noting that FTS of Russia operates the Federal Address Information System, which provides open address data widely used by government and business.
 
-For some economic-geographical and related studies, information detailed down to the level of specific firms is important.
-Such information includes, for example, location, revenue, expenses, number of employees, name and other characteristics.
-The specified data can be used directly, as in the works of K.
-V.
-Rostislav [@Rostislav2021], E.
-A.
-Kolomak and A.
-I.
-Sherubneva [@Kolomak2023], who analyze agglomeration effects based on enterprise indicators.
-In addition, on their basis, statistics with arbitrary spatial or temporal grouping can be considered, including those that are impossible, difficult or inconvenient to extract from Rosstat data.
-For example, I.
-S.
-Gumenyuk operates with calculations of the absolute and relative number of small and medium-sized enterprises in the municipalities of the Kaliningrad region [@Gumenyuk2022]; M.
-A.
-Sarancha includes the number of enterprises in the integral index of the level of development of small businesses in the territory of the municipality [@Sarancha2014]; and M.A.
-Makushin and co-authors use data on enterprises in the cities of BAM to assess the structure of industry [@Makushin2023].
-Information about specific firms acquires additional importance in city-scale studies [@Lachininsky2021; @Nikonorov2021; @Kozhevnikov2023], when studying narrow sectors of the economy that do not appear separately in statistics [@Moiseeva2016; @Pankratov2023], when analyzing a single industry [@Kuzminov2021].
+Researchers in economic geography have traditionally relied on aggregated data.
+However, modern studies often use information disaggregated down to the level of individual firms.
+This disaggregated data can be used directly [@Rostislav2021; @Kolomak2023] or to calculate derived metrics that are not present in Rosstat statistics [@Sarancha2014; @Gumenyuk2022; @Makushin2023].
+In city-scale studies [@Lachininsky2021; @Nikonorov2021; @Kozhevnikov2023], information about specific firms holds additional importance.
+Similarly, in the analysis of narrow sectors of the economy that are not separately represented in aggregated statistics, obtaining such detailed data is of key importance [@Moiseeva2016; @Pankratov2023].
+Getting such detailed data is difficult.
+Researchers typically use commercial or open geoinformation services, as well as business intelligence systems, to obtain data for analysis [@Lachininsky2021; @Moiseeva2016; @Kuzminov2021].
+However, these methods often present two challenges: either the data is provided for a fee, or significant effort is required to format it for analysis.
+Business intelligence systems, such as Spark or Kontur.Focus, accumulate information about various aspects of a company's operations.
+However, they are paid services.
+Public mapping platforms, such as 2GIS and Yandex.Maps, generally prohibit the downloading of data.
+Open platforms, like OpenStreetMap or the SMBs registry, are free, but the data dumps from them are large and complex.
+Converting the gathered data into a ready-to-map format requires a lot of time and effort.
 
-There is no generally accepted source of geo-referenced information about companies in Russia.
-Researchers and applied specialists usually rely on data from commercial [@Lachininsky2021] or open geoinformation services, business intelligence systems [@Moiseeva2016; @Kuzminov2021], as well as the above-mentioned SME register [@Gumenyuk2022].
-Typically, using these methods poses two problems: the data is provided for a fee, or considerable effort is required to bring it into a format suitable for analysis.
-Business intelligence systems such as Spark, Kontur.Focus and Ruslana contain a lot of varied information, but are paid.
-Downloading data from private geographic information systems like 2GIS and Yandex.Maps also costs money, and in addition, there is no convenient interface for it.
-Open platforms like OpenStreetMap or the SME registry are free, but the downloads from them are large and complex, so the researcher has to spend a lot of effort on the technical side of the work, involve third-party specialists, or refuse to use such a data source.
-Formally, a complete list of Russian organizations and individual entrepreneurs is contained in two state registers (respectively, the Unified State Register of Legal Entities and the Unified State Register of Individual Entrepreneurs), however, neither of them is distributed in full - you can only extract individual records manually through the Federal Tax Service web service Russia or automatically for a fee through a software interface, which does not provide access to the general population of companies and is unlikely to allow constructing a representative sample without excessive costs and effort.
-Thus, the question of a convenient tool that would allow free and legal obtaining the most complete set of geo-referenced data on Russian companies is relevant.
+**Data.** The mapping procedure uses a primary data source together with two additional data sources and two lookup tables.
 
-**Source data.** For the mapping, one main data source, two additional data sources, and two lookup tables are used.
-The main data source is SMB registry open data dumps[^1].
-They are published with monthly interval starting from August 2016.
-Each dump contains the full registry for the date of publication, so a substantial portion of the data is duplicated.
-Technically, a dump is a ZIP archive with thousands of XML files.
-Each XML file contains information about 900 organizations or individuals.
-The data attributes of XML file include, for example, name of the business, its registration address (region, district, city, settlement), activity codes according to All-Russia Classifier of Economic Activity Kinds.
+The primary data source is the SMB registry open data dumps[^2], which are published every month starting from August 2016.
+Each dump contains the full registry for the date of publication, resulting in a significant amount of duplicated data.
+Technically, a dump is a ZIP archive containing thousands of XML files, with each file including information about 900 organizations or individuals.
+The XML file's data attributes contain, in particular, the taxpayer number, business name, and registration address (region, district, city, settlement), as well as activity codes based on the All-Russia Classifier of Economic Activity Kinds.
 
-Additional data sources are two FTS open datasets.
-The first of them is called “Information about revenue and expenditure of organizations by their accounting (financial) reporting”[^3], and the second is named “Information about average list count of employees of organizations”[^4].
-Both of them are available starting from approximately 2019 and are updated irregularly, but at least once a year.
-Technically, they are also ZIP archives of XML files, where each archive contains data for a particular date, and each XML file lists 900 organizations.
-Individuals are not included in these datasets due to personal data issues.
+Two FTS open datasets serve as additional data sources.
+The first of them is called “Information about revenue and expenditure of organizations by their accounting (financial) reporting”[^3], and the second is named “Information about the average list count of employees of organizations”[^4].
+Both sources have been available since approximately 2019 and are updated at least once a year, although updates may be irregular.
+Technically, these datasets are collections of ZIP archives containing XML files.
+Each archive contains data for a specific date, and each XML file lists 900 organizations.
+Information about individuals is not included in these datasets due to personal data issues.
 
-[^3]: https://www.nalog.gov.ru/opendata/7707329152-revexp/
-[^4]: https://www.nalog.gov.ru/opendata/7707329152-sshr2019/
+[^2]: Unified registry of small and medium-sized businesses [electronic resource]. URL: https://www.nalog.gov.ru/opendata/7707329152-rsmp/ (access date: 29.02.2024)
+[^3]: Information about revenue and expenditure of organizations by their accounting (financial) reporting [electronic resource].
+URL: https://www.nalog.gov.ru/opendata/7707329152-revexp/ (access date: 29.02.2024)
+[^4]: Information about the average list count of employees of organizations [electronic resource]. URL: https://www.nalog.gov.ru/opendata/7707329152-sshr2019/ (access date: 29.02.2024)
 
-The lookup tables used for the mapping are “Settlements of Russia: population and geographic coordinates” dataset provided by “Infrastructure of the scientific data” project»[^5], and “Cities of Russia” dataset published by commercial company HFLabs[^6].
-Both of them contain addresses and geographical coordinates of Russian cities and settlements.
-In addition, the author created a tiny supplement to the “Cities of Russia” dataset with information about missing cities extracted manually from Federal Address Information System (FIAS)[^7].
+The lookup tables used for the mapping are the "Settlements of Russia: population and geographic coordinates" dataset provided by the "Infrastructure of scientific data" project[^5] and the "Cities of Russia" dataset published by the commercial company HFLabs[^6].
+Both datasets contain addresses and geographical coordinates of Russian cities and settlements.
+Additionally, the author added a small supplement to the 'Cities of Russia' dataset.
+This supplement includes information about missing cities that was manually extracted from the Federal Address Information System[^7].
 
-[^5]: https://data.rcsi.science/data-catalog/datasets/160/
-[^6]: https://github.com/hflabs/city
-[^7]: https://fias.ru
+[^5]: Settlements of Russia: population and geographic coordinates [electronic resource]. URL: (access date: 29.02.2024) https://data.rcsi.science/data-catalog/datasets/160/
+[^6]: Cities of Russia [electronic resource]. URL: https://github.com/hflabs/city (access date: 29.02.2024)
+[^7]: Federal Address Information System [electronic resource]. URL: https://fias.ru (access date: 29.02.2024)
 
-**Methods.** A Python command-line (CLI) application was developed to process the source data.
-Technically, it is based on Python standard library and several third-party packages, including fast XML processing library called lxml, iconic tabular data processing tool named Pandas, popular big data framework Apache Spark, and easy-to-use CLI framework Typer.
-This application was intended to be published as open-source tool for other researchers, so its source code with brief documentation is available at author's personal GitHub account[^8].
-Readers who are familiar with Python programming language may go to the repository and look at the code thoroughly.
-Here, a top-level structure of the application is described, and the data processing flow for the mapping is almost the same.
+A Python command-line (CLI) application was developed to process the source data.
+Technically, it is based on the Python standard library and several third-party packages, including the fast XML processing library called *lxml*, the iconic tabular data processing tool named *Pandas*, the popular big data framework *Apache Spark*, and the easy-to-use CLI framework *Typer*.
+This application was intended to be published as an open-source tool for other researchers, so its source code with brief documentation is available at GitHub public repository[^8].
+Readers who are familiar with Python programming language may go to the repository and look at the details of the code.
+This section describes the top-level structure of the application, and the data processing flow for the mapping is nearly identical.
 
-[^8]:
+[^8]: PavelSyomin/russian-smb-companies: Dataset on small and medium business companies registered in Russia based on Federal Tax Service open data [electronic resource].
+URL: https://github.com/PavelSyomin/russian-smb-companies (access date: 29.02.2024)
 
-*Step 1* is data download.
-The source ZIP archives were downloaded from FTS website and stored locally.
+*Step 1* is data download. The source ZIP archives were downloaded from the FTS website and stored locally.
+
 *Step 2* is data extraction and filtering.
-The data was extracted from ZIP archives and stored in CSV tables, so that one ZIP archive was transformed into 1 CSV table.
-A lot of unused data attributes were dropped, so the resulting size of CSV tables is far smaller than the size of original archives.
-Also, the filtering by activity code was performed on this step to select data only about legal companies.
-According to the classifier, the main activity code equal to 69.10 was used as a filtering clause.
-The application itself allows to filter by other codes or groups of classifier or disable filtering at all, thus providing the opportunity to generate various slices of companies (e.g.
-agricultural, forestry, health services, heavy industries etc) with various structural detalization (i.
+The data was extracted from ZIP archives and stored in CSV tables so that one ZIP archive was transformed into 1 CSV table.
+A lot of unused data attributes were dropped, so the resulting size of the CSV tables is much smaller than the size of the original archives.
+In addition, filtering by activity code took place in this step to select data only about legal companies.
+According to the classifier, the main activity code equal to 69.10 (activity in the area of law) was used as a filtering criterion.
+The application itself allows to filter by other codes or groups of codes or to disable filtering at all, thus providing the opportunity to generate various slices of companies (e.g.
+agricultural, forestry, health services, heavy industries, etc) with various structural resolution (i.
 e.
 up to groups or particular low-level codes of classifier).
+Also, in this step, the additional source datasets are extracted in the same fashion.
+
 *Step 3* is data aggregation: due to the specifics of data publication, a lot of information in archives and corresponding CSV tables is duplicated, and thus duplicates have to be dropped.
 This step dropped the duplicates and thus reduced the volume of data even further.
+Also, the additional data on revenue, expenditure, and employees was deduplicated and filtered to leave only the rows with organizations present in the main data.
+Taxpayer identification number (TIN) was used to find the necessary companies because it acts as a unique persistent company identifier.
+
 *Step 4* is geocoding.
-The addresses were converted to geographical coordinates using the look-up tables, and ≈98% of all addresses were successfully matched.
-Because the address in the source data is detalized up to settlements, the geographical coordinates refer to cities or settlements, and this is the lower bound of spatial resolution of the map.
-Optional *step 5* combines SMB registry data with data on revenue, expenditure and employees and transforms it to produce a yearly panel table.
+The addresses were normalized and converted to geographical coordinates using the look-up tables, and ≈98% of all addresses were successfully matched.
+Because the address in the source data is detailed down to cities or settlements, the geographical coordinates refer to cities or settlements, and this is the lower bound of spatial resolution of the map.
+No action is performed with additional datasets in this step.
+
+Optional *step 5* combines SMB registry data with data on revenue, expenditure, and employees and transforms it to produce a yearly panel table.
 Here, it was not performed, because the geocoded dataset was sufficient for mapping.
 
-After these preprocessing, the tabular dataset was filtered by year (2021 was chosen).
-The maps were plotted with a script written on R programming language with additional packages, including, in particular, ggplot2 and sf.
-The source code for the paper is available at the separate repository[^9].
+After this preprocessing, the tabular dataset was filtered by year (2021 was chosen).
+The maps were plotted with a script written in R programming language with additional packages, including, in particular, ggplot2 and sf.
+The source code for the paper is available in a separate repository[^9].
 
-[^9]:
+[^9]: ru-smb-companies-papers/legal-companies-mapping at main · PavelSyomin/ru-smb-companies-papers [electronic resource].
+URL: https://github.com/PavelSyomin/ru-smb-companies-papers/tree/main/legal-companies-mapping (access date: 29.02.2024)
 
-**Results.** The spatial resolution of the dataset is up to cities or settlements, but regions and their subdivisions are also included, so the aggregation and mapping are also possible.
+**Results.** The spatial resolution of the dataset is up to cities or settlements, but regions and their subdivisions are also included, so aggregation and mapping are also possible.
 Figure 1 shows the count of legal companies by region.
-Some spatial patterns are clearly seen from this visaulization, for instance, the high concentration of legal firms in the capital and the surrounding region, the “Urals-Volga-Krasnodar” belt, and small values in the Central economic region.
+Some spatial patterns are seen from this visualization, for instance, the high concentration of legal firms in the capital of Russia and the surrounding region, the “Urals-Volga-Krasnodar” belt, and small values in the Central economic region.
 
-```{r fig-1-settlements-russia, fig.cap = "Рис.
-1.
-Число субъектов МСП по населённым пунктам России и преобладающий вид деятельности.
-Размер кругов соответствует числу субъектов МСП, цвет — одной из трёх групп видов деятельности по классификатору ОКВЭД (см.
-легенду).
-Показаны только населённые пункты, в которых зарегистрировано более 50 фирм"}
-regions_plot
+```{r fig-1-regions-map, fig.cap = "**Fig. 1.** Count of legal companies by region (constituent entity) of Russia"}
+regions_map
 ```
 
-Figure 2 shows the results of the mapping of legal companies by settlements.
-This is high-resolution map which gives a quick overview of legal business in Russia.
-The spatial distribution a bit resembles the region-scale map, but city-level data allows to notice other tendencies, e.
+Figure 2 displays the mapping of legal companies by cities and rural settlements.
+This high-resolution map provides a quick overview of legal business in Russia.
+The spatial distribution resembles the region-scale map, but city-level data allows us to notice other tendencies, e.
 g.
-concentration of businesses in regional centres.
-Also, the high concentration in Moscow and surroundings is clear, and the similar pattern for St.
+the concentration of businesses in regional centres.
+Also, the high concentration in Moscow and its surroundings is clear, and a similar pattern for St.
 Petersburg is visible.
+A map may be also drawn for a particular region (Sverdlovsk oblast in Figure 3).
 
-```{r fig-2-settlements-russia, fig.cap = "Рис.
-2.
-Число субъектов МСП по населённым пунктам России и преобладающий вид деятельности.
-Размер кругов соответствует числу субъектов МСП, цвет — одной из трёх групп видов деятельности по классификатору ОКВЭД (см.
-легенду).
-Показаны только населённые пункты, в которых зарегистрировано более 50 фирм"}
+```{r fig-2-settlements-map, fig.cap = "**Fig. 2.** Count of legal companies in Russian urban and rural settlements"}
 settlements_map
@@
 
-The similar map may be drawn for a particular region, e. g. for Sverdlovsk oblast, as in the Figure 3. A pattern of high concentration is apparent again.
+<br>
 
-```{r fig-3-settlements_svr, fig.cap = "Рис.
-3.
-Число  субъектов МСП в населённых пунктах Свердловской области и их преобладающий вид деятельности.
-Размер кругов соответствует числу субъектов МСП, цвет — одной из трёх групп видов деятельности по классификатору ОКВЭД (см.
-легенду).
-Показаны все населённые пункты без ограничения по количеству зарегистрированных фирм"}
+```{r fig-3-settlements-map-svr, fig.cap = "**Fig. 3.** Count of legal companies in urban and rural settlements of Sverdlovsk oblast (the Middle Urals)"}
 settlements_map_svr
@@
 The presence of additional data about financial metrics and count of employees allows us to draw more maps.
-On Figure 4, the spatial distribution of legal companies profit is displayed, and on Figure 5, the similar distribution with respect to tne number of employees is shown.
+In Figure 4, the spatial distribution of legal companies' profit is displayed, and in Figure 5, a similar distribution concerning the number of employees is shown.
 Both maps include only companies with non-zero profit or number of employees.
-The maps looks like the counts map (see Figure 2), but exact spatial patterns vary.
+The maps look like the counts map (see Figure 2), but exact spatial patterns vary.
 
-```{r fig-4-profit-map, fig.cap = "Рис.
-4.
-Распределение числа зарегистрированных субъектов МСП, их суммарной прибыли и суммарной численности работников в демонстрационном наборе в 2021 г.
-по регионам (субъектам федерации)"}
+```{r fig-4-profit-map, fig.cap = "**Fig 4.** Profit of legal companies by urban and rural settlements in Russia"}
 settlements_map_profit
@@
 
-```{r fig-5-employees-map, fig.cap = "Рис.
-5.
-Распределение числа зарегистрированных субъектов МСП, их суммарной прибыли и суммарной численности работников в демонстрационном наборе в 2021 г.
-по регионам (субъектам федерации)"}
+```{r fig-5-employees-map, fig.cap = "**Fig 4.** Number of employees in legal companies by urban and rural settlements in Russia"}
 settlements_map_empl
@@
 
-**Conclusion.**
+**Conclusion.** In this paper, a flexible solution for high-resolution mapping of small and medium-sized companies in Russia is proposed.
+It is based on open data provided by Federal Tax Service which is preprocessed and geocoded with a Python CLI application to make a ready-to-mapping tabular spatial dataset.
+The example plots of legal companies and their business metrics (profit and number of employees) are provided to prove the usability of the suggested method.
+The code of the application used to make the dataset and the code of scripts used to draw maps are made open source and published in public repositories on GitHub.
+Despite the large amount of source data (up to several terabytes if unpacked), the tools used for the data processing and mapping are optimized to work on a regular modern PC and do not require to purchase cloud computing resources.
+Also, the geocoding is performed with simple look-up tables instead of commercial APIs, thus the entire workflow is free of charge and offline.
+
+There are several drawbacks and limitations of the described approach.
+First of all, even though the tools are designed to be as simple as possible, they still require some basic understanding of computer programming, because a researcher has to launch a Python CLI application and thus must be able to install Python and work with a command line.
+Next, although many optimization efforts were made, the generation of the dataset (mainly, the extraction of data) still takes a considerable time (up to one or two days if no filtering is used).
+Third, the accuracy of geocoding is high but does not equal 100%.
+Finally, the source data is limited to small and medium-sized companies only, thus any insights based on it must be done with caution.
+Nevertheless, the advised method allows making the most comprehensive mapping of Russian businesses based entirely on open free of charge data and algorithms.
+It may be especially useful for studying the SMB sector as is and for analysis of narrow economic areas that do not fall within the regular statistical reporting.
 

### 2024-02-29 legal-companies-mapping/helpers/Plots.R

@@ -21,7 +21,7 @@ regions <- regions_boundaries %>%
 ru_svr <- st_read("assets/ru_svr.geojson")
 
 # Regional distribution
-regions_plot <- smb_data %>%
+regions_map <- smb_data %>%
   filter(
     start_date <= "2021-12-31",
     end_date >= "2021-12-31") %>% 
@@ -37,7 +37,7 @@ regions_plot <- smb_data %>%
     direction = 1) +
   theme_bw(base_size = 11, base_family = "Times New Roman") +
   theme(legend.position = "bottom")
-regions_plot
+regions_map
 
 settlements_df <- smb_data %>%
   filter(
@@ -50,7 +50,7 @@ settlements_df <- smb_data %>%
 
 settlements_map <- regions %>% ggplot() +
   geom_sf(size = .1) +
-  geom_sf(data = settlements_df, aes(color = log10(n)), size = .1) +
+  geom_sf(data = settlements_df, aes(color = log10(n)), size = .5, shape = 21) +
   coord_sf(crs = ru_crs) +
   scale_color_distiller(
     name = "Count of legal companies (log10)",
@@ -73,9 +73,13 @@ settlements_df_svr <- smb_data %>%
 settlements_map_svr <- ru_svr %>% 
   ggplot() +
   geom_sf() +
-  geom_sf(data = settlements_df_svr, aes(size = n)) +
+  geom_sf(data = settlements_df_svr, aes(color = log10(n)), size = 2) +
   coord_sf(crs = 4326) +
   scale_size_continuous() +
+  scale_color_distiller(
+    name = "Count of legal companies (log10)",
+    palette = "YlGn",
+    direction = 1) +
   theme_bw(base_size = 11, base_family = "Times New Roman")
 settlements_map_svr
 

### 2024-03-04 legal-companies-concentration-and-availability/Paper.Rmd
Remove Zipf law parts from the paper, edit plots, add regression results, edit text, add references

---
title: "UDC 911.3"
subtitle: "Does Concentration of Legal Companies Increases Accessibility of Legal Services: Empirical Evidence from Russian Cities"
output:
  word_document:
    reference: reference.docx
bibliography: biblio.bib
csl: biblio-style.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  dpi = 300,
  fig.id = FALSE,
  fig.width = 6.6,
  fig.height = 5)
```

```{r imports, include=FALSE, cache = TRUE}
source("Script.R")
```

<div custom-style="Author">

Pavel O. Syomin

<br>

Perm State University

<br>

e-mail: ntsp\@ya.ru

<br>

The paper is published for the first time.

</div>

<br>

<div custom-style="Abstract">

**Abstract:** a link between concentration of legal companies and availability of legal services is examined. For the analysis, a dataset on the location of small and medium-sized legal companies and their revenue, expenditure and number of employees was made based on Federal Tax Service open data. The concentration is calculated as a ratio of the number of employees of legal companies in a city and a total number of employees of all legal companies in Russia, and the number of employees of legal companies per 100k population was used to measure the availability. Data visualization and linear regression were used to estimate the link. Positive linear relationship between the concentration and availability is found. However, there are various ways to reach good availability of legal services.

**Keywords:** legal services, concentration, availability, small and medium-sized business, open data

</div>

<br>

For a long time cities have been thought as centers of regions, providing other settlements with goods and services [@Christaller1966]. It is known that professional and business services tend to cluster in large cities, and their spatial distribution is closely linked to urban hierarchies [@Keeble2002; @Rubalcaba2013; @Kekezi2019]. Concentration provides economic benefits for the companies and other becomes self-sustaining. However, the increase in the number of firms do not necessarily leads to the corresponding growth in the accessibility of services they provide. An issue of access to various services also has been of great importance for decades [@Blacksell1990; @Keleher1996], because it is directly connected to the quality of life and human rights. There is also a research on the links between market properties and services availability [@Lu2020]. Typically, the literature focuses on few kinds of services, e. g. medical [@Joseph1982], but other professional services also worth attention, e. g. legal services [@Blacksell1988; @Patel2008].

In this paper, the link between the concentration of legal companies in Russian cities and availability of these services is empirically examined. The analysis is based on information about the location of small and medium-sized legal businesses in Russia obtained from open data published by Federal Tax Service (FTS). A main research question is how the availability of legal professionals depends on the concentration of legal companies. This relationship is tested with respect to city population and relative rank in the region (constituent entity) of Russia. An auxiliary problem is to find (if any) geographical patterns in the spatial distribution of cities by legal services availability.

**Data.** Company-wise disaggregated geo-referenced dataset on small and medium-sized legal companies in Russia was prepared and used in this research. The dataset was created with Python CLI tool *ru-smb-companies* developed by the author of this paper. This tool downloads, extracts, aggregates and geocodes large archived XML dumps from the state registry of small and medium-sized businesses (SMBs) distributed by FTS, producing a tiny CSV dataset. To select legal companies, a filtering by main activity code equal to 69.10 was applied. In addition, the dataset was filtered by the following conditions: year is 2021, company is an organization rather than a sole entrepreneur, company has declared non-zero revenue for the target year, company's address of registration is in a city rather than in a rural area. Also, during the exploratory data analysis, four cities (namely, Innopolis, Kirovsk, Bronnytsy, Krasnoarmeisk) were marked as outliers by the count of companies or employees and excluded from the data.

**Methods.** *Concentration of legal companies* in cities was estimated as a number of employees of legal companies in a given city divided by the total number of employees in legal companies in all the cities. The formula was derived and adapted from the Atlas of Economic Specialization of Russian Regions [@Atlas2021]. *Availability of legal services* was measured as the number of employees of legal companies in a city divided by the city's population. This formula is an adapted version of the legal market saturation metric used in the analysis of Russian legal services market conducted by the Institute for the Rule of Law [@Moiseeva2016]. To categorize the cities by population, the classification of Lappo [@Lappo1997] was used. Data visualization, including mapping, was used as the main method, and linear regression as a helper method. Technically, the analysis was performed with R programming language and additional packages. The source code for this paper is available at the public repository: https://github.com/PavelSyomin/ru-smb-companies-papers/tree/main/legal-companies-concentration-and-availability

**Results and discussion.** The relationship between the concentration of legal companies and availability of legal services (as measured with the metrics proposed in Methods section) is shown in Figure 1. Apart from the general trend, the groups by city population and city rank within the region are also displayed. The overall connection between the concentration and availability is linear. The results of linear modelling proves it: the link between the two variables can be described with the formula *`r ca_model_formula`*, where both the intercept and slope are highly significant *(p < 0.001)*, and the explained variance is relatively high *(R^2^ = `r round(ca_model_summary$adj.r.squared, 2)`)*.

Looking at the regression line common to all the data points, it is easy to notice that despite the general linearity and good linear modelling results, the total distribution is also fan-shaped, that is, the variance of error of the linear model declines as the both dependent and independent variables grow. However, this is mostly the result of several distinct population size groups of cities present in the data. Within each of these groups, the assumptions of linear models are far better satisfied. The link between concentration and availability is similar in all city population groups, and the regression lines are nearly parallel, meaning that the slope coefficient is equal, and only the intercept differs. Also, it can be concluded that in general, the larger is the city, the bigger is the concentration and the bigger is the availability, even though the relationship is relatively weak.

The first (top rank) cities in the regions usually have better availability. This observation is partially explained by the considerably big size of such cities, despite several exceptions. Also, the top rank of the city within the region does not always lead to the highest possible availability of legal services.

Moscow and Saint Petersburg are two remarkable outliers with extremely high concentration and one of the highest levels of availability of legal services. Moscow accounts for almost 1/3, and Saint Petersburg for about 10% of all employees of legal companies. The relative position of the corresponding data points (compared to the main regression lines) implies that there is some sort of “availability saturation”, that is, the level of concentration where the marginal growth of concentration has little effect on availability.

So, high concentration of legal companies, big population, and the status of first city within the region are positively associated with the availability of legal services. On the contrary, the larger is the city, the more legal companies it must have to satisfy the people's needs. As a result, there are different ways to reach high availability of legal services that range from being the business and service center with high concentration that leads to good availability to being a small yet well-served city where a small flock of legal firms satisfy the needs of the local community.

```{r fig-1-concentration-and-accessibility, fig.cap="Figure 1. The link between the concentration of legal companies and the availability of legal services. In addition, city population group and relative city rank within the region are marked. The formulas for the metrics are described in the Methods section. Both X and Y axes are log10-scaled. Dotted regression lines represent the linear fits within city population groups, and the solid gray line represend the overall linear fit", fig.height=5}
ca_plot
```

The spatial distribution of cities by legal services availability is shown on the map below (Figure 2). Regional centers have high relative level of availability. Moscow and its satellite cities form the most prominant cluster of good access to legal services. Other notable centers with high availability are Krasnodar and Stavropol at the southern part of Russia's East, Yekaterinburg in the Urals, Tuymen in Western Siberia. There are regions with low accessibility of legal services, for example, bordering regions in the South of Siberia, Komi and Kalmyk republics.

```{r fig-2-map, fig.cap="Figure 2. Legal service availability in Russian cities on the country's map", fig.height=5}
spatial_plot
```

**Conclusion.** The concentration of legal companies in cities has a positive linear relationship with the legal services availability. The similar linear trend is observed in all the city population groups. The largest cities of regions usually have relatively high availability of legal services, and the same is true for the cities with population above 1 million. Moscow and Saint Petersburg have extremely high concentration and high availability. Probably there is a saturation limit after which the increase of concentration do not have positive impact on availability. Empirical data suggests several paths for reaching high levels of legal services availability, where increasing concentration of legal companies in a city is just one of the options, and cities with relatively small concentration can achieve the level of accessibility similar to the largest ones. Moscow and its surrounding area forms a remarkable cluster of high legal services availability. As for other places, good levels of access to legal services are noticed in Saint Petersburg, Yekaterinburg, Krasnodar, Tyumen, Stavropol.

## References

<div id="refs"></div>

### 2024-03-04 legal-companies-concentration-and-availability/Script.R

@@ -1,5 +1,6 @@
 library(dplyr)
 library(ggplot2)
+library(glue)
 library(readr)
 library(sf)
 library(tidyr)
@@ -57,13 +58,50 @@ data <- left_join(lcc, cp, by = c("oktmo", "city")) %>%
     is_largest_in_region = city_region_rank == 1) %>% 
   ungroup()
 
-concentration_and_accessibility_plot <- ggplot(data, aes(
-    x = empl_per_100k, y = concentration, 
-    color = city_size_group, shape = is_largest_in_region)) +
-  geom_point(size = 2) +
-  scale_x_log10(name = "Lawyers per 100k population (log10)") +
-  scale_y_log10(name = "Concentration (log10)") +
-  scale_color_brewer(name = "City size by population", palette = "Dark2") +
+ca_model <- lm(log(empl_per_100k) ~ log(concentration), data)
+ca_model_summary <- summary(ca_model)
+ca_model_formula = glue(
+  "log10(availability) = {intercept} {sign} {slope} × log10(concentration)",
+  intercept = round(ca_model_summary$coefficients[1, 1], 2), 
+  slope = round(ca_model_summary$coefficients[2, 1], 2),
+  sign = substr(sprintf("%+ .2f", ca_model_summary$coefficients[2, 1]), 1, 1))
+ca_model_formula
+
+ca_plot <- ggplot(
+    data, 
+    aes(x = concentration, y = empl_per_100k)
+  ) +
+  geom_text(
+    aes(label = c("Msk", "SPb")),
+    data = filter(data, city %in% c("Москва", "Санкт-Петербург")),
+    vjust = -1,
+    hjust = 1,
+  ) +
+  geom_point(
+    aes(color = city_size_group, shape = is_largest_in_region), 
+    size = 2
+  ) +
+  geom_smooth(
+    aes(color = city_size_group), 
+    data = filter(data, ! (city %in% c("Москва", "Санкт-Петербург"))),
+    method = "lm", 
+    se = FALSE,
+    linetype = "dotted",
+    size = 1,
+    alpha = .75
+  ) +
+  geom_smooth(
+    data = filter(data, ! (city %in% c("Москва", "Санкт-Петербург"))),
+    method = "lm", 
+    se = FALSE,
+    color = "gray80",
+    size = 1,
+    alpha = .75
+  ) +
+  scale_x_log10(name = "Concentration (log10)") +
[…]

### 2024-03-05 legal-companies-concentration-and-availability/Paper.Rmd
Proofreading, fitting into 5 pages

@@
 title: "UDC 911.3"
-subtitle: "Does Concentration of Legal Companies Increases Accessibility of Legal Services: Empirical Evidence from Russian Cities"
+subtitle: "Concentration of Legal Companies and Accessibility of Legal Services: Empirical Evidence from Russian Cities"
 output:
@@
 
-The paper is published for the first time.
+The paper is being published for the first time.
 
@@
 
-**Abstract:** a link between concentration of legal companies and availability of legal services is examined.
-For the analysis, a dataset on the location of small and medium-sized legal companies and their revenue, expenditure and number of employees was made based on Federal Tax Service open data.
-The concentration is calculated as a ratio of the number of employees of legal companies in a city and a total number of employees of all legal companies in Russia, and the number of employees of legal companies per 100k population was used to measure the availability.
-Data visualization and linear regression were used to estimate the link.
-Positive linear relationship between the concentration and availability is found.
-However, there are various ways to reach good availability of legal services.
+**Abstract:** the article examines the relationship between the concentration of legal companies and the availability of legal services.
+To conduct the analysis, a dataset was created using open data from the Federal Tax Service of Russia.
+The dataset includes information on the location, revenue, expenditure, and number of employees of small and medium-sized legal companies.
+The concentration is calculated by dividing the number of employees of legal companies in the city by the total number of employees of all legal companies in Russia.
+Availability is measured as the number of employees of legal companies per 100,000 population.
+Data visualization and linear regression were utilized to estimate the hypothesized relationship.
+The study showed a positive linear relationship between concentration and availability.
+However, there are multiple approaches to achieving good availability of legal services.
 
@@
 
-For a long time cities have been thought as centers of regions, providing other settlements with goods and services [@Christaller1966].
-It is known that professional and business services tend to cluster in large cities, and their spatial distribution is closely linked to urban hierarchies [@Keeble2002; @Rubalcaba2013; @Kekezi2019].
-Concentration provides economic benefits for the companies and other becomes self-sustaining.
-However, the increase in the number of firms do not necessarily leads to the corresponding growth in the accessibility of services they provide.
-An issue of access to various services also has been of great importance for decades [@Blacksell1990; @Keleher1996], because it is directly connected to the quality of life and human rights.
-There is also a research on the links between market properties and services availability [@Lu2020].
-Typically, the literature focuses on few kinds of services, e.
-g.
-medical [@Joseph1982], but other professional services also worth attention, e.
-g.
-legal services [@Blacksell1988; @Patel2008].
+Cities have traditionally been viewed as regional centers that provide goods and services to smaller surrounding settlements [@Christaller1966].
+It is well-known that professional and business services tend to cluster in large cities, and their spatial distribution is closely linked to urban hierarchies [@Keeble2002; @Rubalcaba2013; @Kekezi2019].
+Concentration of businesses provides economic benefits for the companies and thus is likely to become self-sustaining.
+However, an increase in the number of firms does not necessarily lead to a corresponding growth in the availability of services they provide.
+The issue of access to various services has been of great importance for decades [@Blacksell1990; @Keleher1996], because it is directly connected to the quality of life and human rights.
+Research has also been conducted on the links between market properties and service availability [@Lu2020].
+Typically, the literature focuses on medical services [@Joseph1982], but other professional services are also worth attention, such as legal services [@Blacksell1988; @Patel2008].
 
-In this paper, the link between the concentration of legal companies in Russian cities and availability of these services is empirically examined.
-The analysis is based on information about the location of small and medium-sized legal businesses in Russia obtained from open data published by Federal Tax Service (FTS).
-A main research question is how the availability of legal professionals depends on the concentration of legal companies.
-This relationship is tested with respect to city population and relative rank in the region (constituent entity) of Russia.
-An auxiliary problem is to find (if any) geographical patterns in the spatial distribution of cities by legal services availability.
+This paper empirically examines link between the concentration of legal companies in Russian cities and the availability of legal services.
+The analysis is based on information about the location of small and medium-sized legal businesses in Russia obtained from open data published by the Federal Tax Service (FTS).
+The main research question is how the availability of legal professionals is affected by the concentration of legal companies.
+Additionally, the study aims to identify any geographical patterns in the spatial distribution of cities based on legal services availability.
 
-**Data.** Company-wise disaggregated geo-referenced dataset on small and medium-sized legal companies in Russia was prepared and used in this research.
-The dataset was created with Python CLI tool *ru-smb-companies* developed by the author of this paper.
-This tool downloads, extracts, aggregates and geocodes large archived XML dumps from the state registry of small and medium-sized businesses (SMBs) distributed by FTS, producing a tiny CSV dataset.
+**Data.** In this research, a company-wise geo-referenced dataset on small and medium-sized legal companies in Russia was used.
+The dataset was created using the Python CLI tool *ru-smb-companies* developed by the author of this paper[^1].
+The tool downloads, extracts, aggregates and geocodes large archived XML dumps from the state registry of small and medium-sized businesses (SMBs) distributed by FTS, producing a compact CSV dataset.
 To select legal companies, a filtering by main activity code equal to 69.10 was applied.
-In addition, the dataset was filtered by the following conditions: year is 2021, company is an organization rather than a sole entrepreneur, company has declared non-zero revenue for the target year, company's address of registration is in a city rather than in a rural area.
-Also, during the exploratory data analysis, four cities (namely, Innopolis, Kirovsk, Bronnytsy, Krasnoarmeisk) were marked as outliers by the count of companies or employees and excluded from the data.
+Additionally, the dataset was filtered by the following conditions: the year is 2021, the company is an organization rather than a sole entrepreneur, the company has declared non-zero revenue for the target year, and the company's address of registration is in a city rather than in a rural area.
+During the exploratory data analysis, four cities (Innopolis, Kirovsk, Bronnytsy, Krasnoarmeisk) were identified as outliers based on the number of companies or employees per 100,000 population and were subsequently excluded from the data.
 
-**Methods.** *Concentration of legal companies* in cities was estimated as a number of employees of legal companies in a given city divided by the total number of employees in legal companies in all the cities.
-The formula was derived and adapted from the Atlas of Economic Specialization of Russian Regions [@Atlas2021].
-*Availability of legal services* was measured as the number of employees of legal companies in a city divided by the city's population.
-This formula is an adapted version of the legal market saturation metric used in the analysis of Russian legal services market conducted by the Institute for the Rule of Law [@Moiseeva2016].
-To categorize the cities by population, the classification of Lappo [@Lappo1997] was used.
-Data visualization, including mapping, was used as the main method, and linear regression as a helper method.
-Technically, the analysis was performed with R programming language and additional packages.
-The source code for this paper is available at the public repository: https://github.com/PavelSyomin/ru-smb-companies-papers/tree/main/legal-companies-concentration-and-availability
+[^1]: https://github.com/PavelSyomin/russian-smb-companies
 
-**Results and discussion.** The relationship between the concentration of legal companies and availability of legal services (as measured with the metrics proposed in Methods section) is shown in Figure 1.
-Apart from the general trend, the groups by city population and city rank within the region are also displayed.
-The overall connection between the concentration and availability is linear.
-The results of linear modelling proves it: the link between the two variables can be described with the formula *`r ca_model_formula`*, where both the intercept and slope are highly significant *(p < 0.001)*, and the explained variance is relatively high *(R^2^ = `r round(ca_model_summary$adj.r.squared, 2)`)*.
+**Methods.** The *concentration of legal companies* in cities was calculated by dividing the number of employees of legal companies in a given city by the total number of employees of legal companies in all cities.
+This formula was derived and adapted from the Atlas of Economic Specialization of Russian Regions [@Atlas2021].
+The *availability of legal services* was measured by dividing the number of employees of legal companies in a city divided by the city's population.
+This formula is a modified version of the legal market saturation metric used in the analysis of the Russian legal services market conducted by the Institute for the Rule of Law [@Moiseeva2016].
+To categorize the cities by population, the classification system of Lappo [@Lappo1997] was used.
+The main method employed was data visualization, including mapping, with linear regression used as a supplementary method.
+The analysis was performed using the R programming language and additional packages.
+The source code for this paper can be found at the public repository[^2].
 
-Looking at the regression line common to all the data points, it is easy to notice that despite the general linearity and good linear modelling results, the total distribution is also fan-shaped, that is, the variance of error of the linear model declines as the both dependent and independent variables grow.
-However, this is mostly the result of several distinct population size groups of cities present in the data.
-Within each of these groups, the assumptions of linear models are far better satisfied.
-The link between concentration and availability is similar in all city population groups, and the regression lines are nearly parallel, meaning that the slope coefficient is equal, and only the intercept differs.
-Also, it can be concluded that in general, the larger is the city, the bigger is the concentration and the bigger is the availability, even though the relationship is relatively weak.
+[^2]: https://github.com/PavelSyomin/ru-smb-companies-papers/tree/main/legal-companies-concentration-and-availability
 
-The first (top rank) cities in the regions usually have better availability.
-This observation is partially explained by the considerably big size of such cities, despite several exceptions.
-Also, the top rank of the city within the region does not always lead to the highest possible availability of legal services.
+**Results and discussion.** Figure 1 shows the relationship between the concentration of legal companies and the availability of legal services, as measured by the metrics proposed in Methods section.
+The plot displays the groups by city population and city rank within the region, in addition to the general trend.
+The overall connection between the concentration and availability appears to be linear.
+The linear model results demonstrate that the relationship between the two variables can be described by the formula *`r ca_model_formula`*, where both the intercept and slope are highly significant *(p < 0.001)*, and the explained variance is relatively high *(R^2^ = `r round(ca_model_summary$adj.r.squared, 2)`)*.
 
-Moscow and Saint Petersburg are two remarkable outliers with extremely high concentration and one of the highest levels of availability of legal services.
-Moscow accounts for almost 1/3, and Saint Petersburg for about 10% of all employees of legal companies.
-The relative position of the corresponding data points (compared to the main regression lines) implies that there is some sort of “availability saturation”, that is, the level of concentration where the marginal growth of concentration has little effect on availability.
+Examining the regression line common to all data points, we see that while the linearity is generally good and the linear modeling results are satisfactory, the total distribution is fan-shaped.
+This means that the variance of error of the linear model decreases as both the dependent and independent variables increase.
+However, this is mainly due to the presence of several distinct population size groups of cities in the data.
+Within each of these groups, linear models satisfy the assumptions better.
+The link between concentration and availability is similar across all city population groups, and the regression lines are nearly parallel, indicating equal slope coefficients and differing only in intercepts.
+Additionally, it can be concluded that, in general, larger cities have higher concentrations and availability, although the relationship is relatively weak.
 
-So, high concentration of legal companies, big population, and the status of first city within the region are positively associated with the availability of legal services.
-On the contrary, the larger is the city, the more legal companies it must have to satisfy the people's needs.
-As a result, there are different ways to reach high availability of legal services that range from being the business and service center with high concentration that leads to good availability to being a small yet well-served city where a small flock of legal firms satisfy the needs of the local community.
+<br>
 
 ```{r fig-1-concentration-and-accessibility, fig.cap="Figure 1.
-The link between the concentration of legal companies and the availability of legal services.
-In addition, city population group and relative city rank within the region are marked.
-The formulas for the metrics are described in the Methods section.
-Both X and Y axes are log10-scaled.
-Dotted regression lines represent the linear fits within city population groups, and the solid gray line represend the overall linear fit", fig.height=5}
+The relationship between the concentration of legal companies and the availability of legal services.
+City population group and relative city rank within the region are marked.
+The formulas used to calculate the metrics are described in the Methods section.
+Both the X and Y axes are log10-scaled.
+Dotted regression lines represent the linear fits within city population groups, while the solid gray line represent the overall linear fit", fig.height=4}
 ca_plot
@@
 
-The spatial distribution of cities by legal services availability is shown on the map below (Figure 2).
-Regional centers have high relative level of availability.
-Moscow and its satellite cities form the most prominant cluster of good access to legal services.
-Other notable centers with high availability are Krasnodar and Stavropol at the southern part of Russia's East, Yekaterinburg in the Urals, Tuymen in Western Siberia.
-There are regions with low accessibility of legal services, for example, bordering regions in the South of Siberia, Komi and Kalmyk republics.
+In general, cities with the top rank by the population in their respective regions tend to have better availability of legal services.
+One possible reason for this is a considerably big size of such cities.
 
-```{r fig-2-map, fig.cap="Figure 2. Legal service availability in Russian cities on the country's map", fig.height=5}
+Moscow and Saint Petersburg are two remarkable outliers with extremely high concentration of legal services and a good level of legal services availability.
+Moscow alone accounts for almost one-third of all legal companies employees, and Saint Petersburg accounts for about 10%.
+The position of the corresponding data points relative to the main regression lines suggests that there is an “availability saturation” level, where further concentration has little effect on availability.
+
+A high concentration of legal companies, a large population, and the city's top rank by population within a region are positively associated with the availability of legal services.
+Conversely, larger cities require more legal companies to meet the needs of their residents.
+As a result, there are various ways to achieve high availability of legal services, ranging from being a business and service center with a high concentration of legal firms to being a small yet well-served city where a small number of legal firms meet the needs of the local community.
+
+Figure 2 displays the spatial distribution of cities based on the availability of legal services.
+Regional centers exhibit a high relative level of availability.
+The most prominent cluster of good access to legal services is formed by Moscow and its satellite cities.
+Other notable centers with high availability include Krasnodar and Stavropol in the southern part of Russia's West, Yekaterinburg in the Urals, and Tuymen in Western Siberia.
+There are regions with limited access to legal services, such as the bordering regions in the south of Siberia, as well as the Komi and Kalmyk Republics.
+
+<br>
+
+```{r fig-2-map, fig.cap="Figure 2. Legal services availability in Russian cities on the country's map", fig.height=4}
 spatial_plot
@@
 
-**Conclusion.** The concentration of legal companies in cities has a positive linear relationship with the legal services availability.
-The similar linear trend is observed in all the city population groups.
-The largest cities of regions usually have relatively high availability of legal services, and the same is true for the cities with population above 1 million.
-Moscow and Saint Petersburg have extremely high concentration and high availability.
-Probably there is a saturation limit after which the increase of concentration do not have positive impact on availability.
-Empirical data suggests several paths for reaching high levels of legal services availability, where increasing concentration of legal companies in a city is just one of the options, and cities with relatively small concentration can achieve the level of accessibility similar to the largest ones.
-Moscow and its surrounding area forms a remarkable cluster of high legal services availability.
-As for other places, good levels of access to legal services are noticed in Saint Petersburg, Yekaterinburg, Krasnodar, Tyumen, Stavropol.
+**Conclusion.** The concentration of legal companies in cities has a positive linear relationship with the availability of legal services.
+This trend is observed in all the city population groups.
+The largest cities in regions typically have relatively high availability of legal services, as do cities with population above 1 million.
+Moscow and Saint Petersburg are characterized by the highest concentration and high availability, with Moscow alone accounting for almost one-third of all legal companies employees in the country.
+Empirical data suggests that there may be a saturation limit beyond which the increase of concentration do not have a substantial impact on availability.
+There are multiple paths to achieving high levels of legal services availability, with increasing concentration of legal companies in a city being just one option.
+Even cities with relatively low concentration can achieve similar levels of availability as larger ones.
+Moscow and its region are a cluster of areas with high legal services availability.
+Similarly, Saint Petersburg, Yekaterinburg, Krasnodar, Tyumen, and Stavropol also have good access to legal services.
 

### 2024-03-05 legal-companies-concentration-and-availability/Script.R

@@ -103,7 +103,7 @@ ca_plot <- ggplot(
   scale_color_brewer(
     name = "City size by population", palette = "Dark2") +
   scale_shape_manual(
-    name = "", 
+    name = NULL, 
     values = c("TRUE" = 19, "FALSE" = 21),
     labels = c(
       "TRUE" = "The largest city in a region", 
@@ -112,6 +112,7 @@ ca_plot <- ggplot(
   theme(
     legend.position = c(.99, .01),
     legend.justification = c(1, 0),
+    legend.spacing.y = unit(0, "cm"),
     panel.grid.minor.y = element_blank())
 ca_plot
 
@@ -124,10 +125,5 @@ spatial_plot <- data %>%
   coord_sf(crs = ru_crs) +
   scale_color_binned(name = "Availability", n.breaks = 4, low = "#dadaeb", high = "#3f007d") +
   scale_size_discrete(name = "City size by population", range = c(.2, 2)) +
-  theme_bw(base_size = 14, base_family = "Times New Roman") +
-  theme(
-    legend.position = "bottom",
-    legend.box = "horizontal",
-    legend.direction = "vertical"
-  )
+  theme_bw(base_size = 14, base_family = "Times New Roman")
 spatial_plot

### 2024-03-06 legal-companies-concentration-and-availability/Paper.Rmd
Proofreading (round 2) and other minor improvements

@@
 
-**Abstract:** the article examines the relationship between the concentration of legal companies and the availability of legal services.
+**Abstract:** The article examines the relationship between the concentration of legal companies and the availability of legal services.
 To conduct the analysis, a dataset was created using open data from the Federal Tax Service of Russia.
@@
 The concentration is calculated by dividing the number of employees of legal companies in the city by the total number of employees of all legal companies in Russia.
-Availability is measured as the number of employees of legal companies per 100,000 population.
+Availability is measured as the number of employees of legal companies in the city per 100,000 population.
 Data visualization and linear regression were utilized to estimate the hypothesized relationship.
@@
 
-**Keywords:** legal services, concentration, availability, small and medium-sized business, open data
+**Keywords:** legal services, concentration, availability, small and medium-sized businesses, open data
 
@@
 It is well-known that professional and business services tend to cluster in large cities, and their spatial distribution is closely linked to urban hierarchies [@Keeble2002; @Rubalcaba2013; @Kekezi2019].
-Concentration of businesses provides economic benefits for the companies and thus is likely to become self-sustaining.
+The concentration of businesses provides economic benefits for the companies and thus is likely to become self-sustaining.
 However, an increase in the number of firms does not necessarily lead to a corresponding growth in the availability of services they provide.
-The issue of access to various services has been of great importance for decades [@Blacksell1990; @Keleher1996], because it is directly connected to the quality of life and human rights.
+The issue of access to various services has been of great importance for decades [@Blacksell1990; @Keleher1996] because it is directly connected to the quality of life and human rights.
 Research has also been conducted on the links between market properties and service availability [@Lu2020].
@@
 
-This paper empirically examines link between the concentration of legal companies in Russian cities and the availability of legal services.
+This paper empirically examines the link between the concentration of legal companies in Russian cities and the availability of legal services.
 The analysis is based on information about the location of small and medium-sized legal businesses in Russia obtained from open data published by the Federal Tax Service (FTS).
@@
 The dataset was created using the Python CLI tool *ru-smb-companies* developed by the author of this paper[^1].
-The tool downloads, extracts, aggregates and geocodes large archived XML dumps from the state registry of small and medium-sized businesses (SMBs) distributed by FTS, producing a compact CSV dataset.
+The tool downloads, extracts, aggregates, and geocodes large archived XML dumps from the state registry of small and medium-sized businesses (SMBs) distributed by FTS, producing a compact CSV dataset.
 To select legal companies, a filtering by main activity code equal to 69.10 was applied.
-Additionally, the dataset was filtered by the following conditions: the year is 2021, the company is an organization rather than a sole entrepreneur, the company has declared non-zero revenue for the target year, and the company's address of registration is in a city rather than in a rural area.
+Additionally, the dataset was filtered by the following conditions: the year is 2021, the company is an organization rather than a sole entrepreneur, the company has declared non-zero revenue and expenditure for the target year, and the company's address of registration is in a city rather than in a rural area.
 During the exploratory data analysis, four cities (Innopolis, Kirovsk, Bronnytsy, Krasnoarmeisk) were identified as outliers based on the number of companies or employees per 100,000 population and were subsequently excluded from the data.
@@
 
-**Methods.** The *concentration of legal companies* in cities was calculated by dividing the number of employees of legal companies in a given city by the total number of employees of legal companies in all cities.
+**Methods.** The *concentration of legal companies* in cities was calculated by dividing the number of employees of legal companies in the given city by the total number of employees of legal companies in all cities.
 This formula was derived and adapted from the Atlas of Economic Specialization of Russian Regions [@Atlas2021].
-The *availability of legal services* was measured by dividing the number of employees of legal companies in a city divided by the city's population.
+The *availability of legal services* was measured by dividing the number of employees of legal companies in the city by the city's population.
 This formula is a modified version of the legal market saturation metric used in the analysis of the Russian legal services market conducted by the Institute for the Rule of Law [@Moiseeva2016].
@@
 The analysis was performed using the R programming language and additional packages.
-The source code for this paper can be found at the public repository[^2].
+The source code for this paper can be found in the public repository[^2].
 
@@
 
-**Results and discussion.** Figure 1 shows the relationship between the concentration of legal companies and the availability of legal services, as measured by the metrics proposed in Methods section.
+**Results and discussion.** Figure 1 shows the relationship between the concentration of legal companies and the availability of legal services, as measured by the metrics proposed in the Methods section.
 The plot displays the groups by city population and city rank within the region, in addition to the general trend.
@@
 This means that the variance of error of the linear model decreases as both the dependent and independent variables increase.
-However, this is mainly due to the presence of several distinct population size groups of cities in the data.
+However, this is mainly due to the presence of several distinct population-size groups of cities in the data.
 Within each of these groups, linear models satisfy the assumptions better.
@@
 Both the X and Y axes are log10-scaled.
-Dotted regression lines represent the linear fits within city population groups, while the solid gray line represent the overall linear fit", fig.height=4}
+Dotted regression lines represent the linear fits within city population groups, while the solid gray line represents the overall linear fit.
+The regression line for the millionaire cities is fitted excluding the outliers (Moscow and Saint Petersburg)", fig.height=4}
 ca_plot
@@
 In general, cities with the top rank by the population in their respective regions tend to have better availability of legal services.
-One possible reason for this is a considerably big size of such cities.
+One possible reason for this is the considerably big size of such cities.
 
 Moscow and Saint Petersburg are two remarkable outliers with extremely high concentration of legal services and a good level of legal services availability.
-Moscow alone accounts for almost one-third of all legal companies employees, and Saint Petersburg accounts for about 10%.
+Moscow alone accounts for almost one-third of all legal companies' employees, and Saint Petersburg accounts for about 10%.
 The position of the corresponding data points relative to the main regression lines suggests that there is an “availability saturation” level, where further concentration has little effect on availability.
@@
 The most prominent cluster of good access to legal services is formed by Moscow and its satellite cities.
-Other notable centers with high availability include Krasnodar and Stavropol in the southern part of Russia's West, Yekaterinburg in the Urals, and Tuymen in Western Siberia.
+Other notable centers with high availability include Krasnodar and Stavropol in the southern part of Russia's West, Yekaterinburg and Chelyabinsk in the Urals, and Tuymen and Novosibirsk in Western Siberia.
 There are regions with limited access to legal services, such as the bordering regions in the south of Siberia, as well as the Komi and Kalmyk Republics.
@@
 Moscow and Saint Petersburg are characterized by the highest concentration and high availability, with Moscow alone accounting for almost one-third of all legal companies employees in the country.
-Empirical data suggests that there may be a saturation limit beyond which the increase of concentration do not have a substantial impact on availability.
-There are multiple paths to achieving high levels of legal services availability, with increasing concentration of legal companies in a city being just one option.
+Empirical data suggests that there may be a saturation limit beyond which the increase in concentration does not have a substantial impact on availability.
+There are multiple paths to achieving high levels of legal services availability, with the increasing concentration of legal companies in a city being just one option.
 Even cities with relatively low concentration can achieve similar levels of availability as larger ones.
 Moscow and its region are a cluster of areas with high legal services availability.
-Similarly, Saint Petersburg, Yekaterinburg, Krasnodar, Tyumen, and Stavropol also have good access to legal services.
+Similarly, Saint Petersburg, Krasnodar, Stavropol, Yekaterinburg, Chelyabinsk, Tyumen, and Novosibirsk also have good access to legal services.
 

### 2024-03-06 legal-companies-concentration-and-availability/Script.R

@@ -48,7 +48,7 @@ data <- left_join(lcc, cp, by = c("oktmo", "city")) %>%
       labels = c(
         "Small (<50k)", "Medium (50–100k)", "Big (100–250k)",
         "Large (250–500k)", "Extra-large (500k–1M)", 
-        "Millionare (>1M)")
+        "Millionaire (>1M)")
     )) %>% 
   filter(
     !(city %in% c("Иннополис", "Кировск", "Бронницы", "Красноармейск"))) %>% 

### 2024-03-12 legal-companies-zipf-law/Paper.Rmd
Start writing paper about Zipf law in distribution of legal companies by cities

---
title: "Зависимость «ранг-размер» в распределении городов России по количеству юридических фирм и их работников"
author: Сёмин П. О.
output: 
  word_document:
    reference: reference.docx
bibliography: biblio.bib
csl: biblio-style.csl
---

<div custom-style="Author">

Пермский государственный национальный исследовательский университет, г. Пермь

</div>

<br>

*Аннотация:*

*Ключевые слова:*

Одной из широко известных общественно-географических закономерностей является правило «ранг-размер», также известное как закон Ципфа (Zipf's Law) [@Zipf2016]. Оно характеризует распределение городов по размеру [@Kwok2014]. Согласно правилу «ранг-размер», при ранжировании городов по убыванию численности населения, численность населения обратно пропорциональна рангу [@Rastvortseva2020]. Существуют многочисленные исследования, в которых закон Цифпа эмпирически проверяется на материале отдельных стран [@Jiang2011; @Rastvortseva2020], регионов внутри страны [@Giesen2010; @Makarova2017], нескольких стран [@Manaeva2018] или всего мира в целом [@Jiang2015] либо выступает в качестве одного из компонентов теоретической модели для анализа расселения и его изменения [@Balash2013; @Kwok2014]. В то же время интересно проверить, применимо ли правило Ципфа к другим характеристикам городов, таким как число тех или иных фирм и количество занятых в той или иной сфере деятельности, например, в юридических услугах. Теоретически это предположение не лишено оснований, так как сфера услуг тесно связана с потребностями населения, а значит, её география и сопутствующие закономерности также должны быть похожи.

В данной статье ставятся два вопроса, касающиеся географии юридических услуг в России. Первый: как связана численность населения городов и число юридических фирм и их работников? Второй: соответствует ли распределение числа юридических фирм и их работников в городах России правилу «ранг-размер»?

Для ответа на вопросы используются данные о численности населения городов России, подготовленные коммерческой аналитической фирмой «HFLabs»[^1] и набор данных о российских малых и средних юридических фирмах, собранный автором статьи. Набор данных получен с помощью самостоятельно разработанного консольного приложения на языке программирования Python *ru-smb-companies*[^2]. В качестве источника исходных сведений о фирмах выступают реестр субъектов малого и среднего предпринимательства, свободно распространяемый Федеральной налоговой службой (ФНС) Россиих[^3], а также два дополнительных набора открытых данных ФНС России: сведения о доходах и расходах организаций[^4] и сведения о среднесписочной численности работников организаций[^5]. Использованы сведения о компаниях за 2021 год. Для выбора юридических фирм из всего множества субъектов МСП осуществлена фильтрация по основному коду ОКВЭД, равному 69.10 (деятельность в области права).

[^1]: URL: https://github.com/hflabs/city (дата обращения: 12.03.2024).
[^2]: URL: https://github.com/PavelSyomin/russian-smb-companies (дата обращения: 12.03.2024).
[^3]: URL: https://www.nalog.gov.ru/opendata/7707329152-rsmp/ (дата обращения: 12.03.2024).
[^4]: URL: https://www.nalog.gov.ru/opendata/7707329152-revexp/ (дата обращения: 12.03.2024).
[^5]: URL: https://www.nalog.gov.ru/opendata/7707329152-sshr2019/ (дата обращения: 12.03.2024).

Из компаний были исключены те, которые отчитались о нулевых доходах или расходах. При нулевой отчётной численности работников предполагалось, что в компании работает один человек. Для проверки гипотез использовались линейная регрессия и визуализация данных. Формула модели линейной регрессии — $log(rank) = k \times log(count) + b + \varepsilon$. Анализ выполнен с помощью языка программирования R и дополнительных пакетов к нему. Исходный код доступен в публичном репозитории[^6].

[^6]: URL: (дата обращения: 12.03.2024).

**Литература**

<div id="refs"></div>

<div custom-style="Title">

The “Rank-Size” Rule in the Distribution of Russian Cities by the Count of Legal Companies and their Employees

</div>

<div custom-style="Author">

Syomin P. O.

Perm State University, Perm

</div>

Abstract:

Keywords:

### 2024-03-13 legal-companies-zipf-law/Paper.Rmd
Finish text of paper about Zipf Law

@@
 
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(
+  echo = FALSE,
+  message = FALSE,
+  warning = FALSE,
+  dpi = 300,
+  fig.id = FALSE,
+  fig.width = 6.6,
+  fig.height = 3.3)
+```
+
+```{r imports, include=FALSE, cache = TRUE}
+source("Script.R")
+```
+
 <div custom-style="Author">
@@
 
-*Аннотация:*
+*Аннотация:* в статье на данных по России исследуется зависимость числа юридических фирм и их работников с населением города, а также проверяется правило «ранг-размер» применительно к распределению городов по числу юридических фирм и их работников.
+Предполагается, что зависимость линейная, а правило выполняется, так как правило типично для системы городского расселения в России, а сфера услуг тесно связана с потребностями местного населения и отражает его географические закономерности.
+Используются открытые данные Федеральной налоговой службы о малых и средних предприятиях и среднесписочной численности работников, преобразованные в табличный геопривязанный набор данных.
+Методы анализа — линейная регрессия и визуализация данных.
+Обнаруженная зависимость между населением и числом фирм или работников близка к линейной (R^2^ около 0.7).
+Правило «ранг-размер» в основном выполняется (R^2^ > 0.9).
 
-*Ключевые слова:*
+*Ключевые слова:* города, юридические услуги, закон Ципфа, ранг-размер, население, число работников, число компаний, открытые данные.
+
+<br>
 
@@
 При нулевой отчётной численности работников предполагалось, что в компании работает один человек.
+Иннополис был исключён из списка городов из-за крайне небольшого числа жителей.
 Для проверки гипотез использовались линейная регрессия и визуализация данных.
-Формула модели линейной регрессии — $log(rank) = k \times log(count) + b + \varepsilon$.
+При построении линейных моделей зависимая и независимая переменные логарифмированы.
 Анализ выполнен с помощью языка программирования R и дополнительных пакетов к нему.
@@
 
-**Литература**
+Зависимость между населением города и числом юридических фирм, зарегистрированных в нём, а также числом работников этих фирм, близка к линейной, что видно на графиках (рис.
+1).
+Следует заметить, что обе оси логарифмированы, чтобы уменьшить асимметрию распределения исходных данных.
+Линейная аппроксимация, судя по расположению точек на графике, не является оптимальной, и ошибка модели при небольшой численности населения довольно велика, несмотря на сравнительно хороший коэффициент детерминации (R^2^ = `r round(count_pop_rsq, 2)` у модели числа компаний от числа жителей, и R^2^ = `r round(empl_pop_rsq, 2)` у модели числа работников от числа жителей; все коэффициенты статистически значимы).
+При росте населения города качество модели улучшается.
+
+Правило «ранг-размер» применительно к числу юридических фирм и их работников в основном подтверждается (рис.
+2).
+Коэффициенты детерминации соответствующих линейных моделей близки к единице (R^2^ = `r round(rank_count_rsq, 2)` в случае с компаниями и R^2^ = `r round(rank_empl_rsq, 2)` в случае с работниками; все коэффициенты статистически значимы), и заметные отклонения от линейности характерны только для населённых пунктов с минимальным числом фирм или занятых в них (специфические «зубцы» в начале линий) или для крупнейших городов (правые нижние части графиков).
+
+Таким образом, количество юридических фирм и занятых в них тесно и линейно связано с численностью населения городов.
+Это хорошо согласуется с предположением о зависимости уровня развития сферы услуг от числа жителей.
+Распределение городов России по количеству юридических компаний и их работников в основном соответствует правилу «ранг-размер».
+
+# Литература
 
 <div id="refs"></div>
+
+```{r figure-1-count-empl-by-population, fig.cap="Рисунок 1.
+Зависимость между населением города и (а) числом юридических фирм, (б) числом работников юридических фирм.
+Красный пунктир — линейная аппроксимация.
+Другие пояснения в тексте"}
+counts_and_population_plot
+```
+
+<br>
+
+```{r figure-2-rank-by-count-empl, fig.cap="Рисунок 2.
+Зависимость между (а) число юридических фирм, (б) числом работников юридических фирм и рангом населённого пункта, рассчитанным исходя из этого числа.
+Красный пунктир — линейная аппроксимация.
+Другие пояснения в тексте"}
+rank_size_plot
+```
+
+<br>
 
@@
 
-Abstract:
+<br>
 
-Keywords:
+Abstract: This paper examines whether the distribution of Russian cities by the count of legal companies and their employees follows Zipf's Law (rank-size rule).
+It also investigates the relationship between the city population and the count of legal companies or their workers.Since this rule has been demonstrated for Russian cities and the geography of services is closely related to the geography of population, the study assumes that the rule is followed.
+The analysis utilizes open data from the Federal Tax Service on small and medium-sized businesses, as well as the average yearly count of employees.
+The study employs linear regression and data visualization as methods.
+The relationship between population and the count of legal companies/workers is linear (R^2^ ≈ 0.7), and the distribution of cities by the count of legal companies/employees generally follows Zipf's Law (R^2^ > 0.9).
+
+Keywords: cities, legal services, Zipf's Law, rank-size, population, count of companies, count of employees, open data.

### 2024-03-14 legal-companies-mapping/Slides.Rmd
Make presentation for cartography forum in Pskov

---
title: "Картографирование малых и средних предприятий в России"
output: powerpoint_presentation
author: "Сёмин Павел Олегович, аспирант ПГНИУ"
date: "14.03.2024"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
library(knitr)
```

```{r imports, cache = TRUE, include = FALSE}
source("helpers/Plots.R")
```

# Задача

- Показать на карте российские малые и средние предприятия (МСП)
- Варианты: все сразу или в определённой сфере деятельности (по ОКВЭД)
- Зачем: как часть анализа данных и как визуализация

# Проблема: нет данных

- ЕГРЮЛ — нет прямого доступа ко всей базе, нет координат
- Яндекс/2ГИС — либо за деньги, либо незаконно парсить
- OSM — неполные данные, сложный формат выгрузок
- Реестр МСП — большой, в сложном формате, без координат, зато бесплатный и полностью доступный

# Реестр МСП

- Доступен в открытых данных ФНС
- 200 Гб в сжатом виде за 2016–2023
- Есть ИНН, адрес регистрации до населённого пункта, коды ОКВЭД
- Можно объединять с другими наборами открытых данных по ИНН (доходы/расходы, работники)
- Сложно обработать (многоуровневый XML в архивах, 2 Тб в распакованном виде), нужно геокодировать адреса, чтобы получить координаты

# Обработка реестра МСП

- Разработано консольное приложение на Python
- Загрузка → Извлечение → Агрегация → Геокодирование
- Агрегация убирает дубли (выгрузки идут ежемесячно в полном виде)
- Геокодирование на базе справочных таблиц → нормализованные адреса + координаты фирм с точностью до населённого пункта (распознано 97% адресов)

# Преобразованный набор данных

```{r dataset-sample}
smb_data %>% 
    filter(
    start_date <= "2021-12-31",
    end_date >= "2021-12-31"
  ) %>% 
  select(tin, settlement, lon, lat) %>% 
  slice_head(n = 10) %>% 
  knitr::kable(
    col.names = c("ИНН", "Населённый пункт", "Долгота", "Широта"))
```

# Тестирование

- Вместо всех фирм — только юридические (так быстрее)
- 2021 год
- Вся Россия
- Число фирм, дохода/расходы, число работников

# Россия, число фирм

```{r russia-legal-firms-count}
settlements_map
```

# Россия, прибыль

```{r russia-legal-firms-profit}
settlements_map_profit
```

# Россия, число работников

```{r russia-legal-firms-empl}
settlements_map_empl
```

# Выводы

- Разработана программа, которая из открытых данных ФНС делает набор данных, готовый для картографирования
- Продемонстрирована её работа

### 2024-03-14 legal-companies-mapping/Slides.Rmd
Draw Crimea on Russia map to avoid legal issues in Russia, shorten text on slides to fix overflow

@@
 - Можно объединять с другими наборами открытых данных по ИНН (доходы/расходы, работники)
-- Сложно обработать (многоуровневый XML в архивах, 2 Тб в распакованном виде), нужно геокодировать адреса, чтобы получить координаты
+- Сложно обработать (многоуровневый XML), нужно геокодировать адреса
 
@@
 - Агрегация убирает дубли (выгрузки идут ежемесячно в полном виде)
-- Геокодирование на базе справочных таблиц → нормализованные адреса + координаты фирм с точностью до населённого пункта (распознано 97% адресов)
+- Геокодирование на базе справочных таблиц → нормализованные адреса + координаты фирм (распознано 97% адресов)
 

### 2024-03-15 legal-companies-zipf-law/Paper.Rmd
Minor improvements in styles, text and biblio

@@
 
-Для ответа на вопросы используются данные о численности населения городов России, подготовленные коммерческой аналитической фирмой «HFLabs»[^1] и набор данных о российских малых и средних юридических фирмах, собранный автором статьи.
+Для ответа на вопросы используются данные о численности населения городов России, подготовленные коммерческой аналитической фирмой «HFLabs»[^1], и набор данных о российских малых и средних юридических фирмах, сделанный автором статьи.
 Набор данных получен с помощью самостоятельно разработанного консольного приложения на языке программирования Python *ru-smb-companies*[^2].
-В качестве источника исходных сведений о фирмах выступают реестр субъектов малого и среднего предпринимательства, свободно распространяемый Федеральной налоговой службой (ФНС) Россиих[^3], а также два дополнительных набора открытых данных ФНС России: сведения о доходах и расходах организаций[^4] и сведения о среднесписочной численности работников организаций[^5].
+В качестве источника исходных сведений о фирмах выступают реестр субъектов малого и среднего предпринимательства (МСП), свободно распространяемый Федеральной налоговой службой (ФНС) Россиих[^3], а также два дополнительных набора открытых данных ФНС России: сведения о доходах и расходах организаций[^4] и сведения о среднесписочной численности работников организаций[^5].
 Использованы сведения о компаниях за 2021 год.
@@
 При построении линейных моделей зависимая и независимая переменные логарифмированы.
+Порог статистической значимости выбран равным 0.05.
 Анализ выполнен с помощью языка программирования R и дополнительных пакетов к нему.
@@
 Следует заметить, что обе оси логарифмированы, чтобы уменьшить асимметрию распределения исходных данных.
-Линейная аппроксимация, судя по расположению точек на графике, не является оптимальной, и ошибка модели при небольшой численности населения довольно велика, несмотря на сравнительно хороший коэффициент детерминации (R^2^ = `r round(count_pop_rsq, 2)` у модели числа компаний от числа жителей, и R^2^ = `r round(empl_pop_rsq, 2)` у модели числа работников от числа жителей; все коэффициенты статистически значимы).
+Линейная аппроксимация, судя по расположению точек на графике, не является оптимальной, и ошибка модели при небольшой численности населения довольно велика, несмотря на сравнительно хороший коэффициент детерминации (R^2^ = `r round(count_pop_rsq, 2)` у модели числа компаний от числа жителей, и R^2^ = `r round(empl_pop_rsq, 2)` у модели числа работников от числа жителей; все коэффициенты моделей статистически значимы).
 При росте населения города качество модели улучшается.
@@
 2).
-Коэффициенты детерминации соответствующих линейных моделей близки к единице (R^2^ = `r round(rank_count_rsq, 2)` в случае с компаниями и R^2^ = `r round(rank_empl_rsq, 2)` в случае с работниками; все коэффициенты статистически значимы), и заметные отклонения от линейности характерны только для населённых пунктов с минимальным числом фирм или занятых в них (специфические «зубцы» в начале линий) или для крупнейших городов (правые нижние части графиков).
+Коэффициенты детерминации соответствующих линейных моделей близки к единице (R^2^ = `r round(rank_count_rsq, 2)` в случае с компаниями и R^2^ = `r round(rank_empl_rsq, 2)` в случае с работниками; все коэффициенты моделей статистически значимы), и заметные отклонения от линейности характерны только для населённых пунктов с минимальным числом фирм или занятых в них (специфические «зубцы» в начале линий) или для крупнейших городов (правые нижние части графиков).
 
@@
 <div id="refs"></div>
+
+<br>
 

### 2024-03-27 legal-companies-regional-specialization/Paper.Rmd
Draft paper on regional specialization

---
title: "Legal Services Market in Russia: Specialization of Regions"
author: "Pavel O. Syomin"
date: "PhD Student at Perm State University"
output:
  word_document:
    reference: reference.docx
bibliography: biblio.bib
csl: biblio-style.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  dpi = 300,
  fig.id = FALSE,
  fig.width = 6.6,
  fig.height = 5)
```

```{r imports, include=FALSE, cache = TRUE}
source("Script.R")
```

<br>

<div custom-style="Abstract">

Abstract: ...

Keywords: ...

</div>

<br>

**Introduction.** Analysis of spatial distribution of various sectors of economy is a typical task of a research in economical geography. In Russia, due to its federal structure, this issue often takes a form of studying a regional specialization, where each region equals the constituent entity of the state. There are diverse ways and methods of regional specialization analysis, ranging from descriptive ones to modern GIS or machine learning solutions. In particular, the quantitative methodics proposed in the recent Atlas of Economic Specialization of Russian Regions [@Atlas2021] prepared by the researchers from Higher Schools of Economics (HSE) is of high interest, because it is relatively modern, data-driven, theoretically grounded and reproducible. Unfortunately, only a subset of sectors of Russia's economy has been analyzed by the authors of the Atlas, and some important economic areas are missing. One example of such missing sectors is a legal services market. Despite the absence of research on legal services in Russia (with one notable exception of the analytical review by the members of Institute for the Rule of Law [@Moiseeva2016]), the geographical study of legal market in Russia is important, because the development of legal services, on the one hand, reflects the business development and economic activity, and on the other hand, stimulates the social and economic development by increasing the protection of citizens and business rights and freedoms. So, this short study tries to analyze the regional specialization of Russia with respect to the legal services.

**Objective.** The aim of the research is twofold. First, its purpose is to detect Russia's regions with the specialization on legal services. Second, the goal is to vizualize this specialization on a tile map.

**Data.** For the study, the open data on Russian small and medium-sized businesses (SMBs) distributed by the Federal Tax Service (FTS)[^1] was used. In addition, the FTS open data on the average number of employees of organizations[^2] was utilized. From these source datasets, the information about the legal firms (companies and individual entrepreneurs) that existed in 2021 was extracted, including the region of registration and the number of employees. To select legal firms only, the filter by the activity code equal to 69.10 was applied. The data processing was performed with the command-line Python tool *ru-smb-companies*[^3] developed by the author of this paper. Apart from the firms data, the official statistical information about the average yearly count of employees by region in 2021 provided by the Federal Statistics Service of Russia (Rosstat) was used[^4].

**Methods.** The study follows the technique proposed in the Atlas and uses two metrics to estimate the specialization of regions. The first metric is called concentration, and it is calculated as the count of employees of legal companies in the region divided by the total number of employees of legal companies in all the regions. The second metric is called localization, and it is calculated by dividing the concentration by the ratio of the count of all employees in the region and the count of all employees in Russia. In rough terms, the concentration shows the share of the region in the total number of legal workers, while the localization shows whether this share is greater or less than the share of the region in the Russia's labor force. After the metrics are calculated, the regions are ranked by the localization, and top-16 regions (approximately 20%) are treated as regions where the legal services are the sector of regional importance. Next, the regions are ranked by the concentration, and the cumulative concentration is calculated. The regions with cumulative concentration below 0.8 are marked as the regions where the legal services is a sector of national importance. Some regions are present in both two groups. To draw the result, a tile map template by Dianov [^5] is used with a custom implementation in R programming language based on Yu. Tuckachev's code [^6]. The analysis is done with R as well, and the source code for the research and paper is published in the author's repository on GitHub[^7].

**Results and discussion.** The tile map of regional specialization of Russia with respect to the legal services is shown on Figure 1. Several clusters are distinguishable, including (1) Moscow and (2) Saint Petersburg where legal services are both of national and regional importance, (3) the Western Plain South with a group of regions with national importance of legal services (Rostov, Stavropol, Voronezh, Byelgorod) and one region of joint importance (Krasnodar), (4) the Volga belt formed by a set of regions with regional (Voplgograd, Saratov, Perm) or joint (Samara, Tatarstan, Udmurtia) importance, (5) the Urals cluster with Sverdlovsk and Chelyabinsk regions forming the core and Bashkir republic on the peripheries, (6) Siberian cluster of three regions with both national and regional importance (Tuymen, Omsk, Novosibirsk) and four accompanying regions of regional importance (Altai, Keverovo, Krasnoyarsk, Irkutsk). Moscow region, Nizhniy Novgorod and Primorsky krai also may be treated as single-region clusters. Some regions with local importance of legal services only (e. g. Kaliningrad, Tver, Ivanovo, Kamchatka, Murmansk) does not belong to any of the clusters. The reasons for such a spatial distribution require further investigation. However, it is obvious the Moscow and Saint Petersburg concentrate legal services and localize them due to the top levels of economic developent and concentration. Other four main clusters (3–6) also can be hypothetically explained by the level of economic development. The same hypothesis may be true for the three single-region clusters. It is also important to mention that most of the specialized regions have at least one appeals or cassation court (both general or commercial), thus the concentration or localization of legal services in these regions may be partially explained by the presence of such courts that act as “gravitation centers” for both legal services customers and providers, though this hypothesis needs more thorough testing.

```{r figure-1-map, fig.cap="Figure 1. Tile map of Russia showing the specialization of regions on legal services"}
map
```

**Conclusion.** The specialization of regions on legal services in Russia was measured using concentration and localization metrics based on the number of employees. Six main clusters of legal services and three single-region clusters were distinguished. The result may be linked to the level of economic development of the respective regions and to the presence of appeal or cassation courts in the regions of specialization.

<br>

# References

<div id="refs"></div>

### 2024-03-27 legal-companies-regional-specialization/Script.R

@@ -1,0 +1,52 @@
+library(dplyr)
+library(ggplot2)
+library(readr)
+library(tidyr)
+
+panel <- read_csv("../../ru-smb-companies/legal/panel.csv")
+empl <- read_csv("empl.csv")
+tiles <- read_csv("russia-tiles.csv")
+
+spec <- panel %>% 
+  filter(year == 2021) %>% 
+  drop_na(region) %>% 
+  replace_na(list(employees_count = 0)) %>% 
+  mutate(employees_count = replace(employees_count, employees_count == 0, 1)) %>% 
+  count(region, wt = employees_count, name = "legal_empl") %>% 
+  right_join(empl, by = "region") %>% 
+  rename(total_empl = count) %>% 
+  mutate(
+    concentration = legal_empl / sum(legal_empl, na.rm = TRUE),
+    empl_share = total_empl / sum(total_empl, na.rm = TRUE),
+    localization = concentration / empl_share
+  ) %>% 
+  arrange(-concentration) %>% 
+  mutate(cum_concentraion = cumsum(concentration)) %>% 
+  arrange(-localization) %>% 
+  mutate(loc_rank = row_number()) %>% 
+  mutate(
+    is_national = cum_concentraion <= 0.8,
+    is_regional = loc_rank < nrow(.) * .2,
+    rank = as.numeric(is_national) * 2 + as.numeric(is_regional),
+    rank = factor(rank, labels = c(NA, "Local", "National", "Both"))
+  ) %>% 
+  select(region, is_national, is_regional, rank)
+
+
+map <- left_join(spec, tiles, by = c("region" = "name")) %>% 
+  ggplot(aes(x = col, y = -row, fill = rank)) +
+  geom_raster() +
+  geom_text(aes(label = code), family = "Times New Roman") +
+  scale_fill_brewer(
+    palette = "Blues",
+    na.value = "grey90",
+    name = "Legal services importance",
+    labels = c("Local", "National", "Both", "No")) +
+  coord_equal() +
+  theme_void(base_family = "Times New Roman") +
+  theme(
+    legend.position = c(.9, .1),
+    legend.justification = c(1, 0),
+    legend.direction = "horizontal"
+  )
+map

### 2024-03-29 legal-companies-regional-specialization/Paper.Rmd
Language editing and proofreading, improving introduction, adjusting map

@@
 ---
-title: "Legal Services Market in Russia: Specialization of Regions"
+title: "Legal Services Market in Russia: Regional Specialization"
 author: "Pavel O. Syomin"
@@
   fig.id = FALSE,
-  fig.width = 6.6,
-  fig.height = 5)
+  fig.width = 7,
+  fig.height = 4)
 ```
@@
 
-Abstract: ...
+Abstract: This paper focuses on the spatial distribution of the legal services market in Russia.
+The aim is to identify the most specialized regions (constituent entities) using a quantitative technique based on the relative numbers of employees.
+The research utilizes open data distributed by the Federal Tax Service of Russia and the Federal Statistics Service.
+The regional specialization is measured using two metrics: concentration and localization.
+High concentration indicates regions with legal services of national importance, while high localization marks regions where legal services have local importance.
+The study's results reveal several clusters of legal services.
+These clusters are likely linked to the high level of economic activity in the regions.
+The location of appeal and cassation courts may also partially explain the clusters.
 
-Keywords: ...
+Keywords: specialization, legal services, regions, economic geography, concentration, localization, legal market, open data.
 
@@
 
-**Introduction.** Analysis of spatial distribution of various sectors of economy is a typical task of a research in economical geography.
-In Russia, due to its federal structure, this issue often takes a form of studying a regional specialization, where each region equals the constituent entity of the state.
-There are diverse ways and methods of regional specialization analysis, ranging from descriptive ones to modern GIS or machine learning solutions.
-In particular, the quantitative methodics proposed in the recent Atlas of Economic Specialization of Russian Regions [@Atlas2021] prepared by the researchers from Higher Schools of Economics (HSE) is of high interest, because it is relatively modern, data-driven, theoretically grounded and reproducible.
-Unfortunately, only a subset of sectors of Russia's economy has been analyzed by the authors of the Atlas, and some important economic areas are missing.
-One example of such missing sectors is a legal services market.
-Despite the absence of research on legal services in Russia (with one notable exception of the analytical review by the members of Institute for the Rule of Law [@Moiseeva2016]), the geographical study of legal market in Russia is important, because the development of legal services, on the one hand, reflects the business development and economic activity, and on the other hand, stimulates the social and economic development by increasing the protection of citizens and business rights and freedoms.
-So, this short study tries to analyze the regional specialization of Russia with respect to the legal services.
+**Introduction.** Analysis of economic concentration and regional specialization is a common research task in economic geography.
+It helps to understand the current state of the territory and discover opportunities for its development [@Romanov2015].
+There are various methods for studying concentration and specialization, including calculating coefficients [@Bryantseva2022] and analysis of graphs [@Blanutsa2020].
+In particular, the quantitative technique used in the Atlas of Economic Specialization of Russian Regions [@Atlas2021] is of high interest, because it is modern, data-driven, theoretically grounded, and reproducible.
+The authors of the Atlas have only analyzed a subset of sectors in Russia's economy, and some important economic areas are missing.
+One example of a missing sector is the legal services market.
 
-**Objective.** The aim of the research is twofold.
-First, its purpose is to detect Russia's regions with the specialization on legal services.
-Second, the goal is to vizualize this specialization on a tile map.
+There is a lack of geographical research on legal services in Russia except for one analytical review by the members of the Institute for the Rule of Law [@Moiseeva2016].
+However, some argue that studying the spatial aspects of legal service provision is important [@Joseph1982].
+This is because the development of legal services reflects both business and economic activity and also stimulates social and economic development by increasing the protection of citizens' and businesses' rights and freedoms.
+Therefore, this research aims to analyze the regional distribution of legal services in Russia and identify the regions that specialize in providing such services.
 
-**Data.** For the study, the open data on Russian small and medium-sized businesses (SMBs) distributed by the Federal Tax Service (FTS)[^1] was used.
-In addition, the FTS open data on the average number of employees of organizations[^2] was utilized.
-From these source datasets, the information about the legal firms (companies and individual entrepreneurs) that existed in 2021 was extracted, including the region of registration and the number of employees.
-To select legal firms only, the filter by the activity code equal to 69.10 was applied.
-The data processing was performed with the command-line Python tool *ru-smb-companies*[^3] developed by the author of this paper.
-Apart from the firms data, the official statistical information about the average yearly count of employees by region in 2021 provided by the Federal Statistics Service of Russia (Rosstat) was used[^4].
+**Objective.** The research has a twofold aim.
+Firstly, it aims to identify the regions in Russia that specialize in legal services.
+Secondly, it aims to visualize this specialization on a tile-grid map.
 
-**Methods.** The study follows the technique proposed in the Atlas and uses two metrics to estimate the specialization of regions.
-The first metric is called concentration, and it is calculated as the count of employees of legal companies in the region divided by the total number of employees of legal companies in all the regions.
-The second metric is called localization, and it is calculated by dividing the concentration by the ratio of the count of all employees in the region and the count of all employees in Russia.
-In rough terms, the concentration shows the share of the region in the total number of legal workers, while the localization shows whether this share is greater or less than the share of the region in the Russia's labor force.
-After the metrics are calculated, the regions are ranked by the localization, and top-16 regions (approximately 20%) are treated as regions where the legal services are the sector of regional importance.
-Next, the regions are ranked by the concentration, and the cumulative concentration is calculated.
-The regions with cumulative concentration below 0.8 are marked as the regions where the legal services is a sector of national importance.
+**Data.** This study utilizes the open data on Russian small and medium-sized businesses (SMBs) distributed by the Federal Tax Service (FTS)[^1].
+Additionally, the FTS open data on the average number of employees of organizations[^2] was used.
+The author extracted information about legal firms (companies and individual entrepreneurs) that existed in 2021, including their region of registration and number of employees, from the source datasets.
+To filter out non-legal firms, the activity code was restricted to 69.10.
+The data was processed using the command-line Python tool *ru-smb-companies*[^3], developed by the author of this paper.
+In addition to the company's data, official statistical information from the Federal Statistics Service of Russia (Rosstat) was utilized to determine the average yearly employee count by region in 2021[^4].
+
+[^1]: URL: https://www.nalog.gov.ru/opendata/7707329152-rsmp/ (accessed March 28, 2024).
+[^2]: URL: https://www.nalog.gov.ru/opendata/7707329152-sshr2019/ (accessed March 28, 2024).
+[^3]: URL: https://github.com/PavelSyomin/russian-smb-companies (accessed March 28, 2024).
+[^4]: URL: https://fedstat.ru/indicator/34051 (accessed March 28, 2024).
+
+**Methods.** The study follows the technique proposed in the Atlas [@Atlas2021] and uses two metrics to estimate regional specialization.
+The first metric, *concentration*, is calculated as the number of legal companies employees in the region divided by the total number of legal companies employees in all regions.
+The second metric, *localization*, is calculated by dividing the concentration by the ratio of the number of all employees in the region and the count of all employees in Russia.
+It can be said that concentration indicates the region's share of the total number of legal workers, while localization shows whether this share is greater or less than the region's share in Russia's labor force.
+
+In certain studies, the localization metric is used to conclude about the specialization [@Romashina2019, @Bryantseva2022].
+However, the authors of the Atlas propose a two-dimensional approach that combines concentration and specialization as follows.
+After calculating the metrics, the regions are ranked by localization, and the top 16 regions (approximately 20%) are considered to be areas where the legal services are of local importance.
+Next, the regions are ranked by concentration, and the cumulative concentration is calculated.
+Regions with a cumulative concentration below 0.8 are identified as areas where the legal services are of national importance, that is, these regions collectively account for 80% of the country's legal workers.
 Some regions are present in both two groups.
-To draw the result, a tile map template by Dianov [^5] is used with a custom implementation in R programming language based on Yu.
-Tuckachev's code [^6].
-The analysis is done with R as well, and the source code for the research and paper is published in the author's repository on GitHub[^7].
 
-**Results and discussion.** The tile map of regional specialization of Russia with respect to the legal services is shown on Figure 1.
-Several clusters are distinguishable, including (1) Moscow and (2) Saint Petersburg where legal services are both of national and regional importance, (3) the Western Plain South with a group of regions with national importance of legal services (Rostov, Stavropol, Voronezh, Byelgorod) and one region of joint importance (Krasnodar), (4) the Volga belt formed by a set of regions with regional (Voplgograd, Saratov, Perm) or joint (Samara, Tatarstan, Udmurtia) importance, (5) the Urals cluster with Sverdlovsk and Chelyabinsk regions forming the core and Bashkir republic on the peripheries, (6) Siberian cluster of three regions with both national and regional importance (Tuymen, Omsk, Novosibirsk) and four accompanying regions of regional importance (Altai, Keverovo, Krasnoyarsk, Irkutsk).
-Moscow region, Nizhniy Novgorod and Primorsky krai also may be treated as single-region clusters.
-Some regions with local importance of legal services only (e.
-g.
-Kaliningrad, Tver, Ivanovo, Kamchatka, Murmansk) does not belong to any of the clusters.
-The reasons for such a spatial distribution require further investigation.
-However, it is obvious the Moscow and Saint Petersburg concentrate legal services and localize them due to the top levels of economic developent and concentration.
-Other four main clusters (3–6) also can be hypothetically explained by the level of economic development.
-The same hypothesis may be true for the three single-region clusters.
-It is also important to mention that most of the specialized regions have at least one appeals or cassation court (both general or commercial), thus the concentration or localization of legal services in these regions may be partially explained by the presence of such courts that act as “gravitation centers” for both legal services customers and providers, though this hypothesis needs more thorough testing.
+To visualize the results, a tile-grid map of Russia is used.
+Tile-grid maps are a relatively new phenomenon in Russian cartography, primarily used in popular science and journalism.
+However, they have also been suggested for potential use in spatial research [@Elatskov2020].
+The map is drawn using a tile-grid map template created by I.
+Dianov[^5] and a custom implementation in the R programming language based on Yu.
+Tuckachev's code[^6].
+The analysis was also conducted using R, and the source code for the research and paper has been published in the author's GitHub repository[^7].
 
-```{r figure-1-map, fig.cap="Figure 1. Tile map of Russia showing the specialization of regions on legal services"}
+[^5]: URL: https://dianov.org/all/plitochnaya-karta-rossii/ (accessed March 28, 2024).
+[^6]: URL: https://gist.github.com/tukachev/c3262242a20b3050e0459b4e3afc3a51 (accessed March 28, 2024).
+[^7]: URL: https://github.com/PavelSyomin/ru-smb-companies-papers/tree/main/legal-companies-regional-specialization (accessed March 28, 2024).
+
+**Results and discussion.** Figure 1 shows the tile-grid map of regional specialization in legal services.
+Several clusters are distinguishable based on the importance of legal services.
+These include (1) Moscow (with the surrounding Moscow oblast), (2) Saint Petersburg, (3) the group of southern regions (Krasnodar and Stavropol krai, Voronezh, Belgorod, and Rostov oblasti), (4) the Volga belt formed by Volgograd, Saratov, Samara oblasti, Perm krai, Tatarstan and Udmurt republics, (5) the Urals cluster with Sverdlovsk and Chelyabinsk oblasti as the core and Bashkir republic on the peripheries, (6) Siberian row (Tuymen, Omsk, Novosibirsk, Keverovo oblasti, Altai and Krasnoyarsk krai, Irkutsk oblast).
+Nizhniy Novgorod oblast and Primorsky krai can be treated as single-region clusters.
+Some regions, such as Kaliningrad, Tver, Ivanovo, Murmansk oblasti, and Kamchatka krai, have legal services of local importance that are not part of any cluster.
+
+Further investigation is required to determine the reasons for the observed spatial patterns.
+However, legal services are likely concentrated in Moscow and Saint Petersburg due to their high levels of economic development and concentration.
+The other clusters can also be explained by their level of economic development.
+The same hypothesis may be true for the two single-region clusters.
+Most specialized regions have at least one ordinary or commercial appeals or cassation court.
+This concentration or localization of legal services in these regions may be partially explained by the presence of such courts, which act as “gravitational centers” for both legal services customers and providers.
+However, this hypothesis requires further testing.
+
+<br>
+
+```{r figure-1-map, fig.cap="Figure 1. Tile-grid map of Russia showing the specialization of regions in legal services"}
 map
@@
 
-**Conclusion.** The specialization of regions on legal services in Russia was measured using concentration and localization metrics based on the number of employees.
-Six main clusters of legal services and three single-region clusters were distinguished.
-The result may be linked to the level of economic development of the respective regions and to the presence of appeal or cassation courts in the regions of specialization.
+<br>
+
+**Conclusion.** The concentration and localization of legal services in Russia were measured using metrics based on the number of employees.
+The study identified six main clusters of legal services and two single-region clusters.
+The result may be linked to the level of economic development of the respective regions.
+The presence of an appeal or cassation court may also be connected with the region's specialization in legal services.
 

### 2024-03-29 legal-companies-regional-specialization/Script.R

@@ -36,17 +36,18 @@ spec <- panel %>%
 map <- left_join(spec, tiles, by = c("region" = "name")) %>% 
   ggplot(aes(x = col, y = -row, fill = rank)) +
   geom_raster() +
-  geom_text(aes(label = code), family = "Times New Roman") +
+  geom_text(aes(label = code_en), family = "Times New Roman", size = 3) +
   scale_fill_brewer(
     palette = "Blues",
     na.value = "grey90",
     name = "Legal services importance",
     labels = c("Local", "National", "Both", "No")) +
   coord_equal() +
+  guides(fill = guide_legend(title.position = "top")) +
   theme_void(base_family = "Times New Roman") +
   theme(
     legend.position = c(.9, .1),
     legend.justification = c(1, 0),
     legend.direction = "horizontal"
-  )
+  ) 
 map

### 2024-03-30 legal-companies-regional-specialization/Paper.Rmd
Minor improvements

@@
 The aim is to identify the most specialized regions (constituent entities) using a quantitative technique based on the relative numbers of employees.
-The research utilizes open data distributed by the Federal Tax Service of Russia and the Federal Statistics Service.
+The research utilizes open data distributed by the Federal Tax Service and the Federal Statistics Service of Russia.
 The regional specialization is measured using two metrics: concentration and localization.
@@
 **Data.** This study utilizes the open data on Russian small and medium-sized businesses (SMBs) distributed by the Federal Tax Service (FTS)[^1].
-Additionally, the FTS open data on the average number of employees of organizations[^2] was used.
+Additionally, the FTS open data on the average number of employees of organizations[^2] is used.
 The author extracted information about legal firms (companies and individual entrepreneurs) that existed in 2021, including their region of registration and number of employees, from the source datasets.
-To filter out non-legal firms, the activity code was restricted to 69.10.
+To filter out non-legal firms, the main activity code was restricted to 69.10.
 The data was processed using the command-line Python tool *ru-smb-companies*[^3], developed by the author of this paper.
-In addition to the company's data, official statistical information from the Federal Statistics Service of Russia (Rosstat) was utilized to determine the average yearly employee count by region in 2021[^4].
+In addition to the company's data, official statistical information from the Federal Statistics Service of Russia (Rosstat) was utilized to determine the average yearly employees count by region in 2021[^4].
 
@@
 
-In certain studies, the localization metric is used to conclude about the specialization [@Romashina2019, @Bryantseva2022].
+In certain studies, only the localization metric is used to conclude about the specialization [@Romashina2019, @Bryantseva2022].
 However, the authors of the Atlas propose a two-dimensional approach that combines concentration and specialization as follows.
@@
 Several clusters are distinguishable based on the importance of legal services.
-These include (1) Moscow (with the surrounding Moscow oblast), (2) Saint Petersburg, (3) the group of southern regions (Krasnodar and Stavropol krai, Voronezh, Belgorod, and Rostov oblasti), (4) the Volga belt formed by Volgograd, Saratov, Samara oblasti, Perm krai, Tatarstan and Udmurt republics, (5) the Urals cluster with Sverdlovsk and Chelyabinsk oblasti as the core and Bashkir republic on the peripheries, (6) Siberian row (Tuymen, Omsk, Novosibirsk, Keverovo oblasti, Altai and Krasnoyarsk krai, Irkutsk oblast).
-Nizhniy Novgorod oblast and Primorsky krai can be treated as single-region clusters.
-Some regions, such as Kaliningrad, Tver, Ivanovo, Murmansk oblasti, and Kamchatka krai, have legal services of local importance that are not part of any cluster.
+These include (1) Moscow (with the surrounding Moscow region), (2) Saint Petersburg, (3) the group of southern regions (Krasnodar and Stavropol territories, Voronezh, Belgorod, and Rostov regions), (4) the Volga belt formed by Volgograd, Saratov, Samara regions, Perm territory, Tatarstan and Udmurt republics, (5) the Urals cluster with Sverdlovsk and Chelyabinsk regions as the core and Bashkir republic on the peripheries, (6) Siberian row (Tyumen, Omsk, Novosibirsk, Kemerovo regions, Altai and Krasnoyarsk territories, Irkutsk region).
+Nizhniy Novgorod region and Primorsky territory can be treated as single-region clusters.
+Some regions, such as Kaliningrad, Tver, Ivanovo, Murmansk regions, and Kamchatka territory, have legal services of local importance and are not a part of any cluster.
 
@@
 Most specialized regions have at least one ordinary or commercial appeals or cassation court.
-This concentration or localization of legal services in these regions may be partially explained by the presence of such courts, which act as “gravitational centers” for both legal services customers and providers.
+The concentration or localization of legal services in these regions may be partially explained by the presence of such courts, which act as “gravitational centers” for both legal services customers and providers.
 However, this hypothesis requires further testing.

### 2024-04-04 legal-companies-migration/Paper.Rmd
Edit paper after review: replace two maps with a single tile-grid plot, add some text, change title, other minor improvements

@@
 title: "УДК 911.3"
-subtitle: "Анализ миграции юридических компаний в России в 2016–2023 гг. по открытым данным налоговой службы"
+subtitle: "Релокация юридических компаний в России в 2016–2023 гг."
 author: Сёмин П. О.
@@
 
-**Аннотация.** В статье представлены результаты анализа миграции юридических фирм в России в 2016–2023 гг.
-Миграция компании понимается как изменение адреса регистрации.
+**Аннотация.** В статье представлены результаты анализа релокации юридических фирм в России в 2016–2023 гг.
+Релокация (миграция) компании понимается как изменение адреса регистрации.
 Исследование основано на открытых данных Федеральной налоговой службы (ФНС России), преобразованных в панельный геопривязанный табличный набор данных.
 Методы анализа — описательная статистика, статистический тест Фишера, картографирование, визуализация данных.
-Показано, что миграции юридических фирм в России — редкое явление.
+Показано, что релокация юридических фирм в России — редкое явление.
 Как правило, компании перемещаются на небольшое расстояние (10–50 км).
@@
 Малые предприятия меняют адрес статистически значимо чаще, чем микропредприятия.
-Пространственная картина миграций разнообразная и не характеризуется чёткими тенденциями.
+Пространственная картина релокаций разнообразная и не характеризуется чёткими тенденциями.
 Москва лидирует по оттоку фирм, а Московская область — по их притоку.
@@
 
-**Ключевые слова:** миграция компаний, малое и среднее предпринимательство, юридические компании, юридические услуги, география услуг, анализ данных, открытые данные, ФНС России
+**Ключевые слова:** релокация компаний, миграция компаний, малое и среднее предпринимательство, юридические компании, юридические услуги, география услуг, анализ данных, открытые данные, ФНС России
 
@@
 <div custom-style="Subtitle">
-Analysis of Migration of Russian Legal Companies in 2016–23 Based on Tax Service Open Data
+Relocation of Russian Legal Companies in 2016–23
 </div>
@@
 
-**Abstract.** The paper features the results of analysis of migrations of Russian legal companies in 2016–23.
-Migration here is defined as a change in legal address of a company.
+**Abstract.** The paper features the results of analysis of relocation of Russian legal companies in 2016–23.
+Relocation, or migration, here is defined as a change in legal address of a company.
 The research is based on open data published by Federal Tax Service of Russia that has been processed to make a panel geo-referenced tabular dataset.
 Descriptive statistics, exact Fisher's test, data visualization, and mapping are used as methods.
-It is found that migration of legal companies is relatively a rare phenomenon.
+It is found that relocation of legal companies is relatively a rare phenomenon.
 Typically, businesses move to a short distance (10–50 km), and two thirds of all the migrations are intra-regional, i.e.
 within the constituent entity of Russia.
-Small companies are more likely to migrate compared to microbusinesses.
+Small companies are more likely to relocate compared to microbusinesses.
 The spatial patterns and migration flows are diverse and have few remarkable tendencies.
@@
 
-**Keywords:** companies migration, small and medium-sized businesses, legal companies, legal service, service geography, data analysis, open data, FTS of Russia
+**Keywords:** companies migration, relocation, small and medium-sized businesses, legal companies, legal service, service geography, data analysis, open data, FTS of Russia
 
@@
 Во-вторых, сфера юридических услуг сравнительно слабо связано с набором «традиционных» географических факторов размещения, «привязывающих» фирмы к их первоначальному местонахождению, таких как минеральные, почвенные, лесные, климатические, энергетические ресурсы, и это повышает шанс найти закономерности, отличающиеся от классических.
-В-третьих, юридические услуги важны для бизнеса, и перемещение фирм, оказывающих их, косвенно связано с экономической активностью в целом.
+В-третьих, юридические услуги часто рассматриваются как одна из разновидностей producer services [@Hansen1993], которые положительно влияют на региональное развитие [@Hansen1990] или выступают как лимитирующий фактор [@Hitchens1997], из-за чего без надлежащего юридического обслуживания экономическое развитие, особенно малых фирм, замедляется.
 В-четвёртых, выбор одной конкретной группы компаний упрощает анализ, поскольку уменьшает количество исходных данных, которых довольно много.
@@
 Анализ проводится с помощью методов описательной статистики, статистических тестов, визуализации данных и картографирования.
-Он включает в себя 11 вопросов, некоторые из которых взаимосвязаны.
+Он включает в себя 10 вопросов, некоторые из которых взаимосвязаны.
 Каждому вопросу отводится подраздел в «Результатах и их обсуждении».
@@
 Далее анализируется распределение миграций по расстоянию (3) и по времени (4).
-Следующий шаг — картографирование распределения регионов (5) и населённых пунктов (6) по сальдо миграции.
-После этого изучается соотношение внутри- и межрегиональной миграции (7).
-Далее отдельно рассматриваются внутрирегиональная миграция: определяются пути миграции и их количественные характеристики (8) и анализируется сочетание централизации и децентрализации (9).
-Аналогичные вопросы поднимаются применительно к межрегиональной миграции (10 и 11).
+Следующий шаг — картографирование распределения регионов (5) по сальдо миграции с учётом данных по населённым пунктам на территории соответствующих регионов.
+После этого изучается соотношение внутри- и межрегиональной миграции (6).
+Далее отдельно рассматриваются внутрирегиональная миграция: определяются пути миграции и их количественные характеристики (7) и анализируется сочетание централизации и децентрализации (8).
+Аналогичные вопросы поднимаются применительно к межрегиональной миграции (9 и 10).
 
@@
 Централизация внутри региона рассматривается как перемещение в административный центр субъекта федерации.
+Сальдо релокации рассчитывается как разность числа фирм, появившихся на территории, и числа фирм, покинувших территорию, так что положительное сальдо соответствует притоку фирм, а отрицательное — оттоку.
+За основу картосхемы взята плиточная (tile-grid [@Elatskov2020]) карта России, автором которой является И.
+Дианов[^5], в реализации Ю.
+Тукачева[^6].
 С технической точки зрения исследование реализовано с помощью языка программирования R с дополнительными пакетами.
-Исходный код доступен в репозитории статьи[^5].
+Исходный код доступен в репозитории статьи[^7].
 
-[^5]: https://github.com/PavelSyomin/ru-smb-companies-papers/tree/main/legal-companies-migration
+[^5]: https://dianov.org/all/plitochnaya-karta-rossii/
+[^6]: https://gist.github.com/tukachev/c3262242a20b3050e0459b4e3afc3a51
+[^7]: https://github.com/PavelSyomin/ru-smb-companies-papers/tree/main/legal-companies-migration
 
@@
 
-```{r fig-2-timeline, fig.cap="Рис. 2. Количество миграций за каждый месяц с августа 2016 г. по ноябрь 2023 г. <br> Fig. 2. Migration count for each month from August 2016 to November 2023"}
+```{r fig-2-timeline, fig.cap="Рис. 2. Количество миграций за каждый месяц с октября 2016 г. по октябрь 2023 г. <br> Fig. 2. Migration count for each month from October 2016 to October 2023"}
 mig_timeline_plot
@@
 
-Картограмма (рис.
-4) показывает сальдо миграции по регионам (субъектам федерации).
-Однозначной тенденции не прослеживается, но заметно, что в Московскую область переехало больше всего фирм.
-Сама Москва при этом — абсолютный лидер по оттоку фирм.
+«Плиточная» (tile-grid) карта России (рис.
+4) показывает сводную статистику релокации компаний по регионам и населённым пунктам внутри них.
+Однозначной тенденции не прослеживается: видно, что часть субъектов федерации теряет юридические фирмы, но другая часть их приобретает, причём количество регионов в двух группах приблизительно равное.
+Абсолютные значения сальда релокации невелики (как правило, около нуля, изредка приближаясь к одному-двум десяткам).
+Из этого правила есть два заметных исключения: в Московскую область переехало больше всего фирм, а сама Москва при этом — абсолютный лидер по оттоку фирм.
 Это согласуется, в частности, с данными из Польши, где также характерен уход компаний из столицы в прилегающие районы [@Cybulska2020].
-В остальном сальдо миграции компаний распределено по регионам хаотично, при этом абсолютные значения (число переехавших организаций) невелики — обычно до 25.
 
-```{r fig-4-mig-by-region-plot, fig.cap="Рис. 4. Сальдо миграции юридических фирм по регионам <br> Fig. 4. Net migration rate of legal companies by regions", fig.height=3}
-mig_by_region_plot
-```
+Детализация статистики до населённых пунктов, находящихся внутри субъектов федерации, позволяет понять, что общая тенденция, характерная для региона (приток или отток фирм), как правило, не транслируется на следующий уровень географических объектов: в регионе обычно есть населённые пункты как с положительным, так и с отрицательным сальдо релокации.
+Это свидетельствует о том, что межрегиональные потоки фирм накладываются на внутрирегиональные, из-за чего некоторые населённы региона, в целом теряющего фирмы, могут, тем не менее, привлекать фирмы из других населённых пунктов этого же или других регионов, и наоборот.
 
-## Распределение населённых пунктов по притоку (оттоку) компаний
-
-Картограмма, детализированная до населённых пунктов, показана на 
-рис.
-5.
-Здесь видно, что населённые пункты с положительным сальдо миграции компаний сосредоточены в основном вокруг Москвы и Санкт-Петербурга.
-Кроме того, можно заметить, что в значительной доле субъектов федерации есть как населённые пункты, принимающие фирмы, так и населённые пункты, теряющие их.
-
-```{r fig-5-migration-by-settlement, fig.cap="Рис.
-5.
-Сальдо миграции юридических фирм (положительное или отрицательное) по населённым пунктам <br> Fig.
-5.
-Net migration rate of legal companies (positive or negative) by settlements"}
-migration_by_settlement_plot
+```{r fig-4-mig-by-region-plot, fig.cap="Рис.
+4.
+Региональная (по субъектам России) статистика миграции юридических фирм <br> Fig.
+4.
+Statistics of legal companies migration by region (subject of Russia)"}
+migration_by_region_plot
 ```
@@
 
+Результаты данного исследования могут иметь двоякое практическое применение.
+С одной стороны, их можно использовать в качестве одного из обоснований при определении государственной политики на базе доказательного (data-driven) подхода.
+В частности, из них следует, что масштаб релокации юридических фирм в пределах России, как правило, невелик, поэтому стимулирование развития малого и среднего предпринимательства в этой сфере, вероятно, должно опираться не на привлечение компаний из других регионов (исключая, возможно, самые крупные из них, которые могут открыть филиалы), а на стимулирование создания новых организаций.
+С другой стороны, исследование в сочетании с его полностью воспроизводимой методикой и открытыми источниками данных может использоваться как образец для анализа релокации фирм, ведущих другие виды деятельности.
+
 # Список литературы

### 2024-04-04 legal-companies-migration/Script.R

@@ -12,6 +12,7 @@ empl <- read_csv("../../ru-smb-companies/legal/empl.csv")
 cities <- read_csv("cities.csv")
 regions_geo <- st_read("ru.geojson")
 regions <- read_csv("regions.csv")
+tiles <- read_csv("russia-tiles.csv")
 
 # Number of unique companies
 companies_count <- n_distinct(data[data$kind == 1, "tin"])
@@ -172,9 +173,15 @@ mig_by_month_plot <- migrations %>%
   theme_bw(base_size = 11, base_family = "Times New Roman") +
   theme(panel.grid.minor.y = element_blank(), panel.grid.major.x = element_blank())
 
-# Migration by regions
+# Migration by regions and settlements
 migration_by_region <- migrations %>% 
-  select(region_from, region_to, revenue, empl) %>% 
+  mutate(
+    settlement_from = replace(settlement_from, region_from == "Москва", "Москва"),
+    settlement_from = replace(settlement_from, region_from == "Санкт-Петербург", "Санкт-Петербург"),
+    settlement_to = replace(settlement_to, region_to == "Москва", "Москва"),
+    settlement_to = replace(settlement_to, region_to == "Санкт-Петербург", "Санкт-Петербург"),
+  ) %>% 
+  select(region_from, region_to, settlement_from, settlement_to, revenue, empl) %>% 
   pivot_longer(region_from:region_to, names_to = "type", values_to = "region") %>% 
   drop_na(region) %>% 
   group_by(region, type) %>% 
@@ -189,55 +196,144 @@ migration_by_region <- migrations %>%
   right_join(select(regions_geo, shapeISO), by = c("iso_code" = "shapeISO")) %>% 
   st_as_sf()
 
-ru_crs <- st_crs("+proj=aea +lat_0=0 +lon_0=100 +lat_1=68 +lat_2=44 +x_0=0 +y_0=0 +ellps=krass +towgs84=28,-130,-95,0,0,0,0 +units=m +no_defs")
-mig_by_region_plot <- migration_by_region %>% 
-  ggplot() +
-  geom_sf(aes(fill = cut(count, c(-50, -25, 0, 25, 50, 75, 100))), size = .1) + 
-  coord_sf(crs = ru_crs) +
-  scale_fill_brewer(name = "", palette = "Greys") +
-  theme_bw(base_size = 11, base_family = "Times New Roman")
+legend_texts = data.frame(
+  label = c(
[…]

### 2024-04-10 bel-geo-congress-2024/Slides.qmd
Slides for Belarus geo congress

---
title: "Миграция юридических фирм в России (2016–23)"
author: "Павел Сёмин, Пермский государственный национальный исследовательский университет"
format: 
  revealjs:
    embed-resources: true
---

```{r}
#| include: false
#| cache: true
source("Script.R")
```

# Задачи

1. Оценить количество релокаций.
2. Понять, как решение о релокации связано с размером фирмы.
3. Определить типичные расстояния релокаций.
4. Визуализировать распределение сальдо релокации по регионам.
5. Определить ключевые межрегиональные потоки фирм.

# Данные и методы

1. Табличный датасет на базе реестра МСП.
2. Период — 2016–23.
3. Юридические фирмы — по ОКВЭД 69.10.
4. Релокация = изменение адреса регистрации.

# Результаты

## Релокации редки

```{r}
mig_count_distr_plot
```

## Крупные фирмы перемещаются чуть чаще

```{r}
mig_by_category_plot
```

## Расстояние — или близко, или далеко

```{r}
mig_distance_plot
```

## Сальдо релокации небольшое

```{r}
mig_by_region_plot
```

## Маршруты разнообразные {.scrollable}

```{r}
knitr::kable(
  head(regional_paths_combined_en, 30),
  col.names = c("Path", "Count", "Revenue, M rub", "Employees"),
  caption = "Top 30 paths of inter-regional migration of Russian legal companies in 2016–23",
  digits = 0
)
```

# Выводы

1. Релокация юрфирм — редкое явление.
2. Более крупные фирмы меняют адрес чуть чаще.
3. Расстояние миграции — два варианта: или несколько десятков километров, или тысяча.
4. Московская область «притянула» много фирм, Москва — потеряла, остальные регионы — по-разному и понемногу.

### 2024-04-14 legal-companies-geography/Paper.Rmd
Article template and introduction, #1

---
output: 
  word_document:
    reference_doc: reference.docx
bibliography: refs.bib
#csl: style.csl
link-bibliography: False
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Pavel O. Syomin[^1]**

<br>

<div custom-style="Title">

Legal Business Geography in Russia: Regional Analysis Using Tax Service Open Data

</div>

<br>

# Abstract

Text text text.

<br>

<div custom-style="Block Text">

**KEYWORDS:** w1, w2

</div>

<br>

[^1]: Perm State University, Department of Social and Economic Geography; 15 Bukireva st., Perm, 614068, Russia; e-mail: ntsp\@ya.ru

# Introduction

Geography of services is an important branch of modern economic and social geography. In Russia, it dates back to the 1960s when the pioneering research papers in this field of studies were published [@Tkachenko2016]. In foreign countries, the spatial aspects of services were treated as a part of human geography and touched in the scholarly works about retail [@Reilly1931; @Berry1967] and cities [@Brush1953] usually based on the central places theory [@Christaller1966]. A series of review research articles published in 1985–87 directly uses the term “geography of services” and cites dozens of original papers related to this topic [@Daniels1985; @Daniels1986; @Daniels1987]. Newer foreign studies in this field focus on the processes and phenomena in peripheral or smaller economies [@Hermelin2007] or on the arising and growing sectors like FinTech [@Lai2020; @Wojcik2020]. In Russia, recent research in the geography of services covers spatial properties of banking [@Shustova2015], education [@Shulgina2019], car rental [@Zyrianova2018], food service [@Preobrazhenskiy2018], service sector in general [@Lisin2020]. However, several types of services are rarely studied, e. g. business and legal ones [@Tkachenko2016].

The research on the geography of legal services in Russia is limited. To date, only an analytical review describing the legal market in Russia was published that contains a chapter on its spatial aspects [@Moiseeva2016], and yet another paper briefly addresses the role of Russia in the global legal services market without diving into the internal spatial properties of the Russian legal services itself [@Yastrebov2021]. There are also non-scientific products about Russian legal market that include its regional analysis[^2]. Despite the observed low interest to this issue from other researches, it can be argued that the geographical studies of legal services is necessary. First, legal services, as a subset of other business or producer services, play a substantial role in the development of regions and territories and in the economic growth [@Hansen1993], and sometimes act as a limiting factor, that is, the lack of professional legal advice may cause problems to entrepreneurs, especially to small ones, thus decreasing their performance [@Hitchens1997]. Next, the in-depth analysis of legal companies may give valuable insights about the business sectors they serve [@Faulconbridge2019] or increase the theoretical understanding of markets and industries in general [@Henderson2016]. Last but not least, legal services, apart from their narrow economic function, should be also treated as one of the elements of legal system in a wider sense that protect human rights and freedoms, ensure the rule of law, and act as one of the intermediaries between people, business and government that help to resolve the conflicts, increase the certainty and orderliness of social relations. This aspect of legal services is closely related to the social dimension of geography of services, and from this point of view it is not surprising that the research is also being conducted on the problem of legal services availablity [@Baxter2015].

[^2]: Анализ рынка юридических услуг в России в 2019-2023 гг, прогноз на 2024-2028 гг. Web resource: https://marketing.rbc.ru/research/41683/ (accessed 14.04.2024)

Legal services is quite a narrow and specific area of economy that does not reported separately in the official statistics, so for their analysis, other sources of information are necessary. Company-level georeferenced data is valuable for this task, because it can be easily filtered to leave the firms doing the business in the required activity area only and then aggregated by its spatial attributes to make a geographical overview. Researchers typically extract such the information from public GIS services [@Lachininsky2021] or business intelligence (BI) systems [@Moiseeva2016]. The disadvantage of this method, however, is the complexity of data preprocessing or the relatively high price of the data. Also, GIS and BI systems often do not enable access to historical data, providing real-time picture only, while the geographical studies may benefit from combining spatial and temporal properties. To overcome all these issues, an open administrative data may be used. In particular, the publicly available archives of Registry of Small and Medium-Sized Businesses (SMB registry) distributed by the Federal Tax Service (FTS) of Russia, in combination with a few other open data collections of the same governmental body, look valuable for the task in hand, and they have already been used by geographers [@Rostislav2021; @Gumenyuk2022].

The purpose of this paper is twofold. On the one hand, it aims to provide an overview of geography of legal business in Russia. The scope of the research is intentionally limited due to various organizational and technical constraints: it deals with the service providers only (the supply part of the market) and contains nothing about customers (the demand part), focuses on commercial legal companies with other market participants such as notaries, advocacy firms, non-profit legal organizations left untouched, and operates with only a portion of variables and metrics indicating the characteristics of legal business. On the other hand, the paper proposes a reproducible technique of working with FTS open data for analyzing the geography of services in Russia, particularly the services that do not reported independently from the others in the official statistics (mainly specific types of business and customer services) and thus can hardly be analyzed with the traditional economic geographers' toolkit.

<br>

# Materials and methods

# Results and discussion

# Conclusion

<br>

## References

<div id="refs"></div>

### 2024-04-15 legal-companies-concentration-and-availability/Slides.qmd
Slides for Cities Sustainability conference

---
title: "Концентрация юрфирм и доступность юруслуг: как они связаны?"
author: "Павел Сёмин, Пермский государственный национальный исследовательский университет"
format:
  revealjs:
    embed-resources: true
---

```{r}
#| include: false
#| cache: true
source("Script.R")
```

# Задача

На данных по российским городам понять, как связаны концентрация юридических фирм и доступность юридических услуг (и связаны ли)

# Данные

1. Реестр МСП + сведения о числе работников (ФНС России)
2. 2021 год
3. ОКВЭД 69.10
4. Только юридические лица (не ИП)
5. Только в городах (не в сёлах)
6. Доходы/расходы > 0

# Методы

1. Концентрация *(c)* — число работников юрфирм в городе / общее число работников юрфирм
2. Доступность *(a)* — число работников юрфирм в городе / число жителей города
3. Линейная регрессия *log10(a) ~ log10(c)*, визуализация

# Результаты

## Почти линейная зависимость и чёткие группы по размеру городов

```{r}
ca_plot
```

## Хорошая доступность в региональных центрах

```{r}
spatial_plot
```

# Выводы

1. Концентрация способствует доступности
2. Есть альтернативный путь
3. Территориально доступность выше в региональных центрах

### 2024-04-15 legal-companies-geography/Paper.Rmd
Data and partially methods, #1

@@
 
+**Data.** The research utilizes two primary data sources. All of them are published on the FTS website as open data.
+
+The first dataset is called *“Registry of small and medium-sized enterprises”*[^3].
+It contains full dumps of the respective registry in zipped XML files.
+The first dump was published in August 2016, and since then, new dumps were published every month.
+Each dump contains the whole registry and includes, in particular, company's names, their taxpayer IDs, activity codes according to All-Russia Classifier of Kinds of Economic Activity, address of incorporation (up to a settlement name).
+According to Russian laws, the Registry includes information about all commercial organizations and individual entrepreneurs that meet several criteria, the most important of which are limit on revenue (no more than 2 billion Russian rubles) and average number of employees (no more than 250).
+Thus, the open dumps of this registry can be treated as the most complete publicly available data source on the Russian business entities.
+In addition, this source has geographical reference in the form of human-readable addresses, and historical records from 2016 to now in the form of time series of monthly dumps of the registry.
+
+[^3]: Единый реестр субъектов малого и среднего предпринимательства. Web resource: https://www.nalog.gov.ru/opendata/7707329152-rsmp/ (accessed 15.04.2024)
+
+The second dataset is called *“Information about the average list count of employees of organisations”*[^4].
+As the name suggests, it contains information about the number of company's employees.
+The data was initially published in 2019 and since then was updated with varied frequency, but at least yearly.
+As a result, the data archive includes information for the years from 2019 to 2023.
+For the unspecified reasons, the old versions of the dataset containing information for years from 2019 to 2021 were once deleted from the FTS website, but they had been already downloaded by the author and thus used in this research.
+
+[^4]: Сведения о среднесписочной численности работников организации. Web resource: https://www.nalog.gov.ru/opendata/7707329152-sshr2019/ (accessed 15.04.2024)
+
+In addition to the primary data sources, two lookup tables with the information on Russian cities and settlements are used to normalise and geocode raw addresses contained in the FTS data.
+The first table is called *“Settlement of Russia: population and geographical coordinates”*[^5].
+It was prepared by the non-profit open data initiative called “Research data infrastructure” (RDI).
+The second table called *“Cities of Russia”*[^6] was published by commercial company DaData.
+Both these tables contain information that is actual for the beginning of 2022, thus there may be some inaccuracies in geocoding and normalisation, but they are likely to be negligible.
+
+[^5]: Населенные пункты России: численность населения и географические координаты. Web resource: https://data.rcsi.science/data-catalog/datasets/160/ (accessed 15.04.2024)
+[^6]: Города России. Web resource: https://github.com/hflabs/city (accessed 15.04.2024)
+
+**Methods.** A special Python command-line tool called *ru-smb-companies* was developed by the author of this paper to initially process huge and complex raw datasets from the FTS servers and transform them into a single table.
+The source code of this tool with a brief documentation is published in author's personal GitHub account[^7].
+Here, a concise summary of the data processing workflow is provided.
+The data processing takes several steps.
+
+[^7]: Russian small and medium business dataset generator. Web resource: https://github.com/PavelSyomin/russian-smb-companies (accessed 15.04.2024)
+
+*First,* source datasets were downloaded from the FTS servers and stored on the local machine.
+In addition, their backup was uploaded to author's personal Yandex.Disk account, making it possible for other researchers to access the raw data even if FTS decides to remove it (as indeed happened with data on the number of employees).
+
+*Second,* the filter-on-extract approach was used to extract necessary information from the raw archives into intermediary CSV tables.
+The filter-on-extract approach means that instead of unpacking the entire archive on the disk, the program reads its contents file by file without unpacking, parses the internal XML structure, extracts necessary attributes only, filters the records by activity code to leave legal companies only, and stores the result on the disk.
+Thus, the runtime and disk usage are substantially reduced compared to the traditional extract-then-filter method.
+
+*Third,* the data from the intermediary CSV tables was aggregated to remove duplicated records.
+The reason for the duplicates is the procedure of data publication: as it was mentioned earlier, every version of the source dataset contains the full data for the time of publication rather than updates only, leading to the huge number of repeating records.
+
+*Forth,* the addresses were normalised and geocoded to convert original administrative address format containing region, district, city and settlement name into simpler yet useful triplet of region, area and settlement, and add geographical coordinates and municipal codes (ОКТМО in Russian) for the settlements.
+
+*Fifth,* the data on business entities was combined with the data on employees using taxpayer ID as unique identifier common to both datasets, and also converted into a panel time series table that includes yearly information for each company or individual.
+
+<br>
+
 # Results and discussion

### 2024-04-16 legal-companies-regional-specialization/Slides.qmd
Slides for conference in Stavropol

---
title: "Рынок юридических услуг в России: специализация регионов"
author: "Павел Сёмин, Пермский государственный национальный исследовательский университет"
format:
  revealjs:
    embed-resources: true
---

```{r}
# | include: false
# | cache: true
source("Script.R")
```

# Задача

Определить регионы (субъекты федерации), в которых юридические услуги являются:

- отраслью локальной значимости, 
- отраслью национальной значимости,
- отраслью локальной и национальной значимости

# Данные

1. Реестр МСП + сведения о числе работников (ФНС России)
2. Число работников по субъектам (Росстат)
3. Год — 2021, юрфирмы — по ОКВЭД 69.10.

# Методы

1. «Алтас специализации регионов» (ВШЭ, 2021)
2. Концентрация — доля региона в общей занятости по юруслугам
3. Локализация — концентрация, делённая на долю региона в общей занятости по всем отраслям
4. Топ-16 (20%) по локализации → локальная значимость
5. Накопленная концентрация < 80% → национальная значимость

# Результат

```{r}
map
```

# Выводы

1. Шесть основных кластеров со специализацией на юруслугах
2. Преимущественно это регионы с относительно высоким уровнем экономического развития
3. Во многих из этих регионов есть апелляционные или кассационные суды — возможно, они «притягивают» юридический бизнес

### 2024-04-16 legal-companies-geography/Paper.Rmd
Write methods, #1

@@
 *Second,* the filter-on-extract approach was used to extract necessary information from the raw archives into intermediary CSV tables.
-The filter-on-extract approach means that instead of unpacking the entire archive on the disk, the program reads its contents file by file without unpacking, parses the internal XML structure, extracts necessary attributes only, filters the records by activity code to leave legal companies only, and stores the result on the disk.
+The filter-on-extract approach means that instead of unpacking the entire archive on the disk, the program reads its contents file by file without unpacking, parses the internal XML structure, extracts necessary attributes only, filters the records by activity code equal to 69.10 to leave legal companies only, and stores the result on the disk.
 Thus, the runtime and disk usage are substantially reduced compared to the traditional extract-then-filter method.
@@
 *Fifth,* the data on business entities was combined with the data on employees using taxpayer ID as unique identifier common to both datasets, and also converted into a panel time series table that includes yearly information for each company or individual.
+This resulting table is used in the further analysis.
+
+The research follows the general framework of regional analysis and uses the methods of descriptive statistics and data visualization, including the plotting of data on a tile-grid map of Russia.
+The analysis consists of a few research problems, or questions, that are listed below with the description of the relevant techniques used to find answers to them.
+
+The first problem is the *distribution of legal firms by regions* of Russia.
+To answer this question, legal firms that existed in 2021 are counted by the region, and the result is visualized on the map.
+
+The second problem is the *structure of legal business in regions* with respect to the kind of firms (either individual entrepreneurs or organisations) and the size of organisations (either micro, small, and medium).
+To answer this question, legal firms are counted by region, kind and type, and the result is visualized on a geofacetted plot.
+
+The third problem is the *role of legal business on the regional and national scale*, that is, either its importance within a given region or the importance of region's legal services in the Russia's legal market.
+To discover this issue, two calculations are made.
+The first one is used to better understand the role of the region in the country's legal business and is done by calculating the share of legal companies employees of a region in the total number of country's legal companies employees.
+The second one is aimed at estimating the relative role of legal business in region's economy and is done by dividing the number of legal companies employees in a region by the total number of employees in a region.
+Both two metrics are then plotted on a tile-grid map.
+
+The fourth issue deals with the temporal dimension and reflects the *change of the number of legal firms* in the regions. Legal firms are calculated by region and year, and the results are plotted.
+
+Several aspects are common to all the questions.
+First of all, despite the presence of data from 2016 to 2024, only a part from 2016 to 2021 inclusively is utilized for the analysis.
+The reason for this is the Russia's military operation in Ukraine and the subsequent structural transformation of the economy, international sanctions, substantial political and legal changes in the country that may have significantly impacted the legal market compared to the previous period, and even though all these changes could have been interesting to study, the purpose of the current research is to draw an overall picture of legal market in Russia in “normal situation” without any serious perturbations.
+Statistics for several consequent years, if not processed separately (e.
+g.
+in the case of temporal changes), is averaged using median as a function (e.g.
+number of legal firms for 2016–2021 in the region is the median of counts for each year from 2016 to 2021).
+Next, both organisations and individual entrepreneurs are accounted for.
+If statistical computations require the count of employees, it is assumed that an individual entrepreneur always has exactly one employee, that is, this person itself.
+This is because even though individual entrepreneurs in Russia can legally employ people, the corresponding information is not available in the FTS open data possibly due to personal data protection issues.
+Also, if an organisation has reported zero count of employees, it is still assumed that is has exactly one employee, that is, the company's owner itself.
 

### 2024-04-17 legal-companies-geography/Paper.Rmd
First part of results and discussion, #1

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(
+  echo = FALSE,
+  message = FALSE,
+  warning = FALSE,
+  fig.id = FALSE,
+  fig.width = 6.7,
+  fig.height = 5,
+  dpi = 300)
+```
+
+```{r imports, include=FALSE, cache=TRUE}
+source("Script.R")
 ```
@@
 
-The second problem is the *structure of legal business in regions* with respect to the kind of firms (either individual entrepreneurs or organisations) and the size of organisations (either micro, small, and medium).
-To answer this question, legal firms are counted by region, kind and type, and the result is visualized on a geofacetted plot.
+The second problem is the *structure of legal business in regions* with respect to the kind of firms (either individual entrepreneurs or organisations).
+To answer this question, legal firms are counted by region and king, and the result is visualized on a geofacetted plot.
 
@@
 
+**Average yearly count of legal firms by region** is displayed on a tile-grid colorpleth map in Fig.
+1.
+Several conclusions can be drawn from this visualisation.
+First, Moscow is the only city where the number of legal firms is above 10 thousand, and this is the highest value in Russia.
+Next, a set of regions also has relatively high count of legal companies and individual entrepreneurs.
+In spatial terms, most of these regions form several groups.
+One of these groups is located at the south of the European part of Russia and consists of Krasnodar and Stavropol territories with Rostov, Volgograd and Voronezh regions.
+Tatarstan and Bashkir republics with Samara, Sverdlovsk, Chelyabinsk regions and Perm territory make up another group that may be called Ural-Volga.
+Some Siberian regions, namely Tyumen, Omsk, Novosibirsk, Kemerovo, with Krasnoyarsk territory, form the third group that may be named Siberian by its location.
+Also, Saint Petersburg city, Nizhniy Novgorod region, and Primorie territory act as single-region groups.
+The majority of other regions has relatively low count of legal firms, and in some of them (several republics of North Caucas and Altai and a portion of regions from the North) there are less than a hundred of such companies.
+The co-loocation of regions and groups with different count of legal firms suggests that there may be a sort of central-peripheries relations between them.
+For example, Moscow city and region, Saint Petersburg city, and Nizhniy Novgorod region are surrounded by dozens of regions with relatively small number of legal firms, and thus may act as a centers of legal business for them.
+The same may be true for the pairs of South and Caucasian, as well as for Siberia and Far East.
+It may be hypothesized that economic development of regions is positively associated with the development of legal business.
+Also, the population of a region may also matter.
+
+<br>
+
+```{r fig-1-counts-map, fig.cap="Fig.
+1.
+Median yearly count of legal firms (both organisations and individual entrepreneurs) and shares of ogranisations by regions of Russia for 2016–21.
+Grey is no data"}
+counts_map
+```
+
+<br>
+
+Fig.
+1 also displays the median shares of organisations among all legal firms in regions.
+Legal firms may be either organisations or individual entrepreneurs, so for each region, the rest part of the firms consists of individuals.
+The share of companies among all legal firms varies from less about 25% to almost 70%.
+However, a substantial part of regions has from 40% to 60% of organisations, and thus many regions may be categorized either as with a small domination of companies or as with small domination of individuals on the legal market.
+Several southern regions (Krasnodar and Stavropol territories, Rostov region, Kalmyk republic), as well as several northern ones (Arkhangelsk region, Komi republic, Nenetz and Yamal territories), and some central regions (Moscow and Vladimir) have substantially lower shares of companies compared to the surrounding regions, thus forming the areas of predominantly individual legal business.
+The overall count of legal firms seems to have to connection with the share of organisation among them.
+
 # Conclusion

### 2024-04-17 legal-companies-geography/Script.R

@@ -1,0 +1,58 @@
+library(dplyr)
+library(ggplot2)
+library(readr)
+library(tidyr)
+
+# Load raw data
+data <- read_csv("../../ru-smb-companies/legal/panel.csv")
+tiles <- read_csv("russia-tiles.csv")
+
+# Common preparations
+data <- data %>% 
+  filter(year < 2022) %>% 
+  replace_na(list(employees_count = 0)) %>% 
+  mutate(
+    employees_count = replace(
+      employees_count, employees_count == 0, 1),
+    kind = replace(kind, kind == 3, 2)
+    ) %>% 
+  select(tin, kind, category, region, year, empl = employees_count)
+
+# Count of legal firms by region and shares of companies
+shares <- count(data, region, year, kind) %>% 
+  group_by(region, year) %>%
+  mutate(share = n / sum(n)) %>% 
+  filter(kind == 1) %>% 
+  group_by(region) %>% 
+  summarise(share = median(share))
+
+counts_map <- count(data, region, year) %>% 
+  group_by(region) %>% 
+  summarise(n = median(n)) %>% 
+  left_join(shares) %>% 
+  right_join(tiles, by = c("region" = "name")) %>% 
+  ggplot(aes(x = col, y = -row, fill = n)) +
+  geom_raster() +
+  geom_point(
+    aes(x = col + .25, y = -row - .25, size = share), 
+    shape = "circle open",
+    na.rm = TRUE) +
+  geom_text(
+    aes(x = col - .4, y = -row + .3, label = code_en), 
+    family = "Times New Roman", 
+    size = 3,
+    hjust = 0) +
+  scale_size_binned(name = "Share of organisations", n.breaks = 6) +
+  scale_fill_binned(
+    name = "Count of legal firms", transform = "log10",
+    high = "#eff3ff", low = "#2171b5") +
+  coord_fixed(expand = FALSE) +
+  theme_void(base_family = "Times New Roman", base_size = 11) +
+  theme(
+    legend.position = c(1, .05),
+    legend.justification = c(1, 0),
+    legend.direction = "horizontal",
+    legend.title.position = "top"
+  )
+counts_map
+

### 2024-04-19 legal-companies-geography/Paper.Rmd
Separate counts and shares plots, add history plot, write about employees, make minor fixes and improvements, #1

@@
 
-```{r fig-1-counts-map, fig.cap="Fig.
-1.
-Median yearly count of legal firms (both organisations and individual entrepreneurs) and shares of ogranisations by regions of Russia for 2016–21.
-Grey is no data"}
+```{r fig-1-counts-map, fig.cap="Fig. 1. Median yearly count of legal firms (both organisations and individual entrepreneurs) by regions of Russia for 2016–21. Grey is no data"}
 counts_map
@@
 Fig.
-1 also displays the median shares of organisations among all legal firms in regions.
+2 displays the *median shares of organisations among all legal firms in regions*.
 Legal firms may be either organisations or individual entrepreneurs, so for each region, the rest part of the firms consists of individuals.
@@
 However, a substantial part of regions has from 40% to 60% of organisations, and thus many regions may be categorized either as with a small domination of companies or as with small domination of individuals on the legal market.
-Several southern regions (Krasnodar and Stavropol territories, Rostov region, Kalmyk republic), as well as several northern ones (Arkhangelsk region, Komi republic, Nenetz and Yamal territories), and some central regions (Moscow and Vladimir) have substantially lower shares of companies compared to the surrounding regions, thus forming the areas of predominantly individual legal business.
+Several southern regions (Krasnodar and Stavropol territories, Rostov region, Kalmyk republic), as well as several northern ones (Arkhangelsk region, Komi republic, Nenetz and Yamal territories), and some central regions (Kaluga and Moscow) have substantially lower shares of companies compared to the surrounding regions, thus forming the areas of predominantly individual legal business.
 The overall count of legal firms seems to have to connection with the share of organisation among them.
+
+<br>
+
+```{r fig-2-shares-map, fig.cap="Fig. 2. Median yearly shares of companies among legal firms by regions of Russia for 2016–21. Grey is no data"}
+shares_map
+```
+
+<br>
+
+Fig.
+3 shows the joint results of regional analysis of number of legal companies employees.
+Two main metrics are drawn, the first is the concentration of legal workers in the region, that is, the share of legal companies employees of a region in all country's employees of legal companies, and the second is the share of legal workers in region's labor force.
+The first measure roughly describes the importance of a region in country's legal business, while the second estimates the legal business' impact in regional economy.
+In addition, the number of multi-region courts located in a given region is shown.
+The overall regional distribution of legal companies employees resembles the regional distribution of legal companies (see Fig.
+1) suggesting that legal companies among regions are quite similar to each other, so that there is no significant difference between simple count of companies and weighted count of companies with respect to the number of workers.
+The role of legal business in region's economy is distributed differently, though.
+It's clear that Moscow and Saint Petersburg, the two largest cities concentrating almost 1/3 of legal companies employees, are also characterized by the relatively highest portion of legal workers in total labor force.
+The majority of other regions is surprisingly uniform in having the small share of legal workers (0.1–0.2%) in labor force, with Sverdlovsk, Tyumen and Novosibirsk regions being rare exceptions.
+It is also worth noticing that regions with extremely low share in country's legal business (for example, Tyva republic and Altai territory) also usually have low share of legal workers in regional labor force, with a single exception of Evrei autonomous region.
+In addition, the regional role of legal business seems to have to connection with the presence or absence of multi-regional courts.
+However, looks like that the national role of a region in legal business is somehow connected with the location of such the courts, because almost all regions with relatively high share in country's legal companies employees (>1%) have at least one multi-regional court, and the reverse is true either: other regions typically do not have such the courts.
+Of course, there are regions that do not fall within this tendencies (Bashkortostan republic and Volgograd region, as well as three sub-Moscow regions, Vologda region and Zabaikal territory), but these deviations may be explained by various factors (e.
+g.
+for Tula, Vladimir and Kaluga regions the presence of multi-regional courts is not associated with the development of legal business probably due to the proximity to Moscow, so that the need in local legal firms does not arises).
+
+<br>
+
+```{r fig-3-employees-map, fig.cap="Fig 3. National and regional median yearly shares of legal companies employees in regions of Russia in 2018–2021"}
+employees_map
+```
 

### 2024-04-19 legal-companies-geography/Script.R

@@ -21,14 +21,7 @@ empl <- pivot_longer(
   empl, "2016":"2021", names_to = "year", values_to = "empl_total") %>% 
   mutate(year = as.numeric(year))
 
-# Count of legal firms by region and shares of companies
-shares <- count(data, region, year, kind) %>% 
-  group_by(region, year) %>%
-  mutate(share = n / sum(n)) %>% 
-  filter(kind == 1) %>% 
-  group_by(region) %>% 
-  summarise(share = median(share))
-
+# Count of legal firms by region
 counts_map <- count(data, region, year) %>% 
   group_by(region) %>% 
   summarise(n = median(n)) %>% 
@@ -41,11 +34,6 @@ counts_map <- count(data, region, year) %>%
     family = "Times New Roman", 
     size = 3,
     hjust = 0) +
-  geom_point(
-    aes(x = col + .25, y = -row - .25, size = share), 
-    shape = "circle open",
-    na.rm = TRUE) +
-  scale_size_binned(name = "Share of organisations", n.breaks = 6) +
   scale_fill_binned(
     name = "Count of legal firms", transform = "log10",
     high = "#eff3ff", low = "#2171b5") +
@@ -55,12 +43,45 @@ counts_map <- count(data, region, year) %>%
     legend.position = c(1, .05),
     legend.justification = c(1, 0),
     legend.direction = "horizontal",
-    legend.title.position = "top"
-  )
+    legend.title.position = "top",
+    plot.caption = element_text(size = 11)
+  ) +
+  labs(caption = "Grey is no data")
 counts_map
 
+# Shares of companies among legal firms
+shares_map <- count(data, region, year, kind) %>% 
+  group_by(region, year) %>%
+  mutate(share = n / sum(n)) %>% 
+  filter(kind == 1) %>% 
+  group_by(region) %>% 
+  summarise(share = median(share)) %>% 
+  right_join(tiles, by = c("region" = "name")) %>% 
+  ggplot(aes(x = col, y = -row, fill = share)) +
+  geom_raster() +
+  geom_text(
+    aes(x = col - .4, y = -row + .3, label = code_en), 
+    family = "Times New Roman", 
+    size = 3,
+    hjust = 0) +
+  scale_fill_binned(
+    name = "Share of organizations", transform = "log10",
+    n.breaks = 6,
+    high = "#eff3ff", low = "#2171b5") +
+  coord_fixed(expand = FALSE) +
[…]

### 2024-04-20 legal-companies-geography/Paper.Rmd
Finish results, add conclusion, #1

@@
 
-Text text text.
+This paper provides an exploratory regional analysis of legal business in Russia for 2016–21.
+The analysis is based mainly on the Federal Tax Service (FTS) open data on small and medium-sized businesses and on the number of employees of organisations, as well as on portions of official statistical information.
+The paper proposes a reproducible technique of working with FTS open data for regional analysis in service geography that was partially implemented in a specially designed Python command-line tool.
+The analysis includes questions about the count of legal firms and their employees by region, about the share of organisations in regional legal business, about the role of legal business in region's economy, and about the yearly change of the number of legal companies in regions.
+In addition, it briefly addresses an issue of probable connection between the location of multi-regional courts and the development of legal business.
+It is found out that legal business in Russia is highly concentrated in Moscow and Saint Petersburg that jointly account for 1/3 of legal firms and workers and are also at the top by the role of legal business in the region.
+However, there are other regions with relatively high development of legal business, that are located mainly on the European South of Russia, in the Volga-Urals and in Siberia.
+There is a group of regions with domination of organisations in legal business and the group of regions where individual entrepreneurs prevail over companies.
+District courts are usually located in the regions with high level of legal business development, but the mechanism of this co-location requires further investigation.
+In general, the number of legal firms in regions grown from 2016 to 2021, with the common inverted U-shape pattern probably explained by the coronavirus (covid-19) pandemic that negetively impacted the small and medium business and reverted the initial growth trend.
 
@@
 
-**KEYWORDS:** w1, w2
+**KEYWORDS:** regional analysis, geography of services, legal services, legal business, open data
 
@@
 
+<br>
+
+The changes in the number of legal firms by region are depicted in Fig.
+4.
+It can be clearly seen that the count of such companies increased in the majority of regions, indicating the development of legal business.
+Few regions have lost companies, all of them are located in the European part of Russia, and their net change in the number of legal firms from 2016 to 2021 is relatively low (usually less than 10%) compared to the increase rates in other regions.
+Sometimes neither decline or growth was not clearly expressed (for example, Altai republic, Lipetsk and Tambov regions).
+The regions with the largest growth are some republics at Notrh Caucas, as well as Kalmyk republic and Nenetz territory.
+Their results are probably explained by the low base effect, because the absolute counts of legal firms in all there regions are below 100 (see Fig.
+1).
+As for other regions with high growth, Leningrad and Moscow regions should be remarked, because their results may be partially explained by the relocation of companies from the cities (Saint Petersburg and Moscow) to the surrounding regions.
+However, Moscow and Saint Petersburg also gained legal firms with one of the top net changes.
+The observed patterns of changes indicated by lines of the map can be divided into several groups.
+The first is the stable yet saturating increase typical for Moscow and Saint Petersburg regions, and, less expresses, Khabarovsk territory, Rostov and Ryazan regions.
+The second is the inverted U-shape curve which is charasteristic of the majority of regions with overall growth in the number of legal firms.
+This pattern can be explained by the coronavirus (COVID-19) pandemic effect that has negatively impacted small and medium-sized business and reverted the previous trend of the growth.
+The “height” of this inverted U varies by region (the highest observed in Chukotka territory, Tyva and Adygeya republics).
+It can be hypothesised that the first group is in fact a subset of the second, but the intrinsic growth rate before the pandemic was higher, resulting in the slowing down of the growth rather than in the final decrease.
+The third pattern is typical for the regions with overall decrease in the number of legal firms, and it can be described as “initial short-term growth, then continuous decline”.
+The fourth group is a mixture of increase and decrease, for example Nenetz territory and Northern Osetia republic.
+
+<br>
+
+```{r fig-4-timeline-map, fig.cap="Fig. 4. The change in the number of legal firms by region in 2016–2021"}
+timeline_map
+```
+
+<br>
+
 # Conclusion
+
+The conducted regional analysis discovered several spatial features of legal business in Russia.
+First, the regions are quite different with respect to the count of legal firms or their employees.
+Several regional clusters with relatively high concentration of legal business can be found out, for example, the European South, Volga-Ural and Siberian.
+Unsurprisingly, Moscow and Saint Petersburg cities are characterized by the top count of legal firms and workers, but the overall concentration is really huge, because they jointly account for approximately 1/3 of Russia's legal businesses and legal companies employees.
+Besides, it's again Moscow and Saint Petersburg that have the highest share of legal companies employees in region's labor force (above 0.4% that is 2 or 3 times higher than the average value for other regions).
+However, there are some other regions with relatively high role of legal business in regional economy: Sverdlosvk, Tyumen and Novosibirsk.
+
+Regions can be roughly grouped into two categories: one with the legal organisations prevailing in the legal business, and the other with individual entrepreneurs prevailing over companies.
+The high role of companies is typical for Moscow, Saint Petersburg, some Volga and Siberian regions, as well as a small portion of central regions (Vologda and Ivanovo), while individuals are important in the European South and North, in the majority of Central European regions, in several regions of the Urals, and in the Far East.
+
+The role of region in country's legal business is usually associated with the presence of one or many multi-regional courts in the administrative center of this region.
+Only two regions do not act as a host for any of such the courts but still have relatively high level of development of legal business: Volgograd region and Bashkir republic.
+This observation may be explained in two ways: either the courts act as a magnets for legal business, because courts are necessary for many legal activities, or the courts are placed in regions and cities with highly developer legal business.
+Further research may throw the light on the direction of this court-business relation.
+
+From 2016 to 2021, the number of legal firms in Russia's regions mostly increased.
+The decrease is characteristic only of few regions.
+The top growth was observed in Moscow, Saint Petersburg, as well as in the surrounding regions.
+In addition, some regions (mainly at the European South) also shown very high growth, but probably due to the low base effect.
+The yearly patterns of change may be categorized into four groups, and the inverted U-shape change explained likely by the coronavirus (covid-19) pandemic is the most common among the regions.
+
+The results of this explanatory study provide an overview of Russia's recent legal business geography and may be used as a support for data-driven business and governmental decisions.
+Several research hypothesis also may be conducted from this analysis, thus this study may be treated as a base for further academic development in the field of legal services geography in Russia.
+Apart from this, the reproducible data-driven technique of this research based entirely on open data and open source software tools may be used (with minor changes) for regional analysis and mapping of other narrow service sector areas.
 

### 2024-04-20 legal-companies-geography/Script.R

@@ -25,7 +25,6 @@ empl <- pivot_longer(
 counts_map <- count(data, region, year) %>% 
   group_by(region) %>% 
   summarise(n = median(n)) %>% 
-  left_join(shares) %>% 
   right_join(tiles, by = c("region" = "name")) %>% 
   ggplot(aes(x = col, y = -row, fill = n)) +
   geom_raster() +
@@ -108,7 +107,8 @@ employees_map <- data %>%
     aes(x = col - .4, y = -row - .3, label = n_courts, group = n_courts), 
     family = "Times New Roman", 
     size = 3,
-    hjust = 0) +
+    hjust = 0,
+    na.rm = TRUE) +
   geom_point(
     aes(x = col + .25, y = -row - .25, size = share_regional), 
     shape = "circle open",
@@ -127,13 +127,14 @@ employees_map <- data %>%
     legend.justification = c(1, 0),
     legend.direction = "horizontal",
     legend.title.position = "top",
-    plot.caption = element_text(size = 11)
+    plot.caption = element_text(size = 11),
+    plot.caption.position = 
   ) +
   labs(caption = "Digits are the numbers of mutli-regional courts. Grey is no data")
 employees_map
 
 # Timeline of legal companies count
-count(data, region, year) %>% 
+timeline_map <- count(data, region, year) %>% 
   group_by(region) %>% 
   arrange(year) %>% 
   summarise(
@@ -143,23 +144,43 @@ count(data, region, year) %>%
       abs(first(rel_count) - last(rel_count)) < .05 ~ 0,
       first(n) < last(n) ~ 1,
       first(n) > last(n) ~ -1
-    ), levels = -1:1, ordered = TRUE)) %>% 
+    ), levels = -1:1, ordered = TRUE),
+    change = (last(rel_count) - first(rel_count)) * 100,
+    alpha = .5) %>% 
   right_join(tiles, by = c("region" = "name")) %>% 
   ggplot(aes(x = year, y = rel_count)) + 
+  geom_text(
+    aes(
+      x = 2021,
+      y = 1.25, 
+      label = if_else(!is.na(change), sprintf("%+.0f", change), ""),
+      alpha = alpha),
+    family = "Times New Roman",
+    size = 4,
+    hjust = 1,
+    color = "gray45",
+    key_glyph = "blank") +
   geom_line(aes(color = direction)) +
   geom_text(
-    aes(x = 2016, y = 3, label = code_en),
+    data = tiles,
[…]

### 2024-04-22 legal-companies-geography/Paper.Rmd
Perform proofreading of introduction, materials and methods, add CSL style, make minor fixes in reference docx and in biblio, #1

@@
 bibliography: refs.bib
-#csl: style.csl
+csl: style.csl
 link-bibliography: False
@@
 
-Legal Business Geography in Russia: Regional Analysis Using Tax Service Open Data
+Legal Business Geography in Russia: Regional Analysis and Mapping Using Tax Service Open Data
 
@@
 Geography of services is an important branch of modern economic and social geography.
-In Russia, it dates back to the 1960s when the pioneering research papers in this field of studies were published [@Tkachenko2016].
-In foreign countries, the spatial aspects of services were treated as a part of human geography and touched in the scholarly works about retail [@Reilly1931; @Berry1967] and cities [@Brush1953] usually based on the central places theory [@Christaller1966].
-A series of review research articles published in 1985–87 directly uses the term “geography of services” and cites dozens of original papers related to this topic [@Daniels1985; @Daniels1986; @Daniels1987].
-Newer foreign studies in this field focus on the processes and phenomena in peripheral or smaller economies [@Hermelin2007] or on the arising and growing sectors like FinTech [@Lai2020; @Wojcik2020].
-In Russia, recent research in the geography of services covers spatial properties of banking [@Shustova2015], education [@Shulgina2019], car rental [@Zyrianova2018], food service [@Preobrazhenskiy2018], service sector in general [@Lisin2020].
-However, several types of services are rarely studied, e.
-g.
-business and legal ones [@Tkachenko2016].
+In Russia, the discipline can be traced back to the 1960s, when pioneering research papers were published on this subject [@Tkachenko2016].
+In foreign countries, the spatial aspects of services were treated as a part of human geography and were first touched upon in the scholarly works about retail [@Reilly1931; @Berry1967] and cities [@Brush1953], which were usually based on the central places theory [@Christaller1966].
+A series of review research articles published in 1985–87 directly employs the term “geography of services” and cites numerous original papers related to this topic [@Daniels1985; @Daniels1986; @Daniels1987].
+More recent foreign studies in this field have focused on the processes and phenomena occurring in peripheral or smaller economies [@Hermelin2007] or on the rising sectors such as FinTech [@Lai2020; @Wojcik2020].
+Recent research in the geography of services in Russia addresses spatial properties of banking [@Shustova2015], education [@Shulgina2019], car rental [@Zyrianova2018], food service [@Preobrazhenskiy2018], and the service sector in general [@Lisin2020].
+Nevertheless, several types of services remain understudied by geographers, including business and legal services [@Tkachenko2016].
 
 The research on the geography of legal services in Russia is limited.
-To date, only an analytical review describing the legal market in Russia was published that contains a chapter on its spatial aspects [@Moiseeva2016], and yet another paper briefly addresses the role of Russia in the global legal services market without diving into the internal spatial properties of the Russian legal services itself [@Yastrebov2021].
-There are also non-scientific products about Russian legal market that include its regional analysis[^2].
-Despite the observed low interest to this issue from other researches, it can be argued that the geographical studies of legal services is necessary.
-First, legal services, as a subset of other business or producer services, play a substantial role in the development of regions and territories and in the economic growth [@Hansen1993], and sometimes act as a limiting factor, that is, the lack of professional legal advice may cause problems to entrepreneurs, especially to small ones, thus decreasing their performance [@Hitchens1997].
-Next, the in-depth analysis of legal companies may give valuable insights about the business sectors they serve [@Faulconbridge2019] or increase the theoretical understanding of markets and industries in general [@Henderson2016].
-Last but not least, legal services, apart from their narrow economic function, should be also treated as one of the elements of legal system in a wider sense that protect human rights and freedoms, ensure the rule of law, and act as one of the intermediaries between people, business and government that help to resolve the conflicts, increase the certainty and orderliness of social relations.
-This aspect of legal services is closely related to the social dimension of geography of services, and from this point of view it is not surprising that the research is also being conducted on the problem of legal services availablity [@Baxter2015].
+To date, only an analytical review describing the legal market in Russia has been published, which contains a chapter on its spatial aspects [@Moiseeva2016].
+Another paper briefly addresses the role of Russia in the global legal services market without delving into the internal spatial properties of the Russian legal business [@Yastrebov2021].
+In addition to the scientific publications, there are business products about the legal services market in Russia that include its regional analysis[^2].
+Despite a low observed interest in this issue from other researchers, it can be argued that geographical studies of legal services are essential.
+First, legal services, as a subset of other business or producer services, play a substantial role in the development of regions and territories and in the economic growth [@Hansen1993].
+In some cases, the lack of professional legal advice may even act as a limiting factor, causing problems for entrepreneurs, especially small ones, and decreasing their performance [@Hitchens1997].
+Next, a comprehensive examination of legal companies may yield valuable insights into the business sectors they serve [@Faulconbridge2019] or enhance the theoretical understanding of markets and industries in general [@Henderson2016].
+Finally, legal services should be considered not only in terms of their narrow economic function, but also as an integral element of the broader legal system.
+This system is responsible for protecting human rights and freedoms, ensuring the rule of law, and acting as an intermediary between people, businesses and government.
+It helps to resolve conflicts, increase certainty and orderliness in social relations, and thus contributes to the stability and development of society.
+This aspect of legal services is closely related to the social dimension of the geography of services.
+It is therefore unsurprising that that research is being conducted on the problem of legal services availability [@Baxter2015].
 
@@
 
-Legal services is quite a narrow and specific area of economy that does not reported separately in the official statistics, so for their analysis, other sources of information are necessary.
-Company-level georeferenced data is valuable for this task, because it can be easily filtered to leave the firms doing the business in the required activity area only and then aggregated by its spatial attributes to make a geographical overview.
-Researchers typically extract such the information from public GIS services [@Lachininsky2021] or business intelligence (BI) systems [@Moiseeva2016].
-The disadvantage of this method, however, is the complexity of data preprocessing or the relatively high price of the data.
-Also, GIS and BI systems often do not enable access to historical data, providing real-time picture only, while the geographical studies may benefit from combining spatial and temporal properties.
-To overcome all these issues, an open administrative data may be used.
-In particular, the publicly available archives of Registry of Small and Medium-Sized Businesses (SMB registry) distributed by the Federal Tax Service (FTS) of Russia, in combination with a few other open data collections of the same governmental body, look valuable for the task in hand, and they have already been used by geographers [@Rostislav2021; @Gumenyuk2022].
+The field of legal services represents a narrow and specific area of the economy that is not reported separately in the official statistics.
+Consequently, to analyze this sector, it is necessary to draw upon other sources of information.
+Georeferenced data on distinct firms is of particular value in this context, as it can be readily filtered to exclude firms that are not engaged in the requisite activity and then aggregated according to its spatial attributes to make a geographical summary.
+Researchers typically extract such the information from public geoinformation (GIS) services [@Lachininsky2021] or business intelligence (BI) systems [@Moiseeva2016].
+However, this method has two disadvantages.
+Firstly, the complexity of data preprocessing makes it difficult to use.
+Secondly, the relatively high price of the data makes it unaffordable for many researchers.
+Furthermore, GIS and BI systems often do not enable access to historical data, providing only a real-time picture, while geographical studies may be enhanced by combining spatial and temporal properties.
+To address these challenges, the use of open administrative data may be a viable solution.
+In particular, the publicly available archives of the Registry of Small and Medium-Sized Businesses (SMB registry) distributed by the Federal Tax Service (FTS) of Russia, in combination with a few other open data collections of the same governmental body, appear to be a valuable resource for task at hand.
+These data sources have already been used by geographers [@Rostislav2021; @Gumenyuk2022].
 
-The purpose of this paper is twofold.
-On the one hand, it aims to provide an overview of geography of legal business in Russia.
-The scope of the research is intentionally limited due to various organizational and technical constraints: it deals with the service providers only (the supply part of the market) and contains nothing about customers (the demand part), focuses on commercial legal companies with other market participants such as notaries, advocacy firms, non-profit legal organizations left untouched, and operates with only a portion of variables and metrics indicating the characteristics of legal business.
-On the other hand, the paper proposes a reproducible technique of working with FTS open data for analyzing the geography of services in Russia, particularly the services that do not reported independently from the others in the official statistics (mainly specific types of business and customer services) and thus can hardly be analyzed with the traditional economic geographers' toolkit.
+The objective of this paper is twofold.
+Primarily, it seeks to provide an overview of the geography of legal business in Russia.
+The scope of the research is intentionally limited due to various organizational and technical constraints.
+It focuses exclusively on service providers, or the supply side of the market, and does not address customers, the demand side.
+Furthermore, it concentrates on commercial legal businesses, excluding other market participants such as notaries, advocacy firms, and non-profit legal organizations.
+Finally, only a subset of metrics indicative of characteristics of legal businesses is included in the analysis.
+The specific research questions are presented in the “Materials and methods” section below.
+
+In addition to the primary purpose, the paper proposes a reproducible methodology of working with FTS open data for analyzing the geography of services in Russia, particularly those not reported separately in official statistics, such as specific types of business and customer services.
+Their analysis would be difficult using the traditional tools and data sources of economic geographers.
 
@@
 
-**Data.** The research utilizes two primary data sources. All of them are published on the FTS website as open data.
-
-The first dataset is called *“Registry of small and medium-sized enterprises”*[^3].
-It contains full dumps of the respective registry in zipped XML files.
-The first dump was published in August 2016, and since then, new dumps were published every month.
-Each dump contains the whole registry and includes, in particular, company's names, their taxpayer IDs, activity codes according to All-Russia Classifier of Kinds of Economic Activity, address of incorporation (up to a settlement name).
-According to Russian laws, the Registry includes information about all commercial organizations and individual entrepreneurs that meet several criteria, the most important of which are limit on revenue (no more than 2 billion Russian rubles) and average number of employees (no more than 250).
-Thus, the open dumps of this registry can be treated as the most complete publicly available data source on the Russian business entities.
-In addition, this source has geographical reference in the form of human-readable addresses, and historical records from 2016 to now in the form of time series of monthly dumps of the registry.
+**Data.** The research utilizes two primary data sources: the Unified Registry of Small and Medium-Sized Businesses (SMB Registry)[^3] and the Information about the Average Number of Employees of Organisations (Information about Employees)[^4].
+Both of these are available on the FTS website as open data.
 
 [^3]: Единый реестр субъектов малого и среднего предпринимательства. Web resource: https://www.nalog.gov.ru/opendata/7707329152-rsmp/ (accessed 15.04.2024)
-
-The second dataset is called *“Information about the average list count of employees of organisations”*[^4].
-As the name suggests, it contains information about the number of company's employees.
-The data was initially published in 2019 and since then was updated with varied frequency, but at least yearly.
-As a result, the data archive includes information for the years from 2019 to 2023.
-For the unspecified reasons, the old versions of the dataset containing information for years from 2019 to 2021 were once deleted from the FTS website, but they had been already downloaded by the author and thus used in this research.
-
 [^4]: Сведения о среднесписочной численности работников организации. Web resource: https://www.nalog.gov.ru/opendata/7707329152-sshr2019/ (accessed 15.04.2024)
 
-In addition to the primary data sources, two lookup tables with the information on Russian cities and settlements are used to normalise and geocode raw addresses contained in the FTS data.
-The first table is called *“Settlement of Russia: population and geographical coordinates”*[^5].
-It was prepared by the non-profit open data initiative called “Research data infrastructure” (RDI).
-The second table called *“Cities of Russia”*[^6] was published by commercial company DaData.
-Both these tables contain information that is actual for the beginning of 2022, thus there may be some inaccuracies in geocoding and normalisation, but they are likely to be negligible.
+The SMB Registry dataset contains full dumps of the respective registry in zipped XML files.
+The initial dataset was made available in August 2016, and updates have been released on a monthly basis.
+Each dump contains the entire registry and includes, in particular, the company or individual names, taxpayer IDs, and activity codes according to All-Russia Classifier of Kinds of Economic Activity, as well as addresses of incorporation (up to the settlement name).
+According to Russian laws, the Registry includes information about all commercial organizations and individual entrepreneurs that meet several criteria (SMB subjects).
+The most important criteria are limits on the revenue (no more than 2B Russian rubles) and on the average number of employees (no more than 250).
+Consequently, the open dumps of this registry represent the most comprehensive publicly accessible data source on the Russian business entities.
+Moreover, the archive of these dumps constitutes a time series spanning from August 2016 to the present, thereby providing a history of small and medium-sized businesses changes.
+
+The Information about Employees dataset contains company taxpayer IDs and average yearly numbers of employees for the previous year.
+The data was initially published in 2019 and has been updated with varying frequencies, though at least annually.
+Consequently, the data archive encompasses the period from 2018 to 2023.
+For reasons that are not specified, the older versions of the dataset which included information for the years from 2018 to 2022 are not currently available on the FTS website.
+However, the data had already been downloaded by the author prior to its removal, and thus it is used in this research.
+
+In addition to the primary data sources, two lookup tables with the information on Russian cities and settlements are used to normalize and geocode the raw addresses contained in the FTS data.
+The first table is “Settlements of Russia: Population and Geographical Coordinates”[^5] prepared by the nonprofit open data initiative known as “Research Data Infrastructure” (RDI).
+The second table is “Cities of Russia”[^6] published by the commercial company HFLabs.
+Data on several cities is absent from this table, so it was entered manually.
+Both tables contain information that is actual for the beginning of 2022, and thus there may be some inaccuracies in geocoding and normalization.
+However, these are likely to be negligible.
 
@@
 
-**Methods.** A special Python command-line tool called *ru-smb-companies* was developed by the author of this paper to initially process huge and complex raw datasets from the FTS servers and transform them into a single table.
-The source code of this tool with a brief documentation is published in author's personal GitHub account[^7].
+Apart from these datasets, a small manually collected table with the locations of Russian district-level courts is utilized.
+District-level courts span their activity over two or more regions of Russia that constitute a judicial district, checking and reviewing decisions of lower-level courts.
+These category include appeal and cassation ordinary and commercial courts, with a total of 47 such courts in Russia.
+The location data was obtained from the Supreme Court of Russia website[^7] and the legal information system called “Consultant Plus”[^8].
+While not the primary focus of this paper, the court location data was utilized to supplement the regional analysis by examining the relationship between the legal business geography and the placement of district-level courts.
+
+[^7]: Информация о кассационных и апелляционных судах общей юрисдикции. Web resource: https://vsrf.ru/press_center/news/28148/ (accessed 15.04.2024)
+[^8]: Справочная информация: "Арбитражные суды". Web resource: https://www.consultant.ru/document/cons_doc_LAW_56549/ (accessed 15.04.2024)
+
+**Methods.** A Python command-line, ru-smb-companies, was developed by the author of this paper to initially process large and complex raw XML datasets from the FTS servers and transform them into a single CSV table.
+The source code of this tool, along with a brief documentation, is published in the author's personal GitHub account[^9].
 Here, a concise summary of the data processing workflow is provided.
@@
 
-[^7]: Russian small and medium business dataset generator. Web resource: https://github.com/PavelSyomin/russian-smb-companies (accessed 15.04.2024)
+[^9]: Russian small and medium business dataset generator. Web resource: https://github.com/PavelSyomin/russian-smb-companies (accessed 15.04.2024)
 
-*First,* source datasets were downloaded from the FTS servers and stored on the local machine.
-In addition, their backup was uploaded to author's personal Yandex.Disk account, making it possible for other researchers to access the raw data even if FTS decides to remove it (as indeed happened with data on the number of employees).
+Step 1 is data download.
+The dumps of the SMB Registry and the archives of Information about Employees were downloaded from the FTS servers and stored on the local machine.
+Additionally, their backup was uploaded to the author's personal Yandex.Disk account, thus enabling other researchers to access the raw data even if FTS decides to remove it (as indeed occurred with the data on employees).
 
-*Second,* the filter-on-extract approach was used to extract necessary information from the raw archives into intermediary CSV tables.
-The filter-on-extract approach means that instead of unpacking the entire archive on the disk, the program reads its contents file by file without unpacking, parses the internal XML structure, extracts necessary attributes only, filters the records by activity code equal to 69.10 to leave legal companies only, and stores the result on the disk.
-Thus, the runtime and disk usage are substantially reduced compared to the traditional extract-then-filter method.
+Step 2 is data unpacking.
+A “filter-on-extract” approach was used to facilitate the retrieval of necessary information into intermediary CSV tables.
+This approach entails the program reading the contents of the archive file by file, rather than unpacking the entire archive on the disk.
+Each XML file is temporarily extracted from the archive to the memory and parsed.
+The program filters the records with the data on organizations and individuals contained in the XML file by activity code equal to 69.10 (“activity in the area of law” according to All-Russia Classifier of Kinds of Economic Activity).
+The resulting set of filtered records is then subjected to attribute extraction, and the extracted data is stored on disk.
+Once this process is complete, the memory occupied by the temporarily extracted XML file is freed, and the next file is being processed.
+The preliminary filtering by activity code decreases the runtime of the application, while the direct reading of archive contents reduces the disk space occupied by temporary files.
+Consequently, this step is executed relatively fast in comparison to the less optimized “extract, then filter” approach.
 
-*Third,* the data from the intermediary CSV tables was aggregated to remove duplicated records.
-The reason for the duplicates is the procedure of data publication: as it was mentioned earlier, every version of the source dataset contains the full data for the time of publication rather than updates only, leading to the huge number of repeating records.
+Step 3 is data aggregation.
+The intermediary CSV tables generated in the step 2 were processed to remove duplicate records.
+The reason for the duplicates is the procedure of data publication.
+As previously stated, each version of the source datasets contains the complete data for the time of publication rather than updates only.
+This results in a significant number of repeating records.
 
-*Forth,* the addresses were normalised and geocoded to convert original administrative address format containing region, district, city and settlement name into simpler yet useful triplet of region, area and settlement, and add geographical coordinates and municipal codes (ОКТМО in Russian) for the settlements.
+Step 4 is geocoding.
+This involves the addition of the geographical coordinates and municipal codes of the settlements where the SMB subjects are located to the original data table.
+Furthermore, the addresses were normalized by converting from the original administrative division format, which consists of region, district, city, and settlement names, to a simpler yet useful municipal division format with region, area, and settlement names.
 
-*Fifth,* the data on business entities was combined with the data on employees using taxpayer ID as unique identifier common to both datasets, and also converted into a panel time series table that includes yearly information for each company or individual.
+Step 5 is data panelization.
+The data from on SMB subjects was combined with the data on employees using taxpayer ID as a unique identifier common to both datasets, and also converted into a panel time series table that includes yearly information for each company or individual.
 This resulting table is used in the further analysis.
 
-The research follows the general framework of regional analysis and uses the methods of descriptive statistics and data visualization, including the plotting of data on a tile-grid map of Russia.
-The analysis consists of a few research problems, or questions, that are listed below with the description of the relevant techniques used to find answers to them.
+The research adheres to the general framework of regional analysis and employs the methods of descriptive statistics and data visualization, including the plotting of data on a tile-grid map of Russia.
+The analysis is comprised of four research problems, or questions, which are enumerated below with the description of the relevant techniques used to find the answers.
+The term “legal firm” denotes both companies (organizations) and individuals (individual entrepreneurs).
+The term “legal workers” refers to the employees of these firms.
 
-The first problem is the *distribution of legal firms by regions* of Russia.
-To answer this question, legal firms that existed in 2021 are counted by the region, and the result is visualized on the map.
+Problem 1 is the distribution of legal firms across Russia's regions.
+To discover this distribution, the median number of legal firms in each region is calculated for the period from 2016 to 2021.
+This result is then visualized on a map.
 
-The second problem is the *structure of legal business in regions* with respect to the kind of firms (either individual entrepreneurs or organisations).
-To answer this question, legal firms are counted by region and king, and the result is visualized on a geofacetted plot.
+Problem 2 is the structure of legal business in the regions with respect to the kind of firms (either individuals or companies).
+To address this issue, the median share of legal companies in each region over the 2016–2021 period is calculated and then mapped.
 
-The third problem is the *role of legal business on the regional and national scale*, that is, either its importance within a given region or the importance of region's legal services in the Russia's legal market.
-To discover this issue, two calculations are made.
-The first one is used to better understand the role of the region in the country's legal business and is done by calculating the share of legal companies employees of a region in the total number of country's legal companies employees.
-The second one is aimed at estimating the relative role of legal business in region's economy and is done by dividing the number of legal companies employees in a region by the total number of employees in a region.
-Both two metrics are then plotted on a tile-grid map.
+Problem 3 is the relative importance of regional legal business which is divided into two sub-problems.
+The first is the importance of a region's legal services in the Russian legal market measured by the share of legal workers in a region in the number of country's legal workers.
+The second is the importance of these services within a given region estimated as a share of legal workers in a region in the total number of workers in that region.
+Both metrics are calculated for each year from 2018 to 2021, averaged by a median function, and then plotted on a tile-grid map.
+Additionally, regions with district-level courts are marked on the map to analyze the relationship between the location of such courts and the development of legal business.
 
-The fourth issue deals with the temporal dimension and reflects the *change of the number of legal firms* in the regions. Legal firms are calculated by region and year, and the results are plotted.
-
-Several aspects are common to all the questions.
-First of all, despite the presence of data from 2016 to 2024, only a part from 2016 to 2021 inclusively is utilized for the analysis.
-The reason for this is the Russia's military operation in Ukraine and the subsequent structural transformation of the economy, international sanctions, substantial political and legal changes in the country that may have significantly impacted the legal market compared to the previous period, and even though all these changes could have been interesting to study, the purpose of the current research is to draw an overall picture of legal market in Russia in “normal situation” without any serious perturbations.
-Statistics for several consequent years, if not processed separately (e.
-g.
-in the case of temporal changes), is averaged using median as a function (e.g.
-number of legal firms for 2016–2021 in the region is the median of counts for each year from 2016 to 2021).
-Next, both organisations and individual entrepreneurs are accounted for.
-If statistical computations require the count of employees, it is assumed that an individual entrepreneur always has exactly one employee, that is, this person itself.
-This is because even though individual entrepreneurs in Russia can legally employ people, the corresponding information is not available in the FTS open data possibly due to personal data protection issues.
-Also, if an organisation has reported zero count of employees, it is still assumed that is has exactly one employee, that is, the company's owner itself.
+Problem 4 is the temporal change in the legal business sector across regions.
+To reach conclusions, legal firms are calculated by region and year for the 2016–2021 period.
+The time series of counts are visualized as line plots to identify temporal patterns.
+The regions are divided into the three groups based on the overall growth, decline, or stability of legal business in the observed period, and the group is also marked on the map.
 

### 2024-04-23 legal-companies-geography/Paper.Rmd
Edit paragraphs about Figure 1 and Fugure 2, modify Figure 2, make lexical changes, #1

@@
 The paper proposes a reproducible technique of working with FTS open data for regional analysis in service geography that was partially implemented in a specially designed Python command-line tool.
-The analysis includes questions about the count of legal firms and their employees by region, about the share of organisations in regional legal business, about the role of legal business in region's economy, and about the yearly change of the number of legal companies in regions.
+The analysis includes questions about the count of law firms and their employees by region, about the share of organisations in regional legal business, about the role of legal business in region's economy, and about the yearly change of the number of legal companies in regions.
 In addition, it briefly addresses an issue of probable connection between the location of multi-regional courts and the development of legal business.
@@
 The analysis is comprised of four research problems, or questions, which are enumerated below with the description of the relevant techniques used to find the answers.
-The term “legal firm” denotes both companies (organizations) and individuals (individual entrepreneurs).
-The term “legal workers” refers to the employees of these firms.
+The term “law firm” denotes both companies (organizations) and individuals (individual entrepreneurs).
+The term “lawyers” in this paper refers to the employees of these firms.
 
-Problem 1 is the distribution of legal firms across Russia's regions.
-To discover this distribution, the median number of legal firms in each region is calculated for the period from 2016 to 2021.
+Problem 1 is the distribution of law firms across Russia's regions.
+To discover this distribution, the median number of law firms in each region is calculated for the period from 2016 to 2021.
 This result is then visualized on a map.
@@
 Problem 2 is the structure of legal business in the regions with respect to the kind of firms (either individuals or companies).
-To address this issue, the median share of legal companies in each region over the 2016–2021 period is calculated and then mapped.
+To address this issue, the median share of law firms in each region over the 2016–2021 period is calculated and then mapped.
 
 Problem 3 is the relative importance of regional legal business which is divided into two sub-problems.
-The first is the importance of a region's legal services in the Russian legal market measured by the share of legal workers in a region in the number of country's legal workers.
-The second is the importance of these services within a given region estimated as a share of legal workers in a region in the total number of workers in that region.
+The first is the importance of a region's legal services in the Russian legal market measured by the share of lawyers in a region in the number of country's lawyers.
+The second is the importance of these services within a given region estimated as a share of lawyers in a region in the total number of employees in that region.
 Both metrics are calculated for each year from 2018 to 2021, averaged by a median function, and then plotted on a tile-grid map.
@@
 Problem 4 is the temporal change in the legal business sector across regions.
-To reach conclusions, legal firms are calculated by region and year for the 2016–2021 period.
+To reach conclusions, law firms are calculated by region and year for the 2016–2021 period.
 The time series of counts are visualized as line plots to identify temporal patterns.
 The regions are divided into the three groups based on the overall growth, decline, or stability of legal business in the observed period, and the group is also marked on the map.
+
+Technically, the analysis is carried out using the R programming language (including additional packages).
+A tile grid map of Russia by I.
+Dianov[^10] is used with a modified implementation by Yu.
+Tukachev[^11].
+The source code of this paper is published in the author's repository on GitHub[^12].
+
+[^10]: Плиточная карта России. Web resource: https://dianov.org/all/plitochnaya-karta-rossii/ (accessed 23.04.2024)
+[^11]: Плиточная карта РФ. Web resource: https://gist.github.com/tukachev/c3262242a20b3050e0459b4e3afc3a51 (accessed 23.04.2024)
+[^12]: PavelSyomin / ru-smb-companies-papers. Web resource: https://github.com/PavelSyomin/ru-smb-companies-papers/tree/main/legal-companies-geography (accessed 23.04.2024)
 
@@
 
-**Average yearly count of legal firms by region** is displayed on a tile-grid colorpleth map in Fig.
-1.
-Several conclusions can be drawn from this visualisation.
-First, Moscow is the only city where the number of legal firms is above 10 thousand, and this is the highest value in Russia.
-Next, a set of regions also has relatively high count of legal companies and individual entrepreneurs.
-In spatial terms, most of these regions form several groups.
-One of these groups is located at the south of the European part of Russia and consists of Krasnodar and Stavropol territories with Rostov, Volgograd and Voronezh regions.
-Tatarstan and Bashkir republics with Samara, Sverdlovsk, Chelyabinsk regions and Perm territory make up another group that may be called Ural-Volga.
-Some Siberian regions, namely Tyumen, Omsk, Novosibirsk, Kemerovo, with Krasnoyarsk territory, form the third group that may be named Siberian by its location.
-Also, Saint Petersburg city, Nizhniy Novgorod region, and Primorie territory act as single-region groups.
-The majority of other regions has relatively low count of legal firms, and in some of them (several republics of North Caucas and Altai and a portion of regions from the North) there are less than a hundred of such companies.
-The co-loocation of regions and groups with different count of legal firms suggests that there may be a sort of central-peripheries relations between them.
-For example, Moscow city and region, Saint Petersburg city, and Nizhniy Novgorod region are surrounded by dozens of regions with relatively small number of legal firms, and thus may act as a centers of legal business for them.
-The same may be true for the pairs of South and Caucasian, as well as for Siberia and Far East.
-It may be hypothesized that economic development of regions is positively associated with the development of legal business.
-Also, the population of a region may also matter.
+Figure 1 depicts the average annual number of law firms by region.
+Several conclusions can be drawn from this visualization.
+First, Moscow is the only city where the number of law firms exceeds 10 thousand, and this is the highest value in Russia.
+Second, some regions also have a relatively high number of law firms.
+Geographically, these regions form several groups.
+One of these groups is located in the south of the European part of Russia and consists of Krasnodar and Stavropol Territories with Rostov, Volgograd and Voronezh regions.
+Tatarstan and Bashkir Republics with Samara, Sverdlovsk, and Chelyabinsk Regions, as well as Perm Territory form the Ural-Volga group.
+Tyumen, Omsk, Novosibirsk, Kemerovo Regions with Krasnoyarsk Territory belong to the third group which can be designated as Siberian due to its location.
+The city of St.
+Petersburg, Nizhniy Novgorod Region, and Primorye Territory are separate regions with relatively high number of law firms.
+Third, the majority of other regions have a relatively small number of law firms, with several republics of the North Caucasus and Altai and a portion of regions of the Russian North incorporating less than a hundred of such companies.
+It can be hypothesized that the number of law firms is positively associated with the economic development of a region.
+Furthermore, the population of a region may also be a factor to consider.
+
+The  relative location of regions and groups with different numbers of law firms suggests the possibility of central-peripheries relations between them.
+For instance, Moscow City and Region, St.
+Petersburg City, and Nizhny Novgorod Region may serve as hubs for their surrounding regions of the Central and North-Western Russia.
+A similar phenomenon may be observed in other parts of the country: the Siberian group provides legal services for the Far East, while the regions of the European South serve the republics of the North Caucasus.
 
@@
 
-```{r fig-1-counts-map, fig.cap="Fig. 1. Median yearly count of legal firms (both organisations and individual entrepreneurs) by regions of Russia for 2016–21. Grey is no data"}
+```{r fig-1-counts-map, fig.cap="Fig. 1. The median annual count of law firms by regions of Russia for the period from 2016 to 2021"}
 counts_map
@@
 
-Fig.
-2 displays the *median shares of organisations among all legal firms in regions*.
-Legal firms may be either organisations or individual entrepreneurs, so for each region, the rest part of the firms consists of individuals.
-The share of companies among all legal firms varies from less about 25% to almost 70%.
-However, a substantial part of regions has from 40% to 60% of organisations, and thus many regions may be categorized either as with a small domination of companies or as with small domination of individuals on the legal market.
-Several southern regions (Krasnodar and Stavropol territories, Rostov region, Kalmyk republic), as well as several northern ones (Arkhangelsk region, Komi republic, Nenetz and Yamal territories), and some central regions (Kaluga and Moscow) have substantially lower shares of companies compared to the surrounding regions, thus forming the areas of predominantly individual legal business.
-The overall count of legal firms seems to have to connection with the share of organisation among them.
+Figure 2 illustrates the median annual share of organizations among all law firms in regions.
+The value varies considerably, ranging from approximately 30% to almost 70%.
+However, a substantial proportion of regions exhibit a share of companies ranging from 45% to 55%, which is almost equal to the share of individuals.
+All other regions may be divided into two broad groups: either having a dominance of individuals or organizations in the legal market.
+The former category encompasses the European North and South, as well as certain regions in the Fart East, Central Russia, and Western Siberia.
+The latter category comprises Moscow and St.
+Petersburg cities, along with the majority of the Volga or Siberian regions.
+The comparison of Figure 1 with Figure 2 leads to the conclusion that there is no clear relationship between the number of law firms and the share of companies among them.
 
@@
 
-```{r fig-2-shares-map, fig.cap="Fig. 2. Median yearly shares of companies among legal firms by regions of Russia for 2016–21. Grey is no data"}
+```{r fig-2-shares-map, fig.cap="Fig. 2. The median annual share of companies among legal firms by regions of Russia for the period from 2016 to 2021"}
 shares_map

### 2024-04-23 legal-companies-geography/Script.R

@@ -39,7 +39,7 @@ counts_map <- count(data, region, year) %>%
   coord_fixed(expand = FALSE) +
   theme_void(base_family = "Times New Roman", base_size = 11) +
   theme(
-    legend.position = c(1, .05),
+    legend.position = c(.9, .05),
     legend.justification = c(1, 0),
     legend.direction = "horizontal",
     legend.title.position = "top",
@@ -63,14 +63,16 @@ shares_map <- count(data, region, year, kind) %>%
     family = "Times New Roman", 
     size = 3,
     hjust = 0) +
-  scale_fill_binned(
-    name = "Share of organizations", transform = "log10",
-    n.breaks = 6,
-    high = "#eff3ff", low = "#2171b5") +
+  scale_fill_steps2(
+    name = "Share of organizations",
+    breaks = c(0, 0.45, 0.55),
+    low = "#c2a5cf", mid = "#f7f7f7", high = "#a6dba0",
+    midpoint = 0.5, show.limits = TRUE,
+    limits = function(x) round(x, 2)) +
   coord_fixed(expand = FALSE) +
   theme_void(base_family = "Times New Roman", base_size = 11) +
   theme(
-    legend.position = c(1, .05),
+    legend.position = c(.9, .05),
     legend.justification = c(1, 0),
     legend.direction = "horizontal",
     legend.title.position = "top",
@@ -78,6 +80,15 @@ shares_map <- count(data, region, year, kind) %>%
   ) +
   labs(caption = "Grey is no data")
 shares_map  
+
+count(data, region, year, kind) %>% 
+  group_by(region, year) %>%
+  mutate(share = n / sum(n)) %>% 
+  filter(kind == 1) %>% 
+  group_by(region) %>% 
+  summarise(share = median(share)) %>% 
+  ggplot(aes(x = share)) +
+  geom_density()
 
 # Employees calculations
 employees_map <- data %>% 

### 2024-04-24 legal-companies-geography/Paper.Rmd
Finish proofreading, make other minor improvements, #1

@@
   fig.width = 6.7,
-  fig.height = 5,
+  fig.height = 4,
   dpi = 300)
@@
 
-This paper provides an exploratory regional analysis of legal business in Russia for 2016–21.
-The analysis is based mainly on the Federal Tax Service (FTS) open data on small and medium-sized businesses and on the number of employees of organisations, as well as on portions of official statistical information.
-The paper proposes a reproducible technique of working with FTS open data for regional analysis in service geography that was partially implemented in a specially designed Python command-line tool.
-The analysis includes questions about the count of law firms and their employees by region, about the share of organisations in regional legal business, about the role of legal business in region's economy, and about the yearly change of the number of legal companies in regions.
-In addition, it briefly addresses an issue of probable connection between the location of multi-regional courts and the development of legal business.
-It is found out that legal business in Russia is highly concentrated in Moscow and Saint Petersburg that jointly account for 1/3 of legal firms and workers and are also at the top by the role of legal business in the region.
-However, there are other regions with relatively high development of legal business, that are located mainly on the European South of Russia, in the Volga-Urals and in Siberia.
-There is a group of regions with domination of organisations in legal business and the group of regions where individual entrepreneurs prevail over companies.
-District courts are usually located in the regions with high level of legal business development, but the mechanism of this co-location requires further investigation.
-In general, the number of legal firms in regions grown from 2016 to 2021, with the common inverted U-shape pattern probably explained by the coronavirus (covid-19) pandemic that negetively impacted the small and medium business and reverted the initial growth trend.
+This paper presents an exploratory regional analysis of legal business in Russia over the period from 2016 to 2021.
+The primary data source for the analysis is the Federal Tax Service (FTS) open data on small and medium-sized businesses and the number of employees of organizations.
+The paper proposes a reproducible methodology of working with FTS open data for regional analysis in service geography that was partially implemented in a specially designed Python command-line tool.
+The analysis includes questions regarding the number of law firms and their employees by region, the proportion of organizations engaged in the regional legal business, the role of regions in the country's legal business sector, the role of legal business in the regions' economies, and the change in the number of law firms in regions over the observed period.
+Additionally, it briefly addresses the potential correlation between the location of district-level courts and the development of legal business.
+The analysis revealed that the legal business sector in Russia is highly concentrated in Moscow and St.
+Petersburg.
+These regions also occupy the top position in terms of the role of legal business within the region.
+However, there are other regions with relatively high levels of legal business development, which are primarily located in the European south of Russia, the Volga-Urals, and Siberia.
+A dichotomy emerges between regions where legal business is dominated by organizations and those where individual entrepreneurs prevail over companies.
+District courts are typically situated in regions with a high level of legal business development.
+However, the underlying mechanism of this co-location requires further investigation.
+The number of legal firms increased from 2016 to 2021 in the majority of regions.
+The year-by-year changes within the regions typically follow an inverted U-shaped pattern.
+This pattern is likely explained by the coronavirus (COVID-19) pandemic, which negatively impacted small and medium businesses and reversed the initial growth trend.
 
@@
 Another paper briefly addresses the role of Russia in the global legal services market without delving into the internal spatial properties of the Russian legal business [@Yastrebov2021].
-In addition to the scientific publications, there are business products about the legal services market in Russia that include its regional analysis[^2].
+In addition to scientific publications, there are business products about the legal services market in Russia that include its regional analysis[^2].
 Despite a low observed interest in this issue from other researchers, it can be argued that geographical studies of legal services are essential.
-First, legal services, as a subset of other business or producer services, play a substantial role in the development of regions and territories and in the economic growth [@Hansen1993].
+First, legal services, as a subset of other business or producer services, play a substantial role in the development of regions and territories and economic growth [@Hansen1993].
 In some cases, the lack of professional legal advice may even act as a limiting factor, causing problems for entrepreneurs, especially small ones, and decreasing their performance [@Hitchens1997].
 Next, a comprehensive examination of legal companies may yield valuable insights into the business sectors they serve [@Faulconbridge2019] or enhance the theoretical understanding of markets and industries in general [@Henderson2016].
-Finally, legal services should be considered not only in terms of their narrow economic function, but also as an integral element of the broader legal system.
-This system is responsible for protecting human rights and freedoms, ensuring the rule of law, and acting as an intermediary between people, businesses and government.
-It helps to resolve conflicts, increase certainty and orderliness in social relations, and thus contributes to the stability and development of society.
+Finally, legal services should be considered not only in terms of their narrow economic function but also as an integral element of the broader legal system.
+This system is responsible for protecting human rights and freedoms, ensuring the rule of law, and acting as an intermediary between people, businesses, and government.
+It helps to resolve conflicts, increase certainty and orderliness in social relations and thus contributes to the stability and development of society.
 This aspect of legal services is closely related to the social dimension of the geography of services.
-It is therefore unsurprising that that research is being conducted on the problem of legal services availability [@Baxter2015].
+It is therefore unsurprising that research is being conducted on the problem of legal services availability [@Baxter2015].
 
@@
 Consequently, to analyze this sector, it is necessary to draw upon other sources of information.
-Georeferenced data on distinct firms is of particular value in this context, as it can be readily filtered to exclude firms that are not engaged in the requisite activity and then aggregated according to its spatial attributes to make a geographical summary.
-Researchers typically extract such the information from public geoinformation (GIS) services [@Lachininsky2021] or business intelligence (BI) systems [@Moiseeva2016].
+Georeferenced data on distinct firms is of particular value in this context, as it can be readily filtered to exclude firms that are not engaged in the requisite activity and then aggregated according to their spatial attributes to make a geographical summary.
+Researchers typically extract such information from public geoinformation (GIS) services [@Lachininsky2021] or business intelligence (BI) systems [@Moiseeva2016].
 However, this method has two disadvantages.
@@
 To address these challenges, the use of open administrative data may be a viable solution.
-In particular, the publicly available archives of the Registry of Small and Medium-Sized Businesses (SMB registry) distributed by the Federal Tax Service (FTS) of Russia, in combination with a few other open data collections of the same governmental body, appear to be a valuable resource for task at hand.
+In particular, the publicly available archives of the Registry of Small and Medium-Sized Businesses (SMB registry) distributed by the Federal Tax Service (FTS) of Russia, in combination with a few other open data collections of the same governmental body, appear to be a valuable resource for the task at hand.
 These data sources have already been used by geographers [@Rostislav2021; @Gumenyuk2022].
@@
 The SMB Registry dataset contains full dumps of the respective registry in zipped XML files.
-The initial dataset was made available in August 2016, and updates have been released on a monthly basis.
-Each dump contains the entire registry and includes, in particular, the company or individual names, taxpayer IDs, and activity codes according to All-Russia Classifier of Kinds of Economic Activity, as well as addresses of incorporation (up to the settlement name).
-According to Russian laws, the Registry includes information about all commercial organizations and individual entrepreneurs that meet several criteria (SMB subjects).
+The initial dataset was made available in August 2016, and updates have been released every month.
+Each dump contains the entire registry and includes, in particular, the company or individual names, taxpayer IDs, and activity codes according to the All-Russia Classifier of Kinds of Economic Activity, as well as addresses of incorporation (up to the settlement name).
+According to the Russian laws, the Registry includes information about all commercial organizations and individual entrepreneurs that meet several criteria (SMB subjects).
 The most important criteria are limits on the revenue (no more than 2B Russian rubles) and on the average number of employees (no more than 250).
 Consequently, the open dumps of this registry represent the most comprehensive publicly accessible data source on the Russian business entities.
-Moreover, the archive of these dumps constitutes a time series spanning from August 2016 to the present, thereby providing a history of small and medium-sized businesses changes.
+Moreover, the archive of these dumps constitutes a time series spanning from August 2016 to the present, thereby providing a history of small and medium-sized business changes.
 
@@
 For reasons that are not specified, the older versions of the dataset which included information for the years from 2018 to 2022 are not currently available on the FTS website.
-However, the data had already been downloaded by the author prior to its removal, and thus it is used in this research.
+However, the data had already been downloaded by the author before its removal, and thus it is used in this research.
 
-In addition to the primary data sources, two lookup tables with the information on Russian cities and settlements are used to normalize and geocode the raw addresses contained in the FTS data.
+In addition to the primary data sources, two lookup tables with information on Russian cities and settlements are used to normalize and geocode the raw addresses contained in the FTS data.
 The first table is “Settlements of Russia: Population and Geographical Coordinates”[^5] prepared by the nonprofit open data initiative known as “Research Data Infrastructure” (RDI).
@@
 
-**Methods.** A Python command-line, ru-smb-companies, was developed by the author of this paper to initially process large and complex raw XML datasets from the FTS servers and transform them into a single CSV table.
-The source code of this tool, along with a brief documentation, is published in the author's personal GitHub account[^9].
+Finally, an aggregated official statistic on the annual total number of employees in the regions of Russia[^9] was utilized.
+This data is provided by the Federal Statistics Service (Rosstat) through the publicly available Unified Interdepartmental Information and Statistical System.
+It contains region names and corresponding numbers of employees in all sectors of the economy for each selected year.
+
+[^9]: Численность занятых. Web resource: https://fedstat.ru/indicator/34051 (accessed 24.04.2024)
+
+**Methods.** A Python command line tool, ru-smb-companies, was developed by the author of this paper to initially process large and complex raw XML datasets from the FTS servers and transform them into a single CSV table.
+The source code of this tool, along with a brief documentation, is published in the author's personal GitHub account[^10].
 Here, a concise summary of the data processing workflow is provided.
@@
 
-[^9]: Russian small and medium business dataset generator. Web resource: https://github.com/PavelSyomin/russian-smb-companies (accessed 15.04.2024)
+[^10]: Russian small and medium business dataset generator. Web resource: https://github.com/PavelSyomin/russian-smb-companies (accessed 15.04.2024)
 
@@
 The resulting set of filtered records is then subjected to attribute extraction, and the extracted data is stored on disk.
-Once this process is complete, the memory occupied by the temporarily extracted XML file is freed, and the next file is being processed.
+Once this process is complete, the memory occupied by the temporarily extracted XML file is freed, and the next file is processed.
 The preliminary filtering by activity code decreases the runtime of the application, while the direct reading of archive contents reduces the disk space occupied by temporary files.
@@
 Step 3 is data aggregation.
-The intermediary CSV tables generated in the step 2 were processed to remove duplicate records.
+The intermediary CSV tables generated in Step 2 were processed to remove duplicate records.
 The reason for the duplicates is the procedure of data publication.
@@
 Step 5 is data panelization.
-The data from on SMB subjects was combined with the data on employees using taxpayer ID as a unique identifier common to both datasets, and also converted into a panel time series table that includes yearly information for each company or individual.
+The data on SMB subjects was combined with the data on employees using taxpayer ID as a unique identifier common to both datasets and converted into a panel time series table that includes yearly information for each company or individual.
 This resulting table is used in the further analysis.
@@
 The time series of counts are visualized as line plots to identify temporal patterns.
-The regions are divided into the three groups based on the overall growth, decline, or stability of legal business in the observed period, and the group is also marked on the map.
+The regions are divided into three groups based on the overall growth, decline, or stability of legal business in the observed period, and the group is also marked on the map.
 
@@
 A tile grid map of Russia by I.
-Dianov[^10] is used with a modified implementation by Yu.
-Tukachev[^11].
-The source code of this paper is published in the author's repository on GitHub[^12].
+Dianov[^11] is used with a modified implementation by Yu.
+Tukachev[^12].
+The source code of this paper is published in the author's repository on GitHub[^13].
 
-[^10]: Плиточная карта России. Web resource: https://dianov.org/all/plitochnaya-karta-rossii/ (accessed 23.04.2024)
-[^11]: Плиточная карта РФ. Web resource: https://gist.github.com/tukachev/c3262242a20b3050e0459b4e3afc3a51 (accessed 23.04.2024)
-[^12]: PavelSyomin / ru-smb-companies-papers. Web resource: https://github.com/PavelSyomin/ru-smb-companies-papers/tree/main/legal-companies-geography (accessed 23.04.2024)
+[^11]: Плиточная карта России. Web resource: https://dianov.org/all/plitochnaya-karta-rossii/ (accessed 23.04.2024)
+[^12]: Плиточная карта РФ. Web resource: https://gist.github.com/tukachev/c3262242a20b3050e0459b4e3afc3a51 (accessed 23.04.2024)
+[^13]: PavelSyomin / ru-smb-companies-papers. Web resource: https://github.com/PavelSyomin/ru-smb-companies-papers/tree/main/legal-companies-geography (accessed 23.04.2024)
 
@@
 Geographically, these regions form several groups.
-One of these groups is located in the south of the European part of Russia and consists of Krasnodar and Stavropol Territories with Rostov, Volgograd and Voronezh regions.
-Tatarstan and Bashkir Republics with Samara, Sverdlovsk, and Chelyabinsk Regions, as well as Perm Territory form the Ural-Volga group.
+One of these groups is located in the south of the European part of Russia and consists of Krasnodar and Stavropol Territories with Rostov, Volgograd, and Voronezh regions.
+The Republic of Bashkortostan and the Republic of Tatarstan with Samara, Sverdlovsk, and Chelyabinsk Regions, as well as Perm Territory, form the Ural-Volga group.
 Tyumen, Omsk, Novosibirsk, Kemerovo Regions with Krasnoyarsk Territory belong to the third group which can be designated as Siberian due to its location.
 The city of St.
-Petersburg, Nizhniy Novgorod Region, and Primorye Territory are separate regions with relatively high number of law firms.
+Petersburg, Nizhniy Novgorod Region, and Primorye Territory are separate regions with a relatively high number of law firms.
 Third, the majority of other regions have a relatively small number of law firms, with several republics of the North Caucasus and Altai and a portion of regions of the Russian North incorporating less than a hundred of such companies.
@@
 For instance, Moscow City and Region, St.
-Petersburg City, and Nizhny Novgorod Region may serve as hubs for their surrounding regions of the Central and North-Western Russia.
+Petersburg City, and Nizhny Novgorod Region may serve as hubs for their surrounding regions of Central and North-Western Russia.
 A similar phenomenon may be observed in other parts of the country: the Siberian group provides legal services for the Far East, while the regions of the European South serve the republics of the North Caucasus.
@@
 
-```{r fig-1-counts-map, fig.cap="Fig. 1. The median annual count of law firms by regions of Russia for the period from 2016 to 2021"}
+```{r fig-1-counts-map, fig.cap="Fig. 1. The median annual count of law firms by regions of Russia over the period from 2016 to 2021"}
 counts_map
@@
 All other regions may be divided into two broad groups: either having a dominance of individuals or organizations in the legal market.
-The former category encompasses the European North and South, as well as certain regions in the Fart East, Central Russia, and Western Siberia.
+The former category encompasses the European North and South, as well as certain regions in the Far East, Central Russia, and Western Siberia.
 The latter category comprises Moscow and St.
@@
 
-```{r fig-2-shares-map, fig.cap="Fig. 2. The median annual share of companies among legal firms by regions of Russia for the period from 2016 to 2021"}
+```{r fig-2-shares-map, fig.cap="Fig. 2. The median annual share of companies among legal firms by regions of Russia over the period from 2016 to 2021"}
 shares_map
@@
 
-Fig.
-3 shows the joint results of regional analysis of number of legal companies employees.
-Two main metrics are drawn, the first is the concentration of legal workers in the region, that is, the share of legal companies employees of a region in all country's employees of legal companies, and the second is the share of legal workers in region's labor force.
-The first measure roughly describes the importance of a region in country's legal business, while the second estimates the legal business' impact in regional economy.
-In addition, the number of multi-region courts located in a given region is shown.
-The overall regional distribution of legal companies employees resembles the regional distribution of legal companies (see Fig.
-1) suggesting that legal companies among regions are quite similar to each other, so that there is no significant difference between simple count of companies and weighted count of companies with respect to the number of workers.
-The role of legal business in region's economy is distributed differently, though.
-It's clear that Moscow and Saint Petersburg, the two largest cities concentrating almost 1/3 of legal companies employees, are also characterized by the relatively highest portion of legal workers in total labor force.
-The majority of other regions is surprisingly uniform in having the small share of legal workers (0.1–0.2%) in labor force, with Sverdlovsk, Tyumen and Novosibirsk regions being rare exceptions.
-It is also worth noticing that regions with extremely low share in country's legal business (for example, Tyva republic and Altai territory) also usually have low share of legal workers in regional labor force, with a single exception of Evrei autonomous region.
-In addition, the regional role of legal business seems to have to connection with the presence or absence of multi-regional courts.
-However, looks like that the national role of a region in legal business is somehow connected with the location of such the courts, because almost all regions with relatively high share in country's legal companies employees (>1%) have at least one multi-regional court, and the reverse is true either: other regions typically do not have such the courts.
-Of course, there are regions that do not fall within this tendencies (Bashkortostan republic and Volgograd region, as well as three sub-Moscow regions, Vologda region and Zabaikal territory), but these deviations may be explained by various factors (e.
-g.
-for Tula, Vladimir and Kaluga regions the presence of multi-regional courts is not associated with the development of legal business probably due to the proximity to Moscow, so that the need in local legal firms does not arises).
+Figure 3 presents the combined results of the regional analysis based on the number of lawyers.
+The distribution of lawyers across regions is similar to that of law firms themselves (see Figure 1), suggesting that legal companies in different regions are quite similar to each other in the number of employees.
+However, the role of legal business in regions' economies, as indicated by the share of lawyers in regions' labor forces, is distributed differently.
+Moscow and St.
+Petersburg, the two most populous cities, account for the largest number of lawyers and also exhibit the highest proportion of lawyers in the region's labor force.
+In contrast, the vast majority of other regions is characterized by a relatively uniform low share of lawyers (0.1–0.2%) in the local labor force, with Sverdlovsk, Tyumen, and Novosibirsk regions being remarkable exceptions.
+Regions with an extremely low representation in the country's legal sector (for example, the Republic of Tuva and Altai Territory) also typically have a low proportion of lawyers in their regional labor force, except for the Jewish Autonomous Region.
+
+The regional role of legal business does not appear to be contingent upon the presence or absence of district-level courts.
+However, it seems that the national role of a region in the legal sector is somehow connected with the location of such courts.
+This is evidenced by the fact that almost all regions with a relatively high share of the country's lawyers (>1%) have at least one district-level court, while other regions typically do not have such courts.
+It is important to note that there are exceptions to this trend.
+For instance, the Republic of Bashkortostan and Volgograd Region have relatively high shares of lawyers but no district-level courts.
+In contrast, three sub-Moscow regions, Vologda and Tomsk regions, as well as Khabarovsk and Trans-Baikal Territory, have an opposite situation.
+These discrepancies may be explained by several factors.
+For example, the presence of district-level courts in Tula, Vladimir, and Kaluga regions is not associated with the development of legal business, because the proximity to Moscow reduces the necessity for local law firms.
 
@@
 
-```{r fig-3-employees-map, fig.cap="Fig 3. National and regional median yearly shares of legal companies employees in regions of Russia in 2018–2021"}
+```{r fig-3-employees-map, fig.cap="Fig. 3. The national and regional median annual shares of lawyers in regions of Russia over the period from 2018 to 2021"}
 employees_map
@@
 
-The changes in the number of legal firms by region are depicted in Fig.
-4.
-It can be clearly seen that the count of such companies increased in the majority of regions, indicating the development of legal business.
-Few regions have lost companies, all of them are located in the European part of Russia, and their net change in the number of legal firms from 2016 to 2021 is relatively low (usually less than 10%) compared to the increase rates in other regions.
-Sometimes neither decline or growth was not clearly expressed (for example, Altai republic, Lipetsk and Tambov regions).
-The regions with the largest growth are some republics at Notrh Caucas, as well as Kalmyk republic and Nenetz territory.
-Their results are probably explained by the low base effect, because the absolute counts of legal firms in all there regions are below 100 (see Fig.
-1).
-As for other regions with high growth, Leningrad and Moscow regions should be remarked, because their results may be partially explained by the relocation of companies from the cities (Saint Petersburg and Moscow) to the surrounding regions.
-However, Moscow and Saint Petersburg also gained legal firms with one of the top net changes.
-The observed patterns of changes indicated by lines of the map can be divided into several groups.
-The first is the stable yet saturating increase typical for Moscow and Saint Petersburg regions, and, less expresses, Khabarovsk territory, Rostov and Ryazan regions.
-The second is the inverted U-shape curve which is charasteristic of the majority of regions with overall growth in the number of legal firms.
-This pattern can be explained by the coronavirus (COVID-19) pandemic effect that has negatively impacted small and medium-sized business and reverted the previous trend of the growth.
-The “height” of this inverted U varies by region (the highest observed in Chukotka territory, Tyva and Adygeya republics).
-It can be hypothesised that the first group is in fact a subset of the second, but the intrinsic growth rate before the pandemic was higher, resulting in the slowing down of the growth rather than in the final decrease.
-The third pattern is typical for the regions with overall decrease in the number of legal firms, and it can be described as “initial short-term growth, then continuous decline”.
-The fourth group is a mixture of increase and decrease, for example Nenetz territory and Northern Osetia republic.
+The changes in the number of law firms by region are depicted in Figure 4.
+The number of such firms increased in the majority of regions, indicating the development of the legal business sector.
+In contrast, several regions, all of which are located in the European part of Russia, have experienced a net decrease in the number of law firms, In comparison to the increase rates observed in other regions, the net change in the number of law firms from 2016 to 2021 in these regions is relatively low (usually less than 10%).
+In some cases,  neither decline nor growth was clearly expressed (for example, Republic of Altai, Lipetsk, and Tambov regions).
+The regions with the most significant growth are some republics in the North Caucasus, as well as the Republic of Kalmykia and Nenets Autonomous Area.
+Their results are likely attributable to the low base effect, given that the absolute counts of law firms in all these regions are below 100 (see Figure 1).
+It is noteworthy that Leningrad and Moscow regions also demonstrated one of the highest growth rates.
+It is likely that the relocation of law firms from the cities of Moscow and St.
+Petersburg to the surrounding regions has contributed to this outcome.
+Nevertheless, both Moscow and St.
+Petersburg cities also exhibited high net gain of law firms.
+
+The observed patterns of change indicated by line plots on the map can be divided into several groups.
+The first is the stable yet saturating increase typical for Moscow and Saint Petersburg regions, and, to a lesser extent, Khabarovsk Territory, Rostov, and Ryazan regions.
+The second is the inverted U-shaped curve which is characteristic of the majority of regions with overall growth in the number of legal firms.
+This pattern can be explained by the effect of the coronavirus (COVID-19) pandemic, which has negatively impacted small and medium-sized businesses and reversed the previous trend of growth.
+The height of the inverted U varies by region, with the highest observed in Chukotka Autonomous Area, Republic of Tyva, and Republic of Adygeya.
+It can be hypothesized that the first group is a subset of the second, but that its intrinsic growth rate before the pandemic was higher, resulting in a slowing down of the growth rather than in a final decrease.
+The third pattern is typical of the regions with a general fall in the number of law firms, and it can be described as a continuous decline following an initial period of short-term growth.
+The fourth group is a mixture of increase and decrease.
+For example, Nenets Autonomous Area and Republic of North Ossetia–Alania demonstrate this pattern.
 
@@
 
-```{r fig-4-timeline-map, fig.cap="Fig. 4. The change in the number of legal firms by region in 2016–2021"}
+```{r fig-4-timeline-map, fig.cap="Fig. 4. The change in the number of law firms by region over the period from 2016 to 2021"}
 timeline_map
@@
 
-The conducted regional analysis discovered several spatial features of legal business in Russia.
-First, the regions are quite different with respect to the count of legal firms or their employees.
-Several regional clusters with relatively high concentration of legal business can be found out, for example, the European South, Volga-Ural and Siberian.
-Unsurprisingly, Moscow and Saint Petersburg cities are characterized by the top count of legal firms and workers, but the overall concentration is really huge, because they jointly account for approximately 1/3 of Russia's legal businesses and legal companies employees.
-Besides, it's again Moscow and Saint Petersburg that have the highest share of legal companies employees in region's labor force (above 0.4% that is 2 or 3 times higher than the average value for other regions).
-However, there are some other regions with relatively high role of legal business in regional economy: Sverdlosvk, Tyumen and Novosibirsk.
+The regional analysis revealed several spatial features of legal business in Russia.
+First, the regions exhibited considerable variation in the number of law firms and lawyers.
+Several regional clusters with relatively high development of legal business were identified, including the European South, Volga-Ural, and Siberian.
+As might be expected, the cities of Moscow and St.
+Petersburg are home to the largest number of law firms and lawyers.
+Furthermore, it is again Moscow and St.
+Petersburg cities that have the highest share of lawyers in their total labor force.
+Nevertheless, there are other regions with a relatively high role of legal businesses in their regional economies, including Sverdlovsk, Tyumen, and Novosibirsk.
 
-Regions can be roughly grouped into two categories: one with the legal organisations prevailing in the legal business, and the other with individual entrepreneurs prevailing over companies.
-The high role of companies is typical for Moscow, Saint Petersburg, some Volga and Siberian regions, as well as a small portion of central regions (Vologda and Ivanovo), while individuals are important in the European South and North, in the majority of Central European regions, in several regions of the Urals, and in the Far East.
+Regions can be roughly divided into three categories: one with the legal organizations prevailing in the legal business, another with individual entrepreneurs prevailing over companies, and the last one where the proportion of individuals is almost equal to the proportion of companies.
+The high role of companies is typical for Moscow, Saint Petersburg, Volga, and Siberian regions.
+The opposite is observed in the European North and South, as well as certain regions in the Far East, Central Russia, and Western Siberia
 
-The role of region in country's legal business is usually associated with the presence of one or many multi-regional courts in the administrative center of this region.
-Only two regions do not act as a host for any of such the courts but still have relatively high level of development of legal business: Volgograd region and Bashkir republic.
-This observation may be explained in two ways: either the courts act as a magnets for legal business, because courts are necessary for many legal activities, or the courts are placed in regions and cities with highly developer legal business.
-Further research may throw the light on the direction of this court-business relation.
+The role of the region in the country's legal business is typically associated with the presence of one or more district-level courts in the administrative center of the region.
+Only two regions do not act as hosts for any of these courts but still have a relatively high level of legal business development: Volgograd Region and Republic of Bashkortostan.
+This observation can be explained in two ways.
+One possibility is that the courts act as a magnet for legal business because they are necessary for many legal activities.
+Alternatively, the courts are placed in regions and cities with already highly developed legal business sector.
+Further research may shed light on the direction of this court-business relationship.
 
-From 2016 to 2021, the number of legal firms in Russia's regions mostly increased.
-The decrease is characteristic only of few regions.
-The top growth was observed in Moscow, Saint Petersburg, as well as in the surrounding regions.
-In addition, some regions (mainly at the European South) also shown very high growth, but probably due to the low base effect.
-The yearly patterns of change may be categorized into four groups, and the inverted U-shape change explained likely by the coronavirus (covid-19) pandemic is the most common among the regions.
+From 2016 to 2021, the number of law firms in the majority of Russia's regions generally increased.
+However, a decrease was observed in some regions.
+The cities of Moscow and St.
+Petersburg with the surrounding regions are characterized by one of the highest increase rates.
+Additionally, some regions (mainly in the North Caucasus) exhibited extreme growth, but this may be attributed to the low base effect.
+The yearly patterns of change can be categorized into four groups.
+The inverted U-shaped change, which is likely explained by the coronavirus (COVID-19) pandemic, is the most common pattern among the regions.
 
-The results of this explanatory study provide an overview of Russia's recent legal business geography and may be used as a support for data-driven business and governmental decisions.
-Several research hypothesis also may be conducted from this analysis, thus this study may be treated as a base for further academic development in the field of legal services geography in Russia.
-Apart from this, the reproducible data-driven technique of this research based entirely on open data and open source software tools may be used (with minor changes) for regional analysis and mapping of other narrow service sector areas.
+The results of this exploratory study provide an overview of Russia's recent legal business geography and may be used as support for data-driven business and governmental decisions.
+Several research hypotheses may also be conducted from this analysis, thus this study may be treated as a base for further academic development in the field of legal services geography in Russia.
+Furthermore, the reproducible data-driven methodology employed in this study, which is based entirely on open data and open source software tools may be utilized (with minor modifications) for regional analysis and mapping of other narrow service sector areas.
 

### 2024-04-24 legal-companies-geography/Script.R

@@ -33,19 +33,26 @@ counts_map <- count(data, region, year) %>%
     family = "Times New Roman", 
     size = 3,
     hjust = 0) +
+  geom_text(
+    aes(x = col - .4, y = -row - .3, label = round(n, 0)), 
+    family = "Times New Roman", 
+    size = 3,
+    color = "gray20",
+    hjust = 0,
+    na.rm = TRUE) +
   scale_fill_binned(
-    name = "Count of legal firms", transform = "log10",
-    high = "#eff3ff", low = "#2171b5") +
+    name = "Median count of law firms (2016–21)\nNumbers on the map are exact values\nGrey fill is no data",
+    transform = "log10",
+    high = "#eff3ff", low = "#2171b5"
+  ) +
   coord_fixed(expand = FALSE) +
   theme_void(base_family = "Times New Roman", base_size = 11) +
   theme(
     legend.position = c(.9, .05),
     legend.justification = c(1, 0),
     legend.direction = "horizontal",
-    legend.title.position = "top",
-    plot.caption = element_text(size = 11)
-  ) +
-  labs(caption = "Grey is no data")
+    legend.title.position = "top"
+  )
 counts_map
 
 # Shares of companies among legal firms
@@ -63,9 +70,16 @@ shares_map <- count(data, region, year, kind) %>%
     family = "Times New Roman", 
     size = 3,
     hjust = 0) +
+  geom_text(
+    aes(x = col - .4, y = -row - .3, label = round(100 * share)), 
+    family = "Times New Roman", 
+    size = 3,
+    color = "grey20",
+    hjust = 0) +
   scale_fill_steps2(
-    name = "Share of organizations",
+    name = "Median share of organizations (2016–21)\nNumbers on the map are exact values (%)\nGrey is no data",
     breaks = c(0, 0.45, 0.55),
+    labels = scales::percent,
     low = "#c2a5cf", mid = "#f7f7f7", high = "#a6dba0",
     midpoint = 0.5, show.limits = TRUE,
     limits = function(x) round(x, 2)) +
@@ -75,20 +89,9 @@ shares_map <- count(data, region, year, kind) %>%
     legend.position = c(.9, .05),
     legend.justification = c(1, 0),
     legend.direction = "horizontal",
-    legend.title.position = "top",
-    plot.caption = element_text(size = 11)
-  ) +
-  labs(caption = "Grey is no data")
+    legend.title.position = "top"
+  )
 shares_map  
-
-count(data, region, year, kind) %>% 
[…]

### 2024-04-25 legal-companies-geography/Paper.Rmd
Make final improvements before the first submission, #1

@@
   fig.height = 4,
+  fig.ext = "jpeg",
   dpi = 300)
@@
 
+<div custom-style="Abstract">
+
 This paper presents an exploratory regional analysis of legal business in Russia over the period from 2016 to 2021.
 The primary data source for the analysis is the Federal Tax Service (FTS) open data on small and medium-sized businesses and the number of employees of organizations.
-The paper proposes a reproducible methodology of working with FTS open data for regional analysis in service geography that was partially implemented in a specially designed Python command-line tool.
+The paper proposes a reproducible methodology of working with FTS open data for regional analysis in service geography that was partially implemented in a specially designed Python command line tool.
 The analysis includes questions regarding the number of law firms and their employees by region, the proportion of organizations engaged in the regional legal business, the role of regions in the country's legal business sector, the role of legal business in the regions' economies, and the change in the number of law firms in regions over the observed period.
@@
 This pattern is likely explained by the coronavirus (COVID-19) pandemic, which negatively impacted small and medium businesses and reversed the initial growth trend.
+
+</div>
 
@@
 In some cases, the lack of professional legal advice may even act as a limiting factor, causing problems for entrepreneurs, especially small ones, and decreasing their performance [@Hitchens1997].
-Next, a comprehensive examination of legal companies may yield valuable insights into the business sectors they serve [@Faulconbridge2019] or enhance the theoretical understanding of markets and industries in general [@Henderson2016].
+Next, a comprehensive examination of law firms may yield valuable insights into the business sectors they serve [@Faulconbridge2019] or enhance the theoretical understanding of markets and industries in general [@Henderson2016].
 Finally, legal services should be considered not only in terms of their narrow economic function but also as an integral element of the broader legal system.
@@
 To address these challenges, the use of open administrative data may be a viable solution.
-In particular, the publicly available archives of the Registry of Small and Medium-Sized Businesses (SMB registry) distributed by the Federal Tax Service (FTS) of Russia, in combination with a few other open data collections of the same governmental body, appear to be a valuable resource for the task at hand.
+In particular, the publicly available archives of the Registry of Small and Medium-Sized Businesses (SMB Registry) distributed by the Federal Tax Service (FTS) of Russia, in combination with a few other open data collections of the same governmental body, appear to be a valuable resource for the task at hand.
 These data sources have already been used by geographers [@Rostislav2021; @Gumenyuk2022].
@@
 
-**Data.** The research utilizes two primary data sources: the Unified Registry of Small and Medium-Sized Businesses (SMB Registry)[^3] and the Information about the Average Number of Employees of Organisations (Information about Employees)[^4].
+**Data.** The research utilizes two primary data sources: the Unified Registry of Small and Medium-Sized Businesses (SMB Registry)[^3] and the Information about the Average Number of Employees of Organizations (Information about Employees)[^4].
 Both of these are available on the FTS website as open data.
@@
 Each XML file is temporarily extracted from the archive to the memory and parsed.
-The program filters the records with the data on organizations and individuals contained in the XML file by activity code equal to 69.10 (“activity in the area of law” according to All-Russia Classifier of Kinds of Economic Activity).
+The program filters the records with the data on organizations and individuals contained in the XML file by activity code equal to 69.10 (“activity in the area of law” according to the All-Russia Classifier of Kinds of Economic Activity).
 The resulting set of filtered records is then subjected to attribute extraction, and the extracted data is stored on disk.
@@
 This involves the addition of the geographical coordinates and municipal codes of the settlements where the SMB subjects are located to the original data table.
-Furthermore, the addresses were normalized by converting from the original administrative division format, which consists of region, district, city, and settlement names, to a simpler yet useful municipal division format with region, area, and settlement names.
+Furthermore, the addresses were normalized by converting from the original administrative division format, which consists of region, district, city, and settlement names, to a simpler yet useful format with region, area, and settlement names.
 
@@
 Problem 2 is the structure of legal business in the regions with respect to the kind of firms (either individuals or companies).
-To address this issue, the median share of law firms in each region over the 2016–2021 period is calculated and then mapped.
+To address this issue, the median share of organizations among law firms in each region over the 2016–2021 period is calculated and then mapped.
 
-Problem 3 is the relative importance of regional legal business which is divided into two sub-problems.
+Problem 3 is the relative importance of regional legal business.
+This issue is divided into two sub-problems.
 The first is the importance of a region's legal services in the Russian legal market measured by the share of lawyers in a region in the number of country's lawyers.
@@
 Geographically, these regions form several groups.
-One of these groups is located in the south of the European part of Russia and consists of Krasnodar and Stavropol Territories with Rostov, Volgograd, and Voronezh regions.
-The Republic of Bashkortostan and the Republic of Tatarstan with Samara, Sverdlovsk, and Chelyabinsk Regions, as well as Perm Territory, form the Ural-Volga group.
-Tyumen, Omsk, Novosibirsk, Kemerovo Regions with Krasnoyarsk Territory belong to the third group which can be designated as Siberian due to its location.
+One of these groups is located in the south of the European part of Russia and consists of Krasnodar and Stavropol territories with Rostov, Volgograd, and Voronezh regions.
+The Republic of Bashkortostan and the Republic of Tatarstan with Samara, Sverdlovsk, and Chelyabinsk regions, as well as Perm Territory, form the Ural-Volga group.
+Tyumen, Omsk, Novosibirsk, Kemerovo regions with Krasnoyarsk Territory belong to the third group which can be designated as Siberian due to its location.
 The city of St.
@@
 
-```{r fig-1-counts-map, fig.cap="Fig. 1. The median annual count of law firms by regions of Russia over the period from 2016 to 2021"}
+```{r fig-1-counts-map, fig.cap="Figure 1. The median annual count of law firms by regions of Russia over the period from 2016 to 2021"}
 counts_map
@@
 The latter category comprises Moscow and St.
-Petersburg cities, along with the majority of the Volga or Siberian regions.
+Petersburg cities, along with the majority of the Volga and Siberian regions.
 The comparison of Figure 1 with Figure 2 leads to the conclusion that there is no clear relationship between the number of law firms and the share of companies among them.
@@
 
-```{r fig-2-shares-map, fig.cap="Fig. 2. The median annual share of companies among legal firms by regions of Russia over the period from 2016 to 2021"}
+```{r fig-2-shares-map, fig.cap="Figure 2. The median annual share of companies among legal firms by regions of Russia over the period from 2016 to 2021"}
 shares_map
@@
 For instance, the Republic of Bashkortostan and Volgograd Region have relatively high shares of lawyers but no district-level courts.
-In contrast, three sub-Moscow regions, Vologda and Tomsk regions, as well as Khabarovsk and Trans-Baikal Territory, have an opposite situation.
+In contrast, three sub-Moscow regions, Vologda and Tomsk regions, as well as Khabarovsk and Trans-Baikal territories, have an opposite situation.
 These discrepancies may be explained by several factors.
@@
 
-```{r fig-3-employees-map, fig.cap="Fig. 3. The national and regional median annual shares of lawyers in regions of Russia over the period from 2018 to 2021"}
+```{r fig-3-employees-map, fig.cap="Figure 3. The national and regional median annual shares of lawyers in regions of Russia over the period from 2018 to 2021"}
 employees_map
@@
 In contrast, several regions, all of which are located in the European part of Russia, have experienced a net decrease in the number of law firms, In comparison to the increase rates observed in other regions, the net change in the number of law firms from 2016 to 2021 in these regions is relatively low (usually less than 10%).
-In some cases,  neither decline nor growth was clearly expressed (for example, Republic of Altai, Lipetsk, and Tambov regions).
+In some cases, neither decline nor growth was clearly expressed (for example, the Republic of Altai, Lipetsk, and Tambov regions).
 The regions with the most significant growth are some republics in the North Caucasus, as well as the Republic of Kalmykia and Nenets Autonomous Area.
@@
 The observed patterns of change indicated by line plots on the map can be divided into several groups.
-The first is the stable yet saturating increase typical for Moscow and Saint Petersburg regions, and, to a lesser extent, Khabarovsk Territory, Rostov, and Ryazan regions.
+The first is the stable yet saturating increase typical for Moscow and Leningrad regions, and, to a lesser extent, Khabarovsk Territory, Rostov, and Ryazan regions.
 The second is the inverted U-shaped curve which is characteristic of the majority of regions with overall growth in the number of legal firms.
 This pattern can be explained by the effect of the coronavirus (COVID-19) pandemic, which has negatively impacted small and medium-sized businesses and reversed the previous trend of growth.
-The height of the inverted U varies by region, with the highest observed in Chukotka Autonomous Area, Republic of Tyva, and Republic of Adygeya.
+The height of the inverted U varies by region, with the highest observed in Chukotka Autonomous Area, the Republic of Tyva, and the Republic of Adygeya.
 It can be hypothesized that the first group is a subset of the second, but that its intrinsic growth rate before the pandemic was higher, resulting in a slowing down of the growth rather than in a final decrease.
@@
 The fourth group is a mixture of increase and decrease.
-For example, Nenets Autonomous Area and Republic of North Ossetia–Alania demonstrate this pattern.
+For example, Nenets Autonomous Area and the Republic of North Ossetia–Alania demonstrate this pattern.
 
@@
 
-```{r fig-4-timeline-map, fig.cap="Fig. 4. The change in the number of law firms by region over the period from 2016 to 2021"}
+```{r fig-4-timeline-map, fig.cap="Figure 4. The change in the number of law firms by region over the period from 2016 to 2021"}
 timeline_map
@@
 The role of the region in the country's legal business is typically associated with the presence of one or more district-level courts in the administrative center of the region.
-Only two regions do not act as hosts for any of these courts but still have a relatively high level of legal business development: Volgograd Region and Republic of Bashkortostan.
+Only two regions do not act as hosts for any of these courts but still have a relatively high level of legal business development: Volgograd Region and the Republic of Bashkortostan.
 This observation can be explained in two ways.

### 2024-04-25 legal-companies-geography/Script.R

@@ -167,10 +167,10 @@ timeline_map <- count(data, region, year) %>%
     aes(
       x = 2021,
       y = 1.25, 
-      label = if_else(!is.na(change), sprintf("%+.0f", change), ""),
+      label = ifelse(!is.na(change), sprintf("%+.0f", change), ""),
       alpha = alpha),
     family = "Times New Roman",
-    size = 4,
+    size = 3.5,
     hjust = 1,
     color = "gray45",
     key_glyph = "blank") +
@@ -179,7 +179,7 @@ timeline_map <- count(data, region, year) %>%
     data = tiles,
     aes(x = 2016, y = 2, label = code_en),
     family = "Times New Roman",
-    size = 3,
+    size = 2.5,
     hjust = 0,
     vjust = 1) +
   scale_y_continuous(limits = c(NA, 2)) +
@@ -203,6 +203,7 @@ timeline_map <- count(data, region, year) %>%
     legend.direction = "horizontal",
     legend.title.position = "top",
     legend.text = element_text(size = 11),
+    panel.spacing = unit(1, "mm"),
     strip.background = element_blank(),
     strip.text = element_blank()
   )

### 2024-05-11 legal-companies-concentration-and-availability/conference-paper/Paper.Rmd
Split project files into subfolders to prepare for journal article writing on the same topic, adjust paths using here package, #2

--- 2024-03-06 legal-companies-concentration-and-availability/Paper.Rmd
@@
 knitr::opts_chunk$set(
+  root.dir = "../",
   echo = FALSE,
@@
 ```{r imports, include=FALSE, cache = TRUE}
-source("Script.R")
+library(here)
+source(here("common", "Script.R"))
 ```

### 2024-05-11 legal-companies-concentration-and-availability/common/Script.R

@@ -1,17 +1,18 @@
 library(dplyr)
 library(ggplot2)
 library(glue)
+library(here)
 library(readr)
 library(sf)
 library(tidyr)
 
 # Data on firms, and cities lookup table
-panel <- read_csv("../../ru-smb-companies/legal/panel.csv")
+panel <- read_csv(here("../../ru-smb-companies/legal/panel.csv"))
 cities <- rbind(
-  read_csv("cities.csv"),
-  read_csv("cities_additional.csv")
+  read_csv(here("common", "cities.csv")),
+  read_csv(here("common", "cities_additional.csv"))
 )
-regions_geo <- st_read("ru.geojson")
+regions_geo <- st_read(here("common", "ru.geojson"))
 
 lcc <- panel %>% 
   filter(

### 2024-05-11 legal-companies-concentration-and-availability/journal-article/Article.Rmd
Cover info, introduction, materials and methods for journal article, #2

---
output:
  word_document:
    reference: reference.docx
bibliography: refs.bib
csl: style.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  dpi = 300,
  fig.id = FALSE,
  fig.width = 6.6,
  fig.height = 5)
```

```{r imports, include=FALSE, cache = TRUE}
library(here)
source(here("common", "Script.R"))
```

IRSTI 06.61.53

<div custom-style="Title">

Provision of Commercial Legal Services in Russian Cities

</div>

<div custom-style="Author">

Pavel O. Syomin

<br>

Perm State University, Russia, Perm

<br>

e-mail: ntsp\@ya.ru

</div>

<div custom-style="Abstract">

**Abstract:** The article examines the relationship between the concentration of legal companies and the availability of legal services. To conduct the analysis, a dataset was created using open data from the Federal Tax Service of Russia. The dataset includes information on the location, revenue, expenditure, and number of employees of small and medium-sized legal companies. The concentration is calculated by dividing the number of employees of legal companies in the city by the total number of employees of all legal companies in Russia. Availability is measured as the number of employees of legal companies in the city per 100,000 population. Data visualization and linear regression were utilized to estimate the hypothesized relationship. The study showed a positive linear relationship between concentration and availability. However, there are multiple approaches to achieving good availability of legal services.

**Key words:** legal services, law firms, availability, cities, market saturation, small and medium-sized businesses, open data

</div>

<div custom-style="Title">

Ресей қалаларында коммерциялық заң қызметтерін көрсету

</div>

<div custom-style="Author">

Павел Олегович Сёмин

<br>

Пермь мемлекеттік ұлттық зерттеу университеті, Ресей, Пермь

<br>

e-mail: ntsp\@ya.ru

</div>

<div custom-style="Abstract">

**Аннотация:** The article examines the relationship between the concentration of legal companies and the availability of legal services. To conduct the analysis, a dataset was created using open data from the Federal Tax Service of Russia. The dataset includes information on the location, revenue, expenditure, and number of employees of small and medium-sized legal companies. The concentration is calculated by dividing the number of employees of legal companies in the city by the total number of employees of all legal companies in Russia. Availability is measured as the number of employees of legal companies in the city per 100,000 population. Data visualization and linear regression were utilized to estimate the hypothesized relationship. The study showed a positive linear relationship between concentration and availability. However, there are multiple approaches to achieving good availability of legal services.

**Кілт сөздер:** legal services, concentration, availability, small and medium-sized businesses, open data

</div>

<div custom-style="Title">

Обеспеченность коммерческими юридическими услугами в городах России

</div>

<div custom-style="Author">

Павел Олегович Сёмин

<br>

Пермский государственный национальный исследовательский университет, Россия, Пермь

<br>

e-mail: ntsp\@ya.ru

</div>

<div custom-style="Abstract">

**Аннотация:** The article examines the relationship between the concentration of legal companies and the availability of legal services. To conduct the analysis, a dataset was created using open data from the Federal Tax Service of Russia. The dataset includes information on the location, revenue, expenditure, and number of employees of small and medium-sized legal companies. The concentration is calculated by dividing the number of employees of legal companies in the city by the total number of employees of all legal companies in Russia. Availability is measured as the number of employees of legal companies in the city per 100,000 population. Data visualization and linear regression were utilized to estimate the hypothesized relationship. The study showed a positive linear relationship between concentration and availability. However, there are multiple approaches to achieving good availability of legal services.

**Ключевые слова:** юридические услуги, юридические фирмы, обеспеченность, города, насыщенность рынка, малое и среднее предпринимательство, открытые данные

</div>

# Introduction

Geography of services is an important and well-established field of geographical studies. Its history dates back to the second half of 20th century [@Tkachenko2016]. Two research approaches exist in geography of services: the first can be described as economic, and the second can be designated as social. The economic approach, as its name suggests, treats the services as a regular part of the economy, dealing with their role in GDP or employment, their impact on the growth, the interplay between services and other business activities. For example, a research by D. Wójcik [-@Wojcik2020] provides an overview of FihTech geography in an economic-based manner. The social approach, while closely related to the economic one, focuses on the services as the foundation of people's well-being and studies services provision, availability and accessibility, as well as the diversity and quaility of services. For instance, the study in Indonesia [@Pratiwi2021] uses the various socio-economic data from the national health insurance system grouped by ~500 country's districts to analyze the relationship between a set of public health quality metrics and discuss, among the other problems, the equity problems arising from the geographical differences. The issue of access to various services has been of great importance for decades [@Blacksell1990; @Keleher1996] because it is directly connected to the quality of life and ensuring of human rights. Research has also been conducted on the links between market properties and service availability [@Lu2020].

Cities have traditionally been viewed as regional centers that provide goods and services to smaller surrounding settlements [@Christaller1966]. The urban studies are thus tightly connected with the geography of services [@Reilly1931; @Brush1953]. It is well-known that professional and business services tend to cluster in large cities, and their spatial distribution is closely linked to urban hierarchies [@Keeble2002; @Rubalcaba2013; @Kekezi2019]. The concentration of businesses provides economic benefits for the companies and thus is likely to become self-sustaining. The common sense suggests that such concentration does not only leads to the economic growth, but also positively impacts the service provision, thus being productive in both economic and social dimensions. It is usually the largest cities or the regional centres where the local communities benefit from the variety and quantity of diverse service providers, compared to smaller settlements and rural areas, where the availability of services is limited due to the low number of their providers. However, is this relationship true indeed? The common sense can easily turn out to be wrong, and thus the rigorous data-driven testing of the connection between the number of service providers and the provision of services is crucial.

This paper empirically examines the relationship between the number of law firms in Russian cities and the provision of legal services. Typically, the literature on services availability focuses on medical services [@Joseph1982], but seems that legal services, though being understudied by geographers, are also worth attention [@Blacksell1988; @Patel2008], because they are directly related to the rule of law, protection of human rights, supporting orderliness and stability of social relations. The analysis is based on the open data about small and medium-sized businesses (SMB) published by the Federal Tax Service (FTS) of Russia. The main research question is how the specific availability of legal professionals is affected by the concentration of law firms. Additionally, the study aims to find out geographical patterns in the spatial distribution of legal services availability.

# Materials and methods

**Data.** This research utilizes a geo-referenced dataset on small and medium-sized law firms incorporated in Russia. The dataset is created from FTS open data dumps (archives) of the state SMB Registry[^1] and FTS open data on the average number of employees of ogranizations[^2] using the Python command-line tool *ru-smb-companies*[^3] developed by the author of the paper. It contains information about the particular companies, including company's name, taxpayer ID, region and settlement of incorporation, main activity code. Law firms are selected by the main activity code equal to 69.10, and all the other companies were dropped from the source data. This technique of selection by the main activity code was derived from the analytical paper on the Russian legal services market prepared by the members of the Institute for the Rule of Law [@Moiseeva2016]. Additionally, the dataset was filtered by the following conditions: the year is 2021, the company is an organization rather than a sole entrepreneur, the company has declared non-zero revenue and expenditure for the target year (implying that the company is active), and the company's address of registration is in a city rather than in a rural area. During the exploratory data analysis, four cities (Innopolis, Kirovsk, Bronnytsy, Krasnoarmeisk) were identified as outliers based on the number of law firms or their employees per 100,000 population and were subsequently excluded from the data. In addition to the main dataset, a data on cities population freely distributed by the commercial company HFLabs is used [^4].

[^1]: https://www.nalog.gov.ru/opendata/7707329152-rsmp/
[^2]: https://www.nalog.gov.ru/opendata/7707329152-sshr2019/
[^3]: https://github.com/PavelSyomin/russian-smb-companies
[^4]: https://github.com/hflabs/city

**Methods.** A linear regression is the main method of data analysis used in this study. An independent variable for the regression is the number of employees of law firms in a city denoted as *n*. Theoretically, both number of law firms or their employees may be used, but law firms in different cities may have different size, thus the number of employees seems to be more stable and consistent metric. The value for each city is calculated as a sum of numbers of employees of each law firm incorporated in this city. If a law firm reported zero count of employees, it is assumed that the number of employees is equal to one. The dependent variable in the regression is specific provision of legal services which is denoted as *p* and calculated by dividing the number of employees of law firms in the city by the city's population and multiplying the result by 100,000. In other words, it's a number of law firms employees per 100,000 of population. This formula is a modified version of the legal market saturation metric used in the analysis of the Russian legal services market conducted by the Institute for the Rule of Law [@Moiseeva2016]. In addition, a grouping variable *g* categorizing the cities by population is used to check whether the overall relationship is dependent on the city size. The grouping by population is performed on the basis of the classification system of G. M. Lappo [-@Lappo1997]. Two linear models are fitted, one without grouping and one with grouping. The results are visialized on a plot. The analysis was performed using the R programming language and additional packages. The source code for this paper can be found in the public repository[^2].

[^2]: https://github.com/PavelSyomin/ru-smb-companies-papers/tree/main/legal-companies-concentration-and-availability

**Theoretical models.** Despite the perceived simplicity of the modelling described earlier, it has a theoretical mathematical background, and several possible types of outcome may be laid out right now before the analysis of the data in hand. It is known that a linear model is described by the two coefficients, intercept (a free term of the linear equation showing the relative shift of the regression line along the y axis) and slope (a coefficient before the independent variable showing the angle between the regression line and x axis). The slope can be either positive, negative, or zero. As a special edge case, it can be also infinite (i. e. regression line is vertical), even though, strictly speaking, such a dependency is not a function. Thus, the linear fit described earlier can yield a result of one of four types, each corresponding to one of the slope options and, subsequently, the pattern of relationship between the number of law firms and provision of legal services. These basic theoretical types are shown in Figure 1.

```{r figure-1-theoretical-models, fig.cap="Figure 1 — Four plots showing four types of the theoretical relationship between the number of law firms employees and provision of lwgal services in cities"}
theoretical_models_plot
```

Type 1 corresponds to the zero slope. In this case, the regression line is horizontal. From the substantial point of view, it describes the absolute equality in the provision of legal services among cities, that is, the provision is independent of the number of law firms employees. Despite being quite unlikely, this type also may be described as optimal, because it guarantees equitable access to legal services (at least in spatial terms). Probably this type requires some sort of governmental intervention to form (e. g. encouraging the establishment of law firms in small cities).

Type 2 is characterized by the positive slope and, consequently, positive linear relationship between the number of law firms and the provision of legal services. It seems that at the free market this type is the most probable, even though the specific value of the slope may vary. In such case, the bigger the city, the easier it is to find a lawyer, and vise versa.

Type 3 is the reverse of Type 2: the slope of the regression line is negative, and so is the relationship between the number of law firms and the provision of legal services. It is also quite improbable. One possible situation in which this type may form is the unexpected and rapid migration of people from the smallest cities to the largest with almost uniform (Type 1) relationship as a prerequisite. Under such circumstances, the relative number of law firms in small cities increases fast, but the relative number of law firms in large cities can not grow immediately as fast as the population, and the largest cities start to suffer from the lack of legal services, while the smallest cities are oversaturated with law firms.

Type 4 is purely theoretical, but important in conjunction with the three other types. As Type 1, it also signals about the independence between the number of law firms and provision of legal services, but here the number of law firms employees is fixed rather than their proportion in the population. Again, it is very unlikely to happen in the real world, but it also suggests that some sort of governmental intervention (e. g. strictly controlling the number of law firms) is taking place.

# Results and discussion

Figure 1 shows the relationship between the concentration of legal companies and the availability of legal services, as measured by the metrics proposed in the Methods section. The plot displays the groups by city population and city rank within the region, in addition to the general trend. The overall connection between the concentration and availability appears to be linear. The linear model results demonstrate that the relationship between the two variables can be described by the formula *`r ca_model_formula`*, where both the intercept and slope are highly significant *(p < 0.001)*, and the explained variance is relatively high *(R^2^ = `r round(ca_model_summary$adj.r.squared, 2)`)*.

Examining the regression line common to all data points, we see that while the linearity is generally good and the linear modeling results are satisfactory, the total distribution is fan-shaped. This means that the variance of error of the linear model decreases as both the dependent and independent variables increase. However, this is mainly due to the presence of several distinct population-size groups of cities in the data. Within each of these groups, linear models satisfy the assumptions better. The link between concentration and availability is similar across all city population groups, and the regression lines are nearly parallel, indicating equal slope coefficients and differing only in intercepts. Additionally, it can be concluded that, in general, larger cities have higher concentrations and availability, although the relationship is relatively weak.

<br>

```{r fig-1-concentration-and-accessibility, fig.cap="Figure 1. The relationship between the concentration of legal companies and the availability of legal services. City population group and relative city rank within the region are marked. The formulas used to calculate the metrics are described in the Methods section. Both the X and Y axes are log10-scaled. Dotted regression lines represent the linear fits within city population groups, while the solid gray line represents the overall linear fit. The regression line for the millionaire cities is fitted excluding the outliers (Moscow and Saint Petersburg)", fig.height=4}
ca_plot
```

In general, cities with the top rank by the population in their respective regions tend to have better availability of legal services. One possible reason for this is the considerably big size of such cities.

Moscow and Saint Petersburg are two remarkable outliers with extremely high concentration of legal services and a good level of legal services availability. Moscow alone accounts for almost one-third of all legal companies' employees, and Saint Petersburg accounts for about 10%. The position of the corresponding data points relative to the main regression lines suggests that there is an “availability saturation” level, where further concentration has little effect on availability.

A high concentration of legal companies, a large population, and the city's top rank by population within a region are positively associated with the availability of legal services. Conversely, larger cities require more legal companies to meet the needs of their residents. As a result, there are various ways to achieve high availability of legal services, ranging from being a business and service center with a high concentration of legal firms to being a small yet well-served city where a small number of legal firms meet the needs of the local community.

Figure 2 displays the spatial distribution of cities based on the availability of legal services. Regional centers exhibit a high relative level of availability. The most prominent cluster of good access to legal services is formed by Moscow and its satellite cities. Other notable centers with high availability include Krasnodar and Stavropol in the southern part of Russia's West, Yekaterinburg and Chelyabinsk in the Urals, and Tuymen and Novosibirsk in Western Siberia. There are regions with limited access to legal services, such as the bordering regions in the south of Siberia, as well as the Komi and Kalmyk Republics.

<br>

```{r fig-2-map, fig.cap="Figure 2. Legal services availability in Russian cities on the country's map", fig.height=4}
spatial_plot
```

# Conclusion

The concentration of legal companies in cities has a positive linear relationship with the availability of legal services. This trend is observed in all the city population groups. The largest cities in regions typically have relatively high availability of legal services, as do cities with population above 1 million. Moscow and Saint Petersburg are characterized by the highest concentration and high availability, with Moscow alone accounting for almost one-third of all legal companies employees in the country. Empirical data suggests that there may be a saturation limit beyond which the increase in concentration does not have a substantial impact on availability. There are multiple paths to achieving high levels of legal services availability, with the increasing concentration of legal companies in a city being just one option. Even cities with relatively low concentration can achieve similar levels of availability as larger ones. Moscow and its region are a cluster of areas with high legal services availability. Similarly, Saint Petersburg, Krasnodar, Stavropol, Yekaterinburg, Chelyabinsk, Tyumen, and Novosibirsk also have good access to legal services.

## References

<div id="refs"></div>

### 2024-05-11 legal-companies-concentration-and-availability/common/Script.R

@@ -128,3 +128,60 @@ spatial_plot <- data %>%
   scale_size_discrete(name = "City size by population", range = c(.2, 2)) +
   theme_bw(base_size = 14, base_family = "Times New Roman")
 spatial_plot
+
+# Theoretical models
+facet_labels <- c(
+  "Type 1: s = 0",
+  "Type 2: s > 0",
+  "Type 3: s < 0",
+  "Type 4: s → ∞")
+lines <- data.frame(
+  x = c(0, 1, 0, 1, 0, 1, 0.5, 0.5),
+  y = c(1, 1, 0, 1, 1, 0, 0, 1),
+  group = factor(
+    c(1, 1, 2, 2, 3, 3, 4, 4),
+    levels = 1:4, 
+    labels = facet_labels
+  )
+)
+data_points <- rbind(
+  tibble(
+    x = runif(20),
+    y = jitter(rep(1, 20), 5),
+    group = 1
+  ),
+  tibble(
+    x = runif(20),
+    y = x + jitter(rep(0, 20), 5),
+    group = 2
+  ),
+  tibble(
+    x = runif(20),
+    y = 1 - (x + jitter(rep(0, 20), 5)),
+    group = 3
+  ),
+  tibble(
+    x = jitter(rep(0.5, 20), 5),
+    y = runif(20),
+    group = 4
+  )
+) %>% 
+  mutate(group = factor(group, 1:4, labels = facet_labels))
+theoretical_models_plot <- ggplot(lines, aes(x = x, y = y)) +
+  geom_path() +
+  geom_point(data = data_points, size = .5, alpha = .3) +
+  scale_x_continuous(
+    name = "Number of law firms (LF) employees (n)",
+    labels = function (br) br * 1000,
+    limits = c(-.1, 1.1)
+  ) +
+  scale_y_continuous(
+    name = "LF per 100,000 (p)",
+    labels = function (br) br * 10,
+    limits = c(-.1, 1.1)
+  ) +
+  facet_wrap(vars(group), nrow = 1) +
+  coord_fixed() +
+  theme_minimal(base_size = 16) +
+  labs(caption = "s means slope, dots are hypothetical data points, line is hypothetical linear fit of points")
+theoretical_models_plot

### 2024-05-14 legal-companies-concentration-and-availability/journal-article/Article.Rmd
Modify plots, edit results and discussion, #2

@@
 library(here)
-source(here("common", "Script.R"))
+source(here("journal-article", "Script.R"))
 ```
@@
 Additionally, the dataset was filtered by the following conditions: the year is 2021, the company is an organization rather than a sole entrepreneur, the company has declared non-zero revenue and expenditure for the target year (implying that the company is active), and the company's address of registration is in a city rather than in a rural area.
-During the exploratory data analysis, four cities (Innopolis, Kirovsk, Bronnytsy, Krasnoarmeisk) were identified as outliers based on the number of law firms or their employees per 100,000 population and were subsequently excluded from the data.
 In addition to the main dataset, a data on cities population freely distributed by the commercial company HFLabs is used [^4].
@@
 [^4]: https://github.com/hflabs/city
+
+During the exploratory data analysis, six cities are identified as outliers subsequently excluded from the data.
+Two of these cities, Moscow and Saint Petersburg, are excluded due to the extremely high number of law firms employees: almost 18,866 and 5,550 respectively, while the next city has only less than 2,000 employees.
+The four other cities (Innopolis, Kirovsk, Bronnytsy, Krasnoarmeisk) are characterized by the very high proportion of law firms employees in city's population (from 38 per 10,000 in Krasnoarmeisk to 417 in Innopolis) compared to less than 20 per 10,000 in all the other cities.
 
@@
 
-**Theoretical models.** Despite the perceived simplicity of the modelling described earlier, it has a theoretical mathematical background, and several possible types of outcome may be laid out right now before the analysis of the data in hand.
+**Theoretical model.** Despite the perceived simplicity of the modelling described earlier, it has a theoretical mathematical background, and several possible types of outcome may be laid out right now before the analysis of the data in hand.
 It is known that a linear model is described by the two coefficients, intercept (a free term of the linear equation showing the relative shift of the regression line along the y axis) and slope (a coefficient before the independent variable showing the angle between the regression line and x axis).
@@
 
-Figure 1 shows the relationship between the concentration of legal companies and the availability of legal services, as measured by the metrics proposed in the Methods section.
-The plot displays the groups by city population and city rank within the region, in addition to the general trend.
-The overall connection between the concentration and availability appears to be linear.
-The linear model results demonstrate that the relationship between the two variables can be described by the formula *`r ca_model_formula`*, where both the intercept and slope are highly significant *(p < 0.001)*, and the explained variance is relatively high *(R^2^ = `r round(ca_model_summary$adj.r.squared, 2)`)*.
-
-Examining the regression line common to all data points, we see that while the linearity is generally good and the linear modeling results are satisfactory, the total distribution is fan-shaped.
-This means that the variance of error of the linear model decreases as both the dependent and independent variables increase.
-However, this is mainly due to the presence of several distinct population-size groups of cities in the data.
-Within each of these groups, linear models satisfy the assumptions better.
-The link between concentration and availability is similar across all city population groups, and the regression lines are nearly parallel, indicating equal slope coefficients and differing only in intercepts.
-Additionally, it can be concluded that, in general, larger cities have higher concentrations and availability, although the relationship is relatively weak.
+Figure 1 shows the relationship between the concentration of legal companies and the availability of legal services, as measured by the metrics proposed in the Materials and methods section.
+The overall connection between the number of law firms and provision of legal services appears to be nonlinear and corresponds to a square root formula.
+However, the same relationship within each city size group is fitted well with a linear models.
+The coefficients of these models are different, and the slope decreases when the city population increases.
+While the regression lines for the small and medium cities are close to Type 4 (provision is independent of the number of lawyers because of the small variation in the number of lawers), the modelled relationship in millionaire cities resembles Type 1 (another option for independence, but due to relatively uniform and equal provision regardless the number of lawyers).
+The cities from big to extra-large are characterized by the positive relationship between the number of lawyers and the provision of legal services and can be classified as Type 2.
 
@@
 
-In general, cities with the top rank by the population in their respective regions tend to have better availability of legal services.
-One possible reason for this is the considerably big size of such cities.
+In general, regional centers have relatively high number of lawyers and considerable level of legal services provision.
+This observation is probably explained by the centralization tendencies in the regions and the concentration of business and services in their administrative centers that situate the majority of government bodies and are also usually the most populous cities within a region.
 
-Moscow and Saint Petersburg are two remarkable outliers with extremely high concentration of legal services and a good level of legal services availability.
+Moscow and Saint Petersburg, though excluded from the data analysis and not shown on the main figure, are two remarkable outliers with extremely high concentration of legal services and a good level of legal services availability.
 Moscow alone accounts for almost one-third of all legal companies' employees, and Saint Petersburg accounts for about 10%.
@@
 
-A high concentration of legal companies, a large population, and the city's top rank by population within a region are positively associated with the availability of legal services.
-Conversely, larger cities require more legal companies to meet the needs of their residents.
-As a result, there are various ways to achieve high availability of legal services, ranging from being a business and service center with a high concentration of legal firms to being a small yet well-served city where a small number of legal firms meet the needs of the local community.
+The general tendencies observed in the Figure 1 should be supplemented by the analysis of the source data distribution shown in Figure 2 that extends the Figure 1 and displays the binned count of cities according to the number of lawyers and the provision of legal services.
+The overwhelming majority of cities is characterized by low level of both legal services provision and law firms development.
+Only a limited number (18) of cities is characterized by the high (more than 500) number of lawyers.
+It should be noted that high number of lawyers is good for legal services provision because all these cities are at least at the medium level of provision, and none of them has less than 5 lawyers per 10,000 people.
+However, cities with relatively low number of laywers can nevertheless reach top performance from the point of legal services provision that is on par with the cities with the highest number of lawyers.
+Moreover, it appears that such small yet well-served cities are even more typical than cities with high level of both lawyers and provision.
+The relative location of bin counts in the Figure 2 implies that there — besides the dominant bottom left corner — there are two main clusters of cities showing two different approaches of reaching high level of legal services provision.
+One approach is increase both in absolute number of law firms and relative number of lawyers per population probably charasteristic of the largest cities.
+Another is a relatively small yet sufficient number of lawyers typical probably for small or medium cities.
+Thus, apart from the general tendencies, the two-ways structure is found out.
 

### 2024-05-14 legal-companies-concentration-and-availability/journal-article/Script.R

@@ -1,0 +1,329 @@
+library(dplyr)
+library(ggplot2)
+library(glue)
+library(here)
+library(readr)
+library(sf)
+library(tidyr)
+
+# Load data
+panel <- read_csv(here("../../ru-smb-companies/legal/panel.csv"))
+cities <- rbind(
+  read_csv(here("common", "cities.csv")),
+  read_csv(here("common", "cities_additional.csv"))
+)
+regions_geo <- st_read(here("common", "ru.geojson"))
+tiles <- read_csv(here("journal-article", "russia-tiles.csv"))
+
+# Build table with number of employees by city
+# lfc means Law Firms by City
+lfc <- panel %>% 
+  filter(
+    kind == 1, 
+    year == 2021, 
+    settlement_type == "г",
+    revenue > 0,
+    expenditure > 0
+  ) %>% 
+  mutate(city = case_when(
+    region == "Москва" & is.na(settlement) ~ "Москва",
+    region == "Санкт-Петербург" & is.na(settlement) ~ "Санкт-Петербург",
+    TRUE ~ settlement
+  )) %>% 
+  select(region, city, oktmo, empl = employees_count) %>% 
+  replace_na(list(empl = 0)) %>% 
+  mutate(empl = replace(empl, empl == 0, 1)) %>% 
+  group_by(region, city, oktmo) %>% 
+  summarise(count = n(), empl = sum(empl), .groups = "drop")
+
+# Build table with cities population
+# cp means Cities Population
+cp <- cities %>% 
+  mutate(
+    city = coalesce(cities$city, cities$area, cities$region),
+    is_regional_center = case_when(
+      city == "Москва" ~ TRUE,
+      city == "Санкт-Петербург" ~ TRUE,
+      capital_marker == 2 ~ TRUE,
+      TRUE ~ FALSE
+    )) %>% 
+  select(city, oktmo, population, 
+         lat = geo_lat, lon = geo_lon,
+         is_regional_center)
+
+# Make joint data table for analysis
+data <- left_join(lfc, cp, by = c("oktmo", "city")) %>% 
+  drop_na(count, empl, population) %>% 
+  mutate(
+    provision = 1e4 * empl / population,
+    city_group = cut(
+      population,
+      breaks = c(0, 5e4, 1e5, 2.5e5, 5e5, 1e6, 1e9),
[…]

### 2024-05-16 legal-companies-concentration-and-availability/journal-article/Article.Rmd
Finish results and discussion, edit conclusion, edit map and plots, drop unused code, #2

@@
 ```{r setup, include=FALSE}
+library(dplyr)
+library(ggplot2)
+library(glue)
+library(here)
+library(readr)
+library(sf)
+library(tidyr)
+
 knitr::opts_chunk$set(
@@
 ```{r imports, include=FALSE, cache = TRUE}
-library(here)
 source(here("journal-article", "Script.R"))
@@
 
-<br>
-
 e-mail: ntsp\@ya.ru
@@
 The analysis was performed using the R programming language and additional packages.
-The source code for this paper can be found in the public repository[^2].
+The source code for this paper can be found in the public repository[^5].
 
-[^2]: https://github.com/PavelSyomin/ru-smb-companies-papers/tree/main/legal-companies-concentration-and-availability
+[^5]: https://github.com/PavelSyomin/ru-smb-companies-papers/tree/main/legal-companies-concentration-and-availability
 
@@
 
-```{r figure-1-theoretical-models, fig.cap="Figure 1 — Four plots showing four types of the theoretical relationship between the number of law firms employees and provision of lwgal services in cities"}
+```{r figure-1-theoretical-models, fig.cap="**Figure 1** — Four plots showing four types of the theoretical relationship between the number of law firms employees and provision of lwgal services in cities", fig.height=3}
 theoretical_models_plot
@@
 
-Figure 1 shows the relationship between the concentration of legal companies and the availability of legal services, as measured by the metrics proposed in the Materials and methods section.
+Figure 2 shows the relationship between the concentration of legal companies and the availability of legal services, as measured by the metrics proposed in the Materials and methods section.
 The overall connection between the number of law firms and provision of legal services appears to be nonlinear and corresponds to a square root formula.
@@
 
-```{r fig-1-concentration-and-accessibility, fig.cap="Figure 1.
-The relationship between the concentration of legal companies and the availability of legal services.
-City population group and relative city rank within the region are marked.
-The formulas used to calculate the metrics are described in the Methods section.
-Both the X and Y axes are log10-scaled.
-Dotted regression lines represent the linear fits within city population groups, while the solid gray line represents the overall linear fit.
-The regression line for the millionaire cities is fitted excluding the outliers (Moscow and Saint Petersburg)", fig.height=4}
-ca_plot
+```{r fig-2-empl-provision, fig.cap="**Figure 2** — The relationship between the number of lawyers and provision of legal services", fig.height=5}
+ep_plot
 ```
@@
 
-The general tendencies observed in the Figure 1 should be supplemented by the analysis of the source data distribution shown in Figure 2 that extends the Figure 1 and displays the binned count of cities according to the number of lawyers and the provision of legal services.
+The general tendencies observed in the Figure 2 should be supplemented by the analysis of the source data distribution shown in Figure 3 that extends the Figure 2 and displays the binned count of cities according to the number of lawyers and the provision of legal services.
 The overwhelming majority of cities is characterized by low level of both legal services provision and law firms development.
@@
 
-Figure 2 displays the spatial distribution of cities based on the availability of legal services.
-Regional centers exhibit a high relative level of availability.
-The most prominent cluster of good access to legal services is formed by Moscow and its satellite cities.
-Other notable centers with high availability include Krasnodar and Stavropol in the southern part of Russia's West, Yekaterinburg and Chelyabinsk in the Urals, and Tuymen and Novosibirsk in Western Siberia.
-There are regions with limited access to legal services, such as the bordering regions in the south of Siberia, as well as the Komi and Kalmyk Republics.
+```{r figure-3-empl-provision-binned, fig.cap="**Figure 3** — Count and proportion of cities with variuos combinations of number of lawyers and level of legal services provision"}
+ep_binned_plot
+```
+
+Figure 4 is a map displaying the location of cities with relatively high (more than 5 lawyers per 10,000 people) provision of legal services.
+The most prominent cluster is formed by Moscow and its satellite cities.
+Other notable centers with high availability include Krasnodar and Stavropol in the South of Russia, Yekaterinburg and Chelyabinsk in the Urals, and Tuymen and Novosibirsk in Western Siberia.
+There are several regions with no cities of good legal services provision (especially on Far East, North, and Caucasus), and the majority of cities with relatively high level of legal services provision is located in the Central and Volga economic regions of the country.
 
@@
 
-```{r fig-2-map, fig.cap="Figure 2. Legal services availability in Russian cities on the country's map", fig.height=4}
-spatial_plot
+```{r fig-4-map, fig.cap="**Figure 4** — Legal services provision in Russian cities on the country's map. Only cities with more than 5 lawyers per 10,000 people are shown", fig.height=4}
+map
 ```
@@
 
-The concentration of legal companies in cities has a positive linear relationship with the availability of legal services.
-This trend is observed in all the city population groups.
-The largest cities in regions typically have relatively high availability of legal services, as do cities with population above 1 million.
-Moscow and Saint Petersburg are characterized by the highest concentration and high availability, with Moscow alone accounting for almost one-third of all legal companies employees in the country.
-Empirical data suggests that there may be a saturation limit beyond which the increase in concentration does not have a substantial impact on availability.
-There are multiple paths to achieving high levels of legal services availability, with the increasing concentration of legal companies in a city being just one option.
-Even cities with relatively low concentration can achieve similar levels of availability as larger ones.
-Moscow and its region are a cluster of areas with high legal services availability.
-Similarly, Saint Petersburg, Krasnodar, Stavropol, Yekaterinburg, Chelyabinsk, Tyumen, and Novosibirsk also have good access to legal services.
+The provision of legal services in cities has a positive relationship with the number of lawyers.
+The overall trend can be fitted with a square root curve, while the relationship in distinct city size groups is close to linear with varying slope of the regression line.
+In small and medium cities (with population below 50,000), the increase of number of lawyers has little effect on the level of legal services provision, because the number of lawyers per 10,000 people varies greatly.
+In larger cities, the relationship is linear and positive.
+In the largest cities having the population above 1 million people, the impact of the growth of the number of lawyers on the provision of legal services decreases, because the latter in these cities is relatively high and uniform.
+Regional centers typically have relatively high availability of legal services.
+Moscow and Saint Petersburg are characterized by the highest number of law firms and high provision of legal services.
+Empirical data suggests that there may be a saturation limit beyond which the increase in the number of law firms does not have a substantial impact on provision of legal servicies.
+
+The overwhelming majority (about 85%) of cities is characterized both by low number of law firms and low level of legal services provision.
+There are two groups of cities by their way of achieving high levels of legal services provision.
+The first group is composed of cities with large number of law firms.
+The second group, on the opposite, have small yet number of law firms that nevertheless satisfy the needs of local communitues.
+Moscow and its region is characterized by relatively high level of legal services availability, as well as Central and Volga regions of Russia in general.
+Similarly, Saint Petersburg, Krasnodar, Stavropol, Yekaterinburg, Chelyabinsk, Tyumen, and Novosibirsk also have good provision to legal services.
+On the contrary, in many regions of Far East, North and Caucasus the level of provision of legal services is low.
 

### 2024-05-16 legal-companies-concentration-and-availability/journal-article/Script.R

@@ -56,20 +56,16 @@ data <- left_join(lfc, cp, by = c("oktmo", "city")) %>%
   drop_na(count, empl, population) %>% 
   mutate(
     provision = 1e4 * empl / population,
-    city_group = cut(
+    size = cut(
       population,
-      breaks = c(0, 5e4, 1e5, 2.5e5, 5e5, 1e6, 1e9),
+      breaks = c(0, 5e4, 1e5, 2.5e5, 5e5, 1e6, 1e8),
       labels = c(
         "Small (<50k)", "Medium (50–100k)", "Big (100–250k)",
         "Large (250–500k)", "Extra-large (500k–1M)", 
         "Millionaire (>1M)")
     )) %>% 
   group_by(region) %>% 
-  mutate(
-    city_region_rank = row_number(-population),
-    is_largest_in_region = city_region_rank == 1) %>% 
-  ungroup() %>% 
-  select(-city_region_rank)
+  ungroup()
 
 # Look at the regression variables extreme values
 # to identify outliers
@@ -82,26 +78,41 @@ data %>%
 outliers <- c("Иннополис", "Кировск", "Бронницы", "Красноармейск", 
               "Москва", "Санкт-Петербург")
 
+ep_with_outliers_plot <- data %>% 
+  mutate(
+    is_msk_spb = city %in% c("Москва", "Санкт-Петербург")) %>% 
+  filter(!(city %in% outliers) | is_msk_spb) %>%
+  ggplot(aes(x = empl, y = provision)) +
+  geom_point(
+    aes(color = is_msk_spb), 
+    size = 1,
+    shape = 21) +
+  geom_text(
+    data = ~ filter(.x, is_msk_spb) %>% mutate(city = if_else(city == "Москва", "Msk", "SPb")), 
+    aes(label = city),
+    vjust = "inward",
+    hjust = "inward") +
+  scale_color_manual(
+    name = NULL,
+    values = c("TRUE" = "red1", "FALSE" = "gray50"),
+    guide = guide_none()) +
+  theme_bw() +
+  theme(
+    axis.title = element_blank(),
+    panel.grid.minor = element_blank())
+ep_with_outliers_plot
 
-ca_model <- lm(log(empl_per_100k) ~ log(empl), data)
-ca_model_summary <- summary(ca_model)
-ca_model_formula = glue(
-  "log10(availability) = {intercept} {sign} {slope} × log10(concentration)",
[…]

### 2024-05-17 legal-companies-zipf-law/Slides.qmd
Slides for the Maksakov conference

---
title: "Зависимость «ранг-размер» в распределении городов России по числу юрфирм"
author: "Сёмин Павел Олегович, Пермский государственный национальный исследовательский университет"
format: revealjs
---

```{r}
#| label: imports
#| include: FALSE
#| cache: TRUE
source("Script.R")
```

# Задачи

1. Понять, как число юрфирм и юристов связано с населением города
2. Проверить, выполняется ли правило «ранг-размер» в распределении городов по числу юрфирм и юристов

# Данные

1. Число юрфирм и юристов — открытые данные ФНС России (реестр МСП, сведения о среднесписочной численности работников организаций)
2. Население городов — справочник «Города России» от HFLabs

# Методы

1. Линейная регрессия: log10(y) ~ log10(x)
2. Визуализация данных

# Результаты

## Число юрфирм/юристов ~ население

```{r}
counts_and_population_plot
```

## Ранг-размер

```{r}
rank_size_plot
```

# Выводы

1. Число юрфирм и юристов связано с населением города почти линейно
2. Правило «ранг-размер» в основном выполняется

### 2024-05-17 legal-companies-concentration-and-availability/journal-article/Article.Rmd
Proofreading of introduction, CSL to Chicago, #2

@@
 
-**Аннотация:** The article examines the relationship between the concentration of legal companies and the availability of legal services.
-To conduct the analysis, a dataset was created using open data from the Federal Tax Service of Russia.
-The dataset includes information on the location, revenue, expenditure, and number of employees of small and medium-sized legal companies.
-The concentration is calculated by dividing the number of employees of legal companies in the city by the total number of employees of all legal companies in Russia.
-Availability is measured as the number of employees of legal companies in the city per 100,000 population.
-Data visualization and linear regression were utilized to estimate the hypothesized relationship.
-The study showed a positive linear relationship between concentration and availability.
-However, there are multiple approaches to achieving good availability of legal services.
+**Аннотация:**
 
-**Кілт сөздер:** legal services, concentration, availability, small and medium-sized businesses, open data
+**Кілт сөздер:** заң қызметтері, заң фирмалары, қауіпсіздік, қалалар, нарықтың қанықтығы, шағын және орта бизнес, ашық деректер
 
@@
 
-**Аннотация:** The article examines the relationship between the concentration of legal companies and the availability of legal services.
-To conduct the analysis, a dataset was created using open data from the Federal Tax Service of Russia.
-The dataset includes information on the location, revenue, expenditure, and number of employees of small and medium-sized legal companies.
-The concentration is calculated by dividing the number of employees of legal companies in the city by the total number of employees of all legal companies in Russia.
-Availability is measured as the number of employees of legal companies in the city per 100,000 population.
-Data visualization and linear regression were utilized to estimate the hypothesized relationship.
-The study showed a positive linear relationship between concentration and availability.
-However, there are multiple approaches to achieving good availability of legal services.
+**Аннотация:**
 
@@
 
-Geography of services is an important and well-established field of geographical studies.
-Its history dates back to the second half of 20th century [@Tkachenko2016].
-Two research approaches exist in geography of services: the first can be described as economic, and the second can be designated as social.
-The economic approach, as its name suggests, treats the services as a regular part of the economy, dealing with their role in GDP or employment, their impact on the growth, the interplay between services and other business activities.
-For example, a research by D. Wójcik [-@Wojcik2020] provides an overview of FihTech geography in an economic-based manner.
-The social approach, while closely related to the economic one, focuses on the services as the foundation of people's well-being and studies services provision, availability and accessibility, as well as the diversity and quaility of services.
-For instance, the study in Indonesia [@Pratiwi2021] uses the various socio-economic data from the national health insurance system grouped by ~500 country's districts to analyze the relationship between a set of public health quality metrics and discuss, among the other problems, the equity problems arising from the geographical differences.
-The issue of access to various services has been of great importance for decades [@Blacksell1990; @Keleher1996] because it is directly connected to the quality of life and ensuring of human rights.
-Research has also been conducted on the links between market properties and service availability [@Lu2020].
+Geography of services is a significant and well-established branch of geographical studies.
+The discipline's history can be traced back to the second half of the 20th century, as documented by @Tkachenko2016.
+Two distinct research approaches exist within the field of geography of services.
+The first may be described as economic, while the second may be designated as social.
+The economic approach, as its name suggests, treats the services as a regular part of the economy.
+It considers their role in GDP or employment, their impact on growth, and the interplay between services and other business activities.
+For instance, a study by @Wojcik2020 offers an economic-based overview of FihTech geography.
+The social approach, while closely related to the economic one, focuses on the services as the foundation of people's well-being and studies the provision, availability and accessibility of services, as well as their diversity and quality.
+For example, the study in Indonesia [@Pratiwi2021] employs a variety of socio-economic data from the national health insurance system to discuss, among other issues, the equity problems that arise from geographical differences.
+A problem of access to various services has been a significant concern for decades [@Blacksell1990; @Keleher1996], as it is directly related to the quality of life and the guarantee of human rights.
+Furthermore, research has been conducted on the relationship between geographical properties and service availability [@Holzer2000; @Johnson2006].
 
-Cities have traditionally been viewed as regional centers that provide goods and services to smaller surrounding settlements [@Christaller1966].
-The urban studies are thus tightly connected with the geography of services [@Reilly1931; @Brush1953].
-It is well-known that professional and business services tend to cluster in large cities, and their spatial distribution is closely linked to urban hierarchies [@Keeble2002; @Rubalcaba2013; @Kekezi2019].
-The concentration of businesses provides economic benefits for the companies and thus is likely to become self-sustaining.
-The common sense suggests that such concentration does not only leads to the economic growth, but also positively impacts the service provision, thus being productive in both economic and social dimensions.
-It is usually the largest cities or the regional centres where the local communities benefit from the variety and quantity of diverse service providers, compared to smaller settlements and rural areas, where the availability of services is limited due to the low number of their providers.
-However, is this relationship true indeed?
-The common sense can easily turn out to be wrong, and thus the rigorous data-driven testing of the connection between the number of service providers and the provision of services is crucial.
+Cities have traditionally been regarded as regional hubs that provide goods and services to smaller surrounding settlements [@Christaller1966].
+Consequently, urban studies are closely intertwined with the geography of services [@Reilly1931; @Brush1953].
+It is widely acknowledged that professional and business services tend to cluster, and their spatial distribution is closely linked to urban hierarchies [@Bennett1999; @Keeble2002; @Rubalcaba2013; @Kekezi2019].
+The concentration of businesses provides economic benefits for the companies and, as such, the concentration is likely to become self-sustaining.
+It is reasonable to assume that such concentration does not only leads to the economic growth, but also positively impacts the service provision, thereby being productive in both economic and social dimensions.
+It is a common observation that the largest cities and regional centers offer the greater variety and quantity of diverse service providers than smaller settlements and rural areas, where the availability of services is limited due to the low number of their providers.
+Nevertheless, it must be asked: is this relationship truly valid?
+The common-sense notion that the number of service providers is correlated with the provision of services may, in fact, be incorrect.
+Thus, rigorous data-driven testing of this hypothesis is necessary.
 
 This paper empirically examines the relationship between the number of law firms in Russian cities and the provision of legal services.
-Typically, the literature on services availability focuses on medical services [@Joseph1982], but seems that legal services, though being understudied by geographers, are also worth attention [@Blacksell1988; @Patel2008], because they are directly related to the rule of law, protection of human rights, supporting orderliness and stability of social relations.
+The existing literature on the availability of services typically focuses on medical services [@Joseph1982], with relatively little attention paid to legal services.
+Nevertheless, there is a growing recognition that spatial aspects of legal services, though understudied by geographers, are worthy of investigation [@Blacksell1988; @Patel2008].
+This is because they are directly related to the rule of law, the protection of human rights, and the support of social order and stability.
 The analysis is based on the open data about small and medium-sized businesses (SMB) published by the Federal Tax Service (FTS) of Russia.
-The main research question is how the specific availability of legal professionals is affected by the concentration of law firms.
-Additionally, the study aims to find out geographical patterns in the spatial distribution of legal services availability.
+The primary research question is how the specific provision of legal professionals is influenced by the number of lawyers.
+Additionally, the study aims to identify geographical patterns in the spatial distribution of legal services provision.
 

### 2024-05-18 legal-companies-concentration-and-availability/journal-article/Article.Rmd
Proofread of materials and methods, results and discussion, conclusion; add three references to introduction, #2

@@
 This paper empirically examines the relationship between the number of law firms in Russian cities and the provision of legal services.
-The existing literature on the availability of services typically focuses on medical services [@Joseph1982], with relatively little attention paid to legal services.
+The existing literature on the availability of services typically focuses on medical services [@Joseph1982; @vanDoorslaer2006; @Weiss2020; @Maleki2024], with relatively little attention paid to legal services.
 Nevertheless, there is a growing recognition that spatial aspects of legal services, though understudied by geographers, are worthy of investigation [@Blacksell1988; @Patel2008].
@@
 The analysis is based on the open data about small and medium-sized businesses (SMB) published by the Federal Tax Service (FTS) of Russia.
-The primary research question is how the specific provision of legal professionals is influenced by the number of lawyers.
+The primary research question is how the specific provision of legal services is influenced by the number of lawyers (employees at law firms).
 Additionally, the study aims to identify geographical patterns in the spatial distribution of legal services provision.
@@
 
-**Data.** This research utilizes a geo-referenced dataset on small and medium-sized law firms incorporated in Russia.
-The dataset is created from FTS open data dumps (archives) of the state SMB Registry[^1] and FTS open data on the average number of employees of ogranizations[^2] using the Python command-line tool *ru-smb-companies*[^3] developed by the author of the paper.
-It contains information about the particular companies, including company's name, taxpayer ID, region and settlement of incorporation, main activity code.
-Law firms are selected by the main activity code equal to 69.10, and all the other companies were dropped from the source data.
-This technique of selection by the main activity code was derived from the analytical paper on the Russian legal services market prepared by the members of the Institute for the Rule of Law [@Moiseeva2016].
-Additionally, the dataset was filtered by the following conditions: the year is 2021, the company is an organization rather than a sole entrepreneur, the company has declared non-zero revenue and expenditure for the target year (implying that the company is active), and the company's address of registration is in a city rather than in a rural area.
-In addition to the main dataset, a data on cities population freely distributed by the commercial company HFLabs is used [^4].
+**Data.** This research uses a geocoded dataset on small and medium-sized law firms incorporated in Russia.
+The dataset is derived from FTS open data dumps (archives) of the state SMB Registry[@SMBRegistry2024] and FTS open data on the average number of employees of organizations[@FTSEmployees2024], which were processed using the Python command-line tool *ru-smb-companies* developed by the author of the paper.
+The dataset contains information about distinct companies, including their names, taxpayer IDs, regions and settlements of incorporation, and main activity codes.
 
-[^1]: https://www.nalog.gov.ru/opendata/7707329152-rsmp/
-[^2]: https://www.nalog.gov.ru/opendata/7707329152-sshr2019/
-[^3]: https://github.com/PavelSyomin/russian-smb-companies
-[^4]: https://github.com/hflabs/city
+Law firms were selected based on a main activity code equal of 69.10.
+This selection technique was derived from the analytical paper on the Russian legal services market prepared by the members of the Institute for the Rule of Law [@Moiseeva2016].
+Furthermore, the dataset was subjected to the following filters: the year was set to 2021, the company was identified as an organization rather than a sole entrepreneur, the company had declared non-zero revenue and expenditure for the target year (implying that the company was active), and the company's address of registration was in a city rather than in a rural area.
+In addition to the main dataset, a data on city population was utilized from a freely distributed commercial source [@HFLabs2021].
 
-During the exploratory data analysis, six cities are identified as outliers subsequently excluded from the data.
-Two of these cities, Moscow and Saint Petersburg, are excluded due to the extremely high number of law firms employees: almost 18,866 and 5,550 respectively, while the next city has only less than 2,000 employees.
-The four other cities (Innopolis, Kirovsk, Bronnytsy, Krasnoarmeisk) are characterized by the very high proportion of law firms employees in city's population (from 38 per 10,000 in Krasnoarmeisk to 417 in Innopolis) compared to less than 20 per 10,000 in all the other cities.
+During the exploratory data analysis, six cities were identified as outliers and subsequently excluded from the dataset.
+Two of the cities, Moscow and Saint Petersburg, were removed due to the extremely high number of employees at law firms: 18,866 and 5,550, respectively.
+In contrast, the next city had a considerably smaller number of employees, with less than 2,000.
+The remaining four cities (Innopolis, Kirovsk, Bronnytsy, Krasnoarmeisk) are distinguished by a markedly elevated proportion of law firms' employees in their respective populations.
+This proportion ranges from 38 per 10,000 in Krasnoarmeisk to 417 per 10,000 in Innopolis, in stark contrast to less than 20 per 10,000 in all other cities.
+After applying all the filters and removing the outliers, the dataset contain information about 576 cities.
 
-**Methods.** A linear regression is the main method of data analysis used in this study.
-An independent variable for the regression is the number of employees of law firms in a city denoted as *n*.
-Theoretically, both number of law firms or their employees may be used, but law firms in different cities may have different size, thus the number of employees seems to be more stable and consistent metric.
-The value for each city is calculated as a sum of numbers of employees of each law firm incorporated in this city.
-If a law firm reported zero count of employees, it is assumed that the number of employees is equal to one.
-The dependent variable in the regression is specific provision of legal services which is denoted as *p* and calculated by dividing the number of employees of law firms in the city by the city's population and multiplying the result by 100,000.
-In other words, it's a number of law firms employees per 100,000 of population.
-This formula is a modified version of the legal market saturation metric used in the analysis of the Russian legal services market conducted by the Institute for the Rule of Law [@Moiseeva2016].
-In addition, a grouping variable *g* categorizing the cities by population is used to check whether the overall relationship is dependent on the city size.
-The grouping by population is performed on the basis of the classification system of G. M. Lappo [-@Lappo1997].
-Two linear models are fitted, one without grouping and one with grouping.
-The results are visialized on a plot.
-The analysis was performed using the R programming language and additional packages.
-The source code for this paper can be found in the public repository[^5].
+**Methods.** The primary analytical technique employed in this study is linear regression.
+The independent variable in the regression is the number of lawyers (employees at law firms) in a city, which is denoted as *n*.
+In theory, either the number of law firms or their employees may be used as the independent variable, but law firms in different cities may have different sizes, and thus the number of employees seems to be a more stable and consistent metric.
+The value for each city is calculated as the sum of the number of employees of each law firm incorporated in that city.
+If a law firm reports a zero count of employees, it is assumed that the number of employees is equal to one.
+The dependent variable in the regression is the specific provision of legal services which is denoted as *p* and calculated by dividing the number of employees at law firms in the city by the city's population and multiplying the result by 10,000.
+In other words, it is the number of lawyers per 10,000 of population.
+This formula represents a modified version of the legal market saturation metric used in the analysis of the Russian legal services market conducted by the Institute for the Rule of Law [@Moiseeva2016].
 
-[^5]: https://github.com/PavelSyomin/ru-smb-companies-papers/tree/main/legal-companies-concentration-and-availability
+Cities were classified according to their population size.
+The population-based categorization is conducted in accordance with the classification system proposed by @Lappo1997.
+This is done to ascertain whether the hypothesized relationship is dependent on the city size.
+Linear models were constructed for the entire dataset, without accounting for the groups, or within each group separately.
+If necessary, the variables were log10-transformed.
+The results are then presented in graphical form on the plots.
+The analysis was conducted in the R programming language in conjunction with additional packages.
+The source code for this paper can be accessed via the public repository: https://github.com/PavelSyomin/ru-smb-companies-papers/tree/main/legal-companies-concentration-and-availability.
 
-**Theoretical model.** Despite the perceived simplicity of the modelling described earlier, it has a theoretical mathematical background, and several possible types of outcome may be laid out right now before the analysis of the data in hand.
-It is known that a linear model is described by the two coefficients, intercept (a free term of the linear equation showing the relative shift of the regression line along the y axis) and slope (a coefficient before the independent variable showing the angle between the regression line and x axis).
+**Theoretical model.** Although the modelling described earlier appears straightforward, it is based on a theoretical mathematical framework, and several potential outcomes can be identified before analyzing the data.
+It is known that a simple linear model is described by the two coefficients: the intercept, which is a term of the linear equation showing the relative shift of the regression line along the y-axis, and the slope, which is a coefficient of the independent variable showing the angle between the regression line and the x-axis.
 The slope can be either positive, negative, or zero.
-As a special edge case, it can be also infinite (i.
-e.
-regression line is vertical), even though, strictly speaking, such a dependency is not a function.
-Thus, the linear fit described earlier can yield a result of one of four types, each corresponding to one of the slope options and, subsequently, the pattern of relationship between the number of law firms and provision of legal services.
-These basic theoretical types are shown in Figure 1.
+As a special edge case, it can be also infinite (that is, the regression line is vertical), although, strictly speaking, such a dependency is not a mathematical function.
+Consequently, the linear fit previously described can yield a result of one of four types, each corresponding to one of the slope options and, subsequently, the pattern of relationship between the number of lawyers and the provision of legal services.
+These types are illustrated in Figure 1.
 
-```{r figure-1-theoretical-models, fig.cap="**Figure 1** — Four plots showing four types of the theoretical relationship between the number of law firms employees and provision of lwgal services in cities", fig.height=3}
+```{r figure-1-theoretical-models, fig.cap="**Figure 1** — Four plots showing four types of the theoretical relationship between the number of lawyers and the provision of legal services in cities", fig.height=3}
 theoretical_models_plot
@@
 
-Type 1 corresponds to the zero slope.
-In this case, the regression line is horizontal.
-From the substantial point of view, it describes the absolute equality in the provision of legal services among cities, that is, the provision is independent of the number of law firms employees.
-Despite being quite unlikely, this type also may be described as optimal, because it guarantees equitable access to legal services (at least in spatial terms).
-Probably this type requires some sort of governmental intervention to form (e.
-g.
-encouraging the establishment of law firms in small cities).
+In the context of regression analysis, Type 1 corresponds to the zero slope.
+In this instance, the regression line is horizontal.
+From the standpoint of legal services provision, it describes the absolute equality in the delivery of legal services across cities and its independence of the number of lawyers.
+This type can be also described as optimal, as it ensures equitable access to legal services (at least in spatial terms).
+Type 4, similarly to Type 1, also indicates the independence between the number of lawyers and the provision of legal services.
+However, in this case the number of employees at law firms is fixed, rather than their proportion in the population.
 
-Type 2 is characterized by the positive slope and, consequently, positive linear relationship between the number of law firms and the provision of legal services.
-It seems that at the free market this type is the most probable, even though the specific value of the slope may vary.
-In such case, the bigger the city, the easier it is to find a lawyer, and vise versa.
-
-Type 3 is the reverse of Type 2: the slope of the regression line is negative, and so is the relationship between the number of law firms and the provision of legal services.
-It is also quite improbable.
-One possible situation in which this type may form is the unexpected and rapid migration of people from the smallest cities to the largest with almost uniform (Type 1) relationship as a prerequisite.
-Under such circumstances, the relative number of law firms in small cities increases fast, but the relative number of law firms in large cities can not grow immediately as fast as the population, and the largest cities start to suffer from the lack of legal services, while the smallest cities are oversaturated with law firms.
-
-Type 4 is purely theoretical, but important in conjunction with the three other types.
-As Type 1, it also signals about the independence between the number of law firms and provision of legal services, but here the number of law firms employees is fixed rather than their proportion in the population.
-Again, it is very unlikely to happen in the real world, but it also suggests that some sort of governmental intervention (e.
-g.
-strictly controlling the number of law firms) is taking place.
+Type 2 is defined by the positive slope and, consequently, positive linear relationship between the number of law firms and the provision of legal services.
+It appears that in a free market, this type is the most probable, although the specific value of the slope may vary.
+In such a scenario, the greater the number of lawyers, the easier it is to find a lawyer, and vice versa.
+In contrast to Type 2, Type 3 exhibits a negative regression slope and a negative relationship between the number of lawyers and the provision of legal services.
+It is possible that this type may emerge when there is an unexpected and rapid migration of people from the smallest cities to the largest with a nearly uniform (Type 1) distribution of lawyers as a prerequisite.
+In such circumstances, the specific provision of legal services in the smallest cities increases at a high rate due to the population loss.
+However, in the largest cities the number of lawyers cannot grow as fast as the population, leading to a decline in the specific provision of legal services.
+As a result, the slope of the regression line becomes negative.
 
@@
 
-Figure 2 shows the relationship between the concentration of legal companies and the availability of legal services, as measured by the metrics proposed in the Materials and methods section.
-The overall connection between the number of law firms and provision of legal services appears to be nonlinear and corresponds to a square root formula.
-However, the same relationship within each city size group is fitted well with a linear models.
-The coefficients of these models are different, and the slope decreases when the city population increases.
-While the regression lines for the small and medium cities are close to Type 4 (provision is independent of the number of lawyers because of the small variation in the number of lawers), the modelled relationship in millionaire cities resembles Type 1 (another option for independence, but due to relatively uniform and equal provision regardless the number of lawyers).
-The cities from big to extra-large are characterized by the positive relationship between the number of lawyers and the provision of legal services and can be classified as Type 2.
+Figure 2 illustrates the relationship between the number of lawyers and the provision of legal services.
+The overall connection between the number of law firms and the provision of legal services appears to be nonlinear, with a square root function providing an accurate fit.
+However, the same relationship within each population size group is well-fitted by a linear model.
+The coefficients of these models differ, and the slope declines as the city population increases.
+While the regression lines for the small and medium cities are close to Type 4 (provision is independent of the number of lawyers due to the small variation in the number of lawyers), the modeled relationship in millionaire cities resembles Type 1 (another option for independence, but due to relatively uniform and equal provision regardless of the number of lawyers).
+The cities from big to extra-large are characterized by a positive correlation between the number of lawyers and the provision of legal services, which can be classified as Type 2.
 
@@
 
-```{r fig-2-empl-provision, fig.cap="**Figure 2** — The relationship between the number of lawyers and provision of legal services", fig.height=5}
+```{r fig-2-empl-provision, fig.cap="**Figure 2** — The relationship between the number of lawyers and the provision of legal services in cities", fig.height=5}
 ep_plot
@@
 
-In general, regional centers have relatively high number of lawyers and considerable level of legal services provision.
-This observation is probably explained by the centralization tendencies in the regions and the concentration of business and services in their administrative centers that situate the majority of government bodies and are also usually the most populous cities within a region.
+In general, regional centers have a relatively high number of lawyers and a considerable level of legal services provision.
+This observation is likely attributable to the phenomenon of regional centralization: economic activity and commercial services are concentrated in the administrative centers of regions, which host the governmental bodies and are often the most populous cities within a region.
+Moscow and Saint Petersburg, although excluded from the data analysis and shown separately on a subplot, are two notable outliers with an extremely high number of lawyers and a high level of legal services provision.
+Moscow accounts for approximately one-third of all employees at law firms, while Saint Petersburg accounts for approximately 10%.
+The position of the corresponding data points relative to the main regression lines suggests that there is a saturation point, at which further increase in the number of lawyers has little effect on the provision of legal services.
 
-Moscow and Saint Petersburg, though excluded from the data analysis and not shown on the main figure, are two remarkable outliers with extremely high concentration of legal services and a good level of legal services availability.
-Moscow alone accounts for almost one-third of all legal companies' employees, and Saint Petersburg accounts for about 10%.
-The position of the corresponding data points relative to the main regression lines suggests that there is an “availability saturation” level, where further concentration has little effect on availability.
+The general tendencies observed in Figure 2 should be supplemented by an analysis of the cities distribution shown in Figure 3, which extends Figure 2 and displays the binned count of cities according to the number of lawyers and the provision of legal services.
+The overwhelming majority of cities exhibit a low level of both legal services provision and law firms development.
+A mere 18 cities are distinguished by a high number of lawyers, exceeding 500.
+It is noteworthy that a high number of lawyers is favorable to the legal services provision, as all of these cities are at least at the medium level of provision, and none of them has less than 5 lawyers per 10,000 people.
+However, cities with a relatively low number of lawyers can reach top performance in legal services provision that is on par with the cities with the highest number of lawyers.
+Moreover, it appears that such small yet well-served cities are even more prevalent than cities with a high level of both lawyers and provision.
+The relative location of bin counts in the Figure 2 suggests that, in addition to the dominant bottom left corner, there are two main clusters of cities representing two distinct approaches to achieving a high level of legal services provision.
+One approach is to increase both the absolute number of lawyers and the relative number of lawyers per population, which is characteristic of the largest cities.
+A second approach is to have a relatively small yet sufficient number of lawyers, which may be typical of small and medium-sized cities.
 
-The general tendencies observed in the Figure 2 should be supplemented by the analysis of the source data distribution shown in Figure 3 that extends the Figure 2 and displays the binned count of cities according to the number of lawyers and the provision of legal services.
-The overwhelming majority of cities is characterized by low level of both legal services provision and law firms development.
-Only a limited number (18) of cities is characterized by the high (more than 500) number of lawyers.
-It should be noted that high number of lawyers is good for legal services provision because all these cities are at least at the medium level of provision, and none of them has less than 5 lawyers per 10,000 people.
-However, cities with relatively low number of laywers can nevertheless reach top performance from the point of legal services provision that is on par with the cities with the highest number of lawyers.
-Moreover, it appears that such small yet well-served cities are even more typical than cities with high level of both lawyers and provision.
-The relative location of bin counts in the Figure 2 implies that there — besides the dominant bottom left corner — there are two main clusters of cities showing two different approaches of reaching high level of legal services provision.
-One approach is increase both in absolute number of law firms and relative number of lawyers per population probably charasteristic of the largest cities.
-Another is a relatively small yet sufficient number of lawyers typical probably for small or medium cities.
-Thus, apart from the general tendencies, the two-ways structure is found out.
-
-```{r figure-3-empl-provision-binned, fig.cap="**Figure 3** — Count and proportion of cities with variuos combinations of number of lawyers and level of legal services provision"}
+```{r figure-3-empl-provision-binned, fig.cap="**Figure 3** — Count and proportion of cities with variuos combinations of number of lawyers and the level of legal services provision"}
 ep_binned_plot
@@
 
-Figure 4 is a map displaying the location of cities with relatively high (more than 5 lawyers per 10,000 people) provision of legal services.
+Figure 4 presents a map of cities with relatively high (more than 5 lawyers per 10,000 people) provision of legal services.
 The most prominent cluster is formed by Moscow and its satellite cities.
-Other notable centers with high availability include Krasnodar and Stavropol in the South of Russia, Yekaterinburg and Chelyabinsk in the Urals, and Tuymen and Novosibirsk in Western Siberia.
-There are several regions with no cities of good legal services provision (especially on Far East, North, and Caucasus), and the majority of cities with relatively high level of legal services provision is located in the Central and Volga economic regions of the country.
+Other notable areas with high provision include Krasnodar and Stavropol in the south of Russia, Yekaterinburg and Chelyabinsk in the Urals, and Tuymen and Novosibirsk in Western Siberia.
+There are several regions with no cities that provide a high level of legal services provision (in particular, the Far East, North, and Caucasus).
+Conversely, the majority of cities with a relatively high level of legal services provision are located in the Central and Volga economic regions of the country.
 
@@
 
-```{r fig-4-map, fig.cap="**Figure 4** — Legal services provision in Russian cities on the country's map. Only cities with more than 5 lawyers per 10,000 people are shown", fig.height=4}
+```{r fig-4-map, fig.cap="**Figure 4** — Map of legal services provision in Russian cities. Only cities with more than 5 lawyers per 10,000 people are shown", fig.height=4}
 map
@@
 The provision of legal services in cities has a positive relationship with the number of lawyers.
-The overall trend can be fitted with a square root curve, while the relationship in distinct city size groups is close to linear with varying slope of the regression line.
-In small and medium cities (with population below 50,000), the increase of number of lawyers has little effect on the level of legal services provision, because the number of lawyers per 10,000 people varies greatly.
+The overall trend can be approximated by a square root curve, while the relationship in distinct population size groups is close to linear with varying slope of the regression line.
+In small cities the increase in the number of lawyers has a minimal impact on the level of legal services provision, given the considerable variation of the dependent variable.
 In larger cities, the relationship is linear and positive.
-In the largest cities having the population above 1 million people, the impact of the growth of the number of lawyers on the provision of legal services decreases, because the latter in these cities is relatively high and uniform.
-Regional centers typically have relatively high availability of legal services.
-Moscow and Saint Petersburg are characterized by the highest number of law firms and high provision of legal services.
-Empirical data suggests that there may be a saturation limit beyond which the increase in the number of law firms does not have a substantial impact on provision of legal servicies.
+In the cities with a population above one million, the impact of the growth of the number of lawyers on the provision of legal services is less pronounced, as the latter is already high and uniform in these cities.
+Regional centers typically have a relatively high availability of legal services.
+Moscow and Saint Petersburg are distinguished by the highest number of lawyers and the high provision of legal services.
+Empirical data indicates that there may be a saturation limit beyond which the increase in the number of employees at law firms does not have a substantial impact on the provision of legal services
 
-The overwhelming majority (about 85%) of cities is characterized both by low number of law firms and low level of legal services provision.
-There are two groups of cities by their way of achieving high levels of legal services provision.
-The first group is composed of cities with large number of law firms.
-The second group, on the opposite, have small yet number of law firms that nevertheless satisfy the needs of local communitues.
-Moscow and its region is characterized by relatively high level of legal services availability, as well as Central and Volga regions of Russia in general.
-Similarly, Saint Petersburg, Krasnodar, Stavropol, Yekaterinburg, Chelyabinsk, Tyumen, and Novosibirsk also have good provision to legal services.
-On the contrary, in many regions of Far East, North and Caucasus the level of provision of legal services is low.
+The majority of cities (approximately 85%) are characterized by a paucity of lawyers and a correspondingly limited level of legal services provision.
+Two distinct groups of cities can be identified based on their approach to achieving high level of legal services provision.
+The first group is comprised of cities with large number of lawyers.
+The second group, in contrast, has small number of lawyers that nevertheless meet the needs of local communities.
+Moscow and its surrounding region exhibit a relatively high level of legal services provision, as do the Central and Volga economic regions of Russia.
+Similarly, Saint Petersburg, Krasnodar, Stavropol, Yekaterinburg, Chelyabinsk, Tyumen, and Novosibirsk also have a satisfactory provision of legal services.
+Conversely, in numerous regions of the Far East, North and Caucasus the provision of legal services is notably limited.
 

### 2024-05-18 legal-companies-concentration-and-availability/journal-article/Script.R

@@ -186,14 +186,16 @@ ep_binned_plot
 # Make a map with provision > 5
 ru_crs <- st_crs("+proj=aea +lat_0=0 +lon_0=100 +lat_1=68 +lat_2=44 +x_0=0 +y_0=0 +ellps=krass +towgs84=28,-130,-95,0,0,0,0 +units=m +no_defs")
 map <- data %>% 
-  filter(!(city %in% outliers), provision >= 5) %>% 
+  filter(
+    !(city %in% outliers) | city %in% c("Москва", "Санкт-Петербург"),
+    provision >= 5) %>% 
   st_as_sf(coords = c("lon", "lat"), crs = "EPSG:4326") %>% 
   ggplot() +
   geom_sf(data = regions_geo, color = "gray50", fill = "white") +
   geom_sf(aes(color = provision), shape = 19) +
   coord_sf(crs = ru_crs) +
   scale_color_binned(
-    name = "Lawers per 10,000 people",
+    name = "Lawyers per 10,000 people",
     n.breaks = 4,
     low = "#74c476",
     high = "#00441b") +
@@ -252,7 +254,7 @@ theoretical_models_plot <- ggplot(lines, aes(x = x, y = y)) +
     limits = c(-.1, 1.1)
   ) +
   scale_y_continuous(
-    name = "Laywers per 100,000 (p)",
+    name = "Lawyers per 100,000 (p)",
     labels = function (br) br * 10,
     limits = c(-.1, 1.1)
   ) +
@@ -262,4 +264,6 @@ theoretical_models_plot <- ggplot(lines, aes(x = x, y = y)) +
   labs(caption = "s means slope, dots are hypothetical data points,\nline is hypothetical linear fit of points")
 theoretical_models_plot
 
-
+fit1 <- lm(provision ~ empl, filter(data, size == "Millionaire (>1M)"))
+summary(fit1)
+data$size

### 2024-05-20 legal-companies-concentration-and-availability/journal-article/Article.Rmd
Improve article: add table with regression details, rewrite abstract, add main cities labels and economic regions on the map, other minor changes, #2

@@
 library(here)
+library(knitr)
 library(readr)
@@
 
-**Abstract:** The article examines the relationship between the concentration of legal companies and the availability of legal services.
-To conduct the analysis, a dataset was created using open data from the Federal Tax Service of Russia.
-The dataset includes information on the location, revenue, expenditure, and number of employees of small and medium-sized legal companies.
-The concentration is calculated by dividing the number of employees of legal companies in the city by the total number of employees of all legal companies in Russia.
-Availability is measured as the number of employees of legal companies in the city per 100,000 population.
-Data visualization and linear regression were utilized to estimate the hypothesized relationship.
-The study showed a positive linear relationship between concentration and availability.
-However, there are multiple approaches to achieving good availability of legal services.
+**Abstract:** The article examines the relationship between the number of employees at law firms and the provision of legal services in Russian cities.
+It aims to identify whether the increase in the number of lawyers leads an enhanced level of legal services provision.
+In addition, it attempts to discern geographical patterns in the provision of legal services in Russia.
+To achieve these objectives, open data from the Federal Tax Service of Russia is utilized as the primary source of information.
+A linear regression is employed to estimate the hypothesized relationship, and data plotting and mapping are used to visualize it.
+The study demonstrates an overall positive square root relationship between the number of lawyers and the provision of legal services in cities, as well as a linear relationship between these variables in distinct city population size groups.
+It supports a widely held proposition about the pivotal role of the largest cities in the provision of services and also offers partial confirmation of the commonly accepted view about the positive impact of services concentration on their delivery.
+However, the study argues that there are multiple approaches to achieving high level of legal services provision, with an increase in the number of lawyers being only one of potential options.
+The findings of this research may serve to inform data-driven policymaking decisions of governmental bodies.
 
@@
 
-**Аннотация:**
+**Аннотация:** Мақалада заңгерлік фирмалардың (заңгерлер) қызметкерлерінің саны мен Ресей қалаларында заңгерлік қызмет көрсету арасындағы байланыс қарастырылады.
+Зерттеудің мақсаты – адвокаттар санының артуы заң қызметтерін көрсету деңгейінің артуына әкелетінін анықтау.
+Сонымен қатар, Ресейде заң қызметтерін көрсетудің географиялық заңдылықтарын анықтау әрекеті жасалуда.
+Ақпараттың негізгі көзі ретінде Ресейдің Федералдық салық қызметінің ашық деректері пайдаланылады.
+Сызықтық регрессия болжамды қатынасты бағалау үшін қолданылады, ал графикалық және карталау бақыланатын заңдылықтарды визуализациялау үшін қолданылады.
+Зерттеу қалалардағы заңгерлер саны мен заң қызметтерін көрсету арасындағы оң қисық сызықты (квадрат түбір функциясына жақын) қатынасты, сондай-ақ қалаларды халық саны бойынша топтастыру кезінде осы айнымалылар арасындағы сызықтық байланысты көрсетеді.
+Бұл қызмет көрсетудегі ірі қалалардың шешуші рөлін кеңінен бекітумен сәйкес келеді, сонымен қатар қызметтердің шоғырлануы қызмет көрсетуге оң әсер етеді деген ойды ішінара растайды.
+Дегенмен, зерттеу заң қызметтерін көрсетудің жоғары деңгейіне қол жеткізудің әртүрлі тәсілдері бар екенін көрсетеді, адвокаттар санын көбейту тек бір ғана ықтимал нұсқа болып табылады.
+Зерттеу нәтижелері уәкілетті органдардың деректерге негізделген шешімдер қабылдауына негіз бола алады.
 
@@
 
-**Аннотация:**
+**Аннотация:** В статье исследуется связь между числом сотрудников юридических фирм (юристов) и обеспеченностью юридическими услугами в городах России.
+Цель исследования — определить, приводит ли увеличение числа юристов к повышению уровня обеспеченности юридическими услугами.
+Кроме того, предпринимается попытка выявить географические закономерности обеспеченности юридическими услугами в России.
+В качестве основного источника информации используются открытые данные ФНС России.
+Для оценки предполагаемой взаимосвязи применяется линейная регрессия, а для визуализации обнаруженных закономерностей используются построение графиков и картографирование.
+Исследование показывает положительную криволинейную (близную к функции квадратного корня) связь между количеством юристов и обеспеченностью юридическими услугами в городах, а также линейную связь между этими переменными при группировке городов по численности населения.
+Оно согласуется с широко распространённым утверждением о ключевой роли крупнейших городов в предоставлении услуг, а также частично подтверждает представление о положительном влиянии концентрации услуг на обеспеченность ими.
+Тем не менее, результаты исследования говорят о том, что существуют разные подходы к достижению высокого уровня обеспеченности юридическими услугами, причём увеличение количества юристов является лишь одним из потенциальных вариантов.
+Результаты исследования могут послужить основой для принятия решений органами власти на основе данных.
 
@@
 **Data.** This research uses a geocoded dataset on small and medium-sized law firms incorporated in Russia.
-The dataset is derived from FTS open data dumps (archives) of the state SMB Registry[@SMBRegistry2024] and FTS open data on the average number of employees of organizations[@FTSEmployees2024], which were processed using the Python command-line tool *ru-smb-companies* developed by the author of the paper.
+The dataset is derived from FTS open data dumps (archives) of the state SMB Registry [@SMBRegistry2024] and FTS open data on the average number of employees of organizations [@FTSEmployees2024], which were processed using the Python command-line tool *ru-smb-companies* developed by the author of the paper.
 The dataset contains information about distinct companies, including their names, taxpayer IDs, regions and settlements of incorporation, and main activity codes.
@@
 
-Figure 2 illustrates the relationship between the number of lawyers and the provision of legal services.
+Figure 2 depicts the relationship between the number of lawyers and the provision of legal services.
 The overall connection between the number of law firms and the provision of legal services appears to be nonlinear, with a square root function providing an accurate fit.
-However, the same relationship within each population size group is well-fitted by a linear model.
-The coefficients of these models differ, and the slope declines as the city population increases.
-While the regression lines for the small and medium cities are close to Type 4 (provision is independent of the number of lawyers due to the small variation in the number of lawyers), the modeled relationship in millionaire cities resembles Type 1 (another option for independence, but due to relatively uniform and equal provision regardless of the number of lawyers).
-The cities from big to extra-large are characterized by a positive correlation between the number of lawyers and the provision of legal services, which can be classified as Type 2.
+However, the same relationship within each population size group is well-fitted by a linear model in most of the cases.
+As illustrated in Table 1, the coefficients of these models differ, and the slope declines as the city population increases.
+The regression line for small cities is close to Type 4, and the small coefficient of determination (R² = 0.11) indicates that the provision is almost independent of the number of lawyers, likely due to the low variation of the latter.
+The modeled relationship in millionaire cities resembles Type 1, which is another option for independence caused by relatively uniform and equal provision regardless of the number of lawyers.
+The coefficient of determination here is also relatively low (R² = 0.51), although greater than in the group of small cities.
+Cities of medium to extra-large size are characterized by a positive correlation between the number of lawyers and the provision of legal services, which can be classified as Type 2.
+The coefficient of determination is high (R² is approximately 0.9), indicating that nearly all the variance in the provision of legal services can be explained by the number of lawyers.
 
@@
 ```
+
+<br>
+
+```{r}
+kable(
+  lm_res_for_table,
+  col.names = c(
+    "City size", "Intercept", "Slope", "R²", "# of cities"
+  ),
+  caption = "**Table 1** — Summary of linear models fitted for each city population size group.
+All models and coefficients are statistically significant (p < 0.05), except for intercepts in small and medium cities",
+  digits = c(0, 2, 4, 2, 0)
+)
+```
+
+<br>
 
@@
 The most prominent cluster is formed by Moscow and its satellite cities.
-Other notable areas with high provision include Krasnodar and Stavropol in the south of Russia, Yekaterinburg and Chelyabinsk in the Urals, and Tuymen and Novosibirsk in Western Siberia.
+Other notable cities with high provision include Krasnodar and Stavropol in the south of Russia, Yekaterinburg and Chelyabinsk in the Urals, and Tyumen and Novosibirsk in Western Siberia, as well as Irkutsk in Eastern Siberia.
 There are several regions with no cities that provide a high level of legal services provision (in particular, the Far East, North, and Caucasus).
-Conversely, the majority of cities with a relatively high level of legal services provision are located in the Central and Volga economic regions of the country.
+Conversely, the majority of cities with a relatively high level of legal services provision are located in five economic regions of the country (as categorized by @Leyzerovich2010): Central, Central-Chernozem, Volga-Vyatka, Volga, Ural.
 

### 2024-05-20 legal-companies-concentration-and-availability/journal-article/Script.R

@@ -1,9 +1,11 @@
+library(broom)
 library(dplyr)
 library(ggplot2)
 library(glue)
 library(here)
 library(readr)
 library(sf)
+library(tibble)
 library(tidyr)
 
 # Load data
@@ -12,8 +14,9 @@ cities <- rbind(
   read_csv(here("common", "cities.csv")),
   read_csv(here("common", "cities_additional.csv"))
 )
-regions_geo <- st_read(here("common", "ru.geojson"))
-tiles <- read_csv(here("journal-article", "russia-tiles.csv"))
+economic_regions <- read_csv(here("journal-article", "economic-regions.csv"))
+regions_geo <- st_read(here("common", "ru.geojson")) %>% 
+  left_join(economic_regions, by = c("shapeISO" = "iso_code"))
 
 # Build table with number of employees by city
 # lfc means Law Firms by City
@@ -189,18 +192,40 @@ map <- data %>%
   filter(
     !(city %in% outliers) | city %in% c("Москва", "Санкт-Петербург"),
     provision >= 5) %>% 
+  mutate(city_label = case_when(
+    city == "Москва" ~ "Moscow",
+    city == "Санкт-Петербург" ~ "Saint Petersburg",
+    city == "Краснодар" ~ "Krasnodar",
+    city == "Екатеринбург" ~ "Yekaterinburg",
+    city == "Новосибирск" ~ "Novosibirsk",
+    city == "Иркутск" ~ "Irkutsk",
+    TRUE ~ NA_character_
+  )) %>% 
   st_as_sf(coords = c("lon", "lat"), crs = "EPSG:4326") %>% 
   ggplot() +
-  geom_sf(data = regions_geo, color = "gray50", fill = "white") +
-  geom_sf(aes(color = provision), shape = 19) +
+  geom_sf(aes(fill = economic_region), data = regions_geo, color = "gray50") +
+  geom_sf(aes(size = provision), shape = 21) +
+  geom_sf_label(
+    aes(label = city_label),
+    hjust = "inward",
+    vjust = "inward",
+    label.size = 0,
+    label.padding = unit(0, "mm"),
+    label.r = unit(0, "mm"),
+    na.rm = TRUE) + 
   coord_sf(crs = ru_crs) +
-  scale_color_binned(
+  scale_size_binned(
     name = "Lawyers per 10,000 people",
-    n.breaks = 4,
-    low = "#74c476",
-    high = "#00441b") +
+    range = c(1, 3),
+    n.breaks = 4) +
[…]

### 2024-05-20 legal-services-provision/journal-article/Article.Rmd
Rename folder, #2

--- 2024-05-20 legal-companies-concentration-and-availability/journal-article/Article.Rmd
@@
 
+<br>
+
 ```{r figure-1-theoretical-models, fig.cap="**Figure 1** — Four plots showing four types of the theoretical relationship between the number of lawyers and the provision of legal services in cities", fig.height=3}
@@
 ```
+
+<br>
 
@@
 
+<br>
+
 ```{r figure-3-empl-provision-binned, fig.cap="**Figure 3** — Count and proportion of cities with variuos combinations of number of lawyers and the level of legal services provision"}
@@
 There are several regions with no cities that provide a high level of legal services provision (in particular, the Far East, North, and Caucasus).
-Conversely, the majority of cities with a relatively high level of legal services provision are located in five economic regions of the country (as categorized by @Leyzerovich2010): Central, Central-Chernozem, Volga-Vyatka, Volga, Ural.
+Conversely, the majority of cities with a relatively high level of legal services provision are located in five economic regions of the country, as categorized by @Leyzerovich2010: Central, Central-Chernozem, Volga-Vyatka, Volga, Ural.
 
@@
 
+<br>
+
 # Conclusion
@@
 The overall trend can be approximated by a square root curve, while the relationship in distinct population size groups is close to linear with varying slope of the regression line.
-In small cities the increase in the number of lawyers has a minimal impact on the level of legal services provision, given the considerable variation of the dependent variable.
+In small cities the increase in the number of lawyers has a minimal impact on the level of legal services provision.
 In larger cities, the relationship is linear and positive.
@@
 The second group, in contrast, has small number of lawyers that nevertheless meet the needs of local communities.
-Moscow and its surrounding region exhibit a relatively high level of legal services provision, as do the Central and Volga economic regions of Russia.
-Similarly, Saint Petersburg, Krasnodar, Stavropol, Yekaterinburg, Chelyabinsk, Tyumen, and Novosibirsk also have a satisfactory provision of legal services.
+Moscow and its surrounding region exhibit a relatively high level of legal services provision.
+Similarly, Saint Petersburg, Krasnodar, Stavropol, Yekaterinburg, Chelyabinsk, Tyumen, Novosibirsk, and Irkutsk also have a satisfactory provision of legal services.
+Most of the cities with high level of legal services provision are situated within five economic regions of the country: Central, Central-Chernozem, Volga-Vyatka, Volga, and Ural.
 Conversely, in numerous regions of the Far East, North and Caucasus the provision of legal services is notably limited.

### 2024-05-21 legal-services-provision/journal-article/Article.Rmd
Final edit before submission, #2

@@
 
-**Abstract:** The article examines the relationship between the number of employees at law firms and the provision of legal services in Russian cities.
-It aims to identify whether the increase in the number of lawyers leads an enhanced level of legal services provision.
+**Abstract:** The article examines the relationship between the number of employees at law firms, or lawyers, and the provision of legal services in Russian cities.
+It aims to identify whether the increase in the number of lawyers leads to an enhanced level of legal services provision.
 In addition, it attempts to discern geographical patterns in the provision of legal services in Russia.
 To achieve these objectives, open data from the Federal Tax Service of Russia is utilized as the primary source of information.
-A linear regression is employed to estimate the hypothesized relationship, and data plotting and mapping are used to visualize it.
+Linear regression is employed to estimate the hypothesized relationship, and data plotting and mapping are used to visualize it.
 The study demonstrates an overall positive square root relationship between the number of lawyers and the provision of legal services in cities, as well as a linear relationship between these variables in distinct city population size groups.
-It supports a widely held proposition about the pivotal role of the largest cities in the provision of services and also offers partial confirmation of the commonly accepted view about the positive impact of services concentration on their delivery.
-However, the study argues that there are multiple approaches to achieving high level of legal services provision, with an increase in the number of lawyers being only one of potential options.
+It supports a widely held proposition about the pivotal role of the largest cities in the provision of services and also offers partial confirmation of the commonly accepted view about the positive impact of service concentration on their delivery.
+However, the study argues that there are multiple approaches to achieving a high level of legal services provision, with an increase in the number of lawyers being only one of the potential options.
 The findings of this research may serve to inform data-driven policymaking decisions of governmental bodies.
@@
 For instance, a study by @Wojcik2020 offers an economic-based overview of FihTech geography.
-The social approach, while closely related to the economic one, focuses on the services as the foundation of people's well-being and studies the provision, availability and accessibility of services, as well as their diversity and quality.
+The social approach, while closely related to the economic one, focuses on the services as the foundation of people's well-being and studies the provision, availability, and accessibility of services, as well as their diversity and quality.
 For example, the study in Indonesia [@Pratiwi2021] employs a variety of socio-economic data from the national health insurance system to discuss, among other issues, the equity problems that arise from geographical differences.
-A problem of access to various services has been a significant concern for decades [@Blacksell1990; @Keleher1996], as it is directly related to the quality of life and the guarantee of human rights.
+The problem of access to various services has been a significant concern for decades [@Blacksell1990; @Keleher1996], as it is directly related to the quality of life and the guarantee of human rights.
 Furthermore, research has been conducted on the relationship between geographical properties and service availability [@Holzer2000; @Johnson2006].
@@
 The concentration of businesses provides economic benefits for the companies and, as such, the concentration is likely to become self-sustaining.
-It is reasonable to assume that such concentration does not only leads to the economic growth, but also positively impacts the service provision, thereby being productive in both economic and social dimensions.
-It is a common observation that the largest cities and regional centers offer the greater variety and quantity of diverse service providers than smaller settlements and rural areas, where the availability of services is limited due to the low number of their providers.
+It is reasonable to assume that such concentration not only leads to economic growth but also positively impacts the service provision, thereby being productive in both economic and social dimensions.
+It is a common observation that the largest cities and regional centers offer a greater variety and quantity of diverse services than smaller settlements and rural areas, where the availability of services is limited due to the low number of providers.
 Nevertheless, it must be asked: is this relationship truly valid?
-The common-sense notion that the number of service providers is correlated with the provision of services may, in fact, be incorrect.
+The common-sense notion that the number of service providers is correlated with the provision of services may be incorrect.
 Thus, rigorous data-driven testing of this hypothesis is necessary.
 
-This paper empirically examines the relationship between the number of law firms in Russian cities and the provision of legal services.
+This paper empirically examines the relationship between the number of employees at law firms (designated here for simplicity as lawyers) in Russian cities and the provision of legal services.
 The existing literature on the availability of services typically focuses on medical services [@Joseph1982; @vanDoorslaer2006; @Weiss2020; @Maleki2024], with relatively little attention paid to legal services.
@@
 The primary research question is how the specific provision of legal services is influenced by the number of lawyers (employees at law firms).
-Additionally, the study aims to identify geographical patterns in the spatial distribution of legal services provision.
+Additionally, the study aims to identify geographical patterns of the legal services provision.
 
@@
 
-Law firms were selected based on a main activity code equal of 69.10.
+Law firms were selected based on a main activity code equal to 69.10.
 This selection technique was derived from the analytical paper on the Russian legal services market prepared by the members of the Institute for the Rule of Law [@Moiseeva2016].
 Furthermore, the dataset was subjected to the following filters: the year was set to 2021, the company was identified as an organization rather than a sole entrepreneur, the company had declared non-zero revenue and expenditure for the target year (implying that the company was active), and the company's address of registration was in a city rather than in a rural area.
-In addition to the main dataset, a data on city population was utilized from a freely distributed commercial source [@HFLabs2021].
+In addition to the main dataset, data on city populations was utilized from a freely distributed commercial source [@HFLabs2021].
 
@@
 In contrast, the next city had a considerably smaller number of employees, with less than 2,000.
-The remaining four cities (Innopolis, Kirovsk, Bronnytsy, Krasnoarmeisk) are distinguished by a markedly elevated proportion of law firms' employees in their respective populations.
+The remaining four cities (Innopolis, Kirovsk, Bronnytsy, and Krasnoarmeisk) are distinguished by a markedly elevated proportion of law firms' employees in their respective populations.
 This proportion ranges from 38 per 10,000 in Krasnoarmeisk to 417 per 10,000 in Innopolis, in stark contrast to less than 20 per 10,000 in all other cities.
-After applying all the filters and removing the outliers, the dataset contain information about 576 cities.
+After applying all the filters and removing the outliers, the dataset contains information about 576 cities.
 
@@
 The dependent variable in the regression is the specific provision of legal services which is denoted as *p* and calculated by dividing the number of employees at law firms in the city by the city's population and multiplying the result by 10,000.
-In other words, it is the number of lawyers per 10,000 of population.
+In other words, it is the number of lawyers per 10,000 people.
 This formula represents a modified version of the legal market saturation metric used in the analysis of the Russian legal services market conducted by the Institute for the Rule of Law [@Moiseeva2016].
@@
 Cities were classified according to their population size.
-The population-based categorization is conducted in accordance with the classification system proposed by @Lappo1997.
+The population-based categorization is conducted following the classification system proposed by @Lappo1997.
 This is done to ascertain whether the hypothesized relationship is dependent on the city size.
@@
 
-**Theoretical model.** Although the modelling described earlier appears straightforward, it is based on a theoretical mathematical framework, and several potential outcomes can be identified before analyzing the data.
+**Theoretical model.** Although the modeling described earlier appears straightforward, it is based on a theoretical mathematical framework, and several potential outcomes can be identified before analyzing the data.
 It is known that a simple linear model is described by the two coefficients: the intercept, which is a term of the linear equation showing the relative shift of the regression line along the y-axis, and the slope, which is a coefficient of the independent variable showing the angle between the regression line and the x-axis.
@@
 Type 4, similarly to Type 1, also indicates the independence between the number of lawyers and the provision of legal services.
-However, in this case the number of employees at law firms is fixed, rather than their proportion in the population.
+However, in this case, the number of employees at law firms is fixed, rather than their proportion in the population.
 
-Type 2 is defined by the positive slope and, consequently, positive linear relationship between the number of law firms and the provision of legal services.
+Type 2 is defined by the positive slope and, consequently, a positive linear relationship between the number of lawyers and the provision of legal services.
 It appears that in a free market, this type is the most probable, although the specific value of the slope may vary.
@@
 In contrast to Type 2, Type 3 exhibits a negative regression slope and a negative relationship between the number of lawyers and the provision of legal services.
-It is possible that this type may emerge when there is an unexpected and rapid migration of people from the smallest cities to the largest with a nearly uniform (Type 1) distribution of lawyers as a prerequisite.
+This type may emerge when there is an unexpected and rapid migration of people from the smallest cities to the largest with a nearly uniform (Type 1) distribution of legal services provision as a prerequisite.
 In such circumstances, the specific provision of legal services in the smallest cities increases at a high rate due to the population loss.
-However, in the largest cities the number of lawyers cannot grow as fast as the population, leading to a decline in the specific provision of legal services.
+However, in the largest cities, the number of lawyers cannot grow as fast as the population, leading to a decline in the specific provision of legal services.
 As a result, the slope of the regression line becomes negative.
@@
 Figure 2 depicts the relationship between the number of lawyers and the provision of legal services.
-The overall connection between the number of law firms and the provision of legal services appears to be nonlinear, with a square root function providing an accurate fit.
-However, the same relationship within each population size group is well-fitted by a linear model in most of the cases.
-As illustrated in Table 1, the coefficients of these models differ, and the slope declines as the city population increases.
-The regression line for small cities is close to Type 4, and the small coefficient of determination (R² = 0.11) indicates that the provision is almost independent of the number of lawyers, likely due to the low variation of the latter.
+The overall connection between these variables appears to be nonlinear, with a square root function providing a satisfactory fit.
+However, the same relationship within each population size group is well-fitted by a linear model in most cases.
+As illustrated in Table 1, the coefficients of these linear models differ, and the slope declines as the city population increases.
+The regression line for small cities is close to Type 4, and the small coefficient of determination (R² = 0.11) indicates that the provision is almost independent of the number of lawyers, likely due to the low variance of the latter.
 The modeled relationship in millionaire cities resembles Type 1, which is another option for independence caused by relatively uniform and equal provision regardless of the number of lawyers.
@@
 
-```{r fig-2-empl-provision, fig.cap="**Figure 2** — The relationship between the number of lawyers and the provision of legal services in cities", fig.height=5}
+```{r figure-2-empl-provision, fig.cap="**Figure 2** — The relationship between the number of lawyers and the provision of legal services in cities", fig.height=5}
 ep_plot
@@
 This observation is likely attributable to the phenomenon of regional centralization: economic activity and commercial services are concentrated in the administrative centers of regions, which host the governmental bodies and are often the most populous cities within a region.
-Moscow and Saint Petersburg, although excluded from the data analysis and shown separately on a subplot, are two notable outliers with an extremely high number of lawyers and a high level of legal services provision.
+Moscow and Saint Petersburg, although excluded from the data analysis and shown separately on a subplot (see bottom right frame in Figure 2), are two notable outliers with an extremely high number of lawyers and a high level of legal services provision.
 Moscow accounts for approximately one-third of all employees at law firms, while Saint Petersburg accounts for approximately 10%.
-The position of the corresponding data points relative to the main regression lines suggests that there is a saturation point, at which further increase in the number of lawyers has little effect on the provision of legal services.
+The position of the corresponding data points relative to the main regression lines suggests that there is a saturation point, at which a further increase in the number of lawyers has little effect on the provision of legal services.
 
 The general tendencies observed in Figure 2 should be supplemented by an analysis of the cities distribution shown in Figure 3, which extends Figure 2 and displays the binned count of cities according to the number of lawyers and the provision of legal services.
-The overwhelming majority of cities exhibit a low level of both legal services provision and law firms development.
+The overwhelming majority of cities exhibit a low level of both legal services provision and law firm development.
 A mere 18 cities are distinguished by a high number of lawyers, exceeding 500.
@@
 However, cities with a relatively low number of lawyers can reach top performance in legal services provision that is on par with the cities with the highest number of lawyers.
-Moreover, it appears that such small yet well-served cities are even more prevalent than cities with a high level of both lawyers and provision.
-The relative location of bin counts in the Figure 2 suggests that, in addition to the dominant bottom left corner, there are two main clusters of cities representing two distinct approaches to achieving a high level of legal services provision.
+Moreover, it appears that such small yet well-served cities are even more prevalent than cities with a high level of both law firm development and legal services provision.
+The relative location of bin counts in Figure 2 suggests that, in addition to the dominant bottom left corner, there are two main clusters of cities representing two distinct approaches to achieving a high level of legal services provision.
 One approach is to increase both the absolute number of lawyers and the relative number of lawyers per population, which is characteristic of the largest cities.
-A second approach is to have a relatively small yet sufficient number of lawyers, which may be typical of small and medium-sized cities.
+A second approach is to have a relatively small yet sufficient number of lawyers, which may be typical of small and medium cities.
 
@@
 
-```{r figure-3-empl-provision-binned, fig.cap="**Figure 3** — Count and proportion of cities with variuos combinations of number of lawyers and the level of legal services provision"}
+```{r figure-3-empl-provision-binned, fig.cap="**Figure 3** — Count and proportion of cities with various combinations of number of lawyers and the level of legal services provision"}
 ep_binned_plot
@@
 The provision of legal services in cities has a positive relationship with the number of lawyers.
-The overall trend can be approximated by a square root curve, while the relationship in distinct population size groups is close to linear with varying slope of the regression line.
-In small cities the increase in the number of lawyers has a minimal impact on the level of legal services provision.
+The overall trend can be approximated by a square root curve, while the relationship in distinct population size groups is close to linear with varying slopes of the regression line.
+In small cities, the increase in the number of lawyers has a minimal impact on the level of legal services provision.
 In larger cities, the relationship is linear and positive.
-In the cities with a population above one million, the impact of the growth of the number of lawyers on the provision of legal services is less pronounced, as the latter is already high and uniform in these cities.
-Regional centers typically have a relatively high availability of legal services.
+In cities with a population above one million, the impact of the growth of the number of lawyers on the provision of legal services is less pronounced, as the latter is already high and uniform in these cities.
+Regional centers typically have a relatively high provision of legal services.
 Moscow and Saint Petersburg are distinguished by the highest number of lawyers and the high provision of legal services.
@@
 The majority of cities (approximately 85%) are characterized by a paucity of lawyers and a correspondingly limited level of legal services provision.
-Two distinct groups of cities can be identified based on their approach to achieving high level of legal services provision.
-The first group is comprised of cities with large number of lawyers.
-The second group, in contrast, has small number of lawyers that nevertheless meet the needs of local communities.
+Two distinct groups of cities can be identified based on their approach to achieving a high level of legal services provision.
+The first group is comprised of cities with a large number of lawyers.
+The second group, in contrast, has a small number of lawyers that nevertheless meet the needs of local communities.
 Moscow and its surrounding region exhibit a relatively high level of legal services provision.
 Similarly, Saint Petersburg, Krasnodar, Stavropol, Yekaterinburg, Chelyabinsk, Tyumen, Novosibirsk, and Irkutsk also have a satisfactory provision of legal services.
-Most of the cities with high level of legal services provision are situated within five economic regions of the country: Central, Central-Chernozem, Volga-Vyatka, Volga, and Ural.
-Conversely, in numerous regions of the Far East, North and Caucasus the provision of legal services is notably limited.
+Most of the cities with high levels of legal services provision are situated within five economic regions of the country: Central, Central-Chernozem, Volga-Vyatka, Volga, and Ural.
+Conversely, in numerous regions of the Far East, North, and Caucasus the provision of legal services is notably limited.
 

### 2024-05-21 legal-services-provision/journal-article/Script.R

@@ -279,7 +279,7 @@ theoretical_models_plot <- ggplot(lines, aes(x = x, y = y)) +
     limits = c(-.1, 1.1)
   ) +
   scale_y_continuous(
-    name = "Lawyers per 100,000 (p)",
+    name = "Lawyers per 10,000 (p)",
     labels = function (br) br * 10,
     limits = c(-.1, 1.1)
   ) +

### 2024-05-22 law-firms-mapping/Paper_ru.Rmd
Rename folder, #3

---
output:
  odt_document:
    reference_odt: assets/reference.odt
bibliography: assets/biblio_en.bib
csl: assets/citations.csl
link-bibliography: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE, 
  message = FALSE, 
  echo = FALSE, 
  fig.width = 5.7, 
  fig.height = 4,
  fig.id = FALSE,
  dpi = 300)
```

```{r imports, cache = TRUE, include = FALSE}
source("helpers/Plots.R")
```

УДК 911.3, 519.688

<div custom-style="Author">

**П. О. Сёмин**

Пермский государственный национальный исследовательский университет, Пермь, Россия

E-mail: ntsp\@ya.ru

</div>

<br>

<div custom-style="Title">

Картографирование российских малых и средних предприятий по открытым данным налоговой службы

</div>

<br>

*В статье представлена гибкая и воспроизводимая методика картографирования малых и средних предприятий (МСП) в России. В настоящий момент не хватает качественных наборов данных для визуализации характеристик экономики страны на карте. Существующие официальные источники содержат информацию о местоположении тех или иных объектов в виде человекочитаемых адресов, в то время как ГИС-платформы требуют машиночитаемые пространственные данные с координатами. Представленный в работе набор инструментов — это попытка решить проблему и упростить подготовку данных для картографирования. В качестве источника сведений используются три набора открытых данных Федеральной налоговой службы, в частности, выгрузки реестра МСП. Для их обработки, нормализации адресов и геокодирования было разработано специальное приложение командной строки на языке программирования Python. Несмотря на большой объём исходных данных, приложение работает на среднем современном компьютере и не требует обращения к облачным вычислительным мощностям. Работоспособность инструментов продемонстрирована путём картографирования небольшой части исходного набора данных — сведений о юридических фирмах.*

***Ключевые слова:*** *открытые данные, пространственные данные, картографирование, геокодирование, малое и среднее предпринимательство, набор данных, большие данные, Федеральная налоговая служба.*

<br>

<div custom-style="Author">

**P. O. Syomin**

Perm State University, Perm, Russia

E-mail: ntsp\@ya.ru

</div>

<br>

<div custom-style="Title">

Mapping Russian Small and Medium-Sized Businesses Using Tax Service Open Data

</div>

<br>

*This paper presents a flexible and reproducible workflow for creating highly detailed maps of small and medium-sized businesses (SMBs) in Russia. Currently, there is a lack of data sources for mapping the Russian economy, and existing official datasets often provide location information in human-readable addresses rather than machine-readable geographical coordinates. The proposed set of tools aims to solve this problem and simplify the mapping process. The study utilizes three open datasets published by the Federal Tax Service, including open dumps of SMB registry, as its data source. To transform the source data into tables and normalize and geocode the addresses, a Python CLI application was developed. Despite the large volume of the initial data archives, the application can run on a regular modern personal computer without the need for cloud computing resources. The workflow's usability is demonstrated by mapping a small subset of data, specifically legal companies in Russia.*

***Keywords:*** *open data, spatial data, mapping, geocoding, small and medium-sized business, dataset, big data, Federal Tax Service.*

<br>

**Введение.** Современные картографические инструменты, используемые в науке и практике, обычно предполагают использование источников хорошо структурированных пространственных данных. Под «пространственными» в данном случае понимаются такие данные, которые представлены в одном из распространённых и широко используемых форматов и содержат географические координаты объектов в одной из стандартных систем координат (например, WGS84). Получение таких наборов данных может быть трудной задачей, которая существенно замедлит процесс исследования или увеличит его стоимость. Запрос со стороны исследовательского сообщества и бизнес-кругов на качественные пространственные данные настолько важен, что понемногу стимилирует правительства разных стран создавать геопорталы или упрощать распространение геоданных иными способами [@Jozefowicz2020].

В России официальная информация время от времени публикуется в машиночитаемом виде с координатами, но как правило, открытые геоданные доступны весьма плохо. Относительно слабая доступность таких данных может создавать проблемы при картографировании различных общественных процессов: экономических, географических, культурных. Даже если набор государственных данных содержит географическую привязку в виде «обычного», человекочитаемого адреса, его не получится использовать напрямую для создания карты, так как ГИС-приложениям необходимы машиночитаемые координаты. Для преобразования адресов в координаты служит процедура, известная как геокодирование (geocoding). Для российских адресов обычно используется несколько популярных геокодеров: Geocoding API от Яндекса или Google, сервис геокодирования компании DaData, инструмент OpenStreetMap Nominatim. Они обеспечивают сравнительно хорошее качество, но обычно предоставляются за плату, а в бесплатном варианте сложны в настройке или запрещают сохранять получаемые данные для последующего использования. Соответственно, актуален вопрос об удобном инструменте, который позволил бы облегчить картографирование ценных наборов открытых данных.

Одним из наборов государственных открытых данных в России, который потенциально имеет высокую ценность для исследователей в сфере экономической географии, может являться реестра субъектов малого и среднего предпринимательства (МСП). Указанный реестр ведётся с конца 2015 года на основании ст. 4.1 Федерального закона от 24.07.2007 №207-ФЗ «О развитии малого и среднего предпринимательства в Российской Федерации». Оператор реестра — Федеральная налоговая служба (ФНС России). Реестр автоматически включает в себя сведения обо всех коммерческих юридических лицах и индивидуальных предпринимателях, которые соответствуют установленным критериям. Основных критериев два: среднесписочная численность работников не превышает 250 человек, а доход за год — 2 миллиарда рублей.

Реестр МСП представляет собой наиболее полный общедоступный источник дезагрегированных сведений о российских коммерческих организациях. Другие варианты, такие как системы бизнес-аналитики (например, Спарк-Интерфакс или Контур.Фокус) или веб-портал Единого государственного реестра юридических лиц, доступны за плату или ограничивают бесплатное использование, что ограничивает их полезность. Данные официальной статистики бесплатны, но предоставляются в сводном (агрегированном) виде, то есть без детализации до уровня отдельных фирм. Напротив, полные выгрузки реестра МСП ежемесячно размещаются в разделе открытых данных ФНС России, доступны бесплатно и детализированы до конкретных организаций и предпринимателей. Проблема в том, что координат в них нет — только адреса. Кроме того, выгрузки имеют большой объём (суммарно несколько сотен гигабайт) и сложную структуру.

**Цель статьи.** Данная работа направлена на то, чтобы представить гибкую и воспроизводимую методику картографирования российских малых и средних фирм на основе открытых данных ФНС России. Воспроизводимость достигается за счёт разработки и публикации открытого приложения для обработки и геокодирования данных налоговой службы, которое может быть использовано исследователями для своих задач. В статье приводится пример картографирования российских юридических фирм на основе описанной методики. Аналогичный подход может использоваться для картографирования юридических лиц и индивидуальных предпринимателей в любой сфере деятельности.

**Обзор литературы.** Типичным источником данных для исследований по экономической географии России являются ресурсы Федеральной службы государственной статистики (Росстат) [@Sarancha2014; @Kuznetsova2018; @Petrov2021; @Safronov2021; @Fedorov2022]. К ним относятся официальная статистика, переписи населения и статистические обследования, доступные на официальном сайте Росстата, а также база данных показателей муниципальных образований (БДМО) и единая межведомственная информационно-статистическая система (ЕМИСС). Росстат является важнейшим провайдером информации в исследованиях в области экономической географии, и его данные явно или неявно используются почти во всех исследованиях. Эти данные не пригодны для картографирования напрямую из-за отсутствия координат. Тем не менее, они содержат информацию о местоположении с точностью до регионов, муниципалитетов и городов, которую для создания карты можно объединить с пространственными данными из других источников, таких как GADM или Natural Earth.

Экономико-географы также используют административные данные ФНС России в качестве альтернативного источника информации. Эти данные публикуется на официальном сайте указанного органа власти или его спутниковых веб-ресурсах. В исследованиях применяются, в частности, сведения о доходах физических лиц [@Petrov2021], о налоговых поступлениях по регионам или видам экономической деятельности [@Rostislav2020; @Safronov2021], о среднесписочной численности сотрудников организаций [@Rostislav2021], о численности малого и среднего бизнеса и мерах их поддержки [@Gumenyuk2022]. ФНС России не является специализированным статистическим органом, но собирает огромный объем информации о различных аспектах российской экономики, в том числе о хозяйствующих субъектах, имуществе, доходах и расходах. Географическое положение в данных ФНС, как и у Росстата, описывается с помощью адресных элементов. Стоит отметить, что ФНС России поддерживает Федеральную информационную адресную систему (ФИАС), предоставляющую официальные открытые адресные данные по России, которые широко используются другими государственными органами, а также бизнесом.

В современных исследованиях географы часто оперируют информацией, детализированной до уровня отдельных фирм. Эти дезагрегированные данные можно использовать напрямую [@Rostislav2021; @Kolomak2023] или для расчета производных показателей, которых нет в статистике Росстата [@Sarancha2014; @Gumenyuk2022; @Makushin2023]. В исследованиях на уровне городов [@Lachininsky2021; @Nikonorov2021; @Kozhevnikov2023] информация о конкретных фирмах имеет особую важность. Такая же детализированность крайне ценна при анализе узких секторов экономики, которые отдельно не представлены в агрегированной статистике [@Moiseeva2016; @Pankratov2023]. Для получения подобных данных исследователи обычно используют коммерческие или открытые геоинформационные сервисы, а также системы бизнес-аналитики [@Lachininsky2021; @Moiseeva2016; @Kuzminov2021]. Однако эти источники характеризуются двумя недостатками: они либо платные, либо сложны для использования. Так, системы бизнес-аналитики, такие как Спарк или Контур.Фокус, являются коммерческими, хотя и содержат огромный набор сведений о различных аспектах деятельности компаний. Общедоступные картографические сервисы наподобие 2ГИС и Яндекс.Карт позволяют выгружать данные только за плату, а автоматизированный сбор сведений, доступных бесплатно (так называемый парсинг, или скрейпинг), запрещён их пользовательскими соглашениями. Открытые платформы, такие как OpenStreetMap, бесплатны, но выгрузки (дампы) их данных имеют большой объём и сложную структуру — преобразование в удобный формат требует много времени и сил.

**Данные.** Картографирование предполагает использование одного основного источника данных вместе с двумя дополнительными, а также двух справочных таблиц.

Основным источником данных являются выгрузки (дампы) открытых данных реестра МСП[^1], которые публикуются каждый месяц, начиная с августа 2016 года. Каждая выгрузка содержит полную информацию реестра на дату публикации, что приводит к существенному дублированию. Технически выгрузка представляет собой ZIP-архив, содержащий тысячи XML-файлов, каждый из которых содержит информацию о 900 юридических лицах или индивидуальных предпринимателях. Атрибуты XML-файла включают, в частности, номер налогоплательщика (ИНН), наименование юридического лица или имя физического лица, адрес регистрации (область, район, город, населённый пункт), а также коды видов деятельности (основного и дополнительных) по Общероссийскому классификатору видов экономической деятельности (ОКВЭД) в редакции 2014 года.

[^1]: Единый реестр субъектов малого и среднего предпринимательства [электронный ресурс]. URL: https://www.nalog.gov.ru/opendata/7707329152-rsmp/ (дата обращения: 29.02.2024)

Дополнительными источниками информации выступают два других набора открытых данных ФНС России. Первый из них называется «Сведения о суммах доходов и расходов по данным бухгалтерской (финансовой) отчетности организации»[^2], а второй — «Сведения о среднесписочной численности работников организации»[^3]. Оба доступны приблизительно с 2019 года и обновляются не реже одного раза в год, хотя обновления могут быть нерегулярными. Технически эти наборы данных тоже представляют собой коллекции ZIP-архивов, содержащих файлы XML. Каждый архив содержит данные за определенную дату, а в каждом XML-файле представлены сведения для 900 юридических лиц. Информация о физических лицах не включена в эти наборы данных из-за необходимости защиты персональных данных.

[^2]: Сведения о суммах доходов и расходов по данным бухгалтерской (финансовой) отчетности организации [электронный ресурс]. URL: https://www.nalog.gov.ru/opendata/7707329152-revexp/ (дата обращения: 29.02.2024) 
[^3]: Сведения о среднесписочной численности работников организации [электронный ресурс]. URL: https://www.nalog.gov.ru/opendata/7707329152-sshr2019/ (дата обращения: 29.02.2024)

Справочными таблицами, используемыми для картографирования, являются набор данных «Населенные пункты России: число жителей и географические координаты», предоставленный проектом «Инфраструктура научно-исследовательских данных»[^4] и набор данных «Города России», опубликованный коммерческой компанией HFLabs[^5]. Оба содержат адреса и географические координаты российских городов и населенных пунктов. Поскольку информация о некоторых городах в них отсутствовала, автор статьи добавил её вручную на основе сведений Федеральной информационной адресной системы[^6].

[^4]: Населенные пункты России: численность населения и географические координаты [электронный ресурс]. URL: (дата обращения: 29.02.2024) https://data.rcsi.science/data-catalog/datasets/160/
[^5]: Города России [электронный ресурс]. URL: https://github.com/hflabs/city (дата обращения: 29.02.2024)
[^6]: Федеральная информационная адресная система [электронный ресурс]. URL: https://fias.ru (дата обращения: 29.02.2024)

Для обработки исходных данных было разработано приложение командной строки (CLI) на языке программирования Python. Технически оно основано на стандартной библиотеке Python и нескольких сторонних компонентах, включая быструю библиотеку обработки XML *lxml*, широко известное средство для работы с табличными данными *Pandas*, популярную платформу больших данных *Apache Spark* и простой в использовании фреймворк консольных приложений *Typer*. Исходный код приложения с краткой документацией размещён в общедоступном репозитории на платформе GitHub[^8]. Далее в статье описывается принципиальный (то есть без погружения в детали) алгоритм обработки данных, реализованный в приложении. «Входом» этого алгоритма являются исходные наборы данных ФНС России (справочные таблицы де-факто являются частью приложения). На «выходе» получаются таблицы с геокодированными сведениями, пригодные для использования в любом ГИС-приложении, поддерживающем CSV-файлы в качестве формата данных для векторных слоёв.

[^8]: PavelSyomin/russian-smb-companies: Dataset on small and medium business companies registered in Russia based on Federal Tax Service open data [электронный ресурс]. URL: https://github.com/PavelSyomin/russian-smb-companies (дата обращения: 29.02.2024)

*Шаг 1* — загрузка данных. Исходные ZIP-архивы загружаются с сайта ФНС и сохраняются локально.

*Шаг 2* — извлечение и фильтрация данных. Данные реестра МСП извлекаются из ZIP-архивов и сохраняются в таблицах в формате CSV, так что один ZIP-архив преобразуется в 1 CSV-таблицу. Многие неиспользуемые атрибуты отбрасываются, поэтому размер CSV-таблиц получается намного меньше размера исходных архивов. Кроме того, на этом этапе опционально осуществляется фильтрация по коду основного вида деятельности: это ускоряет процесс и дополнительно уменьшает объём CSV-таблиц. Например, для подготовки демонстрационного примера о юридических фирмах был выбран код основного вида деятельности 69.10 (деятельность в области права по ОКВЭД). Аналогичным образом преобразуются два вспомогательных набора данных, только без фильтрации по ОКВЭД, потому что кодов вида деятельности в них нет.

*Шаг 3* — агрегация данных: она позволяет удалить дублирующиеся записи и собрать сведения из множества CSV-таблиц в одну. Дублирование происходит из-за того, что каждая выгрузка реестра МСП — это сведения реестра целиком, а не только лишь обновления ранее размещённой информации. Кроме того, в ходе агрегирования вспомогательные наборы данных фильтруются так, чтобы в них остались сведения только о тех фирмах, которые были отобраны по коду ОКВЭД на шаге 2. В качестве уникального идентификатора компании используется ИНН.

*Шаг 4* — геокодирование. Адрес нормализуются и преобразуются в географические координаты с использованием справочных таблиц. Полнота геокодирования — около 98%. Поскольку в исходных данных адрес детализирован до городов или населенных пунктов, географические координаты также относятся к городам или населенным пунктам.

Опциональный *шаг 5* — объединение и трансформация результатов, полученных на шагах 3–4, в панельную таблицу. В панельной таблице сведения о каждой фирме за каждый год представлены одной строкой. Таблицы, которые производятся на шагах 3 и 4, построены по-другому: в них для каждой строки указаны даты актуальности сведений, что позволяет добиться большей компактности, но менее удобно для анализа.

Для подготовки демонстрационного примера набор табличных данных, полученный с помощью приложения, был отфильтрован по году (выбран 2021 год). Карты создавались с помощью скрипта, написанного на языке программирования R, с дополнительными пакетами, включая, в частности, ggplot2 и sf. Исходный код статьи доступен в отдельном репозитории[^9].

[^9]: ru-smb-companies-papers/legal-companies-mapping at main · PavelSyomin/ru-smb-companies-papers [электронный ресурс]. URL: https://github.com/PavelSyomin/ru-smb-companies-papers/tree/main/legal-companies-mapping (дата обращения: 29.02.2024)

**Результаты.** Хотя данные детализированы до конкретных фирм, они также содержат информацию о местонахождении с указанием как региона (субъекта федерации), так и населённого пункта, поэтому их можно агрегировать. На рис. 1 показано количество юридических фирм по регионам. Из этой визуализации видны некоторые пространственные закономерности: с одной стороны, концентрация юридических фирм в Москве и Московской области, а также в поясе, протянувшемся от Урала через Поволжье к Краснодару, а с другой стороны, сравнительно небольшое их число, например, в большинстве субъектов, образующих Центральный экономический район, что может быть связано с близостью столицы, оттягивающей на себя большинство компаний.

```{r fig-1-regions-map, fig.cap = "**Рис. 1.** Число юридических фирм по регионам (субъектам федерации) в России"}
regions_map
```

На рис. 2 представлена карта числа юридических фирм по городам и сельским населенным пунктам. По сути это выполненный с довольно высокой детализацией (до населённых пунктов) обзор юридического бизнеса в России. Карта похожа на региональную (рис. 1), однако данные на уровне городов позволяют заметить и другие тенденции, например, концентрацию бизнеса в региональных центрах. Кроме того, очевидна высокая концентрация в Москве и ее окрестностях, и аналогичная картина просматривается в Санкт-Петербурге. Подобную карту можно построить не только для страны в целом, но и для отдельного региона (Свердловская область на рис. 3).

```{r fig-2-settlements-map, fig.cap = "**Рис. 2.** Число юридических фирм в городских и сельских населённых пунктах России"}
settlements_map
```

<br>

```{r fig-3-settlements-map-svr, fig.cap = "**Рис. 3.** Число юридических фирм в городских и сельских населённых пунктах Свердловской области"}
settlements_map_svr
```

Наличие дополнительных данных о финансовых результатах и числе сотрудников позволяет построить дополнительные карты. На рис. 4 показано пространственное распределение прибыли юридических компаний, а на рис. 5 — аналогичное распределение по численности сотрудников. Обе карты включают только компании с ненулевой прибылью или количеством сотрудников. Карты похожи на ту, что изображена на рис. 2, но отличаются в деталях.

```{r fig-4-profit-map, fig.cap = "**Рис. 4.** Суммарная прибыль юридических фирм по городам и сельским населённым пунктам в России"}
settlements_map_profit
```

<br>

```{r fig-5-employees-map, fig.cap = "**Рис. 5.** Суммарное число работников юридических фирм по городам и сельским населённым пунктам в России"}
settlements_map_empl
```

**Заключение.** В данной статье предлагается гибкое решение для картографирования субъектов малого и среднего предпринимательства в России. Оно основано на использовании открытых данных, публикуемых Федеральной налоговой службой, которые предварительно обрабатываются и геокодируются с помощью специально разработанного приложения с открытым исходным кодом на языке программирования Python. В качестве примера приведены карты юридических фирм и их бизнес-показателей (прибыль и количество сотрудников), подтверждающие удобство использования предлагаемого метода. Код приложения, используемого для подготовки набора данных, и код скриптов, используемых для создания карт, опубликованы в общедоступных репозиториях на GitHub. Несмотря на большой объем исходных данных (до нескольких терабайт в распакованном виде), инструменты оптимизированы для работы на обычном современном компьютере средней конфигурации и не требуют приобретения облачных вычислительных ресурсов. Кроме того, геокодирование выполняется с помощью простых справочных таблиц вместо коммерческих API, поэтому весь процесс осуществляется бесплатно и в автономном режиме.

Описанный подход имеет ряд недостатков и ограничений. Прежде всего, несмотря на то, что инструменты разработаны максимально простыми, они все равно требуют некоторых базовых знаний в области программирования, поскольку исследователь должен запустить приложение Python и, следовательно, должен уметь устанавливать Python и работать с командной строкой. Далее, генерация набора данных, несмотря на усилия по оптимизации, по-прежнему занимает значительное время (до одного-двух дней, если не используется фильтрация). В-третьих, точность геокодирования высока, но не равна 100%. Наконец, исходные данные ограничены только субъектами малого и среднего предпринимательства, поэтому любые выводы, основанные на них, следует делать с осторожностью. Тем не менее, предложенный метод позволяет осуществить наиболее полное картографирование российского бизнеса, основанное на открытых данных и алгоритмах. Это может быть особенно полезно для разведочного анализа, для изучения малого и среднего предпринимательства как такового и для анализа узких секторов экономики, которые не попадают в обычную статистическую отчетность.

# Список литературы

<div id="refs"></div>

<br>

# Сведения об авторе

Сёмин Павел Олегович, аспирант кафедры социально-экономической географии Пермского государственного национального исследовательского университета, г. Пермь, Россия; разработчик в ООО «АгроСофтвер», г. Москва, Россия.

E-mail: ntsp\@ya.ru

ORCID: 0000-0002-4015-9206

# About the author

Pavel Olegovich Syomin, PhD Student, Department of Social and Economic Geography, Perm State University, Perm, Russia; Software Developer, AgroSoftware LLC, Moscow, Russia.

E-mail: ntsp\@ya.ru

ORCID: 0000-0002-4015-9206

### 2024-05-22 law-firms-mapping/Slides.Rmd
Rename folder, #3

---
title: "Картографирование малых и средних предприятий в России"
output: powerpoint_presentation
author: "Сёмин Павел Олегович, аспирант ПГНИУ"
date: "14.03.2024"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
library(knitr)
```

```{r imports, cache = TRUE, include = FALSE}
source("helpers/Plots.R")
```

# Постановка задачи

- Показать на карте все российские малые и средние предприятия (МСП).
- Вариант: фильтр по по ОКВЭД для выбора конкретной сферы деятельности.
- Практическая применимость: как часть анализа данных и как визуализация.

# Проблема с данными

- ЕГРЮЛ — нет прямого доступа к базе, нет координат.
- Яндекс/2ГИС/Spark — платно (бесплатный парсинг затруднён).
- OSM — неполные данные, сложный формат выгрузок.

# Реестр МСП

- Доступен в открытых данных ФНС.
- Есть ИНН, адрес регистрации до населённого пункта, коды ОКВЭД.
- Можно объединять с другими наборами открытых данных по ИНН (доходы/расходы, работники).
- Сложно обработать (многоуровневый XML, объём 2Тб, если распаковать), нужно геокодировать адреса.

# Обработка реестра МСП

- Консольное приложение на Python.
- Загрузка → Распаковка и фильтрация → Агрегация → Геокодирование.
- Агрегация убирает дубли (выгрузки идут ежемесячно в полном виде).
- Геокодирование на базе справочных таблиц даёт нормализованные адреса + координаты (распознано 97% адресов).

# Преобразованный набор данных (пример)

```{r dataset-sample}
smb_data %>% 
    filter(
    start_date <= "2021-12-31",
    end_date >= "2021-12-31"
  ) %>% 
  select(tin, settlement, lon, lat) %>% 
  slice_head(n = 10) %>% 
  knitr::kable(
    col.names = c("ИНН", "Населённый пункт", "Долгота", "Широта"))
```

# Тестирование

- Только юридические компании (это демонстрационный вариант, программа позволяет показать любые).
- 2021 год.
- Вся Россия.
- Число фирм, прибыль, число работников.

# Россия, число фирм

```{r russia-legal-firms-count}
settlements_map
```

# Россия, прибыль

```{r russia-legal-firms-profit}
settlements_map_profit
```

# Россия, число работников

```{r russia-legal-firms-empl}
settlements_map_empl
```

# Выводы

- Разработано приложение, которое из открытых данных ФНС делает набор данных о МСП в России, готовый для картографирования.
- Подготовлены демонстрационные карты, сделанные с помощью полученного набора данных.

### 2024-05-23 law-firms-mapping/Paper.Rmd
Make Russian version of plots, modify plot styles, edit Russian text, make Russian references list, #3

--- 2024-02-29 legal-companies-mapping/Paper.Rmd
@@
 ```{r setup, include=FALSE}
+library(dplyr)
+library(forcats)
+library(ggplot2)
+library(readr)
+library(scales)
+library(sf)
+library(tidyr)
+
 knitr::opts_chunk$set(
@@
   fig.width = 5.7, 
-  fig.height = 4,
+  fig.height = 3,
   fig.id = FALSE,
@@
 The state SMB registry was created at the end of 2015 in accordance with Art.
-4.1 of Federal Law of July 24, 2007 No 207-FZ “On the development of small and medium-sized businesses in the Russian Federation”.
+4.1 of Federal Law of July 24, 2007 No 209-FZ “On the development of small and medium-sized businesses in the Russian Federation”.
 The SMB registry is operated by the Federal Tax Service (FTS) of Russia and includes information about all Russian SMBs that meet the inclusion criteria.

### 2024-05-23 law-firms-mapping/Paper_ru.Rmd
Make Russian version of plots, modify plot styles, edit Russian text, make Russian references list, #3

@@
     reference_odt: assets/reference.odt
-bibliography: assets/biblio_en.bib
-csl: assets/citations.csl
+bibliography: assets/biblio_ru.bib
+csl: "assets/gost-r-7-0-100-2018-numeric-alphabetical.csl"
 link-bibliography: false
@@
 ```{r imports, cache = TRUE, include = FALSE}
+RU <- TRUE
 source("helpers/Plots.R")
@@
 
-Картографирование российских малых и средних предприятий по открытым данным налоговой службы
+Картографирование российских субъектов малого и среднего предпринимательства по открытым данным налоговой службы
 
@@
 
-*В статье представлена гибкая и воспроизводимая методика картографирования малых и средних предприятий (МСП) в России.
+*В статье представлена гибкая и воспроизводимая методика картографирования субъектов малого и среднего предпринимательства (МСП) в России.
 В настоящий момент не хватает качественных наборов данных для визуализации характеристик экономики страны на карте.
@@
 Despite the large volume of the initial data archives, the application can run on a regular modern personal computer without the need for cloud computing resources.
-The workflow's usability is demonstrated by mapping a small subset of data, specifically legal companies in Russia.*
+The workflow's usability is demonstrated by mapping a small subset of data, specifically law firms in Russia.*
 
@@
 Получение таких наборов данных может быть трудной задачей, которая существенно замедлит процесс исследования или увеличит его стоимость.
-Запрос со стороны исследовательского сообщества и бизнес-кругов на качественные пространственные данные настолько важен, что понемногу стимилирует правительства разных стран создавать геопорталы или упрощать распространение геоданных иными способами [@Jozefowicz2020].
+Запрос со стороны академического сообщества и бизнес-кругов на качественные пространственные данные настолько важен, что понемногу стимулирует правительства разных стран создавать геопорталы или упрощать распространение геоданных иными способами [@Jozefowicz2020].
 
@@
 Они обеспечивают сравнительно хорошее качество, но обычно предоставляются за плату, а в бесплатном варианте сложны в настройке или запрещают сохранять получаемые данные для последующего использования.
-Соответственно, актуален вопрос об удобном инструменте, который позволил бы облегчить картографирование ценных наборов открытых данных.
+Соответственно, актуален вопрос об удобном инструменте, который позволил бы облегчить картографирование ценных наборов открытых государственных данных.
 
-Одним из наборов государственных открытых данных в России, который потенциально имеет высокую ценность для исследователей в сфере экономической географии, может являться реестра субъектов малого и среднего предпринимательства (МСП).
+Одним из наборов открытых государственных данных в России, который потенциально имеет высокую ценность для исследователей в сфере экономической географии, может являться реестр субъектов малого и среднего предпринимательства (МСП).
 Указанный реестр ведётся с конца 2015 года на основании ст.
-4.1 Федерального закона от 24.07.2007 №207-ФЗ «О развитии малого и среднего предпринимательства в Российской Федерации».
+4.1 Федерального закона от 24.07.2007 №209-ФЗ «О развитии малого и среднего предпринимательства в Российской Федерации».
 Оператор реестра — Федеральная налоговая служба (ФНС России).
 Реестр автоматически включает в себя сведения обо всех коммерческих юридических лицах и индивидуальных предпринимателях, которые соответствуют установленным критериям.
-Основных критериев два: среднесписочная численность работников не превышает 250 человек, а доход за год — 2 миллиарда рублей.
+Основных критериев два: среднесписочная численность работников не превышает 250 человек, а доход за год — 2 миллиардов рублей.
 
 Реестр МСП представляет собой наиболее полный общедоступный источник дезагрегированных сведений о российских коммерческих организациях.
-Другие варианты, такие как системы бизнес-аналитики (например, Спарк-Интерфакс или Контур.Фокус) или веб-портал Единого государственного реестра юридических лиц, доступны за плату или ограничивают бесплатное использование, что ограничивает их полезность.
+Другие варианты, такие как системы бизнес-аналитики (например, Спарк-Интерфакс или Контур.Фокус) или веб-портал Единого государственного реестра юридических лиц, доступны за плату или ограничивают бесплатное использование, что уменьшает их полезность.
 Данные официальной статистики бесплатны, но предоставляются в сводном (агрегированном) виде, то есть без детализации до уровня отдельных фирм.
@@
 Проблема в том, что координат в них нет — только адреса.
-Кроме того, выгрузки имеют большой объём (суммарно несколько сотен гигабайт) и сложную структуру.
+Кроме того, выгрузки имеют большой объём (суммарно несколько сотен гигабайт в сжатом виде и более двух терабайт в несжатом) и сложную структуру.
 
@@
 **Обзор литературы.** Типичным источником данных для исследований по экономической географии России являются ресурсы Федеральной службы государственной статистики (Росстат) [@Sarancha2014; @Kuznetsova2018; @Petrov2021; @Safronov2021; @Fedorov2022].
-К ним относятся официальная статистика, переписи населения и статистические обследования, доступные на официальном сайте Росстата, а также база данных показателей муниципальных образований (БДМО) и единая межведомственная информационно-статистическая система (ЕМИСС).
-Росстат является важнейшим провайдером информации в исследованиях в области экономической географии, и его данные явно или неявно используются почти во всех исследованиях.
+К ним относятся официальная статистика, переписи населения и статистические обследования, доступные на официальном сайте этого государственного органа, а также база данных показателей муниципальных образований (БДМО) и единая межведомственная информационно-статистическая система (ЕМИСС).
+Росстат является важнейшим провайдером информации для исследований в области экономической географии, и его данные явно или неявно используются почти во всех научных работах.
 Эти данные не пригодны для картографирования напрямую из-за отсутствия координат.
@@
 
-Экономико-географы также используют административные данные ФНС России в качестве альтернативного источника информации.
-Эти данные публикуется на официальном сайте указанного органа власти или его спутниковых веб-ресурсах.
-В исследованиях применяются, в частности, сведения о доходах физических лиц [@Petrov2021], о налоговых поступлениях по регионам или видам экономической деятельности [@Rostislav2020; @Safronov2021], о среднесписочной численности сотрудников организаций [@Rostislav2021], о численности малого и среднего бизнеса и мерах их поддержки [@Gumenyuk2022].
+Экономико-географы используют административные данные ФНС России в качестве альтернативного источника информации.
+Эти данные публикуется на официальном сайте указанного органа власти или связанных с ним веб-ресурсах.
+В исследованиях применяются, в частности, сведения о доходах физических лиц [@Petrov2021], о налоговых поступлениях по регионам или видам экономической деятельности [@Rostislav2020; @Safronov2021], о среднесписочной численности работников организаций [@Rostislav2021], о числе субъектов малого и среднего предпринимательства и мерах их поддержки [@Gumenyuk2022].
 ФНС России не является специализированным статистическим органом, но собирает огромный объем информации о различных аспектах российской экономики, в том числе о хозяйствующих субъектах, имуществе, доходах и расходах.
@@
 
-**Данные.** Картографирование предполагает использование одного основного источника данных вместе с двумя дополнительными, а также двух справочных таблиц.
+**Данные.** Картографирование по методике, предложенной в этой статье, предполагает использование одного основного источника данных вместе с двумя дополнительными, а также двух справочных таблиц.
 
-Основным источником данных являются выгрузки (дампы) открытых данных реестра МСП[^1], которые публикуются каждый месяц, начиная с августа 2016 года.
+Основным источником данных являются выгрузки (дампы) реестра МСП[^1], которые публикуются каждый месяц, начиная с августа 2016 года.
 Каждая выгрузка содержит полную информацию реестра на дату публикации, что приводит к существенному дублированию.
@@
 
-[^1]: Единый реестр субъектов малого и среднего предпринимательства [электронный ресурс]. URL: https://www.nalog.gov.ru/opendata/7707329152-rsmp/ (дата обращения: 29.02.2024)
+[^1]: Единый реестр субъектов малого и среднего предпринимательства [Электронный ресурс]. URL: https://www.nalog.gov.ru/opendata/7707329152-rsmp/ (дата обращения: 29.02.2024)
 
@@
 Каждый архив содержит данные за определенную дату, а в каждом XML-файле представлены сведения для 900 юридических лиц.
-Информация о физических лицах не включена в эти наборы данных из-за необходимости защиты персональных данных.
+Информация об индивидуальных предпринимателях не включена в эти наборы из-за необходимости защиты персональных данных.
 
-[^2]: Сведения о суммах доходов и расходов по данным бухгалтерской (финансовой) отчетности организации [электронный ресурс].
+[^2]: Сведения о суммах доходов и расходов по данным бухгалтерской (финансовой) отчетности организации [Электронный ресурс].
 URL: https://www.nalog.gov.ru/opendata/7707329152-revexp/ (дата обращения: 29.02.2024)
-[^3]: Сведения о среднесписочной численности работников организации [электронный ресурс]. URL: https://www.nalog.gov.ru/opendata/7707329152-sshr2019/ (дата обращения: 29.02.2024)
+[^3]: Сведения о среднесписочной численности работников организации [Электронный ресурс]. URL: https://www.nalog.gov.ru/opendata/7707329152-sshr2019/ (дата обращения: 29.02.2024)
 
@@
 
-[^4]: Населенные пункты России: численность населения и географические координаты [электронный ресурс]. URL: (дата обращения: 29.02.2024) https://data.rcsi.science/data-catalog/datasets/160/
-[^5]: Города России [электронный ресурс]. URL: https://github.com/hflabs/city (дата обращения: 29.02.2024)
-[^6]: Федеральная информационная адресная система [электронный ресурс]. URL: https://fias.ru (дата обращения: 29.02.2024)
+[^4]: Населенные пункты России: численность населения и географические координаты [Электронный ресурс]. URL: (дата обращения: 29.02.2024) https://data.rcsi.science/data-catalog/datasets/160/
+[^5]: Города России [Электронный ресурс]. URL: https://github.com/hflabs/city (дата обращения: 29.02.2024)
+[^6]: Федеральная информационная адресная система [Электронный ресурс]. URL: https://fias.ru (дата обращения: 29.02.2024)
 
 Для обработки исходных данных было разработано приложение командной строки (CLI) на языке программирования Python.
-Технически оно основано на стандартной библиотеке Python и нескольких сторонних компонентах, включая быструю библиотеку обработки XML *lxml*, широко известное средство для работы с табличными данными *Pandas*, популярную платформу больших данных *Apache Spark* и простой в использовании фреймворк консольных приложений *Typer*.
+Технически оно основано на стандартной библиотеке этого языка и нескольких сторонних компонентах, включая быструю библиотеку обработки XML *lxml*, широко известное средство для работы с табличными данными *Pandas*, популярную платформу больших данных *Apache Spark* и простой в использовании фреймворк консольных приложений *Typer*.
 Исходный код приложения с краткой документацией размещён в общедоступном репозитории на платформе GitHub[^8].
@@
 
-[^8]: PavelSyomin/russian-smb-companies: Dataset on small and medium business companies registered in Russia based on Federal Tax Service open data [электронный ресурс].
+[^8]: PavelSyomin/russian-smb-companies: Dataset on small and medium business companies registered in Russia based on Federal Tax Service open data [Электронный ресурс].
 URL: https://github.com/PavelSyomin/russian-smb-companies (дата обращения: 29.02.2024)
 
-*Шаг 1* — загрузка данных. Исходные ZIP-архивы загружаются с сайта ФНС и сохраняются локально.
+*Шаг 1* — загрузка данных. Исходные ZIP-архивы загружаются с сайта ФНС России и сохраняются локально.
 
 *Шаг 2* — извлечение и фильтрация данных.
-Данные реестра МСП извлекаются из ZIP-архивов и сохраняются в таблицах в формате CSV, так что один ZIP-архив преобразуется в 1 CSV-таблицу.
+Данные реестра МСП извлекаются из ZIP-архивов и сохраняются в таблицах в формате CSV, так что один ZIP-архив преобразуется в одну CSV-таблицу.
 Многие неиспользуемые атрибуты отбрасываются, поэтому размер CSV-таблиц получается намного меньше размера исходных архивов.
@@
 Кроме того, в ходе агрегирования вспомогательные наборы данных фильтруются так, чтобы в них остались сведения только о тех фирмах, которые были отобраны по коду ОКВЭД на шаге 2.
-В качестве уникального идентификатора компании используется ИНН.
+В качестве уникального идентификатора компании при этом используется ИНН.
 
@@
 
-Для подготовки демонстрационного примера набор табличных данных, полученный с помощью приложения, был отфильтрован по году (выбран 2021 год).
+Для подготовки демонстрационного примера набор табличных данных, полученный с помощью приложения, был отфильтрован по году (выбран 2021-й).
 Карты создавались с помощью скрипта, написанного на языке программирования R, с дополнительными пакетами, включая, в частности, ggplot2 и sf.
@@
 
-[^9]: ru-smb-companies-papers/legal-companies-mapping at main · PavelSyomin/ru-smb-companies-papers [электронный ресурс].
+[^9]: ru-smb-companies-papers/legal-companies-mapping at main · PavelSyomin/ru-smb-companies-papers [Электронный ресурс].
 URL: https://github.com/PavelSyomin/ru-smb-companies-papers/tree/main/legal-companies-mapping (дата обращения: 29.02.2024)
@@
 1 показано количество юридических фирм по регионам.
-Из этой визуализации видны некоторые пространственные закономерности: с одной стороны, концентрация юридических фирм в Москве и Московской области, а также в поясе, протянувшемся от Урала через Поволжье к Краснодару, а с другой стороны, сравнительно небольшое их число, например, в большинстве субъектов, образующих Центральный экономический район, что может быть связано с близостью столицы, оттягивающей на себя большинство компаний.
+Из этой визуализации видны некоторые пространственные закономерности, например: с одной стороны, концентрация юридических фирм в Москве и Московской области, а также в поясе, протянувшемся от Урала через Поволжье к Краснодару, а с другой стороны, сравнительно небольшое их число, например, в субъектах, образующих Центральный экономический район, что может быть связано с близостью столицы, оттягивающей на себя большинство компаний.
 
-```{r fig-1-regions-map, fig.cap = "**Рис. 1.** Число юридических фирм по регионам (субъектам федерации) в России"}
+```{r fig-1-regions-map, fig.cap = "**Рис. 1.** Число юридических фирм по регионам (субъектам федерации) в России в 2021 году"}
 regions_map
@@
 По сути это выполненный с довольно высокой детализацией (до населённых пунктов) обзор юридического бизнеса в России.
-Карта похожа на региональную (рис.
+Карта похожа на региональную (см.
+рис.
 1), однако данные на уровне городов позволяют заметить и другие тенденции, например, концентрацию бизнеса в региональных центрах.
-Кроме того, очевидна высокая концентрация в Москве и ее окрестностях, и аналогичная картина просматривается в Санкт-Петербурге.
+Кроме того, очевидна высокое сосредоточение в Москве и ее окрестностях.
+Аналогичная картина просматривается в Санкт-Петербурге.
 Подобную карту можно построить не только для страны в целом, но и для отдельного региона (Свердловская область на рис.
@@
 
-```{r fig-2-settlements-map, fig.cap = "**Рис. 2.** Число юридических фирм в городских и сельских населённых пунктах России"}
+```{r fig-2-settlements-map, fig.cap = "**Рис. 2.** Число юридических фирм в городских и сельских населённых пунктах России в 2021 году"}
 settlements_map
@@
 
-```{r fig-3-settlements-map-svr, fig.cap = "**Рис. 3.** Число юридических фирм в городских и сельских населённых пунктах Свердловской области"}
+```{r fig-3-settlements-map-svr, fig.cap = "**Рис. 3.** Число юридических фирм в городских и сельских населённых пунктах Свердловской области в 2021 году"}
 settlements_map_svr
@@
 
-```{r fig-4-profit-map, fig.cap = "**Рис. 4.** Суммарная прибыль юридических фирм по городам и сельским населённым пунктам в России"}
+```{r fig-4-profit-map, fig.cap = "**Рис. 4.** Суммарная прибыль юридических фирм по городам и сельским населённым пунктам в России в 2021 году"}
 settlements_map_profit
@@
 
-```{r fig-5-employees-map, fig.cap = "**Рис. 5.** Суммарное число работников юридических фирм по городам и сельским населённым пунктам в России"}
+```{r fig-5-employees-map, fig.cap = "**Рис. 5.** Суммарное число работников юридических фирм по городам и сельским населённым пунктам в России в 2021 году"}
 settlements_map_empl
@@
 Оно основано на использовании открытых данных, публикуемых Федеральной налоговой службой, которые предварительно обрабатываются и геокодируются с помощью специально разработанного приложения с открытым исходным кодом на языке программирования Python.
-В качестве примера приведены карты юридических фирм и их бизнес-показателей (прибыль и количество сотрудников), подтверждающие удобство использования предлагаемого метода.
+В качестве примера приведены карты юридических фирм и их бизнес-показателей (прибыль и количество сотрудников), подтверждающие работоспособность предлагаемого метода.
 Код приложения, используемого для подготовки набора данных, и код скриптов, используемых для создания карт, опубликованы в общедоступных репозиториях на GitHub.
@@
 Описанный подход имеет ряд недостатков и ограничений.
-Прежде всего, несмотря на то, что инструменты разработаны максимально простыми, они все равно требуют некоторых базовых знаний в области программирования, поскольку исследователь должен запустить приложение Python и, следовательно, должен уметь устанавливать Python и работать с командной строкой.
+Прежде всего, несмотря на то, что инструменты сделаны максимально простыми, они все равно требуют некоторых базовых знаний в области программирования, поскольку исследователь должен запустить приложение на Python и, следовательно, должен уметь устанавливать Python и работать с командной строкой.
 Далее, генерация набора данных, несмотря на усилия по оптимизации, по-прежнему занимает значительное время (до одного-двух дней, если не используется фильтрация).
-В-третьих, точность геокодирования высока, но не равна 100%.
+В-третьих, точность геокодирования хотя и высока, но не равна 100%.
 Наконец, исходные данные ограничены только субъектами малого и среднего предпринимательства, поэтому любые выводы, основанные на них, следует делать с осторожностью.

### 2024-05-23 law-firms-mapping/helpers/Plots.R

@@ -32,11 +32,15 @@ regions_map <- smb_data %>%
   geom_sf(aes(fill = n), size = .1) +
   coord_sf(crs = ru_crs) +
   scale_fill_distiller(
-    name = "Count of legal companies",
+    name = ifelse(exists("RU"), "Число юридических фирм", "Count of law firms"),
     palette = "YlGn",
     direction = 1) +
-  theme_bw(base_size = 11, base_family = "Times New Roman") +
-  theme(legend.position = "bottom")
+  theme_void(base_size = 11, base_family = "Times New Roman") +
+  theme(
+    legend.title.position = "top",
+    legend.position = c(.05, 0), 
+    legend.justification = c(0, 0),
+    legend.direction = "horizontal")
 regions_map
 
 settlements_df <- smb_data %>%
@@ -49,15 +53,20 @@ settlements_df <- smb_data %>%
   st_as_sf(coords = c("lon", "lat"), crs = 4326)
 
 settlements_map <- regions %>% ggplot() +
-  geom_sf(size = .1) +
-  geom_sf(data = settlements_df, aes(color = log10(n)), size = .5, shape = 21) +
+  geom_sf(fill = "white", color = "gray50") +
+  geom_sf(data = settlements_df, aes(color = n), size = .5, shape = 21) +
   coord_sf(crs = ru_crs) +
   scale_color_distiller(
-    name = "Count of legal companies (log10)",
+    name = ifelse(exists("RU"), "Число юридических фирм", "Count of law firms"),
     palette = "YlGn",
+    transform = "log10",
     direction = 1) +
-  theme_bw(base_size = 11, base_family = "Times New Roman") +
-  theme(legend.position = "bottom")
+  theme_void(base_size = 11, base_family = "Times New Roman") +
+  theme(
+    legend.title.position = "top",
+    legend.position = c(.05, 0), 
+    legend.justification = c(0, 0),
+    legend.direction = "horizontal")
 settlements_map
 
 settlements_df_svr <- smb_data %>%
@@ -72,15 +81,16 @@ settlements_df_svr <- smb_data %>%
 
 settlements_map_svr <- ru_svr %>% 
   ggplot() +
-  geom_sf() +
-  geom_sf(data = settlements_df_svr, aes(color = log10(n)), size = 2) +
+  geom_sf(fill = "white", color = "gray50") +
+  geom_sf(data = settlements_df_svr, aes(fill = n), size = 2, shape = 21) +
   coord_sf(crs = 4326) +
[…]

### 2024-05-30 law-firms-names/Conference-slides.qmd
Slides about law firm names for the MARS-2024 conferece

---
title: "География названий юридических фирм в России"
author: "Сёмин Павел Олегович, Пермский государственный национальный исследовательский университет"
format: 
  revealjs:
    scrollable: true
---

```{r}
#| label: imports
#| include: false

library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(sf)
library(spatstat)
library(stringr)
library(tidyr)
```

```{r}
#| label: data-preparation
#| cache: true
#| include: false

# load the data
data <- read_csv("../../ru-smb-companies/legal/panel.csv")
cluster_labels <- read_csv("cluster_labels.csv")
vectors <- read_csv("names_vectors.csv")
ru <- st_read("ru.geojson")

# additional variables
ru_crs <- st_crs("+proj=aea +lat_0=0 +lon_0=100 +lat_1=68 +lat_2=44 +x_0=0 +y_0=0 +ellps=krass +towgs84=28,-130,-95,0,0,0,0 +units=m +no_defs")
org_forms <- c(
  "общество с ограниченной ответственностью",
  "общество с дополнительной ответственностью",
  "акционерное общество",
  "открытое акционерное общество",
  "закрытое акционерное общество",
  "публичное акционерное общество",
  "непубличное акционерное общество",
  "хозяйственное партнерство",
  "полное товарищество",
  "товарищество на вере"
)

# prepare firm names
firm_names <- data %>% 
  filter(
    kind == 1, # companies only
    year <= 2021, # 2016–2021
  ) %>% 
  mutate(
    name_id = row_number(),
    settlement = case_when(
      region == "Москва" ~ "Москва",
      region == "Санкт-Петербург" ~ "Санкт-Петербург",
      TRUE ~ settlement)
  ) %>% 
  select(name_id, name = org_name, region, settlement, lat, lon, year) %>% 
  mutate(
    name = str_to_lower(name),
    name = str_remove(name, paste0(org_forms, collapse = "|")),
    name = str_remove_all(name, fixed("\"")),
    name = str_trim(name))
unique_firm_names <- distinct(firm_names, name) %>% 
  left_join(vectors) %>% 
  drop_na(dim_0)

# cluster firm names
set.seed(42)
fit <- kmeans(select(unique_firm_names, dim_0:dim_255), 
              centers = 20)
unique_firm_names$cluster <- fit$cluster

# add cluster groups
firm_names_clusters <- firm_names %>% 
  left_join(unique_firm_names, by = "name") %>% 
  left_join(cluster_labels) %>% 
  select(name_id, name, region, settlement, lat, lon, year, group) %>% 
  drop_na() %>% 
  mutate(group = factor(group))
```

# Задача

Выявить географические различия в именовании юридических фирм в России

# Данные

Реестр МСП с сайта ФНС России (2021 год)

# Обработка данных

## Реестр → таблица

Извлечение и геокодирование сведений о фирмах с помощью специально разработанного консольного приложения на Python

## Пример данных из таблицы {.smaller}

```{r}
kable(
  select(firm_names, name, region, settlement, lat, lon) %>% 
    slice_head(n = 10)
)
```

## Названия → векторы

Векторизация названий с помощью YandexGPT Embeddings API

## Пример вектора

```{r}
vectors[1, -1] %>% as.numeric() %>% round(2)
```

## Кластеризация названий

KMeans на векторах, 20 кластеров

## Примеры кластеров

```{r}
#| label: clusters-sample

nc_sample <- unique_firm_names %>% 
  group_by(cluster) %>% 
  slice_sample(n = 3) %>% 
  select(name)

kable(nc_sample)
```

## Группировка кластеров

Ручной выборочный анализ названий в каждом кластере, присвоение описательной метки каждому кластеру, группировка кластеров

## Ход группировки

```{r}
#| label: groups

kable(cluster_labels)
```

## Результат группировки

Выделено 4 типа именования:

* сфера деятельности — право, юриспруденция (field);
* характер услуг — экспертиза, консультация, посредничество (service);
* организационная форма — группа, партнёрство (form);
* всё остальное (misc)

## Построение карт плотности по группам

Оценка функции плотности на основе набора точек, соответствующих названиям, и картографирование результата

# Результат

---

```{r}
#| label: density-map
#| cache: true

ppp_data <- firm_names_clusters %>% 
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>% 
  st_transform(n, crs = ru_crs) %>% 
  filter(year == 2021, region != "Москва", region != "Санкт-Петербург")
firm_names_ppp <- as.ppp(
  ppp_data$geometry, W = as.owin(st_transform(ru, ru_crs)))
firm_names_ppp$marks <- ppp_data$group
firm_names_spp <- split(firm_names_ppp, marks(firm_names_ppp))
plot(density(firm_names_spp, sigma = 125000))
```

# Выводы

1. Распространение разных типов именования юридических фирм в целом похоже.
2. Тип «сфера деятельности» наиболее широко распространён и ярко выражен. 
3. Тип «организационная форма» преобладает в Центральной России и отчасти на Урале.
4. Тип «характер услуг» наиболее выражен на юге России, а также образует второстепенный «пояс» от Москвы до Урала.

### 2024-06-04 legal-companies-migration/Paper.Rmd
Fix typos

@@
 Во-первых, само понятие миграции фирм в большей степени адаптировано к небольшим организациям, расположенным в одном месте [@Plant2007], и юридические фирмы, как правило, именно такие.
-Во-вторых, сфера юридических услуг сравнительно слабо связано с набором «традиционных» географических факторов размещения, «привязывающих» фирмы к их первоначальному местонахождению, таких как минеральные, почвенные, лесные, климатические, энергетические ресурсы, и это повышает шанс найти закономерности, отличающиеся от классических.
+Во-вторых, сфера юридических услуг сравнительно слабо связана с набором «традиционных» географических факторов размещения, «привязывающих» фирмы к их первоначальному местонахождению, таких как минеральные, почвенные, лесные, климатические, энергетические ресурсы, и это повышает шанс найти закономерности, отличающиеся от классических.
 В-третьих, юридические услуги часто рассматриваются как одна из разновидностей producer services [@Hansen1993], которые положительно влияют на региональное развитие [@Hansen1990] или выступают как лимитирующий фактор [@Hitchens1997], из-за чего без надлежащего юридического обслуживания экономическое развитие, особенно малых фирм, замедляется.
@@
 Он был подготовлен на базе выгрузок из реестра МСП, размещённых на сайте Федеральной налоговой службы в формате открытых данных[^1], а также данных о доходах и расходах[^2] и среднесписочной численности работников организаций[^3].
-Объём исходных выгрузок составляет около 200 Гб в сжатом виде и более 2 Тб в несжатом, поэтому для их обработки автор самостоятельно разработал консольное приложение *ru-smb-companies*[^4] на языке программирования Python, которое формирует на базе открытых компактную таблицу с фирмами, имеющими необходимый код ОКВЭД.
+Объём исходных выгрузок составляет около 200 Гб в сжатом виде и более 2 Тб в несжатом, поэтому для их обработки автор самостоятельно разработал консольное приложение *ru-smb-companies*[^4] на языке программирования Python, которое формирует на базе открытых данных компактную таблицу с фирмами, имеющими необходимый код ОКВЭД.
 Набор данных охватывает период с августа 2016 г.
@@
 Он содержит сведения о `r companies_count` организации.
-Таблица имеет, в частности, такие поля (колонки): наименование компании, её категория (малое, среднее или микропредприятие), ИНН, регион, административный район, населённый пункт регистрации, его географические координаты, доходы, расходы и число работников организации за каждый год с 2019.
+Таблица имеет, в частности, такие поля (колонки): наименование компании, её категория (малое, среднее или микропредприятие), ИНН, регион, административный район, населённый пункт регистрации, его географические координаты, доходы, расходы и число работников организации за каждый год начиная с 2019.
 
@@
 Однозначной тенденции не прослеживается: видно, что часть субъектов федерации теряет юридические фирмы, но другая часть их приобретает, причём количество регионов в двух группах приблизительно равное.
-Абсолютные значения сальда релокации невелики (как правило, около нуля, изредка приближаясь к одному-двум десяткам).
+Абсолютные значения сальдо релокации невелики (как правило, около нуля, изредка приближаясь к одному-двум десяткам).
 Из этого правила есть два заметных исключения: в Московскую область переехало больше всего фирм, а сама Москва при этом — абсолютный лидер по оттоку фирм.
@@
 Детализация статистики до населённых пунктов, находящихся внутри субъектов федерации, позволяет понять, что общая тенденция, характерная для региона (приток или отток фирм), как правило, не транслируется на следующий уровень географических объектов: в регионе обычно есть населённые пункты как с положительным, так и с отрицательным сальдо релокации.
-Это свидетельствует о том, что межрегиональные потоки фирм накладываются на внутрирегиональные, из-за чего некоторые населённы региона, в целом теряющего фирмы, могут, тем не менее, привлекать фирмы из других населённых пунктов этого же или других регионов, и наоборот.
+Это свидетельствует о том, что межрегиональные потоки фирм накладываются на внутрирегиональные, из-за чего некоторые населённые пункты региона, в целом теряющего фирмы, могут, тем не менее, привлекать фирмы из других населённых пунктов этого же или других регионов, и наоборот.
 

### 2024-06-07 legal-services-provision/journal-article/Article.Rmd
Edit after review, #2

@@
 
-Provision of Commercial Legal Services in Russian Cities
+Spatial and Regression Analysis of Provision of Commercial Legal Services in Russian Cities
 
@@
 e-mail: ntsp\@ya.ru
+
+ORCID: 0000-0002-4015-9206
 
@@
 
-**Key words:** legal services, law firms, availability, cities, market saturation, small and medium-sized businesses, open data
+**Key words:** legal services, law firms, availability, cities, spatial analysis, regression analysis, small and medium-sized businesses, open data
 
@@
 
-Ресей қалаларында коммерциялық заң қызметтерін көрсету
+Ресей қалаларында коммерциялық заң қызметтерін көрсетудің кеңістіктік және регрессиялық талдауы
 
@@
 
-**Кілт сөздер:** заң қызметтері, заң фирмалары, қауіпсіздік, қалалар, нарықтың қанықтығы, шағын және орта бизнес, ашық деректер
+**Кілт сөздер:** заң қызметтері, заң фирмалары, қауіпсіздік, қалалар, кеңістіктік талдау, регрессиялық талдау, шағын және орта бизнес, ашық деректер
 
@@
 
-Обеспеченность коммерческими юридическими услугами в городах России
+Географический и регрессионный анализ обеспеченности коммерческими юридическими услугами в городах России
 
@@
 
-**Ключевые слова:** юридические услуги, юридические фирмы, обеспеченность, города, насыщенность рынка, малое и среднее предпринимательство, открытые данные
+**Ключевые слова:** юридические услуги, юридические фирмы, обеспеченность, города, пространственный анализ, регрессионный анализ, малое и среднее предпринимательство, открытые данные
 
@@
 The analysis was conducted in the R programming language in conjunction with additional packages.
-The source code for this paper can be accessed via the public repository: https://github.com/PavelSyomin/ru-smb-companies-papers/tree/main/legal-companies-concentration-and-availability.
+The source code for this paper can be accessed via the public repository: https://github.com/PavelSyomin/ru-smb-companies-papers/tree/main/legal-services-provision.
 
@@
 
+The findings of this study can serve as a scientific foundation for data-driven decision-making by federal and regional government bodies, local authorities, and private actors.
+The federal government can use the results to develop a policy framework for supporting the establishment of small and medium-sized law firms, particularly in the regions and cities where the availability of legal services is limited, or for stimulating the export of legal services from cities with a high level of provision to the peripheries.
+This could be achieved, for instance, by creating local offices of central law firms or providing online consultations.
+Regional and local authorities may utilize the findings to make informed decisions about the actions required for the development of local law business, taking into account the city population size and the identified strategies for achieving a high level of legal services provision.
+It is important to note, however, that in certain instances, a “do nothing” approach may be optimal, because it has been found that some cities with a small number of law firms nevertheless provide a high level of legal services, and therefore no government intervention is necessary.
+Private actors, such as law firms, may use the results of this study to inform their spatial expansion plans.
+For consumers, the results of this study may provide knowledge about the spatial patterns of legal services provision, enabling them to make decisions about where to seek legal advice.
+Finally, the data acquisition technique described in this study can be used to re-obtain information about the individual law firms and perform its additional analysis aligned with specific objectives.
+
 ## References

### 2024-07-05 legal-companies-geography/Paper.Rmd
Editoral changes after first review, #1

@@
 
-[^2]: Анализ рынка юридических услуг в России в 2019-2023 гг, прогноз на 2024-2028 гг. Web resource: https://marketing.rbc.ru/research/41683/ (accessed 14.04.2024)
+[^2]: Analysis of legal services market in Russia in 2019–2023 with forecast for 2024–2028 (in Russian). Web resource: https://marketing.rbc.ru/research/41683/ (accessed 14.04.2024)
 
@@
 To address these challenges, the use of open administrative data may be a viable solution.
-In particular, the publicly available archives of the Registry of Small and Medium-Sized Businesses (SMB Registry) distributed by the Federal Tax Service (FTS) of Russia, in combination with a few other open data collections of the same governmental body, appear to be a valuable resource for the task at hand.
+In particular, the publicly available archives of the Registry of small and medium-sized businesses (SMB registry) distributed by the Federal Tax Service (FTS) of Russia, in combination with a few other open data collections of the same governmental body, appear to be a valuable resource for the task at hand.
 These data sources have already been used by geographers [@Rostislav2021; @Gumenyuk2022].
@@
 
-**Data.** The research utilizes two primary data sources: the Unified Registry of Small and Medium-Sized Businesses (SMB Registry)[^3] and the Information about the Average Number of Employees of Organizations (Information about Employees)[^4].
+**Data.** The research utilizes two primary data sources: the Unified registry of small and medium-sized businesses (SMB registry)[^3] and the Information about the average number of employees of organizations (Information about employees)[^4].
 Both of these are available on the FTS website as open data.
 
-[^3]: Единый реестр субъектов малого и среднего предпринимательства. Web resource: https://www.nalog.gov.ru/opendata/7707329152-rsmp/ (accessed 15.04.2024)
-[^4]: Сведения о среднесписочной численности работников организации. Web resource: https://www.nalog.gov.ru/opendata/7707329152-sshr2019/ (accessed 15.04.2024)
+[^3]: Unified registry of small and medium-sized businesses (in Russian). Web resource: https://www.nalog.gov.ru/opendata/7707329152-rsmp/ (accessed 15.04.2024)
+[^4]: Information about the average number of employees of organizations (in Russian). Web resource: https://www.nalog.gov.ru/opendata/7707329152-sshr2019/ (accessed 15.04.2024)
 
-The SMB Registry dataset contains full dumps of the respective registry in zipped XML files.
+The SMB registry dataset contains full dumps of the respective registry in zipped XML files.
 The initial dataset was made available in August 2016, and updates have been released every month.
 Each dump contains the entire registry and includes, in particular, the company or individual names, taxpayer IDs, and activity codes according to the All-Russia Classifier of Kinds of Economic Activity, as well as addresses of incorporation (up to the settlement name).
-According to the Russian laws, the Registry includes information about all commercial organizations and individual entrepreneurs that meet several criteria (SMB subjects).
+According to the Russian laws, the registry includes information about all commercial organizations and individual entrepreneurs that meet several criteria (SMB subjects).
 The most important criteria are limits on the revenue (no more than 2B Russian rubles) and on the average number of employees (no more than 250).
@@
 In addition to the primary data sources, two lookup tables with information on Russian cities and settlements are used to normalize and geocode the raw addresses contained in the FTS data.
-The first table is “Settlements of Russia: Population and Geographical Coordinates”[^5] prepared by the nonprofit open data initiative known as “Research Data Infrastructure” (RDI).
+The first table is “Settlements of Russia: population and geographical coordinates”[^5] prepared by the nonprofit open data initiative known as “Research Data Infrastructure” (RDI).
 The second table is “Cities of Russia”[^6] published by the commercial company HFLabs.
@@
 
-[^5]: Населенные пункты России: численность населения и географические координаты. Web resource: https://data.rcsi.science/data-catalog/datasets/160/ (accessed 15.04.2024)
-[^6]: Города России. Web resource: https://github.com/hflabs/city (accessed 15.04.2024)
+[^5]: Settlements of Russia: population and geographical coordinates (in Russian). Web resource: https://data.rcsi.science/data-catalog/datasets/160/ (accessed 15.04.2024)
+[^6]: Cities of Russia (in Russian). Web resource: https://github.com/hflabs/city (accessed 15.04.2024)
 
@@
 
-[^7]: Информация о кассационных и апелляционных судах общей юрисдикции. Web resource: https://vsrf.ru/press_center/news/28148/ (accessed 15.04.2024)
-[^8]: Справочная информация: "Арбитражные суды". Web resource: https://www.consultant.ru/document/cons_doc_LAW_56549/ (accessed 15.04.2024)
+[^7]: Information about cassation and appeal ordinary courts (in Russian). Web resource: https://vsrf.ru/press_center/news/28148/ (accessed 15.04.2024)
+[^8]: Reference information: “Commercial courts” (in Russian). Web resource: https://www.consultant.ru/document/cons_doc_LAW_56549/ (accessed 15.04.2024)
 
@@
 
-[^9]: Численность занятых. Web resource: https://fedstat.ru/indicator/34051 (accessed 24.04.2024)
+[^9]: Number of people employed (in Russian). Web resource: https://fedstat.ru/indicator/34051 (accessed 24.04.2024)
 
@@
 Step 1 is data download.
-The dumps of the SMB Registry and the archives of Information about Employees were downloaded from the FTS servers and stored on the local machine.
+The dumps of the SMB registry and the archives of Information about employees were downloaded from the FTS servers and stored on the local machine.
 Additionally, their backup was uploaded to the author's personal Yandex.Disk account, thus enabling other researchers to access the raw data even if FTS decides to remove it (as indeed occurred with the data on employees).
@@
 
-[^11]: Плиточная карта России. Web resource: https://dianov.org/all/plitochnaya-karta-rossii/ (accessed 23.04.2024)
-[^12]: Плиточная карта РФ. Web resource: https://gist.github.com/tukachev/c3262242a20b3050e0459b4e3afc3a51 (accessed 23.04.2024)
+[^11]: Tile grid map of Russia (in Russian). Web resource: https://dianov.org/all/plitochnaya-karta-rossii/ (accessed 23.04.2024)
+[^12]: Tile grid map of Russia. Web resource: https://gist.github.com/tukachev/c3262242a20b3050e0459b4e3afc3a51 (accessed 23.04.2024)
 [^13]: PavelSyomin / ru-smb-companies-papers. Web resource: https://github.com/PavelSyomin/ru-smb-companies-papers/tree/main/legal-companies-geography (accessed 23.04.2024)
@@
 
-Figure 1 depicts the average annual number of law firms by region.
+Fig.
+1 depicts the average annual number of law firms by region.
 Several conclusions can be drawn from this visualization.
@@
 
-```{r fig-1-counts-map, fig.cap="Figure 1. The median annual count of law firms by regions of Russia over the period from 2016 to 2021"}
+```{r fig-1-counts-map, fig.cap="Fig. 1. The median annual count of law firms by regions of Russia over the period from 2016 to 2021"}
 counts_map
@@
 
-Figure 2 illustrates the median annual share of organizations among all law firms in regions.
+Fig.
+2 illustrates the median annual share of organizations among all law firms in regions.
 The value varies considerably, ranging from approximately 30% to almost 70%.
@@
 Petersburg cities, along with the majority of the Volga and Siberian regions.
-The comparison of Figure 1 with Figure 2 leads to the conclusion that there is no clear relationship between the number of law firms and the share of companies among them.
+The comparison of Fig.
+1 with Fig.
+2 leads to the conclusion that there is no clear relationship between the number of law firms and the share of companies among them.
 
@@
 
-```{r fig-2-shares-map, fig.cap="Figure 2. The median annual share of companies among legal firms by regions of Russia over the period from 2016 to 2021"}
+```{r fig-2-shares-map, fig.cap="Fig. 2. The median annual share of companies among legal firms by regions of Russia over the period from 2016 to 2021"}
 shares_map
@@
 
-Figure 3 presents the combined results of the regional analysis based on the number of lawyers.
-The distribution of lawyers across regions is similar to that of law firms themselves (see Figure 1), suggesting that legal companies in different regions are quite similar to each other in the number of employees.
+Fig.
+3 presents the combined results of the regional analysis based on the number of lawyers.
+The distribution of lawyers across regions is similar to that of law firms themselves (see Fig.
+1), suggesting that legal companies in different regions are quite similar to each other in the number of employees.
 However, the role of legal business in regions' economies, as indicated by the share of lawyers in regions' labor forces, is distributed differently.
@@
 
-```{r fig-3-employees-map, fig.cap="Figure 3. The national and regional median annual shares of lawyers in regions of Russia over the period from 2018 to 2021"}
+```{r fig-3-employees-map, fig.cap="Fig. 3. The national and regional median annual shares of lawyers in regions of Russia over the period from 2018 to 2021"}
 employees_map
@@
 
-The changes in the number of law firms by region are depicted in Figure 4.
+The changes in the number of law firms by region are depicted in Fig.
+4.
 The number of such firms increased in the majority of regions, indicating the development of the legal business sector.
@@
 The regions with the most significant growth are some republics in the North Caucasus, as well as the Republic of Kalmykia and Nenets Autonomous Area.
-Their results are likely attributable to the low base effect, given that the absolute counts of law firms in all these regions are below 100 (see Figure 1).
+Their results are likely attributable to the low base effect, given that the absolute counts of law firms in all these regions are below 100 (see Fig.
+1).
 It is noteworthy that Leningrad and Moscow regions also demonstrated one of the highest growth rates.
@@
 
-```{r fig-4-timeline-map, fig.cap="Figure 4. The change in the number of law firms by region over the period from 2016 to 2021"}
+```{r fig-4-timeline-map, fig.cap="Fig. 4. The change in the number of law firms by region over the period from 2016 to 2021"}
 timeline_map

### 2024-07-05 legal-companies-geography/Script.R

@@ -167,10 +167,13 @@ timeline_map <- count(data, region, year) %>%
     aes(
       x = 2021,
       y = 1.25, 
-      label = ifelse(!is.na(change), sprintf("%+.0f", change), ""),
+      label = ifelse(
+        !is.na(change),
+        sub("-", "–", sprintf("%+.0f", change), fixed = TRUE),
+        ""),
       alpha = alpha),
     family = "Times New Roman",
-    size = 3.5,
+    size = 4,
     hjust = 1,
     color = "gray45",
     key_glyph = "blank") +
@@ -179,7 +182,7 @@ timeline_map <- count(data, region, year) %>%
     data = tiles,
     aes(x = 2016, y = 2, label = code_en),
     family = "Times New Roman",
-    size = 2.5,
+    size = 3,
     hjust = 0,
     vjust = 1) +
   scale_y_continuous(limits = c(NA, 2)) +

### 2024-07-06 law-firms-geography/Paper.Rmd
Rename folder, #1
Edit according to reviewer's notes: add custom definition of law business in methods, use per capita counts in Fig. 1 instead of absolute ones, #1

--- 2024-07-05 legal-companies-geography/Paper.Rmd
@@
 This system is responsible for protecting human rights and freedoms, ensuring the rule of law, and acting as an intermediary between people, businesses, and government.
-It helps to resolve conflicts, increase certainty and orderliness in social relations and thus contributes to the stability and development of society.
+It helps to resolve conflicts, increase certainty and orderliness in social relations, and thus contributes to the stability and development of society.
 This aspect of legal services is closely related to the social dimension of the geography of services.
@@
 
-[^2]: Analysis of legal services market in Russia in 2019–2023 with forecast for 2024–2028 (in Russian). Web resource: https://marketing.rbc.ru/research/41683/ (accessed 14.04.2024)
+[^2]: Analysis of the legal services market in Russia in 2019–2023 with forecast for 2024–2028 (in Russian). Web resource: https://marketing.rbc.ru/research/41683/ (accessed 14.04.2024)
 
@@
 Each dump contains the entire registry and includes, in particular, the company or individual names, taxpayer IDs, and activity codes according to the All-Russia Classifier of Kinds of Economic Activity, as well as addresses of incorporation (up to the settlement name).
-According to the Russian laws, the registry includes information about all commercial organizations and individual entrepreneurs that meet several criteria (SMB subjects).
+According to Russian laws, the registry includes information about all commercial organizations and individual entrepreneurs that meet several criteria (SMB subjects).
 The most important criteria are limits on the revenue (no more than 2B Russian rubles) and on the average number of employees (no more than 250).
-Consequently, the open dumps of this registry represent the most comprehensive publicly accessible data source on the Russian business entities.
+Consequently, the open dumps of this registry represent the most comprehensive publicly accessible data source on Russian business entities.
 Moreover, the archive of these dumps constitutes a time series spanning from August 2016 to the present, thereby providing a history of small and medium-sized business changes.
 
-The Information about Employees dataset contains company taxpayer IDs and average yearly numbers of employees for the previous year.
+The Information about employees dataset contains company taxpayer IDs and average yearly numbers of employees for the previous year.
 The data was initially published in 2019 and has been updated with varying frequencies, though at least annually.
@@
 District-level courts span their activity over two or more regions of Russia that constitute a judicial district, checking and reviewing decisions of lower-level courts.
-These category include appeal and cassation ordinary and commercial courts, with a total of 47 such courts in Russia.
+This category include appeal and cassation ordinary and commercial courts, with a total of 47 such courts in Russia.
 The location data was obtained from the Supreme Court of Russia website[^7] and the legal information system called “Consultant Plus”[^8].
@@
 
-Finally, an aggregated official statistic on the annual total number of employees in the regions of Russia[^9] was utilized.
+Finally, an aggregated official statistic on the annual total number of employees in the regions of Russia[^9] and on the average annual resident population[^10] was utilized.
 This data is provided by the Federal Statistics Service (Rosstat) through the publicly available Unified Interdepartmental Information and Statistical System.
@@
 [^9]: Number of people employed (in Russian). Web resource: https://fedstat.ru/indicator/34051 (accessed 24.04.2024)
+[^10]: Average annual resident population (in Russian). Web resource: https://fedstat.ru/indicator/31556 (accessed 24.04.2024)
 
 **Methods.** A Python command line tool, ru-smb-companies, was developed by the author of this paper to initially process large and complex raw XML datasets from the FTS servers and transform them into a single CSV table.
-The source code of this tool, along with a brief documentation, is published in the author's personal GitHub account[^10].
+The source code of this tool, along with a brief documentation, is published in the author's personal GitHub account[^11].
 Here, a concise summary of the data processing workflow is provided.
@@
 
-[^10]: Russian small and medium business dataset generator. Web resource: https://github.com/PavelSyomin/russian-smb-companies (accessed 15.04.2024)
+[^11]: Russian small and medium business dataset generator. Web resource: https://github.com/PavelSyomin/russian-smb-companies (accessed 15.04.2024)
 
@@
 The research adheres to the general framework of regional analysis and employs the methods of descriptive statistics and data visualization, including the plotting of data on a tile-grid map of Russia.
+In this study, the concept of legal business is approached in a relatively narrow sense and with a strong emphasis on formal considerations.
+The term encompasses (1) commercial legal entities or individual entrepreneurs (2) registered in Russia according to the Russian civil law (3) whose main activity is classified under code 69.10 in the All-Russia Classifier of Kinds of Economic Activity.
+This technique is consistent with the selection criterion for law firms used in [@Moiseeva2016].
+The analysis does not include non-profit legal services providers or specialized professionals such as notaries and advocates.
+Due to data availability constraints, it also omits large companies that are not present in the small and medium-sized businesses registry.
+In essence, the legal business, as it is defined in this context, encompasses legal counseling or similar activities.
 The analysis is comprised of four research problems, or questions, which are enumerated below with the description of the relevant techniques used to find the answers.
@@
 Problem 1 is the distribution of law firms across Russia's regions.
-To discover this distribution, the median number of law firms in each region is calculated for the period from 2016 to 2021.
+To discover this distribution, the median number of law firms in each region per 100,000 population is calculated for the period from 2016 to 2021.
 This result is then visualized on a map.
@@
 A tile grid map of Russia by I.
-Dianov[^11] is used with a modified implementation by Yu.
-Tukachev[^12].
-The source code of this paper is published in the author's repository on GitHub[^13].
+Dianov[^12] is used with a modified implementation by Yu.
+Tukachev[^13].
+The source code of this paper is published in the author's repository on GitHub[^14].
 
-[^11]: Tile grid map of Russia (in Russian). Web resource: https://dianov.org/all/plitochnaya-karta-rossii/ (accessed 23.04.2024)
-[^12]: Tile grid map of Russia. Web resource: https://gist.github.com/tukachev/c3262242a20b3050e0459b4e3afc3a51 (accessed 23.04.2024)
-[^13]: PavelSyomin / ru-smb-companies-papers. Web resource: https://github.com/PavelSyomin/ru-smb-companies-papers/tree/main/legal-companies-geography (accessed 23.04.2024)
+[^12]: Tile grid map of Russia (in Russian). Web resource: https://dianov.org/all/plitochnaya-karta-rossii/ (accessed 23.04.2024)
+[^13]: Tile grid map of Russia. Web resource: https://gist.github.com/tukachev/c3262242a20b3050e0459b4e3afc3a51 (accessed 23.04.2024)
+[^14]: PavelSyomin / ru-smb-companies-papers. Web resource: https://github.com/PavelSyomin/ru-smb-companies-papers/tree/main/law-firms-geography (accessed 23.04.2024)
 
@@
 Fig.
-1 depicts the average annual number of law firms by region.
+1 depicts the average relative annual number of law firms by region.
 Several conclusions can be drawn from this visualization.
-First, Moscow is the only city where the number of law firms exceeds 10 thousand, and this is the highest value in Russia.
-Second, some regions also have a relatively high number of law firms.
-Geographically, these regions form several groups.
-One of these groups is located in the south of the European part of Russia and consists of Krasnodar and Stavropol territories with Rostov, Volgograd, and Voronezh regions.
-The Republic of Bashkortostan and the Republic of Tatarstan with Samara, Sverdlovsk, and Chelyabinsk regions, as well as Perm Territory, form the Ural-Volga group.
-Tyumen, Omsk, Novosibirsk, Kemerovo regions with Krasnoyarsk Territory belong to the third group which can be designated as Siberian due to its location.
-The city of St.
-Petersburg, Nizhniy Novgorod Region, and Primorye Territory are separate regions with a relatively high number of law firms.
-Third, the majority of other regions have a relatively small number of law firms, with several republics of the North Caucasus and Altai and a portion of regions of the Russian North incorporating less than a hundred of such companies.
-It can be hypothesized that the number of law firms is positively associated with the economic development of a region.
-Furthermore, the population of a region may also be a factor to consider.
+First, Moscow and Saint Petersburg stand out as the only cities in Russia where the number of law firms exceeds 100 per 100,000 inhabitants.
+Secondly, certain regions also have a relatively high number of law firms per capita.
+Geographically, these regions are situated in different parts of the country, and in some cases, they form clusters.
+The most prominent of these groups is located in the Middle and Southern Urals with Western Siberia and consists of Sverdlovsk, Chelyabinsk, Tyumen, Omsk, and Novosibirsk regions.
+Another group is located in the Far East and includes Primorye territory, Sakhalin region, and, somewhat surprisingly, Kamchatka territory.
+Krasnodar territory, the Republic of Tatarstan, the Republic of Karelia, and Arkhangelsk region are distinct centers with a high number of law firms per capita surrounded by the regions with a more moderate number of such firms.
+Moscow region and its neighboring Ivanovo region are also characterized by a considerable number of law firms per capita, but it seems that their metrics may be the result of a “spillover effect” originating from Moscow city.
+Third, the majority of other regions have a relatively small number of law firms, with several republics of the North Caucasus and Altai and a portion of regions of the Russian North, sometimes incorporating less than ten such companies.
+It can be hypothesized that the number of law firms per capita is contingent upon the economic development of a region.
 
-The relative location of regions and groups with different numbers of law firms suggests the possibility of central-peripheries relations between them.
+The relative location of regions and groups with different numbers of law firms suggests the possibility of central-periphery relations between them.
 For instance, Moscow City and Region, St.
-Petersburg City, and Nizhny Novgorod Region may serve as hubs for their surrounding regions of Central and North-Western Russia.
-A similar phenomenon may be observed in other parts of the country: the Siberian group provides legal services for the Far East, while the regions of the European South serve the republics of the North Caucasus.
+Petersburg City, and the Republic of Tatarstan may serve as hubs for their surrounding areas of Central, North-Western Russia, and Volga region (Povolzhye).
+A similar phenomenon may be observed in other parts of the country: the Ural-Siberian cluster provides legal services for Eastern Siberia and the Far East, while the Krasnodar Territory serves the Rostov and Stavropol.
 
@@
 
-```{r fig-2-shares-map, fig.cap="Fig. 2. The median annual share of companies among legal firms by regions of Russia over the period from 2016 to 2021"}
+```{r fig-2-shares-map, fig.cap="Fig. 2. The median annual share of companies among law firms by regions of Russia over the period from 2016 to 2021"}
 shares_map
@@
 In contrast, several regions, all of which are located in the European part of Russia, have experienced a net decrease in the number of law firms, In comparison to the increase rates observed in other regions, the net change in the number of law firms from 2016 to 2021 in these regions is relatively low (usually less than 10%).
-In some cases, neither decline nor growth was clearly expressed (for example, the Republic of Altai, Lipetsk, and Tambov regions).
+In some cases, neither decline nor growth was clearly expressed (for example, in the Republic of Altai, Lipetsk, and Tambov regions).
 The regions with the most significant growth are some republics in the North Caucasus, as well as the Republic of Kalmykia and Nenets Autonomous Area.
@@
 First, the regions exhibited considerable variation in the number of law firms and lawyers.
-Several regional clusters with relatively high development of legal business were identified, including the European South, Volga-Ural, and Siberian.
+Several areas with a relatively high number of lawyers per capita were identified, including the Urals and Western Siberia cluster, a few regions of the Far East, Krasnodar territory, the Republic of Tatarstan, and a number of other regions.
 As might be expected, the cities of Moscow and St.

### 2024-07-06 law-firms-geography/Script.R

@@ -7,6 +7,7 @@ library(tidyr)
 courts <- read_csv("courts.csv")
 data <- read_csv("../../ru-smb-companies/legal/panel.csv")
 empl <- read_csv("empl.csv")
+pop <- read_csv("population.csv")
 tiles <- read_csv("russia-tiles.csv")
 
 # Common preparations
@@ -20,13 +21,20 @@ courts_by_region <- count(courts, region, name = "n_courts") %>%
 empl <- pivot_longer(
   empl, "2016":"2021", names_to = "year", values_to = "empl_total") %>% 
   mutate(year = as.numeric(year))
+pop <- pivot_longer(
+  pop, "2016":"2021", names_to = "year", values_to = "pop_total") %>% 
+  mutate(year = as.numeric(year))
 
-# Count of legal firms by region
-counts_map <- count(data, region, year) %>% 
+# Relative count of law firms by region
+counts_map_data <- count(data, region, year) %>% 
+  left_join(pop) %>% 
+  mutate(share = 1e5 * n / pop_total) %>% 
   group_by(region) %>% 
-  summarise(n = median(n)) %>% 
-  right_join(tiles, by = c("region" = "name")) %>% 
-  ggplot(aes(x = col, y = -row, fill = n)) +
+  summarise(share = median(share)) %>% 
+  right_join(tiles, by = c("region" = "name"))
+counts_map_breaks <- quantile(counts_map_data$share, seq(0, 1, .2), na.rm = TRUE)
+counts_map <- counts_map_data %>% 
+  ggplot(aes(x = col, y = -row, fill = share)) +
   geom_raster() +
   geom_text(
     aes(x = col - .4, y = -row + .3, label = code_en), 
@@ -34,28 +42,29 @@ counts_map <- count(data, region, year) %>%
     size = 3,
     hjust = 0) +
   geom_text(
-    aes(x = col - .4, y = -row - .3, label = round(n, 0)), 
+    aes(x = col - .4, y = -row - .3, label = round(share, 1)), 
     family = "Times New Roman", 
     size = 3,
     color = "gray20",
     hjust = 0,
     na.rm = TRUE) +
   scale_fill_binned(
-    name = "Median count of law firms (2016–21)\nNumbers on the map are exact values\nGrey fill is no data",
-    transform = "log10",
+    name = "Median number of law firms\nper 100,000 people (2016–21)\nScale breaks are 0.2-spaced quantiles\nNumbers on the map are exact values\nGrey fill is no data",
+    breaks = counts_map_breaks,
[…]

### 2024-07-08 law-firms-geography/Paper.Rmd
Minor edits, #1

@@
 [^9]: Number of people employed (in Russian). Web resource: https://fedstat.ru/indicator/34051 (accessed 24.04.2024)
-[^10]: Average annual resident population (in Russian). Web resource: https://fedstat.ru/indicator/31556 (accessed 24.04.2024)
+[^10]: Average annual resident population (in Russian). Web resource: https://fedstat.ru/indicator/31556 (accessed 06.07.2024)
 
@@
 In this study, the concept of legal business is approached in a relatively narrow sense and with a strong emphasis on formal considerations.
-The term encompasses (1) commercial legal entities or individual entrepreneurs (2) registered in Russia according to the Russian civil law (3) whose main activity is classified under code 69.10 in the All-Russia Classifier of Kinds of Economic Activity.
+The term encompasses commercial legal entities or individual entrepreneurs registered in Russia according to the Russian civil law whose main activity is classified under code 69.10 in the All-Russia Classifier of Kinds of Economic Activity.
 This technique is consistent with the selection criterion for law firms used in [@Moiseeva2016].
@@
 Secondly, certain regions also have a relatively high number of law firms per capita.
-Geographically, these regions are situated in different parts of the country, and in some cases, they form clusters.
+Geographically, these regions are situated in different parts of the country, and in some cases, they form clusters, or groups.
 The most prominent of these groups is located in the Middle and Southern Urals with Western Siberia and consists of Sverdlovsk, Chelyabinsk, Tyumen, Omsk, and Novosibirsk regions.
@@
 Moscow region and its neighboring Ivanovo region are also characterized by a considerable number of law firms per capita, but it seems that their metrics may be the result of a “spillover effect” originating from Moscow city.
-Third, the majority of other regions have a relatively small number of law firms, with several republics of the North Caucasus and Altai and a portion of regions of the Russian North, sometimes incorporating less than ten such companies.
+Third, several republics of the North Caucasus and Altai and a portion of regions of the Russian North have relatively small number of law firms, sometimes incorporating less than ten such businesses.
 It can be hypothesized that the number of law firms per capita is contingent upon the economic development of a region.
 
-The relative location of regions and groups with different numbers of law firms suggests the possibility of central-periphery relations between them.
+The relative location of regions and groups with different numbers of law firms per capita suggests the possibility of central-periphery relations between them.
 For instance, Moscow City and Region, St.
@@
 
-```{r fig-1-counts-map, fig.cap="Fig. 1. The median annual count of law firms by regions of Russia over the period from 2016 to 2021"}
+```{r fig-1-counts-map, fig.cap="Fig. 1. The median annual count of law firms per capita by regions of Russia over the period from 2016 to 2021"}
 counts_map
@@
 The regional analysis revealed several spatial features of legal business in Russia.
-First, the regions exhibited considerable variation in the number of law firms and lawyers.
-Several areas with a relatively high number of lawyers per capita were identified, including the Urals and Western Siberia cluster, a few regions of the Far East, Krasnodar territory, the Republic of Tatarstan, and a number of other regions.
+First, the regions exhibit considerable variation in the number of law firms and lawyers.
+Several areas with a relatively high number of lawyers per capita are identified, including the Urals and Western Siberia cluster, a few regions of the Far East, Krasnodar territory, the Republic of Tatarstan, and a number of other regions.
 As might be expected, the cities of Moscow and St.

### 2024-07-12 law-firms-names/conference-paper/Conference-slides.qmd
Reorganize law firms names folder to separate conference paper from journal article, add current code and data on journal article, #4

--- 2024-05-30 law-firms-names/Conference-slides.qmd
@@
 library(ggplot2)
+library(here)
 library(knitr)
@@
 # load the data
-data <- read_csv("../../ru-smb-companies/legal/panel.csv")
-cluster_labels <- read_csv("cluster_labels.csv")
-vectors <- read_csv("names_vectors.csv")
-ru <- st_read("ru.geojson")
+data <- read_csv(here("../../ru-smb-companies/legal/panel.csv"))
+cluster_labels <- read_csv(
+  here("conference-paper", "cluster-labels-conf.csv"))
+vectors <- read_csv(here("common", "names-vectors.csv"))
+ru <- st_read(here("common", "ru.geojson"))
 

### 2024-07-12 law-firms-names/common/EDA.R

@@ -1,0 +1,568 @@
+library(dplyr)
+library(ggplot2)
+library(glue)
+library(dbscan)
+library(lightgbm)
+library(plotly)
+library(readr)
+library(rjson)
+library(spatstat)
+library(stopwords)
+library(stringr)
+library(tidyr)
+library(tidytext)
+
+data <- read_csv("../../ru-smb-companies/legal/panel.csv")
+tiles <- read_csv("russia-tiles.csv")
+stopwords <- data.frame(
+  word = c(
+    stopwords("ru", "stopwords-iso"),
+    c("общество", "ограниченный", "ответственность", "акционерный")
+  )
+)
+
+# For the conference
+names_2021 <- data %>% 
+  filter(
+    kind == 1,
+    year == 2021) %>% 
+  select(region, area, settlement, settlement_type, name = org_name) %>% 
+  unnest_tokens(word, name) %>% 
+  left_join(hash_lemmas_opencorpora, by = c("word" = "token")) %>% 
+  select(-word, word = lemma) %>% 
+  anti_join(stopwords) %>% 
+  drop_na(region, word)
+
+top_tokens_by_region <- names_2021 %>% 
+  count(region, word) %>% 
+  filter(n >= 5) %>% 
+  group_by(region) %>% 
+  arrange(-n) %>% 
+  #slice_head(n = 10) %>% 
+  ungroup()
+  
+tokens <- count(top_tokens_by_region, word, wt = n, sort = TRUE)
+#write_csv(tokens, "top-10-tokens-by-region.csv")
+token_groups <- read_csv("top-10-tokens-by-region.csv")
+
+token_groups_by_regions <- top_tokens_by_region %>% 
+  left_join(token_groups, by = "word") %>% 
+  count(region, group, wt = n.x) %>% 
+  group_by(region) %>% 
+  mutate(share = n / sum(n)) %>% 
+  ungroup()
+
+token_groups_by_regions %>% 
+  group_by(region) %>% 
+  slice_max(share, with_ties = FALSE) %>% 
+  ungroup() %>% 
+  right_join(tiles, by = c("region" = "name")) %>% 
+  ggplot(aes(x = col, y = -row)) +
+  geom_tile(aes(fill = group, alpha = share)) +
+  geom_text(aes(label = code), size = 3) +
+  coord_equal()
+
+token_groups_by_regions %>% 
+  filter(group == "geo") %>%  
[…]

### 2024-07-12 law-firms-names/common/get_yandex_gpt_vectors.py

@@ -1,0 +1,49 @@
+import csv
+import time
+
+import requests
+import numpy as np
+import tqdm
+
+
+def get_embedding(text: str) -> np.array:
+    query_data = {
+        "modelUri": query_uri,
+        "text": text,
+    }
+
+    return np.array(
+        requests.post(embed_url, json=query_data, headers=headers).json()["embedding"]
+    )
+
+
+FOLDER_ID = "b1guv73jumjli9it2kum"
+IAM_TOKEN = "t1.9euelZqJjcqMjsySz4yakpqbnY3Lju3rnpWam5yPi5WQks2YyMyOiorOzpjl9Pd-EzpN-e9jbBrS3fT3PkI3TfnvY2wa0s3n9euelZrMypqYk5KVmcyPjsmKyIqOkO_8xeuelZrMypqYk5KVmcyPjsmKyIqOkA.A1u_ruhFatIschBArQEDgMiZJVotdF36Tu2IjEw_cJ3UO5ruz8qQSAMbGdwK6afYF_gUg6nqJ44ns04DxVByDw"
+
+query_uri = f"emb://{FOLDER_ID}/text-search-query/latest"
+embed_url = "https://llm.api.cloud.yandex.net:443/foundationModels/v1/textEmbedding"
+headers = {"Content-Type": "application/json", "Authorization": f"Bearer {IAM_TOKEN}", "x-folder-id": f"{FOLDER_ID}"}
+
+with open("names.csv") as f:
+    reader = csv.DictReader(f)
+    names = [row["name"] for row in reader]
+
+missing = []
+with open("names_vectors.csv", "w") as f:
+    fieldnames = ["name"] + [f"dim_{i}" for i in range(256)]
+    writer = csv.DictWriter(f, fieldnames=fieldnames)
+
+    writer.writeheader()
+
+    for name in tqdm.tqdm(names):
+        try:
+            emb = get_embedding(name)
+        except:
+            missing.append(name)
+
+        row = dict(zip(fieldnames, [name] + list(emb)))
+        writer.writerow(row)
+
+with open("missing.csv", "w") as f:
+    for name in missing:
+        print(name, file=f)

### 2024-07-12 law-firms-names/journal-article/Script.R

@@ -1,0 +1,144 @@
+library(dplyr)
+library(ggplot2)
+library(glue)
+library(here)
+library(readr)
+library(readxl)
+library(sf)
+library(stringr)
+library(tidyr)
+
+data <- read_csv(here("../../ru-smb-companies/legal/panel.csv"))
+tiles <- read_csv(here("common", "russia-tiles.csv"))
+ru <- st_read(here("common", "ru.geojson"))
+ru_crs <- st_crs("+proj=aea +lat_0=0 +lon_0=100 +lat_1=68 +lat_2=44 +x_0=0 +y_0=0 +ellps=krass +towgs84=28,-130,-95,0,0,0,0 +units=m +no_defs")
+org_forms <- c(
+  "общество с ограниченной ответственностью",
+  "общество с дополнительной ответственностью",
+  "акционерное общество",
+  "открытое акционерное общество",
+  "закрытое акционерное общество",
+  "публичное акционерное общество",
+  "непубличное акционерное общество",
+  "хозяйственное партнерство",
+  "полное товарищество",
+  "товарищество на вере"
+)
+
+firms <- data %>% 
+  filter(
+    kind == 1, # companies only
+    year <= 2021, # 2016–2021
+  ) %>% 
+  mutate(
+    name_id = row_number(),
+    settlement = case_when(
+      region == "Москва" ~ "Москва",
+      region == "Санкт-Петербург" ~ "Санкт-Петербург",
+      TRUE ~ settlement)
+  ) %>% 
+  select(
+    name_id,
+    tin,
+    name = org_name, 
+    region, 
+    settlement, 
+    lat,
+    lon, 
+    year
+  ) %>% 
+  mutate(
+    name = str_to_lower(name),
+    name = str_remove(name, paste0(org_forms, collapse = "|")),
+    name = str_remove_all(name, fixed("\"")),
+    name = str_trim(name)
+  )
+unique_names <- distinct(firms, name)
+
+# Save unique names to a file for processing with YandexGPT API
+write_csv(
+  unique_names,
+  glue("names-{strftime(Sys.time(), '%Y-%m-%d-%H-%M-%S')}.csv")
+)
+
+# Load vectors obtained from YandexGPT API
+vectors <- read_csv(here("common", "names-vectors.csv"))
+
+# Clustering
+## Find optimal number of clusters
+n_clusters <- 2:100
+scores <- sapply(
+  n_clusters, 
+  function(x) kmeans(
[…]

### 2024-07-16 law-firms-names/Paper.Rmd
Make article template, #4

---
title: "Spatial Analysis of Russian Law Firms Naming"
author: "Pavel Syomin¹"
output:
  word_document:
    reference_doc: journal-article/reference.docx
#bibliography: refs.bib
#csl: style.csl
link-bibliography: False  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<div custom-style="Affiliation">

¹Perm State University, Geographical faculty, Department of social and economic geography, Perm, Russia; e-mail: ntsp\@ya.ru

</div>

<div custom-style="Abstract">

**Abstract:** text.

</div>

<div custom-style="Keywords">

**Keywords:** naming; law firms; text embeddings; economic regions. 
</div>

# 1. Introduction

# 2. Materials and methods

## 2.1. Data

## 2.2. Methods

# 3. Results and discussion

# 4. Conclusion

<div custom-style="References-title">References</div>

<div id="refs"></div>

### 2024-07-16 law-firms-names/journal-article/Script.R
Make distance and neighborhood analysis, #4

@@ -7,6 +7,7 @@ library(readxl)
 library(sf)
 library(stringr)
 library(tidyr)
+library(units)
 
 data <- read_csv(here("../../ru-smb-companies/legal/panel.csv"))
 er <- read_csv(here("journal-article", "economic-regions.csv"))
@@ -105,14 +106,6 @@ clustered <- vectors %>%
   distinct(name, tin, .keep_all = TRUE)
 
 # Distance between regions
-regions_cnt <- vectors %>% 
-  right_join(firms) %>% 
-  right_join(firms_lifetime) %>% 
-  filter(name != "", lifetime >= 3) %>% 
-  drop_na(region) %>% 
-  distinct(name, tin, .keep_all = TRUE) %>% 
-  count(region, sort = TRUE)
-quantile(regions_cnt$n, .05)
 region_vectors <- vectors %>% 
   right_join(firms) %>% 
   right_join(firms_lifetime) %>% 
@@ -123,22 +116,51 @@ region_vectors <- vectors %>%
   summarise(across(dim_0:dim_255, mean), cnt = n()) %>% 
   filter(cnt > quantile(cnt, .05)) %>% 
   select(-cnt)
-region_distances <- as_tibble(cbind(
+
+sf_use_s2(FALSE)
+neighboring_regions <- as_tibble(cbind(
+  iso = ru$shapeISO,
+  as.data.frame(st_intersects(ru, ru, sparse = FALSE, remove_self = TRUE))
+))
+colnames(neighboring_regions) <- c("iso", ru$shapeISO)
+neighboring_regions <- pivot_longer(
+  neighboring_regions,
+  -iso,
+  names_to = "iso_2",
+  values_to = "is_neighbor"
+)
+
+centroids <- st_centroid(ru) %>% select(iso = shapeISO)
+distances <- st_distance(centroids)
+units(distances) <- "km"
+colnames(distances) <- centroids$iso
+geo_distances <- pivot_longer(
+  cbind(iso = centroids$iso, as.data.frame(distances)),
+  cols = -iso, 
+  names_to = "iso_2", 
+  values_to = "geo_distance"
+)
+
+region_names_distances <- as_tibble(cbind(
   select(region_vectors, region),
   as.matrix(dist(select(region_vectors, -region), diag = FALSE))
 ))
-region_distances[upper.tri(region_distances, diag = FALSE)] <- NA 
-colnames(region_distances) <- c("region", pull(select(region_vectors, region)))
-region_distances <- region_distances %>% 
[…]

### 2024-07-17 law-firms-names/Paper.Rmd
Add CSL, write paragraph about geography of services and importance of legal services, #4

@@
   word_document:
-    reference_doc: journal-article/reference.docx
-#bibliography: refs.bib
-#csl: style.csl
+    reference_doc: "journal-article/reference.docx"
+bibliography: "journal-article/refs.bib"
+csl: "journal-article/apa.csl"
 link-bibliography: False  
@@
 
+Geography of services, sometimes referred to as geography of service sector, is one of the essential parts of socio-economic geographical studies in Russia.
+Its history spans for several decades starting from the 1960s.
+Geography of services has economic and social dimensions, where the former is concentrated on the business aspects of the services, while the latter is focused on the issues of human well-being arising from the spatial structure of the relevant services [@Tkachenko2016].
+Existing scientific literature on the geography of services in Russia covers a broad range of topics, e.g.
+education [@Shulgina2019], car rental [@Zyrianova2018], catering [@Preobrazhenskiy2018], medical care [@Kazantcev2022; @Panin2023; @Nikoloski2024], internet access [@Nagirnaya2015; @Zemtsov2022].
+However, research on some problems is limited, with legal services being one of the notable examples of such a problem [@Tkachenko2016; @Moiseeva2016].
+Despite the lack of attention from geographers, spatial aspects of law firms activity worth investigating for the several reasons.
+First, legal services can be categorized as business and producer services, acting as a representative case, and this category of services is important for the economic development in general, both in national or regional scale [@Hansen1993; @Faulconbridge2019].
+Next, despite their relatively small direct economic impact, they have a great indirect importance as the intermediaries and helpers that contribute to the rule of law, social order, stability and predictability of social relations.
+Third, for regular citizens access to legal aid is important for the protection of human rights and ensuring the social justice, and in this context, the spatial aspects of law services are on par with the spatial issues of medical care [@Blacksell1988; @Blacksell1990].
+
 # 2. Materials and methods

### 2024-07-17 law-firms-names/Paper.Rmd
Write a paragraph about firm names and their connection with geography, #4

@@
 
+Geographical research of services may focus on a variety of subjects.
+Some classical pathways of such research include analysis of centers and systems of service or of consumers' behavior, needs and demands, including estimation of influencing factors, investigation on the actual level of services provision, applied scientific work related to the management of social infrastructure [@Tkachenko2016].
+Studies of firm names, sometimes also referred to as company names, brand names or just brands, is relatively unusual yet potentially fruitful method in the geography of services.
+Firm names are designed to distinguish the business from the competitors, to make an emphasis on the field of activity or its geographical area, inform about services supplied by a provider and corresponding benefits.
+Having these functions, brands convey a substantial portion of information in a short text message.
+Such information may be useful to discover the city's business landscape and consumer's perception of a particular city with its characteristic services [@Raento2001], analyze the trends common to a specific industry in a region [@Marjavaara2022] or the role of place as reflected in the toponyms in the business development and consumer attitudes [@Peluso2012; @Zhang2014; @Erdogan2016; @Noorda2019].
+Company names represent the characteristics of businesses and thus may act as “meta-proxy” for a bulk of intertwined economic metrics.
+Such a complex nature makes them suitable for regions identification, especially in the case of vernacular regions [@Reed1976; @Liesch2015; @Vukosav2016; @Holtkamp2018].
+Other interactions between places and brands, e.g.
+the contribution of commercial branding to place branding [@Florek2013] or its compliance with the latter [@BassolsiGardella2023] may also be of interest.
+
 # 2. Materials and methods

### 2024-07-17 law-firms-names/Paper.Rmd
Write paragraphs about the object, scope and goals of the research, #4

@@
 
+There are thousands of law firms in Russia [@Moiseeva2016].
+Their names are likely to reflect some properties of local legal services market.
+The area of Russia is comprised of regions (constituent entities of the federation) that provide some sort of a commonly used spatial unit for geographical studies.
+Besides, economic geographers split the country into 11 economic regions [@Leyzerovich2010].
+Each economic regions includes several constituent entities and is characterized by internal integrity, economic links between the companies, specialization and difference from the others.
+It can be hypothesized that the proximity of regions, as well as internal unity and economic links within the economic regions, leads to the similarity of law firms activity and, correspondingly, to the similarity of their naming.
+The overall aim of this study is to check whether this is true.
+To achieve this goal, a few research questions are asked and then answered based on the data on law firms names.
+These questions, or problems, are specified below.
+
+Question 1 (Q1) is whether the neighboring regions have more similar law firm names that regions that do not share a common border.
+Question 2 (Q2) is a continuous extension of Q1: whether the geographical proximity of regions has an impact on the similarity of law firm names, i.e.
+whether the regions that are close to each other have more similar names, and vice versa.
+Question 3 (Q3) moves up from the level of regions to the level of economic regions: whether names withing the economic region are more similar compared to the names outside the region.
+Question 4 (Q4) is what naming strategies exist in Russia and how they are spatially organized.
+Question 5 (Q5) is a follow-up from the previous: are there strategies common or different within each economic region?
+
 # 2. Materials and methods

### 2024-07-18 ru-smb-data-in-economic-geography/Paper.Rmd
Move all files related to paper about FTS open data to a separate folder, #5

--- 2024-01-08 Paper.Rmd
@@
 output:
-  odt_document:
-    number_sections: true
-    reference_odt: assets/reference.odt
+  word_document:
+    reference_doc: assets/reference.docx
 bibliography: assets/biblio_ru.bib
-csl: "assets/citations.csl"
+csl: "assets/citations-ru.csl"
 link-bibliography: false
@@
   echo = FALSE, 
-  fig.width = 6.5, 
+  fig.width = 6.7, 
   fig.height = 5,
@@
 ```{r imports, cache = TRUE, include = FALSE}
-source("Plots.R")
-source("StatsHelper.R")
-source("Diagram.R")
+source("helpers/Plots.R")
+source("helpers/Stats.R")
+source("helpers/Diagram.R")
 ```
 
-УДК 911.3, 519.688
+Научная статья
+
+УДК 911.3:33+004.42+004.6
 
@@
 
-**Использование открытых данных налоговой службы для создания геопривязанного набора данных о субъектах малого и среднего предпринимательства в России**
+<div custom-style="Title">
+
+Применение открытых данных налоговой службы для экономико-географического анализа
+
+</div>
 
@@
 
-Сёмин Павел Олегович
+***Аннотация.*** В статье представлена технология для создания геопривязанного табличного набора данных о субъектах малого и среднего предпринимательства (МСП) в России на основе открытых данных Федеральной налоговой службы (ФНС России).
+Получаемый набор данных охватывает всю территорию России за 2016–2023 гг.
+Он содержит детализированную до уровня конкретных юридических лиц и индивидуальных предпринимателей информацию о регистрационных номерах, категории, имени или наименовании, месте жительства или местонахождении, виде деятельности, доходах, расходах и численности работников.
+Статья излагает воспроизводимую методику обработки исходных данных и демонстрирует её применение для получения сведений о фирмах в сфере сельского, лесного хозяйства и рыболовства.
+Показано, что открытые данные ФНС России с помощью предложенного алгоритма и его реализации на языке программирования Python можно использовать для решения ряда исследовательских задач в области экономической географии, в особенности тех, где требуются дезагрегированные сведения или пространственная детализация до населённых пунктов.
+Благодаря географическим координатам получаемый набор данных может без дополнительной обработки использоваться для картографирования, а наличие кодов ОКТМО позволяет объединять его с официальными статистическими сведениями.
+
+***Ключевые слова:*** открытые данные, пространственный анализ, геоданные, административные данные, малое и среднее предпринимательство, реестр организаций, Apache Spark, ФНС России
 
@@
 
-Пермский государственный национальный исследовательский университет, Россия, Пермский край, г. Пермь, ул. Букирева, 15, 614068
+<div custom-style="Title">
+
+Application of Tax Service Open Data for Analysis in Economic Geography
+
+</div>
 
@@
 
-Сёмин Павел Олегович: ntsp\@ya.ru, ORCID 0000-0002-4015-9206
+***Abstract.*** The paper features a technology for creating a geo-referenced tabular dataset of small and medium-sized businesses (SMBs) in Russia based on open data from the Federal Tax Service (FTS of Russia).
+The resulting dataset covers the entire territory of Russia for 2016—2023.
+It contains information on registration numbers, category, name, address, activity type, income, expenses, and number of employees.
+The information is detailed up to the specific legal entities or individual entrepreneurs.
+The article outlines a reproducible methodology for raw data processing and demonstrates its application to obtain a dataset of firms in the agriculture, forestry and fisheries sectors.
+The paper shows that with the help of proposed technique and its reference implementation on Python programming language, FTS open data can be used to solve various tasks in economical geography, in particular tasks where the disaggregated information or spatial resolution up to settlements is required.
+In addition, thanks to the geographic coordinates, the generated dataset can be used for mapping without additional processing, and the presence of OKTMO codes allows it to be combined with official statistical information.
+
+***Keywords:*** open data, spatial analysis, spatial data, administrative data, small&medium business, business registry, Apache Spark, FTS of Russia
 
 <br>
-
-Контактное лицо:
-
-Сёмин Павел Олегович
-
-Телефон +7 9222 06 77 01
-
-Email: ntsp\@ya.ru
-
-Адрес: Россия, 622002, Свердловская обл., г. Нижний Тагил, ул. Космонавтов, д. 35, кв. 62
-
-# Аннотация {-}
-
-В статье представлена технология для создания геопривязанного табличного набора данных о субъектах малого и среднего предпринимательства (МСП) в России на основе открытых данных Федеральной налоговой службы (ФНС России).
-Получаемый набор данных охватывает всю территорию России за 2016–2023 гг.
-Он представляет собой полный список субъектов МСП и содержит информацию о регистрационных номерах, категории, имени или наименовании, месте жительства или местонахождении, виде деятельности, доходах, расходах и численности работников.
-В пространственном отношении набор детализирован до населённых пунктов с указанием географических координат и кода муниципального образования (ОКТМО); во временном отношении — до месяцев или лет в зависимости от конкретного признака; в отношении вида деятельности — до отдельных кодов общероссийского классификатора (ОКВЭД).
-В отличие от классической data paper, статья не описывает итоговые данные как таковые, а излагает воспроизводимую методику их получения и демонстрирует её использование для получения набора данных о фирмах в сфере сельского, лесного хозяйства и рыболовства.
-С помощью предложенного алгоритма и его реализации на языке Python исследователь может самостоятельно обработать исходные данные и получить актуальную версию набора данных, адаптированную для его задач.
-Алгоритм оптимизирован так, чтобы для его выполнения было достаточно компьютера средней конфигурации, несмотря на сравнительно большой (около 2 Тб в несжатом виде) объём исходных данных.
-Получаемый с помощью технологии набор данных может применяться как непосредственно для пространственного анализа МСП в России, так и в качестве открытого, бесплатного и относительно репрезентативного источника сведений об экономике России в разрезе конкретных организаций, позволяя отчасти заменить коммерческие системы бизнес-аналитики наподобие Spark и Контур.Фокус.
-Кроме того, благодаря координатам набор данных может без дополнительной обработки использоваться для картографирования, а наличие кодов ОКТМО позволяет объединять его с официальными статистическими сведениями.
-
-**Ключевые слова:** открытые данные, пространственный анализ, геоданные, административные данные, малое и среднее предпринимательство, реестр организаций, набор данных, big data, Apache Spark, ФНС России
 
@@
 Современные общественно-географические исследования часто основаны на работе с данными.
-Некоторые специалисты используют агрегированную экономическую и демографическую статистику.
-Так, О.
-Л.
-Виноградова, анализируя изменение интенсивности земледелия в Европе, опирается на показатели стран в разные годы [@Vinogradova2020]; Т.
-А.
-Балина с соавторам, рассматривая вопрос о территориях опережающего социально-экономического развития, приводит подсчёты по субъектам Российской Федерации [@Balina2021]; Г.
-М.
-Лаппо, описывая свою концепцию разнообразия городов, оперирует сведениями о числе жителей в населённых пунктах в разные периоды времени [@Lappo2019].
-Другие исследователи используют большие массивы дезагрегированных данных, собирая их вручную в общедоступных источниках и используя сложные методы их обработки.
-Например, А.
-И.
-и М.
-В.
-Красновы для своей работы, посвящённой трансформации сети расселения в Псковской области, собирали публичные сведения о местоположении пользователей социальной сети «ВКонтакте» и открытую информацию о поисковых запросах к «Яндексу» [@Krasnov2020].
-Есть работы, опирающиеся на обобщённую статистику, но использующие продвинутые методы её анализа.
-В частности, А.
-В.
-Федорченко взял данные Международной организации автопроизводителей, рассчитал производные индексы и дополнил их картографической визуализацией, что позволило эффективно проанализировать территориальные сдвиги в мировой автомобильной промышленности [@Fedorchenko2017].
-«Большие данные» также проникают в общественно-географическую науку, что отметили ещё десять лет назад М.
-Грэм и Т.
-Шелтон [@Graham2013].
-Недавний пример такого рода исследований — работа Вэй Лю и Вэньцзэ У, которые используют сведения о пользователях социальной сети WeChat для изучения связей между типом землепользования и деятельностью людей [@Liu2020].
-Таким образом, общая тенденция такова: географы используют данные для исследований, хотя объём, тип, качество этих данных и технологии их обработки отличаются.
+К таким данным относится, например, агрегированная (сводная) социально-экономическая статистика [@Fedorchenko2017; @Vinogradova2020; @Balina2021], общедоступные сведения социальных сетей и поисковых систем [@Krasnov2020; @Liu2020], геоданные сотовых операторов [@Babkin2020; @Babkin2021; @Vorobyev2021], информация коммерческих систем бизнес-аналитики [@Moiseeva2016; @Kuzminov2021] или геоинформационных сервисов [@Lachininsky2021], спутниковые снимки [@Lachininsky2023].
+По объёму, разнообразию и скорости обновления некоторые источники информации приближаются к характеристикам «больших данных» [@Graham2013], использование которых для решения традиционных общественно-географических задач — потенциально перспективное направление [@Blanutsa2017].
 
-Прежде чем использовать данные, необходимо их получить.
-Поиск качественных данных в достаточном количестве и их предварительная обработка во многом определяют результативность научной деятельности: проблемы в данных или в их анализе приводят к неправильным выводам, а сделать исследование при нехватке данных зачастую невозможно.
-Следовательно, источники данных и информационные технологии — это важный прикладной аспект общественной географии как науки.
-В.
-Н.
-Лаженцев, например, отмечает, что «совершенствование статистической науки в значительной мере предопределяет уровень экономико-географических исследований» [@Lazhentsev2015], а К.
-Брунсдон рассуждает о важности изложения точных деталей анализа данных в контексте воспроизводимости географических исследований [@Brunsdon2016].
-Практическая важность данных для исследований подтверждается, в частности, различными инициативами, связанными с обеспечением доступа к таким данным.
-Так, П.
-Арцбергер с соавторами ещё 20 лет назад предложили следующий принцип: результаты исследований, финансируемых на публичном уровне, должны быть максимально общедоступными, поскольку это, в частности, способствует новым исследованиям и тем самым увеличивает научный потенциал и отдачу от инвестиций в науку [@Arzberger2006], а Д.
-Кард с коллегами указывают на важность доступа учёных к административным данным, то есть информации, которая накапливается органами исполнительной власти при осуществлении их функций [@Card2010].
-Для географических исследований особую роль играет такая разновидность данных, как пространственные данные, обеспечение доступа к которым в недавние годы также вызывает интерес во многих странах, включая США, Великобританию, государства ЕС, Сингапур [@Jozefowicz2020].
+Для некоторых экономико-географических важны сведения, детализированные до уровня отдельных фирм.
+К таким сведениям относятся, например, местонахождение, выручка, расходы, число работников, наименование и другие характеристики.
+Указанные данные могут использоваться напрямую, как в работах К.В.
+Ростислава [@Rostislav2021], Е.А.
+Коломак и А.И.
+Шерубнёвой [@Kolomak2023], которые анализирует агломерационные эффекты на основе показателей предприятий.
+Кроме того, на их основе можно считать статистику с произвольной пространственной или временной группировкой.
+Например, И.С.
+Гуменюк оперирует подсчётами абсолютного и относительного числа малых и средних предприятий по муниципалитетам Калининградской области [@Gumenyuk2022]; М.А.
+Саранча включает количество предприятий в интегральный индекс уровня развития малого предпринимательства на территории муниципального образования [@Sarancha2014]; а М.А.
+Макушин с соавторами использует данные о предприятиях в городах БАМа, чтобы оценить структуру промышленности [@Makushin2023].
+Дополнительную важность сведения о конкретных фирмах приобретают при исследованиях в масштабе городов [@Lachininsky2021; @Nikonorov2021; @Kozhevnikov2023], при изучении узких секторов экономики, которые не фигурируют обособленно в статистике [@Moiseeva2016; @Pankratov2023], при анализе отдельно взятой отрасли [@Kuzminov2021].
+
+Удобного источника геопривязанной информации о фирмах в России нет.
+Исследователи и прикладные специалисты обычно опираются на данные коммерческих (2ГИС, Яндекс-карты) [@Lachininsky2021] или открытых (OpenStreetMap, Wikimapia) геоинформационных сервисов, систем бизнес-аналитики (например, Spark, Контур.Фокус и Ruslana) [@Moiseeva2016; @Kuzminov2021].
+Как правило, коммерческие фирмы предоставляют сведения за существенную плату, а для выгрузки и обработки информации открытых ресурсов требуется приложить немало усилий.
+Формально полный список российских организаций и индивидуальных предпринимателей содержится в двух государственных реестрах (соответственно Едином государственном реестре юридических лиц и Едином государственном реестре индивидуальных предпринимателей), однако ни один из них не распространяется в полном виде — можно лишь извлекать отдельные записи вручную через веб-сервис ФНС России либо автоматизированно за плату через программный интерфейс.
+Таким образом, актуален вопрос об удобном инструменте, который позволил бы бесплатно и легально получить максимально полный набор геопривязанных данных о российских фирмах.
+
+С недавних пор экономико-географы обратили внимание на такой источник информации, как административные данные ФНС России. 
+Для исследований используются, в частности, сведения о доходах физических лиц [@Petrov2021], о налоговых поступлениях в разрезе субъектов федерации или видов экономической деятельности [@Rostislav2020; @Safronov2021], о среднесписочной численности работников организаций [@Rostislav2021], о количестве учтённых жилых помещений [@Lachininsky2021], о количестве малых и средних предприятий и мерах их поддержки [@Gumenyuk2022].
+Особое внимание привлекает реестр субъектов малого и среднего предпринимательства (реестр МСП), который регулярно и полностью размещается на сайте службы и содержит сведения о конкретных организациях и индивидуальных предпринимателях, включая адреса регистрации.
+Де-факто это наиболее полный из открытых наборов сведений о российских фирмах, детализированный до уровня отдельных фирм и геопривязанный хотя бы каким-то образом.
+Использовать его в исходном виде — непростая задача: выгрузки реестра за несколько лет имеют суммарный сжатый объём 200 Гб и характеризуются сложной многоуровневой структурой, а адреса, записанные в привычном для человека виде, не содержат географических координат или кодов муниципальных образований, в результате чего малопригодны для автоматизированного анализа.
+Цель работы, проведённой при подготовке данной статьи — упростить использование данных реестра МСП для широкого круга экономико-географов.
 
@@
 В качестве основы используется реестр субъектов малого и среднего предпринимательства (реестр МСП), размещённый в форме открытых данных на сайте Федеральной налоговой службы России (ФНС России).
-Набор данных охватывает всю территорию России, все виды экономической деятельности и период с середины 2016 г.
-по настоящее время.
-Его пространственное разрешение — до конкретного населённого пункта, временное — до месяца, структурное — до конкретного кода по общероссийскому классификатору видов экономической деятельности (ОКВЭД).
-Сведения являются дезагрегированными, то есть детализированными до уровня отдельных фирм (организаций и индивидуальных предпринимателей).
-Помимо структурированных адресов, набор содержит географические координаты и коды ОКТМО муниципальных образований.
 В отношении субъектов малого и среднего предпринимательства (МСП) набор данных представляет собой генеральную совокупность.
 Кроме того, в статье обосновывается, что указанный набор данных может использоваться как достаточно репрезентативная выборка для анализа экономики в целом, а не только малого и среднего предпринимательства.
+Статья сочетает в себе элементы методической работы, data paper и software paper: она подробно описывает исходные данные и воспроизводимый алгоритм (методику) работы с ними, в том числе содержит ссылку на код программы, после чего демонстрирует результат обработки данных на тестовом примере, включая потенциальный вариант разведочного анализа.
 
-Статья, помимо введения, включает в себя несколько разделов.
-В обзоре литературы приводится краткая информация об использовании различных источников данных в общественно-географических исследованиях и через это обосновывается необходимость предложенной технологии.
-Раздел «Исходные данные» содержит описание первичных наборов данных, с которыми работает алгоритм, а раздел «Методика обработки данных» излагает саму технологию.
-В разделе «Демонстрация работы технологии» описываются и визуализируются процесс и результаты опробования технологии в отношении сведений о фирмах, ведущих деятельность в сфере сельского и лесного хозяйства, охоты и рыболовства.
-В «Обсуждении» затрагиваются ключевые спорные моменты касательно предложенной технологии и анализируется репрезентативность получаемого с её помощью набора данных.
-«Выводы» подводят итоги работы.
-
-# Обзор литературы: источники данных в общественной географии
-
-Типичным источником данных для исследований общественной географии России являются ресурсы Федеральной службы государственной статистики (Росстата).
-К ним относятся, в частности, размещённые на официальном сайте этого органа данные официальной статистики, переписей и статистических обследований, база данных показателей муниципальных образований (БДМО) и единая межведомственная информационно-статистическая система (ЕМИСС).
-Примеров их использования в географии множество: О.
-В.
-Кузнецова при исследовании структуры экономики российских регионов использует статистику валового регионального продукта и объёма отгруженных товаров [@Kuznetsova2018]; Г.
-М.
-Фёдоров в своём анализе экономики регионов России на Балтике опирается на данные Росстата и Федеральной таможенной службы [@Fedorov2022]; С.
-Г.
-Сафронов изучает третичный сектор экономики России на базе данных Eurostat, Росстата и Федеральной налоговой службы [@Safronov2021]; Ю.
-В.
-Петров с помощью региональной, муниципальной статистики и данных иных органов власти формулирует предложения по социально-экономическому развитию юга Тюменской области [@Petrov2021]; М.
-А.
-Саранча оценивает развитие гостиничного и ресторанного бизнеса, задействуя результаты сплошного наблюдения Росстата за малым и среднем предпринимательством [@Sarancha2014].
-Сведения Росстата также используются для создания производных наборов данных, как в работе И.
-Ю.
-Пекер [@Peker2022].
-Таким образом, большинство общественно-географических исследований не обходятся без применения данных Росстата как единственного источника информации или как одного из источников.
-
-Тем не менее, в ряде случаев данных Росстата недостаточно.
-Как следствие, возникает необходимость в альтернативных источниках информации, которые либо дополняют официальную статистику, либо заменяют её.
-Причины этого бывают разные.
-В некоторых случаях альтернативные источники информации используются для проверки или корректировки Росстата.
-Например, Р.
-А.
-Бабкин, основываясь на данных сотовых операторов, доказывает, что официальная численность населения московской агломерации занижена примерно на 10% из-за погрешности и инертности статистического учёта и «двудомности» значительной части москвичей [@Babkin2020].
-Другая, более частая причина — отсутствие нужных сведений.
-Например, К.
-В.
-Ростислав, анализируя влияние местоположения на прибыльность предприятий, учитывает только организации с численностью работников меньше 250 человек и для проверки этого условия использует открытые данные ФНС России, так как Росстат такие сведения не предоставляет [@Rostislav2021].
-С.
-С.
-Лачининский с соавторами оценивает спрол городских территорий Санкт-Петербурга с опорой на спутниковые снимки Landsat, так как на базе сведений официальной статистики провести необходимые подсчёты невозможно [@Lachininsky2023].
-К преимуществам альтернативных источников данных относят повышенную точность и своевременность [@Vorobyev2021], более высокое пространственное и временное разрешение [@Babkin2021], возможность точного и детального анализа на микроуровне [@Bochkaryov2017], отмечая при этом несвоевременность, неполноту и потенциальную недостоверность официальной статистики [@Kuzminov2021].
-Альтернативные источники данных дают доступ к генеральной совокупности, вследствие чего «отпадает необходимость в использовании выборочных, агрегированных и косвенных» источников сведений [@Blanutsa2017].
-При использовании методов машинного обучения, что также становится актуальным в современной географии, важны объём выборки, дезагрегированность, количество признаков в данных, что также побуждает обращаться к альтернативным источникам [@Rostislav2020].
-Таким образом, хотя традиционные статистические сведения по-прежнему важны в общественно-географических исследованиях, а порой незаменимы, есть нарастающая потребность в других информационных ресурсах.
-Запрос со стороны исследователей идёт на повышение объёма, оперативности предоставления, точности, пространственного и временного разрешения, расширение набора доступных показателей, а также на возможность валидации статистических сведений.
-Он проявляется и в части традиционных научных методов, так и методов, связанных с машинным обучением, хотя более актуален именно для последних, наиболее чувствительных к объёму данных, их дезагрегированности, детализации, удобным форматам представления.
-
-Одним из альтернативных источников информации для географов-обществоведов являются административные данные ФНС России.
-Они публикуются на официальном сайте этого органа власти или связанных с ним ресурсах.
-В географических работах используются, в частности, сведения о доходах физических лиц [@Petrov2021], о налоговых поступлениях в разрезе субъектов федерации или видов экономической деятельности [@Rostislav2020; @Safronov2021], о среднесписочной численности работников организаций [@Rostislav2021], о количестве учтённых жилых помещений [@Lachininsky2021], о количестве малых и средних предприятий и мерах их поддержки [@Gumenyuk2022].
-Внимание исследователей к данным, публикуемым ФНС России, связано, по-видимому, с многофункциональностью этого органа власти, из-за которой он собирает огромный объём информации о разных сторонах российской экономики: хозяйствующих субъектах, имуществе, доходах, расходах.
-Обычно ФНС России, как и Росстат, публикует сводную (агрегированную) информацию.
-Тем не менее, есть несколько наборов данных, содержащих дезагрегированные (исходные, первичные, «сырые») сведения.
-Наибольшее внимание среди них привлекает реестр субъектов малого и среднего предпринимательства (реестр МСП), который регулярно и полностью размещается на сайте службы и содержит сведения о конкретных организациях и индивидуальных предпринимателях.
-
-Для некоторых экономико-географических и смежных исследований важны сведения, детализированные до уровня конкретных фирм.
-К таким сведениям относятся, например, местонахождение, выручка, расходы, число работников, наименование и другие характеристики.
-Указанные данные могут использоваться напрямую, как в работах К.
-В.
-Ростислава [@Rostislav2021], Е.
-А.
-Коломак и А.
-И.
-Шерубнёвой [@Kolomak2023], которые анализирует агломерационные эффекты на основе показателей предприятий.
-Кроме того, на их основе можно считать статистику с произвольной пространственной или временной группировкой, в том числе такой, которую невозможно, сложно или неудобно извлекать из данных Росстата.
-Например, И.
-С.
-Гуменюк оперирует подсчётами абсолютного и относительного числа малых и средних предприятий по муниципалитетам Калининградской области [@Gumenyuk2022]; М.
-А.
-Саранча включает количество предприятий в интегральный индекс уровня развития малого предпринимательства на территории муниципального образования [@Sarancha2014]; а М.
-А.
-Макушин с соавторами использует данные о предприятиях в городах БАМа, чтобы оценить структуру промышленности [@Makushin2023].
-Дополнительную важность сведения о конкретных фирмах приобретают при исследованиях в масштабе городов [@Lachininsky2021; @Nikonorov2021; @Kozhevnikov2023], при изучении узких секторов экономики, которые не фигурируют обособленно в статистике [@Moiseeva2016; @Pankratov2023], при анализе отдельно взятой отрасли [@Kuzminov2021].
-
-Общепринятого источника геопривязанной информации о фирмах в России нет.
-Исследователи и прикладные специалисты обычно опираются на данные коммерческих [@Lachininsky2021] или открытых геоинформационных сервисов, систем бизнес-аналитики [@Moiseeva2016; @Kuzminov2021], а также упомянутого выше реестра МСП [@Gumenyuk2022].
-Как правило, использование этих способов сопряжено с двумя проблемами: данные предоставляются за плату либо требуется приложить значительные усилия для их приведения в подходящий для анализа формат.
-Системы бизнес-аналитики, такие как Spark, Контур.Фокус и Ruslana, содержат много разнообразной информации, но платные.
-Выгрузка данных из частных геоинформационных систем наподобие 2GIS и Яндекс.Карт также стоит денег, а кроме того, для неё нет удобного интерфейса.
-Открытые платформы наподобие OpenStreetMap или реестра МСП бесплатны, но выгрузки из них имеют большой объём и сложную структуру, поэтому исследователю приходится тратить много усилий на техническую сторону работы, привлекать сторонних специалистов или отказываться от использования такого источника данных.
-Формально полный список российских организаций и индивидуальных предпринимателей содержится в двух государственных реестрах (соответственно Едином государственном реестре юридических лиц и Едином государственном реестре индивидуальных предпринимателей), однако ни один из них не распространяется в полном виде — можно лишь извлекать отдельные записи вручную через веб-сервис ФНС России либо автоматизированно за плату через программный интерфейс, что не даёт доступ к генеральной совокупности компаний и вряд ли позволяет построить репрезентативную выборку без чрезмерных затрат и усилий.
-Таким образом, актуален вопрос об удобном инструменте, который позволил бы бесплатно и легально получить максимально полный набор геопривязанных данных о российских фирмах.
-
-# Исходные данные
+# Данные и методы их обработки
 
@@
 
-## Основной источник: реестр МСП
-
-Реестр МСП — это государственная информационная система, содержащая информацию о субъектах малого и среднего предпринимательства.
+**Реестр МСП как основной источник сведений.** Реестр МСП — это государственная информационная система, содержащая информацию о субъектах малого и среднего предпринимательства.
 Она предусмотрена ст.
@@
 Норма о реестре появилась в законе в конце 2015 г.
-На практике реестр размещён на сайте ФНС России в форме открытых данных в виде ежемесячных полных выгрузок, первая из которых датируется августом 2016 г., а последняя доступная на момент подготовки статьи — декабрём 2023 г.[^1]
-
-[^1]: https://www.nalog.gov.ru/opendata/7707329152-rsmp/
+На практике реестр размещён на сайте ФНС России в форме открытых данных в виде ежемесячных полных выгрузок, первая из которых датируется августом 2016 г., а последняя доступная на момент подготовки статьи — декабрём 2023 г.
+(https://www.nalog.gov.ru/opendata/7707329152-rsmp/).
 
@@
 
-На момент подготовки статьи (декабрь 2023 г.) было доступно `r raw_files_count` zip-архивов.
-Их суммарный объём — `r format(raw_files_size_gb, digits = 0)` Гб в сжатом виде и `r format (raw_files_uncompressed_size_tb, digits = 2)` Тб в распакованном.
+На момент подготовки статьи (декабрь 2023 г.) было доступно 88 zip-архивов.
+Их суммарный объём — 193 Гб в сжатом виде и 2.4 Тб в распакованном.
 Из-за большого объёма исходных данных требуется оптимизировать алгоритм их обработки и использовать специализированные инструменты для работы с «большими данными», о которых будет сказано ниже.
@@
 
-## Вспомогательные источники: доходы, расходы, число работников
-
-В качестве вспомогательных источников информации используются ещё два набора открытых данных ФНС России.
-
-1.
-Набор «Сведения о суммах доходов и расходов по данным бухгалтерской (финансовой) отчетности организаций»[^2].
-Содержит ИНН, суммарные доходы и расходы за год, предшествующий году публикации.
+**Вспомогательные источники.** В дополнение к реестру МСП используются ещё два набора открытых данных ФНС России.
+Первый — это «Сведения о суммах доходов и расходов по данным бухгалтерской (финансовой) отчетности организаций» (https://www.nalog.gov.ru/opendata/7707329152-revexp/).
+Он содержит ИНН, суммарные доходы и расходы за год, предшествующий году публикации.
 Обновляется несколько раз в год, но последующие обновления в течение одного года, по-видимому, представляют собой уточнение ранее опубликованных сведений, поэтому практическая периодичность обновления — это год.
-Данные доступны с 2019 г.
-по настоящее время[^3].
 В отличие от реестра МСП, содержит только информацию об организациях, но не об индивидуальных предпринимателях.
 Технически аналогичен реестру МСП: представляет собой временной ряд zip-архивов xml-файлов, где каждый архив соответствует определённой дате, а каждый xml-файл внутри архива содержит информацию о множестве организаций.
-
-[^2]: https://www.nalog.gov.ru/opendata/7707329152-revexp/
-
-[^3]: На момент подготовки статьи на сайте ФНС России была опубликована только последняя версия этого и следующего набора данных.
+Второй — «Сведения о среднесписочной численности работников организации» (https://www.nalog.gov.ru/opendata/7707329152-sshr2019/).
+Содержит ИНН и среднесписочную численность работников за год, предшествующий году публикации сведений.
+По остальным характеристикам идентичен предыдущему.
+Данные обоих наборов доступны с 2019 г.
+по настоящее время.
+На момент подготовки статьи на сайте ФНС России была опубликована только последняя версия этого и следующего набора данных.
 Раньше были доступны предыдущие.
@@
 
-2.
-Набор «Сведения о среднесписочной численности работников организации»[^4].
-Содержит ИНН и среднесписочную численность работников за год, предшествующий году публикации сведений.
-По остальным характеристикам идентичен предыдущему.
-
-[^4]: https://www.nalog.gov.ru/opendata/7707329152-sshr2019/
-
-## Геоданные
-
 Для нормализации адресов, расстановки координат и кодов ОКТМО используются два справочника.
-
-1.
-Набор данных «Населенные пункты России: численность населения и географические координаты»[^5].
+Один из них — «Населенные пункты России: численность населения и географические координаты» (https://data.rcsi.science/data-catalog/datasets/160/).
 Подготовлен проектом «Инфраструктура научно-исследовательских данных» на базе открытых данных Росстата и сведений региональных медицинских информационно-аналитических центров.
 Содержит перечень городских и сельских населённых пунктов с указанием субъекта федерации и муниципального образования, кода ОКТМО и географических координат.
-
-[^5]: https://data.rcsi.science/data-catalog/datasets/160/
-
-2.
-Набор данных «Города России»[^6].
+Второй — «Города России» (https://github.com/hflabs/city).
 Подготовлен компанией «Дадата» на базе собственных разработок.
@@
 
-[^6]: https://github.com/hflabs/city
+**Методика обработки данных.** Предложенный алгоритм обработки данных состоит из пяти шагов, некоторые из которых разбиты на подзадачи.
+Он изображён на схеме (рис.
+1) и описан ниже.
+Обозначения в скобках соответствуют обозначениям на схеме.
 
-# Методика обработки данных
+<br>
 
-Предложенный алгоритм обработки данных состоит из пяти шагов, некоторые из которых разбиты на подзадачи. Он изображён на рис. 1 и описан ниже.
+```{r fig-1-diagram, fig.cap = "Рис. 1. Схема обработки данных / Fig. 1. Data processing diagram"}
+grid::grid.raster(diagram_raster)
+```
 
-## Загрузка данных ФНС России (1)
+<br>
 
+Первый шаг — это загрузка данных ФНС России (1).
 Наборы открытых данных загружаются с сайта ФНС России на устройство пользователя.
@@
 
-## Фильтрующее извлечение данных реестра МСП (2a)
-
-Каждый zip-архив преобразуется в csv-таблицу.
+Второй шаг — фильтрующее извлечение данных (2a).
+Каждый zip-архив реестра МСП преобразуется в csv-таблицу.
 Каждая строка таблицы содержит сведения об одном субъекте МСП.
@@
 Отфильтрованная таблица будет содержать сведения только из необходимой части реестра МСП, что ускоряет работу и сокращает использование дискового пространства.
+Реализация данного шага оптимизирована, чтобы уменьшить время обработки, расход оперативной памяти и места для хранения временных файлов: данные из архивов читаются напрямую без распаковки архива целиком, извлекаются и сохраняются только необходимые атрибуты записей о фирмах.
+Аналогичным образом обрабатываются сведения о доходах, расходах (2b) и о среднесписочной численности работников организаций (2c).
 
-Реализация данного шага оптимизирована, чтобы уменьшить время обработки, расход оперативной памяти и места для хранения временных файлов.
-Наивный подход к преобразованию исходных архивов в таблицу выглядел бы следующим образом: извлечь все xml-файлы из архива, прочитать каждый из них полностью, преобразовать в таблицу, выбрать необходимые признаки и наблюдения.
-Он привёл бы к образованию большого объёма временных файлов (десятки гигабайт) и увеличению времени на их обработку (чтение и конвертация в таблицу осуществлялись бы для всех данных).
-Оптимизированный подход, реализованный автором, предполагает чтение отдельных xml-файлов из архива с распаковкой «на лету» (вместо распаковки всего архива), фильтрацию по коду ОКВЭД в процессе обработки xml-файла (вместо последующей фильтрации) и извлечение нужных признаков напрямую из xml-файла (вместо последующего отбора).
-Он в несколько раз быстрее и не образует временные распакованные файлы.
-
-## Извлечение данных о доходах и расходах (2b)
-
-Каждый zip-архив набора открытых данных о доходах и расходах преобразуется в csv-таблицу.
-Строка таблицы соответствует одному субъекту МСП.
-Извлекаются следующие признаки: ИНН, дата выгрузки сведений, доходы, расходы.
-Фильтрация не производится.
-Реализация оптимизирована: архивы читаются напрямую без распаковки на диск, из xml-файлов сразу берутся необходимые признаки.
-
-## Извлечение данных о среднесписочной численности работников (2c)
-
-Аналогичен шагу 2b, отличается входными данными (zip-архивы из набора о среднесписочной численности работников) и признаками в результирующей таблице (ИНН, дата выгрузки, среднесписочная численность работников).
-
-## Агрегация данных о субъектах МСП (3a)
-
+Третий шаг — агрегация данных о субъектах МСП (3a).
 Таблицы в csv-формате, получившиеся на шаге 2a, сводятся в одну таблицу, из которой удаляются дубли, то есть строки, в которых совпадает набор значимых признаков: ИНН, ОГРН, имя или наименование, вид, категория, код ОКВЭД, адрес регистрации.
@@
 Например, в реестре МСП тип населённого пункта может быть указан полностью («город») или сокращённо («г.»), и вспомогательная обработка перед агрегацией позволяет игнорировать такие различия и воспринимать адрес как неизменный.
+Аналогичным образом производится агрегация и фильтрация данных о доходах, расходах (3b) и агрегация данных о работниках (3c).
 
-## Агрегация и фильтрация данных о доходах, расходах (3b)
-
-Таблицы в csv-формате, получившиеся на шаге 2b, объединяются в одну так, что для каждой организации указаны доходы и расходы за каждый доступный год.
-Объединённая таблица фильтруется по ИНН так, что в ней остаются только те организации, которые есть в агрегированных данных из реестра МСП, полученных на шаге 3a.
-Результат сохраняется в итоговый csv-файл.
-
-## Агрегация данных о работниках (3c)
-
-Аналогичен шагу 3b, но на вход подаются таблицы, полученные на шаге 2c.
-
-## Геопривязка субъектов МСП (4)
-
+Четвёртый шаг — геопривязка субъектов МСП (4).
 Каждая запись в csv-таблице, полученной на этапе 3a, дополняется кодом ОКТМО и географическими координатами населённого пункта, в котором находится субъект МСП.
@@
 
-## Создание панельного представления (5)
-
+Пятый, опциональный шаг — создание панельного представления (5).
 Таблица, полученная на этапе 4, объединяется с таблицами, полученными на этапах 3b и 3c, и преобразуется в панельный временной ряд, где представлены характеристики каждого субъекта МСП за каждый год с 2016 по 2023.
@@
 
-## Реализация алгоритма
+**Характеристики итогового набора данных.** Получаемый набор данных охватывает всю территорию России за период с середины 2016 г.
+по настоящее время.
+В зависимости от настроек фильтрации он содержит информацию обо всех субъектах МСП или только о тех, которые ведут деятельность в выбранных сферах.
+Его пространственное разрешение — до конкретного населённого пункта, временное — до месяца, структурное — до конкретного кода по общероссийскому классификатору видов экономической деятельности (ОКВЭД).
+Сведения являются дезагрегированными, то есть детализированными до уровня отдельных фирм (организаций и индивидуальных предпринимателей).
+Помимо структурированных адресов, набор содержит географические координаты населённых пунктов, в которых находятся фирмы, и коды ОКТМО соответствующих муниципальных образований.
 
-Описанный выше алгоритм реализован как набор скриптов на языке программирования Python.
+**Технические детали реализации.** Описанный выше алгоритм реализован как набор скриптов на языке программирования Python.
 Чтение xml осуществляется с помощью высокоэффективной библиотеки lxml.
@@
 
-# Демонстрация работы технологии
-
-С помощью описанного алгоритма был подготовлен пробный набор данных о субъектах МСП, ведущих деятельность в сфере сельского, лесного хозяйства, охоты, рыболовства и рыбоводства.
+**Демонстрационный вариант.** С помощью описанного алгоритма был подготовлен пробный набор данных о субъектах МСП, ведущих деятельность в сфере сельского, лесного хозяйства, охоты, рыболовства и рыбоводства.
 Фильтрация вида деятельности производилась по следующему условию: основной код ОКВЭД организации или предпринимателя находится в группе А классификатора.
@@
 для сведений о среднесписочной численности работников организаций.
-Далее в подразделах представлена техническая статистика выполнения алгоритма и показаны базовые возможности по анализу полученного набора данных.
-Акцент в технической части делается на изменении количества обрабатываемых данных от шага к шагу алгоритма: благодаря этому видно, что предложенный метод помогает справиться с большим объёмом исходных сведений и превратить сотни гигабайт «сырых» данных в несколько таблиц, размер которых меньше на порядки.
-Аналитическая часть сфокусирована на картографической визуализации, поскольку картографирование — один из наиболее характерных географических методов.
-Она призвана показать пространственную, временную и структурную детализацию полученного набора данных, возможность вычислять на его основе статистические показатели с произвольной группировкой и отслеживать конкретные фирмы.
+Описание и визуализация результата приведены в следующем разделе статьи.
 
-## Техническая статистика работы алгоритма
+# Результаты и обсуждение
 
-Время создания пробного набора данных — около 15 часов. Больше всего занимает извлечение информации из архивов (шаг 2).
+**Техническая статистика работы алгоритма.** Время создания демонстрационного набора данных — около 15 часов.
+Больше всего занимает извлечение информации из архивов (шаг 2).
+Объём исходных данных — около 200 Гб.
+Объём итоговых таблиц — 60 Мб; панельного представления — 634 Мб.
+Итоговый набор данных содержит информацию примерно о `r round(counts[1] / 1000)` тысячах организаций и `r round(counts[2] / 1000)` тысячах индивидуальных предпринимателей.
 
-В таблице 1 приведена информация об уменьшении объёма данных в процессе работы алгоритма, а в таблице 2 — об уменьшении количества записей.
-Из таблиц видно, что технология многократно уменьшает объём обрабатываемых сведений: от сотен гигабайт исходных до нескольких десятков или сотен мегабайт итоговых наборов данных.
-Похожим образом уменьшается и количество записей.
-В результате вместо десятков больших архивов с исходными данными в многомерном xml-формате исследователь получает относительно небольшие двумерные таблицы.
-Панельное представление слегка уступает итоговым таблицам в компактности, но отличается более удобной формой представления временного ряда.
+**Распределение по субъектам федерации.** Нормализованные адреса позволяют агрегировать информацию по субъектам федерации.
+На рис.
+2 представлено количество субъектов МСП, которые вели деятельность в сфере лесного (слева) и сельского (справа) хозяйства в 2021 г.
+Распределение ожидаемо разное: наибольшее количество фирм в сфере сельского хозяйства расположено в Черноземье и на юге России, также заметны животноводческие регионы; лесное хозяйство сосредоточено на севере европейской части страны, в Восточной Сибири и на Дальнем Востоке.
 
-Итоговый набор данных содержит информацию примерно о `r round(out_org_count / 1000)` тысячах организаций и `r round(out_ind_count / 1000)` тысячах индивидуальных предпринимателей.
+<br>
 
-## Временная динамика
+```{r fig-2-regions-agri-forest, fig.cap = "Рис.
+2.
+Распределение количества субъектов МСП, ведущих деятельность в сфере сельского и лесного хозяйства, по регионам (субъектам федерации).
+Цвет — масштабированное количество (0 — минимум, 1 — максимум) / Fig.
+2.
+Regional distribution of SMBs which main activity is agriculture or forestry, in 2021.
+Color represents the min-max scaled value"}
+regions_plot_agri_forestry
+```
 
-Набор данных позволяет отследить необходимые количественные характеристики с точностью до месяца.
-На рис.
-2 показаны 15 субъектов России с наибольшим числом зарегистрированных субъектов МСП и изменение этого числа с августа 2016 г.
-по ноябрь 2023 г.
-
-## Распределение по субъектам федерации
-
-Нормализованные адреса позволяют агрегировать информацию по субъектам федерации.
-На рис.
-3 представлено количество субъектов МСП, которые вели деятельность в сфере лесного (слева) и сельского (справа) хозяйства в 2021 г.
-Распределение ожидаемо разное: наибольшее количество фирм в сфере сельского хозяйства расположено в Черноземье и на юге России, также заметны животноводческие регионы; лесное хозяйство сосредоточено на севере европейской части страны, в Восточной Сибири и на Дальнем Востоке.
+<br>
 
@@
 На рис.
-4 показано количество субъектов МСП, ведущих деятельность по выращиванию пшеницы (код 01.11.11) и картофеля (код 01.13.31) в 2021 г.
-Здесь заметны некоторые странности (например, в лидерах по картофелю оказалась Якутия), которые, видимо, связаны с небольшим количеством субъектов МСП с указанными кодами: обычно у фирм в качестве основного указан более общий код, в то время как для демонстрации картограмма строилась по максимально детализированным.
+3 показано количество субъектов МСП, ведущих деятельность по выращиванию пшеницы (код 01.11.11) и риса (код 01.12) в 2021 г.
+Можно заметить, что пространственное распределение фирм отличается и отчасти согласуется с районами выращивания соответствующих культур, хотя вряд ли полностью соответствует ему, так как номинальный код основного вида деятельности не всегда соответствует фактическому.
 
-## Распределение по населённым пунктам
+<br>
 
-Набор данных помимо названия субъекта содержит географические координаты, название населённого пункта и территориальной единицы (района, муниципалитета) внутри субъекта.
+```{r fig-3-regions-wheat-rice, fig.cap = "Рис.
+3.
+Распределение количества субъектов МСП, ведущих деятельность по выращиванию пшеницы и риса, по регионам (субъектам федерации).
+Цвет — масштабированное количество (0 — минимум, 1 — максимум).
+Серый цвет — отсутствие данных / Fig.
+3.
+Regional distribution of SMBs which main activity is growing of wheat or rice, in 2021.
+Color represents the min-max scaled value, gray is no data"}
+regions_plot_wheat_rice
+```
+
+<br>
+
+**Распределение по населённым пунктам.** Набор данных помимо названия субъекта содержит географические координаты, название населённого пункта и территориальной единицы (района, муниципалитета) внутри субъекта.
 Это позволяет выполнять более детализированные агрегации и показывать их на карте.
 На рис.
-5 показано количество субъектов МСП в населённых пунктах России в 2021 г, а также преимущественная специализация населённых пунктов исходя из того, каких фирм в нём больше всего.
-Показаны только населённые пункты более чем с 50 фирмами.
-Аналогичную картограмму можно построить и для региона: рис.
-6 изображает количество субъектов МСП по видам деятельности в населённых пунктах Свердловской области в 2021 г., причём без ограничения по числу фирм.
+4 показано количество субъектов МСП в населённых пунктах России в 2021 г, а также преимущественная специализация населённых пунктов исходя из того, каких фирм в нём больше всего.
 
-## Отслеживание отдельных фирм
+<br>
 
-Набор данных позволяет анализировать изменения у конкретных организаций и предпринимателей, например, «переезд» (изменение адреса регистрации) из одного региона в другой.
+```{r fig-4-activity-by-settlements-russia, fig.cap = "Рис.
+4.
+Число субъектов МСП по населённым пунктам России и преобладающий вид деятельности.
+Размер кругов соответствует числу субъектов МСП, цвет — одной из трёх групп видов деятельности по классификатору ОКВЭД (см.
+легенду) / Fig.
+4.
+Count of SMBs in some Russian settlements with dominant activity type, in 2021.
+The counts are represented by circle size, and the most popular activity group in the settlement is displayed by color"}
+activity_by_settlements
+```
+
+<br>
+
+**Отслеживание отдельных фирм.** Набор данных позволяет анализировать изменения у конкретных организаций и предпринимателей, например, «переезд» (изменение адреса регистрации) из одного региона в другой.
 На рис.
-7 показано изменение количества субъектов МСП из пробного набора, зарегистрированных в субъектах России, за весь доступный период.
+5 показано изменение количества субъектов МСП из пробного набора, зарегистрированных в субъектах России, за весь доступный период.
 
-## Анализ прибыли и числа работников
+<br>
 
-Вспомогательные данные о доходах, расходах и числе работников позволяют перейти от простого числа компаний к характеристикам их деятельности.
-На трёх картограммах (рис.
-8) показано соответственно общее число фирм, их суммарная прибыль и суммарное число работников по субъектам России.
-Видно, что распределение похожее, но всё-таки разное в зависимости от визуализируемой характеристики.
+```{r fig-5-migrations, fig.cap = "Рис.
+5.
+Распределение количества субъектов МСП из демонстрационного набора, прибывших или покинувших территорию региона (субъекта федерации) за 2016–2023 г.
+/ Fig.
+5.
+Regional distribution of SMBs which have arrived to or left the region during 2016–2023"}
+migrations_plot
+```
 
-# Обсуждение
+<br>
 
-Предложенный алгоритм имеет техническую ценность: он позволяет обработать большой массив данных, имеющих сложную структуру, и сделать на его основе несколько сравнительно компактных таблиц в формате, который традиционно используется исследователями.
+**Обсуждение.** Предложенный алгоритм имеет техническую ценность: он позволяет обработать большой массив данных, имеющих сложную структуру, и сделать на его основе несколько сравнительно компактных таблиц в формате, который традиционно используется исследователями.
 Кроме того, технология нормализует адреса, содержащиеся в исходных данных, и обогащает их географическими координатами и кодами ОКТМО, тем самым делая их пригодными для обработки в геоинформационных системах (ГИС) и пространственного анализа в целом, а также упрощая объединение с муниципальной статистикой.
@@
 Его можно использовать для сплошного исследования малого и среднего предпринимательства в стране.
-В некотором роде это квази-источник первичных статистических сведений, которые, по общему правилу, относятся к информации ограниченного доступа[^7], но в данном случае являются открытыми в силу прямого указания в Налоговом кодексе[^8].
+В некотором роде это квази-источник первичных статистических сведений, которые, по общему правилу, относятся к информации ограниченного доступа, но в данном случае являются открытыми в силу прямого указания в Налоговом кодексе.
 
@@
 Также следует отметить, что анализ сведений об индивидуальных предпринимателях может рассматриваться как обработка персональных данных физических лиц, что создаёт правовые риски, особенно актуальные в свете регулярного ужесточения юридической ответственности за нарушения в этой сфере.
-
-[^7]: Ч. 1 ст. 9 Федерального закона от 29.11.2007 № 282-ФЗ «Об официальном статистическом учете и системе государственной статистики в Российской Федерации».
-
-[^8]: П. 1.1 ст. 102 Налогового кодекса России.
 
@@
 Для оценки репрезентативности реестра МСП можно использовать корреляционный анализ.
-Единая межведомственная информационно-статистическая система (ЕМИСС) содержит данные о числе организаций в разрезе субъектов России и групп ОКВЭД за 2019–2022 гг[^9].
+Единая межведомственная информационно-статистическая система (ЕМИСС) содержит данные о числе организаций в разрезе субъектов России и групп ОКВЭД за 2019–2022 гг (https://fedstat.ru/indicator/58109).
 Такая же статистика рассчитывается на базе реестра МСП.
@@
 
-[^9]: https://fedstat.ru/indicator/58109
-
 Результаты анализа показывают высокую согласованность между числом организаций в целом и числом организаций в реестре МСП.
-Общий коэффициент корреляции в течение рассмотренных четырёх лет составляет чуть больше 0.8 и мало изменяется (таблица 3).
+Общий коэффициент корреляции в течение рассмотренных четырёх лет составляет чуть больше 0.8 и мало изменяется (таблица 1).
 Коэффициенты корреляции по регионам различаются довольно сильно (см.
@@
 9), но в основном составляют от 0.7 до 0.9.
-При анализе отдельных групп ОКВЭД коэффициент корреляции в большинстве случаев не меньше 0.9, а иногда приближается к единице (таблица 4).
+При анализе отдельных групп ОКВЭД коэффициент корреляции в большинстве случаев не меньше 0.9, а иногда приближается к единице (таблица 2).
 Лишь три группы характеризуются сравнительно небольшой корреляцией: госуправление (O), образование (P) и экстерриториальная деятельность (U).
@@
 Если повторить анализ по годам и регионам, исключив три наиболее «проблемные» группы, то общие коэффициенты корреляции по годам повысятся до 0.94–0.95, а по регионам сместятся в сторону 0.9–1.0.
-(таблица 3 и рис.
+(таблица 1 и рис.
 9).
+
+<br>
+
+## Таблица 1
+
+```{r tab-1-corr-by-year}
+knitr::kable(
+  corr_by_year,
+  digits = 2, 
+  col.names = c("Год", "$\\rho_{Spearman}$, все группы ОКВЭД", "$\\rho_{Spearman}$, без O, P, U"),
+  caption = "Корреляция между числом организаций в реестре МСП и общим числом организаций в каждом регионе(субъекте федерации) в каждой группе деятельности по годам / Correlation between the count of SMB organizations and total count of commercial organizations by year")
+```
+
+<br>
+
+```{r fig-6-corr-by-region, fig.cap = "Рис.
+6.
+Распределение регионов (субъектов федерации) по корреляции между числом организаций в реестре МСП и общим числом организаций в каждом году в каждой группе деятельности / Fig.
+6.
+Distribution of regions by the correlation coefficient between the SMB organizations counts and total commercial organizations counts"}
+corr_by_region_plot
+```
+
+<br>
+
+## Таблица 2
+
+```{r tab-2-corr_by_group}
+knitr::kable(
+  corr_by_group,
+  digits = 2,
+  col.names = c("Группа ОКВЭД", "$\\rho_{Spearman}$"),
+  caption = "Корреляция между числом организаций в реестре МСП и общим числом организаций в каждом регионе (субъекте федерации) и каждом году по группам деятельности / Correlation between the count of SMB organizations and total count of commercial organizations by activity group")
+```
+
+<br>
 
@@
 <div id="refs"></div>
-
-\newpage
-
-```{r fig-1-diagram, fig.cap = "Рис. 1. Схема обработки данных"}
-grid::grid.raster(diagram_raster)
-```
-
-<br>
-
-```{r tab-1-sizes-funnel}
-bytes_in_mb <- 2 ** 20
-sizes_funnel <-  sizes_funnel / bytes_in_mb
-knitr::kable(
-  sizes_funnel,
-  digits = 0,
-  format.args = list(scientific = FALSE),
-  col.names = c("Реестр МСП", "Доходы+расходы", "Работники", "Всего"),
-  caption = "Таблица 1. Изменение объёма данных в процессе работы алгоритма. Значения в мегабайтах (Мб). Строки — этапы работы алгоритма, столбцы — вид исходных данных и сумма")
-```
-
-<br>
-
-```{r tab-2-counts-funnel}
-counts_funnel <- counts_funnel / 1000 # in thousands
-knitr::kable(
-  counts_funnel,
-  digits = 0,
-  format.args = list(scientific = FALSE),
-  col.names = c("Реестр МСП", "Доходы+расходы", "Работники"),
-  caption = "Таблица 2. Изменение количества записей в процессе работы алгоритма. Значения в тысячах штук. Строки — этапы работы алгоритма, столбцы — вид исходных данных")
-```
-
-<br>
-
-```{r fig-2-time-series-plot, fig.cap = "Рис.
-2.
-Изменение числа субъектов МСП в демонстрационном наборе данных в 2016–2023 гг.
-с детализацией по месяцам.
-Показаны только первые 15 регионов (субъектов федерации) по количеству записей в соответствующем месяце"}
-time_series_plot
-```
-
-<br>
-
-```{r fig-3-regions-agri-forest, fig.cap = "Рис.
-3.
-Распределение количества субъектов МСП, ведущих деятельность в сфере сельского и лесного хозяйства, по регионам (субъектам федерации).
-Цвет — масштабированное количество (0 — минимум, 1 — максимум)"}
-regions_plot_agri_forestry
-```
-
-<br>
-
-```{r fig-3-regions-wheat-potato, fig.cap = "Рис.
-4.
-Распределение количества субъектов МСП, ведущих деятельность по выращиванию пшеницы и картофеля, по регионам (субъектам федерации).
-Цвет — масштабированное количество (0 — минимум, 1 — максимум).
-Серый цвет — отсутствие данных"}
-regions_plot_wheat_potato
-```
-
-<br>
-
-```{r fig-5-activity-by-settlements-russia, fig.cap = "Рис.
-5.
-Число субъектов МСП по населённым пунктам России и преобладающий вид деятельности.
-Размер кругов соответствует числу субъектов МСП, цвет — одной из трёх групп видов деятельности по классификатору ОКВЭД (см.
-легенду).
-Показаны только населённые пункты, в которых зарегистрировано более 50 фирм"}
-activity_by_settlements
-```
-
-<br>
-
-```{r fig-6-activity-by-settlements_svr, fig.cap = "Рис.
-6.
-Число  субъектов МСП в населённых пунктах Свердловской области и их преобладающий вид деятельности.
-Размер кругов соответствует числу субъектов МСП, цвет — одной из трёх групп видов деятельности по классификатору ОКВЭД (см.
-легенду).
-Показаны все населённые пункты без ограничения по количеству зарегистрированных фирм"}
-activity_by_settlements_svr
-```
-
-<br>
-
-```{r fig-7-migrations, fig.cap = "Рис.
-7.
-Распределение количества субъектов МСП из демонстрационного набора, прибывших или покинувших территорию региона (субъекта федерации) за 2016–2023 г.
-Абсолютные числа масштабированы так, что -1 — максимальная убыль, 1 — максимальный прирост"}
-migrations_plot
-```
-
-<br>
-
-```{r fig-8-count-employees-profit-by-region, fig.cap = "Рис.
-8.
-Распределение числа зарегистрированных субъектов МСП, их суммарной прибыли и суммарной численности работников в демонстрационном наборе в 2021 г.
-по регионам (субъектам федерации)"}
-count_profit_employees_by_region_plot
-```
-
-<br>
-
-```{r tab-3-corr-by-year}
-knitr::kable(
-  corr_by_year,
-  digits = 2, 
-  col.names = c("Год", "$\\rho_{Spearman}$, все группы ОКВЭД", "$\\rho_{Spearman}$, без O, P, U"),
-  caption = "Таблица 3. Корреляция между числом организаций в реестре МСП и общим числом организаций в каждом регионе(субъекте федерации) в каждой группе деятельности по годам")
-```
-
-<br>
-
-```{r fig-9-corr-by-region, fig.cap = "Рис.
-9.
-Распределение регионов (субъектов федерации) по корреляции между числом организаций в реестре МСП и общим числом организаций в каждом году в каждой группе деятельности"}
-corr_by_region_plot
-```
-
-<br>
-
-```{r tab-4-corr_by_group}
-knitr::kable(
-  corr_by_group,
-  digits = 2,
-  col.names = c("Группа ОКВЭД", "$\\rho_{Spearman}$"),
-  caption = "Таблица 4. Корреляция между числом организаций в реестре МСП и общим числом организаций в каждом регионе (субъекте федерации) и каждом году по группам деятельности")
-```

### 2024-07-18 law-firms-names/Paper.Rmd
Write paragraph about data, #5

@@
 
+The issue of obtaining information about firm names is challenging.
+Even though all Russian organisations are registered in the unified state registry operated by Federal Tax Service (FTS of Russia), and an extract about a particular company from this registry can be downloaded online by any person free of charge, there is no straightforward way to get names of all companies from the registry at once without paying a considerable amount of money and without negotiating a contract with the responsible government body.
+There are a few commercial firms that re-cell the data from the registry of companies, but the prices are substantial.
+To overcome these problems, a Unified state registry of small and medium-sized businesses (SMB registry) was used.
+This registry has been maintained since 2016 according to the Article 4.1 of Federal Law of July 24, 2007 No 209-FZ “On the development of small and medium-sized businesses in the Russian Federation”.
+All the companies and individual entrepreneurs that have a yearly revenue below 2 billion roubles and an average yearly number of employees below 250 are automatically included in the SMB registry without any action required on their side.
+In contrast to the main registries of companies and entrepreneurs, SMB registry can be fully downloaded at the FTS open data portal free of charge, as well as the monthly history of its updates since the beginning of the operation.
+Thus, this registry can be treated as the most comprehensive free data source with the information about Russian business entities (organisations and individuals).
+The downloadable version of the registry includes, in particular, firm names, addresses of registration (up to a settlement), and codes of activity kinds according to the All-Russia Classifier of Kinds of Economic Activity.
+Large firms are not present in the registry, so some data is missing.
+However, the majority of commercial law firms in Russia (more than 75%) can be categorized as small and medium-sized businesses and are included in the registry, so the availability of information is relatively high, and it can be used for trustworthy conclusions.
+
 ## 2.2. Methods

### 2024-07-18 law-firms-names/Paper.Rmd
Write a paragraph about FTS data processing abd a paragraph about text embeddings, #5

@@
 
+### 2.2.1. FTS data processing
+
+SMB registry dumps from FTS website have a large volume (≈200 Gb in total) and a complex structure: each dump is an archive of several thousand XML files, and each XML file contain 900 records with various information about organisations and individuals.
+Dumps are monthly, resulting in ≈90 archives for the period from 2016 to the present.
+If unpacked, all archives would occupy more than 2 Tb of disk space, and the extraction and preprocessing of such volume of data would take a large time.
+It should be noted, though, that the majority of information is redundant for the task in hand, because it is duplicated due to data publication specific (each archive contains a full copy of the registry, and if nothing has changed during the month, the data records will be merely the same) and because only the data about law firms is required, while the registry holds information about all the other firms as well.
+To extract necessary data from the raw archives, an open-source Python command-line tool was developed named ru-smb-companies.
+The main stages of data extractions are outlined in the next paragraph, while for the complete information about data processing workflow readers may refer to the source code of the program.
+
+FTS data processing took several steps.
+First, all source archives available at the time of research (November 2023) were downloaded  from FTS servers and stored both locally and in the cloud.
+Next, information about law firms was extracted from each archive to a CSV file.
+Extracted attributes include company or person name, address of incorporation (region, district, city, settlement), taxpayer ID, main activity code and some other properties.
+After this, CSV tables were combined into one table, deduplicated, and raw addresses were normalized and geocoded to fix the inconsistency in location names and add geographical coordinates.
+Finally, the resulting table was transformed in a panel view so that each row represents one business entity at a given year.
+The output table contains data about ≈ 30, 000 law firms for 2016–23.
+
+### 2.2.2. Firm names processing
+
+A major challenge regarding the methods of this study is the working with names.
+The objectives of the research imply that names should be grouped and compared in some way.
+This can easily be accomplished with numbers, but firm names are just short text sequences.
+Of course, a human can easily say whether two names are similar or not and classify them.
+However, the human-based approach does not scale well, and the quantity of names requires automation.
+There are several traditional techniques allowing us to compare text strings.
+For example, edit distance calculates the number of atomic changed required to turn one string to another, and the edit distance between “cat” and “car” is 1, while the edit distance between “car” and “karst” is 3.
+Another method named n-gram similarity splits the strings into subsequences, or tokens, and finds the number of common tokens, so that “law company” and “law bureau” are more close to each other than “law bureau” and “juridical center”.
+Despite their simplicity, they also have a significant disadvantage: they cannot account for the intrinsic variability and flexibility of language and do not consider the meaning of the words and phrases rather than their formal representation.
+For instance, a human understands that “law company” and “legal organisation” mean almost the same, while edit distance between these strings is high, and n-gram similarity is low.
+A modern method called text embeddings (sometimes referred to as vectors) can deal with this issue and therefore it is used for this research.
+
 # 3. Results and discussion

### 2024-07-19 ru-smb-data-in-economic-geography/Paper.Rmd
Make edits after review: draw stats by municipalities instead of settlements in Figure 4, use histogram instead of freqpoly in Figure 6, add activity group names in Table 2, replace 'federation' with 'Russian Federation', #5

@@
 С недавних пор экономико-географы обратили внимание на такой источник информации, как административные данные ФНС России. 
-Для исследований используются, в частности, сведения о доходах физических лиц [@Petrov2021], о налоговых поступлениях в разрезе субъектов федерации или видов экономической деятельности [@Rostislav2020; @Safronov2021], о среднесписочной численности работников организаций [@Rostislav2021], о количестве учтённых жилых помещений [@Lachininsky2021], о количестве малых и средних предприятий и мерах их поддержки [@Gumenyuk2022].
+Для исследований используются, в частности, сведения о доходах физических лиц [@Petrov2021], о налоговых поступлениях в разрезе субъектов России или видов экономической деятельности [@Rostislav2020; @Safronov2021], о среднесписочной численности работников организаций [@Rostislav2021], о количестве учтённых жилых помещений [@Lachininsky2021], о количестве малых и средних предприятий и мерах их поддержки [@Gumenyuk2022].
 Особое внимание привлекает реестр субъектов малого и среднего предпринимательства (реестр МСП), который регулярно и полностью размещается на сайте службы и содержит сведения о конкретных организациях и индивидуальных предпринимателях, включая адреса регистрации.
@@
 Подготовлен проектом «Инфраструктура научно-исследовательских данных» на базе открытых данных Росстата и сведений региональных медицинских информационно-аналитических центров.
-Содержит перечень городских и сельских населённых пунктов с указанием субъекта федерации и муниципального образования, кода ОКТМО и географических координат.
+Содержит перечень городских и сельских населённых пунктов с указанием субъекта Российской Федерации и муниципального образования, кода ОКТМО и географических координат.
 Второй — «Города России» (https://github.com/hflabs/city).
@@
 
-**Распределение по субъектам федерации.** Нормализованные адреса позволяют агрегировать информацию по субъектам федерации.
+**Распределение по субъектам Российской Федерации** Нормализованные адреса позволяют агрегировать информацию по субъектам Российской Федерации.
 На рис.
@@
 2.
-Распределение количества субъектов МСП, ведущих деятельность в сфере сельского и лесного хозяйства, по регионам (субъектам федерации).
+Распределение количества субъектов МСП, ведущих деятельность в сфере сельского и лесного хозяйства, по регионам (субъектам Российской Федерации).
 Цвет — масштабированное количество (0 — минимум, 1 — максимум) / Fig.
@@
 3.
-Распределение количества субъектов МСП, ведущих деятельность по выращиванию пшеницы и риса, по регионам (субъектам федерации).
+Распределение количества субъектов МСП, ведущих деятельность по выращиванию пшеницы и риса, по регионам (субъектам Российской Федерации).
 Цвет — масштабированное количество (0 — минимум, 1 — максимум).
@@
 4.
-Число субъектов МСП по населённым пунктам России и преобладающий вид деятельности.
-Размер кругов соответствует числу субъектов МСП, цвет — одной из трёх групп видов деятельности по классификатору ОКВЭД (см.
+Число субъектов МСП по муниципальным образованиям России и преобладающий вид деятельности.
+Цвет соответствует тому виду деятельности, который указан в качестве основного у наибольшего числа субъектов МСП в данном муниципалитете (см.
 легенду) / Fig.
 4.
-Count of SMBs in some Russian settlements with dominant activity type, in 2021.
-The counts are represented by circle size, and the most popular activity group in the settlement is displayed by color"}
-activity_by_settlements
+Count of SMBs in Russian municipal territories with dominant activity type, in 2021.
+The most popular activity in the municipality is displayed by color (see legend for a mapping of colors and actiity types)"}
+activity_by_municipalities
 ```
@@
 5.
-Распределение количества субъектов МСП из демонстрационного набора, прибывших или покинувших территорию региона (субъекта федерации) за 2016–2023 г.
+Распределение количества субъектов МСП из демонстрационного набора, прибывших или покинувших территорию региона (субъекта Российской Федерации) за 2016–2023 г.
 / Fig.
@@
   col.names = c("Год", "$\\rho_{Spearman}$, все группы ОКВЭД", "$\\rho_{Spearman}$, без O, P, U"),
-  caption = "Корреляция между числом организаций в реестре МСП и общим числом организаций в каждом регионе(субъекте федерации) в каждой группе деятельности по годам / Correlation between the count of SMB organizations and total count of commercial organizations by year")
+  caption = "Корреляция между числом организаций в реестре МСП и общим числом организаций в каждом регионе(субъекте Российской Федерации) в каждой группе деятельности по годам / Correlation between the count of SMB organizations and total count of commercial organizations by year")
 ```
@@
 6.
-Распределение регионов (субъектов федерации) по корреляции между числом организаций в реестре МСП и общим числом организаций в каждом году в каждой группе деятельности / Fig.
+Распределение регионов (субъектов Россиийской Федерации) по корреляции между числом организаций в реестре МСП и общим числом организаций в каждом году в каждой группе деятельности / Fig.
 6.
@@
   digits = 2,
-  col.names = c("Группа ОКВЭД", "$\\rho_{Spearman}$"),
-  caption = "Корреляция между числом организаций в реестре МСП и общим числом организаций в каждом регионе (субъекте федерации) и каждом году по группам деятельности / Correlation between the count of SMB organizations and total count of commercial organizations by activity group")
+  col.names = c("Код группы", "Наименование группы", "$\\rho_{Spearman}$"),
+  caption = "Корреляция между числом организаций в реестре МСП и общим числом организаций в каждом регионе (субъекте Российской Федерации) и каждом году по группам ОКВЭД / Correlation between the count of SMB organizations and total count of commercial organizations by activity group")
 ```

### 2024-07-19 ru-smb-data-in-economic-geography/helpers/Plots.R

@@ -1,6 +1,7 @@
 library(dplyr)
 library(forcats)
 library(ggplot2)
+library(here)
 library(readr)
 library(scales)
 library(sf)
@@ -49,17 +50,18 @@ labels <- data.frame(
 labels <- labels[["ru"]]
 
 # Data
-rsmp_data_path <- "../ru-smb-companies/group_A/smb.csv"
+rsmp_data_path <- here("../../ru-smb-companies/group_A/smb.csv")
 rsmp_data <- read_csv(rsmp_data_path)
 
 # Maps
-regions_boundaries <- st_read("assets/ru.geojson")
+regions_boundaries <- st_read(here("assets", "ru.geojson"))
+municipal_boundaries <- st_read(here("assets", "ru-mun-gadm.geojson"))
 ru_crs <- st_crs("+proj=aea +lat_0=0 +lon_0=100 +lat_1=68 +lat_2=44 +x_0=0 +y_0=0 +ellps=krass +towgs84=28,-130,-95,0,0,0,0 +units=m +no_defs")
-regions <- read_csv("assets/regions.csv")
+regions <- read_csv(here("assets", "regions.csv"))
 regions <- regions_boundaries %>% 
   left_join(regions, by = c("shapeISO" = "iso_code")) %>% 
   select(name, name_en = shapeName)
-ru_svr <- st_read("assets/ru_svr.geojson")
+ru_svr <- st_read(here("assets", "ru_svr.geojson"))
 
 # Regional distribution
 ac_code_mapping <- c("01" = labels[4], "02" = labels[5])
@@ -132,6 +134,56 @@ activity_by_settlements_df <- rsmp_data %>%
   group_by(region, area, settlement, lat, lon) %>% 
   slice_max(n) %>% 
   st_as_sf(coords = c("lon", "lat"), crs = 4326)
+
+sf_use_s2(FALSE)
+
+activity_by_municipalities <- rsmp_data %>%
+  mutate(
+    ac_type = substr(activity_code_main, 2, 4),
+    ac_type = case_when(
+      substr(ac_type, 1, 1) == "3" ~ "Рыболовство и рыбоводство",
+      substr(ac_type, 1, 1) == "2" ~ "Лесоводство и лесозаготовки",
+      ac_type == "1.4" ~ "Животноводство",
+      ac_type == "1.7" ~ "Охота",
+      ac_type %in% c("1.1", "1.2", "1.3") ~ "Растениеводство",
[…]

### 2024-07-19 ru-smb-data-in-economic-geography/helpers/Stats.R

@@ -1,6 +1,8 @@
 library(dplyr)
 library(ggplot2)
+library(here)
 library(readr)
+library(stringr)
 library(tidyr)
 
 labels <- data.frame(
@@ -10,8 +12,8 @@ labels <- data.frame(
     "Итоговые таблицы",
     "Панельное представление",
     "Группы ОКВЭД",
-    "Все",
-    "Все, кроме O, P, U",
+    "Все группы ОКВЭД",
+    "Все группы ОКВЭД, кроме O, P, U",
     "Коэффициент корреляции Спирмена",
     "Число субъектов России"
   ),
@@ -21,8 +23,8 @@ labels <- data.frame(
     "Resulting tables",
     "Panel view",
     "Activity groups",
-    "All",
-    "All except for O, P, U",
+    "All activity groups",
+    "All activity groups except for O, P, U",
     "Spearman's rho",
     "Number of regions"
   )
@@ -30,15 +32,20 @@ labels <- data.frame(
 labels <- labels[[LOCALE]]
 
 # Counts
-rsmp_data <- read_csv("../ru-smb-companies/group_A/smb.csv")
+rsmp_data <- read_csv(here("../../ru-smb-companies/group_A/smb.csv"))
 counts <- rsmp_data %>% 
   mutate(kind = replace(kind, kind == 3, 2)) %>% 
   distinct(tin, .keep_all = TRUE) %>% 
   count(kind) %>% 
   pull(n)
 
+# Activity codes
+ac <- read_csv(here("assets", "activity_codes_classifier.csv"))
+ac_groups <- filter(ac, is.na(code)) %>%
+  mutate(name = str_to_sentence(name))
+
 # Validation
-val_stats <- read_csv("assets/validation-stats.csv")
+val_stats <- read_csv(here("assets", "validation-stats.csv"))
 
 corr_by_year_all <- val_stats %>% 
   group_by(year) %>% 
@@ -67,14 +74,27 @@ corr_by_region <- left_join(
   corr_by_region_all, corr_by_region_filtered,
   by = "region", suffix = c("_all", "_filtered")) %>% 
   pivot_longer(-region, names_to = "option", values_to = "cor")
-corr_by_region_plot <- ggplot(corr_by_region, aes(x = cor, color = option)) +
-  geom_freqpoly(binwidth = .05) +
-  scale_color_discrete(name = labels[5], labels = c(labels[6], labels[7])) +
[…]

### 2024-07-19 law-firms-names/Paper.Rmd
Write paragraphs about vectors, YandexGPT, concrete techniques, and other important consideration about methods, #4

@@
 whether the regions that are close to each other have more similar names, and vice versa.
-Question 3 (Q3) moves up from the level of regions to the level of economic regions: whether names withing the economic region are more similar compared to the names outside the region.
+Question 3 (Q3) moves up from the level of regions to the level of economic regions: whether names within the economic region are more similar compared to the names outside the region.
 Question 4 (Q4) is what naming strategies exist in Russia and how they are spatially organized.
@@
 
+Text embedding, or vector, is a numeric representation of a word, sentence or a document usually designed to catch the meaning of the text.
+An embedding typically looks like just a sequence of decimals with fixed length, i.e.
+the count of elements in this sequence does not depend on the length of the input text, so that a short sentence or an entire paragraph are represented as a sequence of, say, 256 numbers.
+Vectors allow simple computations to find out whether they are more or less similar to each other, and the similarity of vectors corresponds to the similarity of texts.
+Thus, by computing a mathematical distance between embeddings, one can conclude about the semantic distance between texts.
+This is right the case of firm names comparison, similarity, and grouping.
+
+There are various methods of obtaining embeddings for different texts, e.g.
+Word2Vec, GloVe, FastText.
+The newest GPT-based pretrained large language models such as ChatGPT are also capable of producing text vectors.
+For this study, embeddings from YandexGPT model were used.
+These embeddings were obtained for each unique firm name via public API of the model.
+The vector dimensionality is 256, i.e.
+each vector is a sequence of 256 numbers.
+All the concrete hypothesis testing techniques outlined in the next subsection are based on the use of these vectors.
+
+### 2.2.3. Hypothesis testing
+
+To answer Q1, firm names vectors by each region were averaged so that all names of a region are represented with a single vector that equals to the mathematical mean of all firm names vectors within this region.
+All regions are divided into two groups depending on the presence of common border between them.
+Pairwise distances between all the regions are calculated, and then median distance between regions in each of two groups is calculated.
+The results are additionally visualized as a box plot.
+It is expected that median distance between neighboring regions is smaller than median distance between regions that do not have a same border.
+
+The similar technique is used for Q2.
+Again, the mean vector of all firm names for each region is computed, and pairwise semantic distance between regions is calculated.
+Then, a geographical distance between the centroids of regions is also measured.
+A correlation between geographical distance and semantic distance is computed, and the result is plotted.
+It is expected that the coefficient of correlation is positive.
+
+Q3 also builds upon the pairwise semantic distances between regions.
+After these distances are found out, regions are grouped by economic regions.
+For each economic region, two metrics are calculated: the first is a median distance between regions within this econoimic region, and the second is a median distance between regions within and without this economic region.
+It is expected that within-median-distance is smaller that beyond-median-distance.
+The results are additionally illustrated on a boxplot.
+
+Q4 and Q5 follow a slightly different methodics.
+First, all firm names are clustered with KMeans algorithm (number of clusters is 50).
+After clustering, a random sample of names from each cluster is taken, and these names are manually checked, described and tagged with the labels of more broad groups.
+Then these manually defined groups are assigned to names using clusters as links between groups and individual names.
+The number of names belonging to each group in each region is calculated, and the results of this calculation are mapped on a tile-grid map of Russia to analyze the spatial properties of grouping.
+In this way, Q4 is processed.
+For Q5, the regions are marked with the label of the most popular group, and the number of regions with different groups within an economic region is counted.
+It is expected that each economic regions is characterized with only one majority group.
+It may be the same as the group for another economic region, but a small if any variation is expected within an economic region to prove its integrity.
+Also, in general it is expected that there are several clear and meaningful firm naming groups, or naming strategies, and that these groups may have (or not) some clear geographical patterns.
+
+### 2.2.4. Other important considerations
+
+The time range of the source data is from 2016 to 2023.
+Only the firms that have existed for at least 3 years are included in the analysis.
+Only organizations are studied, while individual entrepreneurs are omitted, because they do not have firm names.
+Firm names are preprocessed to remove any parts pointing to the legal form of the business (e.g.
+limited liability company or public stock company), because these parts are common to the majority of names and thus do not add anything valuable.
+Law firms are selected by filtering using the following condition: main activity code equals 69.10 (activity in the area of law according to All-Russia Classifier of Kinds of Economic Activity).
+This filtering technique is derived from @Moiseeva2016.
+
 # 3. Results and discussion

### 2024-07-20 ru-smb-data-in-economic-geography/Paper.Rmd
Edit maps and text to fit problem 1 from the review, update geodata, #5

@@
   word_document:
-    reference_doc: assets/reference.docx
-bibliography: assets/biblio_ru.bib
+    reference_doc: "assets/reference.docx"
+bibliography: "assets/biblio_ru.bib"
 csl: "assets/citations-ru.csl"
@@
 
-**Распределение по субъектам Российской Федерации** Нормализованные адреса позволяют агрегировать информацию по субъектам Российской Федерации.
+**Распределение по субъектам Российской Федерации.** Нормализованные адреса позволяют агрегировать информацию по субъектам Российской Федерации.
 На рис.
@@
 Распределение ожидаемо разное: наибольшее количество фирм в сфере сельского хозяйства расположено в Черноземье и на юге России, также заметны животноводческие регионы; лесное хозяйство сосредоточено на севере европейской части страны, в Восточной Сибири и на Дальнем Востоке.
+Кроме того, видно, что по числу зарегистрированных фирм лесное хозяйство уступает сельскому: даже в регионах, которые специализируются на лесном хозяйстве, количество субъектов МСП, ведущих основную деятельность в этой сфере, едва сравнивается с количеством субъектов МСП, ведущих деятельность в сельском хозяйстве (вероятно, в основном в животноводстве и охоте, которая по ОКВЭД отнесена к тому же коду 01, что и сельское хозяйство).
 
@@
 2.
-Распределение количества субъектов МСП, ведущих деятельность в сфере сельского и лесного хозяйства, по регионам (субъектам Российской Федерации).
-Цвет — масштабированное количество (0 — минимум, 1 — максимум) / Fig.
+Распределение количества субъектов МСП, ведущих деятельность в сфере сельского и лесного хозяйства, по регионам (субъектам Российской Федерации) / Fig.
 2.
-Regional distribution of SMBs which main activity is agriculture or forestry, in 2021.
-Color represents the min-max scaled value"}
+Regional distribution of SMBs which main activity is agriculture or forestry, in 2021", fig.height = 3}
 regions_plot_agri_forestry
@@
 Можно заметить, что пространственное распределение фирм отличается и отчасти согласуется с районами выращивания соответствующих культур, хотя вряд ли полностью соответствует ему, так как номинальный код основного вида деятельности не всегда соответствует фактическому.
+Кроме того, выращивание пшеницы — это намного более распространённое занятие, чем выращивание риса, где даже в регионах специализации общее число фирм весьма невелико.
 
@@
 Распределение количества субъектов МСП, ведущих деятельность по выращиванию пшеницы и риса, по регионам (субъектам Российской Федерации).
-Цвет — масштабированное количество (0 — минимум, 1 — максимум).
 Серый цвет — отсутствие данных / Fig.
@@
 Regional distribution of SMBs which main activity is growing of wheat or rice, in 2021.
-Color represents the min-max scaled value, gray is no data"}
+Grey is no data", fig.height = 3}
 regions_plot_wheat_rice
@@
 
-**Распределение по населённым пунктам.** Набор данных помимо названия субъекта содержит географические координаты, название населённого пункта и территориальной единицы (района, муниципалитета) внутри субъекта.
-Это позволяет выполнять более детализированные агрегации и показывать их на карте.
+**Распределение по муниципальным образованиям.** Набор данных помимо названия субъекта содержит географические координаты, название населённого пункта и территориальной единицы (района, муниципалитета) внутри субъекта.
+Это позволяет выполнять более детализированные агрегации и показывать их на карте — например, спуститься от субъектов Российской Федерации на уровень ниже и перейти к муниципалитетам.
 На рис.
-4 показано количество субъектов МСП в населённых пунктах России в 2021 г, а также преимущественная специализация населённых пунктов исходя из того, каких фирм в нём больше всего.
+4 показано количество субъектов МСП в муниципальных образованиях России в 2021 г, а также преимущественная специализация муниципалитетов исходя из того, каких фирм в нём больше всего.
 
@@
 
-```{r fig-4-activity-by-settlements-russia, fig.cap = "Рис.
+```{r fig-4-activity-by-municipalities-russia, fig.cap = "Рис.
 4.
@@
 Цвет соответствует тому виду деятельности, который указан в качестве основного у наибольшего числа субъектов МСП в данном муниципалитете (см.
-легенду) / Fig.
+легенду).
+Границы муниципалитетов показаны в соответствии с GADM level2 / Fig.
 4.
 Count of SMBs in Russian municipal territories with dominant activity type, in 2021.
-The most popular activity in the municipality is displayed by color (see legend for a mapping of colors and actiity types)"}
+The most popular activity in the municipality is displayed by color (see legend for a mapping of colors and actiity types).
+GADM municipal borders (level2) are used"}
 activity_by_municipalities
@@
 5.
-Regional distribution of SMBs which have arrived to or left the region during 2016–2023"}
+Regional distribution of SMBs which have arrived to or left the region during 2016–2023", fig.height = 3}
 migrations_plot

### 2024-07-20 ru-smb-data-in-economic-geography/helpers/Plots.R

@@ -74,48 +74,42 @@ regions_plot_agri_forestry <- rsmp_data %>%
     end_date >= "2021-12-31",
     !is.na(ac_type)) %>% 
   count(region, ac_type) %>% 
-  group_by(ac_type) %>% 
-  mutate(n = rescale(n)) %>% 
   left_join(regions, by = c("region" = "name")) %>% 
   st_as_sf() %>% 
   ggplot() +
-  geom_sf(aes(fill = n), size = .1) +
+  geom_sf(aes(fill = n / 1000), linewidth = .05) +
   coord_sf(crs = ru_crs) +
   scale_fill_distiller(
-    name = NULL,
-    labels = function(breaks) c("Min", rep("", length(breaks) - 2), "Max"),
+    name = "Количество субъектов МСП, тыс.",
+    breaks = seq(3, 12, 3),
     palette = "YlGn",
     direction = 1) +
   facet_wrap(~ac_type) +
-  theme_bw(base_size = 12, base_family = "Times New Roman") +
+  theme_void(base_size = 12, base_family = "Times New Roman") +
   theme(legend.position = "bottom")
 regions_plot_agri_forestry
 
 ac_code_mapping_wr <- c("01.11.1" = labels[6], "01.12" = labels[7])
 regions_plot_wheat_rice <- rsmp_data %>%
-  mutate(
-    ac_type = ac_code_mapping_wr[activity_code_main]) %>% 
+  mutate(ac_type = ac_code_mapping_wr[activity_code_main]) %>% 
   filter(
     start_date <= "2021-12-31",
     end_date >= "2021-12-31",
     !is.na(ac_type)) %>% 
   count(region, ac_type) %>% 
-  group_by(ac_type) %>% 
-  mutate(n = rescale(n)) %>% 
-  pivot_wider(names_from = ac_type, values_from = n) %>% 
   right_join(regions, by = c("region" = "name")) %>% 
+  pivot_wider(names_from = ac_type, values_from = n) %>% 
   pivot_longer(cols = c(labels[6], labels[7]), names_to = "ac_type", values_to = "n") %>% 
   st_as_sf() %>% 
   ggplot() +
-  geom_sf(aes(fill = n), size = .1) +
+  geom_sf(aes(fill = n / 1000), linewidth = .05) +
   coord_sf(crs = ru_crs) +
   scale_fill_distiller(
-    name = NULL,
[…]

### 2024-07-21 fts-open-data-in-economic-geography/Paper.Rmd
Rename folder, make minor edits in text, add missing classifier file, fix an issue in references, #5

--- 2024-07-20 ru-smb-data-in-economic-geography/Paper.Rmd
@@
 ***Аннотация.*** В статье представлена технология для создания геопривязанного табличного набора данных о субъектах малого и среднего предпринимательства (МСП) в России на основе открытых данных Федеральной налоговой службы (ФНС России).
-Получаемый набор данных охватывает всю территорию России за 2016–2023 гг.
+Получаемый набор данных охватывает всю территорию России за период с 2016 г.
 Он содержит детализированную до уровня конкретных юридических лиц и индивидуальных предпринимателей информацию о регистрационных номерах, категории, имени или наименовании, месте жительства или местонахождении, виде деятельности, доходах, расходах и численности работников.
@@
 
-***Abstract.*** The paper features a technology for creating a geo-referenced tabular dataset of small and medium-sized businesses (SMBs) in Russia based on open data from the Federal Tax Service (FTS of Russia).
-The resulting dataset covers the entire territory of Russia for 2016—2023.
+***Abstract.*** The paper features a technology for creating a geocoded tabular dataset of small and medium-sized businesses (SMBs) in Russia based on open data from the Federal Tax Service (FTS of Russia).
+The resulting dataset covers the entire territory of Russia for the period from 2016.
 It contains information on registration numbers, category, name, address, activity type, income, expenses, and number of employees.
@@
 
-Для некоторых экономико-географических важны сведения, детализированные до уровня отдельных фирм.
+Для части экономико-географических исследований важны сведения, детализированные до уровня отдельных фирм.
 К таким сведениям относятся, например, местонахождение, выручка, расходы, число работников, наименование и другие характеристики.
@@
 Особое внимание привлекает реестр субъектов малого и среднего предпринимательства (реестр МСП), который регулярно и полностью размещается на сайте службы и содержит сведения о конкретных организациях и индивидуальных предпринимателях, включая адреса регистрации.
-Де-факто это наиболее полный из открытых наборов сведений о российских фирмах, детализированный до уровня отдельных фирм и геопривязанный хотя бы каким-то образом.
+Де-факто это наиболее полный набор данных о российских фирмах, детализированный до уровня отдельных фирм, доступный всем без платы и геопривязанный хотя бы каким-то образом.
 Использовать его в исходном виде — непростая задача: выгрузки реестра за несколько лет имеют суммарный сжатый объём 200 Гб и характеризуются сложной многоуровневой структурой, а адреса, записанные в привычном для человека виде, не содержат географических координат или кодов муниципальных образований, в результате чего малопригодны для автоматизированного анализа.
@@
 Норма о реестре появилась в законе в конце 2015 г.
-На практике реестр размещён на сайте ФНС России в форме открытых данных в виде ежемесячных полных выгрузок, первая из которых датируется августом 2016 г., а последняя доступная на момент подготовки статьи — декабрём 2023 г.
+На практике реестр размещён на сайте ФНС России в форме открытых данных в виде ежемесячных полных выгрузок, первая из которых датируется августом 2016 г., а последняя доступная на момент подготовки статьи к публикации — июлем 2024 г.
 (https://www.nalog.gov.ru/opendata/7707329152-rsmp/).
@@
 Коллекция zip-архивов тем самым образует временной ряд (time series).
-Каждый zip-архив содержит множество (обычно 6-7 тысяч) xml-файлов.
+Каждый zip-архив содержит множество (обычно 6–7 тысяч) xml-файлов.
 Каждый xml-файл включает в себя информацию о множестве (обычно ровно 900) субъектов МСП.
@@
 
-На момент подготовки статьи (декабрь 2023 г.) было доступно 88 zip-архивов.
+На момент написания статьи (декабрь 2023 г.) было доступно 88 zip-архивов.
 Их суммарный объём — 193 Гб в сжатом виде и 2.4 Тб в распакованном.
 Из-за большого объёма исходных данных требуется оптимизировать алгоритм их обработки и использовать специализированные инструменты для работы с «большими данными», о которых будет сказано ниже.
-При этом проделанные оптимизации позволяют сделать все необходимые вычисления на обычном персональном компьютере, не задействуя облачные платформы.
+При этом проделанные оптимизации позволяют осуществить все необходимые вычисления на обычном персональном компьютере, не задействуя облачные платформы.
 
@@
 Технически аналогичен реестру МСП: представляет собой временной ряд zip-архивов xml-файлов, где каждый архив соответствует определённой дате, а каждый xml-файл внутри архива содержит информацию о множестве организаций.
-Второй — «Сведения о среднесписочной численности работников организации» (https://www.nalog.gov.ru/opendata/7707329152-sshr2019/).
+Второй вспомогательный источник информации — «Сведения о среднесписочной численности работников организации» (https://www.nalog.gov.ru/opendata/7707329152-sshr2019/).
 Содержит ИНН и среднесписочную численность работников за год, предшествующий году публикации сведений.
@@
 по настоящее время.
-На момент подготовки статьи на сайте ФНС России была опубликована только последняя версия этого и следующего набора данных.
+На момент подготовки статьи на сайте ФНС России была опубликована только последняя версия вспомогательных наборов данных.
 Раньше были доступны предыдущие.
@@
 Содержит перечень городских населённых пунктов с информацией о их месте в административно-территориальном делении, коде ОКТМО и географических координатах.
-Поскольку некоторые населённые пункты в нём отсутствуют, автор статьи добавил их вручную, опираясь на сведения Федеральной информационной адресной системы (ФИАС).
+Поскольку некоторые города в нём отсутствуют, автор статьи добавил их вручную, опираясь на сведения Федеральной информационной адресной системы (ФИАС).
 
@@
 Кроме того, благодаря ей создаётся резервная копия исходных данных, что страхует от риска удаления наборов с сайта налоговой службы.
-Теоретически возможна загрузка и обработка данных в облачном сервисе, таком как Yandex Cloud.
+Теоретически также возможна загрузка и обработка данных в облачном сервисе, таком как Yandex Cloud.
 
@@
 Отфильтрованная таблица будет содержать сведения только из необходимой части реестра МСП, что ускоряет работу и сокращает использование дискового пространства.
-Реализация данного шага оптимизирована, чтобы уменьшить время обработки, расход оперативной памяти и места для хранения временных файлов: данные из архивов читаются напрямую без распаковки архива целиком, извлекаются и сохраняются только необходимые атрибуты записей о фирмах.
+Реализация данного шага оптимизирована, чтобы уменьшить время обработки, расход оперативной памяти и места для хранения временных файлов: данные из архивов читаются напрямую без распаковки архива целиком, при этом извлекаются и сохраняются только необходимые атрибуты записей о фирмах, а всё лишнее отбрасывается.
 Аналогичным образом обрабатываются сведения о доходах, расходах (2b) и о среднесписочной численности работников организаций (2c).
@@
 была известна под одним наименованием, а с 10 апреля 2021 г.
-— под другим, то в агрегированной таблице она будет представлена двумя строками.
-У первой дата начала — 10.08.2016, дата окончания — 10.03.2021; у второй дата начала — 10.04.2021, дата окончания — 10.12.2023 (условная дата последнего архива исходных данных).
+— под другим, то в агрегированной таблице она будет представлена двумя строками, у первой из которых дата начала — 10.08.2016, дата окончания — 10.03.2021; у второй дата начала — 10.04.2021, дата окончания — 10.12.2023 (условная дата последнего архива исходных данных).
 Кроме того, на данном этапе происходят вспомогательные преобразования, цель которых — исправить неточности в исходных данных и корректно определить дубли.
 Например, в реестре МСП тип населённого пункта может быть указан полностью («город») или сокращённо («г.»), и вспомогательная обработка перед агрегацией позволяет игнорировать такие различия и воспринимать адрес как неизменный.
-Аналогичным образом производится агрегация и фильтрация данных о доходах, расходах (3b) и агрегация данных о работниках (3c).
+Аналогичным образом производится агрегация данных о доходах, расходах (3b) и агрегация данных о работниках (3c).
+Кроме того, эти данные не только агрегируются, но и фильтруются так, чтобы в них осталась только информация о тех организациях, которые есть в основном наборе, полученном на шаге 3a.
 
@@
 Для этого используются адрес регистрации и вспомогательные наборы геоданных.
-Кроме того, исходный адрес в наборе данных заменяется на нормализованный и сохраняется как отдельный атрибут каждый записи в виде структурированной текстовой строки.
+Кроме того, исходный адрес в наборе данных заменяется на нормализованный.
 После нормализации адресов набор данных повторно проверяется на дубли.
@@
 Пятый, опциональный шаг — создание панельного представления (5).
-Таблица, полученная на этапе 4, объединяется с таблицами, полученными на этапах 3b и 3c, и преобразуется в панельный временной ряд, где представлены характеристики каждого субъекта МСП за каждый год с 2016 по 2023.
+Таблица, полученная на этапе 4, объединяется с таблицами, полученными на этапах 3b и 3c, и преобразуется в панельный временной ряд, где представлены характеристики каждого субъекта МСП за каждый год с 2016.
 Содержательно такая таблица не отличается от первоначальных, но более удобна для анализа: она одна вместо трёх, и вместо компактного, но не слишком интуитивного и весьма технологизированного формата с датами актуальности информация в ней представлена по годам.
+Если сведения о субъекте МСП в течение года менялись, то строк будет несколько.
 Такой вид более привычен для исследователей, а также упрощает объединение набора данных с официальной статистикой, которая также обычно даётся в разрезе календарных лет.
 
-**Характеристики итогового набора данных.** Получаемый набор данных охватывает всю территорию России за период с середины 2016 г.
+**Характеристики итогового набора данных.** Получаемый набор данных охватывает всю территорию России за период с августа 2016 г.
 по настоящее время.
@@
 
-**Технические детали реализации.** Описанный выше алгоритм реализован как набор скриптов на языке программирования Python.
+**Технические детали реализации.** Описанный выше алгоритм реализован как консольное (CLI) приложение на языке программирования Python.
 Чтение xml осуществляется с помощью высокоэффективной библиотеки lxml.
-Для агрегации используется система обработки больших данных Apache Spark.
-Геопривязка и создание панельного представления реализованы с помощью библиотеки Pandas.
-Исходный код опубликован в личном репозитории автора статьи: https://github.com/PavelSyomin/russian-smb-companies
+Для агрегации, геопривязки и создания панельного представления используются система обработки больших данных Apache Spark и библиотека Pandas.
+Исходный код опубликован в личном репозитории автора статьи: https://github.com/PavelSyomin/ru-smb-companies
 
@@
 Фильтрация вида деятельности производилась по следующему условию: основной код ОКВЭД организации или предпринимателя находится в группе А классификатора.
-Использовались исходные данные за весь доступный период: с августа 2016 г.
+Использовались исходные данные за весь доступный на момент написания статьи период: с августа 2016 г.
 для реестра МСП, с октября 2019 г.
 для сведений о доходах и расходах организаций; с марта 2020 г.
-для сведений о среднесписочной численности работников организаций.
+для сведений о среднесписочной численности работников организаций — и до ноября 2023 г.
 Описание и визуализация результата приведены в следующем разделе статьи.
@@
 2.
-Regional distribution of SMBs which main activity is agriculture or forestry, in 2021", fig.height = 3}
+Regional distribution of SMBs with main activity in agriculture or forestry, in 2021", fig.height = 3}
 regions_plot_agri_forestry
@@
 
-В отличие от официальной статистики, получившийся набор данных позволяет проанализировать пространственное распределение с точностью до конкретных видов деятельности, а не крупных агломератов.
+В отличие от официальной статистики, производимый набор данных позволяет проанализировать пространственное распределение с точностью до конкретных видов деятельности, а не крупных агломератов.
 На рис.
@@
 3.
-Regional distribution of SMBs which main activity is growing of wheat or rice, in 2021.
+Regional distribution of SMBs which main activity code is growing wheat or rice, in 2021.
 Grey is no data", fig.height = 3}
@@
 
-**Распределение по муниципальным образованиям.** Набор данных помимо названия субъекта содержит географические координаты, название населённого пункта и территориальной единицы (района, муниципалитета) внутри субъекта.
+**Распределение по муниципальным образованиям.** Набор данных помимо названия субъекта Российской Федерации содержит географические координаты, название населённого пункта и территориальной единицы (района, муниципалитета) внутри субъекта.
 Это позволяет выполнять более детализированные агрегации и показывать их на карте — например, спуститься от субъектов Российской Федерации на уровень ниже и перейти к муниципалитетам.
 На рис.
-4 показано количество субъектов МСП в муниципальных образованиях России в 2021 г, а также преимущественная специализация муниципалитетов исходя из того, каких фирм в нём больше всего.
+4 показано количество субъектов МСП в муниципальных образованиях России в 2021 г, а также преимущественная специализация муниципалитетов исходя из того, каких фирм в них больше всего.
 
@@
 4.
-Число субъектов МСП по муниципальным образованиям России и преобладающий вид деятельности.
+Число субъектов МСП по муниципальным образованиям и преобладающий вид деятельности.
 Цвет соответствует тому виду деятельности, который указан в качестве основного у наибольшего числа субъектов МСП в данном муниципалитете (см.
@@
 The most popular activity in the municipality is displayed by color (see legend for a mapping of colors and actiity types).
-GADM municipal borders (level2) are used"}
+GADM municipal borders (level2) are used", fig.height = 5}
 activity_by_municipalities
@@
 На рис.
-5 показано изменение количества субъектов МСП из пробного набора, зарегистрированных в субъектах России, за весь доступный период.
+5 показано изменение количества субъектов МСП из пробного набора, зарегистрированных в субъектах Российской Федерации, за весь проанализированный период.
 
@@
 5.
-Regional distribution of SMBs which have arrived to or left the region during 2016–2023", fig.height = 3}
+Regional distribution of SMBs which have arrived to or left the region during 2016–2023", fig.height = 4.5}
 migrations_plot
@@
 
-**Обсуждение.** Предложенный алгоритм имеет техническую ценность: он позволяет обработать большой массив данных, имеющих сложную структуру, и сделать на его основе несколько сравнительно компактных таблиц в формате, который традиционно используется исследователями.
+**Обсуждение.** Предложенный алгоритм имеет техническую ценность: он позволяет обработать большой массив сложноструктурированных данных и сделать на его основе несколько сравнительно компактных таблиц в формате, который традиционно используется исследователями.
 Кроме того, технология нормализует адреса, содержащиеся в исходных данных, и обогащает их географическими координатами и кодами ОКТМО, тем самым делая их пригодными для обработки в геоинформационных системах (ГИС) и пространственного анализа в целом, а также упрощая объединение с муниципальной статистикой.
@@
 На первый взгляд, это разумное предположение, ведь малые и средние предприятия составляют почти 98% в общем числе фирм в России, дают 43% ВВП и обеспечивают занятостью 38% работников [@Barinova2019].
-В то же время очевидно, что реестр МСП не является случайной выборкой из генеральной совокупности хозяйствующих субъектов, поскольку его элементы выбраны не случайно, а исходя из ограничений по финансовым показателям и числу занятых.
+В то же время очевидно, что реестр МСП не является случайной выборкой из генеральной совокупности хозяйствующих субъектов, поскольку его элементы выбраны не случайно, а исходя из ограничений по финансовым показателям и числу сотрудников.
 Следовательно, реестр МСП позволяет делать выводы о небольших фирмах, но не позволяет распространить их на все компании.
@@
 
-Таким образом, в большинстве случаев реестр МСП может рассматриваться как достаточно качественная замена полному списку организаций как минимум в двух аспектах: он корректно отражает географическое распределение количества организаций по субъектам России в подавляющем большинстве групп ОКВЭД и структуру экономики по группам ОКВЭД в значительной части субъектов.
+Таким образом, в большинстве случаев реестр МСП может рассматриваться как достаточно качественная замена полному списку организаций как минимум в двух аспектах: он корректно отражает географическое распределение количества организаций по субъектам Российской Федерации в подавляющем большинстве групп ОКВЭД и структуру экономики по группам ОКВЭД в значительной части субъектов.
 Это не гарантирует репрезентативность на более низком уровне (конкретные коды ОКВЭД, муниципальные образования, населённые пункты) или по другим характеристикам (число работников, прибыль).
@@
 Юридический адрес может отличаться от места ведения деятельности, а ОКВЭД, указанный в документах, может не соответствовать фактической деятельности.
-Сложности усиливаются из-за того, что расхождения могут проявляться в разной степени в зависимости от расположения и отрасли экономики: так, в особых экономических зонах много номинально зарегистрированных организаций, а место деятельности интернет-компаний вряд ли совпадает с местом нахождения офиса.
+Сложности усиливаются из-за того, что расхождения могут проявляться в разной степени в зависимости от расположения и отрасли экономики: так, в особых экономических зонах много номинально зарегистрированных организаций, а место деятельности интернет-компаний обычно не совпадает с местом нахождения офиса.
 Вряд ли эту проблему можно решить системно, поэтому технология презюмирует достоверность данных о местонахождении и виде деятельности и не проверяет их.
@@
 В статье описана технология, позволяющая на основе открытых данных ФНС России создать геопривязанный набор данных о субъектах МСП (организациях и индивидуальных предпринимателях), ведущих деятельность определённого вида.
-Получаемый набор охватывает период с середины 2016 г.
+Получаемый набор охватывает период с августа 2016 г.
 по настоящее время, содержит сведения о регистрационных номерах, наименовании, местонахождении фирмы, её виде и категории, доходах, расходах и числе работников.

### 2024-09-04 law-firms-names/Paper.Rmd
Adjust style and metadata to GeoJournal requirements, #4

--- 2024-07-18 law-firms-names/Paper.Rmd
@@
 ---
-title: "Spatial Analysis of Russian Law Firms Naming"
-author: "Pavel Syomin¹"
+title: "Application of large language models to a geographical analysis of business practices: a case of law firms naming in Russia"
+author: "Pavel Syomin"
 output:
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(
+  warning = FALSE, 
+  message = FALSE, 
+  echo = FALSE, 
+  fig.width = 4.96, 
+  fig.id = FALSE,
+  dpi = 1200)
+IS_PAPER <- TRUE # used in script to disable saving of names and samples
+```
+
+```{r package-imports, include=FALSE}
+library(dplyr)
+library(ggplot2)
+library(glue)
+library(here)
+library(readr)
+library(readxl)
+library(sf)
+library(stringr)
+library(tidyr)
+library(units)
+```
+
+```{r data-imports, cache=TRUE, include=FALSE}
+source(here("journal-article", "Script.R"))
 ```
@@
 
-¹Perm State University, Geographical faculty, Department of social and economic geography, Perm, Russia; e-mail: ntsp\@ya.ru
+Perm State University, Geographical faculty, Department of social and economic geography, Perm, Russia
+
+E-mail: ntsp\@ya.ru
+
+ORCID: 0000-0002-4015-9206
 
@@
 
-**Abstract:** text.
+**Abstract:** The present study analyzes law firm naming in Russia, with a particular focus on the geographical aspects of this phenomenon.
+The primary objective of this research is to ascertain the impact of the spatial relations between the country's regions, specifically those of a neighboring, proximity, and belonging to a parent-level subdivision, on the naming of law firms.
+To conduct the analysis, data regarding law firms is extracted from the open dumps of the state registry of small and medium-sized businesses.
+The names of the firms are transformed into numerical vectors, subsequently clustered with the KMeans method, and then manually grouped to identify naming strategies.
+Numerical representations of law firm names within each region are averaged to get region embeddings, which are then used to compare the names of law firms in different regions.
+The research findings indicate that regions situated closer to one another or sharing a common border tend to adopt slightly more similar law firm names.
+Furthermore, regions that are part of the same economic region also display a greater degree of similarity in law firm naming.
+Three primary strategies of law firm naming are “Form”, “Law“, and “Service”.
+Volga and North Caucasus economic regions are characterized by the predominance of the “Service” strategy, whereas the Volga-Vyatka economic region exhibits a “Law” strategy, and the remaining territories demonstrate a combination of naming strategies.
+The results of the study lend support to the widely used classification of economic regions in Russia.
+The findings also underscore the impact of neighboring, proximity, and interconnectivity on the economic and cultural attributes of the regions.
+The administrative open data processing techniques, as well as the methods of working with firm names based on neural networks proposed in this paper, may be used by other economic geographers for their own research purposes.
 
@@
 
-**Keywords:** naming; law firms; text embeddings; economic regions. 
+**Keywords:** firm names; law firms; text embeddings; economic regions, text similarity.
+
 </div>
+
+# Statements and Declarations
+
+**Competing Interests.** No funding was received for conducting this study. The author has no financial or proprietary interests in any material discussed in this article.
+
+\newpage
 
@@
 
-Geography of services, sometimes referred to as geography of service sector, is one of the essential parts of socio-economic geographical studies in Russia.
-Its history spans for several decades starting from the 1960s.
-Geography of services has economic and social dimensions, where the former is concentrated on the business aspects of the services, while the latter is focused on the issues of human well-being arising from the spatial structure of the relevant services [@Tkachenko2016].
-Existing scientific literature on the geography of services in Russia covers a broad range of topics, e.g.
-education [@Shulgina2019], car rental [@Zyrianova2018], catering [@Preobrazhenskiy2018], medical care [@Kazantcev2022; @Panin2023; @Nikoloski2024], internet access [@Nagirnaya2015; @Zemtsov2022].
-However, research on some problems is limited, with legal services being one of the notable examples of such a problem [@Tkachenko2016; @Moiseeva2016].
-Despite the lack of attention from geographers, spatial aspects of law firms activity worth investigating for the several reasons.
-First, legal services can be categorized as business and producer services, acting as a representative case, and this category of services is important for the economic development in general, both in national or regional scale [@Hansen1993; @Faulconbridge2019].
-Next, despite their relatively small direct economic impact, they have a great indirect importance as the intermediaries and helpers that contribute to the rule of law, social order, stability and predictability of social relations.
-Third, for regular citizens access to legal aid is important for the protection of human rights and ensuring the social justice, and in this context, the spatial aspects of law services are on par with the spatial issues of medical care [@Blacksell1988; @Blacksell1990].
+The geography of services, also known as the geography of the service sector, is a crucial component of geographical studies in Russia.
+This field has a history spanning several decades, with origins dating back to the 1960s.
+The geography of services has two distinct dimensions: economic and social.
+The former is primarily concerned with the business aspects of the services, while the latter addresses the implications for human well-being arising from the geographical organization of services [@Tkachenko2016].
+The existing scientific literature on the geography of services in Russia covers a broad range of topics, including education [@Shulgina2019], car rental [@Zyrianova2018], catering [@Preobrazhenskiy2018], medical care [@Kazantcev2022; @Panin2023; @Nikoloski2024], and internet access [@Nagirnaya2015; @Zemtsov2022].
+However, some areas remain underexplored, with legal services being one such example [@Tkachenko2016; @Moiseeva2016].
 
-Geographical research of services may focus on a variety of subjects.
-Some classical pathways of such research include analysis of centers and systems of service or of consumers' behavior, needs and demands, including estimation of influencing factors, investigation on the actual level of services provision, applied scientific work related to the management of social infrastructure [@Tkachenko2016].
-Studies of firm names, sometimes also referred to as company names, brand names or just brands, is relatively unusual yet potentially fruitful method in the geography of services.
-Firm names are designed to distinguish the business from the competitors, to make an emphasis on the field of activity or its geographical area, inform about services supplied by a provider and corresponding benefits.
-Having these functions, brands convey a substantial portion of information in a short text message.
-Such information may be useful to discover the city's business landscape and consumer's perception of a particular city with its characteristic services [@Raento2001], analyze the trends common to a specific industry in a region [@Marjavaara2022] or the role of place as reflected in the toponyms in the business development and consumer attitudes [@Peluso2012; @Zhang2014; @Erdogan2016; @Noorda2019].
-Company names represent the characteristics of businesses and thus may act as “meta-proxy” for a bulk of intertwined economic metrics.
-Such a complex nature makes them suitable for regions identification, especially in the case of vernacular regions [@Reed1976; @Liesch2015; @Vukosav2016; @Holtkamp2018].
-Other interactions between places and brands, e.g.
-the contribution of commercial branding to place branding [@Florek2013] or its compliance with the latter [@BassolsiGardella2023] may also be of interest.
+Despite the lack of interest from geographers, the spatial aspects of the activity of law firms are worth investigating for a number of reasons.
+Firstly, legal services can be categorized as business or producer services, and this category of services is important for economic development, on both a national and regional scale [@Hansen1993; @Faulconbridge2019].
+Secondly, despite their comparatively minor direct economic impact, these services play a significant role as intermediaries and facilitators, contributing to the establishment and maintenance of the rule of law, social order, stability, and predictability of social relations.
+Thirdly, for the general public, access to legal aid is vital for the protection of human rights and the attainment of social justice.
+In this regard, the spatial aspects of law services are on par with the spatial issues of medical care [@Blacksell1988; @Blacksell1990].
 
-There are thousands of law firms in Russia [@Moiseeva2016].
-Their names are likely to reflect some properties of local legal services market.
-The area of Russia is comprised of regions (constituent entities of the federation) that provide some sort of a commonly used spatial unit for geographical studies.
-Besides, economic geographers split the country into 11 economic regions [@Leyzerovich2010].
-Each economic regions includes several constituent entities and is characterized by internal integrity, economic links between the companies, specialization and difference from the others.
-It can be hypothesized that the proximity of regions, as well as internal unity and economic links within the economic regions, leads to the similarity of law firms activity and, correspondingly, to the similarity of their naming.
-The overall aim of this study is to check whether this is true.
-To achieve this goal, a few research questions are asked and then answered based on the data on law firms names.
-These questions, or problems, are specified below.
+The study of firm names — which are sometimes also referred to as company names, brand names, or simply “brands” — is a relatively uncommon yet potentially fruitful approach to the geographical analysis of services and related problems.
+A firm name is designed to distinguish the business from competitors, to highlight the sector or geographical area in which it operates, and to convey information about the services offered by a provider and the associated benefits.
+A brand is therefore capable of transferring a substantial amount of information in a brief textual format.
+Such information may prove useful in uncovering the city's business landscape and consumer perceptions of a particular city with its distinctive services [@Raento2001].
+It may also help to analyze trends common to a specific sector in a geographic region [@Marjavaara2022].
+The incorporation of place names (toponyms) in brands influences the business development and consumer attitudes [@Peluso2012; @Zhang2014; @Erdogan2016; @Noorda2019].
+Company names encapsulate a multitude of intertwined properties relevant to the economy and culture of a given location.
+This synthesis makes them suitable for the identification of regions, especially in the case of vernacular regions [@Reed1976; @Liesch2015; @Vukosav2016; @Holtkamp2018].
 
-Question 1 (Q1) is whether the neighboring regions have more similar law firm names that regions that do not share a common border.
-Question 2 (Q2) is a continuous extension of Q1: whether the geographical proximity of regions has an impact on the similarity of law firm names, i.e.
-whether the regions that are close to each other have more similar names, and vice versa.
-Question 3 (Q3) moves up from the level of regions to the level of economic regions: whether names withing the economic region are more similar compared to the names outside the region.
-Question 4 (Q4) is what naming strategies exist in Russia and how they are spatially organized.
-Question 5 (Q5) is a follow-up from the previous: are there strategies common or different within each economic region?
+It seems reasonable to posit that the names of Russian law firms reflect some economic and, to a lesser extent, cultural characteristics of the places where these firms are located.
+The substantial number of such firms (tens of thousands according to @Moiseeva2016) appears to be sufficient for a comprehensive and nuanced representation of areas and territories.
+It can be hypothesized that there is a geographical variation in the naming of law firms across the country, which is contingent upon a number of factors.
+The geographical proximity and common borders, for example, should lead to more homogeneous naming practices.
+Similarly, economic links between places and territories are also expected to reduce the diversity of law firm names.
+In addition, it may be hypothesized that there are several law firm naming strategies, or typical models of brand construction, that potentially vary across the country.
+The primary aim of this study is to test these hypotheses.
 
@@
 
-The issue of obtaining information about firm names is challenging.
-Even though all Russian organisations are registered in the unified state registry operated by Federal Tax Service (FTS of Russia), and an extract about a particular company from this registry can be downloaded online by any person free of charge, there is no straightforward way to get names of all companies from the registry at once without paying a considerable amount of money and without negotiating a contract with the responsible government body.
-There are a few commercial firms that re-cell the data from the registry of companies, but the prices are substantial.
-To overcome these problems, a Unified state registry of small and medium-sized businesses (SMB registry) was used.
-This registry has been maintained since 2016 according to the Article 4.1 of Federal Law of July 24, 2007 No 209-FZ “On the development of small and medium-sized businesses in the Russian Federation”.
-All the companies and individual entrepreneurs that have a yearly revenue below 2 billion roubles and an average yearly number of employees below 250 are automatically included in the SMB registry without any action required on their side.
-In contrast to the main registries of companies and entrepreneurs, SMB registry can be fully downloaded at the FTS open data portal free of charge, as well as the monthly history of its updates since the beginning of the operation.
-Thus, this registry can be treated as the most comprehensive free data source with the information about Russian business entities (organisations and individuals).
-The downloadable version of the registry includes, in particular, firm names, addresses of registration (up to a settlement), and codes of activity kinds according to the All-Russia Classifier of Kinds of Economic Activity.
-Large firms are not present in the registry, so some data is missing.
-However, the majority of commercial law firms in Russia (more than 75%) can be categorized as small and medium-sized businesses and are included in the registry, so the availability of information is relatively high, and it can be used for trustworthy conclusions.
+The issue of obtaining information about firm names is a challenging one.
+All Russian organizations are registered in the state registry operated by the Federal Tax Service (FTS of Russia).
+An extract from this registry about a particular company can be searched and downloaded online by any person free of charge.
+However, a considerable sum of money must be paid and a written request submitted to the FTS of Russia to purchase a complete copy of the registry.
+Commercial companies offer data re-cell services, but the prices are similarly high.
+
+To address these challenges, a unified state registry of small and medium-sized businesses (SMB registry) may be used [@SMBRegistry2024].
+The registry has been maintained since 2016 in accordance with Article 4.1 of the Federal Law of 24 July 2007 No.
+209-FZ “On the development of small and medium-sized businesses in the Russian Federation”.
+All companies and individual entrepreneurs with a yearly revenue below 2 billion roubles and an average yearly number of employees below 250 are automatically included in the SMB registry, without any action being required on their part.
+The registry is fully downloadable at the FTS open data portal at no cost.
+Additionally, the portal provides a monthly record of updates made since the registry's inception.
+To sum it up, the SMB registry can be regarded as the most comprehensive free data source offering information on Russian business entities, both organizations and individuals.
+
+The downloadable version of the registry includes, in particular, firm names, addresses of registration (up to a settlement), and codes of activity kinds according to the National Classification of Economic Activities (NCEA).
+It should be noted that the registry does not include large firms, resulting in some data being absent.
+Nevertheless, the majority of commercial law firms in Russia (in excess of 75%) are classified as small and medium-sized businesses and thus are included in the registry.
+Consequently, the completeness of information is relatively high, and it can be used to draw reliable conclusions.
 
@@
 
+### 2.2.1. Law firms selection criteria
+
+In this study, the term “law firm” is defined as a commercial organization registered in Russia in accordance with the Civil Code of Russia, with a main activity code equal to 69.10 (activity in the area of law) according to NCEA.
+The aforementioned definition excludes individual entrepreneurs, foreign companies, non-profit organizations, advocacy firms, and notaries.
+The selection technique based on the main activity code is derived from that proposed by @Moiseeva2016.
+Individual entrepreneurs are excluded from the analysis because they typically operate under their personal names, and even if they have a brand, it is not listed in the SMB registry.
+Other types of law firms are not included in the analysis primarily due to data availability issues.
+
+The analysis covers a period of six years, from 2016 to 2021.
+The initial point in time is defined by the commencement of SMB registry data availability.
+Although data for 2022 and subsequent years are available, they have not been studied as the ongoing war may have had a significant impact on both the economy and the dissemination of information by government bodies.
+The purpose of this research is to provide conclusions about the space and names under “normal” conditions, and it would therefore be unwise to consider data collected during the period when the circumstances are away from normal.
+Organizations with a lifespan of less than three years have been omitted from the research, which are likely to be dummy companies.
+
+### 2.2.2. Firm data acquisition
+
+The processing of SMB registry data is conducted in a series of stages.
+In the initial phase of the research process, all source archives are downloaded.
+Each source archive contains a full copy of the registry data for the specified publication date, with a new archive becoming available on a monthly basis.
+Subsequently, data pertaining to law firms is extracted from the archives and stored in CSV files.
+Each CSV file contains data from one source archive.
+Subsequently, all CSV files are consolidated into a single table, the duplicated records are removed, and the remaining records are augmented with start and end temporal markers.
+Subsequently, the addresses are normalized to resolve the inconsistencies in location names and geocoded to add geographical coordinates.
+Finally, the resulting table is transformed into a panel view, with each row representing a single law firm at a given year.
+All these stages are implemented in an open-source Python command-line tool.
+
+### 2.2.3. Firm name embeddings retrieval
+
+The Yandex Foundation Models cloud service [@Yandex2024] is used to transform the firm names into text vectors, or embeddings.
+The “text-search-query” model is selected because its purpose, as stated in the documentation, is to vectorize short texts.
+Prior to vectorization, the firm names are preprocessed to remove indications of organizational form such as “limited liability company”.
+This is done because such indications are common to the majority of names and thus have little value.
+The results of the queries to the Yandex API are stored separately for reuse, given that API calls are paid and take time to be completed.
+
+### 2.2.4. Region embeddings
+
+The embeddings of firm names are grouped by regions and averaged.
+In this paper, these average vectors are referred to as region embeddings or region vectors.
+Russia is comprised of 83 internationally recognized regions (constituent entities of the Russian Federation), hence 83 region vectors are calculated.
+Each vector represents the combined meaning associated with the names of law firm names within a specific region.
+The mathematical similarity of region vectors is analogous to the similarity of law firm names in regions.
+
+Regions vectors were introduced in this study to reduce computational complexity.
+With thousands of law firms, the mathematical operations between individual names would otherwise have been unduly time-consuming.
+Furthermore, the combined vectors improve the interpretation of the results.
+A region is a prevalent subdivision used in the studies of Russian economic geography.
+Therefore, a single sequence of numbers for a single region is more readily understandable than a big table of individual name vectors.
+
+### 2.2.5. Naming strategies identification
+
+Law firm names are clustered using the KMeans technique.
+A matrix of features for clustering is constructed from the name embeddings.
+As there are 256 items in each vector, the feature matrix has 256 columns.
+A target number of clusters is determined with the elbow method and ultimately set to 50, with the understanding that further manual processing of the clusterization results will be necessary.
+Each name is assigned a cluster label between 1 and 50.
+
+The resulting clusters are subsequently subjected to manual grouping.
+A random sample of 20 names is selected from each cluster.
+With 50 clusters and 20 sampled names allocated in each, the total size of the dataset for manual processing is 1, 000.
+The cluster samples are carefully examined, and a concise description is provided for each, e.g.
+“Names with the word “консалт/консалтинг” (consult/consulting)”.
+After this, the clusters are examined once more and assigned to one or more of the four groups: “Form”, “Law”, “Service”, and “Misc”.
+The mapping of clusters and groups (e.g.
+cluster 1 — “Law”) is employed to link all the firm names to the appropriate groups.
+The concept of naming strategies is derived from the works of @Arora2015 and @BassolsiGardella2023.
+However, the specific groups differ from those used in their respective papers and are drawn from the data-driven analysis.
+
+Each group except for the “Misc” category represents a specific naming strategy.
+The “Form” strategy encompasses names that are based on a word or words denoting a kind of business entity, including “company”, “firm”, “group”, “partnership”, “center”, “alliance”, and “corporation”.
+The “Law” strategy emphasizes the firm's affiliation with the legal services market, employing terms such as “legal”, “juridical”, “lex”, “law”, and “lawyer”.
+Names belonging to the “Service” strategy may indicate a particular service provided by a law firm (“advisor”, “consulting”, “protection”, “support”, “warrant”), describe the sectors of businesses the firm is primarily engaged with (“technology”, “development”, “house management”, “accounting”), or mention services in general.
+The “Misc” category is a catch-all group for names that do not clearly align with any other category.
+
+### 2.2.6. Analysis of firm names similarity
+
+The similarity of law firm names between regions is calculated as a Euclidean distance between the corresponding region vectors.
+Henceforth, this will be referred to as a semantic distance between regions.
+A neighborhood of regions is determined on the basis of the existence of a common border between them, thus allowing for two regions to either be neighboring or not.
+A geographic distance between regions is calculated as a linear distance between the centroids of the region polygons.
+To provide a proxy of economic links between regions, the belonging to the same economic region is utilized.
+According to @Leyzerovich2010, Russia is comprised of 11 economic regions, which themselves consist of a number of regions.
+The defining characteristics of an economic region are its internal unity, distinction from the other economic regions, and the economic links that exist between constituting regions, cities, and enterprises within them.
+
+To estimate the effect of neighboring, regions are divided into pairs of neighbors and non-neighbors.
+Semantic distances are then calculated between the regions in each pair within the two groups.
+A median semantic distance is calculated for each group.
+It is expected that the median semantic distance between neighboring regions would be less than that between regions that do not share a border.
+
+A similar technique is used to ascertain whether geographical distance impacts semantic distance.
+Pairwise semantic and geographical distances between all regions are calculated, and the correlation between geographical distance and semantic distance is estimated.
+It is expected that the coefficient of correlation will be positive.
+
+To analyze the role of economic links, the pairwise semantic distances between regions are split into two groups: distances between regions within an economic region (within-distances) and distances between regions belonging to an economic region and all the other regions (outside-distances).
+The median of both within-distances and outside-distances is calculated for each economic region.
+It is expected that semantic distances between regions within an economic region would be smaller than those between regions belonging to different economic regions.
+
+### 2.2.7. Analysis of naming strategies
+
+A dominant naming strategy is ascribed to each region, defined as the naming strategy used by the majority number of law firms within that region.
+Dominant naming strategies are then mapped and analyzed visually to ascertain whether any geographical patterns emerge.
+In addition, the boundaries of economic regions are illustrated on the same map, and the dominant naming strategies of regions within each economic region are examined.
+No a priori hypotheses are formulated regarding the geographical patterns.
+As for naming strategies within economic regions, it is expected that they will be similar.
+
 # 3. Results and discussion
+
+## 3.1. Law firm names in neighboring regions are more similar
+
+The semantic distance between neighboring regions in comparison to the semantic distance between regions that do not have a common border is illustrated in Fig.
+1.
+Although the difference in medians is relatively minor, it is nevertheless present, and the overall distribution of distances between non-neighboring regions is shifted towards the area of higher values.
+Therefore, neighboring regions have more similar names of law firms than regions that do not border one another.
+
+```{r fig-1-neighbors, fig.cap="**Fig 1.** Semantic distances between neighboring and non-neighboring regions", fig.height=3}
+neighbors_plot
+```
+
+## 3.2. Greater geographical distance reduces the similarity of law firm names
+
+The relationship between geographical and semantic distances is shown in Fig.
+2.
+The correlation coefficient is relatively low but positive.
+Thus, regions that are situated close to each other tend to have a bit more similar names of law firms even in the absence of a common border between them.
+
+```{r fig-2-distance-plot, fig.cap="**Fig. 2.** Relationship between geographical and semantic distances between regions", fig.height=3}
+distance_plot
+```
+
+## 3.3. Law firm names inside an economic region are more homogeneous
+
+Within-distances and outside-distances by economic regions, as outlined in the methods section, are depicted in Fig.
+3.
+In some instances, the semantic distance between regions that are part of the same economic region is slightly less than the semantic distance between these regions and regions of other economic regions.
+However, this pattern does not apply to at least three economic regions: in Volga, North, and North Caucasus, median within-distance is equal to or greater than median outside-distance.
+Consequently, the results of this test are mixed: while links between regions constituting an economic region usually slightly increase the similarity of law firm names, there are several exceptions to this rule.
+A greater within-distance compared to outside-distance may indicate internal heterogeneity within an economic region, thereby suggesting the need for a classification update.
+
+```{r fig-3-economic-regions-distance, fig.cap="**Fig. 3.** Semantic distances between regions within and outside economic regions", fig.height=5}
+economic_regions_distances
+```
+
+## 3.4. Geographical patterns of naming strategies are unclear
+
+The map in Fig.
+4 depicts a dominant naming strategy in regions of Russia.
+As “Form” naming is used by a minority of law firms across all regions, it is not shown.
+European South of Russia and Povolzhye are characterized by a prevalence of the “Service” strategy.
+In other parts of the country, an irregular mixture of both “Law” and “Service” naming strategies is observed.
+
+```{r fig-4-naming-strategies, fig.cap="**Fig. 4.** Map of naming strategies"}
+naming_strategies
+```
+
+## 3.5. Naming strategies within economic regions are inconsistent
+
+As seen from Fig.
+4, few economic regions are characterized by a uniform naming strategy.
+In the Volga-Vyatka economic region, the “Law” strategy is the most prevalent among all the regions, with the exception of the Republic of Mari El.
+In Volga and North Caucasus economic regions, service-based strategy is the prevailing approach in the majority of regions.
+The remaining economic regions exhibit a combination of law firms naming strategies.
 
@@
 
-<div custom-style="References-title">References</div>
+The findings of the study are somewhat contradictory.
+On the one hand, several hypotheses have been proven.
+These include the following: law firm names in neighboring regions are more similar than law firm names in non-neighboring regions; the smaller the geographic distance between regions, the more similar the law firm names; and law firm names within an economic region are typically more similar compared to the law firm names outside an economic region.
+On the other hand, the difference in the metrics of similarity is relatively small in each case, as is the correlation coefficient.
+It is therefore important to note that although the postulated phenomena do indeed exist, their magnitude is weak.
+
+The three principal strategies for naming law firms are as follows: “Form”, “Law”, and “Service”.
+The first of these represents a minority approach, while the remaining majority is divided between two other strategies.
+In the European South of Russia and in the Volga economic region, “Service” is a dominant strategy.
+However, no other spatial patterns in law firms naming have been identified.
+Economic regions are typically characterized by a mixture of naming strategies, except for Volga-Vyatka, where the “Law” strategy prevails, and Volga and North Caucasus, where the “Service” strategy is dominant.
+
+The contribution of this study is primarily theoretical and methodological in nature.
+It underscores the significance of firm names as a source of information for business- and economy-related research.
+It also proposes a technique for comprehensive analysis of branding in a geographical context through the use of modern natural language processing (NLP) solutions based on artificial neural networks.
+Additionally, it advocates for the utilization of disaggregated administrative open data for a spatial analysis.
+The findings demonstrate the effect of geographical proximity, neighboring, and economic relations on the similarity of the economic and, to a certain extent, cultural properties of the territories represented in firm names.
+The results generally align with the widely accepted classification of economic regions in Russia, although they highlight a few instances where a theoretically undesired internal heterogeneity within an economic region is observed.
+Finally, the study provides an illustration of a semi-automated brand classification within a specific economic sector and discovers a geographical variation in naming strategies that could be further investigated.
+
+# Data availability
+
+# Code availability
+
+# References
 

### 2024-09-05 law-firms-names/conference-paper/Conference-slides.qmd
Adjust paths, #4

@@
 # load the data
-data <- read_csv(here("../../ru-smb-companies/legal/panel.csv"))
+data <- read_csv(here("common", "panel.csv"))
 cluster_labels <- read_csv(

### 2024-09-05 law-firms-names/journal-article/Script.R

@@ -12,7 +12,7 @@ library(units)
 sf_use_s2(FALSE) # disable to avoid errors in distances and intersections
 
 # Load the data
-data <- read_csv(here("../../ru-smb-companies/legal/panel.csv"))
+data <- read_csv(here("common", "panel.csv"))
 er <- read_csv(here("journal-article", "economic-regions.csv"))
 er_short_labels <- tibble(
   economic_region = sort(unique(er$economic_region)),
@@ -32,31 +32,30 @@ org_forms <- c(
 )
 ru <- st_read(here("common", "ru.geojson"))
 ru_crs <- st_crs("+proj=aea +lat_0=0 +lon_0=100 +lat_1=68 +lat_2=44 +x_0=0 +y_0=0 +ellps=krass +towgs84=28,-130,-95,0,0,0,0 +units=m +no_defs")
-tiles <- read_csv(here("common", "russia-tiles.csv"))
 
 # Preprocess the data
-firms <- data %>% 
+firms <- data %>%
   filter(
     kind == 1, # companies only
     year <= 2021, # 2016–2021
-  ) %>% 
+  ) %>%
   mutate(
     name_id = row_number(),
     settlement = case_when(
       region == "Москва" ~ "Москва",
       region == "Санкт-Петербург" ~ "Санкт-Петербург",
       TRUE ~ settlement)
-  ) %>% 
+  ) %>%
   select(
     name_id,
     tin,
-    name = org_name, 
-    region, 
-    settlement, 
+    name = org_name,
+    region,
+    settlement,
     lat,
-    lon, 
+    lon,
     year
-  ) %>% 
+  ) %>%
   mutate(
     name = str_to_lower(name),
     name = str_remove(name, paste0(org_forms, collapse = "|")),
@@ -81,7 +80,7 @@ vectors <- read_csv(here("common", "names-vectors.csv"))
 if (!get0("IS_PAPER", ifnotfound = FALSE)) {
   n_clusters <- 2:100
   scores <- sapply(
-    n_clusters, 
+    n_clusters,
     function(x) kmeans(
       slice_sample(select(vectors, dim_0:dim_255), n = 1000),
       centers = x)$tot.withins
@@ -95,10 +94,10 @@ set.seed(42)
 fit <- kmeans(select(vectors, dim_0:dim_255), centers = 50)
 vectors$cluster <- fit$cluster
 if (!get0("IS_PAPER", ifnotfound = FALSE)) {
[…]

### 2024-09-05 law-firms-names/common/get_yandex_gpt_vectors.py
Remove sensitive info, #4

@@ -17,8 +17,8 @@ def get_embedding(text: str) -> np.array:
     )
 
 
-FOLDER_ID = "b1guv73jumjli9it2kum"
-IAM_TOKEN = "t1.9euelZqJjcqMjsySz4yakpqbnY3Lju3rnpWam5yPi5WQks2YyMyOiorOzpjl9Pd-EzpN-e9jbBrS3fT3PkI3TfnvY2wa0s3n9euelZrMypqYk5KVmcyPjsmKyIqOkO_8xeuelZrMypqYk5KVmcyPjsmKyIqOkA.A1u_ruhFatIschBArQEDgMiZJVotdF36Tu2IjEw_cJ3UO5ruz8qQSAMbGdwK6afYF_gUg6nqJ44ns04DxVByDw"
+FOLDER_ID = ""
+IAM_TOKEN = ""
 
 query_uri = f"emb://{FOLDER_ID}/text-search-query/latest"
 embed_url = "https://llm.api.cloud.yandex.net:443/foundationModels/v1/textEmbedding"
@@ -29,7 +29,7 @@ with open("names.csv") as f:
     names = [row["name"] for row in reader]
 
 missing = []
-with open("names_vectors.csv", "w") as f:
+with open("names-vectors.csv", "w") as f:
     fieldnames = ["name"] + [f"dim_{i}" for i in range(256)]
     writer = csv.DictWriter(f, fieldnames=fieldnames)
 

### 2024-09-05 law-firms-names/Paper.Rmd
Add data and code availability statement, #4

@@
 
-**Keywords:** firm names; law firms; text embeddings; economic regions, text similarity.
+**Keywords:** firm names; law firms; text embeddings; economic regions, text similarity; large language models.
 
@@
 
-# Code availability
+A dataset about Russian law firms, vectors of law firm names obtained from Yandex Foundation models, results of manual labelling, additional files used for mapping, as well as all the code used to conduct the study and generate the paper file are available at Zenodo: https://doi.org/10.5281/zenodo.13691418.
 

### 2024-09-30 law-firms-geography/Slides.Rmd
Add slides for the conference, #1

---
title: "Юридический бизнес в России: региональный анализ по данным налоговой службы"
format: revealjs
---

```{r}
#| label: imports
#| include: false
#| cache: true
source("Script.R")
```

# Мотивация и идея исследования

1. Территориальную организацию юридического бизнеса в России (почти) никто не изучал — нужно закрыть этот пробел.
2. В данных Росстата юридические услуги не выделяются — нужен альтернативный источник информации.

# Данные {.smaller}

1. Реестр субъектов малого и среднего предпринимательства (МСП), который ведёт Федеральная налоговая служба (ФНС России):
- открытые данные с августа 2016 года;
- гранулярность — до месяца, конкретной фирмы, населённого пункта, кода ОКВЭД;
- большой (200 Гб в zip-архивах) и сложный (xml-файлы).
2. Сведения о числе работников организаций (ФНС России).
3. Вспомогательно — данные Росстата, данные о судебных органах.

# Методы обработки данных {.smaller}

1. Юридические фирмы: основной ОКВЭД = 69.10.
2. Организации и индивидуальные предприниматели (ИП).
3. Геокодирование на базе справочных таблиц.
4. Поля: наименование, адрес, вид, категория, число работников.
5. Реализация — консольное приложение на Python (https://github.com/PavelSyomin/ru-smb-companies).

# Исследовательские вопросы {.smaller}

1. Территориальная дифференциация по:
- относительному количеству юридических фирм;
- соотношению организаций и ИП;
- доле юристов в общей и региональной занятости.
2. Связь с размещением судов окружного уровня.
3. Динамика числа юридических фирм.

# Результаты

## Число юридических фирм *per capita*

```{r}
counts_map
```

## Организации *vs* ИП

```{r}
shares_map
```

## Доля юристов и суды окружного уровня

```{r}
employees_map
```

## Динамика за 2016–21 годы

```{r}
timeline_map
```

# Выводы {.smaller}

1. Ключевые регионы и районы по развитию юридического бизнеса — Москва, Санкт-Петербург; Краснодарский край, Урал и Западная Сибирь.
2. На юге России преобладает индивидуальный юридический бизнес, а в большинстве ключевых регионов — коллективный.
3. Юридический бизнес, вероятно, тяготеет к местоположению судов окружного уровня (обратное, возможно, тоже верно).
4. Типичная динамика за 2016–21 — куполообразная: рост до пандемии, потом затухание или спад.

### 2024-11-16 fts-open-data-in-economic-geography/Paper.Rmd
Update abstract, #5

@@
 
-***Аннотация.*** В статье представлена технология для создания геопривязанного табличного набора данных о субъектах малого и среднего предпринимательства (МСП) в России на основе открытых данных Федеральной налоговой службы (ФНС России).
-Получаемый набор данных охватывает всю территорию России за период с 2016 г.
-Он содержит детализированную до уровня конкретных юридических лиц и индивидуальных предпринимателей информацию о регистрационных номерах, категории, имени или наименовании, месте жительства или местонахождении, виде деятельности, доходах, расходах и численности работников.
-Статья излагает воспроизводимую методику обработки исходных данных и демонстрирует её применение для получения сведений о фирмах в сфере сельского, лесного хозяйства и рыболовства.
-Показано, что открытые данные ФНС России с помощью предложенного алгоритма и его реализации на языке программирования Python можно использовать для решения ряда исследовательских задач в области экономической географии, в особенности тех, где требуются дезагрегированные сведения или пространственная детализация до населённых пунктов.
-Благодаря географическим координатам получаемый набор данных может без дополнительной обработки использоваться для картографирования, а наличие кодов ОКТМО позволяет объединять его с официальными статистическими сведениями.
+***Аннотация.*** В статье представлена технология создания геокодированного табличного набора данных о субъектах малого и среднего предпринимательства (МСП) в России на основе трёх наборов открытых данных Федеральной налоговой службы (ФНС России).
+Получаемый в результате использования технологии набор данных охватывает всю территорию России и содержит информацию, детализированную до уровня конкретных юридических лиц и индивидуальных предпринимателей.
+Таблица включает следующие поля (колонки): ИНН, ОГРН, вид (юридическое лицо, индивидуальный предприниматель, глава крестьянского (фермерского) хозяйства), категория (микропредприятие, малое, среднее предприятие), имя или наименование, место жительства или местонахождение (субъект федерации, район, город, населённый пункт), код основного вида деятельности по ОКВЭД, доходы, расходы, среднесписочная численность работников.
+Данные о доходах, расходах и численности работников доступны за каждый год с 2018 г., остальные — за каждый месяц с августа 2016 г.
+Статья излагает воспроизводимую методику обработки исходных данных и демонстрирует её применение для получения и разведочного анализа сведений о фирмах в сфере сельского, лесного хозяйства и рыболовства.
+Дополнением к статье является открытое консольное приложение на языке программирования Python, реализующее описанную технологию.
+Показано, что открытые данные ФНС России с помощью предложенного алгоритма можно использовать для решения ряда аналитических и исследовательских задач в области экономической географии, в особенности тех, где требуются дезагрегированные сведения или пространственная детализация до населённых пунктов.
+Благодаря географическим координатам получаемый набор данных может без дополнительной обработки использоваться для картографирования.
+Наличие кодов ОКТМО позволяет объединять его с официальными статистическими сведениями.
 
@@
 
-***Abstract.*** The paper features a technology for creating a geocoded tabular dataset of small and medium-sized businesses (SMBs) in Russia based on open data from the Federal Tax Service (FTS of Russia).
-The resulting dataset covers the entire territory of Russia for the period from 2016.
-It contains information on registration numbers, category, name, address, activity type, income, expenses, and number of employees.
-The information is detailed up to the specific legal entities or individual entrepreneurs.
-The article outlines a reproducible methodology for raw data processing and demonstrates its application to obtain a dataset of firms in the agriculture, forestry and fisheries sectors.
-The paper shows that with the help of proposed technique and its reference implementation on Python programming language, FTS open data can be used to solve various tasks in economical geography, in particular tasks where the disaggregated information or spatial resolution up to settlements is required.
-In addition, thanks to the geographic coordinates, the generated dataset can be used for mapping without additional processing, and the presence of OKTMO codes allows it to be combined with official statistical information.
+***Abstract.*** The paper presents a methodology for the creation of a geocoded tabular dataset of small and medium-sized enterprises (SMEs) in Russia based on three open datasets provided by the Federal Tax Service (FTS of Russia).
+The resulting dataset encompasses the entire territory of Russia.
+The information is provided at the level of specific legal entities or individual entrepreneurs.
+The dataset is structured as a CSV file comprising the following fields: tax number, registration number, legal status (legal entity, individual entrepreneur, head of farming company), SME category (microbusiness, small or medium-sized business), name, address (region, district, city, settlement), main activity code according to NACE, income, expenses, and average list number of employees.
+The dataset includes revenue, expenses, and employee data from 2018 onwards, with yearly granularity, and all other variables from August 2016 onwards, with monthly granularity.
+The article outlines a reproducible methodology for raw data processing and demonstrates its application to obtain a dataset of firms in the agriculture, forestry and fisheries sectors and conduct its exploratory data analysis.
+A reference implementation of the described technology is provided in a form of an open-source Python command-line tool.
+The paper shows that with the help of proposed technique, FTS open data can be used to solve various analytical and academic tasks in economical geography, in particular those where the disaggregated information or spatial resolution up to settlements is required.
+In addition, the incorporation of geographic coordinates into the dataset enables direct mapping without the need for additional processing.
+The presence of OKTMO codes allows for seamless integration with official statistical information.
 
-***Keywords:*** open data, spatial analysis, spatial data, administrative data, small&medium business, business registry, Apache Spark, FTS of Russia
+***Keywords:*** open data, spatial analysis, spatial data, administrative data, small and medium-sized enterprises, business registry, Apache Spark, FTS of Russia
 

### 2024-11-18 fts-open-data-in-economic-geography/Paper.Rmd
Update abstract (once more), #5

@@
 
-***Аннотация.*** В статье представлена технология создания геокодированного табличного набора данных о субъектах малого и среднего предпринимательства (МСП) в России на основе трёх наборов открытых данных Федеральной налоговой службы (ФНС России).
-Получаемый в результате использования технологии набор данных охватывает всю территорию России и содержит информацию, детализированную до уровня конкретных юридических лиц и индивидуальных предпринимателей.
+***Аннотация.*** В статье представлена технология создания геокодированного табличного набора данных о субъектах малого и среднего предпринимательства (МСП) в России на основе открытых данных Федеральной налоговой службы (ФНС России).
+Получаемый в результате применения технологии набор данных охватывает всю территорию России и содержит информацию, детализированную до уровня конкретных юридических лиц и индивидуальных предпринимателей.
 Таблица включает следующие поля (колонки): ИНН, ОГРН, вид (юридическое лицо, индивидуальный предприниматель, глава крестьянского (фермерского) хозяйства), категория (микропредприятие, малое, среднее предприятие), имя или наименование, место жительства или местонахождение (субъект федерации, район, город, населённый пункт), код основного вида деятельности по ОКВЭД, доходы, расходы, среднесписочная численность работников.
@@
 
-***Abstract.*** The paper presents a methodology for the creation of a geocoded tabular dataset of small and medium-sized enterprises (SMEs) in Russia based on three open datasets provided by the Federal Tax Service (FTS of Russia).
-The resulting dataset encompasses the entire territory of Russia.
-The information is provided at the level of specific legal entities or individual entrepreneurs.
-The dataset is structured as a CSV file comprising the following fields: tax number, registration number, legal status (legal entity, individual entrepreneur, head of farming company), SME category (microbusiness, small or medium-sized business), name, address (region, district, city, settlement), main activity code according to NACE, income, expenses, and average list number of employees.
+***Abstract.*** This paper presents a methodology for the creation of a geocoded tabular dataset of small and medium-sized enterprises (SMEs) in Russia based on open data provided by the Federal Tax Service (FTS) of Russia.
+The resulting dataset encompasses the entire territory of the country.
+The data is provided at the level of individual SMEs.
+The dataset is structured as a CSV file comprising the following fields: tax number, registration number, legal status (juridical person, sole trader, head of peasant (farm) enterprise), SME category (microbusiness, small-sized business, medium-sized business), name, address (region, district, city, settlement), main activity code according to NACE, income, expenses, and average list number of employees.
 The dataset includes revenue, expenses, and employee data from 2018 onwards, with yearly granularity, and all other variables from August 2016 onwards, with monthly granularity.
-The article outlines a reproducible methodology for raw data processing and demonstrates its application to obtain a dataset of firms in the agriculture, forestry and fisheries sectors and conduct its exploratory data analysis.
-A reference implementation of the described technology is provided in a form of an open-source Python command-line tool.
-The paper shows that with the help of proposed technique, FTS open data can be used to solve various analytical and academic tasks in economical geography, in particular those where the disaggregated information or spatial resolution up to settlements is required.
-In addition, the incorporation of geographic coordinates into the dataset enables direct mapping without the need for additional processing.
-The presence of OKTMO codes allows for seamless integration with official statistical information.
+The article presents a reproducible methodology for the processing of raw FTS data and illustrates its application in the generation and exploratory data analysis of a dataset comprising firms in the agriculture, forestry and fisheries sectors.
+A reference implementation of the described technology is provided in the form of an open-source Python command-line tool.
+The paper demonstrates that the proposed technique enables the utilisation of FTS open data to address a range of analytical and academic tasks in the field of economic geography, particularly those benefiting from disaggregated information or requiring spatial resolution at the settlement level.
+Furthermore, the incorporation of geographic coordinates into the dataset facilitates direct mapping without the necessity for additional processing.
+The inclusion of municipal codes allows for the seamless integration with official statistical information.
 

### 2024-11-18 fts-open-data-in-economic-geography/Paper.Rmd
Clarify validation aspects, #5

@@
 Вопрос о том, можно ли использовать технологию и создаваемый с её помощью набор данных для изучения не только малого и среднего бизнеса, но и экономики в целом, является сложным.
-Как отмечалось в обзоре литературы, потребность в данных, детализированных до уровня отдельных фирм, существует, но получить такую информацию сложно или дорого.
+Как отмечалось во введении, потребность в данных, детализированных до уровня отдельных фирм, существует, но получить такую информацию сложно или дорого.
 Может быть, сведений о субъектах МСП будет достаточно, чтобы делать правдоподобные выводы об экономике в целом?
-На первый взгляд, это разумное предположение, ведь малые и средние предприятия составляют почти 98% в общем числе фирм в России, дают 43% ВВП и обеспечивают занятостью 38% работников [@Barinova2019].
-В то же время очевидно, что реестр МСП не является случайной выборкой из генеральной совокупности хозяйствующих субъектов, поскольку его элементы выбраны не случайно, а исходя из ограничений по финансовым показателям и числу сотрудников.
-Следовательно, реестр МСП позволяет делать выводы о небольших фирмах, но не позволяет распространить их на все компании.
-Во всяком случае, такой вывод следует, если оценивать репрезентативность реестра МСП как выборки исходя из процедуры её формирования, то есть рассуждать теоретически.
+На первый взгляд, это разумное предположение, ведь малые и средние предприятия дают 43% ВВП, обеспечивают занятостью 38% работников и составляют почти 98% в общем числе фирм в России [@Barinova2019].
+Столь большое пересечение выборки (реестра МСП) и генеральной совокупности (всех российских юридических лиц и индивидуальных предпринимателей) позволяет предположить, что они во многих случаях взаимозаменяемы, но не гарантирует, что в некоторых ситуациях (отдельные субъекты Российской Федерации или виды деятельности) не будут наблюдаться существенные расхождения.
+Следовательно, стоит всё же оценить репрезентативность реестра МСП количественно, в том числе с учётом субъекта Российской Федерации или вида деятельности, а не полагаться исключительно на 98-процентное совпадение с множеством всех субъектов экономической деятельности.
+
+Очевидно, что реестр МСП не является случайной выборкой из генеральной совокупности.
+Причина этого проста: он пополняется не случайно, а исходя из ограничений по финансовым показателям и числу сотрудников.
+Следовательно, если оценивать репрезентативность реестра МСП по процедуре его формирования, то есть рассуждать теоретически, то он нерепрезентативен: позволяет делать выводы о небольших фирмах, но не распространять их на все компании.
 Возможен, однако, и альтернативный, эмпирический путь: чтобы понять, является ли реестр МСП достаточно хорошим «прокси» для генеральной совокупности фирм, надо сравнить распределение фирм в нём с распределением в генеральной совокупности.
@@
 Для оценки репрезентативности реестра МСП можно использовать корреляционный анализ.
-Единая межведомственная информационно-статистическая система (ЕМИСС) содержит данные о числе организаций в разрезе субъектов России и групп ОКВЭД за 2019–2022 гг (https://fedstat.ru/indicator/58109).
+Единая межведомственная информационно-статистическая система (ЕМИСС) содержит данные о числе организаций в разрезе субъектов России и групп ОКВЭД за 2019–2022 гг.
+(https://fedstat.ru/indicator/58109).
 Такая же статистика рассчитывается на базе реестра МСП.
@@
 
-Таким образом, в большинстве случаев реестр МСП может рассматриваться как достаточно качественная замена полному списку организаций как минимум в двух аспектах: он корректно отражает географическое распределение количества организаций по субъектам Российской Федерации в подавляющем большинстве групп ОКВЭД и структуру экономики по группам ОКВЭД в значительной части субъектов.
-Это не гарантирует репрезентативность на более низком уровне (конкретные коды ОКВЭД, муниципальные образования, населённые пункты) или по другим характеристикам (число работников, прибыль).
+Таким образом, в большинстве случаев реестр МСП может рассматриваться как достаточно качественная замена полному списку организаций как минимум в двух аспектах: он корректно отражает географическое распределение количества организаций по субъектам Российской Федерации в подавляющем большинстве групп ОКВЭД и структуру экономики по большинству групп ОКВЭД в значительной части субъектов Российской Федерации.
+Результаты корреляционного анализа выглядят весьма очевидными с учётом того, что субъекты МСП — это 98% всех российских фирм, но позволяют выделить частные случаи, в которых реестр МСП слабо отражает реальность: госуправление, образование, экстерриториальная деятельность.
+Репрезентативность по количеству фирм в разрезе субъектов Российской Федерации и групп ОКВЭД не гарантирует репрезентативности на более низком уровне (конкретные коды ОКВЭД, муниципальные образования, населённые пункты) или по другим характеристикам (число работников, прибыль).
 Тем не менее, технология позволяет подготовить такой набор данных, который можно использовать — в ряде задач, в отсутствие альтернатив и при дополнительных проверках — не только при изучении малого и среднего предпринимательства как такового, но и при географическом исследовании экономической активности всех организаций.

### 2024-11-18 fts-open-data-in-economic-geography/Paper.Rmd
Minor change, #5

@@
 Таким образом, в большинстве случаев реестр МСП может рассматриваться как достаточно качественная замена полному списку организаций как минимум в двух аспектах: он корректно отражает географическое распределение количества организаций по субъектам Российской Федерации в подавляющем большинстве групп ОКВЭД и структуру экономики по большинству групп ОКВЭД в значительной части субъектов Российской Федерации.
-Результаты корреляционного анализа выглядят весьма очевидными с учётом того, что субъекты МСП — это 98% всех российских фирм, но позволяют выделить частные случаи, в которых реестр МСП слабо отражает реальность: госуправление, образование, экстерриториальная деятельность.
+Результаты корреляционного анализа выглядят весьма очевидными с учётом того, что субъекты МСП — это 98% всех российских фирм, но позволяют выделить частные случаи, в которых реестр МСП слабо отражает генеральную совокупность: госуправление, образование, экстерриториальная деятельность.
 Репрезентативность по количеству фирм в разрезе субъектов Российской Федерации и групп ОКВЭД не гарантирует репрезентативности на более низком уровне (конкретные коды ОКВЭД, муниципальные образования, населённые пункты) или по другим характеристикам (число работников, прибыль).

### 2025-01-22 law-firms-names/Paper.Rmd
Edit text, #4

@@
 ---
-title: "Application of large language models to a geographical analysis of business practices: a case of law firms naming in Russia"
+title: "Discovering and Analysing Spatial Heterogenity in Russian Law Firms Naming Using Text Embeddings"
 author: "Pavel Syomin"
@@
 
-It seems reasonable to posit that the names of Russian law firms reflect some economic and, to a lesser extent, cultural characteristics of the places where these firms are located.
-The substantial number of such firms (tens of thousands according to @Moiseeva2016) appears to be sufficient for a comprehensive and nuanced representation of areas and territories.
-It can be hypothesized that there is a geographical variation in the naming of law firms across the country, which is contingent upon a number of factors.
-The geographical proximity and common borders, for example, should lead to more homogeneous naming practices.
-Similarly, economic links between places and territories are also expected to reduce the diversity of law firm names.
-In addition, it may be hypothesized that there are several law firm naming strategies, or typical models of brand construction, that potentially vary across the country.
-The primary aim of this study is to test these hypotheses.
+It seems reasonable to posit that (1) several law firms naming strategies may be identified in Russia and (2) there is a spatial heterogeneity in the distribution of the firms with different naming strategies.
+Law firms naming strategy here is defined as a typical model, or pattern, of brand construction identified by usage of identical or semantically similar words or phrases.
+The potential viability of these two statements cannot be directly conducted from the existing research, but it relies on a common-sense observations about the similarities and differences of law firm names in and between various regions of the country.
+Law firm names are likely to reflect some economic and, to a lesser extent, cultural characteristics of the territories where these firms are located.
+Thus, given the vast area and a material heterogeneity of economical and cultural landscape of Russia, these names should not have a uniform spatial distribution either.
+The primary objective of this paper is to discover these hypothesized naming strategies and their spatial heterogeneity (if any).
+In addition, it aims to test whether the variation of law firms naming between settlements in Russia is contingent upon the geographical proximity of settlements and economic links between them.
+Building upon the Tobler's First Law of Geography (“near things are more related than distant things”) [@Tobler1970], it can be hypothesized that both these factors have a positive impact on the similarity of law firm names, because small geographic distance directly makes settlements “near”, as well as economic links make them to be “near” in economical meaning.
+Finally, due to quite a pilot nature of this paper, it also tries to handle all intriguing side observations about the law firms naming strategies and their spatial distribution in Russia that may arise from the exploratory analysis of the data, even if no robust testing can be performed.
 
@@
 However, a considerable sum of money must be paid and a written request submitted to the FTS of Russia to purchase a complete copy of the registry.
-Commercial companies offer data re-cell services, but the prices are similarly high.
+Commercial companies offer data resell services, but the prices are similarly high.
 
-To address these challenges, a unified state registry of small and medium-sized businesses (SMB registry) may be used [@SMBRegistry2024].
+To address these challenges, a unified state registry of small and medium-sized enterprises (SME registry) may be used [@SMERegistry2024].
 The registry has been maintained since 2016 in accordance with Article 4.1 of the Federal Law of 24 July 2007 No.
 209-FZ “On the development of small and medium-sized businesses in the Russian Federation”.
-All companies and individual entrepreneurs with a yearly revenue below 2 billion roubles and an average yearly number of employees below 250 are automatically included in the SMB registry, without any action being required on their part.
+As a rule of thumb, all companies and individual entrepreneurs with a yearly revenue below 2 billion roubles and an average yearly number of employees below 250 are automatically included in the SMB registry, without any action being required on their part.
+There are additional qualifications for the SME registry, e.
+g.
+absence of non-SME or foreign control such as, but these should not affect a material portion of law firms.
 The registry is fully downloadable at the FTS open data portal at no cost.
@@
 
-The downloadable version of the registry includes, in particular, firm names, addresses of registration (up to a settlement), and codes of activity kinds according to the National Classification of Economic Activities (NCEA).
-It should be noted that the registry does not include large firms, resulting in some data being absent.
-Nevertheless, the majority of commercial law firms in Russia (in excess of 75%) are classified as small and medium-sized businesses and thus are included in the registry.
+The downloadable version of the registry includes, in particular, firm names, addresses of registration (up to a settlement), and codes of activity kinds compatible with NACE Rev.2.
+It should be noted that the registry does not include several types of firms, e.g.
+foreign commercial law firms without a juridical person registered according to Russian law, non-profit organizations, bar (advocacy) establishments, large companies etc.
+This results in some data being absent.
+Nevertheless, this issue should be considered insignificant for a variety of reasons.
+Foreign law firms usually register their affiliated juridical person in Russia and thus are not omitted in the registry.
+Non-profit organisations' share on Russian legal market does not exceed 1% [@Moiseeva2016].
+The names of bar establishments are usually quite uniform, containing the kind of the establishment (cabinet of an advocate, bureau of advocates, college of advocates, juridical consultancy) accompanied with a personal name(s) of advocates or with a mention of geographical location of a particular establishment, thus their potential spatial heterogeneity is not characteristic.
+A matching of the list of the law firms obtained from the SME registry with the extract of law firms from the Russian Financial Statements Database shows that ~3,300 of the firms present in SME registry in 2023 are absent from RFSD, and ~5,200 firms present in RFSD are missing in SME registry.
+The balance of missing records is not that far from the equality, and with the total number of law firms in SME registry (~34000) the share of missing entries is below 15%.
 Consequently, the completeness of information is relatively high, and it can be used to draw reliable conclusions.
@@
 
-In this study, the term “law firm” is defined as a commercial organization registered in Russia in accordance with the Civil Code of Russia, with a main activity code equal to 69.10 (activity in the area of law) according to NCEA.
+In this study, the term “law firm” is defined as a commercial organization registered in Russia in accordance with the Civil Code of Russia, with a NACE Rev.2-compatible main activity code equal to 69.10 (activity in the area of law).
 The aforementioned definition excludes individual entrepreneurs, foreign companies, non-profit organizations, advocacy firms, and notaries.
 The selection technique based on the main activity code is derived from that proposed by @Moiseeva2016.
-Individual entrepreneurs are excluded from the analysis because they typically operate under their personal names, and even if they have a brand, it is not listed in the SMB registry.
+Individual entrepreneurs are excluded from the analysis because they typically operate under their personal names, and even if they have a brand, it is not listed in the SME registry.
 Other types of law firms are not included in the analysis primarily due to data availability issues.
 
-The analysis covers a period of six years, from 2016 to 2021.
-The initial point in time is defined by the commencement of SMB registry data availability.
-Although data for 2022 and subsequent years are available, they have not been studied as the ongoing war may have had a significant impact on both the economy and the dissemination of information by government bodies.
-The purpose of this research is to provide conclusions about the space and names under “normal” conditions, and it would therefore be unwise to consider data collected during the period when the circumstances are away from normal.
-Organizations with a lifespan of less than three years have been omitted from the research, which are likely to be dummy companies.
+The data used in the study covers a period of six years, from 2016 to 2021.
+The initial point in time is defined by the commencement of SME registry data availability.
+The finish point, 2021, is the latest year, is the latest year with “normal” conditions in Russia not impacted by military operation in Ukraine and consequent international sanctions and domestic economic and political perturbations.
+The purpose of temporal dimension is to exclude short-living law firms that are likely to be dummy companies.
+Organizations with a lifespan of less than three years have been omitted from the research, while all the remaining have been included in the further analysis.
 
@@
 
-The processing of SMB registry data is conducted in a series of stages.
+The processing of SME registry data is conducted in a series of stages.
 In the initial phase of the research process, all source archives are downloaded.
@@
 Subsequently, all CSV files are consolidated into a single table, the duplicated records are removed, and the remaining records are augmented with start and end temporal markers.
-Subsequently, the addresses are normalized to resolve the inconsistencies in location names and geocoded to add geographical coordinates.
+Subsequently, the addresses are normalized to resolve the inconsistencies in location names and geocoded up to a city/settlement level to add geographical coordinates.
+The lack of within-city location in the source data prevents from more fine-grained geocoding but does not pose a serious problem because city-level coordinates are enough for large-scale analysis on country or region level.
 Finally, the resulting table is transformed into a panel view, with each row representing a single law firm at a given year.
@@
 The results of the queries to the Yandex API are stored separately for reuse, given that API calls are paid and take time to be completed.
+The dimension of a firm name vector is 256.
+Contextual embeddings (YandexGPT) were used in favor of static ones (e.g.
+FastText), because during the preliminary analysis, they have resulted in far better clusterisation of law firm names (see the description of this step below) which is crucial for naming strategies identification.
 
-### 2.2.4. Region embeddings
+### 2.2.4. Settlement embeddings
 
-The embeddings of firm names are grouped by regions and averaged.
-In this paper, these average vectors are referred to as region embeddings or region vectors.
-Russia is comprised of 83 internationally recognized regions (constituent entities of the Russian Federation), hence 83 region vectors are calculated.
-Each vector represents the combined meaning associated with the names of law firm names within a specific region.
-The mathematical similarity of region vectors is analogous to the similarity of law firm names in regions.
-
-Regions vectors were introduced in this study to reduce computational complexity.
-With thousands of law firms, the mathematical operations between individual names would otherwise have been unduly time-consuming.
-Furthermore, the combined vectors improve the interpretation of the results.
-A region is a prevalent subdivision used in the studies of Russian economic geography.
-Therefore, a single sequence of numbers for a single region is more readily understandable than a big table of individual name vectors.
+The embeddings of firm names are grouped by settlement and averaged.
+In this paper, these average vectors are referred to as settlement embeddings or settlement vectors.
+Each vector represents the combined meaning associated with the names of law firm names within a specific settlement.
+The mathematical similarity of settlement vectors is analogous to the similarity of law firm names in settlements.
+Settlement embeddings were introduced to reduce computational complexity of working with individual vectors.
+Multiple law firms may be located in the same settlement, and with the absence of within-city variation, it is possible to treat their names vectors as a single object.
 
@@
 
-Law firm names are clustered using the KMeans technique.
-A matrix of features for clustering is constructed from the name embeddings.
-As there are 256 items in each vector, the feature matrix has 256 columns.
-A target number of clusters is determined with the elbow method and ultimately set to 50, with the understanding that further manual processing of the clusterization results will be necessary.
-Each name is assigned a cluster label between 1 and 50.
-
-The resulting clusters are subsequently subjected to manual grouping.
+The dimension of law firm names vectors is reduced from 256 to 2 with the UMAP method, and the resulting 2D vectors are clustered using the DBSCAN technique and plotted to illustrate the results.
+The clusters are subsequently subjected to manual analysis.
 A random sample of 20 names is selected from each cluster.
-With 50 clusters and 20 sampled names allocated in each, the total size of the dataset for manual processing is 1, 000.
-The cluster samples are carefully examined, and a concise description is provided for each, e.g.
-“Names with the word “консалт/консалтинг” (consult/consulting)”.
-After this, the clusters are examined once more and assigned to one or more of the four groups: “Form”, “Law”, “Service”, and “Misc”.
-The mapping of clusters and groups (e.g.
-cluster 1 — “Law”) is employed to link all the firm names to the appropriate groups.
+The names samples are carefully examined, and each cluster is attached a concise description (e.g.
+“Consulting companies”), a subcategory, and a category.
+The mapping of clusters and their respective subcategories and categories (e.g.
+cluster N — “Legal bureau” / “Lawyers”) is employed to link all the firm names to the appropriate groups.
 The concept of naming strategies is derived from the works of @Arora2015 and @BassolsiGardella2023.
 However, the specific groups differ from those used in their respective papers and are drawn from the data-driven analysis.
+The specific descriptions, subcategories and categories are provided, explained and analysed in the results section below.
 
@@
 # 3. Results and discussion
+
+## 3.1. Law firms naming strategies
+
+Law firm names clusters resulted from UMAP → DBSCAN pipeline are illustrated using scatterplot in Fig.
+1.
+The visual quality of clusterisation is high: all the clusters are compact, round, and separated by considerable gaps.
+There are no noise points, and all the names have been easily assigned to clusters.
+
+Table 1 shows cluster sizes and sample names from the clusters with manually assigned short cluster descriptions, categories and strategies.
+The quantity of clusters (52) is relatively big, and their size is not even: while the largest group of names consists of items, the smallest one includes just 22 objects.
+A substantial number of clusters (...) is comprised of a mixture of short (a single word) names that cannot be given a reasonable description.
+Their strategy is marked as “Misc” and excluded from further analysis.
+“Misc” strategy encompasses ...
+law firm names.
+All the other names are split between three primary strategies: “Helpers”, “Lawyers”, “Service providers”.
+“Helpers” are law firms that articulate aid, protection, guarantee, advise, agency in their brands.
+Such names may either directly contain words related to legal aid or protection (“Aid” and “Protection” categories) or indirectly cause safety-related connotations by using the words associated with professionals or communities that do not leave a person alone (“Guarantor”, “Expert”, “Agent”, “Partner”, “Association” categories).
+In contrast to “Helpers”, “Lawyers” make a strong emphasis on their belonging to the legal services market by using the words “law”, “lawyer” or various Russian equivalents of “law firm” phrase.
+A distinct category of law firms with this strategy is formed by organisations that use a personal branding that follow the broad pattern “(law firm) <last name of a person> with partners” where (law firm) is an optional part, and <last name of a person> is a pointer to a featured lawyer of this organisation.
+Even if “law firm” subpart of the firm name is optional, the entire pattern is strongly associated in Russia with lawyers, thus this group of names is assigned the “Lawyers” strategy.
+Finally, “Services” strategy encompasses companies with a mention of a target audience (e.g.
+“Business”) or industry (e.g.
+“Construction”), a particular kind of services (e.g.
+“Consulting”, “Expertise”, “Audit” or “Finance&Investment”) or services in general.
+Compared to two other primary strategies, this one usually do not directly mentions law and legal services market.
+It is likely that some of companies following this strategy, in fact, do not provide legal services, focusing instead on paralegal activity or using standard law firms NACE code by a mistake or as a false flag.
+It should be noted that in some cases, a category of names within the “Services” strategy point to a law firm specialization (e.g.
+consulting or real estate), so the further analysis of names may be used to distinguish between full service and specialized law firms and between particular specializations.
 

### 2025-01-22 law-firms-names/common/get_yandex_gpt_vectors.py
Improve script to get text vectors: retrieve folder id and oauth tokens from env vars, get iam token automatically, #4

@@ -1,4 +1,5 @@
 import csv
+import os
 import time
 
 import requests
@@ -6,7 +7,7 @@ import numpy as np
 import tqdm
 
 
-def get_embedding(text: str) -> np.array:
+def get_embedding(query_uri: str, embed_url: str, text: str) -> np.array:
     query_data = {
         "modelUri": query_uri,
         "text": text,
@@ -17,19 +18,35 @@ def get_embedding(text: str) -> np.array:
     )
 
 
-FOLDER_ID = ""
-IAM_TOKEN = ""
+def get_iam_token(oauth_token: str) -> str:
+    resp = requests.post(
+        "https://iam.api.cloud.yandex.net/iam/v1/tokens",
+        data=str({'yandexPassportOauthToken': oauth_token})
+    )
 
-query_uri = f"emb://{FOLDER_ID}/text-search-query/latest"
+    if not resp.ok:
+        raise RuntimeError("Failed to get IAM token for Yandex Cloud")
+
+    return resp.json()["iamToken"]
+
+
+folder_id = os.environ.get("FOLDER_ID")
+oauth_token = os.environ.get("OAUTH_TOKEN")
+
+if not (folder_id and oauth_token):
+    raise RuntimeError("FOLDER_ID and OAUTH_TOKEN env vars must be set")
+
+iam_token = get_iam_token(oauth_token)
+query_uri = f"emb://{folder_id}/text-search-query/latest"
 embed_url = "https://llm.api.cloud.yandex.net:443/foundationModels/v1/textEmbedding"
-headers = {"Content-Type": "application/json", "Authorization": f"Bearer {IAM_TOKEN}", "x-folder-id": f"{FOLDER_ID}"}
+headers = {"Content-Type": "application/json", "Authorization": f"Bearer {iam_token}", "x-folder-id": f"{folder_id}"}
 
-with open("names.csv") as f:
+with open("lemmas.csv") as f:
     reader = csv.DictReader(f)
     names = [row["name"] for row in reader]
 
 missing = []
-with open("names-vectors.csv", "w") as f:
+with open("lemmas-vectors.csv", "w") as f:
     fieldnames = ["name"] + [f"dim_{i}" for i in range(256)]
     writer = csv.DictWriter(f, fieldnames=fieldnames)
 
@@ -37,8 +54,8 @@ with open("names-vectors.csv", "w") as f:
 
     for name in tqdm.tqdm(names):
         try:
-            emb = get_embedding(name)
-        except:
[…]

### 2025-01-22 law-firms-names/journal-article/Script.R
Add more experiments, switch from KMeans to UMAP+DBSCAN, #4

@@ -1,12 +1,18 @@
+library(dbscan)
 library(dplyr)
 library(ggplot2)
 library(glue)
 library(here)
 library(readr)
 library(readxl)
+library(rjson)
+library(rulexicon)
 library(sf)
+library(stopwords)
 library(stringr)
 library(tidyr)
+library(tidytext)
+library(umap)
 library(units)
 
 sf_use_s2(FALSE) # disable to avoid errors in distances and intersections
@@ -32,6 +38,12 @@ org_forms <- c(
 )
 ru <- st_read(here("common", "ru.geojson"))
 ru_crs <- st_crs("+proj=aea +lat_0=0 +lon_0=100 +lat_1=68 +lat_2=44 +x_0=0 +y_0=0 +ellps=krass +towgs84=28,-130,-95,0,0,0,0 +units=m +no_defs")
+stopwords <- data.frame(
+  word = c(
+    stopwords("ru", "stopwords-iso"),
+    c("общество", "ограниченный", "ответственность", "акционерный")
+  )
+)
 
 # Preprocess the data
 firms <- data %>%
@@ -67,37 +79,50 @@ unique_names <- distinct(firms, name)
 # Save unique names to a file for processing with YandexGPT API
 if (!get0("IS_PAPER", ifnotfound = FALSE)) {
   write_csv(
-  unique_names,
-  glue("names-{strftime(Sys.time(), '%Y-%m-%d-%H-%M-%S')}.csv")
-)
+    unique_names,
+    glue("names-{strftime(Sys.time(), '%Y-%m-%d-%H-%M-%S')}.csv"),
+    col_names = FALSE,
+    eol = ". "
+  )
 }
 
 # Load vectors obtained from YandexGPT API
 vectors <- read_csv(here("common", "names-vectors.csv"))
 
 # Clustering
-## Find optimal number of clusters
-if (!get0("IS_PAPER", ifnotfound = FALSE)) {
-  n_clusters <- 2:100
-  scores <- sapply(
-    n_clusters,
-    function(x) kmeans(
-      slice_sample(select(vectors, dim_0:dim_255), n = 1000),
-      centers = x)$tot.withins
-  )
-  ggplot(tibble(x = n_clusters, y = scores), aes(x = x, y = y)) +
-    geom_line()
-}
+## Reduce to 2D with UMAP
+vectors_umap_res <- vectors %>% 
+  select(-name) %>% 
[…]

### 2025-01-22 law-firms-names/misc/EDA.R

@@ -6,6 +6,7 @@ library(lightgbm)
 library(plotly)
 library(readr)
 library(rjson)
+library(rulexicon)
 library(spatstat)
 library(stopwords)
 library(stringr)
@@ -27,11 +28,32 @@ names_2021 <- data %>%
     kind == 1,
     year == 2021) %>% 
   select(region, area, settlement, settlement_type, name = org_name) %>% 
-  unnest_tokens(word, name) %>% 
+  unnest_tokens(word, name, drop = FALSE) %>% 
   left_join(hash_lemmas_opencorpora, by = c("word" = "token")) %>% 
   select(-word, word = lemma) %>% 
   anti_join(stopwords) %>% 
   drop_na(region, word)
+
+count(names_2021, word, sort = TRUE) %>% write_csv("words.csv")
+
+names_2021 %>% 
+  filter(word == "земельный") %>% 
+  pull(name)
+
+names_2021 %>% select(word) %>% distinct(word) %>% write_csv("tokens.csv", col_names = FALSE)
+tvectors <- read_delim("token_vectors.csv", delim = " ") %>% select(1:101)
+tokens_umap <- umap(tvectors %>% select(-1), n_components = 2, random_state = 42)
+tvectors_pc <- data.frame(tokens_umap["layout"]) %>% rename(pc1 = layout.1, pc2 =layout.2)
+
+## Cluster with dbscan
+dbscan_res2 <- dbscan(vectors_pc2, eps = 1)
+vectors_pc2$cluster <- dbscan_res2$cluster
+clustered_names2 <- cbind(select(vectors, name), vectors_pc2)
+count(clustered_names2, cluster, sort = TRUE)
+
+tvectors_pc %>% 
+  ggplot(aes(x = pc1, y = pc2)) +
+  geom_point()
 
 top_tokens_by_region <- names_2021 %>% 
   count(region, word) %>% 
@@ -566,3 +588,120 @@ firm_names_ppp_ivd <- as.ppp(
 firm_names_ppp_ivd$marks <- factor(ppp_data$individual)
 firm_names_spp_ivd <- split(firm_names_ppp_ivd, marks(firm_names_ppp_ivd))
 plot(density(firm_names_spp_ivd, sigma = 125000))
+
+
+
+
+
+excluded_clusters <- c(
[…]

### 2025-01-25 law-firms-names/Paper.Rmd
Edit methods, #4

@@
 ---
-title: "Discovering and Analysing Spatial Heterogenity in Russian Law Firms Naming Using Text Embeddings"
+title: "Identification and Spatial Analysis of Russian Law Firms Naming Using Text Embeddings"
 author: "Pavel Syomin"
@@
 
-### 2.2.4. Settlement embeddings
-
-The embeddings of firm names are grouped by settlement and averaged.
-In this paper, these average vectors are referred to as settlement embeddings or settlement vectors.
-Each vector represents the combined meaning associated with the names of law firm names within a specific settlement.
-The mathematical similarity of settlement vectors is analogous to the similarity of law firm names in settlements.
-Settlement embeddings were introduced to reduce computational complexity of working with individual vectors.
-Multiple law firms may be located in the same settlement, and with the absence of within-city variation, it is possible to treat their names vectors as a single object.
-
-### 2.2.5. Naming strategies identification
+### 2.2.4. Naming strategies identification
 
@@
 
-Each group except for the “Misc” category represents a specific naming strategy.
-The “Form” strategy encompasses names that are based on a word or words denoting a kind of business entity, including “company”, “firm”, “group”, “partnership”, “center”, “alliance”, and “corporation”.
-The “Law” strategy emphasizes the firm's affiliation with the legal services market, employing terms such as “legal”, “juridical”, “lex”, “law”, and “lawyer”.
-Names belonging to the “Service” strategy may indicate a particular service provided by a law firm (“advisor”, “consulting”, “protection”, “support”, “warrant”), describe the sectors of businesses the firm is primarily engaged with (“technology”, “development”, “house management”, “accounting”), or mention services in general.
-The “Misc” category is a catch-all group for names that do not clearly align with any other category.
+### 2.2.5. Spatial regularities identification
 
-### 2.2.6. Analysis of firm names similarity
+Analysis of autocorrelation is used to check whether the spatial distribution of names belonging to each cluster identified on the previous step is random or has some regularities.
+Regions (constituent entries) or the Russian Federation were selected as a spatial subdivision for calculating the autocorrelation measures.
+It could have been more reasonable to use the next spatial level, municipal territories, to perform the computations, but exploratory data analysis has shown that a lion's share of such territories has no law firms, because the legal business is highly concentrated in cities and large settlements.
 
-The similarity of law firm names between regions is calculated as a Euclidean distance between the corresponding region vectors.
-Henceforth, this will be referred to as a semantic distance between regions.
-A neighborhood of regions is determined on the basis of the existence of a common border between them, thus allowing for two regions to either be neighboring or not.
-A geographic distance between regions is calculated as a linear distance between the centroids of the region polygons.
-To provide a proxy of economic links between regions, the belonging to the same economic region is utilized.
-According to @Leyzerovich2010, Russia is comprised of 11 economic regions, which themselves consist of a number of regions.
-The defining characteristics of an economic region are its internal unity, distinction from the other economic regions, and the economic links that exist between constituting regions, cities, and enterprises within them.
+The autocorrelation analysis has two branches differing in the way the spatial weights matrix is constructed.
+In the first branch, neighborhood of the regions is defined based on the adjacency of region's polygons.
+Sakhalin and Kaliningrad regions, being enclaves and thus having zero immediate neighbors, were manually connected with a few closest regions.
+In the second branch, the neighborhood of the regions was identified based on their belonging to the same economic region according to the classification widely used by economic geographers and spatial development practitioners in Russia for decades and outlined, for example, in [@Leyzerovich2010].
+All regions within the same economic region are considered to be connected, and all regions of the different economic regions are disconnected, resulting in spatial weights graph comprised of several independent components.
+Despite the fact that the concept of economic regions in Russia is sometimes considered legacy, it is still implicitly used by researchers and practitioners, and the belonging to the same economic regions implies economic linkage, similarity and cooperation of regions.
+Thus, it makes sense to utilize this classification.
 
-To estimate the effect of neighboring, regions are divided into pairs of neighbors and non-neighbors.
-Semantic distances are then calculated between the regions in each pair within the two groups.
-A median semantic distance is calculated for each group.
-It is expected that the median semantic distance between neighboring regions would be less than that between regions that do not share a border.
+For each cluster, a global Moran's I metric is calculated.
+The variable of interest is a share of law firm names belonging to the cluster in a total quantity of law firm names in a region.
+If there are no law firm names of the cluster in the region, the share is considered to equal zero.
+P-values are adjusted for multiple comparisons with Holm's correction.
+Significance level (alpha) is set to 0.05.
 
-A similar technique is used to ascertain whether geographical distance impacts semantic distance.
-Pairwise semantic and geographical distances between all regions are calculated, and the correlation between geographical distance and semantic distance is estimated.
-It is expected that the coefficient of correlation will be positive.
-
-To analyze the role of economic links, the pairwise semantic distances between regions are split into two groups: distances between regions within an economic region (within-distances) and distances between regions belonging to an economic region and all the other regions (outside-distances).
-The median of both within-distances and outside-distances is calculated for each economic region.
-It is expected that semantic distances between regions within an economic region would be smaller than those between regions belonging to different economic regions.
-
-### 2.2.7. Analysis of naming strategies
-
-A dominant naming strategy is ascribed to each region, defined as the naming strategy used by the majority number of law firms within that region.
-Dominant naming strategies are then mapped and analyzed visually to ascertain whether any geographical patterns emerge.
-In addition, the boundaries of economic regions are illustrated on the same map, and the dominant naming strategies of regions within each economic region are examined.
-No a priori hypotheses are formulated regarding the geographical patterns.
-As for naming strategies within economic regions, it is expected that they will be similar.
+For the clusters with statistically significant global Moran's I, a local Moran's I was subsequently calculated.
+P-values are adjusted with local Holm's correction.
+Cluster classification of local Moran's I are mapped.
 

### 2025-01-25 law-firms-names/Paper.Rmd
Edit methods and results for strategies identification, #4

@@
 The names samples are carefully examined, and each cluster is attached a concise description (e.g.
-“Consulting companies”), a subcategory, and a category.
-The mapping of clusters and their respective subcategories and categories (e.g.
-cluster N — “Legal bureau” / “Lawyers”) is employed to link all the firm names to the appropriate groups.
+“Consulting companies”), a category, and a strategy.
+Clusters with descriptions, categories and strategies form a tree-like hierarchy where clusters make up categories, and categories belong to strategies.
+
 The concept of naming strategies is derived from the works of @Arora2015 and @BassolsiGardella2023.
-However, the specific groups differ from those used in their respective papers and are drawn from the data-driven analysis.
-The specific descriptions, subcategories and categories are provided, explained and analysed in the results section below.
+The latter paper proposes a strategic naming model with four broad naming strategies, namely “Uniqueness”, “Belonging”, “Market”, and “No-strategy”.
+Thee strategies form a 2×2 matrix, where two pairs of “Uniqueness vs Belonging” and “Market vs No-strategy” are axes, and strategies in the pairs are opposite to each other.
+Though invented for classification of names of companies in tourism, this naming framework fits well for the purpose of law firm names grouping.
+The specific descriptions and their mapping to the categories and strategies are provided, explained and analysed in the results section below.
 
@@
 
+```{r fig-1-clusters, fig.cap="**Fig 1.** Law firm names clusters resulting from UMAP → DBSCAN processing of law firm names embeddings. Numbers are assigned automatically by DBSCAN", fig.height=4}
+clusters_plot
+```
+
+```{r table-1-clusters}
+kable(
+  clusters_info_table,
+  col.names = c(
+    "# of cluster", "# of names", "Description",
+    "Category", "Strategy"
+  )
+)
+```
+
 Table 1 shows cluster sizes and sample names from the clusters with manually assigned short cluster descriptions, categories and strategies.
-The quantity of clusters (52) is relatively big, and their size is not even: while the largest group of names consists of items, the smallest one includes just 22 objects.
-A substantial number of clusters (...) is comprised of a mixture of short (a single word) names that cannot be given a reasonable description.
-Their strategy is marked as “Misc” and excluded from further analysis.
-“Misc” strategy encompasses ...
-law firm names.
-All the other names are split between three primary strategies: “Helpers”, “Lawyers”, “Service providers”.
-“Helpers” are law firms that articulate aid, protection, guarantee, advise, agency in their brands.
-Such names may either directly contain words related to legal aid or protection (“Aid” and “Protection” categories) or indirectly cause safety-related connotations by using the words associated with professionals or communities that do not leave a person alone (“Guarantor”, “Expert”, “Agent”, “Partner”, “Association” categories).
-In contrast to “Helpers”, “Lawyers” make a strong emphasis on their belonging to the legal services market by using the words “law”, “lawyer” or various Russian equivalents of “law firm” phrase.
+The quantity of clusters (52) is relatively big, and their size is not even: while the largest group of names consists of items, the smallest one includes just 12 objects.
+Almost 1/3 of clusters (17) is comprised of a mixture of short (usually a single-word) names that cannot be given a reasonable description.
+They are marked as “Misc” and classified as “No-strategy”.
+
+All the other names are split between three primary strategies.
+“Market” strategy is comprised of firms with names having generally positive connotations in Russian language that help to pursue a potential client in a high quality of legal aid and thus act as a some sort of advertisement (“Guarantor”, “Expert”, “Partnership”, “Helper”, "Center” categories and some others).
+“Belonging” strategy, as its name suggests, make a strong emphasis on their belonging to the legal services market by using the words “law”, “lawyer”, “legal” or various Russian equivalents of “law firm” phrase.
 A distinct category of law firms with this strategy is formed by organisations that use a personal branding that follow the broad pattern “(law firm) <last name of a person> with partners” where (law firm) is an optional part, and <last name of a person> is a pointer to a featured lawyer of this organisation.
-Even if “law firm” subpart of the firm name is optional, the entire pattern is strongly associated in Russia with lawyers, thus this group of names is assigned the “Lawyers” strategy.
-Finally, “Services” strategy encompasses companies with a mention of a target audience (e.g.
-“Business”) or industry (e.g.
-“Construction”), a particular kind of services (e.g.
-“Consulting”, “Expertise”, “Audit” or “Finance&Investment”) or services in general.
-Compared to two other primary strategies, this one usually do not directly mentions law and legal services market.
-It is likely that some of companies following this strategy, in fact, do not provide legal services, focusing instead on paralegal activity or using standard law firms NACE code by a mistake or as a false flag.
-It should be noted that in some cases, a category of names within the “Services” strategy point to a law firm specialization (e.g.
-consulting or real estate), so the further analysis of names may be used to distinguish between full service and specialized law firms and between particular specializations.
+Even if “law firm” subpart of the firm name is optional, the entire pattern is strongly associated in Russia with lawyers, thus this type of firm naming marks the belonging to the legal services sphere.
+Finally, “Uniqueness” strategy encompasses companies with names mentioning a particular kind of legal services (e.g.
+“Consulting”), or a target customers group (e.g.
+“Business”), or a related area of specialization (real estate, finance, audit, development), or a specific geographic location.
+
+```{r table-2-strategies-summary}
+kable(
+  strategy_counts,
+  digits = c(NA, 0, 0),
+  col.names = c("Strategy", "# of names", "%")
+)
+```
+
+Among the strategies, “No-strategy” (i.e.
+the absence of a clear strategy) is the most popular (see Table 2).
+It encompasses more than 1/3 of all the studied names.
+Belonging strategy is the second by the popularity holding 30% of law firms.
+Uniqueness strategy is employed by approximately 20% of companies, and the least typical one, market strategy, is used by slightly more than 10% of law firms.
 
@@
 Therefore, neighboring regions have more similar names of law firms than regions that do not border one another.
-
-```{r fig-1-neighbors, fig.cap="**Fig 1.** Semantic distances between neighboring and non-neighboring regions", fig.height=3}
-neighbors_plot
-```
 

### 2025-01-25 law-firms-names/journal-article/Script.R

@@ -3,6 +3,8 @@ library(dplyr)
 library(ggplot2)
 library(glue)
 library(here)
+library(igraph)
+library(RColorBrewer)
 library(readr)
 library(readxl)
 library(rjson)
@@ -110,21 +112,27 @@ vectors_2d <- data.frame(umap_res["layout"]) %>%
 dbscan_res <- dbscan(vectors_2d, eps = 3)
 clustered_names <- data.frame(
   name = vectors$name, 
-  cluster = dbscan_res$cluster
+  cluster = dbscan_res$cluster,
+  vectors_2d
 )
 count(clustered_names, cluster, sort = TRUE)
 
-# Visualize clusters
-vectors_2d %>% 
-  ggplot(aes(x = pc1, y = pc2)) +
-  geom_point()
-
-# Visualize cluster centers with numbers
-clustered_names %>% 
+## Visualize cluster centers with numbers
+clusters_plot <- clustered_names %>% 
   group_by(cluster) %>% 
   summarise(pc1 = mean(pc1), pc2 = mean(pc2)) %>% 
-  ggplot(aes(x = pc1, y = pc2, label = cluster)) +
-  geom_text()
+  ggplot(aes(
+    x = pc1, y = pc2,
+    label = str_pad(cluster, 2, "right")
+  )) +
+  geom_point(
+    data = clustered_names, 
+    aes(x = pc1, y = pc2),
+    size = .5,
+  ) +
+  geom_text(hjust = -.5) +
+  labs(x = "Component # 1", y = "Component #2") +
+  theme_minimal()
 
 # Save a sample for manual analysis
 if (!get0("IS_PAPER", ifnotfound = FALSE)) {
@@ -151,6 +159,40 @@ clustered <- clustered_names %>%
     cluster, strategy) %>%
   drop_na(name, tin, region, settlement, lat, lon, year) %>%
   distinct(name, tin, .keep_all = TRUE)
+
+# Table with clusters info
+clusters_info_table <- count(clustered_names, cluster) %>% 
+  left_join(cluster_labels)
+
+cluster_labels %>% filter(strategy == "Uniqueness")
+
+strategy_counts <- count(clusters_info_table, strategy, wt = n, sort = TRUE) %>% 
+  mutate(share = 100 * n / sum(n))
+
+count(clustered, region, strategy) %>% 
+  complete(region, strategy, fill = list(n = 0)) %>% 
+  group_by(region) %>% 
[…]

### 2025-01-26 law-firms-names/Paper.Rmd
Numerous edits related mainly to autocorrelation tests and their results, #4

@@
 ---
-title: "Identification and Spatial Analysis of Russian Law Firms Naming Using Text Embeddings"
+title: "Identification and Spatial Analysis of Russian Law Firms Naming Patterns Using Text Embeddings"
 author: "Pavel Syomin"
@@
 
+Law firm names may follow a number of patterns.
+A pattern can be defined as as a typical model, or pattern, of brand construction identified by usage of identical or semantically similar words or phrases.
+For example, names with the words “law firm” or “law company” follow one pattern, and names with the words “legal aid” are an example of another pattern.
+Patterns may be grouped multiple times forming a tree-like classification [@Anora2015].
+
 It seems reasonable to posit that (1) several law firms naming strategies may be identified in Russia and (2) there is a spatial heterogeneity in the distribution of the firms with different naming strategies.
-Law firms naming strategy here is defined as a typical model, or pattern, of brand construction identified by usage of identical or semantically similar words or phrases.
+Law firms naming strategy here is defined .
 The potential viability of these two statements cannot be directly conducted from the existing research, but it relies on a common-sense observations about the similarities and differences of law firm names in and between various regions of the country.
@@
 
-## 3.1. Law firm names in neighboring regions are more similar
+## 3.2. Spatial regularities in law firms naming
 
-The semantic distance between neighboring regions in comparison to the semantic distance between regions that do not have a common border is illustrated in Fig.
-1.
-Although the difference in medians is relatively minor, it is nevertheless present, and the overall distribution of distances between non-neighboring regions is shifted towards the area of higher values.
-Therefore, neighboring regions have more similar names of law firms than regions that do not border one another.
+As Tables 3 and 4 demonstrate, there are few law firm names clusters with some regularities in the spatial distribution of law firm names.
+These two tables display results only for the clusters where global Moran's I coefficient is statistically significant.
+The values of Moran's I metric for all the other clusters are close to zero, and their adjusted p-value is above the selected threshold of 0.05, so the results for these clusters are not shown for brevity.
 
-## 3.2. Greater geographical distance reduces the similarity of law firm names
-
-The relationship between geographical and semantic distances is shown in Fig.
-2.
-The correlation coefficient is relatively low but positive.
-Thus, regions that are situated close to each other tend to have a bit more similar names of law firms even in the absence of a common border between them.
-
-```{r fig-2-distance-plot, fig.cap="**Fig. 2.** Relationship between geographical and semantic distances between regions", fig.height=3}
-distance_plot
+```{r table-3-moran-i-by-neighborhood}
+moran_test_by_regions %>% 
+  filter(p_adj < 0.05) %>% 
+  kable(
+    digits = c(0, NA, 2, 3),
+    col.names = c("# of cluster", "Description", "Moran's I", "Adj. p-value")
+  )
 ```
 
-## 3.3. Law firm names inside an economic region are more homogeneous
-
-Within-distances and outside-distances by economic regions, as outlined in the methods section, are depicted in Fig.
-3.
-In some instances, the semantic distance between regions that are part of the same economic region is slightly less than the semantic distance between these regions and regions of other economic regions.
-However, this pattern does not apply to at least three economic regions: in Volga, North, and North Caucasus, median within-distance is equal to or greater than median outside-distance.
-Consequently, the results of this test are mixed: while links between regions constituting an economic region usually slightly increase the similarity of law firm names, there are several exceptions to this rule.
-A greater within-distance compared to outside-distance may indicate internal heterogeneity within an economic region, thereby suggesting the need for a classification update.
-
-```{r fig-3-economic-regions-distance, fig.cap="**Fig. 3.** Semantic distances between regions within and outside economic regions", fig.height=5}
-economic_regions_distances
+```{r table-4-moran-i-by-economic-regions}
+moran_test_by_er %>% 
+  filter(p_adj < 0.05) %>% 
+  kable(
+    digits = c(0, NA, 2, 3),
+    col.names = c("# of cluster", "Description", "Moran's I", "Adj. p-value")
+  )
 ```
 
-## 3.4. Geographical patterns of naming strategies are unclear
+According to the autocorrelation analysis results in both branches, there are in total four law firm names clusters where the spatial distribution is different from the random.
+One of these clusters, “Center or company with geographical location”, is present in both tables and is characterized by the highest value of Moran's I, that is about 0.66.
+Thus, it is likely that law firms with names mentioning a geographical location of the country are not distributed randomly within the country, but situated in a number of neighbouring regions.
+Two maps on the Fig.
+2, despite several differences, demonstrate that it is the regions in Eastern and Western Siberia that are characterized by the highest concentration of “geographically referenced” law firm names.
+A brief manual investigation uncovers that such names usually reference a particular region (e.g.
+Altai), or a specific city (e.g.
+Krasnoyarsk), or entire geographical area as a whole (Siberia).
+This argues for the presence of some sort of local or regional identity which is of highest expression in Siberia.
+It is possible that similar geo-based naming is present in other parts of the country, but the automatic clusterisation of law firm names using their embeddings cannot handle them properly — either because the quantity of geolocated names in other parts of the country is substantially smaller or because the text vectors do not reflect the similarity between the Siberian geographical mentions and geographical mentions in the other parts of Russia.
+Overall, an issue of geolocated law firm names is intriguing and requires further investigation.
 
-The map in Fig.
-4 depicts a dominant naming strategy in regions of Russia.
-As “Form” naming is used by a minority of law firms across all regions, it is not shown.
-European South of Russia and Povolzhye are characterized by a prevalence of the “Service” strategy.
-In other parts of the country, an irregular mixture of both “Law” and “Service” naming strategies is observed.
-
-```{r fig-4-naming-strategies, fig.cap="**Fig. 4.** Map of naming strategies"}
-naming_strategies
-```
-
-## 3.5. Naming strategies within economic regions are inconsistent
-
-As seen from Fig.
-4, few economic regions are characterized by a uniform naming strategy.
-In the Volga-Vyatka economic region, the “Law” strategy is the most prevalent among all the regions, with the exception of the Republic of Mari El.
-In Volga and North Caucasus economic regions, service-based strategy is the prevailing approach in the majority of regions.
-The remaining economic regions exhibit a combination of law firms naming strategies.
+As for other clusters with statistically significant global Moran's I (“Forensic” names, “Experts and advisors”, and personal-branding pattern), the hotspots are situated in the European South of Russia.
+In the case of “Forensic” names, it is likely to be attributed to a tiny set of national republics of North Caucasus that are probably outliers, because their overall number of law firms and the distribution of naming strategies is substantially different from the one in the majority of regions.
+“Experts and advisors” cluster demonstrates hotspots in Rostov, Volgograd and Astrakhan regions.
+They are characterized by a relatively high number of law firms and thus cannot be left to as just outliers.
+However, the exact reasons for the spatial concentration of this naming pattern in these regions remains unclear.
+Finally, a personal branding cluster represents a drastic split of the European South into two groups of regions: the first one comprising of several national republics with high level of personal branding names and the second one consisting of some other national republics and all the regular regions with small share of such names.
+This may be seen as a slight argument for a position that European South of Russia is a conglomerate of two quite distinct groups of regions.
 

### 2025-01-26 law-firms-names/journal-article/Script.R

@@ -23,10 +23,7 @@ sf_use_s2(FALSE) # disable to avoid errors in distances and intersections
 # Load the data
 data <- read_csv(here("common", "panel.csv"))
 er <- read_csv(here("journal-article", "economic-regions.csv"))
-er_short_labels <- tibble(
-  economic_region = sort(unique(er$economic_region)),
-  label = c("C", "CC", "ES", "FE", "N", "NC", "NW", "U", "V", "VV", "WS")
-)
+fd <- read_csv(here("journal-article/federal-districts.csv"))
 org_forms <- c(
   "общество с ограниченной ответственностью",
   "общество с дополнительной ответственностью",
@@ -164,8 +161,6 @@ clustered <- clustered_names %>%
 clusters_info_table <- count(clustered_names, cluster) %>% 
   left_join(cluster_labels)
 
-cluster_labels %>% filter(strategy == "Uniqueness")
-
 strategy_counts <- count(clusters_info_table, strategy, wt = n, sort = TRUE) %>% 
   mutate(share = 100 * n / sum(n))
 
@@ -194,35 +189,49 @@ g <- create_graph(nodes_df = nodes, edges = combine_edfs(edges1, edges2)) %>%
 get_global_graph_attr_info(g)
 render_graph(g)
 
+library(ggraph)
+g <- rbind(
+  count(clusters_info_table, strategy, wt = n) %>% mutate(from = "Parent") %>% select(from, to = strategy, n),
+  count(clusters_info_table, category, strategy, wt = n) %>% select(from = strategy, to = category, n),
+  select(clusters_info_table, from = category, to = cluster, n)
+) %>% 
+  #select(-n) %>% 
+  graph_from_data_frame()
+ggraph(g, layout = 'dendrogram') + 
+  geom_edge_diagonal(aes(color = n)) +
+  geom_node_text(aes(label = name), repel = TRUE) +
+  geom_node_point() +
+  scale_edge_color_stepsn() +
+  theme_void() +
+  coord_flip()
+
 # Spatial autocorrelation
-regions_w <- poly2nb(regions) %>% 
+## Various neighbours identification techniques
+## and corresponding weight matrices
+### By geometry intersection
[…]

### 2025-01-27 law-firms-names/Paper.Rmd
Revert from umap-dbscan to kmeans, reassign strategies and categories, switch from clusters to categories, make a dendrogram of categories with strategies and a map of localmoran i, make other changes, delete useless code from script, #4

@@
 ---
-title: "Identification and Spatial Analysis of Russian Law Firms Naming Patterns Using Text Embeddings"
+title: "Identification and Spatial Analysis of Law Firms Naming Patterns in Russia Using Text Embeddings"
 author: "Pavel Syomin"
@@
 Patterns may be grouped multiple times forming a tree-like classification [@Anora2015].
+Such classification may become a basis for identifying naming strategies.
+Compared to naming patterns, naming strategies are broader groups of firm names that are usually based on sharing a common purpose — e.g.
+emphasis the belonging of a company to a certain activity type.
+@BassolsiGardella2023 argue for a two-dimensional Strategic Naming Model (SNM) that includes two pairs of strategies: belonging and uniqueness, no-strategy and market.
+Pairs may be represented as X and Y axes of a plot, respectively, with paired strategies taking opposite positions on these axes: belonging is an opponent of uniqueness, market is an opponent for no-strategy.
+Authors of the aforementioned paper discover a difference in naming strategies composition of tourist industry firms in three Columbia cities.
+Probably such a variation is not be limited to Columbia cities and tourism and can be found in the case with law firms in Russia.
 
-It seems reasonable to posit that (1) several law firms naming strategies may be identified in Russia and (2) there is a spatial heterogeneity in the distribution of the firms with different naming strategies.
-Law firms naming strategy here is defined .
+The aim of this research is twofold.
+First, it tries to identify law firms naming patterns in Russia and categorize them following the framework of SNM.
+Second, it seeks to check whether there are regularities in the spatial distribution of law firms naming patterns.
 The potential viability of these two statements cannot be directly conducted from the existing research, but it relies on a common-sense observations about the similarities and differences of law firm names in and between various regions of the country.
-Law firm names are likely to reflect some economic and, to a lesser extent, cultural characteristics of the territories where these firms are located.
-Thus, given the vast area and a material heterogeneity of economical and cultural landscape of Russia, these names should not have a uniform spatial distribution either.
-The primary objective of this paper is to discover these hypothesized naming strategies and their spatial heterogeneity (if any).
-In addition, it aims to test whether the variation of law firms naming between settlements in Russia is contingent upon the geographical proximity of settlements and economic links between them.
-Building upon the Tobler's First Law of Geography (“near things are more related than distant things”) [@Tobler1970], it can be hypothesized that both these factors have a positive impact on the similarity of law firm names, because small geographic distance directly makes settlements “near”, as well as economic links make them to be “near” in economical meaning.
-Finally, due to quite a pilot nature of this paper, it also tries to handle all intriguing side observations about the law firms naming strategies and their spatial distribution in Russia that may arise from the exploratory analysis of the data, even if no robust testing can be performed.
+According to the famous Tobler's First Law of Geography, “near things are more related than distant things” [@Tobler1970], so there is a chance that law firms naming patterns may form spatial clusters and hotspots that at least a bit reflect some economic and, to a lesser extent, cultural similarities of the territories where these firms are located.
+Given the vast area and a material heterogeneity of economical and cultural landscape of Russia, these names should not have a uniform spatial distribution.
+However, it is just a hypothesis that needs testing.
 
@@
 The names of bar establishments are usually quite uniform, containing the kind of the establishment (cabinet of an advocate, bureau of advocates, college of advocates, juridical consultancy) accompanied with a personal name(s) of advocates or with a mention of geographical location of a particular establishment, thus their potential spatial heterogeneity is not characteristic.
-A matching of the list of the law firms obtained from the SME registry with the extract of law firms from the Russian Financial Statements Database shows that ~3,300 of the firms present in SME registry in 2023 are absent from RFSD, and ~5,200 firms present in RFSD are missing in SME registry.
+A matching of the list of the law firms obtained from the SME registry and active in 2023 with the extract of law firms from the Russian Financial Statements Database for the same year shows that ~3,300 of the firms present in SME registry are absent from RFSD, and ~5,200 firms present in RFSD are missing in SME registry.
 The balance of missing records is not that far from the equality, and with the total number of law firms in SME registry (~34000) the share of missing entries is below 15%.
@@
 Organizations with a lifespan of less than three years have been omitted from the research, while all the remaining have been included in the further analysis.
+In addition, the utilization of multi-year data acts as a form of averaging, smoothing the potential distortions of spatial structure resulting from the random year-to-year changes of active law firms list.
 
@@
 
-The autocorrelation analysis has two branches differing in the way the spatial weights matrix is constructed.
-In the first branch, neighborhood of the regions is defined based on the adjacency of region's polygons.
+Six spatial weights matrices are used to ensure the robustness of the results.
+Three of these matrices are constructed with K nearest neighbours technique based on the location of region's centroids, where K equals 3, 4 or 6, respectively.
+Another matrix also uses geographical properties of regions, but instead of centroids, it relies on the existence of common border lines or points to identify neighbouring regions.
 Sakhalin and Kaliningrad regions, being enclaves and thus having zero immediate neighbors, were manually connected with a few closest regions.
-In the second branch, the neighborhood of the regions was identified based on their belonging to the same economic region according to the classification widely used by economic geographers and spatial development practitioners in Russia for decades and outlined, for example, in [@Leyzerovich2010].
-All regions within the same economic region are considered to be connected, and all regions of the different economic regions are disconnected, resulting in spatial weights graph comprised of several independent components.
-Despite the fact that the concept of economic regions in Russia is sometimes considered legacy, it is still implicitly used by researchers and practitioners, and the belonging to the same economic regions implies economic linkage, similarity and cooperation of regions.
-Thus, it makes sense to utilize this classification.
+Two remaining matrices treat regions belonging to the same upper-level metaregion as neighbours, and all the other regions as non-neighbours.
+For one of these two matrices, economic regions act as metaregions, and for another, federal districts are used in this fashion.
+Economic regions are defined according to the classification widely used by economic geographers and spatial development practitioners in Russia for decades and outlined, for example, in [@Leyzerovich2010].
+Federal districts composition is defined in the Executive Order of the President of Russia No.
+849, Dated May 13, 2000, “On the Plenipotentiary Representative of the President of the Russian Federation in the Federal District” (with up-to-date amendments).
 
@@
 Cluster classification of local Moran's I are mapped.
+
+### 2.2.6. Data statistics
+
+There are `r initial_firms_count` law firms in the SMB registry in the selected period with `r nrow(unique_names)` unique firm names.
+From the initial dataset, `r valid_firms_count` law firms have existed for at least three years and have been selected for the analysis.
+Their names sometimes are duplicated, and the number of unique law firm names being analysed is `r`.
 

### 2025-01-27 law-firms-names/journal-article/Script.R

@@ -1,28 +1,20 @@
-library(dbscan)
 library(dplyr)
 library(ggplot2)
-library(glue)
 library(here)
 library(igraph)
-library(RColorBrewer)
+library(ggraph)
+library(knitr)
 library(readr)
 library(readxl)
-library(rjson)
-library(rulexicon)
 library(sf)
 library(spdep)
-library(stopwords)
-library(stringr)
 library(tidyr)
-library(tidytext)
-library(umap)
-library(units)
 
 sf_use_s2(FALSE) # disable to avoid errors in distances and intersections
 
 # Load the data
-data <- read_csv(here("common", "panel.csv"))
-er <- read_csv(here("journal-article", "economic-regions.csv"))
+data <- read_csv(here("common/panel.csv"))
+er <- read_csv(here("journal-article/economic-regions.csv"))
 fd <- read_csv(here("journal-article/federal-districts.csv"))
 org_forms <- c(
   "общество с ограниченной ответственностью",
@@ -43,7 +35,6 @@ regions <- st_read(here("common", "regions.geojson")) %>%
   ) %>% 
   arrange(code) %>% 
   select(code, name)
-municipal_boundaries <- st_read(here("ru-mun-gadm.geojson"))
 ru_crs <- st_crs("+proj=aea +lat_0=0 +lon_0=100 +lat_1=68 +lat_2=44 +x_0=0 +y_0=0 +ellps=krass +towgs84=28,-130,-95,0,0,0,0 +units=m +no_defs")
 stopwords <- data.frame(
   word = c(
@@ -52,7 +43,7 @@ stopwords <- data.frame(
   )
 )
 
-# Preprocess the data
+# Prepare firms data
 firms <- data %>%
   filter(
     kind == 1, # companies only
@@ -82,6 +73,8 @@ firms <- data %>%
     name = str_remove_all(name, fixed("\"")),
     name = str_trim(name)
   )
+
+initial_firms_count <- firms %>% distinct(tin) %>% drop_na() %>% nrow()
 unique_names <- distinct(firms, name)
 
 # Save unique names to a file for processing with YandexGPT API
@@ -90,60 +83,45 @@ if (!get0("IS_PAPER", ifnotfound = FALSE)) {
     unique_names,
     glue("names-{strftime(Sys.time(), '%Y-%m-%d-%H-%M-%S')}.csv"),
     col_names = FALSE,
-    eol = ". "
   )
 }
 
 # Load vectors obtained from YandexGPT API
[…]

### 2025-01-28 law-firms-names/Paper.Rmd
Edit introduction and some parts of methods, #4

@@
 
-**Keywords:** firm names; law firms; text embeddings; economic regions, text similarity; large language models.
+**Keywords:** firm names; law firms; text embeddings; spatial analysis, autocorrelation, naming strategies.
 
@@
 
-Law firm names may follow a number of patterns.
-A pattern can be defined as as a typical model, or pattern, of brand construction identified by usage of identical or semantically similar words or phrases.
-For example, names with the words “law firm” or “law company” follow one pattern, and names with the words “legal aid” are an example of another pattern.
-Patterns may be grouped multiple times forming a tree-like classification [@Anora2015].
-Such classification may become a basis for identifying naming strategies.
-Compared to naming patterns, naming strategies are broader groups of firm names that are usually based on sharing a common purpose — e.g.
-emphasis the belonging of a company to a certain activity type.
-@BassolsiGardella2023 argue for a two-dimensional Strategic Naming Model (SNM) that includes two pairs of strategies: belonging and uniqueness, no-strategy and market.
-Pairs may be represented as X and Y axes of a plot, respectively, with paired strategies taking opposite positions on these axes: belonging is an opponent of uniqueness, market is an opponent for no-strategy.
-Authors of the aforementioned paper discover a difference in naming strategies composition of tourist industry firms in three Columbia cities.
-Probably such a variation is not be limited to Columbia cities and tourism and can be found in the case with law firms in Russia.
+Law firm names can follow different patterns.
+A naming pattern can be defined as as a typical way of branding, identified by the use of identical or semantically similar words or phrases.
+For instance, the names “Law firm “Femida” and “Law company “Alibi” follow one pattern, while “Centre of legal aid and support” is an example of another.
+Patterns can be grouped several times to form a hierarchical classification [@Anora2015].
+Such a classification can help to identify naming strategies.
+Compared to patterns, strategies are broader groups of names, usually based on a common purpose, e.g.
+highlighting a company's affiliation to a certain activity (retail, banking, transport, education, etc).
+@BassolsiGardella2023 propose a two-dimensional Strategic Naming Model (SNM) that includes two pairs of opposing strategies: belonging and uniqueness, no-strategy and market.
+They discover a difference between three Colombian cities in the naming strategies of their tourism companies.
+It is likely that such a variation is not be limited to the cities of Colombia and the tourism industry and can also be found in the case of law firms in Russia.
 
 The aim of this research is twofold.
-First, it tries to identify law firms naming patterns in Russia and categorize them following the framework of SNM.
-Second, it seeks to check whether there are regularities in the spatial distribution of law firms naming patterns.
-The potential viability of these two statements cannot be directly conducted from the existing research, but it relies on a common-sense observations about the similarities and differences of law firm names in and between various regions of the country.
-According to the famous Tobler's First Law of Geography, “near things are more related than distant things” [@Tobler1970], so there is a chance that law firms naming patterns may form spatial clusters and hotspots that at least a bit reflect some economic and, to a lesser extent, cultural similarities of the territories where these firms are located.
-Given the vast area and a material heterogeneity of economical and cultural landscape of Russia, these names should not have a uniform spatial distribution.
-However, it is just a hypothesis that needs testing.
+First, it attempts to identify the naming patterns of law firms in Russia and to categorise them according to the SNM framework.
+Second, it seeks to test whether there are any regularities in the spatial distribution of these patterns.
+The potential viability of these two propositions cannot be derived directly from the existing research, but is based on common sense observations of the similarities and differences of law firm naming in and between different regions of the country.
+It is highly unlikely that the names of law firms in Russia are so diverse as to be impossible to classify.
+Their geographical distribution is a much more difficult issue.
+At first glance, there are no obvious factors that could lead lawyers in a particular region to choose a common naming pattern for their firms.
+However, according to the famous Tobler's First Law of Geography, “near things are more related than distant things” [@Tobler1970], there is a possibility that law firm naming patterns may form spatial clusters and hotspots that reflect, at least to some extent, economic and cultural similarities of the territories in which these firms are located.
+There is an evidence of regional variation of branding in food industry [@Goroshko2022] and wine production [@Klimenko2023] in Russia.
+Even though the market of legal services seems to be far less dependant on the properties of place, the question of spatial regularities in naming patterns is also intriguing and worth testing.
 
@@
 
-## 2.1. Data
+## 2.1. Background
+
+## 2.2. Data
 
@@
 209-FZ “On the development of small and medium-sized businesses in the Russian Federation”.
-As a rule of thumb, all companies and individual entrepreneurs with a yearly revenue below 2 billion roubles and an average yearly number of employees below 250 are automatically included in the SMB registry, without any action being required on their part.
-There are additional qualifications for the SME registry, e.
-g.
-absence of non-SME or foreign control such as, but these should not affect a material portion of law firms.
+As a rule of thumb, all companies and individual entrepreneurs with a yearly revenue below 2 billion roubles and an average yearly number of employees below 250 are automatically included in the SME registry, without any action being required on their part.
+There are additional qualifications for the SME registry, such as the absence of non-SME or foreign control, but these are unlikely to affect a significant proportion of law firms.
 The registry is fully downloadable at the FTS open data portal at no cost.
 Additionally, the portal provides a monthly record of updates made since the registry's inception.
-To sum it up, the SMB registry can be regarded as the most comprehensive free data source offering information on Russian business entities, both organizations and individuals.
+To sum it up, the SMB registry can be regarded as the most comprehensive free data source offering information on Russian business entities, both organisations and individuals.
 
-The downloadable version of the registry includes, in particular, firm names, addresses of registration (up to a settlement), and codes of activity kinds compatible with NACE Rev.2.
-It should be noted that the registry does not include several types of firms, e.g.
-foreign commercial law firms without a juridical person registered according to Russian law, non-profit organizations, bar (advocacy) establishments, large companies etc.
-This results in some data being absent.
-Nevertheless, this issue should be considered insignificant for a variety of reasons.
-Foreign law firms usually register their affiliated juridical person in Russia and thus are not omitted in the registry.
-Non-profit organisations' share on Russian legal market does not exceed 1% [@Moiseeva2016].
-The names of bar establishments are usually quite uniform, containing the kind of the establishment (cabinet of an advocate, bureau of advocates, college of advocates, juridical consultancy) accompanied with a personal name(s) of advocates or with a mention of geographical location of a particular establishment, thus their potential spatial heterogeneity is not characteristic.
-A matching of the list of the law firms obtained from the SME registry and active in 2023 with the extract of law firms from the Russian Financial Statements Database for the same year shows that ~3,300 of the firms present in SME registry are absent from RFSD, and ~5,200 firms present in RFSD are missing in SME registry.
-The balance of missing records is not that far from the equality, and with the total number of law firms in SME registry (~34000) the share of missing entries is below 15%.
-Consequently, the completeness of information is relatively high, and it can be used to draw reliable conclusions.
+The downloadable version of the registry includes, in particular, firm names, addresses of registration (up to a settlement), and codes of activity kinds compatible with NACE Rev.
+2 (European Classification of Economic Activities).
+It should be noted that the registry does not include several types of firms such as foreign commercial law firms without a juridical person registered under Russian law, non-profit organisations, bar (advocacy) formations, large enterprises.
+As a result, some data are missing.
+However, for a number of reasons, this problem should be considered insignificant.
+Foreign law firms usually register their affiliated juridical person in Russia and are therefore not omitted from the SME registry.
+The share of non-profit organisations in the Russian legal services market does not exceed 2% [@Moiseeva2016].
+The names of bar (advocacy) formations are usually quite uniform.
+They contain the kind of the formation (advocate's office, advocate's bureau, college of advocates, juridical consultancy) accompanied by the personal name(s) of advocates or a reference to a geographical location.
+Their naming patterns are thus quite apparent and do not require further investigation.
+Large enterprises represent only a tiny minority of law firms.
+
+For a validation, a list of law firms' unique tax IDs obtained from the SME registry for the year 2023 is matched with the list of law firms' tax IDs from the Russian Financial Statements Database (RFSD) [@Bondarkov2025] for the same year.
+The comparison shows that ~3,300 of the firms present in the SME registry are missing from the RFSD, and ~5,200 firms present in the RFSD are missing from the SME registry.
+The balance of missing records is not so far from being equal.
+The SME registry contains ~34000 law firms for the year 2023, so the missing rate is below 15%.
+Consequently, the completeness of the data is relatively high and reliable conclusions can be drawn.
 
@@
 
-In this study, the term “law firm” is defined as a commercial organization registered in Russia in accordance with the Civil Code of Russia, with a NACE Rev.2-compatible main activity code equal to 69.10 (activity in the area of law).
-The aforementioned definition excludes individual entrepreneurs, foreign companies, non-profit organizations, advocacy firms, and notaries.
+In this study, the term “law firm” is defined as a commercial organization registered in Russia in accordance with the Civil Code of Russia, with a NACE Rev.
+2-compatible main activity code equal to 69.10 (activity in the area of law).
+The aforementioned definition excludes individual entrepreneurs, foreign companies, non-profit organizations, bar (advocacy) formations, and notaries.
 The selection technique based on the main activity code is derived from that proposed by @Moiseeva2016.
@@
 
-The data used in the study covers a period of six years, from 2016 to 2021.
+The data used in the study cover a period of six years, from 2016 to 2021.
 The initial point in time is defined by the commencement of SME registry data availability.
-The finish point, 2021, is the latest year, is the latest year with “normal” conditions in Russia not impacted by military operation in Ukraine and consequent international sanctions and domestic economic and political perturbations.
-The purpose of temporal dimension is to exclude short-living law firms that are likely to be dummy companies.
-Organizations with a lifespan of less than three years have been omitted from the research, while all the remaining have been included in the further analysis.
-In addition, the utilization of multi-year data acts as a form of averaging, smoothing the potential distortions of spatial structure resulting from the random year-to-year changes of active law firms list.
+The end point, 2021, is the last year of “normal” conditions in Russia unaffected by the military operation in Ukraine and the resulting international sanctions and domestic economic and political turmoil.
+The purpose of the time dimension is to rule out short-lived law firms, which are likely to be shell companies.
+Organisations with a lifespan of less than three years are excluded from the research, while all the remaining are included in the further analysis.
+In addition, the use of multi-year data acts as a form of averaging, smoothing out the potential biases in the spatial structure caused by random annual changes in the population of active law firms.
 
@@
 The dimension of a firm name vector is 256.
-Contextual embeddings (YandexGPT) were used in favor of static ones (e.g.
-FastText), because during the preliminary analysis, they have resulted in far better clusterisation of law firm names (see the description of this step below) which is crucial for naming strategies identification.
+Contextual embeddings (YandexGPT) were used in preference to static ones (e.g.
+FastText), because in the preliminary analysis, they have resulted in a much better clustering of law firm names, which is crucial for naming patterns identification.
 
-### 2.2.4. Naming strategies identification
+### 2.2.4. Naming patterns identification and classification
 
-The dimension of law firm names vectors is reduced from 256 to 2 with the UMAP method, and the resulting 2D vectors are clustered using the DBSCAN technique and plotted to illustrate the results.
-The clusters are subsequently subjected to manual analysis.
+Law firm names are clustered using the KMeans technique.
+A matrix of features for clustering is constructed from the name embeddings.
+As there are 256 elements in each vector, the feature matrix has 256 columns.
+A target number of clusters is determined with the elbow method and ultimately set to 50, with the understanding that further manual processing of the clustering results will be necessary.
+Each name is assigned a cluster label between 1 and 50.
+Despite the simplicity of the clustering algorithm, which can even be considered old-fashioned compared to (H)DBSCAN, it performs quite well on the data at hand.
+Preliminary tests have shown that the clusters identified by (H)DBSCAN are not significantly better than those obtained by KMeans, but the latter takes much less times and memory.
+It is important to emphasise that automatic clustering is only a auxiliary step in the identification and classification of naming strategies which aims to facilitate the manual data processing by breaking down the entire set of law firm names into the compact groups that can be analysed more easily by a human.
+Therefore, there is no reason to improve the quality of clustering too much.
+
+The resulting clusters are then manually grouped.
 A random sample of 20 names is selected from each cluster.
-The names samples are carefully examined, and each cluster is attached a concise description (e.g.
-“Consulting companies”), a category, and a strategy.
-Clusters with descriptions, categories and strategies form a tree-like hierarchy where clusters make up categories, and categories belong to strategies.
+With 50 clusters and 20 sampled names allocated in each, the total size of the dataset for manual processing is 1, 000.
+The cluster samples are carefully examined, and a concise description is provided for each, e.g.
+“Names with the word “консалт/консалтинг” (consult/consulting)”.
+After this, the clusters are examined again and assigned to a naming pattern, e.g.
+“Consulting” or “Law firm”.
+Finally, the clusters and their naming patterns are divided into four previously mentioned naming strategies derived from @BassolsiGardella2023.
+Each cluster belongs to a single pattern, and each pattern belongs to a single strategy.
+Each strategy consists of several patterns, and each pattern consists of one or many clusters, forming a hierarchical structure.
 
-The concept of naming strategies is derived from the works of @Arora2015 and @BassolsiGardella2023.
-The latter paper proposes a strategic naming model with four broad naming strategies, namely “Uniqueness”, “Belonging”, “Market”, and “No-strategy”.
-Thee strategies form a 2×2 matrix, where two pairs of “Uniqueness vs Belonging” and “Market vs No-strategy” are axes, and strategies in the pairs are opposite to each other.
-Though invented for classification of names of companies in tourism, this naming framework fits well for the purpose of law firm names grouping.
-The specific descriptions and their mapping to the categories and strategies are provided, explained and analysed in the results section below.
+Each group except for the “Misc” category represents a specific naming strategy.
+The “Form” strategy encompasses names that are based on a word or words denoting a kind of business entity, including “company”, “firm”, “group”, “partnership”, “center”, “alliance”, and “corporation”.
+The “Law” strategy emphasizes the firm's affiliation with the legal services market, employing terms such as “legal”, “juridical”, “lex”, “law”, and “lawyer”.
+Names belonging to the “Service” strategy may indicate a particular service provided by a law firm (“advisor”, “consulting”, “protection”, “support”, “warrant”), describe the sectors of businesses the firm is primarily engaged with (“technology”, “development”, “house management”, “accounting”), or mention services in general.
+The “Misc” category is a catch-all group for names that do not clearly align with any other category.
 
@@
 
-There are `r initial_firms_count` law firms in the SMB registry in the selected period with `r nrow(unique_names)` unique firm names.
+There are `r initial_firms_count` law firms in the SME registry in the selected period with `r nrow(unique_names)` unique names.
 From the initial dataset, `r valid_firms_count` law firms have existed for at least three years and have been selected for the analysis.
-Their names sometimes are duplicated, and the number of unique law firm names being analysed is `r`.
+Their names are sometimes duplicated, and the number of unique law firm names being analysed is `r length(unique(clustered$name))`.
 

### 2025-01-28 law-firms-names/Paper.Rmd
Edit methods, results, discussion, conclusion, figures, abstract, and style, before submitting the second revision, #4

@@
 bibliography: "journal-article/refs.bib"
-csl: "journal-article/apa.csl"
+csl: "journal-article/apa_numeric.csl"
 link-bibliography: False  
@@
   echo = FALSE, 
-  fig.width = 4.96, 
+  fig.width = 6.7, 
   fig.id = FALSE,
@@
 ```{r data-imports, cache=TRUE, include=FALSE}
-source(here("journal-article", "Script.R"))
+source(here("journal-article/Script.R"))
 ```
@@
 
-**Abstract:** The present study analyzes law firm naming in Russia, with a particular focus on the geographical aspects of this phenomenon.
-The primary objective of this research is to ascertain the impact of the spatial relations between the country's regions, specifically those of a neighboring, proximity, and belonging to a parent-level subdivision, on the naming of law firms.
-To conduct the analysis, data regarding law firms is extracted from the open dumps of the state registry of small and medium-sized businesses.
-The names of the firms are transformed into numerical vectors, subsequently clustered with the KMeans method, and then manually grouped to identify naming strategies.
-Numerical representations of law firm names within each region are averaged to get region embeddings, which are then used to compare the names of law firms in different regions.
-The research findings indicate that regions situated closer to one another or sharing a common border tend to adopt slightly more similar law firm names.
-Furthermore, regions that are part of the same economic region also display a greater degree of similarity in law firm naming.
-Three primary strategies of law firm naming are “Form”, “Law“, and “Service”.
-Volga and North Caucasus economic regions are characterized by the predominance of the “Service” strategy, whereas the Volga-Vyatka economic region exhibits a “Law” strategy, and the remaining territories demonstrate a combination of naming strategies.
-The results of the study lend support to the widely used classification of economic regions in Russia.
-The findings also underscore the impact of neighboring, proximity, and interconnectivity on the economic and cultural attributes of the regions.
-The administrative open data processing techniques, as well as the methods of working with firm names based on neural networks proposed in this paper, may be used by other economic geographers for their own research purposes.
+**Abstract:** The present study analyzes the naming of law firms in Russia, with a particular focus on the geographical aspects of this phenomenon.
+The main objectives of this research are to identify law firm naming patterns, to categorise them and to test whether there are any spatial regularities in the distribution of law firms according to the identified patterns.
+To conduct the analysis, data regarding law firms is extracted from the open dumps of the state registry of small and medium-sized enterprises.
+The names of the firms are transformed into numerical vectors, subsequently clustered with the KMeans method, and then manually grouped to identify naming patterns and strategies.
+Global and local Moran's *I* statistics are calculated to test for spatial autocorrelation in the proportion of a given pattern in the regions.
+As a result, 22 naming patterns are identified, grouped into four naming strategies.
+Apart from no-strategy (40% of law firms), the most popular strategy is belonging (to the legal services market — 30%), followed by uniqueness (of service or location — 20%) and market (words and phrases with positive or promotional meaning — 10%).
+The research findings indicate that the spatial distribution of law firm naming patterns is generally not significantly different from random, with two exceptions: geographical branding in Siberia and an unusual heighborhood of regions with high and low proportions of “Expert” and “Personal brand” patterns in the European South.
 
@@
 For instance, the names “Law firm “Femida” and “Law company “Alibi” follow one pattern, while “Centre of legal aid and support” is an example of another.
-Patterns can be grouped several times to form a hierarchical classification [@Anora2015].
+Patterns can be grouped several times to form a hierarchical classification [@Arora2015].
 Such a classification can help to identify naming strategies.
@@
 highlighting a company's affiliation to a certain activity (retail, banking, transport, education, etc).
-@BassolsiGardella2023 propose a two-dimensional Strategic Naming Model (SNM) that includes two pairs of opposing strategies: belonging and uniqueness, no-strategy and market.
+Bassols i Gardella et al.
+[@BassolsiGardella2023] propose a two-dimensional Strategic Naming Model (SNM) that includes two pairs of opposing strategies: belonging and uniqueness, no-strategy and market.
 They discover a difference between three Colombian cities in the naming strategies of their tourism companies.
@@
 ## 2.1. Background
+
+Russia belongs to the Germanistic branch of a civil law legal system.
+Legal services (advice, negotiation, representation in court, etc) for organisations and individuals are usually provided by private law companies and individual lawyers.
+The market is divided between ordinary lawyers and advocates (members of the Bar), but there is no bar monopoly.
+There are ~90, 000 law firms in Russia, including ~40, 000 commercial law companies, ~30, 000 sole practitioners, ~20, 000 bar (advocate) offices, and ~10, 000 bar (advocate) organisations.
+There are no geographical restrictions on legal practitioners, so that a law firm registered in one region of the country can provide services in any other region.
+There are full-service and specialised law firms, as well as Russian law firms (the majority) and foreign law firms (the influential minority).
 
@@
 
-## 2.2. Methods
+## 2.3. Methods
 
-### 2.2.1. Law firms selection criteria
+### 2.3.1. Law firms selection criteria
 
@@
 
-### 2.2.2. Firm data acquisition
+### 2.3.2. Firm data acquisition
 
@@
 
-### 2.2.3. Firm name embeddings retrieval
+### 2.3.3. Firm name embeddings retrieval
 
@@
 
-### 2.2.4. Naming patterns identification and classification
+### 2.3.4. Naming patterns identification and classification
 
@@
 “Consulting” or “Law firm”.
-Finally, the clusters and their naming patterns are divided into four previously mentioned naming strategies derived from @BassolsiGardella2023.
+Finally, the clusters and their naming patterns are divided into four previously mentioned naming strategies derived from Bassols i Gardella et al.
+[@BassolsiGardella2023].
 Each cluster belongs to a single pattern, and each pattern belongs to a single strategy.
@@
 
-Each group except for the “Misc” category represents a specific naming strategy.
-The “Form” strategy encompasses names that are based on a word or words denoting a kind of business entity, including “company”, “firm”, “group”, “partnership”, “center”, “alliance”, and “corporation”.
-The “Law” strategy emphasizes the firm's affiliation with the legal services market, employing terms such as “legal”, “juridical”, “lex”, “law”, and “lawyer”.
-Names belonging to the “Service” strategy may indicate a particular service provided by a law firm (“advisor”, “consulting”, “protection”, “support”, “warrant”), describe the sectors of businesses the firm is primarily engaged with (“technology”, “development”, “house management”, “accounting”), or mention services in general.
-The “Misc” category is a catch-all group for names that do not clearly align with any other category.
+### 2.3.5. Testing for spatial regularities
 
-### 2.2.5. Spatial regularities identification
+Spatial autocorrelation analysis is used to check whether the spatial distribution of names following each pattern identified in the previous step is random or has some regularities.
+Regions (constituent entities) of the Russian Federation are chosen as the unit of observation.
+Municipal territories, the next level of spatial division, are not used, because an exploratory data analysis has shown that the lion's share of such the spatial units do not have any law firms.
 
-Analysis of autocorrelation is used to check whether the spatial distribution of names belonging to each cluster identified on the previous step is random or has some regularities.
-Regions (constituent entries) or the Russian Federation were selected as a spatial subdivision for calculating the autocorrelation measures.
-It could have been more reasonable to use the next spatial level, municipal territories, to perform the computations, but exploratory data analysis has shown that a lion's share of such territories has no law firms, because the legal business is highly concentrated in cities and large settlements.
+Six spatial weight matrices are utilized to ensure the robustness of the results.
+Three of these matrices are constructed using the K-nearest neighbors technique, where K is 3, 4 or 6.
+Another matrix relies on the existence of common boundaries or points to detect neighboring regions.
+The Sakhalin and Kaliningrad regions, which are enclaves, are manually linked to a few nearest regions.
+Two remaining matrices treat regions belonging to the same upper-level macro-region as neighbors, and all other regions as non-neighbors.
+One of these matrices uses economic regions act as macro-regions, and the other uses federal districts.
+Economic regions are defined according to the classification that has been widely used by economic geographers and spatial development practitioners in Russia for decades, as outlined, for example, in [@Leyzerovich2010].
+The composition of federal districts is defined by the Decree of the President of the Russian Federation (No.
+849 of May 13, 2000) “On the Plenipotentiary Representative of the President of the Russian Federation in the Federal District”, with respective amendments.
 
-Six spatial weights matrices are used to ensure the robustness of the results.
-Three of these matrices are constructed with K nearest neighbours technique based on the location of region's centroids, where K equals 3, 4 or 6, respectively.
-Another matrix also uses geographical properties of regions, but instead of centroids, it relies on the existence of common border lines or points to identify neighbouring regions.
-Sakhalin and Kaliningrad regions, being enclaves and thus having zero immediate neighbors, were manually connected with a few closest regions.
-Two remaining matrices treat regions belonging to the same upper-level metaregion as neighbours, and all the other regions as non-neighbours.
-For one of these two matrices, economic regions act as metaregions, and for another, federal districts are used in this fashion.
-Economic regions are defined according to the classification widely used by economic geographers and spatial development practitioners in Russia for decades and outlined, for example, in [@Leyzerovich2010].
-Federal districts composition is defined in the Executive Order of the President of Russia No.
-849, Dated May 13, 2000, “On the Plenipotentiary Representative of the President of the Russian Federation in the Federal District” (with up-to-date amendments).
+A global Moran's *I* statistic [@Moran1948] is calculated for each combination of naming pattern and spatial weight matrix.
+The variable of interest is the proportion of law firm names that follow a pattern in a total set of law firm names in a region.
+If there are no law firm names of the pattern in the region, the proportion is considered to be zero.
+For those naming patterns where the global Moran's *I* is statistically significant, and this significance is robust, a local Moran's *I~i~* is calculated for each region.
+In addition, each region is categorised and labelled according to its value and the values of its neighbors.
+Four typical categories are used: “High-High” (high value in a region and its neighbors), “Low-Low” (an opposite), “Low-High”, “High-Low”.
+*P*-values are adjusted for multiple comparisons using Holm's correction [@Holm1979].
+The significance level is set at 0.05.
 
-For each cluster, a global Moran's I metric is calculated.
-The variable of interest is a share of law firm names belonging to the cluster in a total quantity of law firm names in a region.
-If there are no law firm names of the cluster in the region, the share is considered to equal zero.
-P-values are adjusted for multiple comparisons with Holm's correction.
-Significance level (alpha) is set to 0.05.
-
-For the clusters with statistically significant global Moran's I, a local Moran's I was subsequently calculated.
-P-values are adjusted with local Holm's correction.
-Cluster classification of local Moran's I are mapped.
-
-### 2.2.6. Data statistics
+### 2.3.6. Data statistics
 
@@
 
-## 3.1. Law firms naming strategies
+## 3.1. Law firms naming patterns and strategies
 
-Law firm names clusters resulted from UMAP → DBSCAN pipeline are illustrated using scatterplot in Fig.
-1.
-The visual quality of clusterisation is high: all the clusters are compact, round, and separated by considerable gaps.
-There are no noise points, and all the names have been easily assigned to clusters.
+After clustering and manual grouping, 22 naming patterns are identified and assigned to one of four naming strategies (see Fig.
+1 for a dendrogram and Table 1 for a complete list of clusters, their descriptions, patterns, and strategies).
+The most popular is the no-strategy followed by ~18, 700 law firms.
+Slightly less important is the belonging strategy with ~15, 000 companies.
+The uniqueness strategy (~8, 300 law firms) is relatively rare, and the least used is the market strategy (~5, 300 companies).
 
-```{r fig-1-clusters, fig.cap="**Fig 1.** Law firm names clusters resulting from UMAP → DBSCAN processing of law firm names embeddings. Numbers are assigned automatically by DBSCAN", fig.height=4}
-clusters_plot
-```
+Belonging strategy includes law firms that directly state that they are law firms.
+“Law firm” is the most common pattern of this strategy.
+Most of the others (“Law”, “Lawyer”, “Legal service”) are semantically similar to it.
+It may seem questionable to include the “Personal brand” pattern (names such as “Mr.
+Brilliant and partners”) in this category, since personal branding also serves as a differentiation tool and thus implies uniqueness with the corresponding strategy.
+However, in Russia this way of naming is associated with lawyers, and such names often include the mention of law (“Law firm of Mr.
+Brilliant” instead of just “Mr.
+Brilliant”).
+In addition, Bassols i Gardella et al.
+[@BassolsiGardella2023] also categorise the personal branding of tourism firms as the belonging strategy.
 
-```{r table-1-clusters}
-kable(
-  clusters_info_table,
-  col.names = c(
-    "# of cluster", "# of names", "Description",
-    "Category", "Strategy"
-  )
-)
-```
+Market strategy, according to the original definition, is used by firms “having a name directly appealing to markets” [@BassolsiGardella2023].
+In the case of law firms, this can be achieved by mentioning target clients (“Business” pattern) or by using words with positive meanings that attract customers (“Positive”, “Plus”, “Legal aid”, “Guarantor” patterns).
+The “Expert” pattern is complex, as this word can be used to indicate either a high level of service (“expert” as a synonym for “professional”) or forensic services (“expert” as a forensic scientist).
+It is included here into market strategy because the semantic similarity of “experts” and “forensic scientists” in the Russian language is probably used as a marketing tool by forensic companies.
 
-Table 1 shows cluster sizes and sample names from the clusters with manually assigned short cluster descriptions, categories and strategies.
-The quantity of clusters (52) is relatively big, and their size is not even: while the largest group of names consists of items, the smallest one includes just 12 objects.
-Almost 1/3 of clusters (17) is comprised of a mixture of short (usually a single-word) names that cannot be given a reasonable description.
-They are marked as “Misc” and classified as “No-strategy”.
+Uniqueness strategy encompasses law firms whose names mention a particular type of service (“Property&Development”, “Consulting”, “Audit&Finance”) or a specific geographical region (“Siberia”).
+In contrast to belonging, these firms usually do not indicate in their names that they are part of legal services market.
+The opposition between belonging and uniqueness strategies may reflect a division between general and specialised law firms.
+Patterns of this strategy have the potential to be used to identify the specialisation of law firms, although this requires an additional research.
 
-All the other names are split between three primary strategies.
-“Market” strategy is comprised of firms with names having generally positive connotations in Russian language that help to pursue a potential client in a high quality of legal aid and thus act as a some sort of advertisement (“Guarantor”, “Expert”, “Partnership”, “Helper”, "Center” categories and some others).
-“Belonging” strategy, as its name suggests, make a strong emphasis on their belonging to the legal services market by using the words “law”, “lawyer”, “legal” or various Russian equivalents of “law firm” phrase.
-A distinct category of law firms with this strategy is formed by organisations that use a personal branding that follow the broad pattern “(law firm) <last name of a person> with partners” where (law firm) is an optional part, and <last name of a person> is a pointer to a featured lawyer of this organisation.
-Even if “law firm” subpart of the firm name is optional, the entire pattern is strongly associated in Russia with lawyers, thus this type of firm naming marks the belonging to the legal services sphere.
-Finally, “Uniqueness” strategy encompasses companies with names mentioning a particular kind of legal services (e.g.
-“Consulting”), or a target customers group (e.g.
-“Business”), or a related area of specialization (real estate, finance, audit, development), or a specific geographic location.
-
-```{r table-2-strategies-summary}
-kable(
-  strategy_counts,
-  digits = c(NA, 0, 0),
-  col.names = c("Strategy", "# of names", "%")
-)
-```
-
-Among the strategies, “No-strategy” (i.e.
-the absence of a clear strategy) is the most popular (see Table 2).
-It encompasses more than 1/3 of all the studied names.
-Belonging strategy is the second by the popularity holding 30% of law firms.
-Uniqueness strategy is employed by approximately 20% of companies, and the least typical one, market strategy, is used by slightly more than 10% of law firms.
+No-strategy is a catch-all group for names that do not fit into any of the other three.
+Its main component is miscellaneous one-word names that are difficult to categorise or describe.
+This strategy also includes names containing the words “Group”, “Partnership”, “Center” or “Service” (usually without any meaningful clarification of the activity or type of service).
+The “Managing companies” pattern is, after some consideration, classified as no-strategy.
+In Russia, the term “managing companies” may define either property or asset management firms or house management companies, and although some of their services may be treated as paralegal, it is difficult to place them in either the belonging, uniqueness, or market strategy The same logic applies to the “EdSciTech” pattern with probably education or LegalTech companies.
 
@@
 
-As Tables 3 and 4 demonstrate, there are few law firm names clusters with some regularities in the spatial distribution of law firm names.
-These two tables display results only for the clusters where global Moran's I coefficient is statistically significant.
-The values of Moran's I metric for all the other clusters are close to zero, and their adjusted p-value is above the selected threshold of 0.05, so the results for these clusters are not shown for brevity.
+There are few patterns (3 out of 22) with statistically significant regularities in their spatial distribution (Fig.
+2).
+These results are stable regardless of the technique used to construct the spatial weight matrix.
+The “Siberia” pattern is characterised by the highest global Moran's *I*, while the values for the “Expert” and “Personal brand” patterns are much lower, but still statistically significant.
 
-```{r table-3-moran-i-by-neighborhood}
-moran_test_by_regions %>% 
-  filter(p_adj < 0.05) %>% 
-  kable(
-    digits = c(0, NA, 2, 3),
-    col.names = c("# of cluster", "Description", "Moran's I", "Adj. p-value")
-  )
-```
-
-```{r table-4-moran-i-by-economic-regions}
-moran_test_by_er %>% 
-  filter(p_adj < 0.05) %>% 
-  kable(
-    digits = c(0, NA, 2, 3),
-    col.names = c("# of cluster", "Description", "Moran's I", "Adj. p-value")
-  )
-```
-
-According to the autocorrelation analysis results in both branches, there are in total four law firm names clusters where the spatial distribution is different from the random.
-One of these clusters, “Center or company with geographical location”, is present in both tables and is characterized by the highest value of Moran's I, that is about 0.66.
-Thus, it is likely that law firms with names mentioning a geographical location of the country are not distributed randomly within the country, but situated in a number of neighbouring regions.
-Two maps on the Fig.
-2, despite several differences, demonstrate that it is the regions in Eastern and Western Siberia that are characterized by the highest concentration of “geographically referenced” law firm names.
-A brief manual investigation uncovers that such names usually reference a particular region (e.g.
-Altai), or a specific city (e.g.
-Krasnoyarsk), or entire geographical area as a whole (Siberia).
-This argues for the presence of some sort of local or regional identity which is of highest expression in Siberia.
-It is possible that similar geo-based naming is present in other parts of the country, but the automatic clusterisation of law firm names using their embeddings cannot handle them properly — either because the quantity of geolocated names in other parts of the country is substantially smaller or because the text vectors do not reflect the similarity between the Siberian geographical mentions and geographical mentions in the other parts of Russia.
-Overall, an issue of geolocated law firm names is intriguing and requires further investigation.
-
-As for other clusters with statistically significant global Moran's I (“Forensic” names, “Experts and advisors”, and personal-branding pattern), the hotspots are situated in the European South of Russia.
-In the case of “Forensic” names, it is likely to be attributed to a tiny set of national republics of North Caucasus that are probably outliers, because their overall number of law firms and the distribution of naming strategies is substantially different from the one in the majority of regions.
-“Experts and advisors” cluster demonstrates hotspots in Rostov, Volgograd and Astrakhan regions.
-They are characterized by a relatively high number of law firms and thus cannot be left to as just outliers.
-However, the exact reasons for the spatial concentration of this naming pattern in these regions remains unclear.
-Finally, a personal branding cluster represents a drastic split of the European South into two groups of regions: the first one comprising of several national republics with high level of personal branding names and the second one consisting of some other national republics and all the regular regions with small share of such names.
-This may be seen as a slight argument for a position that European South of Russia is a conglomerate of two quite distinct groups of regions.
+A hotspot of the “Siberia” pattern is unsurprisingly located in Siberia (Fig.
+3a).
+It includes four core regions (Novosibirsk, Tomsk, Kemerovo regions, Altai Territory) and a periphery of a few other regions (e.g.
+Omsk region, Republic of Khakassia, Republic of Altai).
+The hotspots of two other patterns (Fig.
+3b, Fig.
+3c) are located in the European South of Russia.
+This part of Russia is characterised by a mixture of regions where the popularity of a particular naming pattern is either high in a similar neighborhood or low in the neighborhood of regions with high values.
+This observation can be partly attributed to the presence of outliers among the national republics of the North Caucasus where the characteristics and composition of the legal services market differ significantly from their nearest neighbors.
+Some of these nearest neighbors (e.g.
+Krasnodar Territory) are, on the contrary, the regions with a relatively developed legal services market.
+In summary, the analysis and mapping of its results revealed two geographical parts of Russia where some naming patterns differ from the rest of the country: Siberia and the European South.
+However, the majority of naming patterns show no spatial regularities, i.e.
+their spatial distribution is not statistically significantly different from random.
 
@@
 The findings of the study are somewhat contradictory.
-On the one hand, several hypotheses have been proven.
-These include the following: law firm names in neighboring regions are more similar than law firm names in non-neighboring regions; the smaller the geographic distance between regions, the more similar the law firm names; and law firm names within an economic region are typically more similar compared to the law firm names outside an economic region.
-On the other hand, the difference in the metrics of similarity is relatively small in each case, as is the correlation coefficient.
-It is therefore important to note that although the postulated phenomena do indeed exist, their magnitude is weak.
-
-The three principal strategies for naming law firms are as follows: “Form”, “Law”, and “Service”.
-The first of these represents a minority approach, while the remaining majority is divided between two other strategies.
-In the European South of Russia and in the Volga economic region, “Service” is a dominant strategy.
-However, no other spatial patterns in law firms naming have been identified.
-Economic regions are typically characterized by a mixture of naming strategies, except for Volga-Vyatka, where the “Law” strategy prevails, and Volga and North Caucasus, where the “Service” strategy is dominant.
+On the one hand, there are 22 law firm naming patterns in Russia that can be attributed to one of the four broad SMN strategies: belonging or uniqueness, market or no-strategy.
+Apart from the no-strategy (~40% of firms), belonging is the most popular strategy used by 30% of law firms, followed by uniqueness (~20%) and market (~10%).
+On the other hand, in gerenal there are no spatial regularities in the distribution of law firms that use a particular pattern: all the parts of the country are generally the same when it comes to law firms naming.
+Two notable exceptions to this rule are Siberia, with a fairly popular geographical branding (“Siberia” pattern), and the European South, with an unusual spatial composition of “Expert” and “Personal brand” patterns.
 
@@
 Additionally, it advocates for the utilization of disaggregated administrative open data for a spatial analysis.
-The findings demonstrate the effect of geographical proximity, neighboring, and economic relations on the similarity of the economic and, to a certain extent, cultural properties of the territories represented in firm names.
-The results generally align with the widely accepted classification of economic regions in Russia, although they highlight a few instances where a theoretically undesired internal heterogeneity within an economic region is observed.
+The findings demonstrate that even though the legal services market looks quite disconnected from the geography, the place can still play some role in its branding.
 Finally, the study provides an illustration of a semi-automated brand classification within a specific economic sector and discovers a geographical variation in naming strategies that could be further investigated.
@@
 <div id="refs"></div>
+
+```{r fig-1-naming-patterns-and-strategies, fig.cap="**Fig 1.** Dendrogram of law firm naming patterns and strategies in Russia", fig.height=8}
+names_classification_plot
+```
+
+```{r table-1-clusters}
+kable(
+  clusters_info_table,
+  col.names = c(
+    "# of cluster", "# of names", "Description",
+    "Pattern", "Strategy"
+  ),
+  caption = "Table 1. Characteristics of clusters of law firm names"
+)
+```
+
+```{r fig-2-global-moran-i, fig.cap="Fig. 2. Global Moran's *I* statistics and p-values for law firm naming patterns", fig.height=4}
+moran_test_global_res_plot
+```
+
+```{r fig-3-localmoran-maps, fig.cap="Fig. 3. Maps of hotspots in the spatial distribution of three law firms naming patterns", fig.height=5}
+localmoran_maps
+```

### 2025-01-28 law-firms-names/journal-article/Script.R

@@ -8,6 +8,7 @@ library(readr)
 library(readxl)
 library(sf)
 library(spdep)
+library(stopwords)
 library(tidyr)
 
 sf_use_s2(FALSE) # disable to avoid errors in distances and intersections
@@ -169,7 +170,7 @@ names_classification_plot <- ggraph(g, layout = "dendrogram") +
     angle_calc = "along", label_dodge = unit(3, "mm")
   ) +
   geom_node_label(aes(label = name)) +
-  scale_y_continuous(expand = expansion(mult = .1)) +
+  scale_y_continuous(expand = expansion(mult = .25)) +
   scale_edge_width_continuous(
     name = "# of names",
     breaks = c(500, 1000, 2500, 5000, 10000), 
@@ -389,7 +390,17 @@ localmoran_res <- rbind(
   right_join(regions %>% select(code) %>% st_drop_geometry()) %>% 
   complete(category, code) %>% 
   drop_na(category) %>% 
-  mutate(category = factor(category, labels = levels(clustered$category)[categories_for_localmoran])) %>% 
+  mutate(
+    category = factor(
+      category, 
+      labels = paste0(
+        "(", 
+        letters[1:3], 
+        ") ",
+        levels(clustered$category)[categories_for_localmoran]
+      )
+    )
+  ) %>% 
   right_join(regions) %>% 
   st_as_sf()
 

### 2025-01-30 law-firms-names/Paper.Rmd
Multiple changes for 3rd revision in SPIR, #4

@@
 ---
-title: "Identification and Spatial Analysis of Law Firms Naming Patterns in Russia Using Text Embeddings"
+title: "Identification and spatial analysis of law firms naming patterns in Russia using text embeddings"
 author: "Pavel Syomin"
@@
 If there are no law firm names of the pattern in the region, the proportion is considered to be zero.
-For those naming patterns where the global Moran's *I* is statistically significant, and this significance is robust, a local Moran's *I~i~* is calculated for each region.
+For those naming patterns where the global Moran's *I* is statistically significant, and this significance is robust, a local Moran's *I~i~* is calculated for each region and spatial weight matrix.
 In addition, each region is categorised and labelled according to its value and the values of its neighbors.
 Four typical categories are used: “High-High” (high value in a region and its neighbors), “Low-Low” (an opposite), “Low-High”, “High-Low”.
-*P*-values are adjusted for multiple comparisons using Holm's correction [@Holm1979].
+The confidence level of a category is determined by the number of spatial weight matrices for which the local Moran's *I~i~* value is found to be statistically significant.
+The regions and their respective categories are then mapped.
+The geoBoundaries [@Runfola2020] dataset with national and subnational boundaries is used for map creation.
+*P*-values for both global and local Moran's statistics are adjusted for multiple comparisons using Holm's correction [@Holm1979].
 The significance level is set at 0.05.
@@
 After clustering and manual grouping, 22 naming patterns are identified and assigned to one of four naming strategies (see Fig.
-1 for a dendrogram and Table 1 for a complete list of clusters, their descriptions, patterns, and strategies).
+1 for a dendrogram and Table 1 for a sample of 10 clusters, their descriptions, patterns, and strategies).
 The most popular is the no-strategy followed by ~18, 700 law firms.
@@
 The uniqueness strategy (~8, 300 law firms) is relatively rare, and the least used is the market strategy (~5, 300 companies).
+
+```{r fig-1-naming-patterns-and-strategies, fig.cap="**Fig 1.** Dendrogram of law firm naming patterns and strategies in Russia", fig.height=8}
+names_classification_plot
+```
 
@@
 [@BassolsiGardella2023] also categorise the personal branding of tourism firms as the belonging strategy.
+
+```{r table-1-clusters}
+clusters_info_table %>% 
+  slice_head(n = 10) %>% 
+  kable(
+    col.names = c(
+      "# of cluster", "# of names", "Description",
+      "Pattern", "Strategy"
+    ),
+    caption = "**Table 1.** Characteristics of clusters of law firm names"
+  )
+write_csv(clusters_info_table, "ESM2.csv")
+```
 
@@
 2).
-These results are stable regardless of the technique used to construct the spatial weight matrix.
-The “Siberia” pattern is characterised by the highest global Moran's *I*, while the values for the “Expert” and “Personal brand” patterns are much lower, but still statistically significant.
+These results are stable regardless of the technique used to construct the spatial weight matrix, with the sole exception of the “Expert” pattern where the spatial weight matrix constructed using economic regions has resulted in an insignificant Moran's *I* value.
+The “Siberia” pattern is characterised by the highest global Moran's *I*, while the values for the “Expert” and “Personal brand” patterns are much lower, but still statistically significant in the majority of the cases.
+The results for all the other naming patterns are statistically insignificant with Moran's *I* statistics close to zero (-0.1…0.1), so they are not visualised.
+
+```{r fig-2-global-moran-i, fig.cap="Fig. 2. Global Moran's *I* statistics and p-values for some law firm naming patterns", fig.height=5, fig.width=4}
+moran_test_global_res_plot
+```
+
+```{r fig-3-localmoran-maps, fig.cap="**Fig. 3.** Maps of hotspots in the spatial distribution of three law firms naming patterns: (a) Expert (b) Personal brand (c) Siberia", fig.height=10}
+localmoran_maps
+```
 
 A hotspot of the “Siberia” pattern is unsurprisingly located in Siberia (Fig.
-3a).
+3c).
 It includes four core regions (Novosibirsk, Tomsk, Kemerovo regions, Altai Territory) and a periphery of a few other regions (e.g.
@@
 The hotspots of two other patterns (Fig.
-3b, Fig.
-3c) are located in the European South of Russia.
+3a, Fig.
+3b) are located in the European South of Russia.
 This part of Russia is characterised by a mixture of regions where the popularity of a particular naming pattern is either high in a similar neighborhood or low in the neighborhood of regions with high values.
@@
 <div id="refs"></div>
-
-```{r fig-1-naming-patterns-and-strategies, fig.cap="**Fig 1.** Dendrogram of law firm naming patterns and strategies in Russia", fig.height=8}
-names_classification_plot
-```
-
-```{r table-1-clusters}
-kable(
-  clusters_info_table,
-  col.names = c(
-    "# of cluster", "# of names", "Description",
-    "Pattern", "Strategy"
-  ),
-  caption = "Table 1. Characteristics of clusters of law firm names"
-)
-```
-
-```{r fig-2-global-moran-i, fig.cap="Fig. 2. Global Moran's *I* statistics and p-values for law firm naming patterns", fig.height=4}
-moran_test_global_res_plot
-```
-
-```{r fig-3-localmoran-maps, fig.cap="Fig. 3. Maps of hotspots in the spatial distribution of three law firms naming patterns", fig.height=5}
-localmoran_maps
-```

### 2025-01-30 law-firms-names/journal-article/Script.R

@@ -305,7 +305,9 @@ moran_test_global_res <- rbind(
   select(category, moran_i, p_adj, weights_type, signif)
 
 moran_test_global_res_plot <- moran_test_global_res %>% 
-  complete(weights_type, category) %>% 
+  filter(signif) %>% 
+  mutate(category = as.character(category)) %>% 
+  complete(weights_type, category, fill = list(p_adj = 1)) %>% 
   mutate(
     p_adj_mark = cut(
       p_adj, 
@@ -318,16 +320,14 @@ moran_test_global_res_plot <- moran_test_global_res %>%
   geom_text(aes(label = round(moran_i, 1))) +
   scale_x_discrete(guide = guide_axis(angle = 45)) +
   scale_fill_brewer(
-    name = "Statistical significance",
-    na.translate = FALSE
+    name = "Statistical\nsignificance"
   ) +
   coord_fixed() +
   labs(
     x = "Law firm names category",
     y = "Type of weights matrix"
   ) +
-  theme_minimal() +
-  theme(legend.position = "bottom")
+  theme_minimal()
 moran_test_global_res_plot
 
 ## Local Moran's I
@@ -393,12 +393,7 @@ localmoran_res <- rbind(
   mutate(
     category = factor(
       category, 
-      labels = paste0(
-        "(", 
-        letters[1:3], 
-        ") ",
-        levels(clustered$category)[categories_for_localmoran]
-      )
+      labels = paste0("(", letters[1:3], ")")
     )
   ) %>% 
   right_join(regions) %>% 
@@ -409,18 +404,16 @@ localmoran_maps <- localmoran_res %>%
   geom_sf() +
   coord_sf(crs = ru_crs) +
   scale_alpha_continuous(
-    name = "Number of significant results"
+    name = "Confidence level"
   ) +
   scale_fill_discrete(
     name = "Region group",
+    labels = c("Low-High", "High-High", "Insignificant"),
     na.value = "gray90"
   ) +
-  facet_wrap(vars(category), ncol = 2) +
+  facet_wrap(vars(category), ncol = 1) +
   theme_void() +
   theme(
-    legend.direction = "horizontal",
-    legend.justification = c(0, 0),
-    legend.position = "inside",
-    legend.position.inside = c(.6, .1),
-    legend.title.position = "top"
-  )
+    strip.text = element_text(hjust = 0)
+  ) +
+  labs(caption = "Boundaries source: www.geoboundaries.org (CC BY 4.0)")

### 2025-02-03 fts-open-data-cli-app/Paper.Rmd
Reorganise files and adjust code for the conference paper about CLI app, #6

--- 2024-01-29 argo-conf-2024-paper/Paper.Rmd
@@
   word_document:
-    reference_docx: reference.docx
-bibliography: refs.bib
-csl: "gost-r-7-0-5-2008-numeric-alphabetical.csl"
+    reference_docx: assets/reference.docx
+bibliography: assets/refs.bib
+csl: assets/gost-r-7-0-5-2008-numeric-alphabetical.csl
 ---
@@
 ```{r imports, include = FALSE, cache = TRUE}
-source("Script.R")
+source("helpers/Script.R")
 ```

### 2025-02-03 fts-open-data-cli-app/helpers/Script.R

@@ -1,50 +1,50 @@
 library(dplyr)
 library(ggplot2)
+library(here)
 library(readr)
-library(scatterpie)
 library(sf)
 library(tidyr)
 
-data_path <- "../../ru-smb-companies/group_A/panel.csv"
+data_path <- here("../large-datasets/ncea-group-a-firms/panel.csv")
 data <- read_csv(data_path)
 
 total <- data %>% distinct(tin) %>% nrow()
 
 sample_row <- data %>%
-  filter(year == 2021, kind == 1) %>% 
+  filter(year == 2021, kind == 1) %>%
   slice(1000:1001)
 sample_table <- data.frame(
   var = colnames(sample_row),
   value = as.character(sample_row[1, ])
 ) %>% replace_na(list(value = "—"))
 
-regions_boundaries <- st_read("ru.geojson")
+regions_boundaries <- st_read(here("assets/ru.geojson"))
 ru_crs <- st_crs("+proj=aea +lat_0=0 +lon_0=100 +lat_1=68 +lat_2=44 +x_0=0 +y_0=0 +ellps=krass +towgs84=28,-130,-95,0,0,0,0 +units=m +no_defs")
 
-plot_1 <- data %>% 
+plot_1 <- data %>%
   filter(
     year == 2021,
-    kind == 1, 
-    substr(activity_code_main, 1, 2) == "01") %>% 
+    kind == 1,
+    substr(activity_code_main, 1, 2) == "01") %>%
   mutate(activity = case_when(
     substr(activity_code_main, 1, 4) %in% c("01.1", "01.2", "01.3") ~ "Растениеводство",
-    substr(activity_code_main, 1, 4) == "01.4" ~ "Животноводство")) %>% 
-  drop_na(activity) %>% 
-  count(lat, lon, activity) %>% 
-  drop_na(lat, lon) %>% 
-  group_by(lat, lon) %>% 
-  arrange(-n) %>% 
-  slice(1) %>% 
-  ungroup() %>% 
-  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>% 
+    substr(activity_code_main, 1, 4) == "01.4" ~ "Животноводство")) %>%
+  drop_na(activity) %>%
+  count(lat, lon, activity) %>%
+  drop_na(lat, lon) %>%
+  group_by(lat, lon) %>%
+  arrange(-n) %>%
+  slice(1) %>%
+  ungroup() %>%
+  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
   ggplot() +
[…]

### 2025-02-05 legal-companies-migration/journal-article/Paper.Rmd
Reorganise files and adjust code for two papers about legal companies migration, #6

--- 2024-06-04 legal-companies-migration/Paper.Rmd
@@
   word_document:
-    reference_docx: reference.docx
-bibliography: biblio-ru.bib
-csl: "biblio-styles-ru.csl"
+    reference_docx: assets/reference.docx
+bibliography: assets/biblio-ru.bib
+csl: assets/biblio-styles-ru.csl
 
@@
 ```{r imports, include=FALSE, cache=TRUE}
-source("Script.R")
+source("helpers/Script.R")
 ```

### 2025-02-05 bel-geo-congress-2024/Script.R

@@ -1,366 +1,0 @@
-library(dplyr)
-library(ggplot2)
-library(glue)
-library(readr)
-library(tidyr)
-library(sf)
-library(units)
-
-# Load the data
-data <- read_csv("../../ru-smb-companies/legal/smb.csv")
-revexp <- read_csv("../../ru-smb-companies/legal/revexp.csv")
-empl <- read_csv("../../ru-smb-companies/legal/empl.csv")
-cities <- read_csv("cities.csv")
-regions_geo <- st_read("ru.geojson")
-regions <- read_csv("regions.csv")
-
-# Number of unique companies
-companies_count <- n_distinct(data[data$kind == 1, "tin"])
-
-# Number of companies as of 2016-08-10 (to compare with Moiseeva2016)
-companies_count_08_2016 <- nrow(filter(data, kind == 1, start_date == "2016-08-10"))
-
-# Calculate the absolute numbers and shares of companies
-# with changes in location
-# Location is defined as change either in region or settlement
-migration_count_distribution <- data %>% 
-  filter(kind == 1) %>% 
-  group_by(tin) %>%
-  distinct(region, settlement) %>% 
-  count() %>% 
-  ungroup() %>% 
-  count(n, name = "count") %>% 
-  mutate(
-    n = n - 1,
-    share = 100 * count / sum(count))
-
-# Draw it on plot
-mig_count_distr_plot <- migration_count_distribution %>% 
-  ggplot(aes(x = n, y = count)) +
-  geom_col() +
-  geom_text(
-    aes(y = 1e6, label = scales::number(share, accuracy = 0.01, suffix = " %")),
-    color = "gray30", size = 3
-    ) + 
-  geom_text(aes(label = count), vjust = -.5) + 
-  scale_y_log10(limits = c(NA, 1e6), breaks = 10^(1:5), labels = ~ .x) +
-  labs(
-    x = "Number of migrations by a company", 
-    y = "Number of companies (log10)") +
-  theme_bw(base_size = 12, base_family = "Times New Roman") +
-  theme(
-    panel.grid.minor.x = element_blank(), 
-    panel.grid.major.x = element_blank(),
-    panel.grid.minor.y = element_blank())
-mig_count_distr_plot
-
-# Prepare the base dataset to analyze migrations of companies
-max_empl <- c(15, 100, 250)
-max_revenue <- c(120e6, 800e6, 2e9)
-migrations <- data %>% 
[…]

### 2025-02-05 law-firms-names/journal-article/Script.R

@@ -417,3 +417,89 @@ localmoran_maps <- localmoran_res %>%
     strip.text = element_text(hjust = 0)
   ) +
   labs(caption = "Boundaries source: www.geoboundaries.org (CC BY 4.0)")
+
+
+aas <- read_csv("aas.csv")
+fas <- read_csv("fas.csv")
+df <- cbind(
+  x1 = sort(aas$region),
+  x2 = sort(regions[regions$name != "Ненецкий автономный округ", ]$name)
+)
+
+w <- regions[regions$name != "Ненецкий автономный округ", ] %>% 
+  arrange(name) %>% 
+  poly2nb() %>% 
+  nb2listw(zero.policy = T)
+
+moran.test(aas %>% arrange(region) %>% pull(appeal_count), w)
+moran.plot(aas %>% arrange(region) %>% pull(appeal_count), w)
+localmoran(aas %>% arrange(region) %>% pull(appeal_count), w) %>% 
+  hotspot(
+    Prname="Pr(z != E(Ii))",
+    cutoff = 2, 
+    p.adjust = "holm",
+    droplevels=FALSE
+  ) %>% 
+  data.frame(
+    group = ., 
+    name = sort(aas$region),
+    geom = regions[regions$name != "Ненецкий автономный округ", ] %>% 
+      arrange(name) %>% 
+      pull(geometry)
+  ) %>% 
+  st_as_sf() %>% 
+  ggplot(aes(fill = group)) +
+  geom_sf() +
+  coord_sf(crs = ru_crs)
+
+d <- cbind(
+  aas %>% arrange(region),
+  geom = regions[regions$name != "Ненецкий автономный округ", ] %>% 
+    arrange(name) %>% 
+    pull(geometry)
+) %>% 
+  st_as_sf()
+
+d %>% 
+  ggplot(aes(fill = appeal_count)) +
+  geom_sf() +
+  coord_sf(crs = ru_crs) +
+  scale_fill_fermenter(
+    name = "",
+    palette = "RdYlBu",
+    breaks = quantile(
+      d$appeal_count,
+      c(.2, .4, .6, .8),
+      na.rm = TRUE
+    ),
+    labels = function(x) round(x, digits = 0),
+    show.limits = TRUE,
+    na.value = "white"
+  )
+
+moran.test(fas %>% arrange(region) %>% pull(fas_count), w)
+moran.plot(fas %>% arrange(region) %>% pull(fas_count), w)
[…]

### 2025-02-05 legal-companies-migration/Script.R

@@ -1,493 +1,0 @@
-library(dplyr)
-library(ggplot2)
-library(glue)
-library(readr)
-library(tidyr)
-library(sf)
-
-# Load the data
-data <- read_csv("../../ru-smb-companies/legal/smb.csv")
-revexp <- read_csv("../../ru-smb-companies/legal/revexp.csv")
-empl <- read_csv("../../ru-smb-companies/legal/empl.csv")
-cities <- read_csv("cities.csv")
-regions_geo <- st_read("ru.geojson")
-regions <- read_csv("regions.csv")
-tiles <- read_csv("russia-tiles.csv")
-
-# Number of unique companies
-companies_count <- n_distinct(data[data$kind == 1, "tin"])
-
-# Number of companies as of 2016-08-10 (to compare with Moiseeva2016)
-companies_count_08_2016 <- nrow(filter(data, kind == 1, start_date == "2016-08-10"))
-
-# Calculate the absolute numbers and shares of companies
-# with changes in location
-# Location is defined as change either in region or settlement
-migration_count_distribution <- data %>% 
-  filter(kind == 1) %>% 
-  group_by(tin) %>%
-  distinct(region, settlement) %>% 
-  count() %>% 
-  ungroup() %>% 
-  count(n, name = "count") %>% 
-  mutate(
-    n = n - 1,
-    share = 100 * count / sum(count))
-
-# Draw it on plot
-mig_count_distr_plot <- migration_count_distribution %>% 
-  ggplot(aes(x = n, y = count)) +
-  geom_col() +
-  geom_text(
-    aes(y = 1e6, label = scales::number(share, accuracy = 0.01, suffix = " %")),
-    color = "gray30", size = 3
-    ) + 
-  geom_text(aes(label = count), vjust = -.5) + 
-  scale_y_log10(limits = c(NA, 1e6), breaks = 10^(1:5), labels = ~ .x) +
-  labs(
-    x = "Number of migrations by a company", 
-    y = "Number of companies (log10)") +
-  theme_bw(base_size = 11, base_family = "Times New Roman") +
-  theme(
-    panel.grid.minor.x = element_blank(), 
-    panel.grid.major.x = element_blank(),
-    panel.grid.minor.y = element_blank())
-mig_count_distr_plot
-
-# Prepare the base dataset to analyze migrations of companies
-max_empl <- c(15, 100, 250)
-max_revenue <- c(120e6, 800e6, 2e9)
[…]

### 2025-02-05 legal-companies-migration/conference-paper/helpers/Script.R

@@ -1,0 +1,367 @@
+library(dplyr)
+library(ggplot2)
+library(glue)
+library(here)
+library(readr)
+library(tidyr)
+library(sf)
+library(units)
+
+# Load the data
+data <- read_csv(here("../large-datasets/law-firms/smb.csv"))
+revexp <- read_csv(here("../large-datasets/law-firms/revexp.csv"))
+empl <- read_csv(here("../large-datasets/law-firms/empl.csv"))
+cities <- read_csv(here("common/cities.csv"))
+regions_geo <- st_read(here("common/ru.geojson"))
+regions <- read_csv(here("common/regions.csv"))
+
+# Number of unique companies
+companies_count <- n_distinct(data[data$kind == 1, "tin"])
+
+# Number of companies as of 2016-08-10 (to compare with Moiseeva2016)
+companies_count_08_2016 <- nrow(filter(data, kind == 1, start_date == "2016-08-10"))
+
+# Calculate the absolute numbers and shares of companies
+# with changes in location
+# Location is defined as change either in region or settlement
+migration_count_distribution <- data %>% 
+  filter(kind == 1) %>% 
+  group_by(tin) %>%
+  distinct(region, settlement) %>% 
+  count() %>% 
+  ungroup() %>% 
+  count(n, name = "count") %>% 
+  mutate(
+    n = n - 1,
+    share = 100 * count / sum(count))
+
+# Draw it on plot
+mig_count_distr_plot <- migration_count_distribution %>% 
+  ggplot(aes(x = n, y = count)) +
+  geom_col() +
+  geom_text(
+    aes(y = 1e6, label = scales::number(share, accuracy = 0.01, suffix = " %")),
+    color = "gray30", size = 3
+    ) + 
+  geom_text(aes(label = count), vjust = -.5) + 
+  scale_y_log10(limits = c(NA, 1e6), breaks = 10^(1:5), labels = ~ .x) +
+  labs(
+    x = "Number of migrations by a company", 
+    y = "Number of companies (log10)") +
+  theme_bw(base_size = 12, base_family = "Times New Roman") +
+  theme(
+    panel.grid.minor.x = element_blank(), 
+    panel.grid.major.x = element_blank(),
+    panel.grid.minor.y = element_blank())
+mig_count_distr_plot
+
+# Prepare the base dataset to analyze migrations of companies
+max_empl <- c(15, 100, 250)
[…]

### 2025-02-05 legal-companies-migration/journal-article/helpers/Script.R

@@ -1,0 +1,493 @@
+library(dplyr)
+library(ggplot2)
+library(glue)
+library(here)
+library(readr)
+library(tidyr)
+library(sf)
+
+# Load the data
+data <- read_csv(here("../large-datasets/law-firms/smb.csv"))
+revexp <- read_csv(here("../large-datasets/law-firms/revexp.csv"))
+empl <- read_csv(here("../large-datasets/law-firms/empl.csv"))
+cities <- read_csv(here("common/cities.csv"))
+regions_geo <- st_read(here("common/ru.geojson"))
+regions <- read_csv(here("common/regions.csv"))
+tiles <- read_csv(here("journal-article/assets/russia-tiles.csv"))
+
+# Number of unique companies
+companies_count <- n_distinct(data[data$kind == 1, "tin"])
+
+# Number of companies as of 2016-08-10 (to compare with Moiseeva2016)
+companies_count_08_2016 <- nrow(filter(data, kind == 1, start_date == "2016-08-10"))
+
+# Calculate the absolute numbers and shares of companies
+# with changes in location
+# Location is defined as change either in region or settlement
+migration_count_distribution <- data %>% 
+  filter(kind == 1) %>% 
+  group_by(tin) %>%
+  distinct(region, settlement) %>% 
+  count() %>% 
+  ungroup() %>% 
+  count(n, name = "count") %>% 
+  mutate(
+    n = n - 1,
+    share = 100 * count / sum(count))
+
+# Draw it on plot
+mig_count_distr_plot <- migration_count_distribution %>% 
+  ggplot(aes(x = n, y = count)) +
+  geom_col() +
+  geom_text(
+    aes(y = 1e6, label = scales::number(share, accuracy = 0.01, suffix = " %")),
+    color = "gray30", size = 3
+    ) + 
+  geom_text(aes(label = count), vjust = -.5) + 
+  scale_y_log10(limits = c(NA, 1e6), breaks = 10^(1:5), labels = ~ .x) +
+  labs(
+    x = "Number of migrations by a company", 
+    y = "Number of companies (log10)") +
+  theme_bw(base_size = 11, base_family = "Times New Roman") +
+  theme(
+    panel.grid.minor.x = element_blank(), 
+    panel.grid.major.x = element_blank(),
+    panel.grid.minor.y = element_blank())
+mig_count_distr_plot
+
[…]

### 2025-02-05 fts-open-data-in-economic-geography/Paper.Rmd
Adjust code and remove useless files for paper about FTS open data, #6

@@
   word_document:
-    reference_doc: "assets/reference.docx"
-bibliography: "assets/biblio_ru.bib"
-csl: "assets/citations-ru.csl"
+    reference_doc: assets/reference.docx
+bibliography: assets/biblio_ru.bib
+csl: assets/citations-ru.csl
 link-bibliography: false

### 2025-02-05 fts-open-data-in-economic-geography/helpers/Plots.R

@@ -50,18 +50,17 @@ labels <- data.frame(
 labels <- labels[["ru"]]
 
 # Data
-rsmp_data_path <- here("../../ru-smb-companies/group_A/smb.csv")
+rsmp_data_path <- here("../large-datasets/ncea-group-a-firms/smb.csv")
 rsmp_data <- read_csv(rsmp_data_path)
 
 # Maps
-regions_boundaries <- st_read(here("assets", "ru.geojson"))
-municipal_boundaries <- st_read(here("assets", "ru-mun-gadm.geojson"))
+regions_boundaries <- st_read(here("assets/ru.geojson"))
+municipal_boundaries <- st_read(here("assets/ru-mun-gadm.geojson"))
 ru_crs <- st_crs("+proj=aea +lat_0=0 +lon_0=100 +lat_1=68 +lat_2=44 +x_0=0 +y_0=0 +ellps=krass +towgs84=28,-130,-95,0,0,0,0 +units=m +no_defs")
-regions <- read_csv(here("assets", "regions.csv"))
+regions <- read_csv(here("assets/regions.csv"))
 regions <- regions_boundaries %>% 
   left_join(regions, by = c("shapeISO" = "iso_code")) %>% 
   select(name, name_en = shapeName)
-ru_svr <- st_read(here("assets", "ru_svr.geojson"))
 
 # Regional distribution
 ac_code_mapping <- c("01" = labels[4], "02" = labels[5])

### 2025-02-05 fts-open-data-in-economic-geography/helpers/Stats.R

@@ -32,7 +32,7 @@ labels <- data.frame(
 labels <- labels[[LOCALE]]
 
 # Counts
-rsmp_data <- read_csv(here("../../ru-smb-companies/group_A/smb.csv"))
+rsmp_data <- read_csv(here("../large-datasets/ncea-group-a-firms/smb.csv"))
 counts <- rsmp_data %>% 
   mutate(kind = replace(kind, kind == 3, 2)) %>% 
   distinct(tin, .keep_all = TRUE) %>% 
@@ -40,12 +40,12 @@ counts <- rsmp_data %>%
   pull(n)
 
 # Activity codes
-ac <- read_csv(here("assets", "activity_codes_classifier.csv"))
+ac <- read_csv(here("assets/activity_codes_classifier.csv"))
 ac_groups <- filter(ac, is.na(code)) %>%
   mutate(name = str_to_sentence(name))
 
 # Validation
-val_stats <- read_csv(here("assets", "validation-stats.csv"))
+val_stats <- read_csv(here("assets/validation-stats.csv"))
 
 corr_by_year_all <- val_stats %>% 
   group_by(year) %>% 

### 2025-02-05 law-firms-geography/Paper.Rmd
Reorganise files and adjust code for paper about law firms geography, #6

@@
   word_document:
-    reference_doc: reference.docx
-bibliography: refs.bib
-csl: style.csl
+    reference_doc: assets/reference.docx
+bibliography: assets/refs.bib
+csl: assets/style.csl
 link-bibliography: False

### 2025-02-05 law-firms-geography/helpers/Script.R

@@ -1,14 +1,15 @@
 library(dplyr)
 library(ggplot2)
+library(here)
 library(readr)
 library(tidyr)
 
 # Load raw data
-courts <- read_csv("courts.csv")
-data <- read_csv("../../ru-smb-companies/legal/panel.csv")
-empl <- read_csv("empl.csv")
-pop <- read_csv("population.csv")
-tiles <- read_csv("russia-tiles.csv")
+courts <- read_csv(here("assets/courts.csv"))
+data <- read_csv(here("../large-datasets/law-firms/panel.csv"))
+empl <- read_csv(here("assets/empl.csv"))
+pop <- read_csv(here("assets/population.csv"))
+tiles <- read_csv(here("assets/russia-tiles.csv"))
 
 # Common preparations
 data <- data %>% 

### 2025-02-06 law-firms-mapping/Paper_ru.Rmd
Fix a typo
Adjust code and remove useless files for paper about law firms mapping, #6

@@
 bibliography: assets/biblio_ru.bib
-csl: "assets/gost-r-7-0-100-2018-numeric-alphabetical.csl"
+csl: assets/gost-r-7-0-100-2018-numeric-alphabetical.csl
 link-bibliography: false
@@
 1), однако данные на уровне городов позволяют заметить и другие тенденции, например, концентрацию бизнеса в региональных центрах.
-Кроме того, очевидна высокое сосредоточение в Москве и ее окрестностях.
+Кроме того, очевидно высокое сосредоточение в Москве и ее окрестностях.
 Аналогичная картина просматривается в Санкт-Петербурге.

### 2025-02-06 law-firms-mapping/helpers/Plots.R

@@ -1,24 +1,25 @@
 library(dplyr)
 library(forcats)
 library(ggplot2)
+library(here)
 library(readr)
 library(scales)
 library(sf)
 library(tidyr)
 
 # Data
-smb_data <- read_csv("assets/smb.csv")
-revexp_data <- read_csv("assets/revexp.csv")
-empl_data <- read_csv("assets/empl.csv")
+smb_data <- read_csv(here("../large-datasets/law-firms/smb.csv"))
+revexp_data <- read_csv(here("../large-datasets/law-firms/revexp.csv"))
+empl_data <- read_csv(here("../large-datasets/law-firms/empl.csv"))
 
 # Maps
-regions_boundaries <- st_read("assets/ru.geojson")
+regions_boundaries <- st_read(here("assets/ru.geojson"))
 ru_crs <- st_crs("+proj=aea +lat_0=0 +lon_0=100 +lat_1=68 +lat_2=44 +x_0=0 +y_0=0 +ellps=krass +towgs84=28,-130,-95,0,0,0,0 +units=m +no_defs")
-regions <- read_csv("assets/regions.csv")
+regions <- read_csv(here("assets/regions.csv"))
 regions <- regions_boundaries %>% 
   left_join(regions, by = c("shapeISO" = "iso_code")) %>% 
   select(name, name_en = shapeName)
-ru_svr <- st_read("assets/ru_svr.geojson")
+ru_svr <- st_read(here("assets/ru_svr.geojson"))
 
 # Regional distribution
 regions_map <- smb_data %>%

### 2025-02-06 legal-companies-regional-specialization/Paper.Rmd
Adjust code and reorganise files for paper about regional specialisation on legal services, #6

@@
   word_document:
-    reference: reference.docx
-bibliography: biblio.bib
-csl: biblio-style.csl
+    reference: assets/reference.docx
+bibliography: assets/biblio.bib
+csl: assets/biblio-style.csl
 ---
@@
 ```{r imports, include=FALSE, cache = TRUE}
-source("Script.R")
+source("helpers/Script.R")
 ```

### 2025-02-06 legal-companies-regional-specialization/helpers/Script.R

@@ -1,11 +1,12 @@
 library(dplyr)
 library(ggplot2)
+library(here)
 library(readr)
 library(tidyr)
 
-panel <- read_csv("../../ru-smb-companies/legal/panel.csv")
-empl <- read_csv("empl.csv")
-tiles <- read_csv("russia-tiles.csv")
+panel <- read_csv(here("../large-datasets/law-firms/panel.csv"))
+empl <- read_csv(here("assets/empl.csv"))
+tiles <- read_csv(here("assets/russia-tiles.csv"))
 
 spec <- panel %>% 
   filter(year == 2021) %>% 

### 2025-02-06 legal-companies-zipf-law/helpers/Script.R
Reorganise files and adjust code for paper about Zipf law in law firms distribution, #6

@@ -1,14 +1,15 @@
 library(dplyr)
 library(ggplot2)
+library(here)
 library(readr)
 library(sf)
 library(tidyr)
 
 # Data on firms, and cities lookup table
-panel <- read_csv("../../ru-smb-companies/legal/panel.csv")
+panel <- read_csv(here("../large-datasets/law-firms/panel.csv"))
 cities <- rbind(
-  read_csv("cities.csv"),
-  read_csv("cities_additional.csv")
+  read_csv(here("assets/cities.csv")),
+  read_csv(here("assets/cities_additional.csv"))
 )
 
 lcc <- panel %>% 
@@ -111,5 +112,3 @@ rank_empl_lm <- data %>%
   lm(log(rank) ~ log(empl), .)
 summary(rank_empl_lm)
 rank_empl_rsq <- summary(rank_empl_lm)$adj.r.squared
-
-

### 2025-02-06 legal-services-provision/conference-paper/Paper.Rmd
Rename folder, #2
Reorganise files and adjust code for paper about legal services provision, #6

--- 2024-05-11 legal-companies-concentration-and-availability/conference-paper/Paper.Rmd
@@
 ```{r imports, include=FALSE, cache = TRUE}
-library(here)
-source(here("common", "Script.R"))
+source("helpers/Script.R")
 ```

### 2025-02-06 legal-services-provision/journal-article/Paper.Rmd
Reorganise files and adjust code for paper about legal services provision, #6

--- 2024-06-07 legal-services-provision/journal-article/Article.Rmd
@@
   word_document:
-    reference: reference.docx
-bibliography: refs.bib
-csl: style.csl
+    reference: assets/reference.docx
+bibliography: assets/refs.bib
+csl: assets/style.csl
 ---
@@
 ```{r imports, include=FALSE, cache = TRUE}
-source(here("journal-article", "Script.R"))
+source("helpers/Script.R")
 ```

### 2025-02-06 legal-services-provision/conference-paper/helpers/Script.R

@@ -7,7 +7,7 @@ library(sf)
 library(tidyr)
 
 # Data on firms, and cities lookup table
-panel <- read_csv(here("../../ru-smb-companies/legal/panel.csv"))
+panel <- read_csv(here("../large-datasets/law-firms/panel.csv"))
 cities <- rbind(
   read_csv(here("common", "cities.csv")),
   read_csv(here("common", "cities_additional.csv"))

### 2025-02-06 legal-services-provision/journal-article/helpers/Script.R

@@ -9,13 +9,13 @@ library(tibble)
 library(tidyr)
 
 # Load data
-panel <- read_csv(here("../../ru-smb-companies/legal/panel.csv"))
+panel <- read_csv(here("../large-datasets/law-firms/panel.csv"))
 cities <- rbind(
-  read_csv(here("common", "cities.csv")),
-  read_csv(here("common", "cities_additional.csv"))
+  read_csv(here("common/cities.csv")),
+  read_csv(here("common/cities_additional.csv"))
 )
-economic_regions <- read_csv(here("journal-article", "economic-regions.csv"))
-regions_geo <- st_read(here("common", "ru.geojson")) %>% 
+economic_regions <- read_csv(here("journal-article/assets/economic-regions.csv"))
+regions_geo <- st_read(here("common/ru.geojson")) %>% 
   left_join(economic_regions, by = c("shapeISO" = "iso_code"))
 
 # Build table with number of employees by city

### 2025-02-07 law-firms-names/conference-paper/Conference-slides.qmd
Reorganise files and adjust code, edit title as a minor revision, #4, #6

@@
 # load the data
-data <- read_csv(here("common", "panel.csv"))
+data <- read_csv(here("../large-datasets/law-firms/panel.csv"))
 cluster_labels <- read_csv(
-  here("conference-paper", "cluster-labels-conf.csv"))
-vectors <- read_csv(here("common", "names-vectors.csv"))
-ru <- st_read(here("common", "ru.geojson"))
+  here("conference-paper/assets/cluster-labels-conf.csv"))
+vectors <- read_csv(here("common/names-vectors.csv"))
+ru <- st_read(here("common/regions.geojson"))
 

### 2025-02-07 law-firms-names/journal-article/Paper.Rmd
Reorganise files and adjust code, edit title as a minor revision, #4, #6

--- 2025-01-30 law-firms-names/Paper.Rmd
@@
 ---
-title: "Identification and spatial analysis of law firms naming patterns in Russia using text embeddings"
+title: "Spatial distribution of naming patterns in Russian law firms through text embeddings"
 author: "Pavel Syomin"
@@
   word_document:
-    reference_doc: "journal-article/reference.docx"
-bibliography: "journal-article/refs.bib"
-csl: "journal-article/apa_numeric.csl"
+    reference_doc: assets/reference.docx
+bibliography: assets/refs.bib
+csl: assets/apa_numeric.csl
 link-bibliography: False  
@@
 ```{r data-imports, cache=TRUE, include=FALSE}
-source(here("journal-article/Script.R"))
+source("helpers/Script.R")
 ```

### 2025-02-07 law-firms-names/journal-article/helpers/Script.R

@@ -14,9 +14,9 @@ library(tidyr)
 sf_use_s2(FALSE) # disable to avoid errors in distances and intersections
 
 # Load the data
-data <- read_csv(here("common/panel.csv"))
-er <- read_csv(here("journal-article/economic-regions.csv"))
-fd <- read_csv(here("journal-article/federal-districts.csv"))
+data <- read_csv(here("../large-datasets/law-firms/panel.csv"))
+er <- read_csv(here("journal-article/assets/economic-regions.csv"))
+fd <- read_csv(here("journal-article/assets/federal-districts.csv"))
 org_forms <- c(
   "общество с ограниченной ответственностью",
   "общество с дополнительной ответственностью",
@@ -29,9 +29,9 @@ org_forms <- c(
   "полное товарищество",
   "товарищество на вере"
 )
-regions <- st_read(here("common", "regions.geojson")) %>% 
+regions <- st_read(here("common/regions.geojson")) %>% 
   left_join(
-    read_csv(here("common", "regions.csv")),
+    read_csv(here("journal-article/assets/regions.csv")),
     by = c("shapeISO" = "iso_code")
   ) %>% 
   arrange(code) %>% 
@@ -88,7 +88,7 @@ if (!get0("IS_PAPER", ifnotfound = FALSE)) {
 }
 
 # Load vectors obtained from YandexGPT API
-vectors <- read_csv(here("common", "names-vectors.csv"))
+vectors <- read_csv(here("common/names-vectors.csv"))
 
 # Cluster the names using vectors
 ## Find optimal number of clusters
@@ -122,7 +122,7 @@ if (!get0("IS_PAPER", ifnotfound = FALSE)) {
 }
 
 # Load the results of manual analysis and join them with names
-cluster_labels <- read_excel(here("journal-article/cluster-labels.xlsx"))
+cluster_labels <- read_excel(here("journal-article/assets/cluster-labels.xlsx"))
 firms_lifetime <- count(firms, tin, name = "lifetime")
 clustered <- clustered_names %>%
   left_join(cluster_labels) %>%
@@ -417,89 +417,3 @@ localmoran_maps <- localmoran_res %>%
     strip.text = element_text(hjust = 0)
   ) +
[…]

### 2025-02-10 law-firms-names/conference-paper/Conference-slides.qmd
Use parquet to store large datasets instead of csv, add and commit large datasets

--- 2024-09-05 law-firms-names/conference-paper/Conference-slides.qmd
@@
 library(knitr)
+library(nanoparquet)
 library(readr)
@@
 # load the data
-data <- read_csv(here("common", "panel.csv"))
+data <- read_parquet(here("../datasets/law-firms/panel.parquet"))
 cluster_labels <- read_csv(
-  here("conference-paper", "cluster-labels-conf.csv"))
-vectors <- read_csv(here("common", "names-vectors.csv"))
-ru <- st_read(here("common", "ru.geojson"))
+  here("conference-paper/assets/cluster-labels-conf.csv"))
+vectors <- read_parquet(here("common/names-vectors.parquet"))
+ru <- st_read(here("common/regions.geojson"))
 

### 2025-02-10 fts-open-data-cli-app/helpers/Script.R

@@ -1,18 +1,16 @@
 library(dplyr)
 library(ggplot2)
 library(here)
-library(readr)
+library(nanoparquet)
 library(sf)
 library(tidyr)
 
-data_path <- here("../large-datasets/ncea-group-a-firms/panel.csv")
-data <- read_csv(data_path)
+data_full <- read_parquet(here("../datasets/ncea-group-a/smb-full-compact.parquet"))
+data <- read_parquet(here("../datasets/ncea-group-a/smb-orgs-2021-panel.parquet"))
 
-total <- data %>% distinct(tin) %>% nrow()
+total <- data_full %>% distinct(tin) %>% nrow()
 
-sample_row <- data %>%
-  filter(year == 2021, kind == 1) %>%
-  slice(1000:1001)
+sample_row <- data %>% slice(1000:1001)
 sample_table <- data.frame(
   var = colnames(sample_row),
   value = as.character(sample_row[1, ])
@@ -22,10 +20,7 @@ regions_boundaries <- st_read(here("assets/ru.geojson"))
 ru_crs <- st_crs("+proj=aea +lat_0=0 +lon_0=100 +lat_1=68 +lat_2=44 +x_0=0 +y_0=0 +ellps=krass +towgs84=28,-130,-95,0,0,0,0 +units=m +no_defs")
 
 plot_1 <- data %>%
-  filter(
-    year == 2021,
-    kind == 1,
-    substr(activity_code_main, 1, 2) == "01") %>%
+  filter(substr(activity_code_main, 1, 2) == "01") %>%
   mutate(activity = case_when(
     substr(activity_code_main, 1, 4) %in% c("01.1", "01.2", "01.3") ~ "Растениеводство",
     substr(activity_code_main, 1, 4) == "01.4" ~ "Животноводство")) %>%
@@ -47,4 +42,3 @@ plot_1 <- data %>%
   theme_bw(base_size = 12, base_family = "Times New Roman") +
   theme(legend.position = "bottom", legend.direction = "horizontal")
 plot_1
-

### 2025-02-10 fts-open-data-in-economic-geography/helpers/Plots.R

@@ -2,6 +2,7 @@ library(dplyr)
 library(forcats)
 library(ggplot2)
 library(here)
+library(nanoparquet)
 library(readr)
 library(scales)
 library(sf)
@@ -50,8 +51,7 @@ labels <- data.frame(
 labels <- labels[["ru"]]
 
 # Data
-rsmp_data_path <- here("../large-datasets/ncea-group-a-firms/smb.csv")
-rsmp_data <- read_csv(rsmp_data_path)
+rsmp_data <- read_parquet(here("../datasets/ncea-group-a/smb-full-compact.parquet"))
 
 # Maps
 regions_boundaries <- st_read(here("assets/ru.geojson"))

### 2025-02-10 fts-open-data-in-economic-geography/helpers/Stats.R

@@ -1,6 +1,7 @@
 library(dplyr)
 library(ggplot2)
 library(here)
+library(nanoparquet)
 library(readr)
 library(stringr)
 library(tidyr)
@@ -32,7 +33,7 @@ labels <- data.frame(
 labels <- labels[[LOCALE]]
 
 # Counts
-rsmp_data <- read_csv(here("../large-datasets/ncea-group-a-firms/smb.csv"))
+rsmp_data <- read_parquet(here("../datasets/ncea-group-a/smb-full-compact.parquet"))
 counts <- rsmp_data %>% 
   mutate(kind = replace(kind, kind == 3, 2)) %>% 
   distinct(tin, .keep_all = TRUE) %>% 

### 2025-02-10 law-firms-geography/helpers/Script.R

@@ -1,12 +1,13 @@
 library(dplyr)
 library(ggplot2)
 library(here)
+library(nanoparquet)
 library(readr)
 library(tidyr)
 
 # Load raw data
 courts <- read_csv(here("assets/courts.csv"))
-data <- read_csv(here("../large-datasets/law-firms/panel.csv"))
+data <- read_parquet(here("../datasets/law-firms/panel.parquet"))
 empl <- read_csv(here("assets/empl.csv"))
 pop <- read_csv(here("assets/population.csv"))
 tiles <- read_csv(here("assets/russia-tiles.csv"))

### 2025-02-10 law-firms-mapping/helpers/Plots.R

@@ -2,22 +2,23 @@ library(dplyr)
 library(forcats)
 library(ggplot2)
 library(here)
+library(nanoparquet)
 library(readr)
 library(scales)
 library(sf)
 library(tidyr)
 
 # Data
-smb_data <- read_csv(here("../large-datasets/law-firms/smb.csv"))
-revexp_data <- read_csv(here("../large-datasets/law-firms/revexp.csv"))
-empl_data <- read_csv(here("../large-datasets/law-firms/empl.csv"))
+smb_data <- read_parquet(here("../datasets/law-firms/smb.parquet"))
+revexp_data <- read_parquet(here("../datasets/law-firms/revexp.parquet"))
+empl_data <- read_parquet(here("../datasets/law-firms/empl.parquet"))
 
 # Maps
 regions_boundaries <- st_read(here("assets/ru.geojson"))
 ru_crs <- st_crs("+proj=aea +lat_0=0 +lon_0=100 +lat_1=68 +lat_2=44 +x_0=0 +y_0=0 +ellps=krass +towgs84=28,-130,-95,0,0,0,0 +units=m +no_defs")
 regions <- read_csv(here("assets/regions.csv"))
-regions <- regions_boundaries %>% 
-  left_join(regions, by = c("shapeISO" = "iso_code")) %>% 
+regions <- regions_boundaries %>%
+  left_join(regions, by = c("shapeISO" = "iso_code")) %>%
   select(name, name_en = shapeName)
 ru_svr <- st_read(here("assets/ru_svr.geojson"))
 
@@ -25,10 +26,10 @@ ru_svr <- st_read(here("assets/ru_svr.geojson"))
 regions_map <- smb_data %>%
   filter(
     start_date <= "2021-12-31",
-    end_date >= "2021-12-31") %>% 
-  count(region) %>% 
-  left_join(regions, by = c("region" = "name")) %>% 
-  st_as_sf() %>% 
+    end_date >= "2021-12-31") %>%
+  count(region) %>%
+  left_join(regions, by = c("region" = "name")) %>%
+  st_as_sf() %>%
   ggplot() +
   geom_sf(aes(fill = n), size = .1) +
   coord_sf(crs = ru_crs) +
@@ -39,7 +40,7 @@ regions_map <- smb_data %>%
   theme_void(base_size = 11, base_family = "Times New Roman") +
   theme(
     legend.title.position = "top",
-    legend.position = c(.05, 0), 
+    legend.position = c(.05, 0),
     legend.justification = c(0, 0),
[…]

### 2025-02-10 law-firms-names/journal-article/helpers/Script.R

@@ -4,6 +4,7 @@ library(here)
 library(igraph)
 library(ggraph)
 library(knitr)
+library(nanoparquet)
 library(readr)
 library(readxl)
 library(sf)
@@ -14,7 +15,7 @@ library(tidyr)
 sf_use_s2(FALSE) # disable to avoid errors in distances and intersections
 
 # Load the data
-data <- read_csv(here("../large-datasets/law-firms/panel.csv"))
+data <- read_parquet(here("../datasets/law-firms/panel.parquet"))
 er <- read_csv(here("journal-article/assets/economic-regions.csv"))
 fd <- read_csv(here("journal-article/assets/federal-districts.csv"))
 org_forms <- c(
@@ -88,7 +89,7 @@ if (!get0("IS_PAPER", ifnotfound = FALSE)) {
 }
 
 # Load vectors obtained from YandexGPT API
-vectors <- read_csv(here("common/names-vectors.csv"))
+vectors <- read_parquet(here("common/names-vectors.parquet"))
 
 # Cluster the names using vectors
 ## Find optimal number of clusters

### 2025-02-10 legal-companies-migration/conference-paper/helpers/Script.R

@@ -2,15 +2,16 @@ library(dplyr)
 library(ggplot2)
 library(glue)
 library(here)
+library(nanoparquet)
 library(readr)
 library(tidyr)
 library(sf)
 library(units)
 
 # Load the data
-data <- read_csv(here("../large-datasets/law-firms/smb.csv"))
-revexp <- read_csv(here("../large-datasets/law-firms/revexp.csv"))
-empl <- read_csv(here("../large-datasets/law-firms/empl.csv"))
+data <- read_parquet(here("../datasets/law-firms/smb.parquet"))
+revexp <- read_parquet(here("../datasets/law-firms/revexp.parquet"))
+empl <- read_parquet(here("../datasets/law-firms/empl.parquet"))
 cities <- read_csv(here("common/cities.csv"))
 regions_geo <- st_read(here("common/ru.geojson"))
 regions <- read_csv(here("common/regions.csv"))
@@ -24,33 +25,33 @@ companies_count_08_2016 <- nrow(filter(data, kind == 1, start_date == "2016-08-1
 # Calculate the absolute numbers and shares of companies
 # with changes in location
 # Location is defined as change either in region or settlement
-migration_count_distribution <- data %>% 
-  filter(kind == 1) %>% 
+migration_count_distribution <- data %>%
+  filter(kind == 1) %>%
   group_by(tin) %>%
-  distinct(region, settlement) %>% 
-  count() %>% 
-  ungroup() %>% 
-  count(n, name = "count") %>% 
+  distinct(region, settlement) %>%
+  count() %>%
+  ungroup() %>%
+  count(n, name = "count") %>%
   mutate(
     n = n - 1,
     share = 100 * count / sum(count))
 
 # Draw it on plot
-mig_count_distr_plot <- migration_count_distribution %>% 
+mig_count_distr_plot <- migration_count_distribution %>%
   ggplot(aes(x = n, y = count)) +
   geom_col() +
   geom_text(
     aes(y = 1e6, label = scales::number(share, accuracy = 0.01, suffix = " %")),
     color = "gray30", size = 3
-    ) + 
-  geom_text(aes(label = count), vjust = -.5) + 
+    ) +
+  geom_text(aes(label = count), vjust = -.5) +
   scale_y_log10(limits = c(NA, 1e6), breaks = 10^(1:5), labels = ~ .x) +
   labs(
-    x = "Number of migrations by a company", 
[…]

### 2025-02-10 legal-companies-migration/journal-article/helpers/Script.R

@@ -2,14 +2,15 @@ library(dplyr)
 library(ggplot2)
 library(glue)
 library(here)
+library(nanoparquet)
 library(readr)
 library(tidyr)
 library(sf)
 
 # Load the data
-data <- read_csv(here("../large-datasets/law-firms/smb.csv"))
-revexp <- read_csv(here("../large-datasets/law-firms/revexp.csv"))
-empl <- read_csv(here("../large-datasets/law-firms/empl.csv"))
+data <- read_parquet(here("../datasets/law-firms/smb.parquet"))
+revexp <- read_parquet(here("../datasets/law-firms/revexp.parquet"))
+empl <- read_parquet(here("../datasets/law-firms/empl.parquet"))
 cities <- read_csv(here("common/cities.csv"))
 regions_geo <- st_read(here("common/ru.geojson"))
 regions <- read_csv(here("common/regions.csv"))
@@ -24,33 +25,33 @@ companies_count_08_2016 <- nrow(filter(data, kind == 1, start_date == "2016-08-1
 # Calculate the absolute numbers and shares of companies
 # with changes in location
 # Location is defined as change either in region or settlement
-migration_count_distribution <- data %>% 
-  filter(kind == 1) %>% 
+migration_count_distribution <- data %>%
+  filter(kind == 1) %>%
   group_by(tin) %>%
-  distinct(region, settlement) %>% 
-  count() %>% 
-  ungroup() %>% 
-  count(n, name = "count") %>% 
+  distinct(region, settlement) %>%
+  count() %>%
+  ungroup() %>%
+  count(n, name = "count") %>%
   mutate(
     n = n - 1,
     share = 100 * count / sum(count))
 
 # Draw it on plot
-mig_count_distr_plot <- migration_count_distribution %>% 
+mig_count_distr_plot <- migration_count_distribution %>%
   ggplot(aes(x = n, y = count)) +
   geom_col() +
   geom_text(
     aes(y = 1e6, label = scales::number(share, accuracy = 0.01, suffix = " %")),
     color = "gray30", size = 3
-    ) + 
-  geom_text(aes(label = count), vjust = -.5) + 
+    ) +
+  geom_text(aes(label = count), vjust = -.5) +
   scale_y_log10(limits = c(NA, 1e6), breaks = 10^(1:5), labels = ~ .x) +
   labs(
-    x = "Number of migrations by a company", 
+    x = "Number of migrations by a company",
[…]

### 2025-02-10 legal-companies-regional-specialization/helpers/Script.R

@@ -1,40 +1,41 @@
 library(dplyr)
 library(ggplot2)
 library(here)
+library(nanoparquet)
 library(readr)
 library(tidyr)
 
-panel <- read_csv(here("../large-datasets/law-firms/panel.csv"))
+panel <- read_parquet(here("../datasets/law-firms/panel.parquet"))
 empl <- read_csv(here("assets/empl.csv"))
 tiles <- read_csv(here("assets/russia-tiles.csv"))
 
-spec <- panel %>% 
-  filter(year == 2021) %>% 
-  drop_na(region) %>% 
-  replace_na(list(employees_count = 0)) %>% 
-  mutate(employees_count = replace(employees_count, employees_count == 0, 1)) %>% 
-  count(region, wt = employees_count, name = "legal_empl") %>% 
-  right_join(empl, by = "region") %>% 
-  rename(total_empl = count) %>% 
+spec <- panel %>%
+  filter(year == 2021) %>%
+  drop_na(region) %>%
+  replace_na(list(employees_count = 0)) %>%
+  mutate(employees_count = replace(employees_count, employees_count == 0, 1)) %>%
+  count(region, wt = employees_count, name = "legal_empl") %>%
+  right_join(empl, by = "region") %>%
+  rename(total_empl = count) %>%
   mutate(
     concentration = legal_empl / sum(legal_empl, na.rm = TRUE),
     empl_share = total_empl / sum(total_empl, na.rm = TRUE),
     localization = concentration / empl_share
-  ) %>% 
-  arrange(-concentration) %>% 
-  mutate(cum_concentraion = cumsum(concentration)) %>% 
-  arrange(-localization) %>% 
-  mutate(loc_rank = row_number()) %>% 
+  ) %>%
+  arrange(-concentration) %>%
+  mutate(cum_concentraion = cumsum(concentration)) %>%
+  arrange(-localization) %>%
+  mutate(loc_rank = row_number()) %>%
   mutate(
     is_national = cum_concentraion <= 0.8,
     is_regional = loc_rank < nrow(.) * .2,
     rank = as.numeric(is_national) * 2 + as.numeric(is_regional),
     rank = factor(rank, labels = c(NA, "Local", "National", "Both"))
-  ) %>% 
+  ) %>%
   select(region, is_national, is_regional, rank)
 
 
-map <- left_join(spec, tiles, by = c("region" = "name")) %>% 
[…]

### 2025-02-10 legal-companies-zipf-law/helpers/Script.R

@@ -1,42 +1,43 @@
 library(dplyr)
 library(ggplot2)
 library(here)
+library(nanoparquet)
 library(readr)
 library(sf)
 library(tidyr)
 
 # Data on firms, and cities lookup table
-panel <- read_csv(here("../large-datasets/law-firms/panel.csv"))
+panel <- read_parquet(here("../datasets/law-firms/panel.parquet"))
 cities <- rbind(
   read_csv(here("assets/cities.csv")),
   read_csv(here("assets/cities_additional.csv"))
 )
 
-lcc <- panel %>% 
+lcc <- panel %>%
   filter(
-    kind == 1, 
-    year == 2021, 
+    kind == 1,
+    year == 2021,
     settlement_type == "г",
     revenue > 0,
     expenditure > 0
-  ) %>% 
+  ) %>%
   mutate(city = case_when(
     region == "Москва" & is.na(settlement) ~ "Москва",
     region == "Санкт-Петербург" & is.na(settlement) ~ "Санкт-Петербург",
     TRUE ~ settlement
-  )) %>% 
-  select(region, city, oktmo, empl = employees_count) %>% 
-  replace_na(list(empl = 0)) %>% 
-  mutate(empl = replace(empl, empl == 0, 1)) %>% 
-  group_by(region, city, oktmo) %>% 
+  )) %>%
+  select(region, city, oktmo, empl = employees_count) %>%
+  replace_na(list(empl = 0)) %>%
+  mutate(empl = replace(empl, empl == 0, 1)) %>%
+  group_by(region, city, oktmo) %>%
   summarise(count = n(), empl = sum(empl), .groups = "drop")
 
-cp <- cities %>% 
-  mutate(city = coalesce(cities$city, cities$area, cities$region)) %>% 
+cp <- cities %>%
+  mutate(city = coalesce(cities$city, cities$area, cities$region)) %>%
   select(city, oktmo, population, lat = geo_lat, lon = geo_lon)
 
-data <- left_join(lcc, cp, by = c("oktmo", "city")) %>% 
-  drop_na(count, empl, population) %>% 
+data <- left_join(lcc, cp, by = c("oktmo", "city")) %>%
+  drop_na(count, empl, population) %>%
   mutate(
     count_per_100k = 1e5 * count / population,
     empl_per_100k = 1e5 * empl / population,
[…]

### 2025-02-10 legal-services-provision/conference-paper/helpers/Script.R

@@ -2,43 +2,44 @@ library(dplyr)
 library(ggplot2)
 library(glue)
 library(here)
+library(nanoparquet)
 library(readr)
 library(sf)
 library(tidyr)
 
 # Data on firms, and cities lookup table
-panel <- read_csv(here("../large-datasets/law-firms/panel.csv"))
+panel <- read_parquet(here("../datasets/law-firms/panel.parquet"))
 cities <- rbind(
   read_csv(here("common", "cities.csv")),
   read_csv(here("common", "cities_additional.csv"))
 )
 regions_geo <- st_read(here("common", "ru.geojson"))
 
-lcc <- panel %>% 
+lcc <- panel %>%
   filter(
-    kind == 1, 
-    year == 2021, 
+    kind == 1,
+    year == 2021,
     settlement_type == "г",
     revenue > 0,
     expenditure > 0
-  ) %>% 
+  ) %>%
   mutate(city = case_when(
     region == "Москва" & is.na(settlement) ~ "Москва",
     region == "Санкт-Петербург" & is.na(settlement) ~ "Санкт-Петербург",
     TRUE ~ settlement
-  )) %>% 
-  select(region, city, oktmo, empl = employees_count) %>% 
-  replace_na(list(empl = 0)) %>% 
-  mutate(empl = replace(empl, empl == 0, 1)) %>% 
-  group_by(region, city, oktmo) %>% 
+  )) %>%
+  select(region, city, oktmo, empl = employees_count) %>%
+  replace_na(list(empl = 0)) %>%
+  mutate(empl = replace(empl, empl == 0, 1)) %>%
+  group_by(region, city, oktmo) %>%
   summarise(count = n(), empl = sum(empl), .groups = "drop")
 
-cp <- cities %>% 
-  mutate(city = coalesce(cities$city, cities$area, cities$region)) %>% 
+cp <- cities %>%
+  mutate(city = coalesce(cities$city, cities$area, cities$region)) %>%
   select(city, oktmo, population, lat = geo_lat, lon = geo_lon)
 
-data <- left_join(lcc, cp, by = c("oktmo", "city")) %>% 
-  drop_na(count, empl, population) %>% 
+data <- left_join(lcc, cp, by = c("oktmo", "city")) %>%
+  drop_na(count, empl, population) %>%
   mutate(
     count_per_100k = 1e5 * count / population,
[…]

### 2025-02-10 legal-services-provision/journal-article/helpers/Script.R

@@ -3,45 +3,46 @@ library(dplyr)
 library(ggplot2)
 library(glue)
 library(here)
+library(nanoparquet)
 library(readr)
 library(sf)
 library(tibble)
 library(tidyr)
 
 # Load data
-panel <- read_csv(here("../large-datasets/law-firms/panel.csv"))
+panel <- read_parquet(here("../datasets/law-firms/panel.parquet"))
 cities <- rbind(
   read_csv(here("common/cities.csv")),
   read_csv(here("common/cities_additional.csv"))
 )
 economic_regions <- read_csv(here("journal-article/assets/economic-regions.csv"))
-regions_geo <- st_read(here("common/ru.geojson")) %>% 
+regions_geo <- st_read(here("common/ru.geojson")) %>%
   left_join(economic_regions, by = c("shapeISO" = "iso_code"))
 
 # Build table with number of employees by city
 # lfc means Law Firms by City
-lfc <- panel %>% 
+lfc <- panel %>%
   filter(
-    kind == 1, 
-    year == 2021, 
+    kind == 1,
+    year == 2021,
     settlement_type == "г",
     revenue > 0,
     expenditure > 0
-  ) %>% 
+  ) %>%
   mutate(city = case_when(
     region == "Москва" & is.na(settlement) ~ "Москва",
     region == "Санкт-Петербург" & is.na(settlement) ~ "Санкт-Петербург",
     TRUE ~ settlement
-  )) %>% 
-  select(region, city, oktmo, empl = employees_count) %>% 
-  replace_na(list(empl = 0)) %>% 
-  mutate(empl = replace(empl, empl == 0, 1)) %>% 
-  group_by(region, city, oktmo) %>% 
+  )) %>%
+  select(region, city, oktmo, empl = employees_count) %>%
+  replace_na(list(empl = 0)) %>%
+  mutate(empl = replace(empl, empl == 0, 1)) %>%
+  group_by(region, city, oktmo) %>%
   summarise(count = n(), empl = sum(empl), .groups = "drop")
 
 # Build table with cities population
 # cp means Cities Population
-cp <- cities %>% 
+cp <- cities %>%
   mutate(
     city = coalesce(cities$city, cities$area, cities$region),
     is_regional_center = case_when(
@@ -49,14 +50,14 @@ cp <- cities %>%
[…]

### 2025-02-14 legal-companies-migration/conference-paper/Paper.Rmd
Rename some files for consistency

--- 2024-02-13 bel-geo-congress-2024/Conference-paper.Rmd
@@
   word_document:
-    reference_docx: conference-paper-reference.docx
-bibliography: refs.bib
-csl: biblio-style.csl
+    reference_docx: assets/conference-paper-reference.docx
+bibliography: assets/refs.bib
+csl: assets/biblio-style.csl
 ---
@@
 
-[^3]: https://github.com/PavelSyomin/tax-service-opendata-paper/tree/main/bel-geo-congress-2024
+[^3]: https://github.com/PavelSyomin/ru-smb-companies-papers/tree/main/bel-geo-congress-2024
 

### 2025-02-28 law-firms-names/journal-article/Paper.Rmd
Minor adjustments before publication, #4

@@
 IS_PAPER <- TRUE # used in script to disable saving of names and samples
+
+# Space as thousands separator
+knit_print.integer <- function(x, ...) {
+  prettyNum(x, big.mark=" ")
+}
+
+registerS3method(
+  "knit_print", "integer", knit_print.integer,
+  envir = asNamespace("knitr")
+)
 ```
@@
 library(dplyr)
-library(ggplot2)
-library(glue)
-library(here)
 library(knitr)
-library(readr)
-library(readxl)
-library(sf)
-library(stringr)
-library(tidyr)
-library(units)
 ```
@@
 The market is divided between ordinary lawyers and advocates (members of the Bar), but there is no bar monopoly.
-There are ~90, 000 law firms in Russia, including ~40, 000 commercial law companies, ~30, 000 sole practitioners, ~20, 000 bar (advocate) offices, and ~10, 000 bar (advocate) organisations.
+There are ~90 000 law firms in Russia, including ~40 000 commercial law companies, ~30 000 sole practitioners, ~20 000 bar (advocate) offices, and ~10 000 bar (advocate) organisations.
 There are no geographical restrictions on legal practitioners, so that a law firm registered in one region of the country can provide services in any other region.
@@
 The issue of obtaining information about firm names is a challenging one.
-All Russian organizations are registered in the state registry operated by the Federal Tax Service (FTS of Russia).
+All Russian organisations are registered in the state registry operated by the Federal Tax Service (FTS of Russia).
 An extract from this registry about a particular company can be searched and downloaded online by any person free of charge.
@@
 Additionally, the portal provides a monthly record of updates made since the registry's inception.
-To sum it up, the SMB registry can be regarded as the most comprehensive free data source offering information on Russian business entities, both organisations and individuals.
+To sum it up, the SME registry can be regarded as the most comprehensive free data source offering information on Russian business entities, both organisations and individuals.
 
@@
 For a validation, a list of law firms' unique tax IDs obtained from the SME registry for the year 2023 is matched with the list of law firms' tax IDs from the Russian Financial Statements Database (RFSD) [@Bondarkov2025] for the same year.
-The comparison shows that ~3,300 of the firms present in the SME registry are missing from the RFSD, and ~5,200 firms present in the RFSD are missing from the SME registry.
+The comparison shows that ~3 300 of the firms present in the SME registry are missing from the RFSD, and ~5 200 firms present in the RFSD are missing from the SME registry.
 The balance of missing records is not so far from being equal.
-The SME registry contains ~34000 law firms for the year 2023, so the missing rate is below 15%.
+The SME registry contains ~34 000 law firms for the year 2023, so the missing rate is below 15%.
 Consequently, the completeness of the data is relatively high and reliable conclusions can be drawn.
@@
 
-In this study, the term “law firm” is defined as a commercial organization registered in Russia in accordance with the Civil Code of Russia, with a NACE Rev.
+In this study, the term “law firm” is defined as a commercial organisation registered in Russia in accordance with the Civil Code of Russia, with a NACE Rev.
 2-compatible main activity code equal to 69.10 (activity in the area of law).
@@
 The “text-search-query” model is selected because its purpose, as stated in the documentation, is to vectorize short texts.
-Prior to vectorization, the firm names are preprocessed to remove indications of organizational form such as “limited liability company”.
+Prior to vectorization, the firm names are preprocessed to remove indications of organisational form such as “limited liability company”.
 This is done because such indications are common to the majority of names and thus have little value.
@@
 A random sample of 20 names is selected from each cluster.
-With 50 clusters and 20 sampled names allocated in each, the total size of the dataset for manual processing is 1, 000.
+With 50 clusters and 20 sampled names allocated in each, the total size of the dataset for manual processing is 1 000.
 The cluster samples are carefully examined, and a concise description is provided for each, e.g.
@@
 Regions (constituent entities) of the Russian Federation are chosen as the unit of observation.
-Municipal territories, the next level of spatial division, are not used, because an exploratory data analysis has shown that the lion's share of such the spatial units do not have any law firms.
+Municipal territories, the next level of spatial division, are not used, because the exploratory data analysis has shown that the lion's share of such the spatial units do not have any law firms.
 
@@
 Two remaining matrices treat regions belonging to the same upper-level macro-region as neighbors, and all other regions as non-neighbors.
-One of these matrices uses economic regions act as macro-regions, and the other uses federal districts.
+One of these matrices uses economic regions as macro-regions, and the other uses federal districts.
 Economic regions are defined according to the classification that has been widely used by economic geographers and spatial development practitioners in Russia for decades, as outlined, for example, in [@Leyzerovich2010].
@@
 
-## 3.1. Law firms naming patterns and strategies
+## 3.1. Law firm naming patterns and strategies
 
@@
 1 for a dendrogram and Table 1 for a sample of 10 clusters, their descriptions, patterns, and strategies).
-The most popular is the no-strategy followed by ~18, 700 law firms.
-Slightly less important is the belonging strategy with ~15, 000 companies.
-The uniqueness strategy (~8, 300 law firms) is relatively rare, and the least used is the market strategy (~5, 300 companies).
+The most popular is the no-strategy followed by ~18 700 law firms.
+Slightly less important is the belonging strategy with ~15 000 companies.
+The uniqueness strategy (~7 500 law firms) is relatively rare, and the least used is the market strategy (~6 100 companies).
 
@@
     col.names = c(
-      "# of cluster", "# of names", "Description",
+      "# of cluster", "# of firms", "Description",
       "Pattern", "Strategy"
     ),
-    caption = "**Table 1.** Characteristics of clusters of law firm names"
+    caption = "**Table 1.** Characteristics of 10 selected clusters of law firm names"
   )
-write_csv(clusters_info_table, "ESM2.csv")
 ```
@@
 Market strategy, according to the original definition, is used by firms “having a name directly appealing to markets” [@BassolsiGardella2023].
-In the case of law firms, this can be achieved by mentioning target clients (“Business” pattern) or by using words with positive meanings that attract customers (“Positive”, “Plus”, “Legal aid”, “Guarantor” patterns).
+In the case of law firms, this can be achieved by mentioning target clients (“Business” pattern) or by using words with positive meaning that attract customers (“Positive”, “Plus”, “Legal aid”, “Guarantor” patterns).
 The “Expert” pattern is complex, as this word can be used to indicate either a high level of service (“expert” as a synonym for “professional”) or forensic services (“expert” as a forensic scientist).
@@
 
-```{r fig-3-localmoran-maps, fig.cap="**Fig. 3.** Maps of hotspots in the spatial distribution of three law firms naming patterns: (a) Expert (b) Personal brand (c) Siberia", fig.height=10}
+```{r fig-3-localmoran-maps, fig.cap="**Fig. 3.** Maps of hotspots in the spatial distribution of three law firm naming patterns: (a) Expert (b) Personal brand (c) Siberia", fig.height=10}
 localmoran_maps
@@
 The findings of the study are somewhat contradictory.
-On the one hand, there are 22 law firm naming patterns in Russia that can be attributed to one of the four broad SMN strategies: belonging or uniqueness, market or no-strategy.
+On the one hand, there are 22 law firm naming patterns in Russia that can be attributed to one of the four broad SNM strategies: belonging or uniqueness, market or no-strategy.
 Apart from the no-strategy (~40% of firms), belonging is the most popular strategy used by 30% of law firms, followed by uniqueness (~20%) and market (~10%).

### 2025-02-28 law-firms-names/journal-article/helpers/Script.R

@@ -10,6 +10,7 @@ library(readxl)
 library(sf)
 library(spdep)
 library(stopwords)
+library(stringr)
 library(tidyr)
 
 sf_use_s2(FALSE) # disable to avoid errors in distances and intersections
@@ -325,7 +326,7 @@ moran_test_global_res_plot <- moran_test_global_res %>%
   ) +
   coord_fixed() +
   labs(
-    x = "Law firm names category",
+    x = "Law firm naming pattern",
     y = "Type of weights matrix"
   ) +
   theme_minimal()
